---
title: "The synpmx AVATAR Algorithm"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The synpmx AVATAR Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
library(synpmx)
```

This article is the full specification of the AVATAR synthetic data generator applied
to pharmacometrics modeling datasets.  The original method [1] was extended to population
pharmacokinetic (PK) datasets in [2], and here further extensions are provided
so that it works well with more realistic datasets.

AVATAR makes no formal privacy claim. It also does not preserve the dose-exposure
relationship.  Use the synthetic datasets to develop and debug analysis code. Do not use
it to estimate parameters.

Because modeling datasets are generally longitudinal, one set of identifying features are
the unique dose or observation times each patient will have.  The default setting of 
`synpmx_avatar` ensure that the set of observation and dose times of the synthetic patient
does not match any one real patient data.   

## Overview of Algorithm

Each synthetic patient is called an avatar.  Each avatar is built from one real 
**anchor** subject's dosing event structure, then filled with measurements blended from several similar real subjects, called **donors**. The generator works on any number of endpoints, so one study can
carry multiple PK profiles alongside multiple pharmacodynamic (PD) measures.

Every patient's data can be thought of as containing two pieces of information which are treated differently

* **event skeleton** is *when* the dose events and the DV observations
happen. 
* **values** are the DV numbers filling those observation slots, plus
the baseline covariates. 

The event skeleton is copied from one anchor patient with some
masking of the times, and the values are blended across several donors. 
The `synpmx_avatar` function runs the following steps.

1. **Declare the columns.** The `roles` variable say which column is the id, time,
   dose, measurement, etc. Only named columns survive.
2. **Coarsen and align the times.** Snap every recorded time onto a shared visit
   grid (often based on nominal time), and align each subject to its own first dose.
3. **Characterize each endpoint**: whether it is discrete, and whether it is
   log transformed. Both are applied later.
4. **Build one numeric profile per source subject**, and reduce it to a handful
   of principal-component coordinates that measure similarity between patients.
5. **Read each subject's route of administration and dosing history**, the two
   facts that decide who may donate to whom.
6. **Restrict the anchor pool and sample anchors.** A patient whose structure
   would give them away, an unusually long follow-up or a schedule nobody else
   shares, is not built upon, though their measurements still contribute as a
   donor. Each cohort (arm) keeps the same size (except for very small cohorts).
7. **Copy the anchor's event skeleton** and redraw the parts of it that identify
   the anchor: which visits carry an observation, and where dosing stopped.
8. **Choose the donors**, nearest first within the anchor's route.
9. **Randomize and cap the donor weights**, so no one real patient is more than
   half of any avatar.
10. **Synthesize the baseline covariates** from those donors.
11. **Synthesize each endpoint trajectory**, then correct the generated value:
    snap a discrete endpoint back onto its levels, and reapply the censoring
    boundary.
12. **Recompute the dose** from the avatar's own blended covariate, where dosing
    is proportional to a covariate (e.g. weight based dosing).
13. **Construct the full PMX dataset**, with new synthetic subject IDs and every
    column back in the shape the source had, and validate it.

## The Masking Mechanisms

Across the steps above are six **masking mechanisms** of the individual patient data.

- **M1. Coarsen the times.** Snap every time onto a shared visit grid (nominal time when available), then add deviations. This prevents identifying a patient from their TIME vector. *Step 2.*
- **M2. Restrict who may be an anchor.** Drop patients whose structure no amount
  of blending would hide, such as a patient followed twice as long as everyone
  else, and patients in a route arm too small to blend within. *Step 6.*
- **M3. Redraw which visits the avatar attended.** Reselect which DV values were
  collected at which time points, in case a patient has a unique set of missed
  visits. *Step 7.*
- **M4. Redraw where dosing stopped.** Where a patient's dose schedule is one
  nobody else has, the avatar stops earlier, at the latest dosing time that several
  patients share. Dose *times* are never moved or invented, so this only ever
  truncates a real schedule earlier. *Step 7.*
- **M5. Blend the values across several donors.** Covariates and DV, capped so
  no one donor dominates, plus noise. *Step 9.*
- **M6. Recompute the dose from the avatar's own covariate.** Done when dosing
  is weight-based or body-surface-area based. Declared with `dose_covariate`, or inferred.
  *Step 12.*

## Step 1: Declare the Meaning of the Columns

The package never guesses critical pharmacometric (PMX) roles from column names.
The user must declare them with `pmx_roles()`:

```{r roles-example}
roles <- pmx_roles(
  id = "ID",
  time = "TIME",
  dv = "DV",
  amt = "AMT",
  evid = "EVID",
  cmt = "CMT",
  dvid = "DVID",
  mdv = "MDV",
  rate = "RATE",
  nominal_time = "NTIME",  # protocol-scheduled time, if the study records it
  cens = "CENS",           # below-limit indicator, if the study has BLQ data
  limit = "LIMIT",         # the other end of an interval, when used
  covariates = c("WT", "AGE", "SEX"),
  dose_covariate = "WT",   # dose is mg/kg; say so rather than let it be inferred
  strata = "ARM",          # assigned arm; groups three mechanisms, see below
  keep = "STUDYID"         # carried verbatim, see the allowlist below
)
```

`id`, `time`, `dv`, and `evid` are required and `?pmx_roles` lists the rest. A
column generally holds one role, with two exceptions. NONMEM's `CMT` routinely
does both jobs, the dosing compartment on event rows and the endpoint key on
observation rows, so `cmt` and `dvid` may name the same column. And
`dose_covariate` must *also* be named in `covariates`, because the avatar's
amount is rebuilt from its own blended value of that column, so the column has
to be one that gets blended.

### The Role Declaration Specifies Which Columns Appear in the Synthetic Dataset

`synpmx_avatar()` keeps only columns a role names, and **drops every other
column**, reporting which ones are dropped. This is because AVATAR builds each synthetic subject
from one real anchor subject, and any column it retains that it does not
actively synthesize is copied from that anchor verbatim. A secondary identifier
such as `USUBJID`, a site number, or a randomization date, left undeclared,
would otherwise follow a real subject's real value straight into the synthetic
data. Dropping by default means a column you forget fails safe, toward removal,
rather than leaking.

That leaves five ways a column is treated:

| Declared as | What AVATAR does to it | Use for |
|---|---|---|
| `dv` (with `cens`/`limit`) | Blended across donors into a new trajectory | the measurement |
| `covariates` | Blended or resampled across donors into a new value | baselines you want *synthesized*: weight, age |
| `strata` | Copied verbatim from the anchor | assigned strata: treatment arm, dose group, cohort |
| `keep` | Copied verbatim from the anchor, and otherwise inert | assigned values you want kept faithful to that subject's dosing, such as a study identifier |
| *(undeclared)* | Dropped | anything you do not need |

**`strata` carries a column the same way `keep` does, and then uses it.** Both
copy the anchor's value verbatim, so the choice between them is not about what
appears in the output. It is about whether that column should also *partition
the cohort*, as will de desrcibed below.  But, `strata` is deliberately **not** 
a blending barrier. Donors are still borrowed across
strata.  Declare a treatment arm as `strata` rather than `keep` whenever the arms differ
in dose rule or visit schedule, and use `keep` for a label that carries no such
structure.

Redundant endpoint labels have their own handling. A dataset that carries both a
numeric `YTYPE` and a character `NAME` for the same endpoint declares **both**
as `dvid`: `dvid = c("YTYPE", "NAME")`. The first is the grouping key.
Validation checks that the rest are a consistent 1:1 mapping with it, catching a
source where the two labels disagree, which `keep` would carry through silently,
and AVATAR carries all of them into the output, aligned.

Immediately before generation, `synpmx_avatar()` calls `validate_pmx(...,
strict = TRUE)`. Among other checks, validation requires:

- nonmissing IDs and EVID values;
- finite numeric TIME values and nondecreasing TIME within subject;
- numeric DV and finite nonmissing observations;
- an endpoint value on every eligible observation when DVID is declared;
- multiple `dvid` columns to be a consistent 1:1 mapping;
- every declared baseline covariate to be constant within subject; and
- when `cens` is declared, each CENS flag to agree with its DV. A left-censored
  value at the limit cannot exceed an ordinary measurement of the same endpoint.

## Step 2: Coarsen the Times and Align to the First Event

### M1. Coarsening the Times, and Putting the Scatter Back

At the default `coarsen_time = TRUE`, source times are collapsed onto a shared
visit grid (`nominal_time` when provided, otherwise a grid is automaticlaly computed). 
This **snapping**, replaces each recorded time with its
visit's grid time. The deviation removed is not discarded. It goes into one pool
shared by the whole cohort.  **Restoring** occurs after each avatar is built, when
a deviation from that pool is drawn and added back to the time.  

### First-Event-Relative Time

Neighbor profiles and donor trajectories are constructed on an internal aligned
time. For subject $i$, the preferred origin is the first nonzero-EVID row having
a positive AMT:

$$
t_{0i} = \min\{t_{ij}: \mathrm{EVID}_{ij}\ne0,\ \mathrm{AMT}_{ij}>0\}.
$$

If no positive-AMT event exists, the first nonzero-EVID row is used. If the
subject has no event row, TIME is left unchanged. The internal aligned time is

$$
t^*_{ij}=t_{ij}-t_{0i}.
$$

This removes a subject-specific calendar or study-time offset. It does **not**
compute time after each dose.

A declared `tad` column is  recomputed in the output, as time since the avatar's own most recent dose.

## Step 3: Characterize Each Endpoint

Blending is a weighted mean of several donors, so before any blending happens
the endpoint must be categorized in two ways.

* What type is it: continuous, binary, ordinal, or integer.  
* Should the endpoint be log transformed (and offset)

Both categories are recorded on the output and used later.

### Endpoint types

- **`continuous`** when any recorded value is not a whole number. Nothing is
  constrained.
- **`binary`** when every value is a whole number and there are two distinct
  levels.
- **`ordinal`** when every value is a whole number and there are at most twelve
  distinct levels.
- **`integer`** when every value is a whole number and there are more than
  twelve.

The endpoint type is inferred from the data, but `pmx_roles(endpoint_types = )` overrides the inference.

### Log-transformation

Transformations are selected per DVID endpoint, or once when the DVID column
is absent. An endpoint is **positive-like** when it has at least one positive value, its median
is positive, and at most 1% of values are negative. The 1% tolerance lets a
concentration endpoint with a couple of small negative assay readings still
count.

PK and PD data can have exact zeros and $\log 0=-\infty$ would make blending impossible. So a small constant $c_e$ set to be half the smallest positive value of that endpoint, giving the transformation of:

$$
g_e(y)=\log\{\max(y,0)+c_e\},
\qquad
g_e^{-1}(z)=\max\{\exp(z)-c_e,0\}.
$$
Transformations and offsets are stored in
`attr(synthetic, "pmx_settings")$endpoint_transforms`.

## Step 4: Build One Numeric Profile per Source Subject for Distance Calculation

Source subject $i$'s profile concatenates their baseline covariates
$\mathbf{c}_i$ and their longitudinal observations $\mathbf{y}_{ie}(G_e)$ on each
of the $E$ endpoints, interpolated onto a common grid $G_e$ of at most fifteen
times, built per endpoint from the observation times pooled across subjects.  This
common grid is not the nominal times, but rather a grid that is later used for calculating
the principal component analysis and distance between subjects.

$$
\mathbf{x}_i=
\left[
  \mathbf{c}_i,\
  \mathbf{y}_{i1}(G_1),\ldots,\
  \mathbf{y}_{iE}(G_E)
\right].
$$

This $\mathbf{x_i}$ profile is used for distance calculations. It does not replace the
source data and is not returned as a synthetic patient.

### The Common Time Grid

For endpoint $e$, the package pools finite aligned observation times and makes a
common grid $G_e$. If there are at most 15 unique times, all are retained. If
there are more, 15 type-8 empirical quantiles are used and duplicate quantiles
are removed.

Each subject's transformed trajectory is linearly interpolated onto $G_e$. There
is no extrapolation during profile construction, so grid locations outside a
subject's observed window become missing profile features. Tied times are
averaged by `stats::approx(..., ties = mean)`.

For PK data a 15-point summary may compresses it. 
The grid points are quantiles of the observed times,
so they land where sampling is dense, early and around absorption and peak,
rather than spreading evenly across a long tail. The grid is also built per
endpoint, so PK and PD each get their own grid times rather than sharing them. For rich, 
 multiple-dose data with several peaks and troughs, 15 points may not be enough to 
 well characterize the shape and when that happens the cost is a worse choice of donors.  Again, this 15 point grid is only for distance calculating between patients and not for the final synthetic data profiles.

### Standardization across Multiple Endpoints

Every endpoint's columns go into one profile and one PCA.  Standardization of $\mathbf{x_i}$ equalizes...

The standardization below is what makes that possible, but it equalizes
**units**, not **endpoints**. Every grid point becomes one unit-variance column,
so an endpoint
sampled at 15 times contributes 15 columns while one sampled at 3 times
contributes 3, and a Euclidean distance summing over columns lets the first
outvote the second. Donor selection is therefore driven mostly by whichever
endpoint was sampled most densely, and baseline covariates are a small minority
of the distance whenever trajectories are rich. There is currently no
way to reweight endpoints.

The profiles are stacked into a matrix with one row per source subject $i$ and one column
per **feature** $l$: one column for each covariate, and one for each grid time of each endpoint, 
so a study with two covariates and two endpoints on fifteen grid points each has 32
columns.

Missing entries are replaced by that column's across-subject median $m_l$:

$$
x^{\mathrm{imp}}_{il}=
\begin{cases}
x_{il}, & x_{il}\text{ is finite},\\
m_l, & \text{otherwise}.
\end{cases}
$$

Each column is then centred on its own across-subject mean and divided
by its own across-subject standard deviation $s_l$:

$$
z_{il}=\frac{x^{\mathrm{imp}}_{il}-\bar{x}^{\mathrm{imp}}_l}{s_l}.
$$

Every column therefore arrives at the PCA with mean 0 and variance 1.  
A column whose standard deviation is not
greater than $\sqrt{\epsilon_{\mathrm{mach}}}$ is treated as constant and is dropped.

### Reducing the Profile to a Few Coordinates by PCA

The standardized matrix $z_{il}$ has $n$ rows, one per source subject, and $p$
columns, one per surviving feature. Its columns are heavily redundant: adjacent
grid times on one trajectory carry nearly the same information, so $p$ columns
hold far fewer than $p$ independent facts about a patient, and a distance
computed straight on those columns counts the repeated ones several times over.

**PCA refresher.** Principal component analysis (PCA) rotates the
$p$ columns into a new set of axes, the **principal components**. Each component
is one fixed weighted combination of the original columns, and the weights are
chosen so that the first component runs along the direction in which subjects
are most spread out, the second along the most spread remaining while being
uncorrelated with the first, and so on. Those directions are the **eigenvectors**
of the correlation matrix of the columns, and the variance of the subjects along
each one, written $\lambda_h$, is the matching **eigenvalue**. 

Because every column was standardized to variance 1 in the previous section, the
total variance is exactly $p$, and $\lambda_h/p$ reads directly as the share of
the whole profile's spread that component $h$ accounts for. The retained
dimension $H$ is the smallest number of components whose shares add up to at
least $v$, which is `pca_variance`, default 0.90:

$$
\frac{\sum_{h=1}^{H}\lambda_h}
     {\sum_{h=1}^{r_{\max}}\lambda_h}
\ge v .
$$

Each subject is then projected into the PCA space of H dimensions, and it is this space that is used to compute the distance between subjects in Step 8:

$$
\boldsymbol{\xi}_i=(\xi_{i1},\ldots,\xi_{iH}),
\qquad
\xi_{ih}=\text{subject } i\text{'s score on principal component } h .
$$

**Rank-safe** is the bound on how many components can exist at all:

$$
r_{\max}=\min(n-1,p).
$$

There can be no more directions than columns, hence $p$; and centering each
column on its own mean spends one degree of freedom, hence $n-1$. A twelve-subject
study therefore has at most eleven components however many grid points it
carries, and a component beyond $r_{\max}$ would be fitting numerical noise. 

## Step 5: Read Each Subject's Route and Dosing History

A subject given a single oral dose should not donate a trajectory to an infusion
template. So two facts are read off each source subject here, once for the whole
cohort, and used twice later: Step 6 needs the route to size the arms it may
anchor in, and Step 8 needs both to pick each avatar's donors.

**Route of administration is an absolute barrier.** A bolus, an infusion and an
oral dose produce categorically different concentration shapes, so blending
across them yields a trajectory no protocol could have produced. Donors are
never drawn from outside the anchor's route. Each subject's route is the set of

$$
\bigl(\mathrm{EVID},\ \mathrm{CMT},\ [\mathrm{RATE}\neq 0]\bigr)
$$

triples on its dosing rows, so the compartment a dose enters and whether it is
delivered over time both bind, while dose *count* does not. Because it is the
whole set, **a patient given more than one route is their own route class**: an
infusion followed later by an oral dose matches neither the oral-only nor the
infusion-only patients, only others with that same combination.

## Step 6: Restrict the Anchor Pool and Sample Anchors

### M2. Which Patients May Be Used as an Anchor

There are two reasons subjects may be excluded from being an anchor.  

1. A subject whose follow-up length or dose count exceeds twice the
cohort's 90th percentile is never used as an anchor (`screen`), so no avatar
inherits a conspicuous skeleton. 
2. A subject in a route arm holding fewer than
`k + 1` subjects has no legal donor set, so by default it is dropped from the
pool too (`on_donor_shortfall`), where `k` is the number of subjects to be chosen
for blending.

When an arm holds nobody who can be masked, individualised dosing in one
cohort of a trial for instance, its avatars keep the anchor they have, the run
raises an alert naming that arm, and rows B1a and B1b of
`vignette("scorecard-synthetic-data-checks")` fail against it.
`unmaskable_strata()` can assess whether any such strata exist.

### Sampling Anchors, and Keeping Each Stratum's Share

Anchors are sampled with replacement from the surviving pool. 
As the default `preserve_strata_balance = TRUE`, and so the target proportions of the
strata in the synthetic dataset match the source dataset.  However, **Strata holding fewer than three source patients are deliberately not balanced**, because reproducing such a stratum's size
exactly would say more about those few patients.

Setting `preserve_strata_balance = FALSE` allows strata sizes to vary.  

## Step 7: Copy the Anchor's Event Skeleton, and Redraw the Visit Schedule

Let subject $i$ have source rows $D_i$. For every requested synthetic subject,
the generator takes its sampled **anchor** $a$ and begins with

$$
S^{(0)} = D_a.
$$

Thus $S^{(0)}$ contains the anchor's **event skeleton**: row-wise combination of
TIME, EVID, AMT, RATE, CMT, DVID, and any other source columns. 

### M3. Redrawing Which Visits the Avatar Attended

After coarsening, every subject sits on the same grid of visit times. What still
differs between them is which of those visits carry an observation, and that
combination can identify a patient on its own.

```
grid            wk1  wk2  wk3  wk4  wk5  wk6
patient A        x    x    x    x    x    x     attended everything
patient B        x    x    x    x                stopped after week 4
patient C        x    x         x    x    x      missed week 3
```

Each row of x's is a **visit set**. If patient C is the only one who ever missed
week 3, then "attended everything except week 3" points at patient C as surely as
an unusual sampling time would.

So the avatar is given a different visit set from its anchor's. The doses and the
grid times are untouched; only which cells carry an observation changes. Three
things are tried in order, and the first that works is used:

1. **Reuse a set.** Take a visit set that at least `min_pattern_share` patients
   hold. The avatar then has a schedule several real patients had.
2. **Build one with a shared shape.** Where no set is held that widely, take a
   *shape* that at least `min_pattern_share` patients share and build a fresh set
   with it.
3. **Build one with a shared arrangement.** Where no shape is shared either, use
   the arrangement alone and take the count from what is typical for it. This is
   what a study with staggered follow-up needs, where every stopping point has a
   single holder.

A **shape** is how many visits were missed paired with how they were arranged. It
says nothing about which ones:

| Arrangement | Meaning |
|---|---|
| `complete` | nothing missed |
| `trailing` | every miss falls in the final visits, so the patient stopped: a discontinuation, or follow-up that has not reached those visits yet |
| `block` | the misses are consecutive but do not run to the end, so the patient was interrupted and came back |
| `scattered` | anything else |

With $n$ visit cells of which $m$ are missed, the arrangement is `trailing` when
every missed index exceeds $n-m$, `block` when the missed indices are consecutive
but not trailing, and `scattered` otherwise. One missed visit that is not the
last is a block of one.

Pairing the arrangement with the count is what makes step 2 work. Two patients
who each missed a single early visit hold different sets, so each is a group of
one and neither set can be reused. As `block|1` they are one group of two, and a
fresh block of one can be built.

When a fresh set is built, the arrangements that fit the wanted count are
enumerated and one that no single patient holds is taken. Trailing misses have
only one arrangement per count, so where that one is already somebody's, the
count moves outward by a visit or two until a free arrangement is found. Misses
are only ever placed on endpoints that some patient actually misses, since
putting one on an endpoint everybody has at every visit would invent missingness
the study does not have.

What survives is how much missingness the study had and of what kind. What is
lost is which weeks any one patient missed.

### M4. Redrawing Where Dosing Stopped

A patient's dose schedule is their list of dose times, and it identifies them
whenever nobody else has the same one. Two terms for talking about it, both
counting doses from the start of therapy rather than describing gaps within it.
A patient's **depth** is how many doses they received, and their **opening at
depth $d$** is their own first $d$ dose times. A patient dosed at 0, 12, 24 and
36 hours has depth 4 and an opening at depth 2 of {0, 12}.

Dose times are never moved or invented, because a drawn regimen could be one no
protocol permits. The one edit allowed is truncation: keeping an opening and
dropping the doses after it leaves a course some patient could have been given,
namely one who stopped early.

So where a patient's schedule is one nobody else has, the avatar's is shortened.
The search starts one dose short of the patient's depth and walks **downward**,
taking the first depth that is safe. It walks down rather than up because
truncation can only remove dose rows, so a deeper target would leave the schedule
where it was. A depth is safe on either of two grounds:

- **Matched.** The opening is exactly the complete schedule of at least
  `min_pattern_share` patients, so the avatar finishes where they finished.
- **Free.** Nobody holds that opening as a complete schedule, and at least
  `min_pattern_share` patients opened with it. The only individual fact such a
  schedule carries is where it stopped, and nobody stopped there.

Worked through, with a floor of two:

```
dose times (h)       0   12   24   36   48
patient A            x    x    x                depth 3
patient B            x    x    x                depth 3
patient C            x    x    x                depth 3
patient D            x    x    x    x    x      depth 5, shared by nobody
```

An avatar anchored on D would carry D's schedule alone. Depth 4 is tried first:
{0, 12, 24, 36} is nobody's complete schedule, but only D ever opened with it, so
it is neither matched nor free. Depth 3 is {0, 12, 24}, which is the complete
schedule of A, B and C, so it is matched and the avatar stops at 24 hours. Had A,
B and C instead run to depths 4, 5 and 6, depth 3 would have been free rather
than matched: nobody would have stopped there, and three patients would have
opened with it.

Both halves of *free* are needed, and the second is the one easily lost. Checking
the dose times one at a time, so that each kept time is one several patients
received, is not enough, because the combination can still belong to one patient.
Getting that wrong leaves avatars truncated to an opening only one real patient
holds, that patient's dosing minus its last dose, and nothing detects it: the
copy-detection checks compare complete schedules, so a set nobody holds complete
is exactly the set that passes. Openings are counted whole.

Not every cohort can be served. Where the patients who stopped early are the sole
holders of their depths, no depth is matched or free, and that regimen cannot be
represented. Those patients fall to M2 and are not built upon, and the masking
report shows the regimen coverage it cost.

Two study shapes are served. In a protocol cohort with staggered discontinuation,
everyone opens on the same schedule, so any depth nobody stopped at is free and a
target is available almost anywhere. The other is routine clinical care, where
every patient begins on the same interval and diverges as their own clinician
decides, so complete schedules rarely repeat while openings do. `pheno_sd` is
that second shape, and truncating each course back to its shared opening keeps a
substantial part of the dosing.

## Step 8: Choose the Donors

### The Distance

For anchor $a$ and candidate donor $r$, the distance is the ordinary
straight-line (Euclidean) distance between their subject coordinates from
Step 4:

$$
d_{ar}=\left\|\boldsymbol{\xi}_a-
                   \boldsymbol{\xi}_r\right\|_2
=\left(\sum_{h=1}^{H}(\xi_{ah}-\xi_{rh})^2\right)^{1/2},
$$

where $h$ runs over the $H$ retained principal components. All the work went
into choosing the coordinates; this is Pythagoras in $H$ dimensions.

This is the only distance in the selection rule. There is no separate structural
metric trading dose differences against schedule differences: structure enters as
the two-stage ordering below rather than as a score.

### The Selection Algorithm

To choose $k$ donors for anchor $a$, using the route and the dosing history of
Step 5:

1. **Same-history stage.** Among the other subjects on the anchor's route that
   also share its dosing history, take the $k$ nearest by $d_{ar}$, breaking
   ties by subject index. If there are at least $k$ of them, stop.
2. **Fallback stage.** Otherwise fill the remainder from the nearest subjects on
   that route whatever their dosing history, and record a warning that donors
   were borrowed across dose groups.
3. **Shortfall.** If the anchor's route holds fewer than $k$ other subjects at
   all, the floor is unreachable, because the only remaining candidates are on
   another route and route is never crossed.

Selection is therefore deterministic under a fixed seed. The fallback stage
matters most for datasets with individualized dosing, where an exact dose magnitude can leave
nearly every subject holding a dosing history of its own and the fallback stage
does essentially all the work.

### What the Fallback Stage Costs

Be clear about what the fallback distance does and does not know. `AMT` is not a
profile feature, so the fallback ranking never compares doses directly. It
compares baseline covariates and the shape and level of the DV trajectory. Dose
does influence it, but only *indirectly*: a subject given a much larger dose has
higher concentrations, and those concentrations are profile features, so it lands
further away. The ranking therefore tends to prefer similar-dose donors without
being told to.

Tendency is not a guarantee, and nothing rescales. A borrowed donor's values are
interpolated onto the anchor's times and blended **raw**, by deliberate design,
choosing structural realism over statistical fidelity. The consequence is the
one stated at the top of this page: a synthetic subject carries its anchor's
`AMT` while its concentrations may be blended from donors given a different
dose, so **the dose-exposure relationship in AVATAR output is not guaranteed to
be preserved**. Fitting a structural model to AVATAR output can recover a
dose-exposure relationship that is an artifact of blending. Use
`synpmx_calibrated()` when the parameters themselves have to mean something.

### When the Donor Floor Cannot Be Reached

Because route is absolute, a route arm holding fewer than $k+1$ subjects can
never supply a legal donor set. There is no good answer in that situation, only a
choice between two bad ones, and which is worse depends on whether the sparse arm
matters more than the disclosure risk of reproducing it. That is the caller's
judgement, so `on_donor_shortfall` makes it explicit:

| `on_donor_shortfall` | Behavior |
|---|---|
| `"drop"` *(default)* | Omit those subjects from the anchor pool. No avatar is built on them, and the synthetic cohort does not represent that arm. |
| `"noise"` | Keep them, blending however many same-route donors exist, possibly none, and relying on `subject_noise_sd` and `residual_noise_sd` for the rest. |
| `"error"` | Refuse to generate, naming the arm and both alternatives. |

Every branch alerts loudly, because each one silently changes something a reader
would want to know: what the cohort covers, or how identifying it is.

`"noise"` is **available but not recommended**, and the reason is the defect this
whole mechanism exists to prevent. A subject blended from one donor, or from
none, is a noised near-copy of a real patient. If you use it because the sparse
arm genuinely matters more, screen the result and treat those subjects as
individually identifying.

One case forces the issue: when *every* arm is below the floor, `"drop"` would
leave nothing to generate, so generation proceeds as if `"noise"` and says so.

## Step 9: Randomize and Cap the Donor Weights

Suppose $K$ donors have been selected. Let $E_r\sim\mathrm{Exponential}(1)$ and
let $R_r$ be a random permutation of $1,\ldots,K$. Importantly, $R_r$ is a
randomized rank rather than the distance order, which is what stops the nearest
neighbour from being systematically the largest contributor. The raw donor weight
is

$$
q_r=\frac{E_r}{\max(d_{ar},\varepsilon)}2^{-R_r},
$$

where

$$
\varepsilon=
\begin{cases}
\max\{10^{-8},10^{-6}\operatorname{median}(d_{ar}:d_{ar}>0)\},
  & \text{if a positive distance exists},\\
10^{-8}, & \text{otherwise}.
\end{cases}
$$

The first normalization is $w_r=q_r/\sum_s q_s$. If the raw weights are
nonfinite or have a nonpositive total, the implementation replaces them with
equal raw values before normalizing.

### Capping Every Donor

Each donor's weight is its **share of the avatar**: five weights that sum to 1.
The floor $k$ says how many real patients go into the blend, but it says nothing
about how much of any one of them comes out. Five donors weighted $(0.95, 0.02,
0.02, 0.005, 0.005)$ satisfy a floor of five while being, to any practical
purpose, a copy of one person. So the cap, rather than $k$, is what controls how
closely an avatar can resemble a single real patient.

The raw formula above produces very uneven weights on purpose, which is the rank
attenuation doing its job. `max_donor_weight` (written $c$ below, default $0.50$)
is the ceiling on any one share.

Enforcing it is less obvious than it sounds, because **weights have to keep
summing to 1**. Cutting the leader down does not remove that weight from the
blend; it has to be given to the others, in proportion to what they already hold:

$$
w_r \leftarrow \Bigl(1-\textstyle\sum_{s\in P}w_s\Bigr)
              \frac{w_r}{\sum_{s\notin P}w_s},
\qquad r\notin P,
$$

where $P$ is the set of donors already pinned at $c$. And that redistribution can
itself push a donor over the ceiling. Starting from $(0.50, 0.25, 0.15, 0.06,
0.04)$ with a tight $c = 0.30$: cutting the leader to $0.30$ frees $0.20$, and
the second donor's share of it takes them from $0.25$ to $0.35$, over the
ceiling. Capping only the largest weight and stopping, which is what the
implementation did throughout the $0.80$ era, leaves exactly that violation in
place: the stated maximum, broken by the donor the redistribution created.

So the rule repeats, pinning everything over $c$, redistributing, and looking
again, which is the standard construction called **water-filling**. Pinned donors
are never topped up, so each pass pins at least one more and it terminates. The
tight example settles at $(0.30, 0.30, 0.24, 0.096, 0.064)$.

### One Pass Suffices at the Default Cap

A second pass is needed only when redistribution lifts the runner-up over the
ceiling, that is when

$$
(1-c)\,\frac{w_2}{1-w_1}>c .
$$

At $c=0.5$ this requires $w_2>1-w_1$, i.e. $w_1+w_2>1$, which no pair of weights
can satisfy. **At any cap of $0.5$ or above the loop runs at most once**, and
simulation agrees exactly: at $c=0.5$ the cap is untouched 33% of the time and
pinned once 67%, never twice. Only tighter caps exercise the iteration; at
$c=0.3$ two passes are needed 53% of the time and three passes 11%.

The general routine is kept because `max_donor_weight` is an argument and a user
may set it below $0.5$, where a single pass is genuinely wrong. But at the
shipped default, the simple thing and the correct thing coincide.

### The Infeasible-Cap Boundary

With $K$ donors the weights average $1/K$, so **no weight vector can satisfy a
ceiling below $1/K$**: asking for $c=0.3$ from three donors is asking three
numbers below $0.3$ to sum to 1. Such a cap relaxes to exactly $1/K$, uniform
weights and the flattest blend available, rather than an error. Two donors under
$c=0.5$ give $(0.5,0.5)$, which meets the cap exactly. With one donor the weight
is necessarily 1, since a cap cannot invent a second donor.

### What an Uncapped Blend Looks Like

It is reasonable to ask whether the cap earns its place, given that $k$ already
forces five donors into every avatar. The answer is that **without a cap, the
floor is largely decorative.** Simulating the raw weight formula at $k=5$ with
representative donor distances:

| uncapped largest donor share | value |
|---|---|
| median | 0.58 |
| interquartile range | 0.47 - 0.72 |
| $P(\text{largest} > 0.5)$ | 0.67 |
| $P(\text{largest} > 0.8)$ | 0.14 |
| effective donors $1/\sum_r w_r^2$ | 2.37 |

So the unconstrained formula typically puts **58% of an avatar into a single
donor**, and one avatar in seven takes more than 80% from one person. Nominally
five patients are blended; effectively about two and a half are. The cap is the
only thing that closes that gap.

### Choosing the Cap

The useful diagnostic is **how often the cap fires**, because that says what role
it is playing:

| `max_donor_weight` | binds on | effective donors |
|---|---|---|
| 0.30 | 99% | 3.92 |
| 0.40 | 89% | 3.28 |
| 0.50 *(default)* | 68% | 2.90 |
| 0.60 | 47% | 2.64 |
| 0.80 | 15% | 2.40 |
| none | 0% | 2.37 |

Both ends of that table are the wrong kind of parameter. A cap that fires on 99%
of subjects is not a guardrail, it *is* the weighting scheme, and the
inverse-distance term underneath it barely matters. A cap at $0.80$ fires on 15%,
trimming only the worst tail while still permitting one patient to be four-fifths
of an avatar; that was the old default, and it was never really protecting
anything.

The default is **$0.50$**, which states as a single checkable sentence: *no
single real patient is more than half of any synthetic patient.* It fires on
about two thirds of subjects, so it genuinely constrains without wholly replacing
the distance weighting.

Its honest cost is that "five donors are blended" is really "about three
effective donors". Tightening to $0.30$ buys about one more at the price of a cap
that fires essentially always. Both are defensible; $0.50$ is the one that can be
explained in a sentence.

### What the Cap Costs

Treating donors as independent, a weighted blend retains $\sum_r w_r^2$ of
individual variance, so the synthetic cohort's between-subject standard deviation
shrinks by $(\sum_r w_r^2)^{1/2}$. The reciprocal $1/\sum_r w_r^2$ is the
effective number of donors, and it reads as a privacy floor and as the
variability cost in one number: $K$ for uniform weights, $1$ for a sole donor.
Because both depend on the cohort, `cap_binding_fraction` and
`mean_effective_donors` are recorded for every run, so the question can be asked
of real data rather than of a simulation.

That independence assumption is pessimistic in the direction that matters.
Donors are *nearest neighbours*, so they are strongly positively correlated, and
averaging them destroys much less variance than the formula implies; the
subject-level perturbation of Step 11 then restores more. Measured on a real
study, tightening the cap raises the effective donor count substantially while
between-subject variability stays flat within noise, where the independence
formula would have predicted a steep fall. This is why the cap can be chosen on
privacy grounds: the fidelity side of the trade is nearly flat.

The same selected subjects and weights are used for every declared covariate and
endpoint. When donor $r$ lacks a usable value at target location $j$, the
available weights are renormalized locally:

$$
w_{rj}^{*}=\frac{w_r I_{rj}}
                 {\sum_s w_s I_{sj}},
$$

where $I_{rj}=1$ when the value is finite and zero otherwise.

### M5. Blending as the Mechanism That Protects the Values

Spanning Steps 8 to 11, this is the mechanism that protects the measurements,
and the only one whose strength grows with cohort size. That is why a source with
fewer than `k + 1` subjects raises a loud alert instead of quietly producing
near-copies.

## Step 10: Synthesize the Baseline Covariates

For a numeric covariate, let $c_r$ be donor $r$'s first nonmissing value. When
all available values are positive and

$$
\frac{\max_r c_r}{\operatorname{median}_r(c_r)}>3,
$$

the blend is formed on the log scale; otherwise it is formed on the original
scale. With working-scale values $h(c_r)$, the donor center is

$$
\mu_c=\sum_r w_r^* h(c_r).
$$

Let $s_c$ be the ordinary sample standard deviation of the available
working-scale donor values. If it is unavailable or zero, the fallback is
$\max(0.05|\mu_c|,0.01)$. The generated working-scale covariate is

$$
c^*=\mu_c+\eta_c,
\qquad
\eta_c\sim\mathrm{N}(0,(\sigma_{\mathrm{subj}}s_c)^2),
$$

where `subject_noise_sd` is $\sigma_{\mathrm{subj}}$ (default 0.15). A log-scale
result is exponentiated, and any all-positive covariate is floored above zero.
The final value is repeated on every row of that synthetic subject.

For factor, character, and logical covariates, one available donor category is
sampled using the locally normalized weights. Categories are not averaged.

### A Rare Category Is Copied Rather Than Blended

A numeric covariate comes out of this step as a weighted mean plus noise, a new
number nobody had. A **categorical** covariate has nothing to average, so a
synthetic patient's category is always **some real patient's actual category,
copied**. Declared `strata` are copied from the anchor exactly, by design, since
they are protocol facts.

So the question is when a copied category reaches the output, and the answer
turns on how many real patients hold the level. Measured on a fixture where one
categorical level is held by a varying number of otherwise-ordinary patients and
everything else is held fixed: **a level held by one patient never reaches the
output; a level held by two does.** The mechanism is geometric rather than
protective. Each categorical level enters the profile as its own feature, so the
sole holder of a level sits alone on that axis, is nobody's nearest neighbour,
and is therefore almost never selected as a donor. Give the level a second holder and the two
become each other's nearest neighbour, the level propagates between them, and it
leaves. By ten holders the output tracks the source frequency closely.

The operative variable is the **number of holders**, not outlier status. Making
the two holders numeric outliers as well changes nothing on the categorical axis.
"Being an outlier is self-protecting" holds for numeric covariates, which are
blended, and does not hold for categories.

Two things follow, and the second is the one that matters.

**This is not a mechanism to rely on.** It is a side effect of nearest-neighbour
donor selection. Nothing enforces it, nothing reports it, and it degrades
smoothly: two holders is already a leak.

**The dangerous shape is a rare level on a *typical* patient.** A patient who is
unusual on their covariates is nobody's nearest neighbour and is self-protecting.
A patient who is ordinary in every way except one rare category, an ultra-rare
mutation status say, sits in the middle of the profile space, is selected as a
donor constantly, and their category is copied out. What to do about that is a
checking question rather than an algorithm one, and it is in
`vignette("scorecard-synthetic-data-checks")` under B5.

### A Covariate and an Endpoint Are Blended Independently

Covariates (this step) and endpoint trajectories (the next) are generated in
separate passes, each with its own donor weighting and its own perturbation. That
is fine when they are unrelated, but some datasets carry a **baseline covariate
that is the same quantity as a longitudinal endpoint**: a `B0` baseline B-cell
count beside a B-cell kinetic endpoint, a baseline biomarker beside its own time
course. AVATAR does not know they are linked, so a synthetic subject's baseline
covariate need not equal the baseline of its own generated trajectory. The two
are individually plausible but not mutually consistent.

This is a known limitation. It is usually harmless for workflow-development use,
where each column need only be realistic on its own. If your analysis relies on a
covariate agreeing with an endpoint's baseline, reconcile them after generation,
or do not declare the redundant covariate and read the baseline from the
trajectory instead.

## Step 11: Synthesize Each Endpoint Trajectory

Only positions that were eligible, nonmissing observations in the anchor template
receive synthesized values. Thus the anchor supplies the number and placement of
observations, including its missing-DV pattern, as redrawn by M3.

For endpoint $e$, donor trajectories are transformed with $g_e$ and interpolated
to the anchor observation times. These are the anchor's own times, not the
fifteen-point grid of Step 4, so the avatar keeps its anchor's sampling
resolution however rich that is.

Suppose target time $t$ lies outside donor $r$'s window. Rather than
extrapolating indefinitely, the method maps the target's relative position in the
whole anchor observation window to the donor window:

$$
q(t)=\frac{t-a_{\mathrm{target}}}
           {b_{\mathrm{target}}-a_{\mathrm{target}}},
\qquad
u_r(t)=a_r+q(t)(b_r-a_r),
$$

where $[a_{\mathrm{target}},b_{\mathrm{target}}]$ and $[a_r,b_r]$ are the target
and donor ranges. The donor trajectory is evaluated at $u_r(t)$. This
normalized-window fallback preserves progress through a trajectory, but it is not
a mechanistic warping model. A donor with only one unique time supplies its mean
at every requested target time.

Let $z_{rj}$ be donor $r$'s transformed value at target position $j$. The
deterministic blend is

$$
\bar z_j=\sum_r w_{rj}^{*}z_{rj}.
$$

If every donor is unavailable at a target, the transformed dataset median for
that endpoint is used and a warning is stored.

The endpoint noise scale $s_e$ is 1 on the offset-log scale. On the identity
scale it is the source endpoint standard deviation, with fallback
$\max(0.1|\operatorname{median}(Y_e)|,0.01)$ when necessary. A single
subject-level shift is shared across all positions of the endpoint:

$$
b_e\sim\mathrm{N}(0,(\sigma_{\mathrm{subj}}s_e)^2).
$$

Within-endpoint residual perturbations follow a stationary-scale first-order
autoregressive, AR(1), process in **observation order**:

$$
\begin{aligned}
\epsilon_{e1} &\sim
  \mathrm{N}(0,(\sigma_{\mathrm{res}}s_e)^2),\\
\epsilon_{ej} &= \phi\epsilon_{e,j-1}+\nu_{ej},\\
\nu_{ej} &\sim
  \mathrm{N}\left(0,
    (\sigma_{\mathrm{res}}s_e)^2(1-\phi^2)\right).
\end{aligned}
$$

Defaults are `residual_noise_sd = 0.05` and `residual_phi = 0.6`. The generated
DV is

$$
Y^*_{ej}=g_e^{-1}(\bar z_j+b_e+\epsilon_{ej}).
$$

Endpoints are processed separately, so PK and PD values cannot be blended with
one another. They nevertheless use the same donor subjects and initial subject
weights, preserving a limited form of cross-endpoint subject coherence. This is a
synthesis device, not a fitted multivariate residual model.

### Snapping a Discrete Endpoint Back onto Its Levels

$Y^*_{ej}$ above is a new number, which is what a continuous endpoint wants and
what a discrete one cannot have. So for an endpoint the value type of Step 3 made
`binary` or `ordinal`, the generated value is replaced by the **nearest observed
level**, deciding ties at the midpoints between adjacent levels. For an `integer`
endpoint it is rounded, and floored at zero where the source held no negative
value. A `continuous` endpoint is left alone.

The snap happens after the noise rather than before, so the subject shift and the
AR(1) sequence still decide *which* level a given observation lands on. What they
cannot do is invent a level between two the source recorded.

The count of generated values that could not be placed on a legal level is
recorded as `endpoint_type_violations`, and should be zero.

### Below-the-Limit-of-Quantification (BLOQ) Data and a Monolix-Style CENS Column

Declare `cens` and, if the source uses one, `limit` in `pmx_roles()`, and AVATAR
reconstructs censoring rather than copying it. The mechanism has three parts, and
the last of them happens here, alongside the snap above: both are corrections
applied to the generated value before it is written.

**The boundary is read from the source.** AVATAR makes no formal privacy claim,
so it may recover the assay limit from the data itself: under the Monolix
convention a censored row reports the boundary in `DV`, with `LIMIT` carrying the
other end of an interval when there is one. Left, right, and interval censoring
are each recognised. A differentially private engine may *not* do this, because
there the limit must be a declared public input: a boundary inferred from
confidential records is itself a release. That asymmetry is the reason the two
engines derive boundaries differently. If a study used more than one assay limit,
the most conservative boundary is taken rather than inventing a per-row rule the
source cannot support.

**Censored donor values are imputed before blending.** Every censored source row
reports the same substituted number, so blending those values directly would put
a floor under the synthetic data that the real study does not have. Each censored
row is instead replaced by a draw inside its censoring region, uniform below the
limit for a left-censored row and uniform within the interval for an
interval-censored one, before the trajectory is transformed and blended. A
uniform draw rather than a fixed LLOQ/2 avoids replacing one artificial spike
with another.

**`DV`, `CENS`, and `LIMIT` are reconstructed together.** The back-transformed
blend is treated as a *latent* value: what the subject would have measured with
no assay limit. The boundary is then applied to that latent, and all three
columns are written from the result, so they cannot disagree. A generated value
below the limit produces `CENS = 1` with `DV` exactly at the limit, reproducing
the spike of identical values that characterises real censored data.

Two honest limitations. The censored *fraction* is reproduced approximately
rather than exactly, because blending shrinks variance toward the middle while
the imputation draws uniformly across the censoring region, so the synthetic rate
typically runs somewhat above the source's. And if a `cens` role is declared but
no boundary can be read from it, rows flagged with `DV` missing for instance,
AVATAR warns and carries the flag through untouched rather than guessing.

## Step 12: Recompute the Dose from the Blended Covariate

### M6. Recomputing the Dose

Under dosing proportional to a baseline covariate, the multiplier is a protocol
property the stratum shares and the covariate is individual, and already blended.
Keeping the multiplier and recomputing the amount from the avatar's own covariate
stops the dose being a verbatim real value, which under mg/kg dosing discloses
the anchor's weight exactly, and repairs a coherence defect: previously every
avatar violated the mg/kg rule its own data claims to follow.

**Declared or inferred, and the difference matters.** Naming the covariate in
`pmx_roles(dose_covariate = )` states the relationship outright; leaving it
`NULL` makes the run infer one. Inference is deliberately conservative: it
requires the dose-to-covariate ratio to collapse onto a handful of levels within
2%, because rewriting amounts on a study that is *not* proportional would be
worse than leaving them alone. Two ordinary things defeat that test, dispensing
in vials and patients escalating by factors that are not identical across the
cohort, so a genuinely weight-based study can come through with its amounts
untouched. Declaring the covariate skips inference and holds **each dose row's
own ratio** rather than snapping to shared levels, so intra-patient escalation
survives exactly. `pmx_masking_report()` says which path ran and, when inference
declined, what it tried.

On a study that dispenses in vials and escalates within patient, which is the
shape inference cannot recover, the difference is the whole mechanism. Left to
infer, the amounts are copied, and each avatar ends up carrying its anchor's
milligrams over its *own* blended weight, a dose per kilogram nobody was
prescribed. Declared, every avatar's course of mg/kg is one the protocol
actually gave, dose row for dose row, and patients who escalated by different
factors come through intact.

## Step 13: Construct the Full PMX Dataset

The synthetic subjects are bound into one dataset, and every column is returned
to the form the source held it in. The point is that the result drops into code
written against the source without anything having to be adjusted: a column that
was an integer is still an integer, a factor still has its levels in their
original order, and a tibble comes back a tibble.

After the subjects are row-bound, the original column order is restored, and each
result column is converted toward its source class:

- factors recover source levels and ordering; new factor IDs add new levels;
- integer columns are rounded and converted to integer;
- double, logical, character, Date, and POSIXct columns are restored; and
- a non-`data.frame` source class, such as a tibble class vector, is reapplied.

New numeric IDs start above the source maximum. New character and factor IDs use
labels such as `syn_001`, with a prefix added repeatedly if needed to avoid
collision.

If MDV is declared and the entire source obeys the standard relationship

$$
[\mathrm{MDV}=0]=[\mathrm{EVID}=0\ \text{and DV is present}],
$$

MDV is re-derived from that rule in the synthetic data and restored to its
original class. For any nonstandard source-wide MDV convention, anchor values are
left as copied. This step neither invents nor removes observation positions,
which ones an avatar has having been settled by M3 in Step 7.

Finally, `validate_pmx(..., strict = TRUE)` is run on the assembled result. A
generation call cannot silently return an output that fails the package's own
structural requirements.

## Reproducibility

All random operations occur inside a local seed scope. `seed` must be an integer
between zero and `.Machine$integer.max`. The caller's random-number generator
kind and `.Random.seed` are restored on exit, whether or not `.Random.seed`
existed before the call. Consequently, identical inputs and arguments produce an
identical complete result without consuming the surrounding analysis RNG stream.

`attr(synthetic, "pmx_settings")` records every argument, the explicit roles, and
what each mechanism did and cost. `pmx_masking_report()` prints it, and that is
the intended way to read it. It does not record the anchor ID, donor IDs,
distances, or realized weights for each synthetic subject.

## Checking a Run

Nothing above is self-certifying. Every mechanism costs something, in discarded
visit sets, excluded anchors, or timing fidelity, and what it buys **depends
entirely on the study**: the same defaults take one cohort to zero identifying
schedules and leave another at all of them, because one study recorded a shared
visit structure and the other did not. There is no headline claim to make, so
every run reports its own numbers instead.

`synpmx_scorecard()` is where those numbers are read. Every mechanism on this
page has a row there, along with the structural checks that a synthetic dataset
is legal PMX data at all.
`vignette("scorecard-synthetic-data-checks")` explains what each row asks and
what counts as passing, and `vignette("public-data-examples")` runs the whole
sequence across eight public datasets, including one where the guarantee cannot
hold and what is done about it.

## Edge Cases and Documented Fallbacks

The implementation favors an explicit warning and structurally valid output over
a silent incompatible blend.

```{r edge-cases-table, echo=FALSE}
edge_cases <- data.frame(
  Situation = c(
    "Fewer than k donors with the anchor's dosing history",
    "Route arm smaller than the k-donor floor",
    "Every route arm below the floor",
    "Cap below 1/K for K donors",
    "Source smaller than the k-donor floor",
    "All selected distances essentially zero",
    "One donor time only",
    "No donor value at a target",
    "No variable profile feature",
    "Identical donors with all noise disabled",
    "Nonstandard MDV convention"
  ),
  Behavior = c(
    "Borrow the nearest donors from other dose/schedule groups on the same route",
    "Follow on_donor_shortfall: drop (default), noise, or error; alert loudly",
    "Generate anyway and raise a loud alert; treat output as identifying",
    "Relax the cap to 1/K, i.e. uniform weights, rather than erroring",
    "Blend all available donors and raise a loud alert",
    "Use epsilon-stabilized randomized weights",
    "Repeat that donor's mean at every target time",
    "Use transformed endpoint median and record a warning",
    "Use a shared zero coordinate; distance cannot separate subjects",
    "A source-shaped trajectory may be mathematically unavoidable; warn",
    "Copy anchor MDV values rather than re-derive them"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(edge_cases, align = c("l", "l"))
```

Other boundaries are deliberate. `synpmx_avatar()` supports only
`event_method = "template"` and `dv_method = "avatar_blend"`. It does not fit a
structural model, reconstruct covariate-parameter relationships, simulate
occasions, enforce a dosing grammar beyond the sampled template, or provide a
censoring model. Fitting an explicit structural model is the job of
`synpmx_calibrated()`, which `vignette("synpmx-4-methods")` compares against this
one.

## Relationship to AVATAR

The method is **AVATAR-inspired**, not an implementation of published AVATAR
software. "AVATAR" is a method name rather than an initialism, from the
patient-centric *avatarization* literature in which each synthetic record is
built from the local neighborhood of real records. The original patient-centric
AVATAR method was reported by Guillaudeux and colleagues [1]. Destere and
colleagues describe a modified AVATAR benchmark for longitudinal population
pharmacokinetic (PopPK) data that first widened the data by subject and then used
PCA, K-nearest neighbors, inverse Euclidean distance, exponential stochastic
weights, and rank attenuation before reverse transformation [2]. The shared
family resemblance is the use of standardized subject features, PCA, local
neighbors, inverse-distance influence, exponential randomization, and rank
attenuation.

`synpmx` adapts those ideas to a PMX event table in several material ways:

- it preserves a sampled longitudinal event template rather than generating all
  fields from a wide vector;
- route of administration is an absolute barrier, while dose and schedule
  differences only make a donor less preferred;
- endpoints are transformed and interpolated separately, and a discrete endpoint
  is returned to its own levels;
- the same donors support covariates and endpoints;
- every multi-donor weight, not only the dominant one, is capped at
  `max_donor_weight` (default 0.50);
- subject and AR(1) perturbations are added on the endpoint working scale; and
- there is no inverse-PCA reconstruction of a full synthetic patient vector.

Destere et al. also demonstrate why method labels are not validation: their
benchmark found materially different PK parameter and residual-error biases
across synthesis algorithms [2]. That paper is a 2026 medRxiv preprint and was
not peer reviewed at the version supplied with this package repository. Its
privacy analyses do not transfer to `synpmx`; this package has undergone no
corresponding attack-based privacy validation.

## Algorithm Summary

What `synpmx_avatar()` executes, in order. These are implementation operations,
finer-grained than the thirteen steps above and following the same order.

1. Validate the source data, the roles, and the generator arguments.
2. Coarsen the recorded times onto a shared visit grid and pool the deviations.
3. Read the censoring boundary and impute censored source values.
4. Compute first-event-relative time and the per-endpoint transformation and
   value type.
5. Assemble baseline covariates and interpolated endpoint trajectories into one
   profile per source subject.
6. Median-impute profile features, remove unusable ones, standardize, and retain
   enough PCA components to meet `pca_variance`.
7. Read each subject's route of administration and dosing history.
8. Resolve anchors whose route arm cannot reach the donor floor per
   `on_donor_shortfall`, apply the structural screen, and sample the remaining
   anchors with replacement, within stratum where strata are declared.
9. For each anchor, copy its whole event skeleton, then redraw its visit set and
   truncate its dose schedule where either is unique.
10. Find `k` donors on the anchor's route, taking those sharing its dosing
    history nearest-first and then the nearest remaining subjects on that route,
    and randomize their inverse-distance weights, water-filling every weight to
    `max_donor_weight`.
11. Generate subject-constant covariates from those donors.
12. For each endpoint, interpolate donor trajectories to the anchor times, blend,
    add a subject shift and AR(1) perturbations, back-transform, snap a discrete
    endpoint to its levels, and reapply the censoring boundary.
13. Recompute dose amounts from the blended covariate where dosing is
    covariate-proportional.
14. Re-derive standard MDV when applicable, give every synthetic subject a new
    ID, preserve tie order, return every column to the form the source held it
    in, and validate the assembled dataset.
15. Attach `pmx_settings` and emit any collected fallback warnings.

## References

1. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).

2. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).
