---
title: "Evaluating AVATAR on public data"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Evaluating AVATAR on public data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")
```

This vignette compares eight source datasets against their synthetic
counterparts, so it needs a few helpers to pull observed rows into a common
shape and plot them. They are ordinary plotting code, not part of the package
API; expand the block if you want to reuse them.

<details>
<summary>Plotting helpers used throughout this vignette</summary>

```{r plot-helpers}
observed_plot_data <- function(data, roles, dataset,
                               clock = "study_time") {
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  if (!is.null(roles$mdv)) {
    observed <- observed & as.character(data[[roles$mdv]]) %in% c("0", "0.0")
  }
  observed <- observed & !is.na(data[[roles$dv]])
  observation_rows <- which(observed)
  occasion <- rep(1L, length(observation_rows))
  tad <- rep(NA_real_, length(observation_rows))
  if (!is.null(roles$occasion)) {
    declared <- suppressWarnings(as.integer(
      data[[roles$occasion]][observation_rows]
    ))
    valid <- !is.na(declared) & declared >= 1L
    occasion[valid] <- declared[valid]
  }
  if (!is.null(roles$tad)) {
    declared <- suppressWarnings(as.numeric(data[[roles$tad]][observation_rows]))
    valid <- is.finite(declared)
    tad[valid] <- pmax(0, declared[valid])
  }
  subject_values <- data[[roles$id]]
  for (id in unique(subject_values[observation_rows])) {
    subject_rows <- which(!is.na(subject_values) & subject_values == id)
    events <- !(as.character(data[[roles$evid]][subject_rows]) %in%
                  c("0", "0.0"))
    if (!is.null(roles$amt)) {
      events <- events & as.numeric(data[[roles$amt]][subject_rows]) > 0
    }
    positions <- which(subject_values[observation_rows] == id)
    event_rows <- subject_rows[events]
    if (length(event_rows) && !is.null(roles$occasion)) {
      event_occasion <- suppressWarnings(as.integer(
        data[[roles$occasion]][event_rows]
      ))
      for (position in positions) {
        candidates <- event_rows[event_occasion == occasion[position]]
        if (length(candidates) && !is.finite(tad[position])) {
          origin <- min(as.numeric(data[[roles$time]][candidates]))
          tad[position] <-
            as.numeric(data[[roles$time]][observation_rows[position]]) - origin
        }
      }
    } else if (length(event_rows)) {
      dose_times <- sort(unique(as.numeric(data[[roles$time]][event_rows])))
      occasion[positions] <- pmax(1L, findInterval(
        as.numeric(data[[roles$time]][observation_rows[positions]]),
        dose_times
      ))
      occasion[positions] <- pmin(occasion[positions], length(dose_times))
      tad[positions] <-
        as.numeric(data[[roles$time]][observation_rows[positions]]) -
        dose_times[occasion[positions]]
    }
  }
  plotted_time <- if (identical(clock, "tad")) tad else
    as.numeric(data[[roles$time]][observation_rows])
  data.frame(
    dataset = factor(dataset, levels = c("Source", "Synthetic")),
    subject = as.character(data[[roles$id]][observation_rows]),
    time = plotted_time,
    dv = as.numeric(data[[roles$dv]][observation_rows]),
    occasion = occasion,
    endpoint = if (is.null(roles$dvid)) "DV" else
      as.character(data[[roles$dvid]][observation_rows]),
    stringsAsFactors = FALSE
  )
}

tad_plot_data <- function(data, roles, dataset) {
  out <- observed_plot_data(data, roles, dataset, clock = "tad")
  names(out)[names(out) == "time"] <- "tad"
  out
}

demo_design_summary <- function(data, roles, dataset, time_bounds,
                                clock = "study_time") {
  plotted <- observed_plot_data(data, roles, dataset, clock)
  if (!identical(clock, "tad")) {
    plotted <- plotted[
      plotted$time >= time_bounds[1L] & plotted$time <= time_bounds[2L],
      , drop = FALSE
    ]
  }
  cohort_ids <- as.character(unique(data[[roles$id]]))
  pieces <- lapply(sort(unique(plotted$endpoint)), function(endpoint) {
    rows <- plotted[plotted$endpoint == endpoint, , drop = FALSE]
    counts <- table(factor(rows$subject, levels = cohort_ids))
    data.frame(
      dataset = dataset,
      endpoint = endpoint,
      patients = length(cohort_ids),
      patients_with_endpoint = sum(counts > 0),
      observations = nrow(rows),
      mean_time_points_per_patient = mean(counts),
      median_time_points_per_patient = stats::median(counts),
      first_time = min(rows$time),
      last_time = max(rows$time),
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, pieces)
}

check_demo_similarity <- function(source, synthetic, roles, time_bounds,
                                  label, clock = "study_time") {
  source_subjects <- length(unique(source[[roles$id]]))
  synthetic_subjects <- length(unique(synthetic[[roles$id]]))
  if (source_subjects != synthetic_subjects) {
    stop(label, ": source and synthetic patient counts differ (",
         source_subjects, " versus ", synthetic_subjects, ").")
  }
  source_summary <- demo_design_summary(
    source, roles, "Source", time_bounds, clock
  )
  synthetic_summary <- demo_design_summary(
    synthetic, roles, "Synthetic", time_bounds, clock
  )
  if (!setequal(source_summary$endpoint, synthetic_summary$endpoint)) {
    stop(label, ": source and synthetic endpoint sets differ.")
  }
  paired <- merge(
    source_summary, synthetic_summary, by = "endpoint",
    suffixes = c("_source", "_synthetic")
  )
  point_difference <- abs(
    paired$mean_time_points_per_patient_synthetic -
      paired$mean_time_points_per_patient_source
  )
  point_allowance <- pmax(
    1, 0.25 * paired$mean_time_points_per_patient_source
  )
  if (any(point_difference > point_allowance)) {
    stop(label, ": mean time points per patient differ materially for ",
         paste(paired$endpoint[point_difference > point_allowance],
               collapse = ", "), ".")
  }
  coverage_allowance <- 0.20 * diff(time_bounds)
  bad_coverage <-
    abs(paired$first_time_synthetic - paired$first_time_source) >
      coverage_allowance |
    abs(paired$last_time_synthetic - paired$last_time_source) >
      coverage_allowance
  if (any(bad_coverage)) {
    stop(label, ": source and synthetic time coverage differs materially for ",
         paste(paired$endpoint[bad_coverage], collapse = ", "), ".")
  }
  summary <- rbind(source_summary, synthetic_summary)
  summary$dataset <- factor(
    summary$dataset, levels = c("Source", "Synthetic")
  )
  summary[order(summary$dataset, summary$endpoint), , drop = FALSE]
}

comparison_facets <- function(data) {
  ggplot2::facet_grid(dataset ~ endpoint, scales = "free_y")
}

source_synthetic_preview <- function(source, synthetic, n = 6L) {
  columns <- intersect(names(source), names(synthetic))
  source_rows <- utils::head(source[, columns, drop = FALSE], n)
  synthetic_rows <- utils::head(synthetic[, columns, drop = FALSE], n)
  source_rows$.dataset <- "Source"
  synthetic_rows$.dataset <- "Synthetic"
  out <- rbind(source_rows, synthetic_rows)
  rownames(out) <- NULL
  out[, c(".dataset", columns), drop = FALSE]
}

# The full masking accounting for one run. `pmx_masking_report()` owns the row
# labels and the explanation next to each one, so this vignette and the study
# reports under `scripts_private/` cannot drift apart; only the caption is local.
masking_table <- function(source, roles, synthetic, label) {
  knitr::kable(
    as.data.frame(pmx_masking_report(synthetic, source, roles)),
    row.names = FALSE, align = c("l", "r", "l"),
    caption = paste0("Everything ", label,
                     "'s run removed, and what was left to build on.")
  )
}
```

</details>

## What this package is for

`synpmx` creates **synthetic pharmacometric data for model-workflow
exploration**: data to develop and debug cleaning, joins, reshaping, plotting,
control-file plumbing, and repeated-dose or longitudinal analysis code, so that
a pipeline runs unchanged against the real data later.

The package offers four generation modes, and the "Introduction to
synpmx" vignette applies all four to one dataset side by side:

- **AVATAR blending** [1, 2] (`synpmx_avatar()`) — build each synthetic subject
  from real subjects. No elicitation, no formal privacy guarantee.
- **Prior only** (`synpmx_prior()` on a public structural model) — read no data
  at all.
- **Calibration** (`synpmx_calibrated()`) — simulate from a public structural
  model whose magnitude is corrected by a small differentially private release.
- **Empirical** (`synpmx_empirical()`) — release a dense set of differentially
  private summaries and rebuild subjects from them.

"AVATAR" is a method name rather than an initialism, from the patient-centric
*avatarization* literature: the original method is due to Guillaudeux and
colleagues [2], and Destere and colleagues benchmark a modified AVATAR for
population PK [1]. This package implements an AVATAR-*inspired* adaptation for
longitudinal event tables, not published AVATAR software.

**Most of this vignette is about the default, AVATAR blending**, run over the
eight public datasets described below. For each generated subject it samples a
compatible source subject's event skeleton as a template, then fills the
covariates and endpoint trajectories with a distance-weighted blend of similar
subjects, plus subject and residual noise. The output has the source schema,
fresh identifiers, and the same cohort size. A late section runs the
model-based path on the same theophylline data, as `scripts/demo_nlmixr2data.R`
does for the first five.

AVATAR output is **not** anonymous and carries **no formal privacy guarantee**.
It is synthetic data that stays under the source data's access controls and
confidentiality obligations, in the same spirit as Novartis's `synadam` (which
resamples each column marginally from the data). Those obligations follow the
data rather than the machine, so working with it on a local workstation covered
by the same controls is a supported use. It is not appropriate for parameter
estimation, inference, model selection, or clinical decisions. When the
generated data would reach anyone the source data could not, use one of the
differentially private modes instead; the [privacy article](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html) works through
that decision and the tradeoff behind it.

## The datasets

Every dataset here is public package data from `nlmixr2data` or `xgxr`, both
`Suggests` of this package. They were not chosen for variety of drug or disease
— `synpmx` never sees the pharmacology — but for variety of *structure*, since
structure is the whole of what the algorithm reads and the whole of what can
expose a patient. "Endpoints" counts **observation** endpoints; a compartment or
`NAME` level used for dosing is not one.

| Dataset | Package | Patients | Rows | Endpoints | What it is here to exercise |
|---|---|---|---|---|---|
| `theo_md` | nlmixr2data | 12 | 348 | 1 | Seven Q24H oral doses with dense profiles on occasions 1 and 7 only. Dosed by weight, but the recorded mg/kg spans 3.1 to 5.9, so dose inference correctly declines — the public case for declaring `dose_covariate` |
| `warfarin` | nlmixr2data | 32 | 515 | 2 | Lower-case schema, one dose, PK (`cp`) and PD (`pca`) running on different time courses, factor covariates |
| `wbcSim` | nlmixr2data | 45 | 280 | 1 | Infusion start/stop pairs and a myelosuppression response with delayed nadir and recovery. Follow-up spreads from 0 to 4580 hours — the only dataset here where the anchor screen fires |
| `nimoData` | nlmixr2data | 12 | 441 | 1 | Ten roughly weekly infusions with declared `OCC`/`TAD` and **recorded actual dose times**. The worked failure: no two subjects were sampled close enough to share a grid visit |
| `mavoglurant` | nlmixr2data | 120 | 2678 | 1 | One- and two-period crossover, `TIME` resetting within `OCC`, an occasion-varying assigned dose, infusion rows. The largest cohort here |
| `pheno_sd` | nlmixr2data | 59 | 744 | 1 | **Real patients**, and the largest real cohort available: neonatal phenobarbital, individualised dosing, sparse irregular sampling, time-varying weight. The honest failure on the dose side |
| `case1_pkpd` | xgxr | 180 | 20820 | 2 | The only public dataset shaped like a study report — declared `NOMTIME`, six treatment arms, a `CENS` column, baseline weight. The case the coarsening design is aimed at |
| `mad` | xgxr | 60 | 4160 | 5 | Multiple ascending dose carrying ordinal, count and binary PD alongside continuous PD and PK. More endpoints than anything else available, and endpoint loss is invisible below three |

Two things about this set are worth knowing before reading any number.

**Half of it is real patient data and half is not.** `warfarin`, `nimoData`,
`mavoglurant` and `pheno_sd` are real studies; `wbcSim`, `case1_pkpd` and `mad`
are model-generated; `theo_md` is real on day 1 with days 2 to 7 simulated. That
distinction earns its place here for one reason: a simulated cohort is sampled on
a tidier schedule than a real one, and the two datasets where the masking fails
below — `nimoData` and `pheno_sd` — are both real, while the one that passes
perfectly is simulated. A result on simulated data is not evidence about a real
study.

**None of these datasets uses steady state, `ADDL`, or `II`.** Every dose is
written out as its own row, so the handling of compressed dose records is
exercised by the package's tests rather than by anything below.

## Shared workflow

Every example follows the same three steps:

1. declare column meanings with `pmx_roles()`;
2. synthesize with `synpmx_avatar()`;
3. validate structure with `validate_pmx()` and compare to the source.

Identifiers, schema, classes, and factor levels are restored on output. The
caller's random-number state is left untouched; reproducibility comes from the
`seed` argument.

## What the masking does, and what it costs

Enough happens by default that it is worth naming the six mechanisms before looking
at any data. The AVATAR algorithm article walks through each in detail. AVATAR blends every avatar's **measurements** from several real
donors, so no single patient's numbers survive. But the **event skeleton** — the
pattern of when a patient was dosed and sampled — is copied verbatim from one
real anchor, and blending never touches it. Six mechanisms act on that
skeleton, and every one of them works by *removing* something:

The article numbers them **M1 to M6**; they are listed here in the order they
run, which puts M6 between M3 and M4.

| | Mechanism | Acts on | Why |
|---|---|---|---|
| **M1** | Snap every time onto a shared visit grid, then add deviations back from a cohort-wide pool (`coarsen_time`) | **all times** — dose events *and* DV observations | As recorded, one patient's list of times is very often unique to them. A *pooled* deviation restores realism without giving the patient their own back |
| **M2** | Drop patients from the **anchor pool** (`screen`, `on_donor_shortfall`) | whole source patients | Follow-up or dose count over **2× the cohort's 90th percentile** would give any avatar a conspicuous skeleton; a route arm under `k + 1` has too few donors to blend with |
| **M3** | Redraw which visits an avatar attended (`min_pattern_share`) | **DV observations only** — dose events untouched | A combination of missed visits held by one patient is a fingerprint made of gaps |
| **M6** | Redraw where dosing stopped (automatic) | **dose events** — truncation only | Stopping at a depth nobody else used is a fingerprint. Dose times are never moved or invented, so this only ever truncates a real schedule earlier |
| **M4** | Blend covariates and DV from donors, capped (`k`, `max_donor_weight`) | measured values | No avatar's numbers come from one real patient |
| **M5** | Recompute the dose amount (automatic) | dose amounts, under **weight-based** (mg/kg) or BSA (mg/m²) dosing | A copied dose under mg/kg discloses the anchor's weight exactly |

`flag_identifiable_subjects()` runs after generation and is deliberately **not**
in this table: it is a plausibility check rather than a masking step, and
`synpmx_avatar()` does not run it for you. The algorithm article explains why.

### How you check whether it worked

Three measurements, because an avatar can point back at a real patient in three
independent ways. A study can pass one and fail another, so none substitutes for
the rest.

| Function | Asks | Run on | Target |
|---|---|---|---|
| `skeleton_uniqueness()` | Is anyone's **observation schedule** unique — the list of times they were sampled at? Separately, is anyone's **dosing** unique? | source, before generating | 0 unique |
| `flag_identifiable_subjects()` | Does anyone **stand out** on follow-up length, dose count, dose size, or peak DV? | synthetic output | nothing flagged |
| `compare_pmx_proximity()` | Did a synthetic patient's **covariates and DV trajectory together** land nearer a real patient than real patients lie to each other? | the pair, after | near 0.5 |

Sampling and dosing are scored separately and only the first is what "schedule"
means here: the observation time vector excludes dose rows entirely. Dosing
uniqueness is a different number, and **coarsening does not touch it** — a study
with weight-based dosing leaves patients unique on dose however good the visit
grid is.

The headline count is **one number split by cause**, and the split says what to
do next:

- **unique observation schedule** — patients whose list of observation times nobody
  else shares. Drive this to zero. It is always the sum of the next two.
  - **unique observation time** — sampled at a clock time nobody else was. *A grid
    can fix this*; that is what `coarsen_time` is for, and a nonzero count after
    coarsening means declaring `nominal_time` is the fix.
  - **unique set of visits** — every time shared, but this combination of
    attended and missed visits is theirs. *No grid can fix this* at any
    resolution. `min_pattern_share` addresses it during generation; otherwise
    remediate or accept.

None of this is free, and the costs differ sharply by dataset. So **each dataset
below ends with its own full accounting** — how many patients were excluded and
why, how many visit sets were lost and who held them, how many subjects
still hold a schedule nobody shares. The tables are deliberately verbose. Reading
one takes a minute; not reading it means shipping a dataset without knowing what
came out of it.

Two quantities are worth learning to read first, because they are the ones that
surprise people:

- **"Unique observation schedule"** — a patient whose list of observation times no
  other patient shares. An avatar anchored on such a patient wears a schedule
  belonging to exactly one real person, however carefully its concentrations
  were blended. The count is split by cause, because the two causes have
  different answers: a *unique observation time* is the grid's job and is fixed by
  declaring `nominal_time`, while a *unique set of visits* means visits are
  missing rather than moved, and no
  grid can fix it.
- **"Patterns discarded"** — a real pattern of missing visits or a dose
  interruption that
  will not appear in the output at all. That loss is the mechanism working, not
  a bug: it is what stops an avatar carrying a schedule traceable to one person.
  Whether the cost is acceptable is a per-study judgment, which is why every run
  reports it.

## Theophylline: repeated dosing, dose-relative PK

`theo_md` has 12 subjects on a repeated oral regimen with a single
concentration endpoint.

```{r theo, eval=has_nlmixr2data}
data("theo_md", package = "nlmixr2data")
theo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", covariates = "WT"
)
theo_synth <- synpmx_avatar(theo_md, theo_roles, seed = 303)
validate_pmx(theo_synth, theo_roles)$valid
knitr::kable(
  source_synthetic_preview(theo_md, theo_synth),
  caption = "Actual Theophylline rows and synthesized rows"
)
```

`compare_pmx_distributions()` summarizes the dependent variable per endpoint and
each baseline covariate, source against synthetic, so you can see at a glance
that the ranges line up. Expect them to be close in magnitude and shape, not
identical — AVATAR does not reproduce the source distribution exactly. Like
every source-derived diagnostic it is marked restricted and stays inside the
source data's access controls.

```{r theo-distributions, eval=has_nlmixr2data}
theo_dist <- compare_pmx_distributions(theo_md, theo_synth, theo_roles)
knitr::kable(theo_dist$endpoints, digits = 2,
             caption = "Concentration distribution, source versus synthetic")
knitr::kable(theo_dist$covariates_numeric, digits = 2,
             caption = "Baseline weight distribution, source versus synthetic")
```

### What the masking did to theophylline

```{r theo-masking, eval=has_nlmixr2data}
masking_table(theo_md, theo_roles, theo_synth, "theophylline")
```

**This is the easy case, and it is worth understanding why.** Nothing was
excluded: no subject is structurally extreme, and with one route there is no arm
below the donor floor, so all twelve subjects were available to anchor on and
twelve avatars were built.

The interesting row is the schedule pair. **All twelve patients start out with a
unique observation schedule** — their observation times are recorded at full
precision and no two patients were sampled at exactly the same moment, so every one of them
holds a list of times nobody else shares. After coarsening the count is **zero**. The inferred
grid found the shared visit structure underneath the recorded times and pulled
everyone onto it. That is the coarsening mechanism doing its entire job.

The visit-set rows follow from it. Before coarsening every subject holds its
own visit set — twelve for twelve subjects, since no two were sampled at
the same moments. Afterwards there are three, each shared by several subjects,
so **nothing had to be discarded** and no arrangement had to be invented. This is
the ideal case: coarsening removes the timing exposure, and visit-set sampling
then has a real pool to draw from and costs nothing.

**Dose is not recomputed here, and this is the case that shows why you should
say so explicitly.** Theophylline *is* dosed by weight, but the recorded
milligrams per kilogram range from 3.1 to 5.9 across subjects — 8 distinct
ratio levels for 11 distinct amounts. Detection deliberately fails closed on
that rather than impose a multiplier the study did not actually use, which is
the right call for an inference engine and the wrong answer for this study.
The run report says so rather than leaving a blank:

```{r theo-dose-note, eval=has_nlmixr2data}
attr(theo_synth, "pmx_settings")$dose_basis_note
```

Naming the covariate in `pmx_roles()` skips the inference and holds each dose
row's own ratio:

```{r theo-dose-declared, eval=has_nlmixr2data}
theo_roles_dosed <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  cmt = "CMT", covariates = "WT", dose_covariate = "WT"
)
theo_declared <- suppressWarnings(
  synpmx_avatar(theo_md, theo_roles_dosed, seed = 303)
)
per_kg <- function(x) {
  dosed <- x[x$EVID != 0, ]
  range(dosed$AMT / dosed$WT)
}
data.frame(
  dataset = c("source", "synthetic, inferred", "synthetic, declared"),
  rbind(per_kg(theo_md), per_kg(theo_synth), per_kg(theo_declared))
) |> setNames(c("dataset", "min mg/kg", "max mg/kg")) |>
  knitr::kable(digits = 2, caption = paste(
    "Dose per kilogram. Declared, each avatar carries its anchor's own ratio,",
    "so the synthetic cohort spans the source's range; inferred, the amount is",
    "the anchor's milligrams over the avatar's blended weight."))
```

Two costs of leaving it inferred, both visible above. The synthetic mg/kg no
longer matches the study's own dosing rule, and the copied milligrams still
encode one real patient's weight exactly.

`skeleton_uniqueness()` is what produced the "before" number. Read it directly
and it summarizes the cohort two ways: how many patients are unique and why,
and how crowded each schedule is. Every column of the returned data frame
counts *how many patients share this property, this patient included*, so `1`
always means "nobody else". By default it scores the recorded times as given;
`coarsen_time = TRUE` scores the shared visit grid `synpmx_avatar()` builds, so
running it both ways shows what coarsening removed.

```{r theo-uniqueness, eval=has_nlmixr2data}
skeleton_uniqueness(theo_md, theo_roles)
```

```{r theo-uniqueness-after, eval=has_nlmixr2data}
skeleton_uniqueness(theo_md, theo_roles, coarsen_time = TRUE)
```

`plot_pmx_schedule()` draws the same cohort: one row per patient, one mark per
event, with the visit grid underneath. A protocol grid reads as vertical
stripes and follow-up ending as a ragged right edge, so a count that looks
alarming and
one that is ordinary look completely different.

```{r theo-schedule-map, eval=has_nlmixr2data, fig.width=8, fig.height=7}
plot_pmx_schedule(theo_md, theo_roles)
```

For a per-subject check on the *generated* data rather than the source,
`flag_identifiable_subjects()` scores each synthetic subject for being a
structural outlier and `remediate_identifiable_subjects()` removes or shortens
the ones it flags. Those find subjects that are *extreme*;
`skeleton_uniqueness()` finds patients whose *schedule is unique*.

```{r theo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
theo_comparison <- rbind(
  observed_plot_data(theo_md, theo_roles, "Source"),
  observed_plot_data(theo_synth, theo_roles, "Synthetic")
)
ggplot2::ggplot(
  theo_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.35) +
  ggplot2::geom_point(alpha = 0.55, size = 0.8) +
  comparison_facets(theo_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "DV", colour = "Dataset",
    title = "Theophylline: source and synthetic data"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

A dose-relative display makes the coarse curve shape easier to see. AVATAR
preserves the rise-and-fall because it blends real subject profiles rather than
inventing a shape.

```{r theo-tad, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
theo_tad <- rbind(
  tad_plot_data(theo_md, theo_roles, "Source"),
  tad_plot_data(theo_synth, theo_roles, "Synthetic")
)
ggplot2::ggplot(
  theo_tad,
  ggplot2::aes(tad, dv, group = interaction(dataset, subject, occasion),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.18) +
  ggplot2::geom_point(alpha = 0.25, size = 0.7) +
  comparison_facets(theo_tad) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Time after dose (hours)", y = "DV", colour = "Dataset",
    title = "Theophylline: dose-relative profiles"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

## Warfarin: separate PK and PD endpoints

`warfarin` has a lower-case schema, one dose, and two endpoints (`cp` and
`pca`) with factor covariates. Both endpoints and all subjects are retained.

```{r warfarin, echo=FALSE, eval=has_nlmixr2data}
data("warfarin", package = "nlmixr2data")
warfarin_roles <- pmx_roles(
  id = "id", time = "time", dv = "dv", amt = "amt", evid = "evid",
  dvid = "dvid", covariates = c("wt", "age", "sex")
)
warfarin_synth <- synpmx_avatar(warfarin, warfarin_roles, seed = 404)
```

```{r warfarin-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
warfarin_comparison <- rbind(
  observed_plot_data(warfarin, warfarin_roles, "Source"),
  observed_plot_data(warfarin_synth, warfarin_roles, "Synthetic")
)
ggplot2::ggplot(
  warfarin_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.3) +
  ggplot2::geom_point(alpha = 0.5, size = 0.8) +
  comparison_facets(warfarin_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Time (hours)", y = "DV", colour = "Dataset",
    title = "Warfarin: PK (cp) and PD (pca) endpoints"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

### What the masking did to warfarin

```{r warfarin-masking, eval=has_nlmixr2data}
masking_table(warfarin, warfarin_roles, warfarin_synth, "warfarin")
```

**This is where the costs start to show.** Again nothing was excluded — all 32
subjects were available to anchor on — but the schedule rows tell a different
story from theophylline's.

**Twelve patients still have a unique schedule after coarsening, and not one of
them for a reason the grid can address.** All twelve share every one of their
observation times with somebody; what singles them out is *which* of those shared
visits they attended. Warfarin has subjects with 13, 14, 16, 17, 18, 19 and 25
observations, and the rarer counts pick those subjects out on their own. No grid
resolution helps, because the times are already shared.

**Four of fourteen visit sets were discarded, each held by one patient.**
Those four patients stay in the cohort and keep contributing measurements as
donors — only their distinctive absences stop being reproduced. Note that
patterns discarded and patients holding them are equal: at the default floor of
2, a pattern is discarded exactly when one patient holds it, so the two numbers
must agree. They diverge only at a floor of 3 or more.

**About a fifth of avatars had their missed visits placed fresh.** The *shape* of
the missingness —
how many visits were missed and whether the misses were terminal, contiguous, or
scattered — came from real patients who share it, but the specific visits missed
were placed fresh, and rejected and redrawn if the placement happened to land on
a pattern too rare to reuse. That is what those avatars get *instead of* one of
the four discarded sets; see the section on pattern strictness below.

**Dose was recomputed from weight.** Warfarin is dosed at 1.5 mg/kg to within
0.1%, tight enough for detection to fire, so each avatar's `amt` comes from its
own blended weight rather than from its anchor. Without this, a copied `amt`
would disclose the anchor's weight exactly.

## WBC: infusion, a delayed response, and structural screening

`wbcSim` has infusion start/stop events and a study-time white-blood-cell
response with a delayed decline, nadir, and recovery. A few subjects are
followed far longer than the rest.

A patient followed much longer than everyone else stands out, so a synthetic
copy built from them could point back to the real person. To lower that risk,
`synpmx_avatar()` by default does not build an avatar from such an extreme
subject, quietly leaving the very-long follow-ups out of the synthetic data
(`screen = TRUE`). This is an informal safeguard against singling someone out,
not a formal privacy guarantee; ordinary long follow-ups are kept, and
`screen = FALSE` keeps every subject.

```{r wbc, echo=FALSE, eval=has_nlmixr2data}
data("wbcSim", package = "nlmixr2data")
wbc_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID", cmt = "CMT"
)
wbc_synth <- suppressWarnings(synpmx_avatar(wbcSim, wbc_roles, seed = 505))
```

```{r wbc-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
wbc_comparison <- rbind(
  observed_plot_data(wbcSim, wbc_roles, "Source"),
  observed_plot_data(wbc_synth, wbc_roles, "Synthetic")
)
ggplot2::ggplot(
  wbc_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.3) +
  ggplot2::geom_point(alpha = 0.4, size = 0.7) +
  comparison_facets(wbc_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "WBC", colour = "Dataset",
    title = "WBC: delayed decline, nadir, and recovery"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

### What the masking did to wbcSim

```{r wbc-masking, eval=has_nlmixr2data}
masking_table(wbcSim, wbc_roles, wbc_synth, "wbcSim")
```

**This is the only public dataset where the anchor screen actually fires.** Two
of the 45 subjects are followed far enough beyond the rest to exceed twice the
cohort's 90th percentile, so no avatar is built on them. Read the next two rows
together: **43 anchors available, 45 avatars built.** The screen does not shrink
the cohort — it removes those two from the pool that avatars are *drawn from*,
and the other 43 are sampled with replacement to fill the same 45 slots. The two
screened subjects also remain donors, so their measurements still contribute.
What is lost is not their data but the possibility of an avatar wearing their
distinctive follow-up length.

**Seventeen patients have a unique schedule, split two and fifteen.** The two
with a unique observation time are the ones worth acting on: `wbcSim` carries no nominal-time column, so
the grid was inferred, and declaring a real one would likely clear them. The
fifteen with a unique set of attended visits are past the grid's reach.

**Five of 25 patterns were discarded.** A larger absolute number than warfarin's
four but a smaller share — 20% against 29% — because 45 subjects on a common
infusion schedule produce more repetition than 32 warfarin subjects do.

Dose is not recomputed: `wbcSim` has no baseline covariate for the amount to be
proportional to, so there is nothing to detect.

## NimoData: infusions, occasions, and a dose group

`nimoData` has ten approximately weekly infusions, declared OCC/TAD, and a
nominal dose group (`DOS`). The dose group is carried through with `keep`, which
copies it verbatim from the same subject that supplied the doses, so it stays
coherent with them. The redundant `WGT` column is simply left undeclared, so
AVATAR drops it.

```{r nimo, echo=FALSE, eval=has_nlmixr2data}
data("nimoData", package = "nlmixr2data")
nimo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
  covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
)
nimo_synth <- suppressWarnings(synpmx_avatar(nimoData, nimo_roles, seed = 606))
```

```{r nimo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
nimo_comparison <- rbind(
  tad_plot_data(nimoData, nimo_roles, "Source"),
  tad_plot_data(nimo_synth, nimo_roles, "Synthetic")
)
ggplot2::ggplot(
  nimo_comparison,
  ggplot2::aes(tad, dv, group = interaction(dataset, subject, occasion),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.2) +
  ggplot2::geom_point(alpha = 0.35, size = 0.7) +
  comparison_facets(nimo_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Time after dose (hours)", y = "DV", colour = "Dataset",
    title = "NimoData: dose-relative profiles across ten infusions"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

### What the masking did to nimoData

```{r nimo-masking, eval=has_nlmixr2data}
masking_table(nimoData, nimo_roles, nimo_synth, "nimoData")
```

**This is the failure case, and it is the most instructive table in the
vignette.** Read the schedule rows: **12 of 12 unique before coarsening, 12 of 12
unique after, every one of them on a unique observation time.** The grid achieved
*nothing*.

The reason is the study design. Ten roughly-weekly infusions per subject over a
long follow-up means no two subjects were ever dosed or sampled close enough
together for the inferred grid to merge them. Every subject is observed at
moments that are theirs alone, so every avatar carries a real visit schedule.

The visit-set rows show the same failure from the other side, and they are the
clearest illustration in the vignette of what the shape fallback does. Twelve
subjects hold twelve distinct visit sets; two of those are held by a single
patient and are discarded. Yet the row that gives the game away is the next one:
**every avatar's arrangement was invented rather than reused, 100%.** Not one
avatar could be given a real patient's set of attended visits, so all twelve got
a set built from a *shape* — how many visits were missed, and whether the misses
were terminal, contiguous, or scattered. The shape survives; which specific
visits each patient missed does not.

**And the dose row fails outright: all twelve avatars carry a dose schedule
nobody else shares**, because the ten infusions were recorded at actual times
rather than nominal ones, so every subject's dosing is theirs alone. Only 7 of
the 12 source regimens are represented at all.

All three failures have one cause and one fix, and `nimoData` gets that fix
demonstrated later in this vignette: constructing a nominal time from the
protocol grid takes the unique-schedule count from 12 to 5, the invented-
arrangement share from 100% to 17%, and the dose-schedule failures from 12 to 0.

A dataset that produces this table should not be shipped as-is. Either declare
or construct `nominal_time`, or treat the avatars as carrying real schedules.

## Mavoglurant: occasion-reset clock and a carried dose column

`mavoglurant` has one- and two-period profiles, a `TIME` axis that resets within
occasion, an occasion-varying `DOSE`, numeric-coded `SEX`, and infusion rows.
The reset clock validates within ID and occasion. AVATAR copies the whole event
template from one anchor, so `DOSE` stays coherent with the doses when carried
through with `keep`.

```{r mavo, echo=FALSE, eval=has_nlmixr2data}
data("mavoglurant", package = "nlmixr2data")
mavo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  cmt = "CMT", rate = "RATE", mdv = "MDV", occasion = "OCC",
  keep = "DOSE", covariates = c("AGE", "SEX", "WT", "HT")
)
mavo_synth <- suppressWarnings(synpmx_avatar(mavoglurant, mavo_roles,
                                              seed = 707))
```

```{r mavo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
mavo_comparison <- rbind(
  tad_plot_data(mavoglurant, mavo_roles, "Source"),
  tad_plot_data(mavo_synth, mavo_roles, "Synthetic")
)
mavo_plot <- ggplot2::ggplot(
  mavo_comparison,
  ggplot2::aes(tad, dv, group = interaction(dataset, subject, occasion),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.2) +
  ggplot2::geom_point(alpha = 0.35, size = 0.7) +
  comparison_facets(mavo_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Time after dose (hours)", y = "DV (log scale)", colour = "Dataset",
    title = "Mavoglurant: dose-relative profiles with occasion reset"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
# Mavoglurant concentrations span orders of magnitude, so a log y axis reads
# better. Use xgxr's log scale when available, otherwise a plain log10 scale.
mavo_plot + if (requireNamespace("xgxr", quietly = TRUE)) {
  xgxr::xgx_scale_y_log10()
} else {
  ggplot2::scale_y_log10()
}
```

### What the masking did to mavoglurant

```{r mavo-masking, eval=has_nlmixr2data}
masking_table(mavoglurant, mavo_roles, mavo_synth, "mavoglurant")
```

**The largest cohort here, and the one where the residual is largest in absolute
terms.** Nothing was excluded from the anchor pool. The grid did most of what it
could: of the 64 patients still unique afterwards, **eleven have a unique
observation time** and the other 53 share every observation time with somebody,
differing only in which of those visits they attended.

Those 53 are the largest absolute residual in the vignette, and no grid setting
moves them. This is what the "unique visit set" row is for: it separates a
problem you can fix by declaring `nominal_time` from one you can only address by
dropping, remediating, or accepting.

The pattern rows are the reassuring part. Mavoglurant has by far the most
distinct visit sets of any dataset here — **73** — and yet only **two** of them
are held by a single patient, so only those two are discarded and **no
arrangement had to be invented at all**: for every one of the 120 avatars, a
real visit set that several patients share was available to reuse. A large
number of distinct patterns is not by itself a problem; a large number of
*singleton* patterns is.

## case1_pkpd: a study report on its own nominal grid

`xgxr::case1_pkpd` is the only public dataset here shaped like a real study
report: 180 patients across six treatment arms, a declared `NOMTIME`, two
endpoints keyed by a character `NAME` column, a baseline weight, and a `CENS`
column. Everything the nlmixr2data five leave untested — a nominal grid, an arm
to stratify on, censoring — is present.

```{r case1, eval=has_xgxr}
case1_pkpd <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
case1_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB",
  keep = "STUDY"
)
case1_synth <- suppressWarnings(
  synpmx_avatar(case1_pkpd, case1_roles, seed = 808)
)
```

**Validate the source too.** Running `validate_pmx()` on the data you are about
to generate from is how a mis-declared role is caught before it becomes a
generation bug, and this dataset is the one where it fires. Declaring
`cens = "CENS"` is refused, because that study's `CENS` is meaningful only for
the PK endpoint — the PD effect is signed, so a "left-censored" PD row reports a
value above uncensored ones and the flag cannot mean what it means for PK:

```{r case1-cens, eval=has_xgxr}
case1_roles_cens <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME", cens = "CENS",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB"
)
validate_pmx(case1_pkpd, case1_roles_cens)$valid
```

Leaving `cens` undeclared is the right answer for this study, and it is what the
run above does.

```{r case1-plot, echo=FALSE, eval=has_xgxr && has_ggplot2}
case1_comparison <- rbind(
  observed_plot_data(case1_pkpd, case1_roles, "Source"),
  observed_plot_data(case1_synth, case1_roles, "Synthetic")
)
ggplot2::ggplot(
  case1_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.15) +
  ggplot2::geom_point(alpha = 0.25, size = 0.6) +
  comparison_facets(case1_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "LIDV", colour = "Dataset",
    title = "case1_pkpd: PK concentration and continuous PD, six arms"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

Six treatment arms, and an avatar never leaves the arm it was anchored in —
`TRTACT` and `DOSE` are declared as `strata`, so they are copied from the anchor
and condition the regimen rather than being sampled independently. The arm
*sizes* also match, which takes a second mechanism: anchors are sampled with
replacement, so `preserve_strata_balance` (the default) gives each stratum its
source share rather than leaving the balance to the draw.

```{r case1-arms, eval=has_xgxr}
knitr::kable(
  as.data.frame(rbind(
    Source = table(case1_pkpd$TRTACT[!duplicated(case1_pkpd$ID)]),
    Synthetic = table(case1_synth$TRTACT[!duplicated(case1_synth$ID)])
  )),
  caption = "Patients per treatment arm, source against synthetic"
)
```

### What the masking did to case1_pkpd

```{r case1-masking, eval=has_xgxr}
masking_table(case1_pkpd, case1_roles, case1_synth, "case1_pkpd")
```

**This is what a clean result looks like, and it is worth seeing once.** All 180
patients start with a unique observation schedule and **none** ends with one. The
difference from every dataset above is a single declared column: with `NOMTIME`
present the grid is `nominal` rather than inferred, so coarsening is reading the
protocol instead of guessing at it. 180 patients collapse onto **six** distinct
visit sets, none of them rare, so nothing was discarded and no arrangement had to
be invented. Both output guarantees are 0.

Read this next to `nimoData`, which is the same mechanism with nothing to work
from. The gap between them is not a tuning difference; it is whether the study
recorded a nominal time.

Dose is not recomputed: the five distinct amounts are the arm's assigned dose,
not a multiple of baseline weight, so detection correctly declines.

## mad: five endpoints, including ordinal, count, and binary

`xgxr::mad` is a multiple-ascending-dose study carrying five observation
endpoints — PK concentration, continuous PD, and ordinal, count, and binary PD —
keyed by `NAME`. Everything else here has one or two, and an endpoint silently
vanishing during generation is invisible with fewer than three.

```{r mad, eval=has_xgxr}
mad <- as.data.frame(get(utils::data(list = "mad", package = "xgxr")))
mad_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", mdv = "MDV", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"),
  covariates = c("WEIGHTB", "SEX")
)
mad_synth <- suppressWarnings(synpmx_avatar(mad, mad_roles, seed = 909))
```

Every endpoint in is an endpoint out. This is a set comparison, not a row count:
row counts stayed plausible in the defect that motivated the check while an
entire compartment disappeared.

```{r mad-endpoints, eval=has_xgxr}
setequal(unique(as.character(mad$NAME)), unique(as.character(mad_synth$NAME)))
```

```{r mad-plot, echo=FALSE, eval=has_xgxr && has_ggplot2, fig.height=6}
mad_comparison <- rbind(
  observed_plot_data(mad, mad_roles, "Source"),
  observed_plot_data(mad_synth, mad_roles, "Synthetic")
)
ggplot2::ggplot(
  mad_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.2) +
  ggplot2::geom_point(alpha = 0.3, size = 0.6) +
  comparison_facets(mad_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "LIDV", colour = "Dataset",
    title = "mad: five endpoints on one time axis"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom",
                 strip.text = ggplot2::element_text(size = 7))
```

The ordinal, count, and binary panels are the ones to look at. AVATAR blends
donor values and then restores the source column's class, so a count endpoint
comes back as counts and an ordinal one keeps its levels — it does not learn that
they are discrete, it never destroys the type in the first place.

### What the masking did to mad

```{r mad-masking, eval=has_xgxr}
masking_table(mad, mad_roles, mad_synth, "mad")
```

Another declared `NOMTIME`, and the same clean outcome: 60 of 60 unique before,
0 after, six visit sets, nothing discarded, no arrangement invented, both
guarantees at 0. Five endpoints do not make masking harder — they make *endpoint
loss* detectable, which is why this dataset is here.

## pheno_sd: real patients whose dosing cannot be masked

`nlmixr2data::pheno_sd` is 59 neonates given phenobarbital for seizure
prevention, and the data come from routine clinical care rather than from a
protocol. Dosing is individualised to the infant, sampling is sparse and
irregular, and weight varies over time. It is the least protocol-like dataset
here, and it is the one that fails.

```{r pheno, eval=has_nlmixr2data}
data("pheno_sd", package = "nlmixr2data")
pheno_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  covariates = c("WT", "APGR")
)
pheno_synth <- suppressWarnings(synpmx_avatar(pheno_sd, pheno_roles, seed = 1010))
```

```{r pheno-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
pheno_comparison <- rbind(
  observed_plot_data(pheno_sd, pheno_roles, "Source"),
  observed_plot_data(pheno_synth, pheno_roles, "Synthetic")
)
ggplot2::ggplot(
  pheno_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.3) +
  ggplot2::geom_point(alpha = 0.5, size = 0.8) +
  comparison_facets(pheno_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "Phenobarbital (ug/mL)", colour = "Dataset",
    title = "pheno_sd: sparse, irregular, individualised"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

### What the masking did to pheno_sd

```{r pheno-masking, eval=has_nlmixr2data}
masking_table(pheno_sd, pheno_roles, pheno_synth, "pheno_sd")
```

**Read the dose rows, not the observation rows.** On the observation side the
mechanism works: the visit-set guarantee holds at 0, so no avatar wears a set of
attended visits that belongs to one real infant. On the dose side it does not,
and the run says so rather than passing quietly — **15 avatars at this seed carry
a dose schedule nobody else in the cohort shares**, which is the row that is
supposed to be 0. In the source, 55 of the 59 infants have a dose schedule
nobody else has, so there is very little to anchor on safely.

The cause is not a setting. Dose events are copied from the anchor verbatim,
because moving a dose is not a protocol-valid edit the way moving a sample is.
When dosing is individualised — a neonate dosed to their own weight, on the day
their own clinician decided — most patients have a dose schedule nobody shares,
and there is no safe patient to anchor on instead. Coarsening cannot help: it
acts on the visit grid, and the exposure here is *which days this infant was
dosed*.

That is what an honest failure looks like, and it is why every run reports both
guarantees separately. A study with this shape either accepts that its avatars
carry real dosing schedules — defensible only inside the source data's own access
controls — or does not use AVATAR.

Dose is not recomputed here either, for the same reason as `theo_md`: the 54
distinct amounts are not a clean multiple of `WT` or `APGR`, so inference
declines rather than imposing a rule the study did not use.

## The model-based path on the same data

Everything above blends real subjects. The alternative is to fit a public model
to the data through an accounted release and simulate from it, which is what
`scripts/demo_nlmixr2data.R` does for all five datasets. Repeating it here for
theophylline shows how much more must be declared, and what is bought.

The empirical engine measures the trajectory shape from the data through noised
summaries. Every clipping range, contribution limit, and budget share is an
explicit public input. `synpmx_empirical()` (like `synpmx_calibrated()`)
refuses to run until `synpmx_enable_dp_engines()` has been called once in the
session, acknowledging the maintenance status covered in
the [privacy article](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html):

```{r enable-dp, eval=has_nlmixr2data}
synpmx_enable_dp_engines()
```

```{r empirical, eval=has_nlmixr2data}
theo_private <- synpmx_empirical(
  data = theo_md, roles = theo_roles,
  endpoints = list(cp = pmx_endpoint(
    alignment = "dose_relative", transform = "log", shape = "occasion", cmt = 2
  )),
  epsilon = 5, delta = 0,
  bounds = pmx_bounds(
    time = c(0, 170), endpoints = list(cp = c(0, 30)), amt = c(0, 500),
    covariates = list(WT = c(40, 130))
  ),
  public_design = pmx_public_design(
    pmx_schema(theo_md), dose_evid = 101, dose_cmt = 1
  ),
  contribution_limits = pmx_contribution_limits(40, 8, 8, 30, 11),
  budget_allocation = pmx_budget_allocation(
    subject_count = 0.10, event = 0.15, timing = 0.15,
    covariates = 0.10, endpoints = 0.50, censoring = 0
  ),
  seed = 707,
  backend = "public", public_source = TRUE   # theo_md is public; no DP claim
)
validate_pmx(theo_private, theo_roles)$valid
sampling_summary(theo_private)
```

The confidential data is read once, at the moment `synpmx_empirical()` is
called. The release it produced travels with the returned dataset, so any
number of further datasets can be drawn from it as post-processing, without
spending more budget:

```{r empirical-again, eval=has_nlmixr2data}
theo_private_2 <- synpmx_generate(theo_private, seed = 708)   # spends nothing
```

Reach for `synpmx_generate()` rather than calling `synpmx_empirical()` a second
time — a second call is a second release, and its budget has to be composed
with the first.

```{r empirical-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
theo_private_comparison <- rbind(
  observed_plot_data(theo_md, theo_roles, "Source"),
  observed_plot_data(theo_private, theo_roles, "Synthetic")
)
ggplot2::ggplot(
  theo_private_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.35) +
  ggplot2::geom_point(alpha = 0.55, size = 0.8) +
  comparison_facets(theo_private_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "DV", colour = "Dataset",
    title = "Theophylline: model-based private generation"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

Because `theo_md` is already public, this demonstration uses the guarded
public-fixture backend, which is **noiseless and makes no DP claim** — exactly
as the demo script does. A confidential fit uses the default OpenDP backend and
fails closed when it is unavailable. The `synpmx_calibrated()` mode, which
asserts curve shape from a public structural model and privately calibrates only
the magnitude, is the better choice at this cohort size; the introduction
vignette runs it on this same dataset.

## How well did the obfuscation work?

Everything above shows that the synthetic data *looks* right. This section asks
the other question: how much of each real patient is still visible in it?

### All eight side by side

Each dataset above carries its own accounting. This section puts the same
quantities in one place, because the *contrast* between datasets is what shows
how much the answer depends on study design rather than on the package.

```{r exposure-table, eval=has_nlmixr2data && has_xgxr}
exposure_row <- function(label, source, roles, synthetic) {
  before <- skeleton_uniqueness(source, roles)
  after <- attr(synthetic, "pmx_settings")
  data.frame(
    Dataset = label,
    Subjects = nrow(before),
    `Unique before` = attr(before, "n_unique_schedule"),
    `Unique after` = after$unique_schedule_n,
    `Unique observation time` = after$unique_obs_time_n,
    `Unique visit set` = after$unique_visit_set_n,
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

exposure <- rbind(
  exposure_row("theo_md", theo_md, theo_roles, theo_synth),
  exposure_row("warfarin", warfarin, warfarin_roles, warfarin_synth),
  exposure_row("wbcSim", wbcSim, wbc_roles, wbc_synth),
  exposure_row("nimoData", nimoData, nimo_roles, nimo_synth),
  exposure_row("mavoglurant", mavoglurant, mavo_roles, mavo_synth),
  exposure_row("pheno_sd", pheno_sd, pheno_roles, pheno_synth),
  exposure_row("case1_pkpd", case1_pkpd, case1_roles, case1_synth),
  exposure_row("mad", mad, mad_roles, mad_synth)
)
knitr::kable(
  exposure,
  caption = paste(
    "Source subjects holding a visit schedule no other subject shares,",
    "before and after time coarsening. The last two columns split the",
    "'after' count by cause."
  )
)
```

Three datasets reach zero and five do not, and the split is not about cohort
size. `case1_pkpd` (180 patients) and `mad` (60) reach it because they declare
`NOMTIME`, so coarsening reads the protocol grid instead of inferring one;
`theo_md` reaches it on an inferred grid because twelve subjects on one dense
protocol leave an obvious grid to find. Everything else is a study whose
recorded times do not collapse, and `nimoData` and `pheno_sd` are the extreme
cases — real studies where the times are genuinely per-patient.

### And what the masking cost

Exposure is only half the story. The mechanisms that reduce it do so by
*removing* things, and the removals are worth seeing next to the numbers above.

```{r exposure-cost, eval=has_nlmixr2data && has_xgxr}
cost_row <- function(label, source, roles, synthetic) {
  settings <- attr(synthetic, "pmx_settings")
  data.frame(
    Dataset = label,
    Patients = settings$source_subjects,
    `Screened out` = settings$anchors_screened_out,
    `Below donor floor` = settings$anchors_route_excluded,
    `Anchors left` = settings$anchors_available,
    `Patterns` = settings$patterns_total,
    `Patterns lost` = settings$patterns_dropped,
    `Patients affected` = settings$subjects_with_dropped_pattern,
    `Dose basis` = ifelse(is.na(settings$dose_basis), "—", settings$dose_basis),
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

knitr::kable(
  rbind(
    cost_row("theo_md", theo_md, theo_roles, theo_synth),
    cost_row("warfarin", warfarin, warfarin_roles, warfarin_synth),
    cost_row("wbcSim", wbcSim, wbc_roles, wbc_synth),
    cost_row("nimoData", nimoData, nimo_roles, nimo_synth),
    cost_row("mavoglurant", mavoglurant, mavo_roles, mavo_synth),
    cost_row("pheno_sd", pheno_sd, pheno_roles, pheno_synth),
    cost_row("case1_pkpd", case1_pkpd, case1_roles, case1_synth),
    cost_row("mad", mad, mad_roles, mad_synth)
  ),
  caption = paste(
    "What the masking removed. The first three counts are patients excluded",
    "from the anchor pool; `Patterns` counts distinct visit sets in",
    "the source and `Patterns lost` those too rare to be reused."
  )
)
```

Three things in this table look wrong at first and are not.

**Only `wbcSim` loses any patient from the anchor pool, and the cohort is still
full size.** Screening removes a subject from the pool avatars are *drawn from*,
not from the cohort: the remaining anchors are sampled with replacement to fill
every slot, and the screened subjects still contribute measurements as donors.
An exclusion costs coverage of a structure, never sample size.

**`Patterns lost` equals `Patients affected` in every row.** That is an identity,
not a coincidence. At the default floor of 2, a pattern is discarded exactly when
*fewer than two* patients share it — that is, when exactly one does. So every
discarded pattern has precisely one holder and the two columns must agree. They
only diverge at a floor of 3 or more, where a pattern held by two patients is
also dropped.

**`Patterns lost` says nothing about how well a dataset did.** `mavoglurant`
loses 2 of 73 and `nimoData` loses 2 of 12, and they are not comparable results:
mavoglurant reused a real visit set for every avatar, while nimoData could reuse
none and invented all twelve from a shape. The discard count has to be read next
to `pattern_generated_fraction`, which the per-dataset tables above print and
this one does not.

### Is a "pattern" too strictly defined?

Reasonably asked, and the answer is yes, deliberately. A pattern here is the
**exact** set of endpoint-and-time pairs a subject was observed at. Two patients
who each missed exactly one visit have *different* patterns if they missed
different visits. Every pattern held by exactly one patient is discarded, and how
many that is depends entirely on the study: 4 of warfarin's 14 after coarsening,
5 of wbcSim's 25, but only 2 of mavoglurant's 73.

The strictness is what makes the guarantee exact: reusing a pattern that one real
patient holds would reproduce that patient's schedule, so nothing short of exact
matching would support the claim. The cost is that the definition cannot see that
"missed one visit, early" and "missed one visit, late" are the same *kind* of
event.

That is why the draw is two-stage. A **shape** is chosen first — how many visits
were missed, and whether the misses were terminal (at the end of the record),
contiguous (an
interruption), or scattered — and both of those patients share one. Within the
shape a real pattern is reused if one clears the floor; only otherwise is an
arrangement generated, and a generated one is rejected and redrawn if it happens
to land on a pattern too rare to have been reusable.

What the fallback rescues is the *avatar*, not the pattern count. A discarded
pattern stays discarded; what the shape gives is a legal set of visits to hand
the avatar that would otherwise have worn it. The per-dataset tables show the
price in the "misses placed fresh" row, and it varies as much as everything else
here: `nimoData` pays it in full at 100%, `pheno_sd` at 41%, `warfarin` at 22%,
`wbcSim` at 2%, and `mavoglurant`, `theo_md`, `case1_pkpd` and `mad` not at all.

What remains lost is *resolution*: how much missingness there was and what kind
survive; which specific visits each patient missed does not.

### And how close the values landed

Everything above concerns *structure*. Blending is the mechanism that protects
the **values**, and `compare_pmx_proximity()` is its measurement: it asks whether
each subject's nearest neighbour lies in its own dataset or the other one. Near
0.5 means a synthetic subject is no more like a real subject than one real
subject is like another, which is the target; toward 0 means memorisation.

```{r proximity, eval=has_nlmixr2data && has_xgxr}
proximity_row <- function(label, source, synthetic, roles) {
  report <- compare_pmx_proximity(source, synthetic, roles, replicates = 30)
  data.frame(
    Dataset = label,
    `Adversarial accuracy` = round(report$adversarial_accuracy, 3),
    `Null lower` = round(report$null_lower, 3),
    `Null upper` = round(report$null_upper, 3),
    `Per side` = report$n_compared,
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

knitr::kable(
  rbind(
    proximity_row("theo_md", theo_md, theo_synth, theo_roles),
    proximity_row("warfarin", warfarin, warfarin_synth, warfarin_roles),
    proximity_row("wbcSim", wbcSim, wbc_synth, wbc_roles),
    proximity_row("mavoglurant", mavoglurant, mavo_synth, mavo_roles),
    proximity_row("pheno_sd", pheno_sd, pheno_synth, pheno_roles),
    proximity_row("case1_pkpd", case1_pkpd, case1_synth, case1_roles),
    proximity_row("mad", mad, mad_synth, mad_roles)
  ),
  caption = paste(
    "Nearest-neighbour adversarial accuracy against a split-half null built",
    "from the source cohort itself. 0.5 is the target."
  )
)
```

Read the null intervals before the point estimates. They are **wide** — these
are cohorts of a few dozen, and the statistic is built from nearest-neighbour
comparisons that are noisy at that size. A value inside the interval means
*nothing was detected*, not that nothing is there. What it would catch is a
blatant leak: a synthetic subject sitting on top of a real one drives the
statistic to zero, which the package's own regression test confirms by handing
the function a verbatim copy and requiring it to object.

### Reading the tables across datasets

**The unique-schedule count after coarsening is the number of patients you would
consider dropping**, and the split by cause decides whether dropping is even the
right response. Each dataset
section above works through its own numbers; the pattern across all eight is what
matters here.

The spread runs from **`case1_pkpd`, where 180 patients on a declared nominal
grid go to 0 and the mechanism has nothing left to do**, through **`mavoglurant`,
where the grid does most of what it can and still leaves 53 of 120 patients
unique on their visit set**, to **`nimoData`, where the grid achieves nothing at all**,
and finally to **`pheno_sd`, where the observation side is clean and the *dosing*
cannot be masked at all**. Same package, same defaults, same seed discipline —
the difference is entirely in how the studies were designed and how their times
were recorded.

That is the point of reporting these per dataset rather than quoting a headline
number. There is no "synpmx removes X% of the exposure." There is only what it
removed from *your* study, which is why every run reports it and why the sequence
below starts with a question about your data rather than about a setting.

### What the other mechanisms leave behind

One residual belongs to a mechanism these tables do not have a column for.

**Dose sequence.** The dose *amount* is handled: under dosing proportional to a
baseline covariate, each avatar's amount is recomputed from its own blended
covariate, as the per-dataset tables report. What is still copied is the
*sequence of levels* an anchor climbed. Where escalation is outcome-adaptive —
driven by a subject's own tolerability — that sequence encodes the subject's
response, and nothing here touches it. No public dataset in this vignette has
that shape, so it does not appear in any table above; a dose-escalation oncology
study would.

**Which visits were attended.** The unique-visit-set row above is what
`min_pattern_share` (default 2) addresses: an avatar's set of attended visits is
drawn from patterns at least that many source subjects share, so no synthetic
patient carries a schedule unique to a real one. No subject is removed from the
cohort to achieve it — a patient with a rare pattern still contributes
measurements as a donor.

What *is* lost is the rare patterns themselves. They are discarded rather than
approximated, so those specific patterns of missing visits and dose
interruptions will not
appear in the output. The loss is a handful of patterns on every dataset here —
two each on `nimoData` and `mavoglurant`, three on `pheno_sd`, four on
`warfarin`, five on `wbcSim` — and it is zero wherever coarsening has already
made the patterns common, on `theo_md` from an inferred grid and on `case1_pkpd`
and `mad` from a declared one.

Every run reports the figures, and the per-dataset tables above print them:
`patterns_total`, `patterns_dropped`, and `subjects_with_dropped_pattern` in the
settings, plus a loud alert. Read them together with
`pattern_generated_fraction`, because a small discard count can be bought with a
large share of invented arrangements — `nimoData` discards only two patterns and
invents the arrangement for every single avatar.

### What to do about it

Dropping 53 of 120 subjects is a real cost, and it is not automatic. Under this
package's governance model — AVATAR output stays inside the source data's own
access controls — a unique schedule is a reason for care, not a blocker. The
sequence worth following is:

1. **Declare `nominal_time` if the study has it.** This is free, and it is the
   only thing that helps the unique-moment column. On `nimoData` it is the
   difference between the mechanism working and not working at all.

   A study that has no nominal-time column can often still *construct* one, and
   `nimoData` is a good example because its inferred grid fails completely. Its
   design is ten roughly-weekly infusions with a declared occasion and
   time-after-dose, so the protocol grid is the occasion number times the
   nominal interval, plus the visit's nominal time after dose:

    ```{r nimo-nominal, eval=has_nlmixr2data}
    nimo_nominal <- nimoData
    nimo_nominal$NTIME <- (nimo_nominal$OCC - 1) * 168 +
      ifelse(nimo_nominal$EVID == 0, round(nimo_nominal$TAD / 24) * 24, 0)

    nimo_roles_nominal <- pmx_roles(
      id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
      rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
      nominal_time = "NTIME",
      covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
    )
    nimo_fixed <- suppressWarnings(
      synpmx_avatar(nimo_nominal, nimo_roles_nominal, seed = 606)
    )
    unlist(attr(nimo_fixed, "pmx_settings")[
      c("time_grid", "unique_schedule_n", "unique_obs_time_n")
    ])
    ```

    Unique schedules fall from 12 of 12 to 5, all in the long washout tail after
    the last dose, where a follow-up visit at 240 or 624 hours belongs to one
    patient. Coarsening that tail further does not help — at a two-week grid the
    count goes back up, because merging genuinely distinct visits creates new
    *pattern* uniqueness.

    Visit-set sampling improves just as much, though the discard count does not
    show it. With the inferred grid no two subjects shared a pattern, so every
    avatar's arrangement had to be invented from a shape; on the nominal grid the
    twelve patterns collapse to eight, real ones become reusable, and the
    invented share falls from **100% to 17%**. Two patterns are discarded either
    way. This is why the discard count should never be read alone.
2. **Then re-run this table.** If the unique-moment count is at or near zero,
   the grid has done its whole job and no amount of tuning will improve it.
3. **Then decide about the pattern column.** These are missing-visit
   patterns. `flag_identifiable_subjects()` and
   `remediate_identifiable_subjects()` can drop or truncate them; whether that
   is worth the lost cohort size is a judgment about who will see the output.

Two limits to keep in view. This counts *schedules*, not dose amounts — a study
with weight-based dosing or per-subject titration leaves subjects unique on dose
regardless of what the grid does, which `skeleton_uniqueness()` reports
separately as `n_share_dosing`. And none of this bounds what an adversary
learns; it reduces the ways a real patient can be singled out without limiting
how often that succeeds. That is what the differentially private modes are for.

## What is preserved in AVATAR, and what is not

Preserved: schema, column classes, factor levels, cohort size, endpoint set,
event structure, coarse regimen and sampling timing, and the broad shape and
magnitude of each endpoint.

Not preserved, by design: exact source distributions, parameter estimates,
covariate-response relationships, and rare individual trajectories. Identifiers
are always freshly generated and never reuse a source value.

Not provided: any formal privacy guarantee. The synthetic data is built by
blending real subject trajectories, so it is appropriate wherever the source
data and synthetic data are accessible by the same users, but — but not for release to anyone outside. 

## References

1. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).

2. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).
