---
title: "Scorecard: Checks of the synthetic data"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Scorecard: Checks of the synthetic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
both <- has_nlmixr2data && has_xgxr
# `compare_pmx_distributions()` composes its panels with patchwork and refuses
# to draw without it, so a missing suggested package is a missing figure rather
# than a failed build.
has_figure <- has_xgxr && requireNamespace("ggplot2", quietly = TRUE) &&
  requireNamespace("patchwork", quietly = TRUE)
```

You have generated a synthetic dataset. Should you use it?  The scorecard, generated by `synpmx_scorecard()` helps assess this and this Vignette explains it.

## The scorecard

The scorecard contains a table of the following items:

* The check of the dataset that is performed.
* That data is read (source data, synthetic data, or both)
* The resulting calculation from the check
* A verdict for whether the check passed, failed or needs review
* An `explore` function to help assess any checks that failed or need review.

Its rows are grouped into four categories, one per question:

* **A. Is the dataset valid**
* **B. Is any individual patient identifiable based on the synthetic data**
* **C. Has the synthetic study design significantly changed from the source**
* **D. How much have the covariate and observation distributions changed**

Every row the card holds has a section of this vignette under the same
identifier.

**Seven checks can say `FAIL`, and no others.** A1, because the output is not a
legal dataset; A3 and A6, because it is not the study that went in; and B1a,
B1b, B4a and B4b, because each means one real patient's structure was
reproduced verbatim. No other row can fail. The rest answer `pass` when there is
nothing to read and `review` when there is something whose meaning depends on
the study: a subject dropped for want of donors, an arm that changed size, a
statistic wandering at a small sample size. One row, D1, is `review` however it
lands, because no threshold on it would be honest.

## The two datasets used here

```{r data, eval=both, message = FALSE}
mad <- as.data.frame(get(utils::data(list = "mad", package = "xgxr")))
mad_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", mdv = "MDV", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"),
  covariates = c("WEIGHTB", "SEX")
)
mad_synth <- suppressWarnings(synpmx_avatar(mad, mad_roles, seed = 909))

data("pheno_sd", package = "nlmixr2data")
pheno_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  covariates = c("WT", "APGR")
)
pheno_synth <- suppressWarnings(
  synpmx_avatar(pheno_sd, pheno_roles, seed = 1010)
)
```

`xgxr::mad` is the worked example: 60 subjects in a multiple-ascending-dose
study, six arms of ten (placebo and five dose levels), a declared nominal time,
a baseline weight and a sex, and **five** endpoints keyed by `NAME` — a
pharmacokinetic (PK) concentration, a continuous pharmacodynamic (PD) effect,
and PD endpoints that are ordinal, count and binary. It is shaped like a real
study report, and most of these checks pass on it.

Five endpoints and a categorical covariate are why it is the example here. On a
study with one continuous endpoint and no category, A6 and B5 both report
that there was nothing to ask, A3 compares a set of size one against itself, and
C3 asks each arm about one endpoint.

`nlmixr2data::pheno_sd` is 59 real neonates given phenobarbital in routine
clinical care, and it is here because **no check on it fails while its dosing
history is visibly shortened**: the median infant received twelve doses and the
median avatar receives five.

## The scorecard, computed

```{r cards, eval=both}
mad_card <- synpmx_scorecard(mad, mad_synth, mad_roles)
synpmx_scorecard_datatable(mad_card)
```

Nothing fails. The rest of this vignette is what each check asks and why its
pass criterion is what it is, using `mad` where a check passes and `pheno_sd`
for further exploration.

Each section below opens with its own rows of this card, so the check being
discussed is on the screen with the discussion. The card is a `data.frame` and
that is all those tables are — a row subset of it, in the form shown under A5a
and A5b.

# A. Is the dataset valid

## A1, A2. Are the synthetic and source data legal PMX datasets

```{r a1a2-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% c("A1", "A2"), ])
```

A1 runs `validate_pmx()` on the synthetic table and A2 runs the same function
on the source. Both must be `TRUE`. The check is structural: schema and column
classes, event grammar, time monotonicity within subject, censoring flags
against their dependent variable (`DV`), infusion start/stop pairing, baseline
covariates that are actually constant, and strata that do not vary within a
subject.

```{r a-validate, eval=both}
validate_pmx(mad_synth, mad_roles)$valid
```

```{r a-source, eval=has_xgxr}
mad_roles_cens <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", mdv = "MDV", nominal_time = "NOMTIME",
  cens = "CENS", strata = c("TRTACT", "DOSE"),
  covariates = c("WEIGHTB", "SEX")
)
validate_pmx(mad, mad_roles_cens)$valid
```

That study's `CENS` column is meaningful only for the PK concentration, and all
five endpoints share the one `LIDV` column, so the flag lands on the four PD
endpoints as well. None of them is a concentration with a lower assay limit: on
the ordinal endpoint a row flagged as below the limit of quantification (BLOQ)
reports a severity of 2 while uncensored rows go down to 1, and the continuous,
count and binary endpoints each do the same. Declaring `cens` here would produce
a coherent-looking dataset with nonsense censoring on four endpoints out of
five; leave it undeclared, or restrict the flag to the rows it describes. A2 is
`review` rather than `pass` in that situation: a real study a validator objects
to is a normal thing to have, and the objection has to be read rather than
scored.

### Time after dose (TAD)

**`tad` is an output, not an input.** `synpmx_avatar()` recomputes it from the
generated times and dose rows and overwrites whatever the source held.
Declaring the role says which column to overwrite and carry through. The source
values are read in one place only, `validate_pmx()` on the source:

## A3. Did every endpoint survive

```{r a3-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "A3", ])
```

A3 compares the set of `dvid` levels observed in the source against the set
observed in the output. They must be equal, and a missing endpoint is a `FAIL`:
it is not the study that went in.

```{r a-endpoints, eval=has_xgxr}
setequal(
  unique(as.character(mad$NAME)),
  unique(as.character(mad_synth$NAME))
)
```

It is a set comparison rather than a row count, because row counts stay
plausible while a whole endpoint disappears. That is also why five endpoints are
worth having in the example: A3 reads `5 of 5` here, and on a one-endpoint study
the check can only ever say that the one endpoint is still there.

## A4. Did the strata (cohort) size survive

```{r a4-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "A4", ])
```

A4 counts distinct subjects on strata. Equal is a pass. Fewer is `review`
rather than `FAIL`, because `on_donor_shortfall = "drop"` removes a subject
that could not be built from enough donors, and declining to build on a patient
the generator cannot protect is the correct answer. It lands on whichever
stratum was thinnest, so a small cohort is where it shows up: a stratum holding
a single patient can legitimately come back empty.

## A5a, A5b. Did the number of observations and doses survive

```{r a5-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% c("A5a", "A5b"), ])
```

A5a and A5b are rows per patient, split by event type: observations in one, dose
events in the other. Each passes when the synthetic count lands within 5% of the
source's — that much movement leaves each patient carrying the same amount of
information, and there is nothing to decide. Further than that is `review` and
never `FAIL`, because the number is permitted to move: a shortened dose course
can be the correct answer, as the rest of this section shows.

On `mad` neither moves — 61.7 observations and 5 doses per patient on both
sides, a fixed protocol grid coming through intact, so both pass. `pheno_sd` is
where they move, and the card is a `data.frame`, so a section of it is an
ordinary row subset:

```{r pheno-card, eval=both}
pheno_card <- synpmx_scorecard(pheno_sd, pheno_synth, pheno_roles)
synpmx_scorecard_datatable(pheno_card[pheno_card$check %in% c("A5a", "A5b"), ])
```

Observations are near-intact at 2.63 against 2.51 per patient — inside the 5%,
so A5a passes — while dosing falls from 9.98 to 5.63 and A5b asks to be read. This
occurs from `synpmx_avatar()` trying to infer a nominal time grid.

```{r a-doses, eval=has_nlmixr2data}
doses_per_patient <- function(data) {
  as.numeric(table(data$ID[data$EVID != 0]))
}
summary(doses_per_patient(pheno_sd))
summary(doses_per_patient(pheno_synth))
```

The median real infant received twelve doses. The median avatar receives five.
`plot_pmx_schedule()` draws it, one row per patient, a grey tick for each dose
and a coloured dot for each observation:

```{r a-schedule, eval=has_nlmixr2data, fig.height=7}
plot_pmx_schedule(pheno_sd, pheno_roles, main = "pheno_sd, source")
plot_pmx_schedule(pheno_synth, pheno_roles, main = "pheno_sd, synthetic")
```

Nothing is invalid here.  No infant's *complete* schedule can be reused — 55 of 59 are held by one patient
— but most infants open the same way, twelve-hourly from time zero, so each
avatar's course is cut back to the deepest opening several infants share. That
is about half the doses. Whether half a course is enough is exactly the
judgement A5b exists to prompt, and it depends on what the dataset is for:
enough to develop a fitting workflow, not enough to characterise the tail of an
individualised regimen. Section B1 is the other half of the story, where the
privacy checks read 0, and A5b is what says at what price.

If it were important to more accurately capture the dosing information, a nominal time grid could be specified by the user, rather than computed by `synpmx_avatar()`, as shown in [Evaluating AVATAR on public data](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html) for the `nimoData` example.

## A6. Did a discrete endpoint stay discrete

```{r a6-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "A6", ])
```

A6 asks whether every generated value on a binary, ordinal or count endpoint is
one the source could have held. It is on every card. Where a study has no
discrete endpoint the result reads `no discrete endpoint` and passes, which is
the true answer rather than a silently absent row.

Blending is a weighted mean, so a weighted mean of several patients' zeros and
ones is a number between them, and a 0/1 endpoint comes back continuous unless
the generated values are put back on the scale the source used. The subject and
residual noise terms then carry it off the level set entirely. Ask what kind of
values each endpoint takes before reading anything else about it:

```{r a-endpoint-types, eval=has_xgxr}
pmx_endpoint_types(mad, mad_roles)
```

Three of `mad`'s five endpoints are discrete, and `LIDV` is one numeric column
carrying all five, so restoring the *column's* class restores nothing about any
of them. `synpmx_avatar()` decides per endpoint and snaps the generated values
back: binary and ordinal onto the source's levels, count to whole numbers. A6
reads the finished table rather than trusting that:

```{r a-discrete, eval=has_xgxr}
discrete <- c("PD - Binary", "PD - Count", "PD - Ordinal")
values_taken <- function(data, endpoint) {
  values <- data$LIDV[data$NAME == endpoint & data$EVID == 0]
  sprintf("%d distinct, %.2f to %.2f", length(unique(values)),
          min(values), max(values))
}
knitr::kable(data.frame(
  endpoint = discrete,
  source = vapply(discrete, values_taken, character(1), data = mad),
  synthetic = vapply(discrete, values_taken, character(1), data = mad_synth),
  row.names = NULL
))
```

**A6 asks whether every generated value is one the source could have held, not
whether every source value came back.** The count endpoint loses its four
highest levels — 20 distinct values become 16, topping out at 15 rather than 19
— and passes, because 16 whole numbers inside the source's range are all values
this endpoint takes. A single generated 2.4 would fail the check. The lost tail is a
D1 question about the distribution.

The observation type is inferred from the source — whether every observed value is a
whole number, and how many distinct ones there are — and
`pmx_roles(endpoint_types = )` overrides it where that inference reads the
study wrongly. 

# B. Is any individual patient identifiable based on the synthetic data

## B1a, B1b. Does any avatar carry one real patient's schedule

```{r b1-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% c("B1a", "B1b"), ])
```

Who was observed when, and who was dosed when, is the axis most specific to
longitudinal data, and the one a general-purpose synthetic-data tool will not
have addressed. Two patterns per avatar are checked:

- **B1a**, the set of visits it attends.
- **B1b**, the times it is dosed at.

Either is a `FAIL` above **0**. An avatar may carry a pattern only if at least
`min_pattern_share` real patients shared it — 2 by default, set in
`synpmx_avatar()`, and the same floor B4a, B4b and B5 use. One avatar carrying
a pattern only one real patient had could be identifying.

The run measures both as it builds and records the answer, so the card reads
them from the run rather than from the finished table:

```{r b1-guarantees, eval=both}
guarantees <- function(synthetic, label) {
  settings <- attr(synthetic, "pmx_settings")
  data.frame(
    dataset = label,
    identifying_visit_sets = settings$identifying_visit_sets,
    identifying_dose_schedules = settings$identifying_dose_schedules
  )
}
knitr::kable(rbind(
  guarantees(mad_synth, "mad"),
  guarantees(pheno_synth, "pheno_sd")
))
```

**Passing is not the same as being fine**, which is what `pheno_sd` is here to
show. Its infants are dosed individually, so 55 of 59 have a dose schedule
nobody else shares. A dose cannot be moved without inventing a regimen the
protocol never allowed, so the only way to 0 was to stop each avatar's dosing
early, at a depth several infants passed through — 10 doses per patient down to
5.6. **Read B1a and B1b next to A5b**, which is where that price is reported.

Where `strata` are declared the two are decided arm by arm, since an avatar is
never moved out of the arm it was anchored in, and the result cell names the arm
that failed, with an example shown below. 

```
B1b   Avatars with a dose schedule nobody else shares   6 in ARM8 (6)   FAIL
```

### Asking before you generate

Both checks can be further explored. `unmaskable_strata(source, roles)`
answers the related question in advance, naming the arms whose patients no
method could mask; an arm with `safe_anchors = 0` fails on every seed.
`skeleton_uniqueness()` is the cohort-wide version, read on the coarsened visit
grid.

```{r b1-uniqueness, eval=has_xgxr}
skeleton_uniqueness(mad, mad_roles, coarsen_time = TRUE)
```

Every row of the above tables are zero.  The dataset `mad` declares a `nominal_time`, so subjects snap onto the
protocol's own grid and two schedules cover the cohort — one for the 50 subjects
who contribute PK, one for the ten placebo subjects who do not. Where a row is
not 0, the screen's own tables say which endpoint drives it, and
`nearest_set_diff` in the returned data frame says how far each patient is from
their nearest neighbour, since one missing sample counts the same as an ad-hoc
schedule in the headline count.

## B2. Does any synthetic patient stand out from its own stratum

```{r b2-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "B2", ])
```

B2 is about **singling out** a synthetic patient based on a record so conspicuous in the
released table that it is the one that might help identify an outlying patient. Nothing else in
section B asks it — B1 and B4 ask whether a real pattern was reproduced exactly,
B3 asks an aggregate question — and B2 is also one of only two rows that read
the synthetic table alone, so whoever receives the data can run it without the
source.

Each patient is screened on four axes: follow-up length, dose count, dose
magnitude and peak `DV`. A patient is flagged on an axis when **both** hold:

1. It is a robust outlier — a modified z above 3.5, the Iglewicz–Hoaglin
   cutoff, measuring its distance from its stratum's median in units of that
   stratum's median absolute deviation.
2. It is **materially alone**: at least `min_relative_gap` of the stratum's
   median away from the nearest other patient. The default is 1, a separation
   larger than the group's whole median value. It is set against ordinary
   between-subject variability rather than against any one cohort — a 30–50%
   coefficient of variation is unremarkable on a PK parameter, so anything
   smaller is inside the study's own noise and could not be picked out of it.

```{r b2-flag, eval=has_xgxr}
sum(flag_identifiable_subjects(mad_synth, mad_roles)$flagged)
sum(flag_identifiable_subjects(mad_synth, mad_roles,
                               min_relative_gap = 0)$flagged)
```

**What to look for.** A synthetic patient followed three times as long as anybody else, a
patient given twice their arm's dose, a patient with Cmax well above everyone else's in that cohort.  
The row is a list of records to read so 0 is a `pass`.  Anything above 0 is `review` and never a `FAIL`,
because the right count is not necessarily 0 — a study can hold a patient who is
genuinely unusual, and an anchor-based generator will hand that patient's shape
to an avatar.

**What a hit looks like.** No synthetic dataset in the [public-data
evaluation](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html)
reports anybody, so the example has to come from a source table. `nlmixr2data::wbcSim` is a cohort whose follow-up ends by
672 hours, with three exceptions:

```{r b2-wbc, eval=has_nlmixr2data}
data("wbcSim", package = "nlmixr2data")
wbc_roles <- pmx_roles(id = "ID", time = "TIME", dv = "DV", amt = "AMT",
                       evid = "EVID", cmt = "CMT")
wbc_flags <- flag_identifiable_subjects(wbcSim, wbc_roles)
as.data.frame(wbc_flags)[wbc_flags$flagged, ]
```

Two patients, followed to 1730 and 4580 hours. The third, at 1130, is not
reported: its nearest neighbour is at 672, and 458 hours is just under the
cohort's median follow-up of 480. That is the line the default draws, and where
you want it drawn is a study-level decision — `min_relative_gap = 0.5` adds the
merely striking, and 0 gives you the modified z alone.

That the synthetic cohorts report nobody is the screen working rather than
idling. An avatar is a blend of several donors, so producing a subject that
extreme takes a defect, and this is the row that would say so.
`remediate_identifiable_subjects()` drops or truncates whatever it reports.

## B3. Is the synthetic cohort as a whole too close to the real one

```{r b3-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "B3", ])
```

B3 is adversarial accuracy, and it is a **cohort-level** statistic. It asks
whether the synthetic set as a whole sits closer to the real set than real
patients sit to each other, taking covariates and trajectory together.

The verdict is one-sided, because the two ways out of the interval are two
different findings. **Below** it is memorisation, which is the question this
section asks, and that is the only reading marked `review`. **Above** it the two
sets have separated — a classifier could tell them apart, which costs utility
and discloses nothing, so it passes. Inside is "nothing detected". The result
says `in`, `above` or `below` in every case, so the direction is readable
whatever the verdict says.

It asks, of every subject on both sides, whether its nearest neighbour lies in
its own dataset or in the other one, and reports the fraction that were
correctly placed. 0.5 means a synthetic subject is on average no more like a
real subject than two real subjects are like each other; toward 0 means
memorisation.

**One avatar sitting on top of one real patient will not move it.** The
statistic is an average over the whole cohort, so one memorised subject shifts
it by about `1/n` while the noise it is read against falls only as
`1/sqrt(n)` — the two never meet at pharmacometric cohort sizes. B3 detects a
*systematic* leak, one that affects a large share of the cohort. The per-record
questions are B1, B2 and B4, and B3 does not stand in for them.

The null interval is the same statistic computed on two random halves of the
**source**, `replicates` times, keeping the central 95% of those values. It is
therefore what the statistic does when both sides are genuinely real patients
from one cohort, at this cohort's size.

```{r b3-proximity, eval=both, message=FALSE}
proximity <- function(source, synthetic, roles, label) {
  report <- compare_pmx_proximity(source, synthetic, roles, replicates = 30)
  data.frame(
    dataset = label,
    adversarial_accuracy = round(report$adversarial_accuracy, 3),
    null_lower = round(report$null_lower, 3),
    null_upper = round(report$null_upper, 3),
    per_side = report$n_compared
  )
}
knitr::kable(rbind(
  proximity(mad, mad_synth, mad_roles, "mad"),
  proximity(pheno_sd, pheno_synth, pheno_roles, "pheno_sd")
))
```

A value inside the interval means *nothing was detected*, never *nothing is
there*. What it reliably catches is a blatant leak: replace the synthetic table
with a copy of the source when calling this function and it reports 0, well
below the interval, and objects.


**Read the direction before reacting to a miss.** Falling below the interval indicates 
synthetic subjects sitting closer to real ones than real
ones sit to each other, and is the failure this statistic exists to find. 

## B4a, B4b. Is any generated vector a copy of a real one

```{r b4-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% c("B4a", "B4b"), ])
```

The observation times (B4a), and the dosing times (B4b) should not be exactly reproduced from a patient.  Both must be **0** and either above 0 is
a `FAIL`. On a study with one protocol grid and good adherence, B4a and B4b are close to silent.

## B5. Did a rare category reach the output

```{r b5-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "B5", ])
```

B5 asks whether a level too few *source* patients held reached the output, and
it is `review`.

Numeric covariates are blended into a number nobody had. Categorical ones cannot
be — there is no average of `Female` and `Male` — so a synthetic patient's
category is `sample()`d from its donors and is always some real patient's actual
category, copied, and `strata` are copied from the anchor exactly as protocol
facts. A rare category therefore reaches the output whenever one of its holders
is chosen as a donor. The patient at risk is the one who is ordinary in every
way *except* that category: they sit in the middle of the profile space and are
chosen constantly, where a patient unusual on everything is nobody's nearest
neighbour and is self-protecting. `vignette("avatar-algorithm")` measures it
under step 8 — a level held by one patient almost never reaches the output and a
level held by two does — but that is geometry rather than protection, so nothing
enforces it and this row is what reports it.

`compare_pmx_rare_levels()` censuses every categorical axis the roles declare,
reporting the union of both sides, because a level the source had and the output
lost is a coverage failure a one-sided census cannot see:

```{r b5-census, eval=has_xgxr}
compare_pmx_rare_levels(mad, mad_synth, mad_roles)
```

A level is **exposed** when fewer than `min_pattern_share` real patients held
it, the same floor that protects visit sets. The row to act on is an exposed
level that `reached` the output, because that level is one real patient's
attribute, copied — and the card carries the list of them, since which levels
they were is what decides what to do. On `mad` nothing is exposed: the two arm
columns, `TRTACT` and the numeric `DOSE` that repeats it, hold ten subjects per
level and `SEX` holds thirty each. A study with a two-patient
stratum would show it here and nowhere else. `pheno_sd` declares no categorical
axis at all, so its B5 says the question could not be asked, which is different
from a study that was asked and came back clean.

The remedies are upstream of generation: drop the covariate from `covariates`,
or collapse its rare levels before generating.

Two limits. **The risk is rarity in the world, not in this dataset.** If five
people alive carry a mutation, a synthetic dataset containing it discloses that
someone with that mutation was in this study, which for a named trial with
public inclusion criteria can be nearly identifying on its own, and no cohort
size helps. And the census reads one column at a time, so a combination — arm x
sex x age band x mutation — is a quasi-identifier it does not enumerate. "What
is still missing" at the end says what closing that would take.

## These checks do not meet formal differential privacy criteria

It should be emphasized that these checks prevent some of the obvious forms of disclosure, but tehy do not meet the same criteria as differential privacy. 

# C. Has the synthetic study design significantly changed from the source

Failing these produces synthetic data that may not adequately describe the study.  

## C1. Did every arm keep its size

```{r c1-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "C1", ])
```

C1 counts how many declared strata hold the number of patients their source
stratum held. Every one matching is a pass; a shifted size is `review`, since a
dropped subject has to land somewhere. `compare_pmx_strata_sizes()` is the
per-stratum reading behind that count:

```{r c-sizes, eval=has_xgxr}
mad_sizes <- compare_pmx_strata_sizes(mad, mad_synth, mad_roles)
knitr::kable(
  as.data.frame(mad_sizes)[mad_sizes$column == "TRTACT x DOSE", -1],
  row.names = FALSE, digits = 1
)
```

One deliberate exception: a stratum holding fewer than three source patients is
left unbalanced on purpose, because reproducing its size exactly would disclose
that size. `strata_balanced` and `strata_stochastic` in the run settings say how
many strata fell on each side.

## C2. What features of the study were lost

```{r c2-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "C2", ])
```

C2 counts how many of the source's distinct dose-time schedules are represented
in the output. Every one still represented is a pass; fewer is `review`, because
shortening a regimen nobody else shares — or declining to build on it at all —
is the correct answer and the check is there to say what it cost. The rest of the question,
what else was lost, has no pass mark and is read from
`pmx_masking_report()`. The dosing section is the one C2 scores; drop the
`section` argument for the other six.

```{r c-masking, eval=has_nlmixr2data}
knitr::kable(as.data.frame(
  pmx_masking_report(pheno_synth, pheno_sd, pheno_roles,
                     section = "dose_schedules")
), row.names = FALSE, align = c("l", "r", "l"))
```

A "dose schedule", here and in C2, is the set of dose **times** a patient
received, and nothing else. Amounts are deliberately not part of it: on a
weight-based study every patient's milligrams are their own, so counting
amounts would make every patient their own schedule. Two patients on the same
visit days at different doses are one schedule; a patient who stopped after two
of three doses is another, and if nobody shares it, declining to anchor on them
removes it from the output.

On `pheno_sd` 35 of 56 distinct dose schedules are represented, and 92% of
avatars had their dosing re-truncated to reach that. It is the A5b finding
again, from the generator's side: the schedules survive, their tails do not.

## C3. Did every arm keep its endpoints

```{r c3-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "C3", ])
```

A3 compares endpoint sets across the whole cohort, so an arm that came back
without an endpoint the rest of the study still holds passes it. C3 asks the
same question arm by arm, and every arm matching is a pass.
`compare_pmx_strata_endpoints()` is the reading behind the count, one row per
arm and endpoint. Six arms and five endpoints make thirty rows, so this is the
PK concentration alone:

```{r c-endpoints, eval=has_xgxr}
mad_by_arm <- as.data.frame(
  compare_pmx_strata_endpoints(mad, mad_synth, mad_roles)
)
knitr::kable(mad_by_arm[mad_by_arm$endpoint == "PK Concentration", ],
             row.names = FALSE)
```

The cells are patients contributing the endpoint rather than measurements. A
measurement count moves whenever a sampling schedule was truncated, with every
arm and every endpoint intact, so it would report that movement and miss a
genuine loss.

The placebo arm holds no PK concentration on either side, which is the check
passing: an avatar never leaves the arm it was anchored in, so it cannot
acquire an endpoint its arm never had. The row to act on is a nonzero source
count against a zero synthetic count, and C3 is `review` rather than `FAIL`
when it appears, because an endpoint held by one patient in an arm leaves with
that patient when the generator declines to build on them.

This is the check that five endpoints buy. A3 above reads `5 of 5` on the
cohort, and would still read `5 of 5` if one arm had lost one of them entirely.

# D. How much have the covariate and observation distributions changed

## D1. Do the values land in the same range

```{r d1-card, eval=both, echo=FALSE}
synpmx_scorecard_datatable(mad_card[mad_card$check %in% "D1", ])
```

D1 reports the endpoint or covariate whose standard deviation moved furthest
between the two tables, in either direction. It is `review` on every study,
because spread shrinking is what the algorithm does rather than something it
got wrong, and spread growing has no threshold either.

**Look at the distributions before the number.** `compare_pmx_distributions()`
draws every endpoint and covariate the roles declare, source against synthetic
on shared axes, which is the reading a standard deviation cannot give you: one
mode and two modes with the same mean and spread produce identical rows.

```{r d-figure, eval=has_figure, fig.height=compare_pmx_distributions_height(mad, mad_roles)}
compare_pmx_distributions(mad, mad_synth, mad_roles)
```

A panel holding eight or fewer distinct values is drawn as proportion bars
rather than as a curve, which is why `PD - Binary` and `PD - Ordinal` come out
as bars: a smoothed density over `{0, 1}` would be a shape this study does not
have.

The same call with `output = "tables"` returns the numbers behind those panels,
and D1 is one cell of them:

```{r d-distributions, eval=has_xgxr}
mad_dist <- compare_pmx_distributions(mad, mad_synth, mad_roles,
                                      output = "tables")
knitr::kable(mad_dist$endpoints, digits = 2,
             caption = "The five endpoints, source against synthetic")
knitr::kable(mad_dist$covariates_numeric, digits = 2,
             caption = "Baseline weight, source against synthetic")
knitr::kable(mad_dist$covariates_categorical, digits = 2,
             caption = "Sex, source against synthetic")
```

**Between-subject variability shrinks, necessarily.** An avatar is an average of
several donors, and averaging reduces variance. Baseline weight has a source
standard deviation of about 14.5 kg and a synthetic one of about 8.8. D1 reports
whichever endpoint or covariate moved furthest, `sd x0.6` on that weight here,
and is always `review`, since neither direction has a threshold that would be
right on a second study.

**A categorical covariate moves for a different reason.** `SEX` is not blended —
there is no average of `Female` and `Male` — it is drawn from the donors, so its
proportions follow the anchor draw rather than shrinking toward a mean. An even
30/30 split comes back 40/20, visible as the last panel of the figure and the
last table above. Nothing in D1 scores it; declaring the column as `strata`
instead is what holds a split exactly, at the cost of treating it as a protocol
fact.

The quantities that explain how much the numeric side shrank are reported with
the run:

```{r d-blending, eval=has_xgxr}
unlist(attr(mad_synth, "pmx_settings")[
  c("k", "mean_effective_donors", "max_donor_weight", "cap_binding_fraction")
])
```

`mean_effective_donors` is how many donors an avatar is genuinely averaged over
once the weights are accounted for, about 2.9 here out of `k = 5`, and
`cap_binding_fraction` is how often the single-donor weight cap bound. Fewer
effective donors means more fidelity and less masking; that trade is the whole
design.

**Plot `DV` against time as well.** The figure above draws each variable's
distribution, which is not the same as its time course: two cohorts can hold the
same concentrations and reach them on different days. No function here overlays
source and synthetic profiles, because every group has plotting code it already
trusts for its own study — the [public-data
evaluation](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html)
shows one such overlay for each of eight datasets.

**Do the covariate–response relationships survive?** The question a modeller
actually cares about, and the least covered by anything above: a
weight–exposure slope, a dose–response relationship, a treatment effect.
Marginal distributions can match while the relationship between them is
destroyed. The
[covariate and treatment effect article](https://iamstein.github.io/synpmx/articles/example-avatar-PKPD-covariate-treatment-effect.html)
is the worked evidence.

# What these checks cannot tell you

**Nothing here bounds what an adversary learns.** These checks reduce the ways
a real patient can be singled out. They do not limit how often that succeeds,
and they compose no better than the weakest one. That is what the
differentially private modes are for.

**The guarantees are about reproduction, not similarity.**
`identifying_visit_sets` counts avatars with a visit set that is *exactly* one
real patient's. An avatar whose visits sit merely *near* a real patient's is
not covered by it, and `nearest_set_diff` is the only thing that will tell you
how near.

**Most checks read the source, and are therefore restricted output.** They may
not leave the environment the source data lives in. `compare_pmx()` reports
this per component:

```{r e-release, eval=has_xgxr}
compare_pmx(mad, mad_synth, mad_roles)$release_status
```

Only two things in this vignette do not read the source: `validate_pmx()` and
`flag_identifiable_subjects()`, both on the synthetic table. Every uniqueness
count, distribution comparison and proximity statistic is computed from real
patient data and inherits its handling obligations.

# What is still missing

**A1: a sample can cross a dose.** Generated observation times move and dose
times do not, so a sample drawn before a dose can land at or after it: a trough
becomes a time-zero sample, and a declared `occasion` column no longer runs in
step with `time`. Counting the avatars holding an observation labelled occasion
`k` that lands at or after the `k+1`th dose, `nlmixr2data::nimoData` on the
constructed nominal grid of the [public-data
evaluation](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html)
gives 12 of 12,
against 0 of 12 in its source. Closing it takes a post-generation invariant over
the finished table, checking that each observation keeps the number of doses
that preceded it and that `occasion` is non-decreasing within a subject.

**B5: rare combinations, and rarity in the world.** The source-side census
exists as `compare_pmx_rare_levels()` and B5. Joint combinations do not, since
with `d` covariates there are `2^d` of them and the census is per column;
closing that takes a decision about which combinations are worth enumerating.
One mitigation is accidentally present: each covariate is sampled
independently, so sex may come from one donor and race from another, and a rare
joint combination is less likely to be reproduced whole than any of its parts.
That is a fidelity cost, real correlations between covariates broken, doing
double duty as a weak privacy benefit.

The combination that matters most is a level that is rare *in the population*
rather than in this study, which is not measurable from the study at all. Its
realistic form is a declaration on `pmx_roles()`, or not carrying that
covariate. It is also the only gap here that would change generation rather
than reporting, by suppressing or coarsening a level.

Two further gaps are whole measurement approaches the package does not attempt,
both covered in the
[checking literature review](https://iamstein.github.io/synpmx/articles/synthetic-data-checking-review.html):

- **No control group.** Every check that reads the source compares the
  synthetic data against patients the generator was allowed to use. That
  confounds *the generator captured the population* with *the generator
  memorized a patient*, and the standard remedy — holding patients out of
  generation and asking the question twice — is not implemented. It is why B3's
  null interval says less than it appears to.
- **Nothing measures linkability, and nothing is per patient on the privacy
  side.** *Linkability* is an adversary connecting two records that belong to
  the same person, here an avatar to the real patient it was anchored on,
  without necessarily naming anybody. The AVATAR literature's two per-patient
  measures of it apply directly to this generator and are not computed:
  *local cloaking* is, for one real patient, how many other avatars lie closer
  to them than their own avatar does — 0 is the worst case and the published
  medians are 11 and 24 — and *hidden rate* is the percentage of real patients
  whose local cloaking is at least 1, published at 93% and 94%.

And one check cannot be on this list: **does the pipeline that will consume the
real study run unchanged against this?** Only you can answer it. Run your own
code against the output and see whether the joins, reshapes, derivations and
control streams behave.

## Where to go next

- `vignette("avatar-demo")` — generating a dataset and reading its scorecard, on one
  study end to end.
- `vignette("avatar-algorithm")` — how the default generator works, and the six
  masking mechanisms these checks are measuring.
- [Evaluating AVATAR on public data](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html) — these checks run across eight
  public datasets, with the masking cost for each.
- [Literature: checking synthetic data](https://iamstein.github.io/synpmx/articles/synthetic-data-checking-review.html) — the published methods for
  checking synthetic data, what each one asks, and which of them this package
  does not attempt.
- [Literature: generating synthetic data](https://iamstein.github.io/synpmx/articles/synthetic-data-generation-review.html) — the four families of
  generation method, and where `synpmx` sits among them.
- [Privacy](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html) — when the enumerated protections are not enough,
  and what a formal guarantee buys instead.
