---
title: "The synpmx AVATAR Algorithm"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The synpmx AVATAR Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
library(synpmx)
```

This article is the full specification of the AVATAR synthetic data generator, 
which builds synthetic records from the
local neighborhood of real records. The original method [2] was extended to 
population PK datasets in [1] and here, further extensions are provided so that
it works well with more realistic datasets.

## The process at a glance

Each synthetic subject is built from one real *anchor* subject's dosing event structure, 
then filled with measurements blended from several similar real subjects. The generator works on 
**any number of endpoints** — one can have multiple PK profiles
alongside multiple PD measures, declared through the `dvid` role. 
The pipeline has six **stages**, lettered (a) to (f). There are 5 different masking methods employed and 
they are denoted **M1** to **M5**.

- **(a) Declare the columns** — *Step 1*. Roles say which column is the id, the
  time, the dose, the measurement, and so on. Only named columns survive.
- **(b) Start each avatar from one real patient** — *Step 2*. Sample an
  **anchor patient** with replacement from the eligible source patients and copy its
  rows entirely: every TIME, EVID, AMT, RATE and CMT. Later steps
  will overwrite the DV and covariate columns and also alter TIME and AMT. 
  Starting from an actual patient's event table rather  than drawing the event fields independently 
  is a deliberate choice because independently
  drawn fields produce impossible records, such as an infusion rate on an oral
  dose. TIME is snapped to a grid of times, based on nominal time when available, to reduce
  the ability to identify a subject from the time of dosing and data collection (**M1**).  At the end,
  time is refined by adding some variability back to each time point.
- **(c) Put every subject on a comparable footing** — *Steps 3, 4, and 5*. Align
  time to the first event, choose a transformation for each DVID, and reduce each
  subject to one numeric *profile* (baseline covariates plus its trajectory on
  a common time-grid), then reduce the the data further then to a handful of principal-component 
  coordinates for measuring the similarity between two patients.
- **(d) Build each synthetic subject** — *Steps 6, 7, 8, and 9*. Copy a real
  anchor's event skeleton. Do not
  anchor on a source subject whose structure is an outlier, nor on one whose
  route arm is too small to blend within (**M2**). Draw which visits the avatar
  attended from sets at least two real patients share (**M3**). Pick the `k` (default 5)
  nearest donors, borrowing across dose/schedule groups when the anchor's own
  group is too small — but never across route of administration — and blend
  their covariates and trajectories onto the skeleton, capping any one donor's
  share at `max_donor_weight` (default 0.50) and adding subject-level and
  within-subject noise (**M4**). Recompute the dose from the avatar's own
  blended covariate when in the original dataset dosing is proportional to a covariate, as it is
  for weight-based dosing (**M5**).  Then reconstruct any below-limit (BLOQ) censoring, using
  an automatically detected LOQ for each DVID
- **(e) Recompile the dataset** — *Step 10*. Restore the source's
  column order, each column's type and factor levels, and the container class,
  so a tibble comes back a tibble. Fresh subject IDs are assigned, `MDV` is
  re-derived and a record of every
  choice made is attached as `attr(x, "pmx_settings")`.
- **(f) Screen the result** — *Step 11*. Because the event skeleton is
  copied from one anchor, a structurally unusual source subject yields a
  structurally unusual — and identifiable — avatar. A per-subject screen flags
  those, and a remediation step truncates, drops, or replaces them.  This is a separate
  step run manually by the user.  

Underneath all six stages is one idea. Every patient's rows hold two very
different things: the **event skeleton** (*when* the
dose events and DV observations happen) and the **values**, 
the DV numbers filling those observation slots plus the baseline covariates.

The skeleton is copied from one real patient with some masking and the values are blended from
several. The **anchor** supplies event skeleton, and multiple **donors** help inform DV values. 

## The masking, in the order it happens

The stages above describe how a synthetic dataset gets *built*. Cutting across
them are five **masking mechanisms**, numbered in the order they run. Each one
is explained in full at the step that fires it.

- **M1 — coarsen the times.** Snap every time onto a shared visit grid.  Then later add
  deviations back from a cohort-wide pool. This prevents identifying a patient from their
  TIME vector.
- **M2 — restrict who can be an anchor.** Drop patients whose structure no
  amount of blending would hide (e.g. patients followed twice as long as everyone else) 
  and patients in a route arm too small to blend within. *Step 2.*
- **M3 — redraw which visits the avatar attended.** Reselect which DV values at which time
points were collected, in case a patient has a unique set of missing visits *Step 2.*
- **M4 — blend the values across several donors.** Covariates and DV, capped so
  no one donor dominates, plus noise. *Step 7.*
- **M5 — recompute the dose from the avatar's own covariate.** Done when the dosing
  is weight- or BSA-based, so that subjects do not have uniquely identifying doses. *Step 9.*

## Step 1: declare the meaning of the columns

The package never guesses critical PMX roles from column names. The user must
declare them with `pmx_roles()`:

```{r roles-example}
roles <- pmx_roles(
  id = "ID",
  time = "TIME",
  dv = "DV",
  amt = "AMT",
  evid = "EVID",
  cmt = "CMT",
  dvid = "DVID",
  mdv = "MDV",
  rate = "RATE",
  cens = "CENS",           # below-limit indicator, if the study has BLQ data
  limit = "LIMIT",         # the other end of an interval, when used
  covariates = c("WT", "AGE", "SEX"),
  keep = "ARM"             # carried verbatim; see the allowlist below
)
roles
```

`id`, `time`, `dv`, and `evid` are required; the rest are optional. `amt`,
`cmt`, `mdv`, `rate`, `cens`, and `limit` each name one column. `dvid` names one
column, or several that encode the same endpoint (see below). `covariates` and
`keep` each take any number of columns. One source column cannot be assigned to
two roles. The roles `subject_properties`, `assigned_dose`, and `exclude` belong
to the differentially private engines only; `synpmx_avatar()` rejects them and
points at the role that does the job here.

### The role declaration is the allowlist

`synpmx_avatar()` keeps only columns a role names, and **drops every other
column**, reporting which ones. This is deliberate. AVATAR builds each synthetic
subject from one real *anchor* subject, and any column it retains that it does
not actively synthesize is copied from that anchor verbatim. A secondary
identifier — `USUBJID`, a site number, a randomization date — left undeclared
would otherwise ride a real subject's real value straight into the synthetic
data. Dropping by default means a column you forget fails safe, toward removal,
rather than leaking.

That leaves five ways a column is treated, and the difference matters:

| Declared as | What AVATAR does to it | Use for |
|---|---|---|
| `dv` (with `cens`/`limit`) | Blended across donors into a new trajectory | the measurement |
| `covariates` | Blended/resampled across donors into a new value | baselines you want *synthesized* — weight, age |
| `subject_properties` | Copied verbatim from the anchor, **and used to group two mechanisms** | assigned strata — treatment arm, dose group, cohort |
| `keep` | Copied verbatim from the anchor, and otherwise inert | assigned values you want kept faithful to that subject's dosing — a redundant endpoint label, a study identifier |
| *(undeclared)* | Dropped | anything you do not need |

`covariates` and `keep` are opposites, and choosing wrongly is the common
mistake. A **covariate is blended**: the synthetic subject gets a genuinely new
value interpolated from several neighbours, decoupled from any one person. A
**kept column is copied**: the synthetic subject gets one real subject's real
value, unchanged. For a treatment arm, a dose group, or a randomization sequence
— anything that must stay consistent with the doses AVATAR copied from the same
anchor — `keep` is correct, precisely because it never leaves that anchor's
side. Because a kept value is a real subject's real value, output containing it
must stay within the source data's own access controls and obligations.

**`subject_properties` carries a column the same way `keep` does, and then uses
it.** Both copy the anchor's value verbatim, so the choice between them is not
about what appears in the output — it is about whether that column should also
*partition the cohort*. A `subject_properties` column must be constant within a
subject, and the combination across such columns defines a stratum that two
mechanisms group by:

- **M5**, the dose recomputation, looks for a covariate-proportional dose
  *within each stratum*. A study dosing one arm at 1 mg/kg and another at
  2 mg/kg only looks proportional once the arms are separated; pooled, the
  ratios do not collapse onto a few levels and detection fails closed.
- **M3** draws each avatar's visit set from patterns held by other patients *in
  the same stratum*, so an avatar on a weekly schedule is never handed a daily
  arm's pattern.

It is deliberately **not** a blending barrier — donors are still borrowed across
strata to reach the floor, since only route of administration blocks that. So
declare a treatment arm as `subject_properties` rather than `keep` whenever the
arms differ in dose rule or visit schedule, and use `keep` for a label that
carries no such structure.

Redundant endpoint labels have their own handling. A dataset that carries both a
numeric `YTYPE` and a character `NAME` for the same endpoint declares **both**
as `dvid`: `dvid = c("YTYPE", "NAME")`. The first is the grouping key; validation
checks that the rest are a consistent 1:1 mapping with it — catching a source
where the two labels disagree, which `keep` would carry through silently — and
AVATAR carries all of them into the output, aligned.

Immediately before generation, `synpmx_avatar()` calls `validate_pmx(...,
strict = TRUE)`. Among other checks, validation requires:

- nonmissing IDs and EVID values;
- finite numeric TIME values and nondecreasing TIME within subject;
- numeric DV and finite nonmissing observations;
- an endpoint value on every eligible observation when DVID is declared;
- multiple `dvid` columns to be a consistent 1:1 mapping;
- every declared baseline covariate to be constant within subject; and
- when `cens` is declared, each CENS flag to agree with its DV — a
  left-censored value at the limit cannot exceed an ordinary measurement of the
  same endpoint. A pooled dataset with two assays at different limits will trip
  this legitimately; synthesize one study at a time, or drop the `cens` role.

Every error names the role and the column it maps to, and the count and an
example of the offending rows, so a failure says exactly what to fix.

Rows are eligible observations when EVID is zero and, if MDV is declared, MDV
is also zero. A nonmissing DV is additionally required when a value is used to
build a profile or donor trajectory. In symbols, with bracket notation for a
logical condition,

$$
O_{ij} = [\mathrm{EVID}_{ij}=0]
         [\mathrm{MDV}_{ij}=0\;\text{if MDV is declared}]
         [Y_{ij}\;\text{is present}].
$$

Rows with EVID equal to zero but missing DV, or with nonzero MDV, remain in the
sampled template but do not contribute a measurement value.

## Step 2: copy one anchor's rows

Let subject $i$ have source rows $D_i$. For every requested synthetic subject, the
generator samples an **anchor** $a$ with replacement and begins with

$$
S^{(0)} = D_a.
$$

Thus $S^{(0)}$ contains the anchor's full row count and row-wise combination of
TIME, EVID, AMT, RATE, CMT, DVID, and any other source columns. This is the
event skeleton. By copying these rows together, the method avoids impossible
records that can result from generating event fields independently.

Only a few parts of this skeleton are subsequently changed:

- TIME may be jittered if `time_jitter > 0`;
- declared baseline covariates are replaced for all rows of the synthetic subject;
- eligible, originally present observation DVs are synthesized;
- standard MDV may be re-derived;
- ID is replaced; and
- rows are sorted by TIME, retaining original order within ties.

Event-row DV values and undeclared columns otherwise remain anchor-template
values. Consequently, users should not place a semantically important column
in the input and assume the package understands it merely because it is
present.

### M2 — which patients may be used as an anchor

Runs before generation. Two
exclusions, both acting on the pool of source subjects an avatar may be built
from. A subject whose follow-up length or dose count exceeds twice the cohort's
90th percentile is never used as an anchor (`screen`), so no avatar inherits a
conspicuous skeleton. And a subject in a route arm holding fewer than `k + 1`
subjects has no legal donor set, so by default it is dropped from the pool too
(`on_donor_shortfall`).

Neither shrinks the cohort: the remaining anchors are sampled with replacement to
fill every slot, and excluded subjects still contribute measurements as donors.
What is lost is coverage of that structure, not sample size. This catches
subjects who are **extreme**; the post-generation screen below is a different thing, and not a masking step.

### M3 — redrawing which visits the avatar attended

Once
coarsening has put every subject on a shared visit grid, what remains of a
schedule is which of those visits each subject attended. Every time is then
shared, so no single visit is identifying — the *combination of absences* is,
and a patient who missed weeks 2 and 3 is singled out by a fingerprint made of
gaps. No grid fixes this at any resolution, because the grid decides where the
visits are and not which ones a subject has. So the pattern is drawn from ones at
least `min_pattern_share` subjects hold.

The default of 2 is the smallest value that means anything, and what it means is
exact: **no synthetic patient carries a schedule unique to a real patient.** An
attacker who links a reproduced pattern to a participant gets at least two
candidates, never one. Three is more conservative but no more defensible as a
sentence, and it costs sharply more — see the section on what this loses below.

"Sample which visits were attended" is the most easily misread row in that
table, so here it is concretely. **It concerns DV observations only. Dose events
are never redrawn** — an avatar's doses, their times and their amounts come
verbatim from its anchor, because a drawn regimen could be one no protocol
permits.

After M1, every patient sits on the same grid of visit times. What still differs
between patients is which of those visits each one actually turned up for:

```
grid            wk1  wk2  wk3  wk4  wk5  wk6
patient A        x    x    x    x    x    x     attended everything
patient B        x    x    x    x                dropped out after week 4
patient C        x    x         x    x    x      missed week 3
```

Patient B's and patient C's rows are *visit sets*. If only patient C ever missed
week 3, then "attended everything except week 3" points at patient C as surely
as an unusual sampling time would — a fingerprint made of gaps rather than of
times.

So M3 redraws that row. The avatar keeps its anchor's doses and the grid's visit
times, and only the **pattern of x's** is replaced by one drawn from patients who
share it, or built fresh with the same *shape* (how many misses, and whether
they run to the end, sit in a block, or scatter). What survives is how much
missingness the study had and of what kind; what is lost is which specific weeks
any one patient missed.

## Step 3: align time without pretending every subject was sampled identically

### First-event-relative time

Neighbor profiles and donor trajectories are constructed on an internal
aligned time. For subject $i$, the preferred origin is the first nonzero-EVID
row having a positive AMT:

$$
t_{0i} = \min\{t_{ij}: \mathrm{EVID}_{ij}\ne0,\ \mathrm{AMT}_{ij}>0\}.
$$

If no positive-AMT event exists, the first nonzero-EVID row is used. If the
subject has no event row, TIME is left unchanged. The internal aligned time is

$$
t^*_{ij}=t_{ij}-t_{0i}.
$$

This removes a subject-specific calendar or study-time offset. It does **not**
compute time after each dose.

### TIME is not TAD

A conventional time-after-dose (TAD) variable resets after every dose. AVATAR
does not use a declared `tad` role for alignment and never recomputes a TAD
column. With multiple doses, $t^*$ continues to increase from the first
qualifying dose; there is no later reset. A declared `tad` column is retained
and copied from the anchor template unchanged; an undeclared one is dropped, as
any undeclared column is.

This has two practical consequences:

1. Use the actual analysis-time column as `time`. Do not expect a TAD column to
   drive neighbor construction or interpolation.
2. If `time_jitter > 0`, a declared TAD column is copied, not recalculated, and
   can become inconsistent with the jittered TIME. Recompute TAD downstream if
   the workflow requires it.

### Different observation times across subjects

Two subjects sampled at different times cannot be compared as vectors until
they are put on shared coordinates. That is what the common grid is for — and
it is worth saying up front what it is *not* for, because the 15-point cap
below invites a misreading.

> **The grid is a ruler, not a resampler.** It is used only to build the
> profile that measures which subjects resemble each other (Step 5). It never
> limits the resolution of the generated data. A synthetic subject carries
> every one of its anchor's observation times, however many that is, and the
> donor values placed there are interpolated from each donor's **full** observed
> trajectory — not from its 15-point summary. A richly sampled source gives a
> richly sampled avatar.

For endpoint $e$, the package pools finite aligned observation times and makes
a common grid $G_e$. If there are at most 15 unique times, all are retained. If
there are more, 15 type-8 empirical quantiles are used and duplicate quantiles
are removed.

Each subject's transformed trajectory is linearly interpolated onto $G_e$.
There is no extrapolation during profile construction, so grid locations
outside a subject's observed window become missing profile features. Tied
times are averaged by `stats::approx(..., ties = mean)`.

The cap does bind on ordinary PK data: `theo_md` pools 156 distinct aligned
times, so donor ranking there runs on a 15-point summary of trajectories with
about 22 observations each, compressed further by the PCA of Step 5. Two things
limit the damage. The grid points are *quantiles of the observed times*, so they
land where sampling is dense — early, around absorption and peak — rather than
spreading evenly across a long tail. And the grid is built per endpoint, so PK
and PD each get their own 15 rather than sharing them. The residual risk is
worst for multiple-dose data with several peaks and troughs, where 15 points
across a long window can blur the shape that distinguishes subjects. When that
happens the cost is a *worse choice of donors*, never a coarser output
trajectory. `max_points` is currently fixed at 15 and is not an argument.

During final synthesis, donors are first interpolated directly to the anchor's
aligned observation times. Suppose target time $t$ lies outside donor $r$'s
window. Rather than extrapolating indefinitely, the method maps the target's
relative position in the whole anchor observation window to the donor window:

$$
q(t)=\frac{t-a_{\mathrm{target}}}
           {b_{\mathrm{target}}-a_{\mathrm{target}}},
\qquad
u_r(t)=a_r+q(t)(b_r-a_r),
$$

where $[a_{\mathrm{target}},b_{\mathrm{target}}]$ and $[a_r,b_r]$ are the
target and donor ranges. The donor trajectory is evaluated at $u_r(t)$. This
normalized-window fallback preserves progress through a trajectory, but it is
not a mechanistic warping model. A donor with only one unique time supplies its
mean at every requested target time.

### Optional coherent time jitter

The default `time_jitter = 0` copies template times exactly. Otherwise, one
normal perturbation is drawn for each unique template time $u_l$,

$$
u'_l=u_l+\delta_l, \qquad
\delta_l\sim\mathcal{N}(0,\sigma_t^2),
$$

and constrained between the midpoints of adjacent nominal times. Nonnegative
time is enforced at the first time. All rows tied at $u_l$ receive the same
$u'_l$, so a dose and observation tied at a nominal time remain tied and time
ordering cannot cross.

### M1 — coarsening the times, and putting the scatter back

(`coarsen_time = TRUE`,
before Step 5). Source times are collapsed onto a shared visit grid, and the
per-visit deviations are pooled across the cohort and resampled independently
onto each avatar. This protects the **schedule**, which blending never touches:
the skeleton is copied verbatim from one anchor, and under actual recorded times
almost every subject holds the only copy of their visit vector. The order is the
mechanism — snapping is many-to-one and *destroys* the deviation, whereas
perturbing a time in place leaves it recoverable. Note that `time_jitter` is
therefore not an alternative: its clamp holds every time inside its own Voronoi
cell, so at any magnitude a visit stays within half a gap of the source value.

M1 is the only mechanism that both removes something and then restores part of
it, so it is worth separating the two halves.

**Snapping**, before any avatar exists, replaces each recorded time with its
visit's grid time. The deviation removed is not discarded — it goes into one pool
shared by the whole cohort.

**Restoring**, while each avatar is built, draws a deviation from that pool and
adds it back. Without this the output would sit exactly on a tidy grid, which no
real study does, and any analysis code that reconciles actual against nominal
time would have nothing to reconcile.

The restoration is **not itself masking**, and it does not undo the masking
either. That distinction is the whole point: an avatar gets a deviation drawn
from the cohort, never the one its own anchor had. The scatter comes back; the
fingerprint does not.

Each drawn offset is clamped inside its own visit's cell, so no time moves more
than half the gap to the neighbouring visit, and resampling can neither reorder
visits nor stretch follow-up. A dose event and a DV observation recorded at the
same time receive the same offset and stay together.

**This is why `time_jitter` is not an alternative.** Jitter perturbs times
without ever destroying the original, and its clamp holds every value within half
a gap of *the source patient's own* value at any setting — so the source schedule
stays recoverable. Jitter is a realism control. M1 is the privacy one, and the
difference is that M1 throws the real deviation away before drawing a new one.

## Step 4: choose an endpoint transformation

### Why there is a transformation at all

Blending happens *on the transformed scale*. So this step is really the choice
of **what "average" means** when five patients are mixed together.

Concentrations span orders of magnitude. Averaging 0.5, 0.8, and 12.0 on the
raw scale gives 4.4 — a number dominated by the single high patient and
representative of none of them. Averaging on the log scale gives about 1.6, the
geometric mean, which is the pharmacometric convention for precisely this
reason. Blend concentrations on the raw scale instead and one high-exposure
donor drags every avatar it touches upward.

Logs have one problem with PK data: exact zeros. Pre-dose rows and imputed BLQ
values are genuinely 0, and $\log 0=-\infty$ would poison the entire blend. So a
small constant $c_e$ is added before taking logs. Choosing $c_e$ to be **half
the smallest positive value in that endpoint** puts zeros just below the
smallest real measurement — near the bottom of the observed range, where a zero
belongs — and scales it to the data rather than fixing an arbitrary $10^{-6}$.

Not every endpoint wants logs. A PD score, a change from baseline, or a
temperature difference can legitimately be negative, where a log is undefined.
The "positive-like" test below is what separates the two cases; its 1% tolerance
lets a concentration endpoint with a couple of small negative assay readings
still count as positive-like.

### The rule

Transformations are selected separately for each DVID endpoint, or once for
the implicit endpoint `"DV"` when DVID is absent. After discarding nonfinite
values, an endpoint is considered positive-like when it has at least one
positive value, its median is positive, and at most 1% of values are negative.

For a positive-like endpoint, define

$$
c_e=\max\left\{\frac{1}{2}\min(Y_e[Y_e>0]),\sqrt{\epsilon_{\mathrm{mach}}}\right\}
$$

and use the offset-log pair

$$
g_e(y)=\log\{\max(y,0)+c_e\},
\qquad
g_e^{-1}(z)=\max\{\exp(z)-c_e,0\}.
$$

Other endpoints use $g_e(y)=y$. The truncations mean a positive-like endpoint
cannot generate a negative DV, whatever the noise of Step 9 does — a guarantee
that comes free with the back-transformation rather than needing a separate
clamp. Transformation choices and offsets are stored in
`attr(synthetic, "pmx_settings")$endpoint_transforms`, so the scale an endpoint
was blended on is always recoverable from the output.

## Step 5: build one numeric profile per source subject

A profile concatenates baseline and longitudinal information:

$$
\mathbf{x}_i=
\left[
  \mathbf{c}_i,\
  \mathbf{y}_{i1}(G_1),\ldots,\
  \mathbf{y}_{iE}(G_E)
\right].
$$

Here $\mathbf{c}_i$ contains the first nonmissing value of each declared
baseline covariate. Numeric covariates contribute one feature. Factors and
character covariates contribute one indicator for every factor level or
observed character value. The vector $\mathbf{y}_{ie}(G_e)$ is subject $i$'s
transformed endpoint-$e$ trajectory interpolated on its common grid.

This profile is used only for distance calculations. It does not replace the
source data and is never returned as a synthetic patient.

### Multiple endpoints

`synpmx_avatar()` handles any number of endpoints — a PK concentration and one
or more PD measures, declared through `dvid`. Each is processed *separately*
first and only then merged, in this order:

1. **Transform, per endpoint.** Step 4 runs once per endpoint, so PK may be
   blended on the log scale while a PD score stays on the identity scale.
2. **Interpolate, per endpoint.** Each endpoint gets its *own* common grid,
   built from the pooled observation times of that endpoint alone. A sparsely
   sampled PD is not padded out to match a rich PK.
3. **Concatenate.** Covariate features, then each endpoint's grid columns, are
   laid side by side into the single row $\mathbf{x}_i$ above.
4. **Impute, drop, standardize.** Median imputation, removal of constant
   columns, then column-by-column standardization.
5. **One PCA over everything**, giving a single $\boldsymbol{\xi}_i$ that mixes
   all endpoints — not one PCA per endpoint.

Consequently there is **one distance, one donor set, and one set of weights**,
shared by every endpoint. A donor is chosen for overall similarity; a donor that
matches PK closely but PD poorly still donates its PD.

Standardization is what makes endpoints commensurable at all: a log
concentration and a raw neutrophil count cannot be compared until each column is
divided by its across-subject standard deviation.

But note carefully what standardization does *not* do. It equalizes **units**,
not **endpoints**. Every grid point becomes one unit-variance column, so an
endpoint sampled at 15 times contributes 15 columns while one sampled at 3 times
contributes 3, and a Euclidean distance summing over columns lets the first
outvote the second by roughly five to one. On a worked two-endpoint example with
7 usable PK columns, 3 PD columns, and 1 covariate, the retained variance splits

| block | features | share of retained variance |
|---|---|---|
| covariate | 1 | 4.4% |
| PD | 3 | 28.1% |
| PK | 7 | 67.5% |

so donor selection is driven mostly by whichever endpoint was sampled most
densely, and baseline covariates are a small minority of the distance whenever
trajectories are rich. **This weighting is an accident of the sampling design,
not a modeling choice**, and there is currently no way to reweight endpoints.
If donors must match on a sparse endpoint, that is a limitation to know about
before trusting the selection.

Two related behaviors follow. A subject missing an endpoint entirely is
median-imputed on those columns, so it looks *average* there and sits nearer the
middle of the space than it deserves to. And at generation, if a chosen donor
has no value for the endpoint being filled, the weights renormalize locally over
the donors that do — so a missing endpoint thins the blend rather than punching
a hole in it.

### Missing features and scaling

An entirely missing feature is removed. Otherwise, a missing profile entry is
replaced with its across-subject feature median $m_l$:

$$
x^{\mathrm{imp}}_{il}=
\begin{cases}
x_{il}, & x_{il}\text{ is finite},\\
m_l, & \text{otherwise}.
\end{cases}
$$

This imputation exists only inside neighbor finding. It does not fill source or
output DV rows. Features whose standard deviation is not greater than
$\sqrt{\epsilon_{\mathrm{mach}}}$ are removed. The remaining columns are
standardized as

$$
z_{il}=\frac{x^{\mathrm{imp}}_{il}-\bar{x}^{\mathrm{imp}}_l}{s_l}.
$$

### Rank-safe PCA

When there are at least two subjects and two usable features, PCA is applied to
the standardized matrix with maximum rank

$$
r_{\max}=\min(n-1,p).
$$

Let $\lambda_h$ be the variance of principal component $h$. The retained
dimension $H$ is the smallest value satisfying

$$
\frac{\sum_{h=1}^{H}\lambda_h}
     {\sum_{h=1}^{r_{\max}}\lambda_h}
\ge v,
$$

where $v$ is `pca_variance` (default 0.90).

The **subject coordinate** is the output of this whole step and the input to
every distance in Step 6, so it gets its own symbol:

$$
\boldsymbol{\xi}_i=(\xi_{i1},\ldots,\xi_{iH}),
\qquad
\xi_{ih}=\text{subject } i\text{'s score on principal component } h .
$$

Read $\boldsymbol{\xi}$ as the Greek letter *xi*. It is one point per subject in
an $H$-dimensional space, and $H$ is typically a handful — the number of
components needed to reach $v$. Everything upstream exists to produce it: the
raw profile $\mathbf{x}_i$ is imputed to $x^{\mathrm{imp}}_{il}$, standardized
to $z_{il}$ so that a weight in kilograms and a log-concentration compete on
equal footing rather than by the size of their units, and finally rotated by
PCA into $\boldsymbol{\xi}_i$.

If PCA is not available but standardized features exist, those features are used
directly as $\boldsymbol{\xi}_i$. If none exist, all subjects receive the same
one-dimensional zero coordinate.

The term **rank-safe** means the attempted PCA rank cannot exceed either the
feature rank or $n-1$. It does not imply that sparse or nearly duplicated
profiles contain useful separation.

## Below-the-limit-of-quantification (BLOQ) data and a Monolix-style CENS column

Declare `cens` and, if the source uses one, `limit` in `pmx_roles()`, and AVATAR
reconstructs censoring rather than copying it. The mechanism has three parts.

**The boundary is read from the source.** AVATAR makes no formal privacy claim,
so it may recover the assay limit from the data itself: under the Monolix
convention a censored row reports the boundary in `DV`, with `LIMIT` carrying
the other end of an interval when there is one. Left, right, and interval
censoring are each recognised. A differentially private engine may *not* do
this — there the limit must be a declared public input, because a boundary
inferred from confidential records is itself a release. That asymmetry is the
reason the two engines derive boundaries differently.

If a study used more than one assay limit, the most conservative boundary is
taken rather than inventing a per-row rule the source cannot support.

**Censored donor values are imputed before blending.** Every censored source row
reports the same substituted number, so blending those values directly would put
a floor under the synthetic data that the real study does not have. Each
censored row is instead replaced by a draw inside its censoring region — uniform
below the limit for a left-censored row, uniform within the interval for an
interval-censored one — before the trajectory is transformed and blended. A
uniform draw rather than a fixed LLOQ/2 avoids replacing one artificial spike
with another.

**`DV`, `CENS`, and `LIMIT` are reconstructed together.** The back-transformed
blend is treated as a *latent* value: what the subject would have measured with
no assay limit. The boundary is then applied to that latent, and all three
columns are written from the result, so they cannot disagree. A generated value
below the limit produces `CENS = 1` with `DV` exactly at the limit, reproducing
the spike of identical values that characterises real censored data.

Two honest limitations. The censored *fraction* is reproduced approximately, not
exactly: blending shrinks variance toward the middle while the imputation draws
uniformly across the censoring region, so the synthetic rate typically runs
somewhat above the source's. And if a `cens` role is declared but no boundary
can be read from it — rows flagged with `DV` missing, for instance — AVATAR
warns and carries the flag through untouched rather than guessing.

## Step 6: choose the donors

Similarity in PCA space is not sufficient on its own. A subject receiving a
single oral dose should not donate a trajectory to an infusion or repeat-dose
template. But the two ways a donor can be unsuitable are not the same kind of
problem, and the generator treats them differently.

### Two structural axes, one of them absolute

**Route of administration is an absolute barrier.** A bolus, an infusion, and an
oral dose produce categorically different concentration shapes. Blending across
them does not yield a noisier version of the truth; it yields a trajectory no
protocol could have produced. Donors are therefore *never* drawn from outside
the anchor's route, at any stage, for any shortfall — including when that leaves
the anchor with no legal donor at all. Each subject's route key is read from its
dosing rows only, as the set of

$$
\bigl(\mathrm{EVID},\ \mathrm{CMT},\ \mathbb{1}[\mathrm{RATE}\neq 0]\bigr)
$$

triples appearing on them, so that the compartment a dose enters and whether it
is delivered over time both bind, while dose *count* does not: three oral doses
and five oral doses share a route. NONMEM's `RATE < 0` (modeled rate or
duration) counts as an infusion.

**Everything else is a preference.** Each subject also receives an event
signature containing:

- the ordered EVID values on event rows;
- optional CMT and DVID values on those rows;
- the sign and rounded numeric magnitude of optional AMT and RATE;
- the number of positive-dose starts, or all events if no positive AMT exists;
- successive start-time gaps rounded to two significant digits; and
- the set of endpoints on eligible observation rows.

Notice what is not in the signature: the observation-time schedule. Event values
still come from the anchor template, while unequal observation times are handled
by interpolation. Dose magnitude *is* in the signature, so same-dose subjects
are preferred — but unlike route it is not a barrier, and the fallback below
will cross it.

### The distance

For anchor $a$ and candidate donor $r$, the distance is the ordinary
straight-line (Euclidean) distance between their subject coordinates from
Step 5:

$$
d_{ar}=\left\|\boldsymbol{\xi}_a-
                   \boldsymbol{\xi}_r\right\|_2
=\sqrt{\sum_{h=1}^{H}(\xi_{ah}-\xi_{rh})^2},
$$

where

- $h$ indexes the **retained principal components**, running from 1 to $H$ —
  not subjects and not time points;
- $\xi_{ah}$ is the **anchor's** score on component $h$, and $\xi_{rh}$ is
  **candidate $r$'s** score on that same component, both defined in Step 5;
- each squared term is one component's disagreement between the two subjects,
  and the sum over all $H$ components, square-rooted, is how far apart they sit.

It is Pythagoras in $H$ dimensions; all the work went into choosing the
coordinates. Because the features were standardized before PCA, no single
original measurement dominates the distance merely by being recorded in larger
units.

This is the only distance in the selection rule. There is no separate structural
metric trading dose differences against schedule differences: structure enters
as the two-stage ordering below, not as a score.

### The selection algorithm

Write $\mathcal{R}_a=\{r\neq a: \mathrm{route}_r=\mathrm{route}_a\}$ for the
route-compatible pool and $\mathcal{S}_a=\{r\in\mathcal{R}_a:
\mathrm{sig}_r=\mathrm{sig}_a\}$ for the exact-signature subset. To choose $k$
donors for anchor $a$:

1. **Exact stage.** Sort $\mathcal{S}_a$ by increasing $d_{ar}$, breaking ties
   by subject index, and take the first $k$. If $|\mathcal{S}_a|\ge k$, stop.
2. **Fallback stage.** Sort $\mathcal{R}_a\setminus\mathcal{S}_a$ the same way
   and take enough of it to bring the total to $k$, whatever the dose or
   schedule. Record a warning that donors were borrowed across dose groups.
3. **Shortfall.** If $|\mathcal{R}_a|<k$ the floor is unreachable, because the
   only remaining candidates are on another route and route is never crossed.

Ties break by subject index, so selection is deterministic under a fixed seed.
Stage-2 donors' measurements are mapped onto the anchor's own observation times
by interpolation, so the avatar keeps its anchor's regimen while its values are
averaged across at least $k$ real patients. This borrowing matters most for
datasets with individualized dosing, where an exact dose magnitude can make
nearly every subject its own signature group: in `theo_md`, weight-based dosing
yields 11 signature groups across 12 subjects, so stage 2 does essentially all
the work.

### What stage 2 costs, stated plainly

Be clear about what the fallback distance does and does not know. `AMT` is not a
profile feature, so the stage-2 ranking never compares doses directly — it
compares baseline covariates and the shape and level of the DV trajectory.
Dose does influence it, but only *indirectly*: a subject given a much larger
dose has higher concentrations, and those concentrations are profile features,
so it lands further away. The ranking therefore tends to prefer similar-dose
donors without being told to.

Tendency is not a guarantee, and nothing rescales. A borrowed donor's values are
interpolated onto the anchor's times and blended **raw**, by deliberate design
(structural realism over statistical fidelity). The consequence is worth stating
directly rather than leaving implicit: a synthetic subject carries its anchor's
`AMT` while its concentrations may be blended from donors given a different
dose, so **the dose–exposure relationship in AVATAR output is not guaranteed to
be preserved**. For developing and debugging model code — the use this generator
exists for — that is acceptable and understood. For estimating parameters, it is
not: fitting a structural model to AVATAR output can recover a dose–exposure
relationship that is an artifact of blending. Use `synpmx_calibrated()` when the
parameters themselves have to mean something.

### When the floor cannot be reached

Because route is absolute, a route arm holding fewer than $k+1$ subjects can
never supply a legal donor set. There is no good answer in that situation, only
a choice between two bad ones, and which is worse depends on whether the sparse
arm matters more than the disclosure risk of reproducing it. That is the
caller's judgement, so `on_donor_shortfall` makes it explicit:

| `on_donor_shortfall` | Behavior |
|---|---|
| `"drop"` *(default)* | Omit those subjects from the anchor pool. No avatar is built on them, and the synthetic cohort does not represent that arm. |
| `"noise"` | Keep them, blending however many same-route donors exist — possibly none — and relying on `subject_noise_sd` and `residual_noise_sd` for the rest. |
| `"error"` | Refuse to generate, naming the arm and both alternatives. |

Every branch alerts loudly, because each one silently changes something a
reader would want to know: what the cohort covers, or how identifying it is.

`"noise"` is **available but not recommended**, and the reason is the defect
this whole mechanism exists to prevent. A subject blended from one donor, or
from none, is a noised near-copy of a real patient — exactly the behavior
`REV-025` was raised about. If you use it because the sparse arm genuinely
matters more, screen the result with `flag_identifiable_subjects()` and treat
those subjects as individually identifying.

One case forces the issue: when *every* arm is below the floor, `"drop"` would
leave nothing to generate, so generation proceeds as if `"noise"` and says so.

## Step 7: randomize and cap donor weights

Suppose $K$ donors have been selected. Let $E_r\sim\mathrm{Exponential}(1)$
and let $R_r$ be a random permutation of $1,\ldots,K$. Importantly, $R_r$ is a
randomized rank, not the distance order. The raw donor weight is

$$
q_r=\frac{E_r}{\max(d_{ar},\varepsilon)}2^{-R_r},
$$

where

$$
\varepsilon=
\begin{cases}
\max\{10^{-8},10^{-6}\operatorname{median}(d_{ar}:d_{ar}>0)\},
  & \text{if a positive distance exists},\\
10^{-8}, & \text{otherwise}.
\end{cases}
$$

The first normalization is $w_r=q_r/\sum_s q_s$. If the raw weights are
nonfinite or have a nonpositive total, the implementation replaces them with
equal raw values before normalizing.

### Capping every donor, not just the largest

Each donor's weight is its **share of the avatar**: five weights that sum to 1.
The floor $k$ says how many real patients go into the blend, but it says nothing
about how much of any one of them comes out. Five donors weighted $(0.95, 0.02,
0.02, 0.005, 0.005)$ satisfy a floor of five while being, to any practical
purpose, a copy of one person. So the cap, not $k$, is what controls how closely
an avatar can resemble a single real patient.

The raw formula above produces very uneven weights on purpose — that is the
rank attenuation doing its job. Left alone it hands one donor a median 58% of
the avatar. `max_donor_weight` (written $c$ below, default $0.50$) is the
ceiling on any one share.

Enforcing it is less obvious than it sounds, because **weights have to keep
summing to 1**. Cutting the leader down does not remove that weight from the
blend; it has to be given to the others, in proportion to what they already
hold:

$$
w_r \leftarrow \Bigl(1-\textstyle\sum_{s\in P}w_s\Bigr)
              \frac{w_r}{\sum_{s\notin P}w_s},
\qquad r\notin P,
$$

where $P$ is the set of donors already pinned at $c$. And that redistribution
can itself push a donor over the ceiling. Starting from $(0.50, 0.25, 0.15,
0.06, 0.04)$ with a tight $c = 0.30$: cutting the leader to $0.30$ frees $0.20$,
and the second donor's share of it takes them from $0.25$ to $0.35$ — over the
ceiling. Capping only the largest weight and stopping, which is what the
implementation did throughout the $0.80$ era, leaves exactly that violation in
place: the stated maximum, broken by the donor the redistribution created.

So the rule repeats — *pin everything over $c$, redistribute, look again* — the
standard construction called **water-filling**. Pinned donors are never topped
up, so each pass pins at least one more and it terminates. The tight example
settles at $(0.30, 0.30, 0.24, 0.096, 0.064)$.

### At the default cap, one pass is provably enough

Worth knowing, because it means the loop above is not the complexity it looks
like. A second pass is needed only when redistribution lifts the runner-up over
the ceiling, that is when

$$
(1-c)\,\frac{w_2}{1-w_1}>c .
$$

At $c=0.5$ this requires $w_2>1-w_1$, i.e. $w_1+w_2>1$, which no pair of weights
can satisfy. **At any cap of $0.5$ or above the loop runs at most once**, and
simulation agrees exactly: at $c=0.5$ the cap is untouched 33% of the time and
pinned once 67%, never twice. Only tighter caps exercise the iteration — at
$c=0.3$ two passes are needed 53% of the time and three passes 11%.

The general routine is kept because `max_donor_weight` is an argument and a user
may set it below $0.5$, where a single pass is genuinely wrong. But at the
shipped default, the simple thing and the correct thing coincide.

### The infeasible-cap boundary

With $K$ donors the weights average $1/K$, so **no weight vector can satisfy a
ceiling below $1/K$** — asking for $c=0.3$ from three donors is asking three
numbers below $0.3$ to sum to 1. Such a cap relaxes to exactly $1/K$: uniform
weights, the flattest blend available, rather than an error. Two donors under
$c=0.5$ give $(0.5,0.5)$, which meets the cap exactly. With one donor the weight
is necessarily 1; a cap cannot invent a second donor.

### Why a cap is needed at all

It is reasonable to ask whether the cap earns its place, given that $k$ already
forces five donors into every avatar. The answer is that **without a cap, the
floor is largely decorative.** Simulating the raw weight formula at $k=5$ with
representative donor distances:

| uncapped largest donor share | value |
|---|---|
| median | 0.58 |
| interquartile range | 0.47 – 0.72 |
| $P(\text{largest} > 0.5)$ | 0.67 |
| $P(\text{largest} > 0.8)$ | 0.14 |
| effective donors $1/\sum_r w_r^2$ | 2.37 |

So the unconstrained formula typically puts **58% of an avatar into a single
donor**, and one avatar in seven takes more than 80% from one person. Nominally
five patients are blended; effectively about two and a half are. The cap is the
only thing that closes that gap.

### Choosing the cap

The useful diagnostic is **how often the cap fires**, because that says what
role it is playing:

| `max_donor_weight` | binds on | effective donors |
|---|---|---|
| 0.30 | 99% | 3.92 |
| 0.40 | 89% | 3.28 |
| 0.50 *(default)* | 68% | 2.90 |
| 0.60 | 47% | 2.64 |
| 0.80 | 15% | 2.40 |
| none | 0% | 2.37 |

Both ends of that table are the wrong kind of parameter. A cap that fires on
99% of subjects is not a guardrail — it *is* the weighting scheme, and the
inverse-distance term underneath it barely matters. A cap at $0.80$ fires on
15%, trimming only the worst tail while still permitting one patient to be
four-fifths of an avatar; that was the old default, and it was never really
protecting anything.

The default is **$0.50$**, which states as a single checkable sentence: *no
single real patient is more than half of any synthetic patient.* It fires on
about two thirds of subjects, so it genuinely constrains without wholly
replacing the distance weighting.

Its honest cost is that "five donors are blended" is really "about three
effective donors" ($2.90$). Tightening to $0.30$ buys about one more effective
donor at the price of a cap that fires essentially always. Both are defensible;
$0.50$ is the one that can be explained in a sentence.

Because these numbers depend on the cohort, `pmx_settings` records
`cap_binding_fraction` and `mean_effective_donors` for every run, so the same
question can be asked of real data rather than of a simulation.

### What the cap costs

Treating donors as independent, a weighted blend retains $\sum_r w_r^2$ of
individual variance, so the synthetic cohort's between-subject standard
deviation shrinks by $\sqrt{\sum_r w_r^2}$. The reciprocal
$1/\sum_r w_r^2$ — the effective number of donors, reported as
`mean_effective_donors` in `pmx_settings` — reads as a privacy floor and as the
variability cost in one number. It equals $K$ for uniform weights and $1$ for a
sole donor.

That independence assumption is, however, pessimistic in the direction that
matters. Donors are *nearest neighbours*, so they are strongly positively
correlated and averaging them destroys much less variance than the formula
implies; the subject-level perturbation of Step 9 then restores more. Measured
on `theo_md` — between-subject SD of $\log\mathrm{AUC}$, averaged over 20 seeds,
against a source SD of $0.273$ — the cap turns out to be nearly free:

| `max_donor_weight` | effective donors | BSV retained |
|---|---|---|
| 0.80 | 2.50 | 72% |
| 0.50 | 2.93 | 75% |
| 0.30 | 3.91 | 78% |
| 0.25 | 4.43 | 74% |
| 0.20 (uniform) | 5.00 | 68% |

Tightening the cap raises the effective donor count substantially — 2.4 with no
cap, 2.9 at the default $0.50$, 3.9 at $0.30$ — while between-subject
variability stays flat within noise at a 12-subject source. The independence
formula would have predicted a fall from 81% to 49% over that range; correlation
among neighbours is why it does not happen. This is one dataset and one summary,
so the table is evidence that capping is affordable here, not a general result.
It is also the reason the cap can be chosen on privacy grounds: the fidelity
side of the trade is nearly flat.

The same selected subjects and weights are used for every declared covariate
and endpoint. When donor $r$ lacks a usable value at target location $j$, the
available weights are renormalized locally:

$$
w_{rj}^{*}=\frac{w_r I_{rj}}
                 {\sum_s w_s I_{sj}},
$$

where $I_{rj}=1$ when the value is finite and zero otherwise.

### M4 — why blending is the mechanism that protects the values

Spanning Steps 6-9. No avatar's measurements come from
one real patient: each is a distance-weighted blend of at least `k` = 5
compatible donors, with `max_donor_weight` = 0.50 capping any single donor at
half the blend. This is the only mechanism whose strength grows with cohort
size, which is why a source with fewer than `k + 1` subjects raises a loud
alert instead of quietly producing near-copies.

## Step 8: synthesize baseline covariates

For a numeric covariate, let $c_r$ be donor $r$'s first nonmissing value. When
all available values are positive and

$$
\frac{\max_r c_r}{\operatorname{median}_r(c_r)}>3,
$$

the blend is formed on the log scale; otherwise it is formed on the original
scale. With working-scale values $h(c_r)$, the donor center is

$$
\mu_c=\sum_r w_r^* h(c_r).
$$

Let $s_c$ be the ordinary sample standard deviation of the available
working-scale donor values. If it is unavailable or zero, the fallback is
$\max(0.05|\mu_c|,0.01)$. The generated working-scale covariate is

$$
c^*=\mu_c+\eta_c,
\qquad
\eta_c\sim\mathcal{N}(0,(\sigma_{\mathrm{subj}}s_c)^2),
$$

where `subject_noise_sd` is $\sigma_{\mathrm{subj}}$ (default 0.15). A log-scale
result is exponentiated, and any all-positive covariate is floored above zero.
The final value is repeated on every row of that synthetic subject.

For factor, character, and logical covariates, one available donor category is
sampled using the locally normalized weights. Categories are not averaged.

### A covariate and an endpoint are blended independently

Covariates (this step) and endpoint trajectories (the next) are generated in
separate passes, each with its own donor weighting and its own perturbation.
That is fine when they are unrelated, but some datasets carry a **baseline
covariate that is the same quantity as a longitudinal endpoint** — a `B0`
baseline B-cell count beside a B-cell kinetic endpoint, a baseline biomarker
beside its own time course. AVATAR does not know they are linked, so a synthetic
subject's baseline covariate need not equal the baseline of its own generated
trajectory. The two are individually plausible but not mutually consistent.

This is a known limitation (`REV-022`). It is usually harmless for
workflow-development use, where each column need only be realistic on its own.
If your analysis relies on a covariate agreeing with an endpoint's baseline,
reconcile them after generation, or do not declare the redundant covariate and
read the baseline from the trajectory instead.

## Step 9: synthesize each endpoint trajectory

Only positions that were eligible, nonmissing observations in the anchor
template receive synthesized values. Thus the anchor supplies the number and
placement of observations, including its missing-DV pattern.

For endpoint $e$, donor trajectories are transformed with $g_e$ and
interpolated to the anchor observation times. Let $z_{rj}$ be donor $r$'s
transformed value at target position $j$. The deterministic blend is

$$
\bar z_j=\sum_r w_{rj}^{*}z_{rj}.
$$

If every donor is unavailable at a target, the transformed dataset median for
that endpoint is used and a warning is stored.

The endpoint noise scale $s_e$ is 1 on the offset-log scale. On the identity
scale it is the source endpoint standard deviation, with fallback
$\max(0.1|\operatorname{median}(Y_e)|,0.01)$ when necessary. A single
subject-level shift is shared across all positions of the endpoint:

$$
b_e\sim\mathcal{N}(0,(\sigma_{\mathrm{subj}}s_e)^2).
$$

Within-endpoint residual perturbations follow a stationary-scale first-order
autoregressive, AR(1), process in **observation order**:

$$
\begin{aligned}
\epsilon_{e1} &\sim
  \mathcal{N}(0,(\sigma_{\mathrm{res}}s_e)^2),\\
\epsilon_{ej} &= \phi\epsilon_{e,j-1}+\nu_{ej},\\
\nu_{ej} &\sim
  \mathcal{N}\left(0,
    (\sigma_{\mathrm{res}}s_e)^2(1-\phi^2)\right).
\end{aligned}
$$

Defaults are `residual_noise_sd = 0.05` and `residual_phi = 0.6`. The generated
DV is

$$
Y^*_{ej}=g_e^{-1}(\bar z_j+b_e+\epsilon_{ej}).
$$

Endpoints are processed separately, so PK and PD values cannot be blended
with one another. They nevertheless use the same donor subjects and initial
subject weights, preserving a limited form of cross-endpoint subject
coherence. This is a synthesis device, not a fitted multivariate residual
model.

## A transparent worked example

The small dataset below has six subjects, one dose each, one positive endpoint,
two baseline covariates, and deliberately different observation times. It is
large enough to have non-anchor neighbors but small enough to inspect.

```{r worked-source}
make_subject <- function(id, observation_time, wt, sex, peak) {
  time <- c(0, observation_time)
  evid <- c(1L, rep(0L, length(observation_time)))
  data.frame(
    ID = as.integer(id),
    TIME = time,
    DV = c(0, peak * exp(-0.55 * observation_time) + 0.03 * id),
    AMT = c(100, rep(0, length(observation_time))),
    EVID = evid,
    CMT = c(1L, rep(2L, length(observation_time))),
    MDV = ifelse(evid == 0L, 0L, 1L),
    WT = rep(c(58, 64, 70, 76, 82, 88)[id], length(time)),
    SEX = factor(
      rep(sex, length(time)),
      levels = c("female", "male")
    )
  )
}

worked_source <- do.call(rbind, list(
  make_subject(1, c(0.5, 1.0, 2.0, 4.0), 58, "female", 9.0),
  make_subject(2, c(0.25, 1.5, 3.0, 6.0), 64, "male", 10.0),
  make_subject(3, c(0.75, 1.25, 2.5, 5.0), 70, "female", 11.0),
  make_subject(4, c(0.4, 1.8, 3.5, 5.5), 76, "male", 12.0),
  make_subject(5, c(0.6, 1.4, 2.8, 4.5), 82, "female", 13.0),
  make_subject(6, c(0.3, 1.1, 3.2, 6.5), 88, "male", 14.0)
))
rownames(worked_source) <- NULL

worked_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", mdv = "MDV",
  covariates = c("WT", "SEX")
)

knitr::kable(
  worked_source[worked_source$ID %in% 1:2, ],
  digits = 3,
  caption = "The first two source subjects"
)
```

The next chunk uses internal helpers solely to expose a reproducible teaching
trace. These helpers are not public API. Ordinary analysis code should call
`synpmx_avatar()` rather than depend on them.

```{r worked-trace, echo=FALSE}
worked_seed <- 2026L
# The package default. With six source subjects exactly five donors are
# available, so the floor is met and the default cap is feasible -- at k = 3 a
# cap below 1/3 would flatten the weights to uniform, hiding the very
# randomization this trace exists to show.
worked_k <- 5L
worked_profiles <- synpmx:::.build_profiles(
  worked_source, worked_roles, pca_variance = 0.90
)

set.seed(worked_seed)
anchor_index <- sample.int(
  length(worked_profiles$subjects), 1L, replace = TRUE
)
trace_warnings <- synpmx:::.warning_collector()
worked_donors <- synpmx:::.select_donors(
  anchor_index, worked_profiles, worked_k, trace_warnings
)

trace_summary <- data.frame(
  role = c("anchor", rep("donor", length(worked_donors$indices))),
  source_ID = c(
    worked_profiles$subjects[anchor_index],
    worked_profiles$subjects[worked_donors$indices]
  ),
  profile_distance = c(NA_real_, worked_donors$distances),
  weight = c(NA_real_, worked_donors$weights),
  stringsAsFactors = FALSE
)
```

With `seed = 2026`, the anchor and its selected compatible donors are:

```{r worked-neighbors, echo=FALSE}
knitr::kable(trace_summary, digits = 4)
```

This trace shows the cap of Step 7 doing its work, and shows why the cap is not
optional. One donor is pinned at exactly `max_donor_weight` $=0.50$; without the
cap that donor would have taken more. Notice too that the pinned donor is *not*
the nearest one — the randomized rank $R_r$, not the distance order, decides who
dominates a given avatar, which is what stops the nearest neighbour from being
systematically the largest contributor.

The recorded `mean_effective_donors`, $1/\sum_r w_r^2$, is
`r round(1 / sum(worked_donors$weights^2), 2)` against a floor of
`r worked_k` — the gap between "five donors were used" and "the blend is worth
about `r round(1 / sum(worked_donors$weights^2), 1)` independent donors". That
gap is the honest cost of a cap chosen for simplicity, and it is why
`cap_binding_fraction` and `mean_effective_donors` are both recorded rather than
left to be assumed.

The anchor contributes the event skeleton. Donors contribute transformed DVs
after interpolation to the anchor times. The following table shows the exact
pre-noise blend used by the implementation.

```{r worked-blend, echo=FALSE}
anchor_skeleton <- worked_source[
  worked_profiles$subject_rows[[anchor_index]], , drop = FALSE
]
target_rows <- which(
  synpmx:::.observation_rows(
    anchor_skeleton, worked_roles, require_present = TRUE
  )
)
target_time <- synpmx:::.aligned_time(
  anchor_skeleton, worked_roles
)[target_rows]
worked_transform <- worked_profiles$transforms[["DV"]]

worked_donor_matrix <- vapply(worked_donors$indices, function(donor) {
  trajectory <- synpmx:::.donor_trajectory(
    worked_source,
    worked_roles,
    worked_profiles$subject_rows[[donor]],
    "DV",
    worked_transform
  )
  synpmx:::.interpolate_trajectory(trajectory, target_time)
}, numeric(length(target_time)))
if (is.null(dim(worked_donor_matrix))) {
  worked_donor_matrix <- matrix(worked_donor_matrix, ncol = 1L)
}
colnames(worked_donor_matrix) <- paste0(
  "donor_", worked_profiles$subjects[worked_donors$indices], "_z"
)
worked_blend <- apply(
  worked_donor_matrix, 1L,
  synpmx:::.weighted_available,
  weights = worked_donors$weights
)

worked_synthetic <- synpmx_avatar(
  worked_source,
  worked_roles,
  n_subjects = 1,
  seed = worked_seed,
  k = worked_k
)
worked_synthetic_rows <- synpmx:::.observation_rows(
  worked_synthetic, worked_roles, require_present = TRUE
)

blend_table <- data.frame(
  anchor_TIME = target_time,
  worked_donor_matrix,
  blended_z = worked_blend,
  deterministic_DV = synpmx:::.inverse_dv(
    worked_blend, worked_transform
  ),
  final_synthetic_DV = worked_synthetic$DV[worked_synthetic_rows],
  check.names = FALSE
)
knitr::kable(
  blend_table,
  digits = 3,
  caption = paste(
    "Interpolation and blending for the anchor endpoint; z is the",
    "endpoint working scale"
  )
)
```

`deterministic_DV` is the back-transformed donor blend. `final_synthetic_DV` also
contains the one subject shift and AR(1) residual sequence. The event rows,
new ID, and subject-constant synthesized covariates can be inspected directly:

```{r worked-result}
knitr::kable(worked_synthetic, digits = 3)
```

The result records enough settings to audit the public generator call:

```{r worked-settings, echo=FALSE}
worked_settings <- attr(worked_synthetic, "pmx_settings")
settings_table <- data.frame(
  setting = c(
    "seed", "n_subjects", "k", "pca_variance",
    "subject_noise_sd", "residual_noise_sd", "residual_phi",
    "time_jitter", "alignment", "compatible_event_groups", "routes",
    "on_donor_shortfall", "max_donor_weight", "cap_binding_fraction",
    "mean_effective_donors", "warnings"
  ),
  value = c(
    worked_settings$seed,
    worked_settings$n_subjects,
    worked_settings$k,
    worked_settings$pca_variance,
    worked_settings$subject_noise_sd,
    worked_settings$residual_noise_sd,
    worked_settings$residual_phi,
    worked_settings$time_jitter,
    worked_settings$alignment,
    worked_settings$compatible_event_groups,
    worked_settings$routes,
    worked_settings$on_donor_shortfall,
    worked_settings$max_donor_weight,
    round(worked_settings$cap_binding_fraction, 3),
    round(worked_settings$mean_effective_donors, 3),
    if (length(worked_settings$warnings)) {
      paste(worked_settings$warnings, collapse = "; ")
    } else {
      "<none>"
    }
  ),
  stringsAsFactors = FALSE
)
transform_table <- do.call(rbind, lapply(
  names(worked_settings$endpoint_transforms),
  function(endpoint) {
    transform <- worked_settings$endpoint_transforms[[endpoint]]
    data.frame(
      endpoint = endpoint,
      method = transform$method,
      offset = transform$offset,
      positive = transform$positive,
      stringsAsFactors = FALSE
    )
  }
))
knitr::kable(settings_table, caption = "Recorded generator settings")
knitr::kable(
  transform_table,
  digits = 5,
  caption = "Recorded endpoint transformation"
)
```

### M5 — recomputing the dose from the blended covariate

Runs automatically. Under dosing
proportional to a baseline covariate, the multiplier is a protocol property the
stratum shares and the covariate is individual — and already blended. Keeping the
multiplier and recomputing the amount from the avatar's own covariate stops the
dose being a verbatim real value (which under mg/kg dosing discloses the anchor's
weight exactly) and repairs a coherence defect: previously every avatar violated
the mg/kg rule its own data claims to follow. See `REV-027`.

**The post-generation screen. Screening the finished data, and remediating** (`flag_identifiable_subjects()`,
after generation). The only mechanism `synpmx_avatar()` does **not** run for you.
It scores each synthetic subject on four axes — follow-up time, dose count, dose
magnitude, peak DV — and `remediate_identifiable_subjects()` truncates, drops, or
replaces what it flags, refilling the cohort so the size is preserved. M2 keeps extreme *source* patients from becoming anchors; this catches whatever
still looks implausible once everything else has run. Run it before the data leaves the source's
access controls.

## Step 10: restore PMX schema and deterministic conventions

After all synthetic subjects are row-bound, the original column order is restored.
Each result column is converted toward its source class:

- factors recover source levels and ordering; new factor IDs add new levels;
- integer columns are rounded and converted to integer;
- double, logical, character, Date, and POSIXct columns are restored; and
- a non-`data.frame` source class, such as a tibble class vector, is reapplied.

New numeric IDs start above the source maximum. New character and factor IDs
use labels such as `syn_001`, with a prefix added repeatedly if needed to
avoid collision.

If MDV is declared and the entire source obeys the standard relationship

$$
[\mathrm{MDV}=0]=[\mathrm{EVID}=0\ \text{and DV is present}],
$$

MDV is re-derived from that rule in the synthetic data and restored to its original
class. For any nonstandard source-wide MDV convention, anchor values are left
as copied. *This step* neither invents nor removes observation positions --
which ones an avatar has was already settled by mechanism 5 above.

Finally, `validate_pmx(..., strict = TRUE)` is run on the assembled result. A
generation call cannot silently return an output that fails the package's own
structural requirements.

## Reproducibility and diagnostics

All random operations occur inside a local seed scope. `seed` must be an
integer between zero and `.Machine$integer.max`. The caller's RNG kind and
`.Random.seed` are restored on exit, whether or not `.Random.seed` existed
before the call. Consequently, identical inputs and arguments produce an
identical complete result without consuming the surrounding analysis RNG
stream.

`attr(synthetic, "pmx_settings")` records arguments, explicit roles, endpoint
transformations, the alignment description, the number of compatible event
groups and distinct routes, the donor weight cap and how often it bound, the
mean and minimum effective donor count, and unique fallback warnings. It does not currently record the
anchor ID, donor IDs, distances, or realized weights for each synthetic subject.

Use the public checks for different questions:

- `validate_pmx()` asks whether one dataset is structurally usable under the
  declared roles.
- `compare_pmx()` reports source/synthetic row counts, subjects, event counts,
  endpoints, column classes, validations, and exploratory trajectories when
  `ggplot2` is installed.
- `compare_pmx_distributions()` summarizes the dependent variable per endpoint
  and each baseline covariate, source against synthetic, so their ranges can be
  compared at a glance.

None of these establishes distributional equivalence, privacy, or model
fidelity.

## After generation: screening for identifiable subjects

### It is a plausibility check, not a masking step

`flag_identifiable_subjects()` and `remediate_identifiable_subjects()` exist, run
after generation, and are **not** in the table above. That is a deliberate
demotion, and worth explaining because an earlier version of this article listed
them as a seventh mechanism.

The screen scores each finished avatar on four axes — follow-up time, dose count,
dose magnitude, peak DV — and flags outliers on a robust median/MAD statistic.
The argument for calling it masking was that blending might *create* a
conspicuous avatar even from unremarkable donors. On inspection that argument
does not hold up:

- **On the two structural axes, M2 has already done the work.** An avatar's
  follow-up length and dose count come from its anchor, unchanged. If no extreme
  patient was used as an anchor, no avatar has an extreme skeleton.
- **On the two value axes, being conspicuous does not identify anyone.** A high
  peak DV in an avatar is a blend of several donors plus noise. It can exceed
  every real value in the study precisely *because* it belongs to nobody. An
  attacker who spots it learns that the generator added noise.

What it does catch is **implausibility**: an avatar that looks wrong sitting next
to the rest of the cohort. That is worth catching — output that looks extreme
undermines confidence in the whole dataset — but it is a realism check, not a
privacy one.

**And the two use different statistics, which is a trap.** M2 cuts at twice the
cohort's 90th percentile, deliberately loose, so only gross outliers are excluded.
The screen uses median/MAD at 3.5, which is much tighter and behaves badly on a
bimodal cohort. On `mavoglurant` — half the patients have one dosing period and
half have two — the MAD is tiny and **75 of 120 avatars are flagged**, on
follow-up time and dose count, the very axes M2 already screened. Almost none of
that is exposure; it is the statistic reacting to a two-humped distribution.

That number is also the answer to "why not just run it automatically": on this
dataset it would discard 63% of the cohort, for a reason that is largely an
artefact. Run it, read what it flags, and decide — which is why it takes a
threshold argument and why nothing runs it for you.


The blending in Steps 6–9 protects the *measurements*: every avatar's DV values
are averaged from at least `k` real subjects, so no one person's numbers survive
verbatim. It does **not** protect the *structure*. Stage (d) copies the event
skeleton — the number of doses, their sizes, and the observation times — from a
single anchor. So a source subject with an unusual structure (a lone
long-followed patient, a one-off dose level) produces an avatar with that same
unusual structure, and a structural oddity is enough to single someone out even
when their values are blended.

Structural exposure comes in two forms, and they need different remedies.

A subject can be **extreme** — followed far longer than anyone else, dosed at a
level nobody else received. The screen and the two functions below handle that.

A subject can also be **alone** without being extreme at all. Sitting squarely
in the middle of every distribution is no protection if nobody else shares your
exact vector of observation times, and under actual recorded times — as opposed
to nominal, protocol-scheduled ones — that is very nearly everybody. Verbatim
copying then reproduces one real patient's visit pattern exactly. This is the
case `coarsen_time = TRUE` addresses, by collapsing every subject onto a shared
visit grid *before* any avatar is built and resampling pooled deviations back
afterwards; `skeleton_uniqueness()` is how you see how much of it there was and
how much survived. Note that `time_jitter` cannot substitute: every jittered
time is clamped inside its own Voronoi cell, so at any magnitude a visit stays
within half a gap of where the source subject's visit was.

Two functions address the extreme case after the data is generated.

`flag_identifiable_subjects()` scores every subject, one axis at a time, with a
robust median/MAD outlier statistic (the Iglewicz–Hoaglin modified z-score,
default cutoff 3.5) on four structural features:

- **follow-up time** — the last observation time, which catches the lone
  long-followed subject;
- **number of doses** — an unusual dosing-history length;
- **dose magnitude** — a rare dose level; and
- **DV value** — an extreme peak measurement.

A subject is flagged when it is an outlier on any axis, and the report names the
axes. Scoring is per subject on one summary per axis, so a single long-followed
patient is one flag on the time axis — not one flag per late row — and one
extreme patient is one flag on the DV axis. The screen is a heuristic, not a
privacy proof: a flag is a candidate to remove, and an empty list is not a
certificate of anonymity.

`remediate_identifiable_subjects()` acts on those flags. Its default policy
splits by what a value-level edit can honestly fix:

- a subject flagged **only** for a long follow-up is **truncated** back to the
  cohort's longest ordinary follow-up — shortening a long timeline leaves a
  shorter but ordinary subject;
- a subject flagged for **any other** reason is **dropped**, because an
  extreme-DV subject is elevated throughout (trimming points only mangles it)
  and a rare dose cannot be trimmed without breaking the regimen;
- a subject flagged for both is dropped, since truncation would not resolve the
  other reason.

When a `source` is supplied, each dropped subject is **replaced**: fresh avatars
are generated, screened by the same policy, and appended with new ids until the
cohort is back to its original size — so screening does not shrink the dataset.
The `time` and `other` arguments expose the policy.

This detect-and-remediate path is a deliberate alternative to preventing
structural outliers at generation time. A more thorough approach would sample
each avatar's event skeleton from the cohort, so no avatar carries any one real
subject's unique structure; that is a possible future direction, not current
behavior.

## Edge cases and documented fallbacks

The implementation favors an explicit warning and structurally valid output
over a silent incompatible blend.

```{r edge-cases-table, echo=FALSE}
edge_cases <- data.frame(
  Situation = c(
    "Fewer than k same-signature donors",
    "Route arm smaller than the k-donor floor",
    "Every route arm below the floor",
    "Cap below 1/K for K donors",
    "Source smaller than the k-donor floor",
    "All selected distances essentially zero",
    "One donor time only",
    "No donor value at a target",
    "No variable profile feature",
    "Identical donors with all noise disabled",
    "Nonstandard MDV convention"
  ),
  Behavior = c(
    "Borrow the nearest donors from other dose/schedule groups on the same route",
    "Follow on_donor_shortfall: drop (default), noise, or error; alert loudly",
    "Generate anyway and raise a loud alert; treat output as identifying",
    "Relax the cap to 1/K, i.e. uniform weights, rather than erroring",
    "Blend all available donors and raise a loud alert",
    "Use epsilon-stabilized randomized weights",
    "Repeat that donor's mean at every target time",
    "Use transformed endpoint median and record a warning",
    "Use a shared zero coordinate; distance cannot separate subjects",
    "A source-shaped trajectory may be mathematically unavoidable; warn",
    "Copy anchor MDV values rather than re-derive them"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(edge_cases, align = c("l", "l"))
```

Other boundaries are deliberate. `synpmx_avatar()` supports only
`event_method = "template"` and `dv_method = "avatar_blend"`. It does not fit
a structural model, reconstruct covariate-parameter relationships, simulate
occasions, enforce a dosing grammar beyond the sampled template, or provide a
censoring model. Fitting an explicit structural model is the job of
`synpmx_calibrated()`, described below.

## Relationship to AVATAR

The method is **AVATAR-inspired**, not an implementation of published AVATAR
software. "AVATAR" is a method name rather than an initialism, from the
patient-centric *avatarization* literature in which each synthetic record is
built from the local neighborhood of real records. The shared family resemblance
is the use of standardized subject features, PCA, local neighbors,
inverse-distance influence, exponential randomization, and rank attenuation.
Destere and colleagues describe a modified AVATAR benchmark for longitudinal
population pharmacokinetic (PopPK) data that first widened the data by subject
and then used PCA, K-nearest neighbors, inverse Euclidean distance, exponential
stochastic weights, and rank attenuation before reverse transformation [1]. The original patient-centric AVATAR method was reported by
Guillaudeux and colleagues [2].

`synpmx` adapts those ideas to a PMX event table in several material ways:

- it preserves a sampled longitudinal event template rather than generating
  all fields from a wide vector;
- route of administration is an absolute barrier, while dose and schedule
  differences only make a donor less preferred;
- endpoints are transformed and interpolated separately;
- the same donors support covariates and endpoints;
- every multi-donor weight, not only the dominant one, is capped at
  `max_donor_weight` (default 0.50);
- subject and AR(1) perturbations are added on the endpoint working scale; and
- there is no inverse-PCA reconstruction of a full synthetic patient vector.

Destere et al. also demonstrate why method labels are not validation: their
benchmark found materially different PK parameter and residual-error biases
across synthesis algorithms [1]. That paper is a 2026 medRxiv preprint and was
not peer reviewed at the version supplied with this package repository. Its
privacy analyses do not transfer to `synpmx`; this package has undergone
no corresponding attack-based privacy validation.

## Algorithm summary

What `synpmx_avatar()` actually executes, in order. These are implementation
operations, finer-grained than the ten Steps and not in one-to-one
correspondence with them, so the last column says where each is explained rather
than leaving two numberings to be guessed at.

| # | Operation | Explained in |
|---|---|---|
| 1 | Validate source data, roles, and generator arguments | Step 1 |
| 2 | Compute first-event-relative time and endpoint-specific transformations | Steps 3–4 |
| 3 | Assemble baseline covariates and interpolate transformed endpoint trajectories into one profile per source subject | Step 5 |
| 4 | Median-impute profile features, remove unusable features, standardize, and retain enough PCA components to meet `pca_variance` when possible | Step 5 |
| 5 | Construct event/schedule signatures and route keys | Step 6 |
| 6 | Resolve anchors whose route arm cannot reach the donor floor per `on_donor_shortfall`, apply the structural screen, and sample the remaining anchors with replacement | Step 6 |
| 7 | For each anchor, copy its whole event template | Step 2 |
| 8 | Find `k` donors on the anchor's route — exact-signature nearest-first, then the nearest remaining route-compatible subjects — and randomize their inverse-distance weights, water-filling every weight to `max_donor_weight` | Steps 6–7 |
| 9 | Generate subject-constant covariates from those donors | Step 8 |
| 10 | For each endpoint, interpolate donor trajectories to anchor times, blend, add a subject shift and AR(1) perturbations, and back-transform | Step 9 |
| 11 | Re-derive standard MDV when applicable, assign a new ID, preserve tie order, restore the source schema, and validate the assembled result | Step 10 |
| 12 | Attach `pmx_settings` and emit any collected fallback warnings | Step 10 |

## How you know whether it worked

Because there are no formal guarantees with this method, after the synthetic dataset is created, it is useful to check back whether the synthetic patients can point back to the real patient.  There are there different tests for this: 

**Unique, extreme, or close.** A patient is *unique* when something about their
schedule belongs to nobody else. A patient is *extreme* when they stand out from
the cohort — the lone very long follow-up. A synthetic patient is *close* when
their measurements landed nearer a real patient than real patients lie to each
other. These are independent failures: a study can pass any one and fail
another, so all three are reported and none substitutes for the rest.

| Function | Asks | Run on | Target |
|---|---|---|---|
| `skeleton_uniqueness()` | Is anyone's **DV observation schedule** unique? Separately, is anyone's **dosing** unique? | the **source**, before generating | 0 unique |
| `flag_identifiable_subjects()` | Does anyone **stand out** from the cohort on follow-up length, dose count, dose size, or peak DV? | the **synthetic output** (or the source) | nothing flagged |
| `compare_pmx_proximity()` | Did a synthetic patient's **covariates and DV trajectory together** land nearer a real patient than real patients lie to each other? | the **pair**, after generating | near 0.5 |


**Sampling and dosing are scored separately** `skeleton_uniqueness()` builds two different keys per patient:

- the **observation time vector** — the sorted list of times that patient was
  *sampled* at. Dose rows are excluded. This is what `unique_schedule_n` counts
  and what `coarsen_time` acts on.
- the **event signature** — the dosing side: how many doses, their amounts and
  rates, the intervals between them, and which endpoints were measured.
  Observation times are deliberately *not* in it. This is what
  `n_unique_dose_signature` counts, and **coarsening does not touch it**: a
  study with weight-based dosing or per-patient titration leaves patients unique
  on dose however good the visit grid is. `REV-027`'s dose recomputation is the
  mechanism that addresses that side.

So a patient can be unique on sampling, on dosing, on both, or on neither, and
the two have different remedies.

**"Values" means the whole profile, not just DV.** `compare_pmx_proximity()`
compares patients in the space `.build_profiles()` builds, which is each
patient's **baseline covariates together with their DV trajectory** on a common
grid, reduced to principal components. So a synthetic patient can be flagged as
too close because their weight and age nearly match a real patient's, not only
because their concentrations do.

### The exposure counts

`skeleton_uniqueness()` produces the numbers that `synpmx_avatar()` also records
in `attr(x, "pmx_settings")`. There is **one headline count, split by cause**,
and the split is what tells you what to do next:

- **`unique_schedule_n`** — patients whose list of observation times no other
  patient shares. An avatar anchored on one wears a schedule belonging to
  exactly one real person, however carefully its measurements were blended. This
  is the number to drive to zero. It always equals the sum of the next two.
  - **`unique_obs_time_n`** — of those, the ones sampled at a clock time
    nobody else was sampled at. **A grid can fix this**, and `coarsen_time`
    exists to; a count above zero after coarsening means the inferred grid could
    not find a shared schedule, and declaring `nominal_time` is the fix.
  - **`unique_visit_set_n`** — of those, the ones who share every individual
    time with somebody but attended a combination of visits nobody else did:
    dropout, an interruption, a missed week. **No grid can fix this** at any
    resolution, because the times are already shared. `min_pattern_share`
    addresses it during generation; otherwise the response is to remediate or
    accept.

Two counts sit alongside these and answer narrower questions:
`n_unique_dose_signature` (a dose structure or amount nobody else has, which
coarsening cannot touch) and `n_unique_obs_count` (an observation count nobody
else has, the residual left for the outlier screen).

### Why measure rather than assert

Every mechanism in this article costs something — discarded visit sets, excluded
anchors, timing fidelity — and what it buys **depends entirely on the study**.
The same defaults take `theo_md` from twelve unique schedules to zero and leave
`nimoData` at twelve of twelve, because one study recorded a shared visit
structure and the other did not. There is no headline "synpmx removes X% of the
exposure", and any such claim would be about the datasets it was measured on
rather than about yours.

So every run reports its own numbers, loudly, and
`scripts/explore_nimodata_coarsening.qmd` works one failure through to a fix to
show what reading them looks like in practice.

## Glossary

Collected here rather than up front, since each term is introduced where it
first matters. This is for looking one back up.

| Term | Meaning |
|---|---|
| **dose event** | A row where `EVID` is not 0: when drug was given, how much, at what rate. |
| **DV observation** | A row where `EVID` is 0 and a measurement is present: when the patient was observed and what came back. |
| **avatar** | One synthetic patient in the output. |
| **anchor** | The single real patient whose **event skeleton** an avatar copies. Each avatar has exactly one, and it is chosen by sampling with replacement from the pool of eligible source patients. |
| **event skeleton** | *When* things happened to a patient, with no measured values: the dose events and the times of the DV observations. This is what gets copied from the anchor; the values that fill it do not. |
| **donor** | A real patient whose *measurements* contribute to an avatar. Each avatar blends at least `k` (default 5) of them, and no single donor may exceed `max_donor_weight` (default 0.50) of the blend. An avatar's anchor is usually also one of its donors, but the two roles are separate: the anchor supplies structure, donors supply values. |
| **visit set** | Which of the study's shared visits one patient actually had — the pattern of attended and missed DV observations, once every patient is on a common grid. |

## References

1. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).

2. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).
