---
title: "Demo: Using synpmx_avatar"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx_avatar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
run <- has_xgxr
plots <- has_xgxr && has_ggplot2
# `compare_pmx_distributions()` composes its panels with patchwork, so its
# figure needs one package more than the trajectory plots do, and refuses to
# draw without it rather than degrading. Guarding the chunk is what keeps a
# missing suggested package a missing figure instead of a failed build.
distributions <- plots && requireNamespace("patchwork", quietly = TRUE)
```

Generate a synthetic dataset, look at it, and read the checks of the data, sumamrized in a scorecard. 

`vignette("avatar-scorecard")` is the reference for better understanding the synthetic data checks.  

The dataset is `xgxr::case1_pkpd`: 180 patients, six arms from placebo to 300 mg, with a pharmacokinetic (PK) concentration endpoint and a continuous pharmacodynamic (PD) endpoint.

## Configuration

To run this on your own study, edit the two chunks below that:

1. Read in the dataset.
2. Define the column roles. 

The rest of the code can be kept as is.

```{r data, eval=run, message = FALSE}
library(dplyr)
raw <- as.data.frame(get(utils::data(list = "case1_pkpd", package = "xgxr"))) |>
  mutate(CENS = ifelse(NAME == "PD - Continuous", 0, CENS))
SEED <- 808
```

The column meanings.  Columns not specified here are dropped from the synthetic dataset.

```{r roles, eval=run}
roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  cens         = "CENS",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",        # which endpoint each observation row is
  nominal_time = "NOMTIME",     # the protocol grid
  strata       = c("TRTACT", "DOSE"), #treatment arms.
  covariates   = "WEIGHTB",     # blended like everything else
  keep         = "STUDY"        # carried through verbatim
)
```

## Generate Synthetic Data

```{r synthesize, eval=run, message = FALSE}
synthetic <- synpmx_avatar(raw, roles, seed = SEED)
```

## Plot synthetic data and original data

```{r overlay-setup, eval=plots}
library(ggplot2)
library(xgxr)
xgx_theme_set()
comparison_colours <- c(source = "#1B6CA8", synthetic = "#D95F02")


both <- rbind(
  cbind(raw[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$EVID == 0 & !is.na(both$LIDV), ]
obs$TRTACT <- factor(obs$TRTACT,
                     levels = c("Placebo", "3 mg", "10 mg", "30 mg",
                                "100 mg", "300 mg"))
```

```{r overlay-pk, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PK Concentration" & obs$TIME <= 24, ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA~TRTACT) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 24, by = 6)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "PK concentration", colour = NULL) +
  theme(legend.position = "top") + 
  ggtitle("Day 1 Conc. Profile")
```

```{r overlay-pd, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PD - Continuous", ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.35) +
  xgx_scale_x_time_units(units_dataset = "hours", units_plot = "weeks") +
  facet_grid(DATA~TRTACT) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "PD (continuous)", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("PD Response")
```

## Distributions of Synthetic and Original Data

One panel per endpoint and per baseline covariate, source against synthetic.
`compare_pmx_distributions_height()` sizes the figure, since it grows a row of
panels at a time; `output = "tables"` gives the numbers behind it.

```{r distributions, eval=distributions, fig.height=compare_pmx_distributions_height(raw, roles)}
compare_pmx_distributions(raw, synthetic, roles)
```

The curves sit on top of each other on both endpoints. Look at their widths
rather than their positions: the orange curve is visibly narrower than the blue
one for the PD endpoint and for baseline weight, and slightly narrower for PK.
That narrowing is between-subject variability shrinking, which is what the
AVATAR blending algorithm does rather than a fault in the run. `output =
"tables"` puts a number on it, and the scorecard's D1 row reports the variable
that moved furthest.

## The scorecard

The scorecard computes checks of the synthetic dataset.  It is described further in `vignette("avatar-scorecard")`.

```{r scorecard, eval=run}
scorecard <- synpmx_scorecard(raw, synthetic, roles)
synpmx_scorecard_datatable(scorecard)
```

**`review` is not a soft `pass`.** Five rows here have no pass mark because no
threshold would be honest. Doses per patient is the clearest: on this study it
is unchanged, but on a study with individualised dosing it can halve.

**Every row names the call that explains it.** When a row reads oddly, run what
its `explore` column says: the numbers are a summary, and the call behind each
one is where the answer is.

## Where to go next

- `vignette("avatar-scorecard")` — every check, what it asks, and what
  passing means.
- [Evaluating AVATAR on public data](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html) - AVATAR algorithm applied to 8 public datasets.
