---
title: "The synpmx AVATAR Algorithm"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The synpmx AVATAR Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
library(synpmx)
```

This article is the full specification of the AVATAR synthetic data generator applied
to pharmacometrics modeling datasets.  The original method [1] was extended to population
pharmacokinetic (PK) datasets in [2], and here further extensions are provided
so that it works well with more realistic datasets.

AVATAR makes no formal privacy claim. It also does not preserve the dose-exposure
relationship.  Use the synthetic datasets to develop and debug analysis code. Do not use
it to estimate parameters.

Because modeling datasets are generally longitudinal, one set of identifying features are
the unique dose or observation times each patient will have.  The default setting of 
`synpmx_avatar` ensure that the set of observation and dose times of the synthetic patient
does not match any one real patient data.   

## Overview of Algorithm

Each synthetic patient is called an avatar.  Each avatar is built from one real 
**anchor** subject's dosing event structure, then filled with measurements blended from several similar real subjects, called **donors**. The generator works on any number of endpoints, so one study can
carry multiple PK profiles alongside multiple pharmacodynamic (PD) measures.

Every patient's data can be thought of as containing two pieces of information which are treated differently

* **event skeleton** is *when* the dose events and the DV observations
happen. 
* **values** are the DV numbers filling those observation slots, plus
the baseline covariates. 

The event skeleton is copied from one anchor patient with some
masking of the times, and the values are blended across several donors. 
The `synpmx_avatar` function runs the following steps.

1. **Declare the columns.** The `roles` variable say which column is the id, time,
   dose, measurement, etc. Only named columns survive.
2. **Coarsen and align the times.** Snap every recorded time onto a shared visit
   grid (often based on nominal time), and align each subject to its own first dose.
3. **Choose an endpoint transformation and value type.** Decide what "average"
   means for each endpoint, and whether it is continuous or discrete.
4. **Build one numeric profile per source subject**, and reduce it to a handful
   of principal-component coordinates that measure similarity between patients.
5. **Decide who may donate data to whom**, by reading each subject's route of
   administration and dosing history.
6. **Restrict the anchor pool and sample anchors**, keeping each cohort (arm) the
   same size (except for very small cohorts)
7. **Copy the anchor's rows** and redraw the parts of its schedule that identify
   it: every TIME, EVID, AMT, RATE and CMT is copied, then the visit set and the
   stopping point are replaced.
8. **Choose the donors**, nearest first within the anchor's route.
9. **Randomize and cap the donor weights**, so no one real patient is more than
   half of any avatar.
10. **Synthesize the baseline covariates** from those donors.
11. **Synthesize each endpoint trajectory**, then correct the generated value:
    snap a discrete endpoint back onto its levels, and reapply the censoring
    boundary.
12. **Recompute the dose** from the avatar's own blended covariate, where dosing
    is proportional to one.
13. **Restore the PMX schema and validate**: column order, types, factor levels,
    the container class, fresh subject IDs, and a record of every choice made.

## The Masking Mechanisms

Cutting across the steps are six **masking mechanisms**, numbered in the order
they run. Each is explained in full at the step that fires it.

- **M1. Coarsen the times.** Snap every time onto a shared visit grid, then add
  deviations back from a cohort-wide pool. This prevents identifying a patient
  from their TIME vector. *Step 2.*
- **M2. Restrict who may be an anchor.** Drop patients whose structure no amount
  of blending would hide, such as a patient followed twice as long as everyone
  else, and patients in a route arm too small to blend within. *Step 6.*
- **M3. Redraw which visits the avatar attended.** Reselect which DV values were
  collected at which time points, in case a patient has a unique set of missed
  visits. *Step 7.*
- **M4. Redraw where dosing stopped.** Where a patient's dose schedule is one
  nobody else has, the avatar stops earlier, at the deepest opening several
  patients share. Dose *times* are never moved or invented, so this only ever
  truncates a real schedule earlier. *Step 7.*
- **M5. Blend the values across several donors.** Covariates and DV, capped so
  no one donor dominates, plus noise. *Step 9.*
- **M6. Recompute the dose from the avatar's own covariate.** Done when dosing
  is weight-based or body-surface-area based, so that subjects do not carry
  uniquely identifying doses. Declared with `dose_covariate`, or inferred.
  *Step 12.*

**The post-generation screen is not a seventh mechanism.**
`flag_identifiable_subjects()` scores each finished avatar on follow-up time,
dose count, dose magnitude and peak DV, and it is a realism check rather than a
masking step. M2 has already handled the two structural axes, since an avatar's
follow-up length and dose count come from an anchor that was screened. On the
two value axes, being conspicuous identifies nobody: a high peak DV is a blend
of several donors plus noise, and it can exceed every real value in the study
precisely because it belongs to no one.

## Step 1: Declare the Meaning of the Columns

The package never guesses critical pharmacometric (PMX) roles from column names.
The user must declare them with `pmx_roles()`:

```{r roles-example}
roles <- pmx_roles(
  id = "ID",
  time = "TIME",
  dv = "DV",
  amt = "AMT",
  evid = "EVID",
  cmt = "CMT",
  dvid = "DVID",
  mdv = "MDV",
  rate = "RATE",
  nominal_time = "NTIME",  # protocol-scheduled time, if the study records it
  cens = "CENS",           # below-limit indicator, if the study has BLQ data
  limit = "LIMIT",         # the other end of an interval, when used
  covariates = c("WT", "AGE", "SEX"),
  dose_covariate = "WT",   # dose is mg/kg; say so rather than let it be inferred
  strata = "ARM",          # assigned arm; groups two mechanisms, see below
  keep = "STUDYID"         # carried verbatim; see the allowlist below
)
roles
```

A **dose event** is a row where EVID is not zero: when drug was given, how much,
at what rate. A **DV observation** is a row where EVID is zero and a measurement
is present. Rows are eligible observations when EVID is zero and, if MDV is
declared, MDV is also zero. A nonmissing DV is additionally required when a
value is used to build a profile or a donor trajectory. In symbols, with bracket
notation for a logical condition,

$$
O_{ij} = [\mathrm{EVID}_{ij}=0]
         [\mathrm{MDV}_{ij}=0\;\text{if MDV is declared}]
         [Y_{ij}\;\text{is present}].
$$

Rows with EVID equal to zero but missing DV, or with nonzero MDV, remain in the
sampled template but do not contribute a measurement value.

**`dose_covariate` is worth declaring wherever it applies.** It names the
covariate the dose is a fixed multiple of, and it is the one role that replaces
an *inference* rather than adding information the package could not have. Left
`NULL`, M6 has to work out for itself whether the study is dose-proportional,
and it does that conservatively on purpose: the dose-to-covariate ratio must
collapse onto a handful of levels within 2%, because recomputing amounts on a
study that is not proportional would be a worse error than leaving them alone.
Two entirely ordinary things defeat that test, dispensing in vials and patients
escalating by factors that are not identical across the cohort, so a genuinely
weight-based study can come through with M6 silently doing nothing. Declaring
the covariate also changes *how* the amount is rebuilt, since each dose row's
own ratio is held rather than snapped to a shared level, so intra-patient
escalation survives exactly.

`id`, `time`, `dv`, and `evid` are required; the rest are optional. `amt`,
`cmt`, `mdv`, `rate`, `nominal_time`, `tad`, `occasion`, `addl`, `ii`, `cens`,
and `limit` each name one column. `dvid` names one column, or several that
encode the same endpoint. `covariates`, `strata` and `keep` each take any number
of columns, and `endpoint_types` takes a named vector declaring what kind of
values an endpoint holds (Step 3). One source column cannot be assigned to two
roles. The roles `assigned_dose` and `exclude` belong to the differentially
private engines only; `synpmx_avatar()` rejects them and points at the role that
does the job here.

### The Role Declaration Is the Allowlist

`synpmx_avatar()` keeps only columns a role names, and **drops every other
column**, reporting which ones. This is deliberate. AVATAR builds each avatar
from one real anchor subject, and any column it retains that it does not
actively synthesize is copied from that anchor verbatim. A secondary identifier
such as `USUBJID`, a site number, or a randomization date, left undeclared,
would otherwise ride a real subject's real value straight into the synthetic
data. Dropping by default means a column you forget fails safe, toward removal,
rather than leaking.

That leaves five ways a column is treated, and the difference matters:

| Declared as | What AVATAR does to it | Use for |
|---|---|---|
| `dv` (with `cens`/`limit`) | Blended across donors into a new trajectory | the measurement |
| `covariates` | Blended or resampled across donors into a new value | baselines you want *synthesized*: weight, age |
| `strata` | Copied verbatim from the anchor, **and used to group three mechanisms** | assigned strata: treatment arm, dose group, cohort |
| `keep` | Copied verbatim from the anchor, and otherwise inert | assigned values you want kept faithful to that subject's dosing, such as a study identifier |
| *(undeclared)* | Dropped | anything you do not need |

`covariates` and `keep` are opposites, and choosing wrongly is the common
mistake. A **covariate is blended**: the synthetic subject gets a genuinely new
value interpolated from several neighbours, decoupled from any one person. A
**kept column is copied**: the synthetic subject gets one real subject's real
value, unchanged. For a treatment arm, a dose group, or a randomization
sequence, anything that must stay consistent with the doses AVATAR copied from
the same anchor, `keep` is correct, precisely because it never leaves that
anchor's side. Because a kept value is a real subject's real value, output
containing it must stay within the source data's own access controls and
obligations.

**`strata` carries a column the same way `keep` does, and then uses it.** Both
copy the anchor's value verbatim, so the choice between them is not about what
appears in the output. It is about whether that column should also *partition
the cohort*. A `strata` column must be constant within a subject, and the
combination across such columns defines a stratum that three mechanisms group
by:

- **M3** draws each avatar's visit set from patterns held by other patients *in
  the same stratum*, so an avatar on a weekly schedule is never handed a daily
  arm's pattern.
- **M6**, the dose recomputation, looks for a covariate-proportional dose
  *within each stratum*. A study dosing one arm at 1 mg/kg and another at
  2 mg/kg only looks proportional once the arms are separated; pooled, the
  ratios do not collapse onto a few levels and detection fails closed.
- **Anchor sampling** gives each stratum its source share of the synthetic
  cohort, rather than leaving the balance to chance (Step 6).

It is deliberately **not** a blending barrier. Donors are still borrowed across
strata to reach the floor, since only route of administration blocks that. So
declare a treatment arm as `strata` rather than `keep` whenever the arms differ
in dose rule or visit schedule, and use `keep` for a label that carries no such
structure.

Redundant endpoint labels have their own handling. A dataset that carries both a
numeric `YTYPE` and a character `NAME` for the same endpoint declares **both**
as `dvid`: `dvid = c("YTYPE", "NAME")`. The first is the grouping key.
Validation checks that the rest are a consistent 1:1 mapping with it, catching a
source where the two labels disagree, which `keep` would carry through silently,
and AVATAR carries all of them into the output, aligned.

Immediately before generation, `synpmx_avatar()` calls `validate_pmx(...,
strict = TRUE)`. Among other checks, validation requires:

- nonmissing IDs and EVID values;
- finite numeric TIME values and nondecreasing TIME within subject;
- numeric DV and finite nonmissing observations;
- an endpoint value on every eligible observation when DVID is declared;
- multiple `dvid` columns to be a consistent 1:1 mapping;
- every declared baseline covariate to be constant within subject; and
- when `cens` is declared, each CENS flag to agree with its DV. A left-censored
  value at the limit cannot exceed an ordinary measurement of the same endpoint.
  A pooled dataset with two assays at different limits will trip this
  legitimately; synthesize one study at a time, or drop the `cens` role.

Every error names the role and the column it maps to, and the count and an
example of the offending rows, so a failure says exactly what to fix.

## Step 2: Coarsen the Times and Align to the First Event

### M1. Coarsening the Times, and Putting the Scatter Back

At the default `coarsen_time = TRUE`, source times are collapsed onto a shared
visit grid, and the per-visit deviations are pooled across the cohort and
resampled independently onto each avatar. This protects the **schedule**, which
blending never touches: the skeleton is copied verbatim from one anchor, and
under actual recorded times almost every subject holds the only copy of their
visit vector.

M1 is the only mechanism that both removes something and then restores part of
it, so the two halves are separate.

**Snapping**, before any avatar exists, replaces each recorded time with its
visit's grid time. The deviation removed is not discarded. It goes into one pool
shared by the whole cohort.

**Restoring**, while each avatar is built, draws a deviation from that pool and
adds it back. Without this the output would sit exactly on a tidy grid, which no
real study does, and any analysis code that reconciles actual against nominal
time would have nothing to reconcile.

The restoration is **not itself masking**, and it does not undo the masking
either. That distinction is the whole point: an avatar gets a deviation drawn
from the cohort, never the one its own anchor had. The scatter comes back; the
fingerprint does not.

Each drawn offset is clamped inside its own visit's cell, so no time moves more
than half the gap to the neighbouring visit, and resampling can neither reorder
visits nor stretch follow-up. A dose event and a DV observation recorded at the
same time receive the same offset and stay together.

The order is the mechanism. Snapping is many-to-one and *destroys* the
deviation, whereas perturbing a time in place leaves it recoverable. This is why
`time_jitter` is not an alternative at any magnitude: its clamp holds every time
inside its own Voronoi cell, so a visit stays within half a gap of the source
patient's own value and the source schedule stays recoverable. Jitter is a
realism control. M1 is the privacy one.

Where the study records a protocol-scheduled time, declaring `nominal_time`
gives coarsening the grid outright instead of making it infer one.

### First-Event-Relative Time

Neighbor profiles and donor trajectories are constructed on an internal aligned
time. For subject $i$, the preferred origin is the first nonzero-EVID row having
a positive AMT:

$$
t_{0i} = \min\{t_{ij}: \mathrm{EVID}_{ij}\ne0,\ \mathrm{AMT}_{ij}>0\}.
$$

If no positive-AMT event exists, the first nonzero-EVID row is used. If the
subject has no event row, TIME is left unchanged. The internal aligned time is

$$
t^*_{ij}=t_{ij}-t_{0i}.
$$

This removes a subject-specific calendar or study-time offset. It does **not**
compute time after each dose.

### TIME Is Not TAD

A conventional time-after-dose (TAD) variable resets after every dose. AVATAR
does not use a declared `tad` role for alignment and never recomputes a TAD
column. With multiple doses, $t^*$ continues to increase from the first
qualifying dose; there is no later reset. A declared `tad` column is retained
and copied from the anchor template unchanged; an undeclared one is dropped, as
any undeclared column is.

This has two practical consequences:

1. Use the actual analysis-time column as `time`. Do not expect a TAD column to
   drive neighbor construction or interpolation.
2. If `time_jitter > 0`, a declared TAD column is copied, not recalculated, and
   can become inconsistent with the jittered TIME. Recompute TAD downstream if
   the workflow requires it.

### Optional Coherent Time Jitter

The default `time_jitter = 0` copies template times exactly. Otherwise, one
normal perturbation is drawn for each unique template time $u_l$,

$$
u'_l=u_l+\delta_l, \qquad
\delta_l\sim\mathcal{N}(0,\sigma_t^2),
$$

and constrained between the midpoints of adjacent nominal times. Nonnegative
time is enforced at the first time. All rows tied at $u_l$ receive the same
$u'_l$, so a dose and observation tied at a nominal time remain tied and time
ordering cannot cross.

## Step 3: Choose an Endpoint Transformation and Value Type

Two decisions are made per endpoint here, and both are recorded on the output.
The transformation decides what "average" means when donors are mixed. The value
type decides whether the mixed result is allowed to be a new number at all. Both
are applied in Step 11.

### Why There Is a Transformation

Blending happens *on the transformed scale*, so this step is really the choice of
what "average" means when five patients are mixed together.

Concentrations span orders of magnitude. Averaging 0.5, 0.8, and 12.0 on the raw
scale gives 4.4, a number dominated by the single high patient and
representative of none of them. Averaging on the log scale gives about 1.6, the
geometric mean, which is the pharmacometric convention for precisely this
reason. Blend concentrations on the raw scale instead and one high-exposure
donor drags every avatar it touches upward.

Logs have one problem with PK data: exact zeros. Pre-dose rows and imputed
below-the-limit-of-quantification (BLQ) values are genuinely 0, and
$\log 0=-\infty$ would poison the entire blend. So a small constant $c_e$ is
added before taking logs. Choosing $c_e$ to be **half the smallest positive
value in that endpoint** puts zeros just below the smallest real measurement,
near the bottom of the observed range where a zero belongs, and scales it to the
data rather than fixing an arbitrary $10^{-6}$.

Not every endpoint wants logs. A PD score, a change from baseline, or a
temperature difference can legitimately be negative, where a log is undefined.
The positive-like test below is what separates the two cases, and its 1%
tolerance lets a concentration endpoint with a couple of small negative assay
readings still count as positive-like.

### The Transformation Rule

Transformations are selected separately for each DVID endpoint, or once for the
implicit endpoint `"DV"` when DVID is absent. After discarding nonfinite values,
an endpoint is considered positive-like when it has at least one positive value,
its median is positive, and at most 1% of values are negative.

For a positive-like endpoint, define

$$
c_e=\max\left\{\frac{1}{2}\min(Y_e[Y_e>0]),\sqrt{\epsilon_{\mathrm{mach}}}\right\}
$$

and use the offset-log pair

$$
g_e(y)=\log\{\max(y,0)+c_e\},
\qquad
g_e^{-1}(z)=\max\{\exp(z)-c_e,0\}.
$$

Other endpoints use $g_e(y)=y$. The truncations mean a positive-like endpoint
cannot generate a negative DV, whatever the noise of Step 11 does, a guarantee
that comes free with the back-transformation rather than needing a separate
clamp. Transformation choices and offsets are stored in
`attr(synthetic, "pmx_settings")$endpoint_transforms`, so the scale an endpoint
was blended on is always recoverable from the output.

### The Endpoint Value Type

A weighted mean of five donors' zeros and ones is a number between them, and the
subject and residual noise then carry it outside the source range entirely. A
binary endpoint comes back as hundreds of distinct values, an ordinal 1/2/3
scale comes back with values above 3, and an integer count comes back as
fractions. Nothing downstream catches it, because class restoration works on the
DV *column* and a PMX table holds every endpoint in that one numeric column, so
there is nothing per-endpoint left to restore.

Each endpoint is therefore classified as one of four **value types**:

- **`continuous`** when any recorded value is not a whole number. Nothing is
  constrained.
- **`binary`** when every value is a whole number and there are two distinct
  levels.
- **`ordinal`** when every value is a whole number and there are at most twelve
  distinct levels.
- **`integer`** when every value is a whole number and there are more than
  twelve.

**Inferring this from the data is safe in a way that inferring the dose basis is
not**, and the difference is worth being explicit about, because the two look
like the same kind of guess. Dose-basis detection asks whether a ratio
*clusters*, which a study can fail to do for many ordinary reasons, and a wrong
answer rewrites every dose amount. This one asks only whether every recorded
value of an endpoint is a whole number, which the data answers outright, and the
response to a false positive is to reproduce granularity the source already had.
It still fails closed on too little evidence.

`pmx_roles(endpoint_types = )` overrides the inference in either direction, and
accepts `"count"` as a synonym for `"integer"`. `pmx_endpoint_types()` reports
what was decided for each endpoint and why, and the same decision is recorded as
`endpoint_value_types` on the output.

**What this costs, stated where it is met.** Snapping a discrete endpoint back
onto its levels (Step 11) means generated values on that endpoint **are** source
values. A 0/1 endpoint has no third value available to emit. So the blend, which
is what protects the measurements everywhere else, protects nothing here: what
protects a patient on a discrete endpoint is the visit-set and dose-schedule
machinery of M1 through M4, never the distinctness of the number.

## Step 4: Build One Numeric Profile per Source Subject

A profile concatenates baseline and longitudinal information:

$$
\mathbf{x}_i=
\left[
  \mathbf{c}_i,\
  \mathbf{y}_{i1}(G_1),\ldots,\
  \mathbf{y}_{iE}(G_E)
\right].
$$

Here $\mathbf{c}_i$ contains the first nonmissing value of each declared
baseline covariate. Numeric covariates contribute one feature. Factors and
character covariates contribute one indicator for every factor level or observed
character value. The vector $\mathbf{y}_{ie}(G_e)$ is subject $i$'s transformed
endpoint-$e$ trajectory interpolated on its common grid.

This profile is used only for distance calculations. It does not replace the
source data and is never returned as a synthetic patient.

### The Common Time Grid

Two subjects sampled at different times cannot be compared as vectors until they
are put on shared coordinates. That is what the common grid is for, and it is
worth saying up front what it is *not* for, because the 15-point cap below
invites a misreading.

> **The grid is a ruler, not a resampler.** It is used only to build the profile
> that measures which subjects resemble each other. It never limits the
> resolution of the generated data. A synthetic subject carries every one of its
> anchor's observation times, however many that is, and the donor values placed
> there are interpolated from each donor's **full** observed trajectory, not
> from its 15-point summary. A richly sampled source gives a richly sampled
> avatar.

For endpoint $e$, the package pools finite aligned observation times and makes a
common grid $G_e$. If there are at most 15 unique times, all are retained. If
there are more, 15 type-8 empirical quantiles are used and duplicate quantiles
are removed.

Each subject's transformed trajectory is linearly interpolated onto $G_e$. There
is no extrapolation during profile construction, so grid locations outside a
subject's observed window become missing profile features. Tied times are
averaged by `stats::approx(..., ties = mean)`.

The cap binds on ordinary PK data, where a rich sampling schedule is compressed
to a 15-point summary before the PCA of this step compresses it further. Two
things limit the damage. The grid points are *quantiles of the observed times*,
so they land where sampling is dense, early and around absorption and peak,
rather than spreading evenly across a long tail. And the grid is built per
endpoint, so PK and PD each get their own 15 rather than sharing them. The
residual risk is worst for multiple-dose data with several peaks and troughs,
where 15 points across a long window can blur the shape that distinguishes
subjects. When that happens the cost is a *worse choice of donors*, never a
coarser output trajectory. `max_points` is fixed at 15 and is not an argument.

During final synthesis, donors are first interpolated directly to the anchor's
aligned observation times. Suppose target time $t$ lies outside donor $r$'s
window. Rather than extrapolating indefinitely, the method maps the target's
relative position in the whole anchor observation window to the donor window:

$$
q(t)=\frac{t-a_{\mathrm{target}}}
           {b_{\mathrm{target}}-a_{\mathrm{target}}},
\qquad
u_r(t)=a_r+q(t)(b_r-a_r),
$$

where $[a_{\mathrm{target}},b_{\mathrm{target}}]$ and $[a_r,b_r]$ are the target
and donor ranges. The donor trajectory is evaluated at $u_r(t)$. This
normalized-window fallback preserves progress through a trajectory, but it is
not a mechanistic warping model. A donor with only one unique time supplies its
mean at every requested target time.

### Multiple Endpoints

`synpmx_avatar()` handles any number of endpoints, a PK concentration and one or
more PD measures, declared through `dvid`. Each is processed *separately* first
and only then merged, in this order:

1. **Transform, per endpoint.** Step 3 runs once per endpoint, so PK may be
   blended on the log scale while a PD score stays on the identity scale.
2. **Interpolate, per endpoint.** Each endpoint gets its *own* common grid, built
   from the pooled observation times of that endpoint alone. A sparsely sampled
   PD is not padded out to match a rich PK.
3. **Concatenate.** Covariate features, then each endpoint's grid columns, are
   laid side by side into the single row $\mathbf{x}_i$ above.
4. **Impute, drop, standardize.** Median imputation, removal of constant
   columns, then column-by-column standardization.
5. **One PCA over everything**, giving a single $\boldsymbol{\xi}_i$ that mixes
   all endpoints, not one PCA per endpoint.

Consequently there is **one distance, one donor set, and one set of weights**,
shared by every endpoint. A donor is chosen for overall similarity, and a donor
that matches PK closely but PD poorly still donates its PD.

Standardization is what makes endpoints commensurable at all: a log
concentration and a raw neutrophil count cannot be compared until each column is
divided by its across-subject standard deviation.

But note carefully what standardization does *not* do. It equalizes **units**,
not **endpoints**. Every grid point becomes one unit-variance column, so an
endpoint sampled at 15 times contributes 15 columns while one sampled at 3 times
contributes 3, and a Euclidean distance summing over columns lets the first
outvote the second. Donor selection is therefore driven mostly by whichever
endpoint was sampled most densely, and baseline covariates are a small minority
of the distance whenever trajectories are rich. **This weighting is an accident
of the sampling design rather than a modeling choice**, and there is currently
no way to reweight endpoints. If donors must match on a sparse endpoint, that is
a limitation to know about before trusting the selection.

Two related behaviors follow. A subject missing an endpoint entirely is
median-imputed on those columns, so it looks *average* there and sits nearer the
middle of the space than it deserves to. And at generation, if a chosen donor
has no value for the endpoint being filled, the weights renormalize locally over
the donors that do, so a missing endpoint thins the blend rather than punching a
hole in it.

### Missing Features and Scaling

An entirely missing feature is removed. Otherwise, a missing profile entry is
replaced with its across-subject feature median $m_l$:

$$
x^{\mathrm{imp}}_{il}=
\begin{cases}
x_{il}, & x_{il}\text{ is finite},\\
m_l, & \text{otherwise}.
\end{cases}
$$

This imputation exists only inside neighbor finding. It does not fill source or
output DV rows. Features whose standard deviation is not greater than
$\sqrt{\epsilon_{\mathrm{mach}}}$ are removed. The remaining columns are
standardized as

$$
z_{il}=\frac{x^{\mathrm{imp}}_{il}-\bar{x}^{\mathrm{imp}}_l}{s_l}.
$$

### Rank-Safe PCA

When there are at least two subjects and two usable features, principal
component analysis (PCA) is applied to the standardized matrix with maximum rank

$$
r_{\max}=\min(n-1,p).
$$

Let $\lambda_h$ be the variance of principal component $h$. The retained
dimension $H$ is the smallest value satisfying

$$
\frac{\sum_{h=1}^{H}\lambda_h}
     {\sum_{h=1}^{r_{\max}}\lambda_h}
\ge v,
$$

where $v$ is `pca_variance` (default 0.90).

The **subject coordinate** is the output of this whole step and the input to
every distance in Step 8, so it gets its own symbol:

$$
\boldsymbol{\xi}_i=(\xi_{i1},\ldots,\xi_{iH}),
\qquad
\xi_{ih}=\text{subject } i\text{'s score on principal component } h .
$$

Read $\boldsymbol{\xi}$ as the Greek letter *xi*. It is one point per subject in
an $H$-dimensional space, and $H$ is typically a handful, the number of
components needed to reach $v$. Everything upstream exists to produce it: the
raw profile $\mathbf{x}_i$ is imputed to $x^{\mathrm{imp}}_{il}$, standardized to
$z_{il}$ so that a weight in kilograms and a log-concentration compete on equal
footing rather than by the size of their units, and finally rotated by PCA into
$\boldsymbol{\xi}_i$.

If PCA is not available but standardized features exist, those features are used
directly as $\boldsymbol{\xi}_i$. If none exist, all subjects receive the same
one-dimensional zero coordinate.

The term **rank-safe** means the attempted PCA rank cannot exceed either the
feature rank or $n-1$. It does not imply that sparse or nearly duplicated
profiles contain useful separation.

## Step 5: Decide Who May Donate to Whom

Similarity in PCA space is not sufficient on its own. A subject receiving a
single oral dose should not donate a trajectory to an infusion or repeat-dose
template. So before any donor is chosen, each source subject is read for the two
things that decide whether it may donate to a given anchor, and Step 8 uses them
to pick each avatar's actual donors. The two are not the same kind of
restriction, and the generator treats them differently.

**Route of administration is an absolute barrier.** A bolus, an infusion, and an
oral dose produce categorically different concentration shapes. Blending across
them does not yield a noisier version of the truth; it yields a trajectory no
protocol could have produced. Donors are therefore *never* drawn from outside
the anchor's route, at any stage, for any shortfall, including when that leaves
the anchor with no legal donor at all.

Each subject's route is read from its dosing rows only, as the set of

$$
\bigl(\mathrm{EVID},\ \mathrm{CMT},\ \mathbb{1}[\mathrm{RATE}\neq 0]\bigr)
$$

triples appearing on them, so that the compartment a dose enters and whether it
is delivered over time both bind, while dose *count* does not: three oral doses
and five oral doses share a route. NONMEM's `RATE < 0`, a modeled rate or
duration, counts as an infusion.

**Everything else is a preference.** Each subject also receives a **dosing
pattern**, a summary of the course it was given and what was measured on it,
holding:

- the ordered EVID values on event rows;
- optional CMT and DVID values on those rows;
- the sign and rounded numeric magnitude of optional AMT and RATE;
- the number of positive-dose starts, or all events if no positive AMT exists;
- successive start-time gaps rounded to two significant digits; and
- the set of endpoints on eligible observation rows.

Two subjects with the same dosing pattern were, as far as the data records, put
through the same course. Step 8 prefers such a donor, and where too few exist it
takes the nearest subject on the same route whatever its pattern.

Notice what is not in the pattern: the observation-time schedule. Event values
still come from the anchor template, while unequal observation times are handled
by interpolation. Dose magnitude *is* in it, so same-dose subjects are preferred,
but unlike route it is not a barrier.

**Assigned strata do not enter this at all.** A treatment arm groups visit
patterns, dose-basis detection and anchor sampling, and it never restricts who
may donate: donors are borrowed across arms freely, because route is the only
thing that makes a blend physiologically impossible. An avatar therefore stays in
its anchor's arm while its measurements may come from patients in another.

## Step 6: Restrict the Anchor Pool and Sample Anchors

### M2. Which Patients May Be Used as an Anchor

Two exclusions, both acting on the pool of source subjects an avatar may be
built from. A subject whose follow-up length or dose count exceeds twice the
cohort's 90th percentile is never used as an anchor (`screen`), so no avatar
inherits a conspicuous skeleton. And a subject in a route arm holding fewer than
`k + 1` subjects has no legal donor set, so by default it is dropped from the
pool too (`on_donor_shortfall`).

A third exclusion enforces the guarantee stated at the top of this page. A
patient whose set of observation times, or set of dose times, is shared by
nobody and cannot be exchanged for a safe one is not built upon, since any
avatar anchored there would carry that one person's schedule. This one is
decided **per avatar rather than per patient**: the avatar is re-anchored
somewhere else, and the patient stays in the pool for every other avatar. An
earlier version excluded such patients outright and removed almost half a cohort
to fix a single avatar, which is why.

**Somewhere else in the same stratum.** An anchor carries its own `strata`
values into the output, so moving an avatar to another arm would put a patient in
an arm they were never in and leave the run reporting a balance it did not
build. Where an arm holds nobody who can be masked, individualised dosing in one
cohort of a trial for instance, its avatars keep the anchor they have, the run
raises an alert naming that arm, and the scorecard's B1a and B1b rows fail
against it. `unmaskable_strata()` answers the same question from the source
before anything is generated.

None of the three shrinks the cohort: the remaining anchors are sampled with
replacement to fill every slot, and excluded subjects still contribute
measurements as donors. What is lost is coverage of that structure rather than
sample size, and `pmx_masking_report()` says how much, as the share of the
source's distinct dose regimens that survive into the output. Where only one
patient stopped after a single dose, "stopped after one dose" cannot be
represented at all without pointing at that person.

The exclusion catches subjects who are **extreme**. The post-generation screen
is a different thing, and not a masking step.

### Sampling Anchors, and Keeping Each Stratum's Share

Anchors are sampled with replacement from the surviving pool. An avatar never
leaves its anchor's stratum, but sampling with replacement leaves the *balance*
to chance, so a balanced design can come back materially unbalanced on one seed
and differently unbalanced on the next, and any downstream summary grouped by
arm inherits that.

At the default `preserve_strata_balance = TRUE`, each declared stratum is
therefore given the share of the synthetic cohort it holds in the source.
Targets are proportions rather than counts, so they still hold when `n_subjects`
differs from the source size. **Strata holding fewer than three source patients
are deliberately not balanced**, because reproducing such a stratum's size
exactly would say more about those few patients than the rest of the cohort
does; those subjects are pooled and sampled as before. Setting the argument
`FALSE` returns to pure sampling with replacement.

## Step 7: Copy the Anchor's Rows, and Redraw the Schedule

Let subject $i$ have source rows $D_i$. For every requested synthetic subject,
the generator takes its sampled **anchor** $a$ and begins with

$$
S^{(0)} = D_a.
$$

Thus $S^{(0)}$ contains the anchor's full row count and row-wise combination of
TIME, EVID, AMT, RATE, CMT, DVID, and any other source columns. This is the
event skeleton. By copying these rows together, the method avoids impossible
records that can result from generating event fields independently, such as an
infusion rate on an oral dose.

Only a few parts of this skeleton are subsequently changed:

- which visits carry an observation is redrawn by M3, and where dosing stops by
  M4, both below;
- TIME may be jittered if `time_jitter > 0`;
- declared baseline covariates are replaced for all rows of the synthetic
  subject (Step 10);
- eligible, originally present observation DVs are synthesized (Step 11);
- dose amounts may be recomputed (Step 12);
- standard MDV may be re-derived, ID is replaced, and rows are sorted by TIME
  retaining original order within ties (Step 13).

Event-row DV values and undeclared columns otherwise remain anchor-template
values. Consequently, users should not place a semantically important column in
the input and assume the package understands it merely because it is present.

### M3. Redrawing Which Visits the Avatar Attended

Once coarsening has put every subject on a shared visit grid, what remains of a
schedule is which of those visits each subject attended. Every time is then
shared, so no single visit is identifying. The *combination of absences* is, and
a patient who missed weeks 2 and 3 is singled out by a fingerprint made of gaps.
No grid fixes this at any resolution, because the grid decides where the visits
are and not which ones a subject has. So the pattern is drawn from ones at least
`min_pattern_share` subjects hold.

The default of 2 is the smallest value that means anything, and what it means is
exact: **no synthetic patient carries a schedule unique to a real patient.** An
attacker who links a reproduced pattern to a participant gets at least two
candidates, never one. Three is more conservative but no more defensible as a
sentence, and it costs sharply more.

**This concerns DV observations only. Dose events are never redrawn**, since an
avatar's doses, their times and their amounts come verbatim from its anchor,
because a drawn regimen could be one no protocol permits.

After M1, every patient sits on the same grid of visit times. What still differs
between patients is which of those visits each one actually turned up for:

```
grid            wk1  wk2  wk3  wk4  wk5  wk6
patient A        x    x    x    x    x    x     attended everything
patient B        x    x    x    x                dropped out after week 4
patient C        x    x         x    x    x      missed week 3
```

Patient B's and patient C's rows are *visit sets*. If only patient C ever missed
week 3, then "attended everything except week 3" points at patient C as surely
as an unusual sampling time would.

So M3 redraws that row. The avatar keeps its anchor's doses and the grid's visit
times, and only the **pattern of x's** is replaced by one drawn from patients who
share it, or built fresh with the same *shape*: how many misses, and whether they
run to the end, sit in a block, or scatter. What survives is how much
missingness the study had and of what kind. What is lost is which specific weeks
any one patient missed.

Three details of how the shape is placed are worth stating, because each was a
defect first and each is now a regression test:

- **A shape is placed by enumerating the legal arrangements, not by resampling.**
  Misses that run to the end have exactly *one* arrangement for a given count, so
  redrawing at random proposes the same rejected arrangement over and over. Where
  nothing at the wanted count is free, the count moves outward by a visit or two.
- **When no (kind, count) shape is shared either**, the kind alone is used.
  "These patients have visits missing at the end" is plainly shared even when
  every stopping depth has a single holder, which is what a study with staggered
  follow-up looks like.
- **Misses are only ever placed on endpoints that actually go missing.** An
  endpoint every patient has at every visit is complete by construction, and
  putting a miss on one of its cells invents missingness the study does not have
  rather than masking any.

### M4. Redrawing Where Dosing Stopped

M3's counterpart on the dose side, and the reason it is separate is the sentence
above: dose events are never redrawn, because a drawn regimen could be one no
protocol permits. There is exactly one edit that escapes that objection.
Truncating a schedule at one of its own dose times yields a regimen the study did
give someone, a patient who stopped early. Moving a dose, or inventing one, does
not.

So where a patient's schedule is one nobody else has, the avatar stops earlier,
at the deepest point that satisfies one of two conditions. Either **several
patients stopped exactly there**, so the avatar wears a regimen those patients
share; or **nobody stopped there and several patients passed through it**, so the
stopping point identifies nobody while the doses leading up to it are ones many
patients received. The walk is **downward only**, since truncation removes dose
rows and cannot add one, so a deeper target would silently leave the schedule
exactly where it was.

The second condition's two halves are both necessary, and the second half is the
one that is easy to lose. Judging the doses one at a time, so that every dose
time in the kept set is a time several patients received, is not enough, because
the *combination* can still be one patient's. Getting that wrong leaves avatars
truncated to an opening only one real patient holds, one real patient's dosing
minus its last dose, and no check catches it: every copy-detection row compares
complete schedules, so a set nobody holds complete is precisely the set that
passes. Openings are counted whole.

This does not rescue every cohort. Where the only patients who stopped early are
the sole holders of their stopping depths, neither depth is free, and that
regimen cannot be represented at all. Such a cohort falls back to M2, so those
patients are not built upon, and the masking report shows the regimen coverage it
cost.

Two shapes are fully served rather than one. A study with many patients stopping
at many depths, the ordinary oncology shape, has free depths throughout. So does
a study where **no complete schedule repeats but the openings do**, which is
routine clinical care, where every patient begins on the same interval and
diverges as their own clinician decides. `pheno_sd` is that second shape:
almost every infant has a dose schedule nobody shares, and truncating each course
back to its shared opening still keeps a substantial part of the dosing.

## Step 8: Choose the Donors

### The Distance

For anchor $a$ and candidate donor $r$, the distance is the ordinary
straight-line (Euclidean) distance between their subject coordinates from
Step 4:

$$
d_{ar}=\left\|\boldsymbol{\xi}_a-
                   \boldsymbol{\xi}_r\right\|_2
=\sqrt{\sum_{h=1}^{H}(\xi_{ah}-\xi_{rh})^2},
$$

where

- $h$ indexes the **retained principal components**, running from 1 to $H$, not
  subjects and not time points;
- $\xi_{ah}$ is the **anchor's** score on component $h$, and $\xi_{rh}$ is
  **candidate $r$'s** score on that same component, both defined in Step 4;
- each squared term is one component's disagreement between the two subjects,
  and the sum over all $H$ components, square-rooted, is how far apart they sit.

It is Pythagoras in $H$ dimensions, and all the work went into choosing the
coordinates. Because the features were standardized before PCA, no single
original measurement dominates the distance merely by being recorded in larger
units.

This is the only distance in the selection rule. There is no separate structural
metric trading dose differences against schedule differences: structure enters as
the two-stage ordering below rather than as a score.

### The Selection Algorithm

Write $\mathcal{R}_a=\{r\neq a: \mathrm{route}_r=\mathrm{route}_a\}$ for the
subjects sharing the anchor's route and $\mathcal{S}_a=\{r\in\mathcal{R}_a:
\mathrm{pat}_r=\mathrm{pat}_a\}$ for those of them sharing its dosing pattern
too, both from Step 5. To choose $k$ donors for anchor $a$:

1. **Same-pattern stage.** Sort $\mathcal{S}_a$ by increasing $d_{ar}$, breaking
   ties by subject index, and take the first $k$. If $|\mathcal{S}_a|\ge k$,
   stop.
2. **Fallback stage.** Sort $\mathcal{R}_a\setminus\mathcal{S}_a$ the same way
   and take enough of it to bring the total to $k$, whatever the dose or
   schedule. Record a warning that donors were borrowed across dose groups.
3. **Shortfall.** If $|\mathcal{R}_a|<k$ the floor is unreachable, because the
   only remaining candidates are on another route and route is never crossed.

Ties break by subject index, so selection is deterministic under a fixed seed.
Fallback donors' measurements are mapped onto the anchor's own observation times
by interpolation, so the avatar keeps its anchor's regimen while its values are
averaged across at least $k$ real patients. This borrowing matters most for
datasets with individualized dosing, where an exact dose magnitude can leave
nearly every subject holding a dosing pattern of its own and the fallback stage
does essentially all the work.

### What the Fallback Stage Costs

Be clear about what the fallback distance does and does not know. `AMT` is not a
profile feature, so the fallback ranking never compares doses directly. It
compares baseline covariates and the shape and level of the DV trajectory. Dose
does influence it, but only *indirectly*: a subject given a much larger dose has
higher concentrations, and those concentrations are profile features, so it lands
further away. The ranking therefore tends to prefer similar-dose donors without
being told to.

Tendency is not a guarantee, and nothing rescales. A borrowed donor's values are
interpolated onto the anchor's times and blended **raw**, by deliberate design,
choosing structural realism over statistical fidelity. The consequence is the
one stated at the top of this page: a synthetic subject carries its anchor's
`AMT` while its concentrations may be blended from donors given a different
dose, so **the dose-exposure relationship in AVATAR output is not guaranteed to
be preserved**. Fitting a structural model to AVATAR output can recover a
dose-exposure relationship that is an artifact of blending. Use
`synpmx_calibrated()` when the parameters themselves have to mean something.

### When the Donor Floor Cannot Be Reached

Because route is absolute, a route arm holding fewer than $k+1$ subjects can
never supply a legal donor set. There is no good answer in that situation, only a
choice between two bad ones, and which is worse depends on whether the sparse arm
matters more than the disclosure risk of reproducing it. That is the caller's
judgement, so `on_donor_shortfall` makes it explicit:

| `on_donor_shortfall` | Behavior |
|---|---|
| `"drop"` *(default)* | Omit those subjects from the anchor pool. No avatar is built on them, and the synthetic cohort does not represent that arm. |
| `"noise"` | Keep them, blending however many same-route donors exist, possibly none, and relying on `subject_noise_sd` and `residual_noise_sd` for the rest. |
| `"error"` | Refuse to generate, naming the arm and both alternatives. |

Every branch alerts loudly, because each one silently changes something a reader
would want to know: what the cohort covers, or how identifying it is.

`"noise"` is **available but not recommended**, and the reason is the defect this
whole mechanism exists to prevent. A subject blended from one donor, or from
none, is a noised near-copy of a real patient. If you use it because the sparse
arm genuinely matters more, screen the result and treat those subjects as
individually identifying.

One case forces the issue: when *every* arm is below the floor, `"drop"` would
leave nothing to generate, so generation proceeds as if `"noise"` and says so.

## Step 9: Randomize and Cap the Donor Weights

Suppose $K$ donors have been selected. Let $E_r\sim\mathrm{Exponential}(1)$ and
let $R_r$ be a random permutation of $1,\ldots,K$. Importantly, $R_r$ is a
randomized rank rather than the distance order, which is what stops the nearest
neighbour from being systematically the largest contributor. The raw donor weight
is

$$
q_r=\frac{E_r}{\max(d_{ar},\varepsilon)}2^{-R_r},
$$

where

$$
\varepsilon=
\begin{cases}
\max\{10^{-8},10^{-6}\operatorname{median}(d_{ar}:d_{ar}>0)\},
  & \text{if a positive distance exists},\\
10^{-8}, & \text{otherwise}.
\end{cases}
$$

The first normalization is $w_r=q_r/\sum_s q_s$. If the raw weights are
nonfinite or have a nonpositive total, the implementation replaces them with
equal raw values before normalizing.

### Capping Every Donor

Each donor's weight is its **share of the avatar**: five weights that sum to 1.
The floor $k$ says how many real patients go into the blend, but it says nothing
about how much of any one of them comes out. Five donors weighted $(0.95, 0.02,
0.02, 0.005, 0.005)$ satisfy a floor of five while being, to any practical
purpose, a copy of one person. So the cap, rather than $k$, is what controls how
closely an avatar can resemble a single real patient.

The raw formula above produces very uneven weights on purpose, which is the rank
attenuation doing its job. `max_donor_weight` (written $c$ below, default $0.50$)
is the ceiling on any one share.

Enforcing it is less obvious than it sounds, because **weights have to keep
summing to 1**. Cutting the leader down does not remove that weight from the
blend; it has to be given to the others, in proportion to what they already hold:

$$
w_r \leftarrow \Bigl(1-\textstyle\sum_{s\in P}w_s\Bigr)
              \frac{w_r}{\sum_{s\notin P}w_s},
\qquad r\notin P,
$$

where $P$ is the set of donors already pinned at $c$. And that redistribution can
itself push a donor over the ceiling. Starting from $(0.50, 0.25, 0.15, 0.06,
0.04)$ with a tight $c = 0.30$: cutting the leader to $0.30$ frees $0.20$, and
the second donor's share of it takes them from $0.25$ to $0.35$, over the
ceiling. Capping only the largest weight and stopping, which is what the
implementation did throughout the $0.80$ era, leaves exactly that violation in
place: the stated maximum, broken by the donor the redistribution created.

So the rule repeats, pinning everything over $c$, redistributing, and looking
again, which is the standard construction called **water-filling**. Pinned donors
are never topped up, so each pass pins at least one more and it terminates. The
tight example settles at $(0.30, 0.30, 0.24, 0.096, 0.064)$.

### One Pass Suffices at the Default Cap

A second pass is needed only when redistribution lifts the runner-up over the
ceiling, that is when

$$
(1-c)\,\frac{w_2}{1-w_1}>c .
$$

At $c=0.5$ this requires $w_2>1-w_1$, i.e. $w_1+w_2>1$, which no pair of weights
can satisfy. **At any cap of $0.5$ or above the loop runs at most once**, and
simulation agrees exactly: at $c=0.5$ the cap is untouched 33% of the time and
pinned once 67%, never twice. Only tighter caps exercise the iteration; at
$c=0.3$ two passes are needed 53% of the time and three passes 11%.

The general routine is kept because `max_donor_weight` is an argument and a user
may set it below $0.5$, where a single pass is genuinely wrong. But at the
shipped default, the simple thing and the correct thing coincide.

### The Infeasible-Cap Boundary

With $K$ donors the weights average $1/K$, so **no weight vector can satisfy a
ceiling below $1/K$**: asking for $c=0.3$ from three donors is asking three
numbers below $0.3$ to sum to 1. Such a cap relaxes to exactly $1/K$, uniform
weights and the flattest blend available, rather than an error. Two donors under
$c=0.5$ give $(0.5,0.5)$, which meets the cap exactly. With one donor the weight
is necessarily 1, since a cap cannot invent a second donor.

### What an Uncapped Blend Looks Like

It is reasonable to ask whether the cap earns its place, given that $k$ already
forces five donors into every avatar. The answer is that **without a cap, the
floor is largely decorative.** Simulating the raw weight formula at $k=5$ with
representative donor distances:

| uncapped largest donor share | value |
|---|---|
| median | 0.58 |
| interquartile range | 0.47 - 0.72 |
| $P(\text{largest} > 0.5)$ | 0.67 |
| $P(\text{largest} > 0.8)$ | 0.14 |
| effective donors $1/\sum_r w_r^2$ | 2.37 |

So the unconstrained formula typically puts **58% of an avatar into a single
donor**, and one avatar in seven takes more than 80% from one person. Nominally
five patients are blended; effectively about two and a half are. The cap is the
only thing that closes that gap.

### Choosing the Cap

The useful diagnostic is **how often the cap fires**, because that says what role
it is playing:

| `max_donor_weight` | binds on | effective donors |
|---|---|---|
| 0.30 | 99% | 3.92 |
| 0.40 | 89% | 3.28 |
| 0.50 *(default)* | 68% | 2.90 |
| 0.60 | 47% | 2.64 |
| 0.80 | 15% | 2.40 |
| none | 0% | 2.37 |

Both ends of that table are the wrong kind of parameter. A cap that fires on 99%
of subjects is not a guardrail, it *is* the weighting scheme, and the
inverse-distance term underneath it barely matters. A cap at $0.80$ fires on 15%,
trimming only the worst tail while still permitting one patient to be four-fifths
of an avatar; that was the old default, and it was never really protecting
anything.

The default is **$0.50$**, which states as a single checkable sentence: *no
single real patient is more than half of any synthetic patient.* It fires on
about two thirds of subjects, so it genuinely constrains without wholly replacing
the distance weighting.

Its honest cost is that "five donors are blended" is really "about three
effective donors". Tightening to $0.30$ buys about one more effective donor at
the price of a cap that fires essentially always. Both are defensible; $0.50$ is
the one that can be explained in a sentence.

Because these numbers depend on the cohort, `pmx_settings` records
`cap_binding_fraction` and `mean_effective_donors` for every run, so the same
question can be asked of real data rather than of a simulation.

### What the Cap Costs

Treating donors as independent, a weighted blend retains $\sum_r w_r^2$ of
individual variance, so the synthetic cohort's between-subject standard deviation
shrinks by $\sqrt{\sum_r w_r^2}$. The reciprocal $1/\sum_r w_r^2$, the effective
number of donors, reported as `mean_effective_donors` in `pmx_settings`, reads as
a privacy floor and as the variability cost in one number. It equals $K$ for
uniform weights and $1$ for a sole donor.

That independence assumption is pessimistic in the direction that matters.
Donors are *nearest neighbours*, so they are strongly positively correlated, and
averaging them destroys much less variance than the formula implies; the
subject-level perturbation of Step 11 then restores more. Measured on a real
study, tightening the cap raises the effective donor count substantially while
between-subject variability stays flat within noise, where the independence
formula would have predicted a steep fall. This is why the cap can be chosen on
privacy grounds: the fidelity side of the trade is nearly flat.

The same selected subjects and weights are used for every declared covariate and
endpoint. When donor $r$ lacks a usable value at target location $j$, the
available weights are renormalized locally:

$$
w_{rj}^{*}=\frac{w_r I_{rj}}
                 {\sum_s w_s I_{sj}},
$$

where $I_{rj}=1$ when the value is finite and zero otherwise.

### M5. Blending as the Mechanism That Protects the Values

Spanning Steps 8 to 11. No avatar's measurements come from one real patient:
each is a distance-weighted blend of at least `k` = 5 compatible donors, with
`max_donor_weight` = 0.50 capping any single donor at half the blend. This is
the only mechanism whose strength grows with cohort size, which is why a source
with fewer than `k + 1` subjects raises a loud alert instead of quietly producing
near-copies. It is also the mechanism that does **not** apply to a discrete
endpoint, where the snap of Step 11 puts the value back onto a source level.

## Step 10: Synthesize the Baseline Covariates

For a numeric covariate, let $c_r$ be donor $r$'s first nonmissing value. When
all available values are positive and

$$
\frac{\max_r c_r}{\operatorname{median}_r(c_r)}>3,
$$

the blend is formed on the log scale; otherwise it is formed on the original
scale. With working-scale values $h(c_r)$, the donor center is

$$
\mu_c=\sum_r w_r^* h(c_r).
$$

Let $s_c$ be the ordinary sample standard deviation of the available
working-scale donor values. If it is unavailable or zero, the fallback is
$\max(0.05|\mu_c|,0.01)$. The generated working-scale covariate is

$$
c^*=\mu_c+\eta_c,
\qquad
\eta_c\sim\mathcal{N}(0,(\sigma_{\mathrm{subj}}s_c)^2),
$$

where `subject_noise_sd` is $\sigma_{\mathrm{subj}}$ (default 0.15). A log-scale
result is exponentiated, and any all-positive covariate is floored above zero.
The final value is repeated on every row of that synthetic subject.

For factor, character, and logical covariates, one available donor category is
sampled using the locally normalized weights. Categories are not averaged.

### A Rare Category Is Copied Rather Than Blended

That last sentence has a consequence worth measuring rather than asserting,
because it inverts the intuition that blending protects everything. It is the
covariate half of the same property the discrete endpoints of Step 3 have on the
observation side: where there is nothing to average, a synthetic patient's value
is some real patient's value.

A numeric covariate comes out of this step as a weighted mean plus noise, a new
number nobody had. A **categorical** covariate has nothing to average, so a
synthetic patient's category is always **some real patient's actual category,
copied**. Declared `strata` are copied from the anchor exactly, by design, since
they are protocol facts.

So the question is when a copied category reaches the output. Here is the
measurement, on a 40-patient fixture where one categorical level is held by a
varying number of otherwise-ordinary patients and everything else is held fixed:

```{r rare-level, eval=TRUE, message=FALSE}
rare_level_fixture <- function(n_holders) {
  set.seed(1)
  subject <- function(i) {
    times <- c(0, 1, 2, 4, 8, 24)
    data.frame(
      ID = i, TIME = c(0, times), NTIME = c(0, times),
      DV = c(NA, round(10 * exp(-0.1 * times) +
                         rnorm(length(times), 0, 0.3), 3)),
      AMT = c(100, rep(0, length(times))),
      EVID = c(1L, rep(0L, length(times))),
      CMT = c(1L, rep(2L, length(times))),
      WT = round(runif(1, 60, 80), 1),
      RARE = if (i <= n_holders) "rare-level" else "common",
      stringsAsFactors = FALSE
    )
  }
  do.call(rbind, lapply(1:40, subject))
}

propagation <- function(n_holders) {
  source <- rare_level_fixture(n_holders)
  roles <- pmx_roles(
    id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
    cmt = "CMT", nominal_time = "NTIME", covariates = c("WT", "RARE")
  )
  synthetic <- suppressWarnings(
    synpmx_avatar(source, roles, n_subjects = 200, seed = 9)
  )
  carrying <- length(unique(synthetic$ID[synthetic$RARE == "rare-level"]))
  data.frame(
    source_holders = paste0(n_holders, " of 40"),
    source_pct = round(100 * n_holders / 40, 1),
    avatars_carrying = carrying,
    avatar_pct = round(100 * carrying / 200, 1)
  )
}

knitr::kable(
  do.call(rbind, lapply(c(1, 2, 3, 5, 10, 20), propagation)),
  caption = paste(
    "A categorical level held by n real patients, and how many of 200",
    "avatars carry it."
  )
)
```

**A level held by one patient never reaches the output; a level held by two
does.** The mechanism is geometric rather than protective. Step 4 one-hot encodes
each categorical level as its own feature of the profile, so the sole holder of a
level sits alone on that axis, is nobody's nearest neighbour, and is therefore
almost never selected as a donor. Give the level a second holder and the two
become each other's nearest neighbour, the level propagates between them, and it
leaves. By ten holders the output tracks the source frequency closely.

The operative variable is the **number of holders**, not outlier status. Making
the two holders numeric outliers as well changes nothing on the categorical axis.
"Being an outlier is self-protecting" holds for numeric covariates, which are
blended, and does not hold for categories.

Two things follow, and the second is the one that matters.

**This is not a mechanism to rely on.** It is a side effect of nearest-neighbour
donor selection. Nothing enforces it, nothing reports it, and it degrades
smoothly: two holders is already a leak.

**The dangerous shape is a rare level on a *typical* patient.** A patient who is
unusual on their covariates is nobody's nearest neighbour and is self-protecting.
A patient who is ordinary in every way except one rare category, an ultra-rare
mutation status say, sits in the middle of the profile space, is selected as a
donor constantly, and their category is copied out. What to do about that is a
checking question rather than an algorithm one, and it is in
`vignette("scorecard-synthetic-data-checks")` under B5.

### A Covariate and an Endpoint Are Blended Independently

Covariates (this step) and endpoint trajectories (the next) are generated in
separate passes, each with its own donor weighting and its own perturbation. That
is fine when they are unrelated, but some datasets carry a **baseline covariate
that is the same quantity as a longitudinal endpoint**: a `B0` baseline B-cell
count beside a B-cell kinetic endpoint, a baseline biomarker beside its own time
course. AVATAR does not know they are linked, so a synthetic subject's baseline
covariate need not equal the baseline of its own generated trajectory. The two
are individually plausible but not mutually consistent.

This is a known limitation. It is usually harmless for workflow-development use,
where each column need only be realistic on its own. If your analysis relies on a
covariate agreeing with an endpoint's baseline, reconcile them after generation,
or do not declare the redundant covariate and read the baseline from the
trajectory instead.

## Step 11: Synthesize Each Endpoint Trajectory

Only positions that were eligible, nonmissing observations in the anchor template
receive synthesized values. Thus the anchor supplies the number and placement of
observations, including its missing-DV pattern, as redrawn by M3.

For endpoint $e$, donor trajectories are transformed with $g_e$ and interpolated
to the anchor observation times. Let $z_{rj}$ be donor $r$'s transformed value at
target position $j$. The deterministic blend is

$$
\bar z_j=\sum_r w_{rj}^{*}z_{rj}.
$$

If every donor is unavailable at a target, the transformed dataset median for
that endpoint is used and a warning is stored.

The endpoint noise scale $s_e$ is 1 on the offset-log scale. On the identity
scale it is the source endpoint standard deviation, with fallback
$\max(0.1|\operatorname{median}(Y_e)|,0.01)$ when necessary. A single
subject-level shift is shared across all positions of the endpoint:

$$
b_e\sim\mathcal{N}(0,(\sigma_{\mathrm{subj}}s_e)^2).
$$

Within-endpoint residual perturbations follow a stationary-scale first-order
autoregressive, AR(1), process in **observation order**:

$$
\begin{aligned}
\epsilon_{e1} &\sim
  \mathcal{N}(0,(\sigma_{\mathrm{res}}s_e)^2),\\
\epsilon_{ej} &= \phi\epsilon_{e,j-1}+\nu_{ej},\\
\nu_{ej} &\sim
  \mathcal{N}\left(0,
    (\sigma_{\mathrm{res}}s_e)^2(1-\phi^2)\right).
\end{aligned}
$$

Defaults are `residual_noise_sd = 0.05` and `residual_phi = 0.6`. The generated
DV is

$$
Y^*_{ej}=g_e^{-1}(\bar z_j+b_e+\epsilon_{ej}).
$$

Endpoints are processed separately, so PK and PD values cannot be blended with
one another. They nevertheless use the same donor subjects and initial subject
weights, preserving a limited form of cross-endpoint subject coherence. This is a
synthesis device, not a fitted multivariate residual model.

### Snapping a Discrete Endpoint Back onto Its Levels

$Y^*_{ej}$ above is a new number, which is what a continuous endpoint wants and
what a discrete one cannot have. So for an endpoint the value type of Step 3 made
`binary` or `ordinal`, the generated value is replaced by the **nearest observed
level**, deciding ties at the midpoints between adjacent levels. For an `integer`
endpoint it is rounded, and floored at zero where the source held no negative
value. A `continuous` endpoint is left alone.

The snap happens after the noise rather than before, so the subject shift and the
AR(1) sequence still decide *which* level a given observation lands on. What they
cannot do is invent a level between two the source recorded.

The count of generated values that could not be placed on a legal level is
recorded as `endpoint_type_violations`, and should be zero.

### Below-the-Limit-of-Quantification (BLOQ) Data and a Monolix-Style CENS Column

Declare `cens` and, if the source uses one, `limit` in `pmx_roles()`, and AVATAR
reconstructs censoring rather than copying it. The mechanism has three parts, and
the last of them happens here, alongside the snap above: both are corrections
applied to the generated value before it is written.

**The boundary is read from the source.** AVATAR makes no formal privacy claim,
so it may recover the assay limit from the data itself: under the Monolix
convention a censored row reports the boundary in `DV`, with `LIMIT` carrying the
other end of an interval when there is one. Left, right, and interval censoring
are each recognised. A differentially private engine may *not* do this, because
there the limit must be a declared public input: a boundary inferred from
confidential records is itself a release. That asymmetry is the reason the two
engines derive boundaries differently. If a study used more than one assay limit,
the most conservative boundary is taken rather than inventing a per-row rule the
source cannot support.

**Censored donor values are imputed before blending.** Every censored source row
reports the same substituted number, so blending those values directly would put
a floor under the synthetic data that the real study does not have. Each censored
row is instead replaced by a draw inside its censoring region, uniform below the
limit for a left-censored row and uniform within the interval for an
interval-censored one, before the trajectory is transformed and blended. A
uniform draw rather than a fixed LLOQ/2 avoids replacing one artificial spike
with another.

**`DV`, `CENS`, and `LIMIT` are reconstructed together.** The back-transformed
blend is treated as a *latent* value: what the subject would have measured with
no assay limit. The boundary is then applied to that latent, and all three
columns are written from the result, so they cannot disagree. A generated value
below the limit produces `CENS = 1` with `DV` exactly at the limit, reproducing
the spike of identical values that characterises real censored data.

Two honest limitations. The censored *fraction* is reproduced approximately
rather than exactly, because blending shrinks variance toward the middle while
the imputation draws uniformly across the censoring region, so the synthetic rate
typically runs somewhat above the source's. And if a `cens` role is declared but
no boundary can be read from it, rows flagged with `DV` missing for instance,
AVATAR warns and carries the flag through untouched rather than guessing.

## Step 12: Recompute the Dose from the Blended Covariate

### M6. Recomputing the Dose

Under dosing proportional to a baseline covariate, the multiplier is a protocol
property the stratum shares and the covariate is individual, and already blended.
Keeping the multiplier and recomputing the amount from the avatar's own covariate
stops the dose being a verbatim real value, which under mg/kg dosing discloses
the anchor's weight exactly, and repairs a coherence defect: previously every
avatar violated the mg/kg rule its own data claims to follow.

**Declared or inferred, and the difference matters.** Naming the covariate in
`pmx_roles(dose_covariate = )` states the relationship outright; leaving it
`NULL` makes the run infer one. Inference is deliberately conservative: it
requires the dose-to-covariate ratio to collapse onto a handful of levels within
2%, because rewriting amounts on a study that is *not* proportional would be
worse than leaving them alone. Two ordinary things defeat that test, dispensing
in vials and patients escalating by factors that are not identical across the
cohort, so a genuinely weight-based study can come through with its amounts
untouched. Declaring the covariate skips inference and holds **each dose row's
own ratio** rather than snapping to shared levels, so intra-patient escalation
survives exactly. `pmx_masking_report()` says which path ran and, when inference
declined, what it tried.

The difference on a study that dispenses in vials and escalates within patient,
which is the shape inference cannot recover:

```{r m6-declared}
# Three doses per patient at 1, 2 and 4 mg/kg -- but dispensed in 25 mg vials,
# so the recorded mg/kg is smeared and the ratios do not cluster.
m6_source <- do.call(rbind, lapply(1:12, function(i) {
  wt <- 52 + i * 2.3
  amt <- round(c(1, 2, 4) * wt / 25) * 25
  out <- rbind(
    data.frame(ID = i, TIME = c(0, 336, 1008), DV = NA_real_, AMT = amt,
               EVID = 1L, CMT = 1L, WT = wt),
    data.frame(ID = i, TIME = c(24, 360, 1032), DV = 5 + i / 10, AMT = 0,
               EVID = 0L, CMT = 2L, WT = wt))
  out[order(out$TIME, -out$EVID), ]
}))

m6_roles <- function(declare) {
  args <- list(id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
               cmt = "CMT", covariates = "WT")
  if (declare) args$dose_covariate <- "WT"
  do.call(pmx_roles, args)
}

# Each patient's course as a dose-per-kilogram profile. The question M6 exists
# to answer is whether a synthetic patient's profile is one a real patient
# could have been given -- not whether the numbers look plausible.
m6_profiles <- function(x) {
  dosed <- x[x$EVID != 0, ]
  vapply(split(round(dosed$AMT / dosed$WT, 2), dosed$ID),
         function(r) paste(r, collapse = " / "), character(1))
}

m6_coherent <- function(declare) {
  synthetic <- suppressWarnings(suppressMessages(
    synpmx_avatar(m6_source, m6_roles(declare), seed = 8)
  ))
  mean(m6_profiles(synthetic) %in% m6_profiles(m6_source))
}

data.frame(
  `dose_covariate` = c("NULL (inferred)", "\"WT\" (declared)"),
  `avatars on a real dosing profile` = paste0(
    round(100 * c(m6_coherent(FALSE), m6_coherent(TRUE))), "%"),
  check.names = FALSE
) |> knitr::kable(caption = paste(
  "Whether each avatar's course of mg/kg is one the protocol actually gave.",
  "Inference declines here, so the amounts are copied and each avatar ends up",
  "at its anchor's milligrams over its *own* blended weight -- a dose per",
  "kilogram nobody was prescribed."))
```

## Step 13: Restore the PMX Schema and Validate

After all synthetic subjects are row-bound, the original column order is
restored. Each result column is converted toward its source class:

- factors recover source levels and ordering; new factor IDs add new levels;
- integer columns are rounded and converted to integer;
- double, logical, character, Date, and POSIXct columns are restored; and
- a non-`data.frame` source class, such as a tibble class vector, is reapplied.

New numeric IDs start above the source maximum. New character and factor IDs use
labels such as `syn_001`, with a prefix added repeatedly if needed to avoid
collision.

If MDV is declared and the entire source obeys the standard relationship

$$
[\mathrm{MDV}=0]=[\mathrm{EVID}=0\ \text{and DV is present}],
$$

MDV is re-derived from that rule in the synthetic data and restored to its
original class. For any nonstandard source-wide MDV convention, anchor values are
left as copied. This step neither invents nor removes observation positions,
which ones an avatar has having been settled by M3 in Step 7.

Finally, `validate_pmx(..., strict = TRUE)` is run on the assembled result. A
generation call cannot silently return an output that fails the package's own
structural requirements.

## Reproducibility

All random operations occur inside a local seed scope. `seed` must be an integer
between zero and `.Machine$integer.max`. The caller's random-number generator
kind and `.Random.seed` are restored on exit, whether or not `.Random.seed`
existed before the call. Consequently, identical inputs and arguments produce an
identical complete result without consuming the surrounding analysis RNG stream.

`attr(synthetic, "pmx_settings")` records every argument, the explicit roles, and
what each mechanism did and cost. `pmx_masking_report()` prints it, and that is
the intended way to read it. It does not record the anchor ID, donor IDs,
distances, or realized weights for each synthetic subject.

## Checking a Run

Nothing above is self-certifying. Every mechanism costs something, in discarded
visit sets, excluded anchors, or timing fidelity, and what it buys **depends
entirely on the study**: the same defaults take one cohort to zero identifying
schedules and leave another at all of them, because one study recorded a shared
visit structure and the other did not. There is no headline claim to make, so
every run reports its own numbers instead.

`synpmx_scorecard()` is where those numbers are read. Every mechanism on this
page has a row there, along with the structural checks that a synthetic dataset
is legal PMX data at all.
`vignette("scorecard-synthetic-data-checks")` explains what each row asks and
what counts as passing, and `vignette("public-data-examples")` runs the whole
sequence across eight public datasets, including one where the guarantee cannot
hold and what is done about it.

## Edge Cases and Documented Fallbacks

The implementation favors an explicit warning and structurally valid output over
a silent incompatible blend.

```{r edge-cases-table, echo=FALSE}
edge_cases <- data.frame(
  Situation = c(
    "Fewer than k donors with the anchor's dosing pattern",
    "Route arm smaller than the k-donor floor",
    "Every route arm below the floor",
    "Cap below 1/K for K donors",
    "Source smaller than the k-donor floor",
    "All selected distances essentially zero",
    "One donor time only",
    "No donor value at a target",
    "No variable profile feature",
    "Identical donors with all noise disabled",
    "Nonstandard MDV convention"
  ),
  Behavior = c(
    "Borrow the nearest donors from other dose/schedule groups on the same route",
    "Follow on_donor_shortfall: drop (default), noise, or error; alert loudly",
    "Generate anyway and raise a loud alert; treat output as identifying",
    "Relax the cap to 1/K, i.e. uniform weights, rather than erroring",
    "Blend all available donors and raise a loud alert",
    "Use epsilon-stabilized randomized weights",
    "Repeat that donor's mean at every target time",
    "Use transformed endpoint median and record a warning",
    "Use a shared zero coordinate; distance cannot separate subjects",
    "A source-shaped trajectory may be mathematically unavoidable; warn",
    "Copy anchor MDV values rather than re-derive them"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(edge_cases, align = c("l", "l"))
```

Other boundaries are deliberate. `synpmx_avatar()` supports only
`event_method = "template"` and `dv_method = "avatar_blend"`. It does not fit a
structural model, reconstruct covariate-parameter relationships, simulate
occasions, enforce a dosing grammar beyond the sampled template, or provide a
censoring model. Fitting an explicit structural model is the job of
`synpmx_calibrated()`, which `vignette("synpmx-4-methods")` compares against this
one.

## Relationship to AVATAR

The method is **AVATAR-inspired**, not an implementation of published AVATAR
software. "AVATAR" is a method name rather than an initialism, from the
patient-centric *avatarization* literature in which each synthetic record is
built from the local neighborhood of real records. The original patient-centric
AVATAR method was reported by Guillaudeux and colleagues [1]. Destere and
colleagues describe a modified AVATAR benchmark for longitudinal population
pharmacokinetic (PopPK) data that first widened the data by subject and then used
PCA, K-nearest neighbors, inverse Euclidean distance, exponential stochastic
weights, and rank attenuation before reverse transformation [2]. The shared
family resemblance is the use of standardized subject features, PCA, local
neighbors, inverse-distance influence, exponential randomization, and rank
attenuation.

`synpmx` adapts those ideas to a PMX event table in several material ways:

- it preserves a sampled longitudinal event template rather than generating all
  fields from a wide vector;
- route of administration is an absolute barrier, while dose and schedule
  differences only make a donor less preferred;
- endpoints are transformed and interpolated separately, and a discrete endpoint
  is returned to its own levels;
- the same donors support covariates and endpoints;
- every multi-donor weight, not only the dominant one, is capped at
  `max_donor_weight` (default 0.50);
- subject and AR(1) perturbations are added on the endpoint working scale; and
- there is no inverse-PCA reconstruction of a full synthetic patient vector.

Destere et al. also demonstrate why method labels are not validation: their
benchmark found materially different PK parameter and residual-error biases
across synthesis algorithms [2]. That paper is a 2026 medRxiv preprint and was
not peer reviewed at the version supplied with this package repository. Its
privacy analyses do not transfer to `synpmx`; this package has undergone no
corresponding attack-based privacy validation.

## Algorithm Summary

What `synpmx_avatar()` executes, in order. These are implementation operations,
finer-grained than the thirteen steps above and following the same order.

1. Validate the source data, the roles, and the generator arguments.
2. Coarsen the recorded times onto a shared visit grid and pool the deviations.
3. Read the censoring boundary and impute censored source values.
4. Compute first-event-relative time and the per-endpoint transformation and
   value type.
5. Assemble baseline covariates and interpolated endpoint trajectories into one
   profile per source subject.
6. Median-impute profile features, remove unusable ones, standardize, and retain
   enough PCA components to meet `pca_variance`.
7. Read each subject's route of administration and dosing pattern.
8. Resolve anchors whose route arm cannot reach the donor floor per
   `on_donor_shortfall`, apply the structural screen, and sample the remaining
   anchors with replacement, within stratum where strata are declared.
9. For each anchor, copy its whole event template, then redraw its visit set and
   truncate its dose schedule where either is unique.
10. Find `k` donors on the anchor's route, taking those sharing its dosing
    pattern nearest-first and then the nearest remaining subjects on that route,
    and randomize their inverse-distance weights, water-filling every weight to
    `max_donor_weight`.
11. Generate subject-constant covariates from those donors.
12. For each endpoint, interpolate donor trajectories to the anchor times, blend,
    add a subject shift and AR(1) perturbations, back-transform, snap a discrete
    endpoint to its levels, and reapply the censoring boundary.
13. Recompute dose amounts from the blended covariate where dosing is
    covariate-proportional.
14. Re-derive standard MDV when applicable, assign a new ID, preserve tie order,
    restore the source schema, and validate the assembled result.
15. Attach `pmx_settings` and emit any collected fallback warnings.

## References

1. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).

2. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).
