---
title: "Demo: Using synpmx_calibrated"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx_calibrated}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_opendp <- requireNamespace("opendp", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
plots <- has_opendp && has_ggplot2
```

One study through the calibrated path end to end: decide whether the release is
worth making, spend the budget once, read what was released, and generate from
it for free.

The study is `mixroute_sim`, a simulated 90-subject dataset shipped with the
package. It is public, so nothing here is a real disclosure — but it is run
through the genuine OpenDP backend rather than the no-noise fixture, because
the point of this document is what a real release looks like.

**The numbers below change every time this page is built.** Privacy noise is
never user-seeded; a `seed` argument controls ordinary generation only.
Two runs of the same call give different releases, by design.

## The acknowledgment

`synpmx_calibrated()` refuses to run until this has been called once in the
session. It is a deliberate speed bump, not a technical safeguard.

```{r enable}
synpmx_enable_dp_engines()
```

## Configuration

```{r inputs, eval=has_opendp}
study <- as.data.frame(mixroute_sim)

study_roles <- pmx_roles(
  id = "ID", time = "TIME", nominal_time = "NTIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", cens = "CENS"
)
```

The model and the design are public inputs, and each records where it came
from. These are the same objects prior-only generation takes; what calibration
adds is a correction to the magnitude.

```{r public-inputs, eval=has_opendp}
public_model <- pmx_structural_model(
  pk = "1cmt_oral", typical = c(cl = 2, v = 10, ka = 0.5),
  source = "the documented generating truth of the mixroute_sim fixture"
)
public_design <- pmx_trial_design(
  dose_levels = 100, cohort_sizes = 90, sampling = c(0, 1, 2, 3, 7),
  n_doses = 3, dose_interval = 7,
  source = "the fixture's protocol: 100 mg on days 0, 7 and 14"
)
priors <- pmx_priors(pk = pmx_prior(
  c(1 / 4, 4),
  source = "scaling literature: the prediction is believed good to four-fold"
))
```

The prior is the input worth pausing on. It says the model's prediction is
believed accurate to about four-fold in either direction, which is a claim
anyone can make without knowing the clearance. It is also the dominant
sensitivity term: the noise is measured against its width.

## Is this release worth its budget?

Ask before spending. `pmx_preflight()` reads no data and spends nothing.

```{r preflight}
pmx_preflight(priors, epsilon = 0.1, n_subjects = 90)
```

At 90 subjects an epsilon of 0.1 is worth making. The same question at a
phase 1 cohort size gets a different answer:

```{r preflight-small}
pmx_preflight(priors, epsilon = 0.1, n_subjects = 12)
```

That is not a marginal call — the function says the release would tell you
nothing you did not already assume, and recommends prior-only generation
instead. **Run this before choosing epsilon, not after.**

## Spend the budget, once

```{r fit, eval=has_opendp}
synthetic <- synpmx_calibrated(
  data = study, roles = study_roles, model = public_model,
  design = public_design, priors = priors,
  epsilon = 0.1, seed = 404, backend = "opendp"
)
```

A warning here is expected rather than exceptional, and this page shows
whichever one this build's draw produced. At an epsilon of 0.1 the noise is a
third of the prior's width, so a fair proportion of draws land pressed against
a prior boundary — and when one does, the function says so, because the
generated data then reflects the boundary rather than the study. That is the
mechanism working: the release is still valid, and it is still nearly
uninformative. The
[public-data evaluation](https://iamstein.github.io/synpmx/articles/calibrated-public-data-examples.html)
draws this release two hundred times and shows the whole distribution, which is
the honest way to read a single one.

## What was released

```{r release, eval=has_opendp}
release <- attr(synthetic, "synpmx_release")
release
```

Two numbers left the study: a correction multiplying the assumed clearance, and
a noisy subject count. Everything else in the printed object is a public input
or an accounting entry.

```{r accounting, eval=has_opendp}
release$privacy$accounting$entries
```

The correction is not an estimate of anything, and the noisy count is not the
study's size — this study has exactly 90 subjects and the released count will
not say so. Compare the corrected clearance against the 2 L/day that was
assumed: whatever multiple this build drew, a single release at this epsilon is
one sample from a wide distribution, not a measurement.

## The generated data

```{r generated, eval=has_opendp}
str(synthetic)
```

The table wears the columns `study_roles` declared, so it drops into code
written against the source study without renaming anything.

```{r figure, eval=plots, fig.height=4}
observed <- function(data, label) {
  keep <- data$EVID == 0
  value <- suppressWarnings(as.numeric(data$DV[keep]))
  out <- data.frame(set = label, subject = paste(label, data$ID[keep]),
                    time = as.numeric(data$TIME[keep]), dv = value)
  out[is.finite(out$dv) & out$dv > 0, ]
}
frame <- rbind(observed(study, "Source"), observed(synthetic, "Synthetic"))
ggplot2::ggplot(frame, ggplot2::aes(time, dv, group = subject, colour = set)) +
  ggplot2::geom_line(alpha = 0.3) +
  ggplot2::geom_point(alpha = 0.35, size = 0.7) +
  ggplot2::facet_wrap(~ set) +
  ggplot2::scale_colour_manual(
    values = c(Source = "#1B6CA8", Synthetic = "#D95F02"), guide = "none") +
  ggplot2::labs(x = "Study time (days)", y = "Concentration (mg/L)",
                title = "mixroute_sim, calibrated at epsilon = 0.1") +
  ggplot2::theme_minimal()
```

## Further datasets are free

Generation from an existing release is post-processing. It reads no
confidential data and spends nothing, so any number of datasets can be drawn
from the one payment.

```{r again, eval=has_opendp}
another <- synpmx_generate(synthetic, seed = 405)
identical(names(another), names(synthetic))
```

Draw them this way rather than by calling `synpmx_calibrated()` again. A second
fit spends the budget a second time, and the function warns when it notices
one:

```{r repeat-warning, eval=has_opendp}
second_fit <- tryCatch(
  synpmx_calibrated(
    data = study, roles = study_roles, model = public_model,
    design = public_design, priors = priors,
    epsilon = 0.1, seed = 406, backend = "opendp"
  ),
  warning = function(w) conditionMessage(w)
)
cat(if (is.character(second_fit)) second_fit else "no warning", "\n")
```

## What this run does and does not support

The event structure, the schedule, the occasions and the censoring flag are
correct because the design said so. The concentration *level* has been pulled
toward the study by one released number. Everything else — the curve shape,
the between-subject spread, the residual error — is the public model's
assertion, uncorrected and uncosted.

So the output supports developing code against a dataset shaped like this
study, with a formal guarantee attached to the one quantity that was measured.
It does not support estimating a parameter, comparing arms, or concluding
anything about the compound.

## Where to go next

- `vignette("calibrated-algorithm")` — every step, what each one spends, and
  where the sensitivity bound comes from.
- [Evaluating calibration on public data](https://iamstein.github.io/synpmx/articles/calibrated-public-data-examples.html)
  — what this epsilon produces across repeated draws, and the cohort size below
  which it produces noise.
- [Feasibility by cohort size](https://iamstein.github.io/synpmx/articles/feasibility.html)
  — where `f` comes from.
- [What differential privacy does and does not guarantee](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html)
  — the trust-boundary decision rule, and what a production release needs.
