---
title: "Demo: one dataset, end to end"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: one dataset, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
run <- has_xgxr
plots <- has_xgxr && has_ggplot2
```

One pharmacometric (PMX) dataset taken from raw event table to synthetic event
table, in the order you would actually do it. Every number below is produced by
the run; nothing is quoted from a previous one.

The dataset is `xgxr::case1_pkpd`: 180 patients, six arms from placebo to
300 mg, with both a pharmacokinetic (PK) concentration endpoint and a
continuous pharmacodynamic (PD) endpoint. It is a good demonstration case
because its recorded sample times are *all* distinct — every patient's
observation schedule is unique before anything is done to it — which is the
condition the visit-grid machinery exists for.

To run this on your own study, edit the two chunks under **Configuration** and
leave the rest alone. Every check made along the way is collected into a
**scorecard** at the end, with the answer that counts as passing for each.
Companion reading: `vignette("synpmx-4-methods")` for why AVATAR blending is the
default method, `vignette("avatar-algorithm")` for what each step does
internally, and `vignette("synthetic-data-checks")` for the full list of
questions to ask of any synthetic dataset.

## Configuration

The data, and one look at it before touching anything.

```{r data, eval=run}
raw <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
SEED <- 808

table(raw$NAME, raw$TRTACT)
```

The column meanings. This is the whole configuration: AVATAR blending needs
no structural model, no priors, and no design — only which column plays which
role. Columns not named here are dropped, which the run says out loud.

```{r roles, eval=run}
roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",        # which endpoint each observation row is
  nominal_time = "NOMTIME",     # the protocol grid; see below for why it matters
  strata       = c("TRTACT", "DOSE"),
  covariates   = "WEIGHTB",     # baseline covariates, blended like everything else
  keep         = "STUDY"        # carried through verbatim
)
```

Two role choices are worth pausing on, because they are the ones that change
the result rather than describe the data.

- **`nominal_time`** is the declared protocol grid. Left out, the run infers a
  grid and reports it as inferred. Here the study has a real one, so declare it.
- **`strata`** are the treatment arms. Declaring them keeps the arm balance and
  stops a 3 mg patient being blended with a 300 mg patient, but it also splits
  the donor pool six ways. On a small cohort that trade can go the other way;
  the run report is how you tell.

`CENS` is deliberately *not* declared as `cens`. The column exists, but it is
all zeros with no censored records, and `validate_pmx()` refuses a censoring
column that never fires — see `vignette("synthetic-data-checks")`.

## Does the source validate?

Structure first. A dataset that fails here will fail confusingly later.

```{r validate, eval=run}
report <- validate_pmx(raw, roles, strict = FALSE)
stopifnot(report$valid)
report$valid
```

## How identifying is the visit schedule?

Run this *before* generating. Every column answers one question — **how many
patients share this property with me, me included** — so `1` means nobody else.

Scored as recorded:

```{r uniqueness-before, eval=run}
skeleton_uniqueness(raw, roles)
```

All 180 patients have an observation schedule nobody else has, and the cause is
a one-off observation time rather than a one-off pattern of attendance. That
distinction decides whether it is fixable: a one-off *time* snaps onto the
protocol grid, a one-off *attendance pattern* does not snap onto anything.

Scored on the grid `synpmx_avatar()` actually generates on:

```{r uniqueness-after, eval=run}
skeleton_uniqueness(raw, roles, coarsen_time = TRUE)
```

100% to 0%. Coarsening onto `NOMTIME` collapses 180 distinct schedules into two
— one for the 150 patients with PK sampling, one for the 30 placebo patients
without. This is the best case for the mechanism, and it is why declaring
`nominal_time` is worth the effort of finding the column.

```{r schedule-map, eval=run, fig.height=7}
plot_pmx_schedule(raw, roles)
```

Vertical stripes mean patients share visits; a ragged right edge is follow-up
ending. In the lower panel a bar of height one is a moment only one patient was
ever sampled at — exactly what would be copied onto an avatar — drawn in red.

## Synthesize

```{r synthesize, eval=run}
synthetic <- synpmx_avatar(raw, roles, seed = SEED)

stopifnot(validate_pmx(synthetic, roles)$valid)
c(rows = nrow(synthetic),
  subjects = length(unique(synthetic$ID)),
  shared_ids = length(intersect(synthetic$ID, raw$ID)))
```

Same size as the source, and no identifier is shared with it.

## What the masking mechanisms did

Each row carries the sentence that says what its number means, so this section
is deliberately short.

```{r masking, eval=run}
pmx_masking_report(synthetic, raw, roles)
```

Four rows carry the verdict, and on this run all four are clean.

- **Avatars carrying a visit set nobody else shares — 0%.** This is the row
  that must be zero. A non-zero value means some avatar wears one real
  patient's pattern of missed visits.
- **Avatars carrying a dose schedule nobody else shares — 0%.** The same
  guarantee for dosing, which is copied from the anchor verbatim.
- **Visit sets held by fewer than 2 patients, so not reused — 0 of 6.** This is
  where the fidelity cost would show, and here there is none: no real pattern
  of absences had to be discarded to get the two rows above to zero.
- **Effective donors per avatar — about 2.9**, against a floor of `k = 5`. This
  is the honest number: `k` is how many patients were blended, `1 / sum(w^2)`
  is how many of them actually mattered. The cap binds on 71% of avatars, so
  `max_donor_weight` is doing real work rather than sitting unused.

One row is a genuine limitation of this run: **amounts recomputed from a
covariate — no.** The five dose levels are not a fixed multiple of `WEIGHTB`,
so `AMT` is copied from the anchor and each avatar's implied mg/kg is its
anchor's rather than its own. For a weight- or body-surface-area-based study,
declare `dose_covariate` and the report will say it succeeded.

## Do the values line up?

Ranges and shapes, source against synthetic. These will not match exactly, and
they are not meant to: blending averages, so it shrinks spread.

```{r distributions, eval=run}
compare_pmx_distributions(raw, synthetic, roles)
```

Medians track closely on both endpoints. The standard deviations are visibly
smaller in the synthetic cohort — for the PD endpoint and for baseline weight
alike — which is blending doing exactly what blending does. Plan for it: this
data is for exercising software, not for estimating a variance component.

## Is any avatar too close to a real patient?

Everything above concerns structure. This measures the values.

```{r proximity, eval=run}
prox <- compare_pmx_proximity(raw, synthetic, roles)
prox
```

Read the two numbers together. `0.5` is the target rather than the maximum: it
means a synthetic patient is no more like a real patient than one real patient
is like another. The expected interval is not assumed — it is the same
statistic recomputed on halves of the real cohort, so small-sample artefacts
cancel. Landing inside it means *nothing was detected*, never that nothing is
there.

## Post-generation plausibility screen

`flag_identifiable_subjects()` scores each **synthetic** patient on follow-up
time, dose count, dose magnitude, and peak dependent variable, and flags robust
outliers. It is a plausibility check, not a masking step: a conspicuous blended
value identifies nobody, because it belongs to nobody.

```{r flag, eval=run}
flagged <- flag_identifiable_subjects(synthetic, roles)
sum(flagged$flagged)
knitr::kable(flagged[flagged$flagged, ])
```

Look at what it flags before acting on it. On a cohort with two clearly
different groups it will flag the split rather than a leak.
`remediate_identifiable_subjects()` will truncate, drop, or replace what it
returns.

## Source against synthetic

```{r overlay-setup, eval=plots}
library(ggplot2)

both <- rbind(
  cbind(raw[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$EVID == 0 & !is.na(both$LIDV), ]
obs$TRTACT <- factor(obs$TRTACT,
                     levels = c("Placebo", "3 mg", "10 mg", "30 mg",
                                "100 mg", "300 mg"))
arms <- c(source = "grey30", synthetic = "#C1272D")
```

The day-1 concentration profile, by arm, on a log scale:

```{r overlay-pk, eval=plots, fig.height=3.6}
ggplot(obs[obs$NAME == "PK Concentration" & obs$TIME <= 24, ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_wrap(~TRTACT, nrow = 1) +
  scale_y_log10() +
  scale_colour_manual(values = arms) +
  labs(x = "Time (hours)", y = "PK concentration", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

Dose ordering, absorption, and elimination all survive, and the avatars sit
inside the source cloud rather than beside it. Where they visibly differ is at
the edges of each arm — the most extreme source profiles have no synthetic
counterpart, which is the shrinkage again.

The full-duration pharmacodynamic response:

```{r overlay-pd, eval=plots, fig.height=3.6}
ggplot(obs[obs$NAME == "PD - Continuous", ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.35) +
  facet_wrap(~TRTACT, nrow = 1) +
  scale_colour_manual(values = arms) +
  labs(x = "Time (hours)", y = "PD (continuous)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The dose-ordered separation between placebo and 300 mg is preserved; the red
band is narrower than the grey one in every arm. Identifiers are disjoint by
construction, so grouping on `ID` never joins a real patient's line to an
avatar's.

## The scorecard

Every check above, with its answer and whether that answer passes. This is the
table from `vignette("synthetic-data-checks")` filled in by this run — the point
of collecting it is that the pass criteria are otherwise scattered through the
prose above, and deciding whether to ship a dataset should not require rereading
the document.

Nothing here is an exported helper. It is about forty lines of ordinary R
against the run report and the two tables, and it is written out rather than
hidden so you can lift it into your own study and change what it asks.

```{r scorecard, eval=run}
settings <- attr(synthetic, "pmx_settings")

# One row per check. `ok = NA` means the check has no pass mark and has to be
# read -- which is a real answer, not a missing one.
row <- function(check, question, reads, result, ok = NA) {
  data.frame(
    check = check, question = question, reads = reads, result = result,
    verdict = if (isTRUE(ok)) "pass" else if (isFALSE(ok)) "FAIL" else "review",
    stringsAsFactors = FALSE
  )
}

subjects <- function(data) length(unique(data[[roles$id]]))
# Endpoints are the `dvid` values carried by *observation* rows. This study's
# NAME column also has a "Dosing" level on its dose rows, which is not an
# endpoint and would inflate the count.
endpoints <- function(data) {
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  sort(unique(as.character(data[[roles$dvid]][observed])))
}
per_patient <- function(data, which) {
  events <- !as.character(data[[roles$evid]]) %in% c("0", "0.0")
  rows <- if (which == "dose") sum(events) else sum(!events)
  round(rows / subjects(data), 1)
}
# Each subject's sorted observation times as one string: two subjects share a
# string only if their schedules are identical.
time_vectors <- function(data) {
  observed <- data[as.character(data[[roles$evid]]) %in% c("0", "0.0"), ]
  tapply(as.numeric(observed[[roles$time]]),
         as.character(observed[[roles$id]]),
         function(times) paste(sort(times), collapse = ","))
}
# One baseline value per subject, as character so a factor cannot collapse to
# its integer codes.
holders <- function(data, column) {
  as.character(tapply(as.character(data[[column]]),
                      as.character(data[[roles$id]]),
                      function(x) x[1]))
}
# Of the levels that reached the output, how many source patients held the
# rarest one. This is scorecard row B5, by hand, over strata and any
# non-numeric covariate.
categorical <- c(roles$strata, roles$covariates)
categorical <- categorical[vapply(categorical,
                                  function(v) !is.numeric(raw[[v]]),
                                  logical(1))]
rarest_level <- min(vapply(categorical, function(column) {
  source_holders <- holders(raw, column)
  present <- intersect(unique(holders(synthetic, column)),
                       unique(source_holders))
  min(as.integer(table(factor(source_holders, levels = present))))
}, integer(1)))

arm_size <- function(data) {
  first <- !duplicated(as.character(data[[roles$id]]))
  table(as.character(data[[roles$strata[1]]])[first])
}
source_arms <- arm_size(raw)
synth_arms <- arm_size(synthetic)[names(source_arms)]
arms_matched <- sum(source_arms == synth_arms)

copies <- length(intersect(time_vectors(raw), time_vectors(synthetic)))
inside_null <- prox$adversarial_accuracy >= prox$null_lower &&
  prox$adversarial_accuracy <= prox$null_upper

card <- rbind(
  row("A1", "Synthetic table is a legal PMX dataset", "synthetic",
      as.character(validate_pmx(synthetic, roles)$valid),
      validate_pmx(synthetic, roles)$valid),
  row("A2", "Source is legal under the declared roles", "source",
      as.character(report$valid), report$valid),
  row("A3", "Every endpoint survived", "both",
      paste(length(endpoints(synthetic)), "of", length(endpoints(raw))),
      setequal(endpoints(raw), endpoints(synthetic))),
  row("A4", "Cohort size survived", "both",
      paste(subjects(raw), "->", subjects(synthetic)),
      subjects(raw) == subjects(synthetic)),
  row("A5", "Observations per patient", "both",
      paste(per_patient(raw, "obs"), "->", per_patient(synthetic, "obs"))),
  row("A5", "Doses per patient", "both",
      paste(per_patient(raw, "dose"), "->", per_patient(synthetic, "dose"))),
  row("B1a", "Avatars wearing one real patient's visit set", "run report",
      settings$identifying_visit_sets, settings$identifying_visit_sets == 0),
  row("B1b", "Avatars wearing one real patient's dose schedule", "run report",
      settings$identifying_dose_schedules,
      settings$identifying_dose_schedules == 0),
  row("B2", "Synthetic patients unusual within their own arm", "synthetic",
      paste(sum(flagged$flagged), "of", nrow(flagged))),
  row("B3", "Adversarial accuracy inside its null interval", "both",
      sprintf("%.3f in [%.3f, %.3f]", prox$adversarial_accuracy,
              prox$null_lower, prox$null_upper),
      inside_null),
  row("B4", "Generated time vectors copying a real one", "both",
      copies, copies == 0),
  row("B5", "Source patients holding the rarest exported level", "both",
      paste(rarest_level, "(floor", paste0(settings$min_pattern_share, ")")),
      rarest_level >= settings$min_pattern_share),
  row("C3", "Arms keeping their source size", "both",
      paste(arms_matched, "of", length(source_arms)),
      arms_matched == length(source_arms)),
  row("C4", "Dose regimens represented", "both",
      paste(settings$dose_regimens_represented, "of",
            settings$dose_regimens_source)),
  row("D2", "Effective donors per avatar", "run report",
      paste(round(settings$mean_effective_donors, 1), "of k =", settings$k))
)

knitr::kable(card, row.names = FALSE)
```

Three things to notice about how this reads.

**`review` is not a soft `pass`.** Five rows have no pass mark because no
threshold would be honest. Doses per patient is the clearest: on this study it
is unchanged, and on a study with individualised dosing it can halve while
every guarantee above it still reads 0. A checklist that scored that row would
have to pick a number, and any number picked would be wrong on some dataset.

**The `reads` column decides where the table can go.** Every row marked `source`
or `both` was computed from real patient data and inherits its handling
obligations, so the filled-in scorecard is itself restricted output. Only the
`synthetic` and `run report` rows can travel with the data.

**B5 is the row doing the least work.** It reports the rarest categorical level
that reached the output, which on a six-arm study with thirty patients per arm
is a comfortable 30. On a study with a two-patient stratum it would read 2, and
that is the whole finding — but nothing in the package computes it, so the lines
above are what you have to write. See `vignette("synthetic-data-checks")`,
section B5.

## What this run does and does not license

The output is fit for software development, package testing, reproducible
examples, teaching, and giving a coding agent something realistic to work
against. It reproduces the *structure* of a clinical trial event table:
dosing histories, multiple endpoints, arm balance, and the visit schedule.

It is not fit for estimating population parameters, and this run showed why:
variance is shrunk by blending.

The masking report's dose row is not a second reason here. `case1_pkpd` gives
every patient in an arm the same milligram dose, so M5 found no covariate the
amounts are proportional to and left them alone — which is right, and discloses
nothing, because thirty patients share each amount. On a weight- or
body-surface-area-based study that row matters: declare `dose_covariate` in
`pmx_roles()`, or each avatar carries its anchor's milligrams and therefore its
anchor's weight.

Two of the tables here (`compare_pmx_distributions()` and
`compare_pmx_proximity()`) are computed from the source data and are labelled
restricted in their own output. They belong in the trusted environment with the
real data, not in a document that leaves it.
