---
title: "Demo: one dataset, end to end"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: one dataset, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
run <- has_xgxr
plots <- has_xgxr && has_ggplot2
```

Generate a synthetic dataset, look at it, and read the scorecard. That is the
whole document.

`vignette("synthetic-data-checks")` is the reference for the checking half: it
states every check, what each one asks, what counts as passing, and which
questions no function can answer. Read it once. This demo runs the checks it can
and does not re-explain them.

The dataset is `xgxr::case1_pkpd`: 180 patients, six arms from placebo to
300 mg, with a pharmacokinetic (PK) concentration endpoint and a continuous
pharmacodynamic (PD) endpoint.

## Configuration

Two chunks. To run this on your own study, edit them and leave the rest alone.

```{r data, eval=run}
raw <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
SEED <- 808
```

The column meanings. This is the whole configuration: AVATAR blending needs no
structural model, no priors, and no design — only which column plays which role.
Columns not named here are dropped, which the run says out loud.

```{r roles, eval=run}
roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",        # which endpoint each observation row is
  nominal_time = "NOMTIME",     # the protocol grid
  strata       = c("TRTACT", "DOSE"),
  covariates   = "WEIGHTB",     # blended like everything else
  keep         = "STUDY"        # carried through verbatim
)
```

Two role choices change the result rather than describe the data.

- **`nominal_time`** is the declared protocol grid. Left out, the run infers a
  grid and reports it as inferred. Here the study has a real one, so declare it.
- **`strata`** are the treatment arms. Declaring them keeps the arm balance and
  stops a 3 mg patient being blended with a 300 mg patient, but it also splits
  the donor pool six ways. On a small cohort that trade can go the other way.

## Generate

```{r synthesize, eval=run}
synthetic <- synpmx_avatar(raw, roles, seed = SEED)

c(rows = nrow(synthetic),
  subjects = length(unique(synthetic$ID)),
  shared_ids = length(intersect(synthetic$ID, raw$ID)))
```

Same size as the source, and no identifier is shared with it.

## Look at it

Source and synthetic overlaid, by arm. Identifiers are disjoint by
construction, so grouping on `ID` never joins a real patient's line to an
avatar's.

```{r overlay-setup, eval=plots}
library(ggplot2)

both <- rbind(
  cbind(raw[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$EVID == 0 & !is.na(both$LIDV), ]
obs$TRTACT <- factor(obs$TRTACT,
                     levels = c("Placebo", "3 mg", "10 mg", "30 mg",
                                "100 mg", "300 mg"))
arms <- c(source = "grey30", synthetic = "#C1272D")
```

The day-1 concentration profile, on a log scale:

```{r overlay-pk, eval=plots, fig.height=3.6}
ggplot(obs[obs$NAME == "PK Concentration" & obs$TIME <= 24, ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_wrap(~TRTACT, nrow = 1) +
  scale_y_log10() +
  scale_colour_manual(values = arms) +
  labs(x = "Time (hours)", y = "PK concentration", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The full-duration pharmacodynamic response:

```{r overlay-pd, eval=plots, fig.height=3.6}
ggplot(obs[obs$NAME == "PD - Continuous", ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.35) +
  facet_wrap(~TRTACT, nrow = 1) +
  scale_colour_manual(values = arms) +
  labs(x = "Time (hours)", y = "PD (continuous)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

Dose ordering, absorption, elimination and the placebo-to-300 mg separation all
survive, and the avatars sit inside the source cloud rather than beside it.
Where they visibly differ is at the edges of each arm: the red band is narrower
than the grey one everywhere, because blending averages and averaging shrinks
spread.

The same thing as numbers, source against synthetic:

```{r distributions, eval=run}
compare_pmx_distributions(raw, synthetic, roles)
```

Medians track closely on both endpoints. The standard deviations are visibly
smaller in the synthetic cohort — for the PD endpoint and for baseline weight
alike. Plan for it: this data is for exercising software, not for estimating a
variance component.

## The scorecard

One line. Every check from `vignette("synthetic-data-checks")` that can be run
automatically, with the answer and whether that answer passes.

```{r scorecard, eval=run}
pmx_scorecard(raw, synthetic, roles)
```

Three things to notice about how this reads.

**`review` is not a soft `pass`.** Four rows have no pass mark because no
threshold would be honest. Doses per patient is the clearest: on this study it
is unchanged, and on a study with individualised dosing it can halve while
every guarantee above it still reads 0. Any number picked for it would be wrong
on some dataset.

**The `reads` column decides where the table can go.** Every row marked `source`
or `both` was computed from real patient data and inherits its handling
obligations, so the filled-in scorecard is itself restricted output. Only the
`synthetic` and `run report` rows can travel with the data.

**The scorecard is not the whole checklist.** It omits what no function can
produce: dose and observation ordering, the census of how many *real* patients
held each categorical level, and the one check that decides whether this dataset
is any use — does the pipeline that will consume the real study run unchanged
against it? Run your own code against the output and see.

## Where to go next

- `vignette("synthetic-data-checks")` — every check, what it asks, and what
  passing means.
- `vignette("synpmx-4-methods")` — why AVATAR blending is the default.
- `vignette("avatar-algorithm")` — what each generation step does internally.
