---
title: "Demo: one dataset, end to end"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: one dataset, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
run <- has_xgxr
plots <- has_xgxr && has_ggplot2
```

One pharmacometric (PMX) dataset taken from raw event table to synthetic event
table, in the order you would actually do it. Every number below is produced by
the run; nothing is quoted from a previous one.

The dataset is `xgxr::case1_pkpd`: 180 patients, six arms from placebo to
300 mg, with both a pharmacokinetic (PK) concentration endpoint and a
continuous pharmacodynamic (PD) endpoint. It is a good demonstration case
because its recorded sample times are *all* distinct — every patient's
observation schedule is unique before anything is done to it — which is the
condition the visit-grid machinery exists for.

To run this on your own study, edit the two chunks under **Configuration** and
leave the rest alone. Companion reading: `vignette("synpmx-4-methods")` for why
AVATAR blending is the default method, `vignette("avatar-algorithm")` for what
each step does internally, and `vignette("synthetic-data-checks")` for the full
list of questions to ask of any synthetic dataset.

## Configuration

The data, and one look at it before touching anything.

```{r data, eval=run}
raw <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
SEED <- 808

table(raw$NAME, raw$TRTACT)
```

The column meanings. This is the whole configuration: AVATAR blending needs
no structural model, no priors, and no design — only which column plays which
role. Columns not named here are dropped, which the run says out loud.

```{r roles, eval=run}
roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",        # which endpoint each observation row is
  nominal_time = "NOMTIME",     # the protocol grid; see below for why it matters
  strata       = c("TRTACT", "DOSE"),
  covariates   = "WEIGHTB",     # baseline covariates, blended like everything else
  keep         = "STUDY"        # carried through verbatim
)
```

Two role choices are worth pausing on, because they are the ones that change
the result rather than describe the data.

- **`nominal_time`** is the declared protocol grid. Left out, the run infers a
  grid and reports it as inferred. Here the study has a real one, so declare it.
- **`strata`** are the treatment arms. Declaring them keeps the arm balance and
  stops a 3 mg patient being blended with a 300 mg patient, but it also splits
  the donor pool six ways. On a small cohort that trade can go the other way;
  the run report is how you tell.

`CENS` is deliberately *not* declared as `cens`. The column exists, but it is
all zeros with no censored records, and `validate_pmx()` refuses a censoring
column that never fires — see `vignette("synthetic-data-checks")`.

## Does the source validate?

Structure first. A dataset that fails here will fail confusingly later.

```{r validate, eval=run}
report <- validate_pmx(raw, roles, strict = FALSE)
stopifnot(report$valid)
report$valid
```

## How identifying is the visit schedule?

Run this *before* generating. Every column answers one question — **how many
patients share this property with me, me included** — so `1` means nobody else.

Scored as recorded:

```{r uniqueness-before, eval=run}
skeleton_uniqueness(raw, roles)
```

All 180 patients have an observation schedule nobody else has, and the cause is
a one-off observation time rather than a one-off pattern of attendance. That
distinction decides whether it is fixable: a one-off *time* snaps onto the
protocol grid, a one-off *attendance pattern* does not snap onto anything.

Scored on the grid `synpmx_avatar()` actually generates on:

```{r uniqueness-after, eval=run}
skeleton_uniqueness(raw, roles, coarsen_time = TRUE)
```

100% to 0%. Coarsening onto `NOMTIME` collapses 180 distinct schedules into two
— one for the 150 patients with PK sampling, one for the 30 placebo patients
without. This is the best case for the mechanism, and it is why declaring
`nominal_time` is worth the effort of finding the column.

```{r schedule-map, eval=run, fig.height=7}
plot_pmx_schedule(raw, roles)
```

Vertical stripes mean patients share visits; a ragged right edge is follow-up
ending. In the lower panel a bar of height one is a moment only one patient was
ever sampled at — exactly what would be copied onto an avatar — drawn in red.

## Synthesize

```{r synthesize, eval=run}
synthetic <- synpmx_avatar(raw, roles, seed = SEED)

stopifnot(validate_pmx(synthetic, roles)$valid)
c(rows = nrow(synthetic),
  subjects = length(unique(synthetic$ID)),
  shared_ids = length(intersect(synthetic$ID, raw$ID)))
```

Same size as the source, and no identifier is shared with it.

## What the masking mechanisms did

Each row carries the sentence that says what its number means, so this section
is deliberately short.

```{r masking, eval=run}
pmx_masking_report(synthetic, raw, roles)
```

Four rows carry the verdict, and on this run all four are clean.

- **Avatars carrying a visit set nobody else shares — 0%.** This is the row
  that must be zero. A non-zero value means some avatar wears one real
  patient's pattern of missed visits.
- **Avatars carrying a dose schedule nobody else shares — 0%.** The same
  guarantee for dosing, which is copied from the anchor verbatim.
- **Visit sets held by fewer than 2 patients, so not reused — 0 of 6.** This is
  where the fidelity cost would show, and here there is none: no real pattern
  of absences had to be discarded to get the two rows above to zero.
- **Effective donors per avatar — about 2.9**, against a floor of `k = 5`. This
  is the honest number: `k` is how many patients were blended, `1 / sum(w^2)`
  is how many of them actually mattered. The cap binds on 71% of avatars, so
  `max_donor_weight` is doing real work rather than sitting unused.

One row is a genuine limitation of this run: **amounts recomputed from a
covariate — no.** The five dose levels are not a fixed multiple of `WEIGHTB`,
so `AMT` is copied from the anchor and each avatar's implied mg/kg is its
anchor's rather than its own. For a weight- or body-surface-area-based study,
declare `dose_covariate` and the report will say it succeeded.

## Do the values line up?

Ranges and shapes, source against synthetic. These will not match exactly, and
they are not meant to: blending averages, so it shrinks spread.

```{r distributions, eval=run}
compare_pmx_distributions(raw, synthetic, roles)
```

Medians track closely on both endpoints. The standard deviations are visibly
smaller in the synthetic cohort — for the PD endpoint and for baseline weight
alike — which is blending doing exactly what blending does. Plan for it: this
data is for exercising software, not for estimating a variance component.

## Is any avatar too close to a real patient?

Everything above concerns structure. This measures the values.

```{r proximity, eval=run}
compare_pmx_proximity(raw, synthetic, roles)
```

Read the two numbers together. `0.5` is the target rather than the maximum: it
means a synthetic patient is no more like a real patient than one real patient
is like another. The expected interval is not assumed — it is the same
statistic recomputed on halves of the real cohort, so small-sample artefacts
cancel. Landing inside it means *nothing was detected*, never that nothing is
there.

## Post-generation plausibility screen

`flag_identifiable_subjects()` scores each **synthetic** patient on follow-up
time, dose count, dose magnitude, and peak dependent variable, and flags robust
outliers. It is a plausibility check, not a masking step: a conspicuous blended
value identifies nobody, because it belongs to nobody.

```{r flag, eval=run}
flagged <- flag_identifiable_subjects(synthetic, roles)
sum(flagged$flagged)
knitr::kable(flagged[flagged$flagged, ])
```

Look at what it flags before acting on it. On a cohort with two clearly
different groups it will flag the split rather than a leak.
`remediate_identifiable_subjects()` will truncate, drop, or replace what it
returns.

## Source against synthetic

```{r overlay-setup, eval=plots}
library(ggplot2)

both <- rbind(
  cbind(raw[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$EVID == 0 & !is.na(both$LIDV), ]
obs$TRTACT <- factor(obs$TRTACT,
                     levels = c("Placebo", "3 mg", "10 mg", "30 mg",
                                "100 mg", "300 mg"))
arms <- c(source = "grey30", synthetic = "#C1272D")
```

The day-1 concentration profile, by arm, on a log scale:

```{r overlay-pk, eval=plots, fig.height=3.6}
ggplot(obs[obs$NAME == "PK Concentration" & obs$TIME <= 24, ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_wrap(~TRTACT, nrow = 1) +
  scale_y_log10() +
  scale_colour_manual(values = arms) +
  labs(x = "Time (hours)", y = "PK concentration", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

Dose ordering, absorption, and elimination all survive, and the avatars sit
inside the source cloud rather than beside it. Where they visibly differ is at
the edges of each arm — the most extreme source profiles have no synthetic
counterpart, which is the shrinkage again.

The full-duration pharmacodynamic response:

```{r overlay-pd, eval=plots, fig.height=3.6}
ggplot(obs[obs$NAME == "PD - Continuous", ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.35) +
  facet_wrap(~TRTACT, nrow = 1) +
  scale_colour_manual(values = arms) +
  labs(x = "Time (hours)", y = "PD (continuous)", colour = NULL) +
  theme_bw() +
  theme(legend.position = "top")
```

The dose-ordered separation between placebo and 300 mg is preserved; the red
band is narrower than the grey one in every arm. Identifiers are disjoint by
construction, so grouping on `ID` never joins a real patient's line to an
avatar's.

## What this run does and does not license

The output is fit for software development, package testing, reproducible
examples, teaching, and giving a coding agent something realistic to work
against. It reproduces the *structure* of a clinical trial event table:
dosing histories, multiple endpoints, arm balance, and the visit schedule.

It is not fit for estimating population parameters, and this run showed both
reasons why — variance is shrunk by blending, and dose amounts were not
rescaled to each avatar's own weight.

Two of the tables here (`compare_pmx_distributions()` and
`compare_pmx_proximity()`) are computed from the source data and are labelled
restricted in their own output. They belong in the trusted environment with the
real data, not in a document that leaves it.
