---
title: "What a trial summary contains"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{What a trial summary contains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_dt <- requireNamespace("DT", quietly = TRUE)
run <- has_xgxr
plots <- has_xgxr && has_ggplot2
tables <- run && has_dt

# Anything more than a few rows is a sortable, scrollable widget rather than a
# console dump; short things are printed.
show <- function(x, caption = NULL, paged = FALSE) {
  DT::datatable(
    x, rownames = FALSE, caption = caption,
    options = list(
      dom = if (paged) "tip" else "t", paging = paged,
      pageLength = 10, scrollX = TRUE, ordering = TRUE
    )
  )
}
```

`synpmx_pca_summarize()` returns a **trial summary**: everything a synthetic
copy of the study will be built from, and nothing else.
`synpmx_pca_generate()` reads that object and no patient row, so what is listed
here is the complete account of what the generated dataset descends from.

This vignette walks through the object group by group. It is a reference: read
`vignette("pca-demo")` first for a run end to end, and
`vignette("pca-algorithm")` for how each quantity is estimated and why.

The object is a plain list, and the names used below are its own, so a quantity
can be reached directly when no accessor covers it.

## The study

```{r data, eval=run, message=FALSE}
library(dplyr)
raw <- as.data.frame(get(utils::data(list = "case1_pkpd", package = "xgxr"))) |>
  mutate(CENS = ifelse(NAME == "PD - Continuous", 0, CENS))

roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  cens         = "CENS",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",
  nominal_time = "NOMTIME",
  strata       = c("TRTACT", "DOSE"),
  covariates   = "WEIGHTB",
  keep         = "STUDY"
)

trial_summary <- synpmx_pca_summarize(raw, roles, seed = 808)
trial_summary
```

## The inventory

`pca_report()` is the top-level account: one row per released quantity, how many
numbers it holds, and the smallest number of patients standing behind any one of
them.

```{r report, eval=tables}
show(as.data.frame(pca_report(trial_summary)),
     "Every quantity read out of the source data")
```

```{r report-plain, eval=run && !has_dt}
pca_report(trial_summary)
```

Every section below expands one of those rows.

## The settings that produced it

Short, so printed.

```{r settings, eval=run}
unlist(trial_summary$settings)
c(patients = trial_summary$n_source, components = trial_summary$basis$k)
```

## The arms

Sizes are what the generated cohort is split by, in the same proportions.

```{r arms, eval=run}
data.frame(
  arm = trial_summary$arms$arms,
  patients = as.integer(trial_summary$arms$sizes[trial_summary$arms$arms])
)
```

## The feature grid

Every column of the matrix the components are built on: one per baseline
covariate, and one per endpoint per retained nominal time. `patients` is how
many subjects hold an observation in that cell, and cells held by fewer than
`min_column_patients` were dropped rather than modelled.

```{r grid, eval=tables}
show(as.data.frame(pca_features(trial_summary)),
     "Every feature, with its mean and standard deviation", paged = TRUE)
```

`center` and `scale` are on the modelling scale, which the `transform` column
names: the log scale for a positive endpoint, the original scale otherwise.

## The endpoint transforms and assay limits

```{r transforms, eval=run}
basis <- trial_summary$basis
data.frame(
  endpoint = names(basis$transforms),
  transform = vapply(basis$transforms, function(t) t$method, character(1)),
  offset = signif(vapply(basis$transforms, function(t) t$offset, numeric(1)), 4),
  lloq = vapply(trial_summary$schema$censoring, function(c) {
    if (is.null(c$left)) NA_real_ else c$left
  }, numeric(1)),
  row.names = NULL
)
```

The offset keeps `log()` finite near zero. The lower limit of quantification is
what a generated value below it is reported at, with `CENS` set.

## The components

`pca_components()` gives one loading per component and retained cell. Plotted
against time the components become curves: flat and positive is overall
magnitude, and crossing zero separates early visits from late ones.

```{r variance, eval=tables}
variance <- attr(pca_components(trial_summary), "variance_explained")
variance$variance_explained <- round(variance$variance_explained, 4)
variance$cumulative <- round(variance$cumulative, 4)
show(variance, "Variance explained, per component")
```

```{r loadings-plot, eval=plots, fig.height=4.5}
library(ggplot2)
library(xgxr)
xgx_theme_set()

components <- pca_components(trial_summary)
ggplot(subset(components, component %in% c("PC1", "PC2", "PC3") &
                 !is.na(time)),
       aes(time, loading, colour = component)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_line() + geom_point(size = 1) +
  facet_wrap(~endpoint, scales = "free_x") +
  labs(x = "Nominal time (hours)", y = "Loading", colour = NULL) +
  theme(legend.position = "top")
```

```{r loadings-table, eval=tables}
loadings <- components
loadings$loading <- round(loadings$loading, 4)
show(loadings, "Every loading, by component and cell", paged = TRUE)
```

One matrix carries the covariates and every endpoint together, so a single draw
keeps the relationships between them. The loading mass per component says where
each one is concentrated:

```{r loading-mass, eval=run}
mass <- tapply(
  components$loading^2,
  list(components$component,
       ifelse(is.na(components$endpoint), "covariate", components$endpoint)),
  sum
)
round(mass[paste0("PC", seq_len(min(4, nrow(mass)))), , drop = FALSE], 3)
```

Each row sums to one. On this study the first two components are almost entirely
the PK profile, and the covariate only enters further down --- alongside the PD
endpoint, which is the joint structure that a separate decomposition per
endpoint would have thrown away.

## The score model

A new subject's scores are their arm's mean plus a fresh draw whose spread is
that arm's residual standard deviation.

```{r scores, eval=tables}
scores <- as.data.frame(pca_scores(trial_summary))
scores$mean <- round(scores$mean, 3)
scores$sd <- round(scores$sd, 3)
show(scores, "Mean score and residual spread, per arm and component",
     paged = TRUE)
```

The `sd` column is the whole of the between-subject variability the synthetic
data will have. It differs by arm on purpose: an arm sitting on the assay limit
is genuinely tighter than one well above it, and giving every arm the pooled
spread would smear the low arms upward.

## The dosing model

`pca_dosing()` is the planned schedule each arm starts from, and
`pca_dose_rates()` is how its patients departed from it.

```{r dosing, eval=tables}
show(head(pca_dosing(trial_summary), 12),
     "The planned schedule, first cycles of each arm", paged = TRUE)
```

```{r rates, eval=tables}
show(pca_dose_rates(trial_summary), "How patients departed from the plan")
```

```{r rates-plain, eval=run && !has_dt}
pca_dose_rates(trial_summary)
```

All three rates are zero here and every arm has a single dose level, so each
generated patient receives the planned schedule exactly. On a study with dose
modifications the rates are not zero, and generated patients then differ from
one another as the source patients did.

## The visit model

One probability per arm, endpoint and modelled time. Attendance is drawn
independently per visit, so no real patient's set of attended visits is reused.

```{r visits-plot, eval=plots, fig.height=4}
ggplot(pca_visits(trial_summary), aes(time, probability, colour = arm)) +
  geom_line() +
  facet_wrap(~endpoint, scales = "free_x") +
  labs(x = "Nominal time (hours)", y = "P(observation)", colour = NULL) +
  theme(legend.position = "top")
```

```{r visits-table, eval=tables}
visits <- pca_visits(trial_summary)
visits$probability <- round(visits$probability, 3)
show(visits, "Probability of an observation, per arm, endpoint and time",
     paged = TRUE)
```

## The schema

What the generated table needs in order to come back in the source's shape.
Column prototypes are zero-length vectors: they carry class and factor levels
and no values.

```{r schema, eval=tables}
schema <- trial_summary$schema
show(data.frame(
  column = schema$columns,
  class = vapply(schema$prototypes, function(p) class(p)[[1L]], character(1)),
  levels = vapply(schema$prototypes, function(p) {
    if (is.factor(p)) paste(levels(p), collapse = ", ") else ""
  }, character(1))
), "Column order and type")
```

Two entries are read from real values rather than averaged, and both are worth
knowing about. **Endpoint level sets**: a discrete endpoint has no value between
its levels, so a generated value is a source value.

```{r endpoint-types, eval=run}
data.frame(
  endpoint = names(schema$endpoint_specs),
  type = vapply(schema$endpoint_specs, function(s) s$type, character(1)),
  levels = vapply(schema$endpoint_specs, function(s) {
    if (is.null(s$levels)) "" else paste(s$levels, collapse = ", ")
  }, character(1)),
  reason = vapply(schema$endpoint_specs, function(s) s$reason, character(1)),
  row.names = NULL
)
```

**Arm constants**: the strata and kept columns, one value per arm, carried
verbatim onto every generated patient in it.

```{r arm-values, eval=run}
do.call(rbind, lapply(names(schema$arm_values), function(arm) {
  values <- schema$arm_values[[arm]]
  data.frame(arm = gsub("\r", " / ", arm, fixed = TRUE),
             as.data.frame(values, stringsAsFactors = FALSE))
}))
```

Subject identifiers are not in the schema. They are minted at generation, and
the only thing kept about the source's identifiers is the largest one, so a
synthetic identifier cannot collide with a real one.

```{r ids, eval=run}
c(class = schema$id_class, offset = schema$id_offset)
```

## Generating from it

Nothing beyond the object above is read.

```{r generate, eval=run}
synthetic <- synpmx_pca_generate(trial_summary, seed = 808)
c(rows = nrow(synthetic),
  subjects = length(unique(synthetic$ID)),
  valid = validate_pmx(synthetic, roles)$valid)
```

## Where to go next

- `vignette("pca-demo")` — one study end to end, with the source and synthetic
  data plotted against each other.
- `vignette("pca-algorithm")` — how each quantity above is estimated, and what
  it does and does not preserve.
