---
title: "Demo: Using synpmx_pca"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx_pca}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
run <- has_xgxr
plots <- has_xgxr && has_ggplot2
distributions <- plots && requireNamespace("patchwork", quietly = TRUE)
```

Generate a synthetic dataset from a fitted model rather than from real patients'
values, look at it, and read the checks of the data.

`synpmx_pca_summarize()` reduces the study to summaries and
`synpmx_pca_generate()` builds a dataset from those summaries alone. No number a
patient measured reaches the output. What it carries out of
the source is a mean, a scale, a set of principal-component loadings, one mean
score vector per arm, and a residual covariance. `synpmx_avatar()` blends real
values from neighbouring patients and is the subject of `vignette("avatar-demo")`,
which runs the same study through that algorithm.

It makes no formal privacy claim, and it is not for estimation. The full
specification is in `vignette("pca-algorithm")` and the checks in
`vignette("avatar-scorecard")`.

The dataset is `xgxr::case1_pkpd`: 180 patients, six arms from placebo to 300
mg, with a pharmacokinetic (PK) concentration endpoint and a continuous
pharmacodynamic (PD) endpoint. It is the dataset `vignette("avatar-demo")` uses,
so the two runs can be read against each other.

## Configuration

To run this on your own study, edit the two chunks below that:

1. Read in the dataset.
2. Define the column roles.

The rest of the code can be kept as is.

```{r data, eval=run, message = FALSE}
library(dplyr)
raw <- as.data.frame(get(utils::data(list = "case1_pkpd", package = "xgxr"))) |>
  mutate(CENS = ifelse(NAME == "PD - Continuous", 0, CENS))
SEED <- 808
```

The column meanings.  Columns not specified here are dropped from the synthetic dataset.

```{r roles, eval=run}
roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  cens         = "CENS",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",        # which endpoint each observation row is
  nominal_time = "NOMTIME",     # required: the grid the model is built on
  strata       = c("TRTACT", "DOSE"), # treatment arms, and the score model's groups
  covariates   = "WEIGHTB",     # drawn jointly with the trajectories
  keep         = "STUDY"        # carried through verbatim
)
```

## Summarize the Study

Generation happens in two steps, and they are separate so that the first can be
looked at before the second runs. `synpmx_pca_summarize()` is the only step that
reads patient data.

```{r summarize, eval=run, message = FALSE}
model <- synpmx_pca_summarize(raw, roles)
model
```

**`nominal_time` is required.** The grid it names is the axis every feature sits
on, and dose rows and observation rows are placed on it together, so the timing
comes through at its source values: doses per patient, the last dose, the end of
follow-up, the baseline sample before the first dose. `synpmx_avatar()` will
derive a grid when none is declared; `synpmx_pca()` refuses, because choosing
where a visit belongs is a statement about the protocol rather than something to
infer from recorded times.

`strata` does more work here than it does in `synpmx_avatar()`. It names the
arms, and the arms are the groups the score model is fitted against: each one
gets its own mean score vector and its own residual covariance. An arm of fewer
than three patients has no spread of its own to model, so the function refuses
rather than generating from one or two people. `min_arm_patients` is where that
floor is set.

## What the summary read out of the source

```{r report, eval=run}
pmx_pca_report(model)
```

`min_patients` is the column to read. A grid cell or a covariate mean is backed
by most of the cohort; a per-arm quantity is backed by one arm. Nothing in this
table is a patient record, but a quantity standing on few patients is where
disclosure risk sits, not in the loadings.

## The dosing model

Dosing is not copied from anybody. Each arm contributes the schedule its
patients hold in common, and that schedule is what every generated subject in
the arm receives.

```{r dosing, eval=run}
dosing <- pmx_pca_dosing(model)
aggregate(cbind(doses = dose) ~ arm + amt + share + distinct, dosing, length)
```

`share` is the fraction of the arm holding that schedule and `distinct` is how
many schedules the arm actually contained. Both at their best here: one schedule
per arm, held by everyone. A study recording actual dose times rather than
planned ones would show `distinct` close to the arm size and `share` close to
`1/n`, and the generated data would hold one schedule where the source held
thirty. The timing would still be right; the variety would be gone.

## The visit model

```{r plot-setup, eval=plots}
library(ggplot2)
library(xgxr)
xgx_theme_set()
comparison_colours <- c(source = "#1B6CA8", synthetic = "#D95F02")
```

```{r visits, eval=plots, fig.height=4}
ggplot(pmx_pca_visits(model), aes(time, probability, colour = arm)) +
  geom_line() +
  facet_wrap(~endpoint, scales = "free_x") +
  labs(x = "Nominal time (hours)", y = "P(observation)", colour = NULL) +
  theme(legend.position = "top")
```

Attendance is drawn per visit from these probabilities, so no real patient's set
of attended visits is reused.

## Generate Synthetic Data

Everything above is what the synthetic data will be built from.
`synpmx_pca_generate()` reads the model and a subject count, and no patient row
is in scope while it runs.

```{r generate, eval=run, message = FALSE}
synthetic <- synpmx_pca_generate(model, seed = SEED)
```

`synpmx_pca(raw, roles, seed = SEED)` is the same two calls in one, for when
there is no reason to stop and look.

## Plot synthetic data and original data

```{r overlay-setup, eval=plots}
both <- rbind(
  cbind(raw[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$EVID == 0 & !is.na(both$LIDV), ]
obs$TRTACT <- factor(obs$TRTACT,
                     levels = c("Placebo", "3 mg", "10 mg", "30 mg",
                                "100 mg", "300 mg"))
```

```{r overlay-pk, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PK Concentration" & obs$TIME <= 24, ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA~TRTACT) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 24, by = 6)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "PK concentration", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("Day 1 Conc. Profile")
```

```{r overlay-pd, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PD - Continuous", ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.35) +
  xgx_scale_x_time_units(units_dataset = "hours", units_plot = "weeks") +
  facet_grid(DATA~TRTACT) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "PD (continuous)", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("PD Response")
```

The synthetic profiles are smoother than the source ones. The model holds a
handful of components, and everything outside them, including measurement noise
visit to visit, is not reproduced.

The low-dose arms sit flat on the assay limit in both panels. That is the
censoring being put back: the drawn value is the latent one, and where the
source declared a lower limit of quantification the reported value, `CENS` and
`LIMIT` are rebuilt from it together. The basis itself is fitted on values drawn
inside the censoring region rather than on the stack of identical limits, so it
describes the patients rather than the assay.

## Reading the components

A principal component is a direction in the space of whole trajectories, and its
loadings say how much each visit contributes to it. Plotted against time rather
than tabulated, the components become curves: a component that is flat and
positive across a profile is overall magnitude, and one that crosses zero
separates early visits from late ones.

```{r components, eval=plots, fig.height=4.5}
components <- pmx_pca_components(model)
ggplot(subset(components, component %in% c("PC1", "PC2", "PC3")),
       aes(time, loading, colour = component)) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_line() + geom_point(size = 1) +
  facet_wrap(~endpoint, scales = "free_x") +
  labs(x = "Nominal time (hours)", y = "Loading", colour = NULL) +
  theme(legend.position = "top")
```

```{r variance, eval=run}
attr(pmx_pca_components(model), "variance_explained")
```

Describe what the curve shows rather than naming a mechanism for it. A component
that separates early from late looks like a difference in decline rate, and
calling it a half-life would be a claim the fit cannot support: no clearance was
estimated, and a score standard deviation is a variance decomposition rather
than a pharmacokinetic parameter. That is the price of this basis, and
`vignette("pca-algorithm")` states what the alternative would cost.

## Distributions of Synthetic and Original Data

```{r distributions, eval=distributions, fig.height=compare_pmx_distributions_height(raw, roles)}
compare_pmx_distributions(raw, synthetic, roles)
```

## The scorecard

The scorecard computes checks of the synthetic dataset.  It is described further in `vignette("avatar-scorecard")`.

```{r scorecard, eval=run}
scorecard <- synpmx_scorecard(raw, synthetic, roles)
synpmx_scorecard_datatable(scorecard)
```

**Fifteen of the eighteen rows mean the same thing here as they do for
`synpmx_avatar()`.** They read the source and synthetic tables and compute the
same way whatever produced the data, which is what the scorecard was for.

**Three read `unavailable`, and that is not the same as unanswered.** B1a, B1b
and C2 read the AVATAR run record rather than the tables, and this table does not
carry one. For B1a and B1b the guarantee is structural instead: no individual's
visit set and no individual's dose schedule exists in the model to be copied.
B4a and B4b ask the copy question directly, on the finished tables, and both
read zero.

**C2 is the one to be careful about.** It asks how many distinct dose-time
schedules survived, and this is the row where the loss is largest — one schedule
per arm, against one per patient in a study recording actual dose times. Read
`pmx_pca_dosing()` in its place, where `distinct` and `share` say the same thing
per arm.

## Where to go next

- `vignette("pca-algorithm")` — every step of this algorithm, and what each one
  reads out of the source.
- `vignette("avatar-demo")` — the same study through `synpmx_avatar()`, which
  blends real values instead.
- `vignette("avatar-scorecard")` — every check, what it asks, and what passing
  means.
