---
title: "Evaluating PCA on public data"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Evaluating PCA on public data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
# `compare_pmx_distributions()` composes its panels with patchwork and refuses
# to draw without it, rather than degrading the way the rest of the plotting
# does. Guarding those chunks keeps a missing suggested package a missing
# figure instead of a failed build.
has_distributions <- has_ggplot2 &&
  requireNamespace("patchwork", quietly = TRUE)
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")
```

This vignette runs `synpmx_pca()` on the same eight public datasets that
`vignette("avatar-public-data-examples")` runs `synpmx_avatar()` on, with the
role declarations copied from that document unchanged, so that the generator is
the only thing that differs between the two.

`synpmx_pca()` refuses to run on a dataset it cannot model, and the first
question this survey answers is how many of the eight it accepts. The answer is
in "What ran, and what did not" at the end, computed by the runs below rather
than written here.

**This is not a fidelity ranking of the two generators.** A dataset PCA refuses
is not a dataset AVATAR handles better; it is a dataset PCA declines to guess
about. Where both run, the two scorecards are comparable and the figures are
drawn the same way.

<details>
<summary>Plotting and run helpers used throughout this vignette</summary>

```{r plot-helpers}
observed_plot_data <- function(data, roles, dataset,
                               clock = "study_time") {
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  if (!is.null(roles$mdv)) {
    observed <- observed & as.character(data[[roles$mdv]]) %in% c("0", "0.0")
  }
  observed <- observed & !is.na(data[[roles$dv]])
  observation_rows <- which(observed)
  occasion <- rep(1L, length(observation_rows))
  tad <- rep(NA_real_, length(observation_rows))
  if (!is.null(roles$occasion)) {
    declared <- suppressWarnings(as.integer(
      data[[roles$occasion]][observation_rows]
    ))
    valid <- !is.na(declared) & declared >= 1L
    occasion[valid] <- declared[valid]
  }
  if (!is.null(roles$tad)) {
    declared <- suppressWarnings(as.numeric(data[[roles$tad]][observation_rows]))
    valid <- is.finite(declared)
    tad[valid] <- pmax(0, declared[valid])
  }
  subject_values <- data[[roles$id]]
  for (id in unique(subject_values[observation_rows])) {
    subject_rows <- which(!is.na(subject_values) & subject_values == id)
    events <- !(as.character(data[[roles$evid]][subject_rows]) %in%
                  c("0", "0.0"))
    if (!is.null(roles$amt)) {
      events <- events & as.numeric(data[[roles$amt]][subject_rows]) > 0
    }
    positions <- which(subject_values[observation_rows] == id)
    event_rows <- subject_rows[events]
    if (length(event_rows) && !is.null(roles$occasion)) {
      event_occasion <- suppressWarnings(as.integer(
        data[[roles$occasion]][event_rows]
      ))
      for (position in positions) {
        candidates <- event_rows[event_occasion == occasion[position]]
        if (length(candidates) && !is.finite(tad[position])) {
          origin <- min(as.numeric(data[[roles$time]][candidates]))
          tad[position] <-
            as.numeric(data[[roles$time]][observation_rows[position]]) - origin
        }
      }
    } else if (length(event_rows)) {
      dose_times <- sort(unique(as.numeric(data[[roles$time]][event_rows])))
      occasion[positions] <- pmax(1L, findInterval(
        as.numeric(data[[roles$time]][observation_rows[positions]]),
        dose_times
      ))
      occasion[positions] <- pmin(occasion[positions], length(dose_times))
      tad[positions] <-
        as.numeric(data[[roles$time]][observation_rows[positions]]) -
        dose_times[occasion[positions]]
    }
  }
  plotted_time <- if (identical(clock, "tad")) tad else
    as.numeric(data[[roles$time]][observation_rows])
  data.frame(
    dataset = factor(dataset, levels = c("Source", "Synthetic")),
    subject = as.character(data[[roles$id]][observation_rows]),
    time = plotted_time,
    dv = as.numeric(data[[roles$dv]][observation_rows]),
    occasion = occasion,
    endpoint = if (is.null(roles$dvid)) "DV" else
      as.character(data[[roles$dvid]][observation_rows]),
    stringsAsFactors = FALSE
  )
}

# Every dataset below gets the same figure: observation rows only, source beside
# synthetic on a shared y axis, one row per endpoint. `clock = "tad"` plots time
# after dose instead of study time, which is the readable view once a study
# doses more than once.
overlay_plot <- function(source, synthetic, roles, title,
                         clock = "study_time",
                         x_label = "Study time (hours)", y_label = "DV",
                         log_y = FALSE, alpha = 0.3) {
  plotted <- rbind(
    observed_plot_data(source, roles, "Source", clock),
    observed_plot_data(synthetic, roles, "Synthetic", clock)
  )
  # One line per patient on a study-time axis, and one line per occasion on a
  # dose-relative one, where the profiles are meant to lie on top of each other.
  plotted$series <- if (identical(clock, "tad")) {
    interaction(plotted$dataset, plotted$subject, plotted$occasion)
  } else {
    interaction(plotted$dataset, plotted$subject)
  }
  figure <- ggplot2::ggplot(
    plotted,
    ggplot2::aes(time, dv, group = series, colour = dataset)
  ) +
    ggplot2::geom_line(alpha = alpha) +
    ggplot2::geom_point(alpha = alpha, size = 0.7) +
    # Endpoint down the side, source and synthetic across. `facet_grid()` frees
    # a scale per ROW, so this orientation is the one that gives each endpoint
    # its own y while holding source and synthetic on a shared one -- which is
    # the only arrangement the eye can compare.
    (if (length(unique(plotted$endpoint)) > 1L) {
      ggplot2::facet_grid(endpoint ~ dataset, scales = "free_y",
                          switch = "y")
    } else {
      ggplot2::facet_wrap(~dataset)
    }) +
    ggplot2::scale_colour_manual(values = comparison_colours) +
    ggplot2::labs(x = x_label, y = y_label, colour = "Dataset", title = title) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "none",
                   strip.placement = "outside")
  if (isTRUE(log_y)) figure + ggplot2::scale_y_log10() else figure
}

# Facet rows are endpoints now, so a five-endpoint study needs five times the
# height a one-endpoint study does. Chunks pass this to `fig.height` rather than
# every figure inheriting the one-endpoint default and arriving unreadable.
overlay_height <- function(data, roles, per_endpoint = 2.2, minimum = 3.4) {
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  endpoints <- if (is.null(roles$dvid)) 1L else {
    length(unique(data[[roles$dvid]][observed]))
  }
  max(minimum, per_endpoint * endpoints)
}

# `synpmx_pca_summarize()` refuses a dataset it cannot model rather than
# guessing, so every run here is an attempt whose outcome is part of the
# result. The refusal message is kept and printed; it names the gate and what
# to do about it, and this vignette does not restate either.
pca_run <- function(data, roles, seed) {
  summary <- tryCatch(
    synpmx_pca_summarize(data, roles, seed = seed),
    error = function(e) structure(conditionMessage(e), class = "pca_refusal")
  )
  if (inherits(summary, "pca_refusal")) {
    return(list(ok = FALSE, refusal = as.character(summary)))
  }
  list(ok = TRUE, summary = summary,
       synthetic = synpmx_pca_generate(summary, seed = seed))
}

# Chunk options are evaluated even for chunks that will not run, so a guard has
# to hold when the data package is missing and the run object was never
# created.
ran <- function(name) {
  environment <- knitr::knit_global()
  exists(name, envir = environment) &&
    isTRUE(get(name, envir = environment)$ok)
}

# Both height helpers take the data and roles lazily, so a chunk option
# referring to an object the missing data package never created is never
# forced.
pca_height <- function(name, data, roles, default = 4) {
  if (ran(name)) overlay_height(data, roles) else default
}

dist_height <- function(name, data, roles, default = 4) {
  if (ran(name)) compare_pmx_distributions_height(data, roles) else default
}

# What the fit read out of the source, as a table rather than as the printed
# object, so it sits beside the figures at the same weight. `pca_report()` owns
# the row labels and the explanation next to each one.
release_table <- function(run, label) {
  knitr::kable(
    as.data.frame(pca_report(run$summary)),
    row.names = FALSE, align = c("l", "l", "r", "r"),
    caption = paste0(
      "What ", label, "'s fit read out of the source data. `min_patients` is ",
      "the smallest number of patients standing behind any one number in ",
      "that row."
    )
  )
}

# A refused dataset keeps its section and its steps; the steps that cannot run
# say so rather than being left out.
refusal_note <- function(run) {
  cat("synpmx_pca_summarize() refused this dataset:\n\n  ",
      run$refusal, "\n\n",
      "Steps 3 to 6 have no synthetic dataset to run on: no generation, no ",
      "figure, no distributions, no scorecard.\n", sep = "")
}
```

</details>

## The datasets

Every dataset here is public package data from `nlmixr2data` or `xgxr`, both
`Suggests` of this package. The table is the one in
`vignette("avatar-public-data-examples")`, unchanged.

<!-- Pandoc sizes these columns in proportion to the dashes in the row below,
     so the last column's width is set by the dashes under it rather than by
     its contents. Keep them uneven. -->

| Dataset | Package | Patients | Rows | Endpoints | What it is here to exercise |
|----------|-----------|-----|-----|-----|--------------------------------------------------------|
| `case1_pkpd` | xgxr | 180 | 20820 | 2 | Six arms, a declared `NOMTIME`, PK and PD, baseline weight, `CENS`. The shape a study report usually arrives in |
| `mad` | xgxr | 60 | 4160 | 5 | Multiple ascending dose carrying ordinal, count and binary PD alongside continuous PD and PK |
| `theo_md` | nlmixr2data | 12 | 348 | 1 | Seven Q24H oral doses with dense profiles on occasions 1 and 7 only. Dosed by weight. |
| `warfarin` | nlmixr2data | 32 | 515 | 2 | One dose, PK and PD running on different time courses, categorical covariates |
| `wbcSim` | nlmixr2data | 45 | 280 | 1 | Infusion start/stop pairs and a delayed WBC response. Follow-up time varies. |
| `nimoData` | nlmixr2data | 12 | 441 | 1 | Ten roughly weekly infusions, recorded at the times they happened |
| `mavoglurant` | nlmixr2data | 120 | 2678 | 1 | One- and two-period crossover, `TIME` resetting within `OCC`, an occasion-varying assigned dose, infusion rows. |
| `pheno_sd` | nlmixr2data | 59 | 744 | 1 | Neonatal phenobarbital, individualised dosing, sparse irregular sampling, time-varying weight. |

## What PCA requires that AVATAR does not

`synpmx_avatar()` runs on all eight. `synpmx_pca_summarize()` has three gates,
and a dataset that fails any one of them is refused rather than modelled.

| Gate | Requirement | Why it exists |
|---|---|---|
| Nominal grid | `nominal_time` declared in `pmx_roles()`, present on every dose and observation row | Every feature the method fits is a cell on that grid. AVATAR derives a grid when none is declared, as a masking step; here the grid is the model's axis, and deriving it would be a statement about the protocol that only the caller can make. |
| Cohort size | 10 subjects | Below it there is no spread for a basis to describe. |
| Arm size | 3 patients in every arm named by `strata` | An arm of one or two has a mean score vector that is that patient and a covariance that is noise around them. |

The gates are what makes the survey short. `vignette("pca-algorithm")` states
each one where it is enforced.

## Shared workflow

Every example follows the same six steps. Steps 1, 4, 5 and 6 are the AVATAR
survey's steps 1, 3, 4 and 5; steps 2 and 3 are its step 2, split, because the
model this generator fits is a thing to look at before generating from it.

1. declare column meanings with `pmx_roles()`, copied from the AVATAR survey;
2. summarize with `synpmx_pca_summarize()`, and inventory the fit with
   `pca_report()`;
3. generate with `synpmx_pca_generate()`;
4. plot the real and synthetic data;
5. plot the observation (DV) and baseline covariate distributions, source
   against synthetic;
6. report the score card for the synthetic dataset.

A refused dataset keeps steps 1 and 2 and says at step 2 that steps 3 to 6 have
nothing to run on.

The scorecard marks each check `pass`, `FAIL`, or `review`. A `review` row is
one where the pharmacometrician could evaluate whether the output is acceptable
for use. If the synthetic dataset will not be crossing a trust boundary, items
marked `review` are acceptable.

## case1_pkpd: a declared nominal grid

**Start here.** 180 patients across six treatment arms, with a declared
`NOMTIME`, two endpoints keyed by a character `NAME` column, a baseline weight,
and a `CENS` column. `vignette("pca-demo")` runs this generator end to end.

The nominal grid this dataset declares is the one thing PCA cannot do without,
so this is the dataset the method is easiest on and the one to read the other
seven against.

```{r case1, eval=has_xgxr, message=FALSE}
case1_pkpd <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
case1_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB",
  keep = "STUDY"
)
case1_pca <- pca_run(case1_pkpd, case1_roles, seed = 808)
```

```{r case1-report, eval=has_xgxr, echo=FALSE}
if (ran("case1_pca")) release_table(case1_pca, "case1_pkpd") else
  refusal_note(case1_pca)
```

```{r case1-plot, echo=FALSE, eval=ran("case1_pca") && has_ggplot2, fig.height=pca_height("case1_pca", case1_pkpd, case1_roles)}
overlay_plot(case1_pkpd, case1_pca$synthetic, case1_roles,
             "case1_pkpd: PK concentration and continuous PD, six arms",
             y_label = "LIDV", alpha = 0.2)
```

```{r case1-distributions, eval=ran("case1_pca") && has_distributions, fig.height=dist_height("case1_pca", case1_pkpd, case1_roles)}
compare_pmx_distributions(case1_pkpd, case1_pca$synthetic, case1_roles)
```

```{r case1-scorecard, eval=ran("case1_pca")}
synpmx_scorecard_datatable(
  synpmx_scorecard(case1_pkpd, case1_pca$synthetic, case1_roles)
)
```

Read the timing rows against the AVATAR card for the same dataset. AVATAR
coarsens recorded times onto a grid and puts scatter back; this generator places
every value on the nominal grid itself, so the synthetic times are exactly
on-protocol and the source times are not. That is a difference in what the two
methods claim about time, and it shows up on the card rather than in the
figures.

## mad: five endpoints, three of them discrete

Multiple ascending dose, 60 patients, five endpoints on one time axis: PK, a
continuous PD, and ordinal, count and binary PD alongside them. `NOMTIME` is
declared, so the gate that stops the six datasets below does not apply here.

The question this dataset asks of PCA is what a basis fitted on a common scale
does with endpoints that are not continuous. `pmx_endpoint_types()` classifies
them and `synpmx_pca_summarize()` records the classification; the scorecard's
endpoint rows are where a discrete endpoint that came back continuous would
show.

```{r mad, eval=has_xgxr, message=FALSE}
mad <- as.data.frame(get(utils::data(list = "mad", package = "xgxr")))
mad_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", mdv = "MDV", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"),
  covariates = c("WEIGHTB", "SEX")
)
mad_pca <- pca_run(mad, mad_roles, seed = 909)
```

```{r mad-report, eval=has_xgxr, echo=FALSE}
if (ran("mad_pca")) release_table(mad_pca, "mad") else refusal_note(mad_pca)
```

```{r mad-plot, echo=FALSE, eval=ran("mad_pca") && has_ggplot2, fig.height=pca_height("mad_pca", mad, mad_roles)}
overlay_plot(mad, mad_pca$synthetic, mad_roles,
             "mad: five endpoints on one time axis",
             y_label = "LIDV", alpha = 0.2) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 7))
```

```{r mad-distributions, eval=ran("mad_pca") && has_distributions, fig.height=dist_height("mad_pca", mad, mad_roles)}
compare_pmx_distributions(mad, mad_pca$synthetic, mad_roles)
```

```{r mad-scorecard, eval=ran("mad_pca")}
synpmx_scorecard_datatable(
  synpmx_scorecard(mad, mad_pca$synthetic, mad_roles)
)
```

## theo_md: no nominal grid

Twelve patients, seven Q24H oral doses, dense profiles on the first and last
occasions and a single trough between. Dosed by weight, so no two patients share
a dose amount.

```{r theo, eval=has_nlmixr2data, message=FALSE}
data("theo_md", package = "nlmixr2data")
theo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", covariates = "WT"
)
theo_pca <- pca_run(theo_md, theo_roles, seed = 303)
```

```{r theo-report, eval=has_nlmixr2data, echo=FALSE}
if (ran("theo_pca")) release_table(theo_pca, "theo_md") else
  refusal_note(theo_pca)
```

```{r theo-plot, echo=FALSE, eval=ran("theo_pca") && has_ggplot2}
overlay_plot(theo_md, theo_pca$synthetic, theo_roles,
             "theo_md: dose-relative concentration profiles",
             clock = "tad", x_label = "Time after dose (hours)", alpha = 0.25)
```

```{r theo-distributions, eval=ran("theo_pca") && has_distributions, fig.height=dist_height("theo_pca", theo_md, theo_roles)}
compare_pmx_distributions(theo_md, theo_pca$synthetic, theo_roles)
```

```{r theo-scorecard, eval=ran("theo_pca")}
synpmx_scorecard_datatable(
  synpmx_scorecard(theo_md, theo_pca$synthetic, theo_roles)
)
```

The dense-then-sparse sampling this dataset is here to exercise is a protocol
statement, and the columns do not carry it. A `NOMTIME` column reconstructed
from the recorded times would be this vignette deciding what the protocol was.

## warfarin: two endpoints on different time courses

Thirty-two patients, one dose, a PK and a PD endpoint running on different time
courses, and categorical covariates. The dose is 1.5 mg/kg, declared through
`dose_covariate` rather than left to be inferred.

```{r warfarin, eval=has_nlmixr2data, message=FALSE}
data("warfarin", package = "nlmixr2data")
warfarin_roles <- pmx_roles(
  id = "id", time = "time", dv = "dv", amt = "amt", evid = "evid",
  dvid = "dvid", covariates = c("wt", "age", "sex"),
  dose_covariate = "wt"   # 1.5 mg/kg; say so rather than let it be inferred
)
warfarin_pca <- pca_run(warfarin, warfarin_roles, seed = 404)
```

```{r warfarin-report, eval=has_nlmixr2data, echo=FALSE}
if (ran("warfarin_pca")) release_table(warfarin_pca, "warfarin") else
  refusal_note(warfarin_pca)
```

```{r warfarin-plot, echo=FALSE, eval=ran("warfarin_pca") && has_ggplot2, fig.height=pca_height("warfarin_pca", warfarin, warfarin_roles)}
overlay_plot(warfarin, warfarin_pca$synthetic, warfarin_roles,
             "warfarin: PK (cp) and PD (pca) endpoints")
```

```{r warfarin-distributions, eval=ran("warfarin_pca") && has_distributions, fig.height=dist_height("warfarin_pca", warfarin, warfarin_roles)}
compare_pmx_distributions(warfarin, warfarin_pca$synthetic, warfarin_roles)
```

```{r warfarin-scorecard, eval=ran("warfarin_pca")}
synpmx_scorecard_datatable(
  synpmx_scorecard(warfarin, warfarin_pca$synthetic, warfarin_roles)
)
```

## wbcSim: infusion pairs and a delayed response

Forty-five patients, infusion start and stop pairs, a white-blood-cell count
that declines to a nadir and recovers, and a follow-up time that varies by
patient.

```{r wbc, eval=has_nlmixr2data, message=FALSE}
data("wbcSim", package = "nlmixr2data")
wbc_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID", cmt = "CMT"
)
wbc_pca <- pca_run(wbcSim, wbc_roles, seed = 505)
```

```{r wbc-report, eval=has_nlmixr2data, echo=FALSE}
if (ran("wbc_pca")) release_table(wbc_pca, "wbcSim") else refusal_note(wbc_pca)
```

```{r wbc-plot, echo=FALSE, eval=ran("wbc_pca") && has_ggplot2}
overlay_plot(wbcSim, wbc_pca$synthetic, wbc_roles,
             "wbcSim: delayed decline, nadir, and recovery",
             y_label = "WBC")
```

```{r wbc-distributions, eval=ran("wbc_pca") && has_distributions, fig.height=dist_height("wbc_pca", wbcSim, wbc_roles)}
compare_pmx_distributions(wbcSim, wbc_pca$synthetic, wbc_roles)
```

```{r wbc-scorecard, eval=ran("wbc_pca")}
synpmx_scorecard_datatable(
  synpmx_scorecard(wbcSim, wbc_pca$synthetic, wbc_roles)
)
```

## nimoData: infusions recorded as actuals

Twelve patients, ten roughly weekly infusions recorded at the times they
happened rather than at the times they were planned. This is the dataset the
AVATAR survey uses to show what a derived grid costs when there is nothing to
derive it from.

```{r nimo, eval=has_nlmixr2data, message=FALSE}
data("nimoData", package = "nlmixr2data")
nimo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
  covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
)
nimo_pca <- pca_run(nimoData, nimo_roles, seed = 606)
```

```{r nimo-report, eval=has_nlmixr2data, echo=FALSE}
if (ran("nimo_pca")) release_table(nimo_pca, "nimoData") else
  refusal_note(nimo_pca)
```

```{r nimo-plot, echo=FALSE, eval=ran("nimo_pca") && has_ggplot2}
overlay_plot(nimoData, nimo_pca$synthetic, nimo_roles,
             "nimoData: ten infusions over a long follow-up", alpha = 0.2)
```

```{r nimo-distributions, eval=ran("nimo_pca") && has_distributions, fig.height=dist_height("nimo_pca", nimoData, nimo_roles)}
compare_pmx_distributions(nimoData, nimo_pca$synthetic, nimo_roles)
```

```{r nimo-scorecard, eval=ran("nimo_pca")}
synpmx_scorecard_datatable(
  synpmx_scorecard(nimoData, nimo_pca$synthetic, nimo_roles)
)
```

`TAD` and `OCC` are declared here and describe the dosing structure, but neither
is a nominal grid: they say where a sample sat relative to a dose that happened,
not where the protocol planned to put it.

## mavoglurant: a crossover with time resetting

One hundred and twenty patients, one- and two-period crossover, `TIME` resetting
within `OCC`, an occasion-varying assigned dose, and infusion rows. The largest
cohort in the survey.

```{r mavo, eval=has_nlmixr2data, message=FALSE}
data("mavoglurant", package = "nlmixr2data")
mavo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  cmt = "CMT", rate = "RATE", mdv = "MDV", occasion = "OCC",
  keep = "DOSE", covariates = c("AGE", "SEX", "WT", "HT")
)
mavo_pca <- pca_run(mavoglurant, mavo_roles, seed = 707)
```

```{r mavo-report, eval=has_nlmixr2data, echo=FALSE}
if (ran("mavo_pca")) release_table(mavo_pca, "mavoglurant") else
  refusal_note(mavo_pca)
```

```{r mavo-plot, echo=FALSE, eval=ran("mavo_pca") && has_ggplot2}
overlay_plot(mavoglurant, mavo_pca$synthetic, mavo_roles,
             "mavoglurant: dose-relative profiles with occasion reset",
             clock = "tad", x_label = "Time after dose (hours)", alpha = 0.2)
```

```{r mavo-distributions, eval=ran("mavo_pca") && has_distributions, fig.height=dist_height("mavo_pca", mavoglurant, mavo_roles)}
compare_pmx_distributions(mavoglurant, mavo_pca$synthetic, mavo_roles)
```

```{r mavo-scorecard, eval=ran("mavo_pca")}
synpmx_scorecard_datatable(
  synpmx_scorecard(mavoglurant, mavo_pca$synthetic, mavo_roles)
)
```

## pheno_sd: sparse, irregular, individualised

Fifty-nine neonates on phenobarbital, dosed individually, sampled sparsely and
irregularly, with a weight that changes over the study.

```{r pheno, eval=has_nlmixr2data, message=FALSE}
data("pheno_sd", package = "nlmixr2data")
pheno_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  covariates = c("WT", "APGR")
)
pheno_pca <- pca_run(pheno_sd, pheno_roles, seed = 1010)
```

```{r pheno-report, eval=has_nlmixr2data, echo=FALSE}
if (ran("pheno_pca")) release_table(pheno_pca, "pheno_sd") else
  refusal_note(pheno_pca)
```

```{r pheno-plot, echo=FALSE, eval=ran("pheno_pca") && has_ggplot2}
overlay_plot(pheno_sd, pheno_pca$synthetic, pheno_roles,
             "pheno_sd: sparse, irregular, individualised",
             y_label = "Phenobarbital (ug/mL)")
```

```{r pheno-distributions, eval=ran("pheno_pca") && has_distributions, fig.height=dist_height("pheno_pca", pheno_sd, pheno_roles)}
compare_pmx_distributions(pheno_sd, pheno_pca$synthetic, pheno_roles)
```

```{r pheno-scorecard, eval=ran("pheno_pca")}
synpmx_scorecard_datatable(
  synpmx_scorecard(pheno_sd, pheno_pca$synthetic, pheno_roles)
)
```

Sampling this sparse and this irregular is the case a nominal grid would matter
most for, because there is little else to align the patients on.

## What ran, and what did not

```{r survey-outcome, echo=FALSE}
survey_runs <- c(
  case1_pkpd = "case1_pca", mad = "mad_pca", theo_md = "theo_pca",
  warfarin = "warfarin_pca", wbcSim = "wbc_pca", nimoData = "nimo_pca",
  mavoglurant = "mavo_pca", pheno_sd = "pheno_pca"
)
outcome <- data.frame(
  Dataset = names(survey_runs),
  Outcome = unname(vapply(survey_runs, function(name) {
    if (!exists(name, envir = knitr::knit_global())) "not attempted"
    else if (ran(name)) "modelled" else "refused"
  }, character(1))),
  Gate = unname(vapply(survey_runs, function(name) {
    if (!exists(name, envir = knitr::knit_global())) {
      return("source package not installed")
    }
    run <- get(name, envir = knitr::knit_global())
    if (isTRUE(run$ok)) return("")
    # The refusal names its gate in the first sentence; the rest is the
    # remedy, which belongs in the section rather than in a summary row.
    sub("\\..*$", ".", run$refusal)
  }, character(1))),
  stringsAsFactors = FALSE
)
knitr::kable(outcome, row.names = FALSE,
             caption = "Every dataset in the survey, and what synpmx_pca() did with it.")
```

The datasets this generator refuses are refused for one reason far more often
than the other two, and it is the reason `vignette("pca-algorithm")`, Step 2
gives for requiring the grid rather than deriving one.

**What to do with a refused dataset.** Three options, in the order to try them.
Add the protocol's planned times as a column and declare them as
`nominal_time`, which is the only one that makes PCA work on the dataset rather
than around it. Failing that, use `synpmx_avatar()`, which derives a grid as a
masking step and runs on all eight. The third option, reconstructing a nominal
grid by rounding the recorded times, is a statement about the protocol, and the
package refuses to make it on the caller's behalf for the same reason this
vignette does not make it here.

## Where to go next

- `vignette("pca-algorithm")` for what the generator does, step by step.
- `vignette("pca-demo")` for one dataset end to end.
- `vignette("pca-trial-summary")` for what a fitted model contains.
- `vignette("avatar-public-data-examples")` for the same eight datasets under
  `synpmx_avatar()`.
- `vignette("avatar-scorecard")` for what each scorecard row asks.
