---
title: "The synpmx PMX Model Algorithm"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The synpmx PMX Model Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
library(synpmx)
stored <- system.file("extdata", "warfarin-model-fit.rds", package = "synpmx")
fit <- if (nzchar(stored)) readRDS(stored) else NULL
has_fit <- !is.null(fit)
```

This article is the full specification of `synpmx_model()`, which fits a small
set of linear pharmacokinetic (PK) models to a study and generates new subjects
by simulating from the fit. Use of the algorithm is demonstrated in
`vignette("pmxmodel-demo")`.

**The fitted parameters are not estimates to report.** They exist to make
simulated profiles look like the source study. The candidate set is five linear
models and the covariate model is allometric scaling or nothing, which is too
little to answer a scientific question. This restriction is the same one the
rest of the package carries and it binds harder here, because the output of the
fit looks exactly like the output of a real population analysis. The fitted
object prints the warning with itself for that reason.

No formal privacy guarantee is offered. No patient's measured value reaches the
output — the claim `synpmx_pca()` makes, and stronger than `synpmx_avatar()`'s —
but the fixed effects and the covariance matrix are functions of the individuals
in the source, and neither is noised.

## Where this sits among the three generators

`synpmx_avatar()` blends real neighbouring patients' values. `synpmx_pca()`
reduces each subject to principal-component scores and models the scores. This
one estimates a structural model and simulates from it. None of the three is
retired and none is a replacement for another; they differ in what they carry
out of the source and in what they can therefore reproduce.

The one thing this generator does that neither of the others does is connect the
dosing to the values. In `synpmx_pca()` the schedule and the profile are drawn
from separate models, so a synthetic subject's dose reduction appears in the
dosing records and not in the concentrations. Here the schedule is drawn first
and the concentration is computed from it, so a subject who steps down a level
has a lower exposure from that cycle on. That is the largest fidelity gain on
offer, and Step 6 is where it happens.

## Three exported functions

| Function | Reads patient data | Needs `nlmixr2` | Returns |
|---|---|---|---|
| `synpmx_model_estimate(data, roles, ...)` | yes | yes | `pmx_fitted_model` |
| `synpmx_model_generate(fitted_model, n_subjects, seed)` | no | no | data frame |
| `synpmx_model(data, roles, n_subjects, seed, ...)` | yes | yes | data frame |

`model_report()` inventories everything the fitted object carries,
`model_candidates()` returns the comparison table the selection was made from,
and `model_parameters()` returns the fixed effects, the between-subject
covariance matrix and the residual error.

`nlmixr2` is in `Suggests` and is loaded only by `synpmx_model_estimate()`.
Generation runs on base R from a stored fit, which is why the second stage needs
neither the package nor a compiler.

## `nominal_time` is required, and there are two time axes

Two different questions need two different clocks, and this generator is the
only one in the package that reads both.

**Estimation reads recorded times.** A population PK fit is a statement about
time after the dose that was actually given, so the subject-level dosing records
go to the fitter as recorded, reductions and skipped cycles included. Fitting
against the planned schedule where a patient's dose was reduced would push the
drop in concentration that followed into clearance, and the model would report a
population that eliminates the drug faster than the real one.

**Everything else reads the nominal grid.** The dosing model, the visit model,
and every signal in Steps 1 and 2 sit on the protocol's planned times. On
recorded times a "median profile" is one point per time, because no two patients
share a clock reading, and a study sampled at six times looks like a study
sampled at two hundred. A shape, a peak position and a count of sampling times
are all statements about the protocol.

Inferring that grid is a statement about the protocol only the caller can make,
so `nominal_time` is required rather than derived. `synpmx_pca()` refuses on the
same grounds.

## Step 1: Which endpoint is the drug concentration

`pmx_endpoint_types()` already infers continuous, integer, binary and ordinal
from the values, and that inference is reused unchanged. What this generator
adds is a second classification on top of it. Four signals, computed for every
endpoint that is a time course:

1. **Compartment.** The `cmt` role puts the endpoint in the compartment the
   doses go to, or one above a dosing compartment nobody observes — the
   depot-and-central convention.
2. **Post-dose only.** Observations before the first dose are absent, or sit at
   or below the censoring limit, in most subjects.
3. **Shape.** The median profile within the richest dose interval rises to a
   maximum and declines without rising again.
4. **Dose proportionality.** Between the highest and lowest dose level, the
   ratio of median subject maxima is within a factor of two of the dose ratio.

**Signals 2 and 4 are required; 1 and 3 break ties.** That split is a statement
about what each signal is worth. Being absent before the first dose and scaling
with the dose are properties only a drug concentration has. A compartment number
is a convention the dataset's author chose, and a rise-and-fall shape is one many
biomarkers also have.

Signal 4 is not computable on a single-level study, and it is also not computable
on one dosed by body weight. A study that gives every patient a slightly
different amount has no dose levels: reading each amount as its own level
compares a median of one patient against a median of one patient over a dose
ratio near 1, which almost any endpoint passes — including the biomarker the
signal exists to reject. Where the distinct amounts outnumber half the cohort the
signal reports "not computable" and signals 1 to 3 decide.

Where no endpoint passes, or two pass and neither tie-break separates them, the
function errors and names `endpoint_roles` as the way through. That is the
inference-versus-declaration fork, answered the way `dose_covariate` answers it:
infer where the data settles the question, offer the declaration as an override.

```{r signals, eval=has_fit}
model_report(fit)$endpoints$signals
```

**What this cannot do.** On a study with one continuous endpoint and one dose
level, the classification rests on signal 2 alone, and signal 2 passes for any
endpoint with no pre-dose observation. `nlmixr2data::wbcSim` is such a study: its
white-cell count is classified as the drug concentration, because nothing in the
data says otherwise. The remedy is `endpoint_roles`, and the evidence is in the
table above — a row reading "not computable" under `proportional` is the
classification telling you how little it had to go on.

## Step 2: What design produced it

Detection prunes the candidate set; the fit picks among what survives. Neither
property below decides the model on its own.

### Route

A `rate` role carrying a nonzero value is an infusion, and `1cmt_infusion` is
the only candidate. Otherwise there is one property that separates the two
remaining routes, and it is not how many patients peak early.

A drug given by mouth cannot be in the blood at the moment it is swallowed. So a
concentration observed at time zero after a dose is intravenous, and a median
profile that rises before it falls is oral. A profile that declines from a first
sample drawn some time after the dose is both of those at once — an intravenous
bolus, or an oral dose whose absorption finished before anybody looked — and no
amount of reading the source separates them. There both routes are offered and
AIC settles it.

Counting subjects instead is the mistake this replaces, and `warfarin` is why.
Twenty-two of its thirty-two patients are first sampled at 24 hours, well past
the peak, so every one of their profiles declines from its first point. A
per-subject vote reads 69% intravenous from a study that is oral: it counts the
sampling schedule rather than the drug. The share of subjects whose own profile
rises is still reported, because it says how much of the cohort the median
profile speaks for, but it decides nothing.

```{r route, eval=has_fit}
model_report(fit)$design$reason
```

### Sampling richness

**Across the cohort**, distinct nominal times after a dose holding an
observation. Below six, no model is offered and the function errors, pointing at
`synpmx_avatar()` and `synpmx_pca()`, which need no identifiable structure.

**Per subject**, distinct nominal times after a dose within the richest dose
interval. Four or more, with two after the median peak, is enough to identify a
distribution phase. A sample *before* the peak is required only where there is
an ascending limb to sample — an intravenous bolus peaks at the dose, and asking
for a sample before it would refuse every such study.

This count decides nothing. It is reported, and `model_report()` says so when
the sampling would support a two-compartment model, because asking for one is
your move rather than the search's.

### One compartment, and one fit

The default is one-compartment: `1cmt_iv`, `1cmt_oral` or `1cmt_infusion`. Where
the route is ambiguous both are fitted and AIC chooses, which is the only case
the default runs more than one fit.

A distribution phase is a refinement of a shape the one-compartment model
already has, and this generator exists to make simulated profiles resemble the
source study rather than to characterise it. The cost is not small. On a
thirty-subject oral study a two-compartment fit takes 49 seconds against 12 for
a one-compartment one, and on data a one-compartment model describes it spends
that time against a flat likelihood and reports the worse AIC anyway.

Better starting values shorten it but do not rescue it — the same fit from
non-compartmental starts ran 49 seconds against 60 from cruder ones, which is a
17% saving on a fit that is five times too slow.

**It remains available.** `pk = "2cmt_oral"` or `pk = "2cmt_iv"` forces the
model and skips the search, and the generator simulates from it exactly as it
does from any other — the closed-form solution is already there.

## Step 3: Estimate the candidates

The candidate set is at most the five models in the package's closed-form set,
and never more. A candidate the fitter could estimate and the generator could not
simulate would be a model that fits and then generates nothing, so the two lists
are one list. That is also why no dose effect on clearance is fitted: it would
mean a candidate set that is no longer five closed-form models, and generation
would need a solver.

Each candidate carries log-normal between-subject variability on every
structural parameter and a proportional residual error. An endpoint holding
values at or below zero after censoring is handled gets an additive error
instead, and the substitution is recorded on the fit.

**Censored observations are imputed, and the fit is told how many.** A value
below the assay limit is replaced by a uniform draw inside the censoring region
— a draw rather than a fixed LLOQ/2, which would swap one artificial spike for
another — and the boundary is put back when the synthetic data is emitted. This
is deliberate rather than a shortcut past the M3 likelihood: the same imputation
is what lets the visit model, the PD shapes and the covariate model read a
latent value instead of a stack of identical boundary substitutions, and it is
the same thing `synpmx_pca()` does.

It is an assumption all the same, and its weight is the share of the endpoint
that carries it, so that share is measured and reported with the fit. On
`xgxr::case1_pkpd`, 46% of concentrations sit below the limit — 95% of them in
the lowest dose arm — and there the fitted parameters are substantially a
statement about the draw rather than about measurements. Read that line in
`model_report()` before trusting the concentrations of a heavily censored study.

Starting values come from a non-compartmental reading of the source rather than
from a guess. A population fit started far from the answer either converges
slowly or reports the starting values back.

The terminal slope is fitted by log-linear regression over the points after the
peak, and everything leans on it. Clearance is the dose over the area
**extrapolated to infinity**, not over the trapezoid alone, which on a study
sampled to four half-lives understates the area by about a tenth and overstates
clearance by the same. Volume is clearance over the terminal slope, which is the
quantity the terminal phase identifies — dose over the peak is not a volume of
any kind for an oral dose. Absorption is solved from where the peak falls: for a
one-compartment oral model the peak sits at `log(ka/ke)/(ka - ke)`, one equation
in one unknown once the slope has given `ke`.

Where a two-compartment model is asked for, the steady-state volume comes from
the mean residence time, with the absorption mean taken back off for an oral
dose, and is split between the central and peripheral compartments.

**The default fits one model and stops.** Not a search that happens to have one
candidate: one model, chosen by the route detection above, with allometric
scaling folded into it where a weight-like covariate is declared. A whole
`synpmx_model_estimate()` call on a thirty-subject study takes about eleven
seconds, and every part of that is the one fit.

`pk` is how you buy accuracy with time. Naming one model forces it; naming
several searches over them and picks on AIC.

```r
# One fit, the default.
synpmx_model_estimate(data, roles)
# One fit, your choice of model.
synpmx_model_estimate(data, roles, pk = "2cmt_oral")
# Two fits, compared on AIC.
synpmx_model_estimate(data, roles, pk = c("1cmt_oral", "2cmt_oral"))
```

**How long it takes is set by the dose records, not the patients.** Every
likelihood evaluation superposes one contribution per dose per subject, so the
cost scales with the number of dose events rather than with cohort size. A
single-dose study of thirty-two patients fits in about eleven seconds. A
twelve-week daily regimen carries eighty-five dose records per subject, and a
hundred and fifty such patients is two orders of magnitude more work: the same
call on `xgxr::case1_pkpd` had not finished after twenty-five minutes. Nothing
is wrong when that happens — it is what population fitting costs on that design
— but it is worth knowing before you start rather than after.

**Where there is a search, selection is on AIC**, and the estimation method is
`focei` for that reason.
SAEM's log-likelihood is a Gaussian-quadrature step run after the fit, and at
phase 1 cohort sizes it returns a non-finite value: `theo_sd` fits perfectly well
under SAEM — clearance 2.75, volume 32.3, absorption 1.51, the textbook values —
and reports `AIC = Inf`, so a search over two candidates has nothing to compare.
`estimation = "saem"` remains available for a study large enough to give it a
likelihood. A candidate whose AIC is not finite is recorded as not converged
whichever method produced it.

A candidate that fails drops out carrying its reason and stays in the table, so
that a search which came down to one survivor does not look like a search that
had one candidate. Where nothing converges the function errors rather than
returning the least bad fit.

```{r candidates, eval=has_fit}
model_candidates(fit)
```

### Covariates

The default `"auto"` applies allometric scaling on clearance and volume where a
weight-like covariate is declared, and fits nothing else. The exponents are the
standard 0.75 and 1 rather than estimated ones, and the effect is **asserted
rather than tested**: it is folded into the one fit, not compared against a
model without it. Testing it would double the cost of the whole call, and
allometry on a body weight is a shape this generator assumes rather than a
hypothesis it examines. `covariate_effects = "none"` switches it off.

The covariate is recognised by name — `wt`, `weight`, `bw` and the like — and
must be numeric and positive. There is no way to recognise a body weight from
its values alone, and guessing from a distribution would be worse than not
guessing.

**The cost of this default is explicit.** A covariate that influences the real
profiles and is not in the model is generated independently of them, so the
synthetic data carries no relationship between the two. `synpmx_avatar()`
preserves those relationships without modelling them, because a blended
subject's covariates and profile come from the same donors. `model_report()`
reports the correlation between each declared covariate and the individual
random effects, which is where an unmodelled relationship shows up.

### Pharmacodynamics

Every remaining continuous endpoint is fitted against a constant, a linear and
an exponential time course and selected on AIC, with between-subject variability
on the baseline.

All three are fitted by least squares rather than through `nlmixr2`, so all
three together cost no measurable time — they are three-parameter curves on one
endpoint, the fit is well conditioned, and routing them through a population
fitter would put a compiler in the path of every PD endpoint for no gain in what
the generator then draws. **The one compiled fit in the whole call is the PK
one.**

**These shapes have no exposure dependence.** A PD endpoint driven by
concentration is reproduced as a time course that happens to resemble the average
subject's response. A dataset whose point is the exposure-response relationship
is not served by this generator.

Binary and ordinal endpoints are not fitted at all. They are drawn from the level
frequencies their arm holds at each nominal time, which is what the visit model
already does for attendance.

## Step 4: The dosing model and the visit model

Everything that is not the concentration-time curve comes from the models
`synpmx_pca_summarize()` already builds, unchanged in what they represent. Dose
reductions, interrupted cycles, discontinuation and missed visits are modelled
here exactly as they are there, and so are arm sizes, the covariate
distributions, the censoring boundary and the schema. The two generators call the
same code rather than a copy of it; what differs is only the adapter that hands
it the grid, and this generator writes its own over the nominal times the source
holds.

Per arm:

| Model | What it holds | Drawn at generation as |
|---|---|---|
| Planned schedule | The nominal dose times enough of the arm reached, and the modal amount at each cycle among patients still on their starting dose | The cycle grid every subject starts from |
| Dose ladder | The levels patients dropped to, as ratios to their own starting dose, built from within-patient decreases | The amount multiplier in force at a cycle |
| Reduction rate | Discrete-time hazard of stepping down a level | Decided before the cycle is dosed |
| Interruption rate | Discrete-time hazard of skipping a cycle without ending treatment | Decided at the cycle |
| Discontinuation rate | Discrete-time hazard of stopping treatment | Decided after the cycle is dosed |
| Visit model | Per endpoint and per retained nominal time, the fraction of the arm holding an observation there | Attendance, drawn per visit |

A study where nobody reduces, skips or stops early has all three rates at zero
and one level, and the model then reproduces the planned schedule exactly. No
detector decides which kind of study this is, because the rates already say it.

A grid cell is kept only where at least `min_arm_patients` distinct patients hold
an observation there. A nominal time one patient attended is that patient, and
generating from it would put them back.

## Step 5: What the fitted model carries

Estimated by `nlmixr2`: the selected structural model, the candidate comparison
table, fixed effects, the between-subject covariance matrix, the residual error,
the covariate effects that survived, and the PD shape and parameters per
endpoint.

Not estimated: the per-arm dosing model and its three rates, the visit model, arm
sizes, the covariate distributions, the censoring boundary, the schema and the
roles.

**No individual estimates.** Empirical Bayes estimates are per-subject
quantities, and a fitted model that carried them would be writing out a
description of each real patient. They are not stored on the object. Generation
draws random effects from the covariance matrix instead, and the correlation
report above computes from them and keeps only the correlation.

```{r report, eval=has_fit}
model_report(fit)
```

## Step 6: Generate

Per synthetic subject: assign an arm keeping the source arm shares, draw
covariates from the arm's covariate model, draw random effects from the
covariance matrix, apply the covariate effects to the typical parameters, draw
the dose schedule from the arm's dosing model, draw the visits attended from the
visit model, evaluate the profile at the attended times against the drawn
schedule, add residual error, apply the censoring boundary and emit.

**The order of the two draws is fixed.** The schedule is drawn first and the
profile is computed from it, so a synthetic subject who steps down a dose level
has a lower exposure from that cycle on, and one who skips a cycle has the
trough that implies. Drawing a profile and then a schedule would reproduce
`synpmx_pca()`'s disconnect with extra steps.

A concentration is floored at zero. A proportional error on a value near the
assay limit produces a negative one often enough to matter — on `warfarin` it
puts the minimum at -1.4 against a source minimum of 0 — and the value is floored
rather than redrawn, because redrawing until the value is positive is a truncated
distribution nobody declared.

## Gates

| Gate | Threshold | Reason |
|---|---|---|
| Cohort size | 20 subjects | A covariance matrix fitted to a handful of subjects describes those subjects. `synpmx_pca()`'s floor is 10 and this one is higher because a parameter estimate concentrates on its cohort faster than a score does. |
| Cohort time coverage | 6 distinct nominal times after a dose | Below it no linear model is identifiable and the fit reports the starting values. |
| Arm size | 3 patients | Inherited from the dosing and visit models, which are summaries of an arm. An arm of one or two has no rates to pool. |
| `nominal_time` undeclared | — | The grid is a statement about the protocol only the caller can make. |
| No grid cell shared | `min_arm_patients` | A nominal time one patient attended is that patient. |
| No PK endpoint identified | — | Errors and names `endpoint_roles`. |
| No candidate converged | — | Errors rather than returning the least bad fit. |
| A fitted model as a public input | — | `pmx_prior()`, `synpmx_prior()` and `synpmx_calibrated()` refuse a `pmx_fitted_model`. Its parameters were estimated from the confidential study, so accepting one as a public structural input would charge no privacy budget for information taken from the data. |

That last gate is the one whose absence would be a privacy defect rather than a
quality defect. What no check reaches is a caller reading numbers off a fitted
model and typing them into `pmx_structural_model(typical = )` by hand.

## What this generator does not do

- **No exposure-response.** The PD shapes are time courses with no concentration
  term.
- **No dose-dependent PK.** Dose and arm are apparatus, so a nonlinearity in the
  source is generated away rather than reported.
- **No occasion-varying parameters.** Occasions are pooled and the `occasion`
  role is not read. A dataset whose clock resets within occasion has times that
  are already dose-relative, and putting it on one cumulative axis before calling
  is the caller's work.
- **No goodness-of-fit verdict.** The generator is usable only where the selected
  model describes the source, and nothing here scores that. `model_report()`
  shows what the search had to go on; it does not tell you the answer was right.

## See also

- `vignette("pmxmodel-demo")` — the same algorithm run on a study end to end.
- `vignette("pca-algorithm")` — the generator this one shares its dosing and
  visit models with.
- `vignette("avatar-algorithm")` — the blending generator, and the six masking
  mechanisms it uses.
- `vignette("avatar-scorecard")` — the checks that score any generated dataset
  against its source.
