---
title: "Demo: Using synpmx_model"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx_model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_dplyr <- requireNamespace("dplyr", quietly = TRUE)
has_dt <- requireNamespace("DT", quietly = TRUE)
stored <- system.file("extdata", "mad-model-fit.rds", package = "synpmx")
fit <- if (nzchar(stored)) readRDS(stored) else NULL
run <- has_xgxr && has_dplyr && !is.null(fit)
plots <- run && has_ggplot2
distributions <- plots && requireNamespace("patchwork", quietly = TRUE)
tables <- run && has_dt
# Three significant digits on every measured quantity. Times, counts and
# identifiers are exempt: `signif(2015.9, 3)` is 2020, which would move a
# nominal visit onto a time the study never had.
exact <- c("time", "ntime", "nomtime", "nominal_time", "cycle", "id", "index",
           "n", "observations", "censored", "patients", "min_patients",
           "numbers", "planned_cycles", "distinct", "source_doses", "offset",
           "count")
digits3 <- function(x) {
  x <- as.data.frame(x)
  for (column in names(x)) {
    if (tolower(column) %in% exact || !is.numeric(x[[column]])) next
    x[[column]] <- signif(x[[column]], 3)
  }
  x
}
show <- function(x, caption = NULL) {
  if (!has_dt) return(digits3(x))
  DT::datatable(
    digits3(x), rownames = FALSE, caption = caption,
    options = list(dom = "t", paging = FALSE, scrollX = TRUE, ordering = TRUE)
  )
}
```

Fit a population model to a study, look at what the fit carries, and generate a
synthetic dataset by simulating from it. Generation uses fitted population
parameters and study summaries: structural curves, between-subject variability,
residual error, endpoint frequencies, a planned dose schedule per arm, and dose
changes and one attendance rate per endpoint, pooled over the arms.
It does not read the original patient rows.

It makes no formal privacy claim, and **it is not for estimation** — the fitted
parameters exist to make simulated profiles look like the source study, and the
object prints that warning with itself. The full specification is in
`vignette("pmxmodel-algorithm")`.

The dataset is `xgxr::mad`: 60 patients in six treatment arms, with a
pharmacokinetic (PK) concentration endpoint and continuous, count, ordinal and
binary pharmacodynamic (PD) endpoints.

## Configuration

To run this on your own study, edit the two chunks below that:

1. Read in the dataset.
2. Define the column roles.

The rest of the code can be kept as is.

```{r data, eval=run, message = FALSE}
library(dplyr)
raw <- as.data.frame(get(utils::data(list = "mad", package = "xgxr")))
SEED <- 808
```

The column meanings.  Columns not specified here are dropped from the synthetic dataset.

```{r roles, eval=run}
roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  mdv          = "MDV",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",        # which endpoint each observation row is
  nominal_time = "NOMTIME",     # the grid the visit model is built on
  strata       = c("TRTACT", "DOSE"), # treatment arms, fitted one at a time
  covariates   = c("WEIGHTB", "SEX") # drawn independently of the profiles
)
```

## Fitting

`synpmx_model_estimate()` is the only stage that reads patient data, and the only
one that needs `nlmixr2`. It works out which endpoint is the drug, what design
produced it, and tries two compartments with checked fallback to one.

```{r estimate-shown, eval=FALSE}
fit <- synpmx_model_estimate(raw, roles, seed = 1)
```

```{r estimate-note, echo=FALSE, results="asis", eval=run}
cat("The chunk above is shown rather than run. Fitting compiles a model, so this",
    "document reads a stored `synpmx_model_estimate()` fit and",
    "`R CMD check` never needs a compiler.\n")
```

The fitted clearance is
`r if (run) signif(fit$parameters$fixed[["cl"]], 3) else NA` L/h and the volume
`r if (run) signif(fit$parameters$fixed[["v"]], 3) else NA` L. They put the
simulated profiles where the source's are, and are not estimates to report.

## What the fit carries

`model_report()` prints everything the generator will simulate from, and the
diagnostics beside it. `vignette("pmxmodel-fingerprint")` goes through each
part, and `vignette("pmxmodel-algorithm")` says how each is estimated.

```{r report, eval=run}
model_report(fit)
```

```{r candidates, eval=run}
show(model_candidates(fit), "Models attempted and acceptance results")
```

## Generating

`synpmx_model_generate()` reads no patient data. Its arguments are the fit and a
subject count, so everything about the source that reaches the output has already
passed through the fit.

```{r generate, eval=run}
synthetic <- synpmx_model_generate(fit, seed = SEED)
```

The generated table follows the declared schema: retained columns and classes,
the same compartment numbers, and new identifiers.

```{r head, eval=run}
show(head(synthetic, 10), "The first ten rows")
```

## Plot synthetic data and original data

Compare each endpoint on the recorded time scale, with source and synthetic
subjects shown separately within each treatment arm.

```{r overlay-setup, eval=plots}
library(ggplot2)
library(xgxr)
xgx_theme_set()
comparison_colours <- c(source = "#1B6CA8", synthetic = "#D95F02")

both <- rbind(
  cbind(raw[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$EVID == 0 & !is.na(both$LIDV), ]
obs$TRTACT <- factor(obs$TRTACT,
                     levels = c("Placebo", "100 mg", "200 mg", "400 mg",
                                "800 mg", "1600 mg"))
```

```{r overlay-pk, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PK Concentration" & obs$TIME <= 24, ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA~TRTACT) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 24, by = 6)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "PK concentration", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("Day 1 Conc. Profile")
```

```{r overlay-pk-last, eval=plots, fig.height=5}
last_dose <- obs[obs$NAME == "PK Concentration" &
                   obs$TIME >= 120 & obs$TIME <= 144, ]
last_dose$TAD <- last_dose$TIME - 120

ggplot(last_dose, aes(TAD, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA~TRTACT) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 24, by = 6)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time after dose (hours)", y = "PK concentration", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("Last Dose (120 h) Conc. Profile")
```

Every synthetic concentration profile uses the selected structural curve with
new parameter draws and residual error. Compare the rise, decline and spread
within each arm: pooled generation checks can miss discrepancies confined to
one dose group or part of the sampling window.

```{r overlay-pd, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PD - Continuous", ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.35) +
  xgx_scale_x_time_units(units_dataset = "hours", units_plot = "days") +
  facet_grid(DATA~TRTACT) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (days)", y = "PD (continuous)", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("PD Response")
```

The PD time course has no exposure term, so a synthetic patient's arm does
not reach their response. The `mad` section of the
[public-data evaluation](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html#mad-five-endpoints-only-one-of-them-a-concentration)
compares the late response by arm.

## Distributions of Synthetic and Original Data

One panel per endpoint and per baseline covariate, source against synthetic.
`compare_pmx_distributions_height()` sizes the figure, since it grows a row of
panels at a time; `output = "tables"` gives the numbers behind it.

```{r distributions, eval=distributions, fig.height=compare_pmx_distributions_height(raw, roles)}
compare_pmx_distributions(raw, synthetic, roles)
```

## Scoring it

`synpmx_scorecard()` does not know what produced the dataset it scores, so it
reads this output the same way it reads an AVATAR or PCA one.

```{r scorecard, eval=run}
card <- synpmx_scorecard(raw, synthetic, roles)
synpmx_scorecard_datatable(card)
```

Checks requiring an AVATAR run record are not applicable to this generator.
`vignette("scorecard")` documents what each row asks and what its pass
criterion is.

## One call

`synpmx_model()` is the two stages together, for when you do not need to look at
the fit first. The result carries the fit's release as an attribute, which is
what generation read; the candidate table and the other diagnostics are only on
the fit itself.

```{r one-call, eval=FALSE}
synthetic <- synpmx_model(raw, roles, seed = SEED)
attr(synthetic, "pmx_fitted_model")
```

## See also

- `vignette("pmxmodel-algorithm")` — what each step does and why.
- `vignette("pmxmodel-fingerprint")` — every quantity the fit carries out of the
  study, in detail.
- [Evaluating the model generator on public data](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html)
  — the same generator over the public studies the other two surveys use.
- `vignette("pca-demo")` — a worked study through the principal-component
  generator.
- `vignette("avatar-demo")` — a worked study through blending.
- `vignette("scorecard")` — the checks above, in detail.
- [Privacy protections in the PMX model generator](https://iamstein.github.io/synpmx/articles/pmxmodel-privacy.html)
  — what leaves the study, and the checks on it.
