---
title: "Demo: Using synpmx_model"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx_model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_dplyr <- requireNamespace("dplyr", quietly = TRUE)
has_dt <- requireNamespace("DT", quietly = TRUE)
stored <- system.file("extdata", "mad-model-fit.rds", package = "synpmx")
fit <- if (nzchar(stored)) readRDS(stored) else NULL
run <- has_xgxr && has_dplyr && !is.null(fit)
plots <- run && has_ggplot2
distributions <- plots && requireNamespace("patchwork", quietly = TRUE)
tables <- run && has_dt
# Three significant digits on every measured quantity. Times, counts and
# identifiers are exempt: `signif(2015.9, 3)` is 2020, which would move a
# nominal visit onto a time the study never had.
exact <- c("time", "ntime", "nomtime", "nominal_time", "cycle", "id", "index",
           "n", "observations", "censored", "patients", "min_patients",
           "numbers", "planned_cycles", "distinct", "source_doses", "offset",
           "count")
digits3 <- function(x) {
  x <- as.data.frame(x)
  for (column in names(x)) {
    if (tolower(column) %in% exact || !is.numeric(x[[column]])) next
    x[[column]] <- signif(x[[column]], 3)
  }
  x
}
show <- function(x, caption = NULL) {
  if (!has_dt) return(digits3(x))
  DT::datatable(
    digits3(x), rownames = FALSE, caption = caption,
    options = list(dom = "t", paging = FALSE, scrollX = TRUE, ordering = TRUE)
  )
}
```

Fit a population model to a study, look at what the fit carries, and generate a
synthetic dataset by simulating from it. Generation uses fitted population
parameters and study summaries: structural curves, between-subject variability,
residual error, endpoint frequencies, and dosing and visit models per arm.
It does not read the original patient rows.

It makes no formal privacy claim, and **it is not for estimation** — the fitted
parameters exist to make simulated profiles look like the source study, and the
object prints that warning with itself. The full specification is in
`vignette("pmxmodel-algorithm")`.

The dataset is `xgxr::mad`: 60 patients in six treatment arms, with a
pharmacokinetic (PK) concentration endpoint and continuous, count, ordinal and
binary pharmacodynamic (PD) endpoints. The two-compartment fit passed the
convergence and generation checks. The public-data survey also shows a fit
accepted with a false-convergence warning.

## Configuration

To run this on your own study, edit the two chunks below that:

1. Read in the dataset.
2. Define the column roles.

The rest of the code can be kept as is.

```{r data, eval=run, message = FALSE}
library(dplyr)
raw <- as.data.frame(get(utils::data(list = "mad", package = "xgxr")))
SEED <- 808
```

The column meanings.  Columns not specified here are dropped from the synthetic dataset.

```{r roles, eval=run}
roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  mdv          = "MDV",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",        # which endpoint each observation row is
  nominal_time = "NOMTIME",     # the grid the visit model is built on
  strata       = c("TRTACT", "DOSE"), # treatment arms, fitted one at a time
  covariates   = c("WEIGHTB", "SEX") # drawn independently of the profiles
)
```

## Fitting

`synpmx_model_estimate()` is the only stage that reads patient data, and the only
one that needs `nlmixr2`. It works out which endpoint is the drug, what design
produced it, and tries two compartments with checked fallback to one.

```{r estimate-shown, eval=FALSE}
fit <- synpmx_model_estimate(raw, roles, seed = 1)
```

```{r estimate-note, echo=FALSE, results="asis", eval=run}
cat("The chunk above is shown rather than run. Fitting compiles a model, so this",
    "document reads a stored `synpmx_model_estimate()` fit and",
    "`R CMD check` never needs a compiler.\n")
```

The fit report gives the elapsed time and the number of subjects used. Each
likelihood evaluation reads the dosing history, so repeated-dose studies cost
more than single-dose studies. `max_fit_subjects` caps the population fit;
the dosing, visit and covariate models still read every subject.

`pk = "1cmt_oral"` requests one compartment directly. The requested fit must
pass the same convergence, parameter and generation checks, and is never
replaced by another model automatically.

A clearance of `r if (run) signif(fit$parameters$fixed[["cl"]], 3) else NA` L/h
and a volume of `r if (run) signif(fit$parameters$fixed[["v"]], 3) else NA` L.
Whether those are the right numbers for this compound is not the question the
generator asks: they exist to put the simulated profiles where the source's
are, and the object prints that warning with itself.

## What the fit carries

Two halves. `model_report()` separates them, because they answer to different
things — one half is an estimate with all an estimate's caveats, the other is a
summary of the study's apparatus. Everything the generator simulates from is in
here, which is why printing the fitted object shows the same account.

```{r report, eval=run}
model_report(fit)
```

The concentration has a proportional residual error. The continuous and count
PD endpoints have fitted time courses with no exposure term. Binary and ordinal
endpoints are drawn from arm-by-visit frequencies rather than from the
compartment model. The signals table explains why `PK Concentration` was
identified as the concentration endpoint.

The correlation block is the other thing to read. A covariate that moves with a
random effect and is not in the model above is generated independently of the
profiles, so the synthetic data carries no relationship between them.
`synpmx_avatar()` keeps those relationships without modelling them.

## The candidates

```{r candidates, eval=run}
show(model_candidates(fit), "Models attempted and acceptance results")
```

The table contains only models actually attempted. An accepted two-compartment
fit needs no one-compartment comparison. A rejected fit retains its convergence
status and rejection reason, and the fallback has its own row.
`model_report()` explains the selection. A vector of model names in `pk`
requests an Akaike information criterion (AIC) comparison of those models after
acceptance checks.

A study containing only pharmacodynamic observations can instead declare
`endpoint_roles = c(pd = "response")`, using its endpoint name. That skips
compartment fitting and uses the existing PD time courses. A positive baseline
can also identify the study as PD-only automatically, as the public-data
survey demonstrates on `wbcSim`.

## Generating

`synpmx_model_generate()` reads no patient data. Its arguments are the fit and a
subject count, so everything about the source that reaches the output has already
passed through the fit.

```{r generate, eval=run}
synthetic <- synpmx_model_generate(fit, seed = SEED)
```

The generated table follows the declared schema: retained columns and classes,
the same compartment numbers, and new identifiers.

```{r head, eval=run}
show(head(synthetic, 10), "The first ten rows")
```

## Plot synthetic data and original data

Compare each endpoint on the recorded time scale, with source and synthetic
subjects shown separately within each treatment arm.

```{r overlay-setup, eval=plots}
library(ggplot2)
library(xgxr)
xgx_theme_set()
comparison_colours <- c(source = "#1B6CA8", synthetic = "#D95F02")

both <- rbind(
  cbind(raw[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$EVID == 0 & !is.na(both$LIDV), ]
obs$TRTACT <- factor(obs$TRTACT,
                     levels = c("Placebo", "100 mg", "200 mg", "400 mg",
                                "800 mg", "1600 mg"))
```

```{r overlay-pk, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PK Concentration" & obs$TIME <= 24, ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA~TRTACT) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 24, by = 6)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "PK concentration", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("Day 1 Conc. Profile")
```

```{r overlay-pk-last, eval=plots, fig.height=5}
last_dose <- obs[obs$NAME == "PK Concentration" &
                   obs$TIME >= 120 & obs$TIME <= 144, ]
last_dose$TAD <- last_dose$TIME - 120

ggplot(last_dose, aes(TAD, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA~TRTACT) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 24, by = 6)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time after dose (hours)", y = "PK concentration", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("Last Dose (120 h) Conc. Profile")
```

Every synthetic concentration profile uses the selected structural curve with
new parameter draws and residual error. Compare the rise, decline and spread
within each arm: pooled generation checks can miss discrepancies confined to
one dose group or part of the sampling window.

```{r overlay-pd, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PD - Continuous", ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.35) +
  xgx_scale_x_time_units(units_dataset = "hours", units_plot = "days") +
  facet_grid(DATA~TRTACT) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (days)", y = "PD (continuous)", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("PD Response")
```

The continuous PD endpoint has an exponential time course. Its parameters are
fitted to the pooled study, and subjects differ in their baseline draws and
residual error.

**What is missing is the dose ordering.** One shape is fitted to the pooled
observations and each subject gets a draw on its baseline, so the arm a subject
was assigned to does not reach its response.

```{r pd-by-arm, eval=run}
late_pd <- function(data, label) {
  rows <- data$NAME == "PD - Continuous" & data$EVID == 0 &
    data$TIME > 144 & !is.na(data$LIDV)
  out <- aggregate(list(mean_pd = data$LIDV[rows]),
                   list(arm = data$TRTACT[rows]), function(x) round(mean(x)))
  stats::setNames(out, c("arm", paste0("mean_pd_", label)))
}
pd_table <- merge(late_pd(raw, "source"), late_pd(synthetic, "synthetic"),
                  by = "arm")
show(pd_table, "Mean PD response after 144 h, by arm")
```

This table compares late responses by arm. The continuous PD model carries no
exposure term, so any source relationship between dose and response is not
preserved by construction. A dataset whose purpose is exposure-response needs
that limitation considered separately from numerical fit acceptance.

## Distributions of Synthetic and Original Data

One panel per endpoint and per baseline covariate, source against synthetic.
`compare_pmx_distributions_height()` sizes the figure, since it grows a row of
panels at a time; `output = "tables"` gives the numbers behind it.

```{r distributions, eval=distributions, fig.height=compare_pmx_distributions_height(raw, roles)}
compare_pmx_distributions(raw, synthetic, roles)
```

## Scoring it

`synpmx_scorecard()` does not know what produced the dataset it scores, so it
reads this output the same way it reads an AVATAR or PCA one.

```{r scorecard, eval=run}
card <- synpmx_scorecard(raw, synthetic, roles)
synpmx_scorecard_datatable(card)
```

Read the fidelity checks alongside the plots. A passing acceptance screen
excludes gross fitting and generation failures; it does not guarantee that the
simulated spread, attendance or endpoint relationships match the source.
Checks requiring an AVATAR run record are not applicable to this generator.

`vignette("scorecard")` documents what each row asks and what its pass
criterion is.

## One call

`synpmx_model()` is the two stages together, for when you do not need to look at
the fit first. It is on the result either way, as an attribute.

```{r one-call, eval=FALSE}
synthetic <- synpmx_model(raw, roles, seed = SEED)
attr(synthetic, "pmx_fitted_model")
```

## See also

- `vignette("pmxmodel-algorithm")` — what each step does and why.
- `vignette("pmxmodel-fingerprint")` — every quantity the fit carries out of the
  study, in detail.
- [Evaluating the model generator on public data](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html)
  — the same generator over the public studies the other two surveys use.
- `vignette("pca-demo")` — a worked study through the principal-component
  generator.
- `vignette("avatar-demo")` — a worked study through blending.
- `vignette("scorecard")` — the checks above, in detail.
