---
title: "Demo: Using synpmx_model"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx_model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_dt <- requireNamespace("DT", quietly = TRUE)
stored <- system.file("extdata", "warfarin-model-fit.rds", package = "synpmx")
fit <- if (nzchar(stored)) readRDS(stored) else NULL
run <- has_data && !is.null(fit)
plots <- run && has_ggplot2 && has_xgxr
distributions <- plots && requireNamespace("patchwork", quietly = TRUE)
# Three significant digits on every measured quantity. Times, counts and
# identifiers are exempt: `signif(2015.9, 3)` is 2020, which would move a
# nominal visit onto a time the study never had.
exact <- c("time", "ntime", "nomtime", "nominal_time", "cycle", "id", "index",
           "n", "observations", "imputed", "patients", "min_patients",
           "numbers", "planned_cycles", "distinct", "source_doses", "offset",
           "count")
digits3 <- function(x) {
  x <- as.data.frame(x)
  for (column in names(x)) {
    if (tolower(column) %in% exact || !is.numeric(x[[column]])) next
    x[[column]] <- signif(x[[column]], 3)
  }
  x
}
show <- function(x, caption = NULL) {
  if (!has_dt) return(digits3(x))
  DT::datatable(
    digits3(x), rownames = FALSE, caption = caption,
    options = list(dom = "t", paging = FALSE, scrollX = TRUE, ordering = TRUE)
  )
}
```

Fit a population model to a study, look at what the fit carries, and generate a
synthetic dataset by simulating from it. No number a patient measured reaches
the output: what leaves the source is a structural model, a handful of fixed
effects, a covariance matrix, a residual error, and a dosing and visit model per
arm.

It makes no formal privacy claim, and **it is not for estimation** — the fitted
parameters exist to make simulated profiles look like the source study, and the
object prints that warning with itself. The full specification is in
`vignette("pmxmodel-algorithm")`.

## The study

`nlmixr2data::warfarin` is thirty-two patients given a single oral dose, with a
concentration endpoint `cp` and a pharmacodynamic one `pca`.

```{r data, eval=run}
data("warfarin", package = "nlmixr2data")
warfarin <- as.data.frame(warfarin)
str(warfarin)
```

Its sixteen recorded observation times are the protocol's, shared across
patients, so declaring `nominal_time` from `time` asserts something true about
this study rather than constructing a grid. A study whose recorded times are
per-patient clock readings needs a grid built by hand first; the public-data
survey for `synpmx_pca()` works several of those.

```{r roles, eval=run}
sort(unique(warfarin$time[warfarin$evid == 0]))

warfarin$ntime <- warfarin$time
warfarin_roles <- pmx_roles(
  id = "id", time = "time", nominal_time = "ntime", dv = "dv", amt = "amt",
  evid = "evid", dvid = "dvid", covariates = c("wt", "age", "sex")
)
```

## Fitting

`synpmx_model_estimate()` is the only stage that reads patient data, and the only
one that needs `nlmixr2`. It works out which endpoint is the drug, what design
produced it, fits the candidates that design admits and picks one on AIC.

```{r estimate-shown, eval=FALSE}
fit <- synpmx_model_estimate(warfarin, warfarin_roles, seed = 1)
```

That is one population fit: a one-compartment oral model with allometric scaling
on `wt`, chosen by the route detection rather than by a search. Ask for more
where you want it — `pk = "2cmt_oral"`, or `pk = c("1cmt_oral", "2cmt_oral")` to
compare the two on AIC.

```{r estimate-note, echo=FALSE, results="asis", eval=run}
cat("The chunk above is shown rather than run. Fitting compiles a model, so this",
    "document reads a stored fit built by `scripts/build-model-fits.R` and",
    "`R CMD check` never needs a compiler.\n")
```

```{r fit-print, eval=run}
fit
```

Clearance of `r if (run) round(fit$parameters$fixed[["cl"]], 3) else NA` L/h and a
volume of `r if (run) round(fit$parameters$fixed[["v"]], 1) else NA` L are the
values this dataset is known for, which is the point: the fit is good enough to
simulate from. It is still not a result to report, for the reason the object
prints.

## What the fit carries

Two halves. `model_report()` separates them, because they answer to different
things — one half is an estimate with all an estimate's caveats, the other is a
summary of the study's apparatus.

```{r report, eval=run}
model_report(fit)
```

The signals table says how the concentration endpoint was decided. Here `cp` is
the only endpoint with a rise-and-fall shape, and `proportional` reads "not
computable" because `warfarin` gives every patient the same dose — there are no
dose levels to compare. That is the classification telling you how much it had to
go on, and it is worth reading before trusting the answer.

The correlation block is the other thing to read. A covariate that moves with a
random effect and is not in the model above is generated independently of the
profiles, so the synthetic data carries no relationship between them.
`synpmx_avatar()` keeps those relationships without modelling them.

## The candidates

```{r candidates, eval=run}
show(model_candidates(fit), "Candidates the design admitted")
```

One row, because the default fits one model and stops. A distribution phase is a
refinement of a shape the one-compartment model already has and costs about five
times as long to fit, so it is available rather than routine: `pk = "2cmt_oral"`
forces it, `pk = c("1cmt_oral", "2cmt_oral")` fits both and picks on AIC, and
`model_report()` says when the sampling would support one.

The whole call takes about eleven seconds on a study this size, and effectively
all of it is this one fit — the PD shapes are least squares and cost nothing
measurable.

## Generating

`synpmx_model_generate()` reads no patient data. Its arguments are the fit and a
subject count, so everything about the source that reaches the output has already
passed through the fit.

```{r generate, eval=run}
synthetic <- synpmx_model_generate(fit, n_subjects = 32, seed = 7)
str(synthetic)
```

The generated table is in the source's shape: same columns, same classes, same
compartment numbers, new identifiers.

```{r head, eval=run}
show(head(synthetic, 10), "The first ten rows")
```

## Plot synthetic data and original data

The same overlay the other two demos draw, so the three can be read against each
other: source and synthetic on the same axes, one row each.

```{r overlay-setup, eval=plots}
library(ggplot2)
library(xgxr)
xgx_theme_set()
comparison_colours <- c(source = "#1B6CA8", synthetic = "#D95F02")

both <- rbind(
  cbind(warfarin[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$evid == 0 & !is.na(both$dv), ]
```

```{r overlay-pk, eval=plots, fig.height=5}
ggplot(obs[obs$dvid == "cp", ],
       aes(time, dv, group = id, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA ~ .) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 120, by = 24)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "Warfarin concentration", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("Concentration Profile")
```

The synthetic profiles are smoother than the source's, and that is the
generator's shape rather than a fault in the run: every one of them is the same
one-compartment curve evaluated at a different draw of clearance, volume and
absorption, with residual error on top. A real cohort's profiles wander in ways
one structural model does not reproduce.

```{r overlay-pd, eval=plots, fig.height=5}
ggplot(obs[obs$dvid == "pca", ],
       aes(time, dv, group = id, colour = DATA)) +
  geom_line(alpha = 0.35) +
  facet_grid(DATA ~ .) +
  xgx_scale_x_time_units("hours", breaks = seq(0, 120, by = 24)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "Prothrombin complex activity",
       colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("PD Response")
```

The PD endpoint is a straight line per subject, which is the whole of what the
shape search offers: a constant, a line, or an exponential, with variability on
the baseline and no exposure term. The source's response falls and then flattens;
the synthetic one keeps falling. A dataset whose point is that curvature is not
served by this generator.

## Distributions of Synthetic and Original Data

One panel per endpoint and per baseline covariate, source against synthetic.
`compare_pmx_distributions_height()` sizes the figure, since it grows a row of
panels at a time; `output = "tables"` gives the numbers behind it.

```{r distributions, eval=distributions, fig.height=compare_pmx_distributions_height(warfarin, warfarin_roles)}
compare_pmx_distributions(warfarin, synthetic, warfarin_roles)
```

## Scoring it

`synpmx_scorecard()` does not know what produced the dataset it scores, so it
reads this output the same way it reads an AVATAR or PCA one.

```{r scorecard, eval=run}
card <- synpmx_scorecard(warfarin, synthetic, warfarin_roles)
show(as.data.frame(card), "Scorecard")
```

The three rows that read `unavailable` need `synpmx_avatar()`'s own run record,
which no other generator writes. `vignette("avatar-scorecard")` documents what
each row asks and what its pass criterion is.

**B4a fails, and it is the row rather than the generator that is at fault.** It
asks whether a generated subject's set of observation times is one that fewer
than two real patients held. Two of the thirty-two are.

That sounds like a disclosure and is not one. `warfarin` has 32 patients holding
15 distinct visit sets, and **14 of those 15 are held by exactly one patient** —
so "matches a set only one patient held" is very nearly "matches any real
patient at all", which every generator that reproduces the study's design will
do. Over 200 seeds this generator reproduces 1.26 such sets on average and fires
on 83% of them; the two seen here are what chance predicts, not an excess.

The same holds wherever you look. Across four public studies the share of source
visit sets held by exactly one patient is 93% for `warfarin`, 83% for `wbcSim`,
and 100% for both `theo_sd` and `theo_md`. A threshold that selects almost the
whole source is measuring how wide the study's visit grid is, not what the
generator did with it.

There is a sharper demonstration. The generator writes its observations onto the
nominal grid, so a synthetic subject's times are always drawn from those sixteen
slots. `warfarin`'s recorded times happen to *equal* its nominal grid, which is
what makes a collision possible at all. Move the source's clock a few minutes
off that grid — changing nothing about the synthetic data, and nothing about
privacy — and the row reads 0 instead of 2. B4a is answering a question about
how this study recorded its visit times.

Nor would redrawing help. Attendance here is drawn independently per visit, so a
generated set was never taken from the patient it happens to match — rejecting
collisions would truncate the attendance distribution and drive the row to zero
by construction, which fixes the number rather than any risk.

**B4b is the row that answers the disclosure question, and it reads 0.** No value
any patient measured is reproduced. That is the claim this generator makes, and
it is the one worth checking.

## One call

`synpmx_model()` is the two stages together, for when you do not need to look at
the fit first. It is on the result either way, as an attribute.

```{r one-call, eval=FALSE}
synthetic <- synpmx_model(warfarin, warfarin_roles, n_subjects = 32, seed = 7)
attr(synthetic, "pmx_fitted_model")
```

## See also

- `vignette("pmxmodel-algorithm")` — what each step does and why.
- `vignette("pca-demo")` — the same study through the principal-component
  generator.
- `vignette("avatar-demo")` — and through blending.
- `vignette("avatar-scorecard")` — the checks above, in detail.
