---
title: "The PMX model study fingerprint"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The PMX model study fingerprint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_dt <- requireNamespace("DT", quietly = TRUE)
stored <- system.file("extdata", "warfarin-model-fit.rds", package = "synpmx")
fit <- if (nzchar(stored)) readRDS(stored) else NULL
run <- has_data && !is.null(fit)
tables <- run && has_dt

# Three significant digits on measured quantities. Times and counts are exempt:
# rounding a nominal visit time would move it onto a time the study never had.
exact <- c("time", "ntime", "nomtime", "nominal_time", "cycle", "id", "index",
           "n", "observations", "censored", "patients", "min_patients",
           "numbers", "planned_cycles", "distinct", "source_doses", "offset",
           "count")
digits3 <- function(x) {
  for (column in names(x)) {
    if (column %in% exact || !is.numeric(x[[column]])) next
    x[[column]] <- signif(x[[column]], 3)
  }
  x
}
show <- function(x, caption = NULL, paged = FALSE) {
  if (!has_dt) return(digits3(as.data.frame(x)))
  DT::datatable(
    digits3(as.data.frame(x)), rownames = FALSE, caption = caption,
    options = list(dom = if (paged) "tip" else "t", paging = paged,
                   pageLength = 10, scrollX = TRUE, ordering = TRUE)
  )
}
```

A study's **fingerprint** is the set of numbers a synthetic study is generated
from. For `synpmx_model()` it is a population PK model, a time course for each
PD endpoint, and a summary of the study's design: its arms, dose schedules,
attendance and covariate distributions. Synthetic data is drawn from the
fingerprint and random numbers alone, so the fingerprint is everything about
the source that can reach the output.

`synpmx_model_estimate()` returns a fit. The fit holds the fingerprint plus
diagnostics for whoever ran it, such as the candidate models and the
correlations between covariates and random effects. `model_release()` extracts
the fingerprint, and that is the object to carry out of the environment that
holds the study. `synpmx_model_generate()` accepts either and reads only the
fingerprint.

This vignette walks through the fingerprint of one study. Read
`vignette("pmxmodel-demo")` first for a run end to end, and
`vignette("pmxmodel-algorithm")` for how each quantity is estimated.

## The study

`nlmixr2data::warfarin`: 32 patients, one oral dose scaled to body weight, a
warfarin concentration and a prothrombin complex activity (PCA).

```{r data, eval=run}
data("warfarin", package = "nlmixr2data")
warfarin <- as.data.frame(warfarin)
warfarin$ntime <- warfarin$time
roles <- pmx_roles(
  id = "id", time = "time", nominal_time = "ntime", dv = "dv", amt = "amt",
  evid = "evid", dvid = "dvid", covariates = c("wt", "age", "sex")
)
```

```{r estimate-shown, eval=FALSE}
fit <- synpmx_model_estimate(warfarin, roles, seed = 1)
```

This document reads a stored copy of that fit, so it needs no compiler.

```{r release, eval=run}
release <- model_release(fit)
names(release)
setdiff(names(fit), names(release))  # diagnostics that stay with the fit
```

`model_report()` summarizes the whole fit in one printout:

```{r report, eval=run}
model_report(fit)
```

## The PK model

### Which endpoint is the concentration

The fit decides which endpoint is the drug concentration from four signals:
whether it is observed in a dosing compartment, whether it appears only after
dosing, whether its median profile rises and falls, and whether it scales with
dose. This is a diagnostic rather than part of the fingerprint, but everything
in the PK model depends on it.

```{r signals, eval=run}
fit$endpoints$signals
fit$design$reason
```

`proportional` is `NA` because `warfarin` is dosed by body weight, so there are
no dose levels to compare and the other three signals decide.
`endpoint_roles` in `synpmx_model_estimate()` overrides the decision.

### Structural model

```{r structural, eval=run}
release$structural
model_candidates(fit)
```

By default the fit tries two compartments and falls back to one only if two
fail the acceptance checks. `pk` names a model, or several to compare by AIC.

### Parameters

Typical values on the natural scale, the between-subject covariance on the log
scale, and the residual error. Every PK and PD estimate in the fingerprint is
rounded to one significant figure.

```{r parameters, eval=run}
model_parameters(fit)
```

No covariate is in the model: `covariate_effects = "none"` is the default, so
`release$covariate_effects` is empty and every synthetic patient's parameters
are drawn from the covariance matrix alone.

```{r covariate-effects, eval=run}
release$covariate_effects
```

The fingerprint holds no patient's own random effects (empirical Bayes
estimates). They are read during estimation and dropped.

`warfarin` reports four concentrations at or below zero, which is what an assay
returns near its limit. They are fitted at half the smallest positive value, and
the residual error stays proportional.

### Censoring

```{r censoring, eval=run}
fit$censoring
```

`warfarin` declares no assay limit, so this is empty. Where a study declares
one, its below-limit concentrations enter the population fit as censored, and
this table gives the share of each endpoint below the limit. On
`xgxr::case1_pkpd` it is 46% of the concentrations, so much of what the PK
model knows there is a bound rather than a measurement.

## The PD time courses

Each continuous endpoint other than the concentration gets a time course with
no exposure term, chosen by AIC from a constant, a linear and an exponential
shape. It reproduces the average response over time, which is enough for
testing analysis code, and is not an exposure-response model.

```{r pd, eval=run}
lapply(release$pd, function(shape) c(shape = shape$pd, shape$typical))
```

```{r pd-candidates, eval=tables}
show(fit$pd[[1L]]$candidates, "The three shapes, compared on AIC")
```

## Covariates

Each baseline covariate has one distribution for the whole study, and
synthetic patients draw their covariates from it independently of everything
else. A continuous covariate is a mean and SD on the log scale, taken after the
highest and lowest 5% of patients are set aside. A categorical one is a set of
level frequencies, keeping only levels at least `min_category_patients`
patients (default 3) hold.

```{r covariate-model, eval=run}
str(release$covariates, max.level = 2)
```

Because no covariate is in the PK model, a relationship the real study had,
such as heavier patients having a larger volume, is absent from the synthetic
data. The fit reports where that happens: the correlation between each
covariate and each patient's random effects, as a diagnostic that stays with
the fit.

```{r correlations, eval=tables}
show(fit$correlations[order(-abs(fit$correlations$correlation)), ],
     "Each covariate against the random effects")
```

Here weight and sex both move with volume. If the synthetic data needs that
relationship, `covariate_effects = "auto"` adds allometric scaling of clearance
and volume on body weight. `synpmx_avatar()` keeps such relationships without
modelling them, because each synthetic patient's covariates and profile come
from the same real patients.

## The study design

The design half of the fingerprint is the same as the one
`synpmx_pca_summarize()` builds, from the same code.

### Arms

```{r arms, eval=run}
data.frame(arm = release$arms$arms, patients = as.integer(release$arms$sizes))
```

### Dosing

Each arm has its planned schedule. The levels doses were reduced to, and the
rates of reducing, skipping and stopping, are pooled over the arms.

```{r dosing, eval=tables}
arm <- release$arms$arms[[1L]]
show(release$dosing[[arm]]$planned, "The planned schedule")
```

```{r rates, eval=run}
dosing <- release$dosing[[arm]]
c(levels = length(dosing$levels), reduction = dosing$reduction,
  interruption = dosing$interruption, discontinuation = dosing$discontinuation)
```

`warfarin` is a single dose everyone received, so all three rates are zero and
the planned schedule is reproduced exactly. Where patients do reduce or skip
doses, the synthetic schedule is drawn before the profile is simulated, so a
synthetic patient who steps down gets the lower exposure.

### Visits

The visits each arm has, and one attendance rate per endpoint. Attendance is
drawn visit by visit at generation. The rates, then the number of visits, per
endpoint:

```{r visits, eval=run}
rate <- do.call(pmax, unname(lapply(release$visits, function(v) v$probability)))
signif(tapply(rate, release$cells$endpoint, max), 3)
table(release$cells$endpoint)
```

A visit is kept only where at least `min_arm_patients` patients were observed
there.

### Schema

The shape of the table to generate: its columns and their classes, the
compartment numbers, and any assay limit per endpoint.

```{r schema, eval=run}
release$schema$columns
unlist(release$schema$cmt_obs)
```

## Generating from it

```{r generate, eval=run}
synthetic <- synpmx_model_generate(release, n_subjects = 32, seed = 7)
c(rows = nrow(synthetic),
  subjects = length(unique(synthetic$id)),
  valid = validate_pmx(synthetic, roles)$valid)
```

## Where to go next

- `vignette("pmxmodel-demo")`: this generator run on a study end to end.
- `vignette("pmxmodel-algorithm")`: how each quantity above is estimated.
- `vignette("pca-fingerprint")`: the same walk for the PCA generator, which
  shares the study design and describes the profiles differently.
- `vignette("scorecard")`: the checks that compare generated data with its
  source.
- [Privacy protections in the PMX model generator](https://iamstein.github.io/synpmx/articles/pmxmodel-privacy.html):
  what protects the fingerprint, and what risk remains.
