---
title: "The synpmx PMX Model Algorithm"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The synpmx PMX Model Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
library(synpmx)
stored <- system.file("extdata", "warfarin-model-fit.rds", package = "synpmx")
fit <- if (nzchar(stored)) readRDS(stored) else NULL
has_fit <- !is.null(fit)
```

This article is the full specification of `synpmx_model()`, which fits a small
set of linear pharmacokinetic (PK) models to a study and generates new subjects
by simulating from the fit. Use of the algorithm is demonstrated in
`vignette("pmxmodel-demo")`.

**The aim is a simple model that roughly reproduces the properties of the
study, not one that characterises it.** The synthetic profiles it generates are
for prototyping analysis code. The goodness of fit of the selected model is not
assessed: `model_report()` says what the search had to go on, not whether the
model describes the source. The candidate set is five linear models and the
covariate model is allometric scaling or nothing, which is too little to answer
a scientific question, so the fitted parameters are not estimates to report,
even though the printed object looks exactly like the output of a real
population analysis.

No formal privacy guarantee is offered, although no patient's measured value
reaches the output.

## Three Exported Functions

| Function | Reads patient data | Needs `nlmixr2` | Returns |
|---|---|---|---|
| `synpmx_model_estimate(data, roles, ...)` | yes | yes | `pmx_fitted_model` |
| `synpmx_model_generate(fitted_model, n_subjects, seed)` | no | no | data frame |
| `synpmx_model(data, roles, n_subjects, seed, ...)` | yes | yes | data frame |

`model_report()` inventories everything the fitted object carries,
`model_candidates()` returns the comparison table the selection was made from,
and `model_parameters()` returns the fixed effects, the between-subject
covariance matrix and the residual error.

## Two Time Axes, and `nominal_time` Is Required

**Estimation reads recorded times.** A population PK fit is a statement about
time after the dose that was actually given, so the subject-level dosing records
go to the fitter as recorded, reductions and skipped cycles included. Fitting
against the planned schedule where a patient's dose was reduced would push the
drop in concentration that followed into clearance, and the model would report a
population that eliminates the drug faster than the real one.

**Everything else reads the nominal grid.** The dosing model, the visit model,
and every signal in Steps 1 and 2 sit on the protocol's planned times. On
recorded times a "median profile" is one point per time, because no two patients
share a clock reading, and a study sampled at six times looks like a study
sampled at two hundred. A shape, a peak position and a count of sampling times
are all statements about the protocol.

Inferring that grid is a statement about the protocol only the caller can make,
so `nominal_time` is required rather than derived.

**Repeated doses are written out before anything reads them.** A regimen
recorded as one record plus `ADDL` and `II` — "and 364 more like it, every day"
— is expanded by `pmx_expand_doses()` on the way in, so that every dose the
patient received is a row to the dosing model, the derived time after dose and
the estimation table alike. Before this, every reader counted rows: `onc_sim`
showed the fitter 237 dose records standing for 71,180 doses and got back a
clearance six orders of magnitude off, from a fit that had moved and so tripped
no gate (`REV-051`). The fitter is handed the compact form again, because
`nlmixr2` reads it natively and is fastest on it, and the generated study is
folded back by `pmx_compress_doses()` on the way out, so it comes back in the
encoding its source used.

## Overview of Algorithm

1. **Classify the endpoints.** Four signals decide which endpoint is the drug
   concentration. Everything else continuous becomes a pharmacodynamic (PD)
   endpoint.
2. **Detect the design.** The route of administration and the sampling richness
   prune the set of PK models that will be offered to the fitter.
3. **Estimate the parameters.** Fit each surviving PK model as a population
   nonlinear mixed-effects model through `nlmixr2` and select on AIC, with
   allometric scaling folded in where a weight-like covariate is declared. Fit
   each PD endpoint's time course by least squares.
4. **Fit a dosing model and a visit model per arm**: a planned dose schedule
   with rates for reduction, interruption and discontinuation, and the
   probability of a visit at each nominal time.
5. **Store the fit**: the structural model, the parameters, the arm models and
   the schema, and no per-subject quantity.
6. **Generate new subjects.** Draw covariates and random effects, draw the dose
   schedule, evaluate the profile against that schedule, add residual error and
   put the assay limit back.

Steps 1 to 4 read the source. Step 6 reads only what Step 5 stored, which is why
generation runs on base R and needs neither `nlmixr2` nor a compiler.

## Step 1: Identify the Drug Concentration Endpoint

`pmx_endpoint_types()` already infers continuous, integer, binary and ordinal
from the values, and that inference is reused unchanged. What this generator
adds is a second classification on top of it. Four signals, computed for every
endpoint that is a time course:

1. **Compartment.** The `cmt` role puts the endpoint in the compartment the
   doses go to, or one above a dosing compartment nobody observes, which is the
   depot-and-central convention.
2. **Post-dose only.** Observations before the first dose are absent, or sit at
   or below the censoring limit, in most subjects.
3. **Shape.** The median profile within the richest dose interval rises to a
   maximum and declines without rising again.
4. **Dose proportionality.** Between the highest and lowest dose level, the
   ratio of median subject maxima is within a factor of two of the dose ratio.

**Signals 2 and 4 are required; 1 and 3 break ties.** Being absent before the
first dose and scaling with the dose are properties only a drug concentration
has. A compartment number is a convention the dataset's author chose, and a
rise-and-fall shape is one many biomarkers also have.

Signal 4 is not computable on a single-level study, nor on one dosed by body
weight. A study that gives every patient a slightly different amount has no dose
levels: reading each amount as its own level compares a median of one patient
against a median of one patient over a dose ratio near 1, which almost any
endpoint passes. Where the distinct amounts outnumber half the cohort the signal
reports "not computable" and signals 1 to 3 decide.

Where no endpoint passes, or two pass and neither tie-break separates them, the
function errors and names `endpoint_roles` as the way through. That is the
inference-versus-declaration fork, answered the way `dose_covariate` answers it:
infer where the data settles the question, offer the declaration as an override.

```{r signals, eval=has_fit}
model_report(fit)$endpoints$signals
```

**What this cannot do.** On a study with one continuous endpoint and one dose
level the classification rests on signal 2 alone, and signal 2 passes for any
endpoint with no pre-dose observation, including a white-cell count. The remedy
is `endpoint_roles`, and the evidence is in the table above: a row reading "not
computable" under `proportional` is the classification saying how little it had
to go on.

## Step 2: Detect the Design

Detection prunes the candidate set; the fit picks among what survives. Neither
property below decides the model on its own.

### Route

**A declared route is not detected at all.** `pmx_roles(adm = "ADM", routes =
c("1" = "iv", "2" = "extravascular"))` names the administration column and says
what its values mean, and the two are declared together or not at all: which id
is which route is a convention of the dataset and cannot be read off the
numbers, so a guess would put every dose in the wrong compartment without
failing.

Where a study declares both routes, the route is a property of each dose record
rather than of the study, and the candidate is `1cmt_mixed` — the intravenous
form for the doses given intravenously and the extravascular form for the rest,
summed. That sum is exact rather than an approximation, because every model in
the closed-form set is linear in dose, and it is what lets one patient receive
both. Bioavailability scales the extravascular doses and not the intravenous
ones, which is what bioavailability means and why it is identifiable in a study
dosed both ways and not in one dosed a single way. It is fitted as a fixed
effect with no between-subject term: a design with one dose each way identifies
the contrast between the routes, not a per-subject distribution over it.

**A mixed model is fitted with explicit compartments rather than the closed
form**, and bioavailability is the whole reason. Under `linCmt()` the `f(depot)`
directive does not reach the doses the data sends to the depot: evaluated at
known parameters it scaled the intravenous doses instead, so `f` was estimated
as its own reciprocal and clearance and volume as `cl/f` and `v/f` — a fit whose
own predictions tracked the data while reporting parameters that meant something
else (`SIM-077`). Written as `d/dt(depot)` and `d/dt(central)`, the compartments
are real, `f()` binds to the one it names, and the predictions match the
analytic profile on both routes. Only a mixed study pays the solver cost; a
study dosed one way has no `f` to place and keeps the closed form.

Starting values for a mixed study are read one route at a time, for the same
reason the fit has to be: pooled, the cohort's median profile is an intravenous
decline and an extravascular rise averaged together, and no non-compartmental
quantity read off it describes either route. Disposition comes from the
intravenous records, where `f` is not in the way; absorption from the
extravascular ones, where there is a peak to find; and their two clearances give
`f` itself a starting value, since an extravascular read returns `cl/f`.

Detection below runs only where nothing was declared. It answers with one route
for the whole study, which is the right answer for a study dosed one way and no
answer at all for a study dosed two ways — the `rate` test reads a single
nonzero rate anywhere as "this is an infusion study", so a mixed study came back
as an infusion model with no absorption in it.

A `rate` role carrying a nonzero value is an infusion, and `1cmt_infusion` is
the only candidate. Otherwise one property separates the two remaining routes,
and it is not how many patients peak early.

A drug given by mouth cannot be in the blood at the moment it is swallowed. So a
concentration observed at time zero after a dose is intravenous, and a median
profile that rises before it falls is oral. A profile that declines from a first
sample drawn some time after the dose is both of those at once, an intravenous
bolus or an oral dose whose absorption finished before anybody looked, and no
amount of reading the source separates them. There both routes are offered and
AIC settles it.

Counting subjects instead is the mistake this replaces. In a study whose
sampling starts well past the peak, every subject's profile declines from its
first point, and a per-subject vote reads an oral study as intravenous: it counts
the sampling schedule rather than the drug. The share of subjects whose own
profile rises is still reported, because it says how much of the cohort the
median profile speaks for, but it decides nothing.

```{r route, eval=has_fit}
model_report(fit)$design$reason
```

### Sampling Richness

**Across the cohort**, distinct nominal times after a dose holding an
observation. Below six the fit still runs and warns: a one-compartment model is
not identifiable from that sampling, so its parameters stay near their starting
values and the simulated profiles are a plausible shape rather than this
study's. There is no simpler structural model to drop to --- the default
candidate set is already one compartment. `synpmx_avatar()` and `synpmx_pca()`
carry sparse sampling without fitting a structure to it. No post-dose
observation at all is an error, since there is then no curve to fit.

**Per subject**, distinct nominal times after a dose within the richest dose
interval. Four or more, with two after the median peak, is enough to identify a
distribution phase. A sample *before* the peak is required only where there is
an ascending limb to sample, since an intravenous bolus peaks at the dose.

This count decides nothing. It is reported, and `model_report()` says so when
the sampling would support a two-compartment model, because asking for one is
your move rather than the search's.

### One Compartment by Default

The default is one-compartment: `1cmt_iv`, `1cmt_oral` or `1cmt_infusion`. Where
the route is ambiguous both are fitted and AIC chooses, which is the only case
the default runs more than one fit.

A distribution phase is a refinement of a shape the one-compartment model
already has, and this generator exists to make simulated profiles resemble the
source study rather than to characterise it. The cost is not small: a
two-compartment fit takes several times as long, and on data a one-compartment
model describes it spends that time against a flat likelihood and reports the
worse AIC anyway.

**It remains available.** `pk = "2cmt_oral"` or `pk = "2cmt_iv"` forces the
model and skips the search, and the generator simulates from it exactly as it
does from any other, since the closed-form solution is already there.

## Step 3: Estimate the PK and PD Parameters

**The candidates are the structural PK models Step 2 left standing.** Each one
is fitted as a population nonlinear mixed-effects model through `nlmixr2`, which
estimates the fixed effects, the between-subject covariance matrix and the
residual error, and where more than one candidate is offered they are compared
on AIC. The PD endpoints are estimated separately, and not as mixed-effects
models.

The candidate set is at most the five models in the package's closed-form set,
and never more. A candidate the fitter could estimate and the generator could not
simulate would be a model that fits and then generates nothing, so the two lists
are one list. That is also why no dose effect on clearance is fitted: it would
mean a candidate set that is no longer five closed-form models, and generation
would need a solver.

**A subject the endpoint was never measured in is not fitted.** It adds no term
to the likelihood, and a record carrying doses and no samples is at best ignored
by the solver. Its dosing and its visits are still real and still reach the
models of Step 4, which read the study rather than the table handed to the
fitter. How many subjects that left out is said before the fit, and it warns
where it takes the fitted cohort under `min_subjects`.

Each candidate carries log-normal between-subject variability on every
structural parameter and a proportional residual error. An endpoint holding
values at or below zero after censoring is handled gets an additive error
instead, and the substitution is recorded on the fit.

**The concentration is fitted with its censored observations censored.** Where
the study declares a censoring column, the rows below the assay limit reach
`nlmixr2` as censored -- the limit in `DV`, the flag in `CENS`, and the other
bound in `LIMIT` where the study reports one -- so each contributes the
probability of falling below the limit rather than a value nobody measured. On
30 subjects of `case1_pkpd`, 45% of them below the limit, that moves the
proportional residual from 0.517 to 0.290 against imputing the same rows, and
the residual is what sets the scatter of every generated observation. It costs
about a factor of 1.7 in fitting time there, and less where less of the endpoint
is censored.

**Everything else reads an imputed value**, because nothing else has a
likelihood to put censoring in: the visit model, the covariate model and the PD
shapes are least-squares fits, and a stack of identical boundary substitutions
would bend each of them toward the limit. Those rows are replaced by a uniform
draw inside the censoring region -- a draw rather than a fixed LLOQ/2, which
would swap one artificial spike for another -- and the boundary is put back when
the synthetic data is emitted. The share of each endpoint sitting below the
limit is reported with the fit either way.

Starting values come from a non-compartmental reading of the source rather than
from a guess, because a population fit started far from the answer either
converges slowly or reports the starting values back. The terminal slope is
fitted by log-linear regression over the points after the peak, and everything
leans on it. Clearance is the dose over the area **extrapolated to infinity**,
not over the trapezoid alone, which on a study sampled to four half-lives
understates the area by about a tenth and overstates clearance by the same.
Volume is clearance over the terminal slope, which is the quantity the terminal
phase identifies, since dose over the peak is not a volume of any kind for an
oral dose. Absorption is solved from where the peak falls: for a one-compartment
oral model the peak sits at `log(ka/ke)/(ka - ke)`, one equation in one unknown
once the slope has given `ke`. For a two-compartment model the steady-state
volume comes from the mean residence time, with the absorption mean taken back
off for an oral dose, and is split between the central and peripheral
compartments.

**The default fits one model and stops.** Not a search that happens to have one
candidate: one model, chosen by the route detection above, with allometric
scaling folded into it where a weight-like covariate is declared. `pk` is how
you buy accuracy with time.

```r
# One fit, the default.
synpmx_model_estimate(data, roles)
# One fit, your choice of model.
synpmx_model_estimate(data, roles, pk = "2cmt_oral")
# Two fits, compared on AIC.
synpmx_model_estimate(data, roles, pk = c("1cmt_oral", "2cmt_oral"))
```

**How long it takes is set by the dose records, not the patients.** Every
likelihood evaluation superposes one contribution per dose per subject, so the
cost scales with the number of dose events rather than with cohort size. A
single-dose study of a few dozen patients fits in seconds. A daily regimen
carrying scores of dose records per subject over a cohort several times larger is
two orders of magnitude more work, and a call that has not returned after half an
hour is that cost rather than a fault.

**What it did take is reported.** Each candidate's fitting time is a column of
the candidate table, the fit prints the total with itself, and the run says so
as it finishes, because the number a caller weighs a rerun against is how long
the last one took.

**Where there is a search, selection is on AIC**, and the estimation method is
`focei` for that reason. SAEM's log-likelihood is a Gaussian-quadrature step run
after the fit, and at phase 1 cohort sizes it returns a non-finite value, so a
search over two candidates has nothing to compare. `estimation = "saem"` remains
available for a study large enough to give it a likelihood. A candidate whose
AIC is not finite is recorded as not converged whichever method produced it.

A candidate that fails drops out carrying its reason and stays in the table, so
that a search which came down to one survivor does not look like a search that
had one candidate. Where nothing converges the function errors rather than
returning the least bad fit.

```{r candidates, eval=has_fit}
model_candidates(fit)
```

### Covariates

The default `"auto"` applies allometric scaling on clearance and volume where a
weight-like covariate is declared, and fits nothing else. The exponents are the
standard 0.75 and 1 rather than estimated ones, and the effect is **asserted
rather than tested**: it is folded into the one fit, not compared against a
model without it. Testing it would double the cost of the whole call.
`covariate_effects = "none"` switches it off.

The covariate is recognised by name, `wt`, `weight`, `bw` and the like, and must
be numeric and positive. There is no way to recognise a body weight from its
values alone.

**The cost of this default is explicit.** A covariate that influences the real
profiles and is not in the model is generated independently of them, so the
synthetic data carries no relationship between the two. `model_report()` reports
the correlation between each declared covariate and the individual random
effects, which is where an unmodelled relationship shows up.

### Pharmacodynamics

Every remaining continuous endpoint is fitted against a constant, a linear and
an exponential time course and selected on AIC, with between-subject variability
on the baseline. `pd = c(endpoint = "linear")` names the shape and skips the
search for that endpoint; a named shape that will not fit is an error rather
than a silent fallback. There is no minimum number of observations: a
shape is a candidate where it has **one residual degree of freedom**, which is
what stops a line being fitted exactly through two points and then winning the
comparison it was never tested by. Below any candidate the endpoint is a
constant at its mean, which is what an endpoint measured once is, and the report
says that rather than naming a search that did not run.

These shapes have no concentration term, so a PD endpoint driven by exposure is
reproduced as a time course that happens to resemble the average subject's
response.

**One shape for the whole cohort by default, and one per arm on request.** The
pooled fit predicts a single number for every arm, so a synthetic patient's dose
does not reach their response: on `case1_pkpd` the source's mean PD after 1500 h
runs 83, 68, 70, 126, 237, 341 across placebo and five ascending doses, and the
pooled shape answers 149 to all six. `pd_by_arm = TRUE` fits the shape within
each arm instead, which is the same per-arm summary the dosing and visit models
already are and asserts no dose-response form -- an arm is fitted on its own
observations, on the same three-candidate ladder, or not at all. On that study
it answers 47, 76, 77, 126, 217, 363. It is not the default because it is a
different claim about the study: the pooled shape says these endpoints are a
time course, the per-arm shape says each arm has its own. An arm holding too
little of an endpoint to fit keeps the pooled shape, so no arm is left with
nothing to generate.

**The between-subject spread and the residual are read around each subject's
own curve, not around the population one.** Taking the residual as the spread of
every point about the typical curve puts all of the between-subject variation
into it, and generation then emits that as independent noise per observation. On
a study where subjects share a shape and sit at different levels, almost the
whole of the structure comes back as scatter and no synthetic subject has a
profile at all. `.pd_profile()` is linear in `baseline` for every shape, so a
subject's own baseline is a least-squares projection needing no second
optimizer, and it is the same quantity generation draws: the between-subject
term is the spread of those baselines on the log scale, and the residual is what
is left around each subject's own curve.

**A shape that fails to fit stays in the table.** The exponential is tried from
four start sets rather than one, because the single median-based set fails on
any response that falls and then recovers, which is the shape a turnover
endpoint has. Where none of the four converges the candidate is reported with
`converged` false and the solver's message rather than dropped, so a flat line
winning on AIC against two other flat lines is visible as that.

All three shapes are fitted by least squares rather than through `nlmixr2`, so
all three together cost no measurable time, and **the only compiled fits in the
whole call are the PK candidates.** `model_report()` breaks the wait down by
fit: one line per candidate the search compiled, and one for the least-squares
shapes. A search that spent nine of its ten minutes in a second candidate is a
run whose answer is to name `pk`.

Binary and ordinal endpoints are not fitted at all. They are drawn from the level
frequencies their arm holds at each nominal time, which is what the visit model
already does for attendance.

## Step 4: Fit the Dosing Model and the Visit Model

Everything that is not the concentration-time curve comes from the models
`synpmx_pca_summarize()` already builds, unchanged in what they represent. Dose
reductions, interrupted cycles, discontinuation and missed visits are modelled
here exactly as they are there, and so are arm sizes, the covariate
distributions, the censoring boundary and the schema. The two generators call the
same code rather than a copy of it; what differs is only the adapter that hands
it the grid, and this generator writes its own over the nominal times the source
holds.

Per arm:

| Model | What it holds | Drawn at generation as |
|---|---|---|
| Planned schedule | The nominal dose times enough of the arm reached, and the modal amount at each cycle among patients still on their starting dose | The cycle grid every subject starts from |
| Dose ladder | The levels patients dropped to, as ratios to their own starting dose, built from within-patient decreases | The amount multiplier in force at a cycle |
| Reduction rate | Discrete-time hazard of stepping down a level | Decided before the cycle is dosed |
| Interruption rate | Discrete-time hazard of skipping a cycle without ending treatment | Decided at the cycle |
| Discontinuation rate | Discrete-time hazard of stopping treatment | Decided after the cycle is dosed |
| Visit model | Per endpoint and per retained nominal time, the fraction of the arm holding an observation there | Attendance, drawn per visit |

A study where nobody reduces, skips or stops early has all three rates at zero
and one level, and the model then reproduces the planned schedule exactly. No
detector decides which kind of study this is, because the rates already say it.

A grid cell is kept only where at least `min_arm_patients` distinct patients hold
an observation there. A nominal time one patient attended is that patient, and
generating from it would put them back.

## Step 5: Store the Fit

Estimated by `nlmixr2`: the selected structural model, the candidate comparison
table, fixed effects, the between-subject covariance matrix, the residual error,
the covariate effects that survived, the PD shape and parameters per endpoint,
and how long the fitting and the whole call took.

Not estimated: the per-arm dosing model and its three rates, the visit model, arm
sizes, the covariate distributions, the censoring boundary, the schema and the
roles.

**Everything on the object is an input to Step 6, so printing it reports all of
them** — the dose reduction, skipped cycle and early stop rates per arm, the
attendance frequency behind a missed observation, how each covariate is drawn,
which cells are drawn from recorded values rather than simulated, the assay
limit and how many observations sit below it, the floor nothing is emitted
below, and the
columns the generated table will carry. `model_report()` returns the same
account as a list.

**No individual estimates.** Empirical Bayes estimates are per-subject
quantities, and a fitted model that carried them would be writing out a
description of each real patient. They are not stored on the object. Generation
draws random effects from the covariance matrix instead, and the correlation
report above computes from them and keeps only the correlation.

```{r report, eval=has_fit}
model_report(fit)
```

## Step 6: Generate New Subjects

Per synthetic subject: assign an arm keeping the source arm shares, draw
covariates from the arm's covariate model, draw random effects from the
covariance matrix, apply the covariate effects to the typical parameters, draw
the dose schedule from the arm's dosing model, draw the visits attended from the
visit model, evaluate the profile at the attended times against the drawn
schedule, add residual error, apply the censoring boundary and emit.

**The order of the two draws is fixed.** The schedule is drawn first and the
profile is computed from it, so a synthetic subject who steps down a dose level
has a lower exposure from that cycle on, and one who skips a cycle has the
trough that implies. This is the one thing the generator does that neither
`synpmx_avatar()` nor `synpmx_pca()` does. Drawing a profile and then a schedule
would put a subject's dose reduction in the dosing records and not in the
concentrations.

**The proportional multiplier is lognormal.** Written as `1 + N(0, cv)` it is
non-positive with probability `pnorm(-1 / cv)`, which is 2.3% of draws at a
coefficient of variation of 0.5 and 7.7% at 0.7, and each of those is a
concentration clamped to zero. A coefficient of variation that high is a
statement that the structural model does not describe the data, and it belongs
in the output as a wide band rather than as a scatter of zeros on the floor of a
log axis. `exp(N(0, sqrt(log(1 + cv^2))))` has the same coefficient of variation
and a median of one, so the spread the fit estimated is preserved and no draw
reaches zero.

**A value is floored at the smallest one the study reported, halved.** The
residual is not the only thing that can put a synthetic value below anything the
assay could return: a one-compartment profile evaluated late in a long dose
interval underflows on its own. A study that declares a censoring column says
where its assay stopped, and that boundary is put back. A study that declares
none still had an assay, and its smallest reported value is the only evidence of
where the limit sat, so half that value becomes a floor. An endpoint that reports
a zero is given no floor, since a zero is a value a floor would contradict.
`model_report()` carries the floor per endpoint.

**A floor that catches a large share of the output is a warning, not a
repair.** Raising a value is still changing it, and a floor doing that to much
of a dataset is hiding a fitted model that does not describe the low end of the
data. What the floor caught is recorded on the generated dataset as the
`pmx_floored` attribute, and above one observation in twenty the generator says
so.

## Gates

Only a study the generator cannot fit at all is refused. A study it can fit
badly is fitted, and told about.

| Gate | Threshold | Result |
|---|---|---|
| Cohort size | 20 subjects | Warns and fits. A covariance matrix fitted to a handful of subjects describes those subjects, and nothing downstream will say so: the scorecard asks whether the output copies anybody or changed the study's shape, and a small-cohort fit does neither. `synpmx_pca()`'s floor is 10 and this one is higher because a parameter estimate concentrates on its cohort faster than a score does. |
| Cohort time coverage | 6 distinct nominal times after a dose | Warns and fits. Below it a one-compartment model is not identifiable and the parameters sit close to their starting values, which is the caller's call to accept. |
| No observation after a dose | — | Errors, naming which of the role columns the empty count came from: nothing selected as an observation, nothing selected as a dose, or the two never meeting in one subject. |
| Arm size | 3 patients | Warns and drops those patients before anything is fitted, so the arm is absent from the model and from the data generated from it. Inherited from the dosing and visit models, which are summaries of an arm: an arm of one or two has no rates to pool. |
| `nominal_time` undeclared | — | Errors. The grid is a statement about the protocol only the caller can make. |
| No grid cell shared | `min_arm_patients` | The cell is dropped. A nominal time one patient attended is that patient. |
| Administration column | `adm` and `routes` together | Errors on either alone. What an administration id means is a convention of the dataset, and reading it wrong routes every dose to the wrong compartment silently. |
| PD shape candidacy | 1 residual degree of freedom | The shape is dropped from the comparison, not the endpoint from the study. Below every candidate the endpoint is generated as a constant at its mean. A shape fitted exactly through its own points has `AIC` `-Inf` and would win any comparison it entered. |
| No PK endpoint identified | — | Errors and names `endpoint_roles`. |
| No candidate converged | — | Errors rather than returning the least bad fit. |
| The fit did not move | 1% of every starting value | Warns, and prints `!! THE FIT DID NOT MOVE !!` above the parameter block in `model_report()`. A converged-looking fit is not necessarily an estimated one: where the optimizer takes no effective step, `nlmixr2` returns an objective, an AIC and a full table of starting values, and nothing else downstream contradicts them — the generator simulates from them and the scorecard passes, because it asks whether the output copies anybody or changed the study's shape and a fit that never moved does neither. Each parameter's start, estimate and percent change are listed, because "did not move" is a claim the reader has to be able to check. |
| A fitted model as a public input | — | `pmx_prior()`, `synpmx_prior()` and `synpmx_calibrated()` refuse a `pmx_fitted_model`, because its parameters were estimated from the confidential study. |
