---
title: "Demo: Using synpmx_model"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx_model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_dplyr <- requireNamespace("dplyr", quietly = TRUE)
has_dt <- requireNamespace("DT", quietly = TRUE)
stored <- system.file("extdata", "case1-pkpd-model-fit.rds", package = "synpmx")
fit <- if (nzchar(stored)) readRDS(stored) else NULL
run <- has_xgxr && has_dplyr && !is.null(fit)
plots <- run && has_ggplot2
distributions <- plots && requireNamespace("patchwork", quietly = TRUE)
tables <- run && has_dt
# Three significant digits on every measured quantity. Times, counts and
# identifiers are exempt: `signif(2015.9, 3)` is 2020, which would move a
# nominal visit onto a time the study never had.
exact <- c("time", "ntime", "nomtime", "nominal_time", "cycle", "id", "index",
           "n", "observations", "imputed", "patients", "min_patients",
           "numbers", "planned_cycles", "distinct", "source_doses", "offset",
           "count")
digits3 <- function(x) {
  x <- as.data.frame(x)
  for (column in names(x)) {
    if (tolower(column) %in% exact || !is.numeric(x[[column]])) next
    x[[column]] <- signif(x[[column]], 3)
  }
  x
}
show <- function(x, caption = NULL) {
  if (!has_dt) return(digits3(x))
  DT::datatable(
    digits3(x), rownames = FALSE, caption = caption,
    options = list(dom = "t", paging = FALSE, scrollX = TRUE, ordering = TRUE)
  )
}
```

Fit a population model to a study, look at what the fit carries, and generate a
synthetic dataset by simulating from it. No number a patient measured reaches
the output: what leaves the source is a structural model, a handful of fixed
effects, a covariance matrix, a residual error, and a dosing and visit model per
arm.

It makes no formal privacy claim, and **it is not for estimation** — the fitted
parameters exist to make simulated profiles look like the source study, and the
object prints that warning with itself. The full specification is in
`vignette("pmxmodel-algorithm")`.

The dataset is `xgxr::case1_pkpd`: 180 patients, six arms from placebo to 300
mg, with a pharmacokinetic (PK) concentration endpoint and a continuous
pharmacodynamic (PD) endpoint. `vignette("avatar-demo")` and
`vignette("pca-demo")` run the other two generators over the same study.

## Configuration

To run this on your own study, edit the two chunks below that:

1. Read in the dataset.
2. Define the column roles.

The rest of the code can be kept as is.

```{r data, eval=run, message = FALSE}
library(dplyr)
raw <- as.data.frame(get(utils::data(list = "case1_pkpd", package = "xgxr"))) |>
  mutate(CENS = ifelse(NAME == "PD - Continuous", 0, CENS))
SEED <- 808
```

The column meanings.  Columns not specified here are dropped from the synthetic dataset.

```{r roles, eval=run}
roles <- pmx_roles(
  id           = "ID",
  time         = "TIME",
  dv           = "LIDV",
  cens         = "CENS",
  amt          = "AMT",
  evid         = "EVID",
  cmt          = "CMT",
  dvid         = "NAME",        # which endpoint each observation row is
  nominal_time = "NOMTIME",     # the grid the visit model is built on
  strata       = c("TRTACT", "DOSE"), # treatment arms, fitted one at a time
  covariates   = "WEIGHTB",     # drawn, and available to the structural model
  keep         = "STUDY"        # carried through verbatim
)
```

## Fitting

`synpmx_model_estimate()` is the only stage that reads patient data, and the only
one that needs `nlmixr2`. It works out which endpoint is the drug, what design
produced it, fits the candidates that design admits and picks one on AIC.

```{r estimate-shown, eval=FALSE}
fit <- synpmx_model_estimate(raw, roles, seed = 1)
```

```{r estimate-note, echo=FALSE, results="asis", eval=run}
cat("The chunk above is shown rather than run. Fitting compiles a model, so this",
    "document reads a stored fit built by `scripts/build-model-fits.R` and",
    "`R CMD check` never needs a compiler.\n")
```

That call takes about eleven minutes on this study, and says so before it
starts: every likelihood evaluation sums a contribution per dose per subject,
and twelve weeks of daily dosing for 150 patients is 12,750 dose records. Where
a study's doses are on an exact interval they are compressed to one record per
patient first, which the message also reports; `case1_pkpd` records its dose
times as actuals — 0, 24.22, 48.28 — so there is no exact interval to compress
to, and the wait is the honest cost of the design.

```{r fit-print, eval=run}
fit
```

A clearance of `r if (run) signif(fit$parameters$fixed[["cl"]], 3) else NA` L/h
and a volume of `r if (run) signif(fit$parameters$fixed[["v"]], 3) else NA` L.
Whether those are the right numbers for this compound is not the question the
generator asks: they exist to put the simulated profiles where the source's
are, and the object prints that warning with itself.

## What the fit carries

Two halves. `model_report()` separates them, because they answer to different
things — one half is an estimate with all an estimate's caveats, the other is a
summary of the study's apparatus.

```{r report, eval=run}
model_report(fit)
```

Three lines in the estimated half are worth reading before anything else. The
residual error is proportional, which follows from every observed concentration
here being positive; a study recording zeros gets an additive error instead,
because a proportional one has nothing to scale there. The PD endpoint gets an
exponential shape from the least-squares search, with no exposure term. And 46%
of the concentrations sit below the assay limit and were imputed inside the
censoring region before fitting — a study this censored asks the fit to describe
mostly-censored arms, and the arm-by-arm comparison further down is where that
shows.

The signals table says how the concentration endpoint was decided: `PK
Concentration` is post-dose, dose-proportional and rise-and-fall, and the PD
endpoint is none of the three.

The correlation block is the other thing to read. A covariate that moves with a
random effect and is not in the model above is generated independently of the
profiles, so the synthetic data carries no relationship between them.
`synpmx_avatar()` keeps those relationships without modelling them.

## The candidates

```{r candidates, eval=run}
show(model_candidates(fit), "Candidates the design admitted")
```

One row, because the default fits one model and stops. A distribution phase is a
refinement of a shape the one-compartment model already has and costs several
times as long to fit, so it is available rather than routine: `pk = "2cmt_oral"`
forces it, `pk = c("1cmt_oral", "2cmt_oral")` fits both and picks on AIC, and
`model_report()` says when the sampling would support one.

## Generating

`synpmx_model_generate()` reads no patient data. Its arguments are the fit and a
subject count, so everything about the source that reaches the output has already
passed through the fit.

```{r generate, eval=run}
synthetic <- synpmx_model_generate(fit, seed = SEED)
```

The generated table is in the source's shape: same columns, same classes, same
compartment numbers, new identifiers.

```{r head, eval=run}
show(head(synthetic, 10), "The first ten rows")
```

## Plot synthetic data and original data

The same overlays the other two demos draw, so the three can be read against
each other.

```{r overlay-setup, eval=plots}
library(ggplot2)
library(xgxr)
xgx_theme_set()
comparison_colours <- c(source = "#1B6CA8", synthetic = "#D95F02")

both <- rbind(
  cbind(raw[, names(synthetic)], DATA = "source"),
  cbind(synthetic, DATA = "synthetic")
)
obs <- both[both$EVID == 0 & !is.na(both$LIDV), ]
obs$TRTACT <- factor(obs$TRTACT,
                     levels = c("Placebo", "3 mg", "10 mg", "30 mg",
                                "100 mg", "300 mg"))
```

```{r overlay-pk, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PK Concentration" & obs$TIME <= 24, ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA~TRTACT) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 24, by = 6)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "PK concentration", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("Day 1 Conc. Profile")
```

```{r overlay-pk-last, eval=plots, fig.height=5}
last_dose <- obs[obs$NAME == "PK Concentration" &
                   obs$TIME >= 2016 & obs$TIME <= 2040, ]
last_dose$TAD <- last_dose$TIME - 2016

ggplot(last_dose, aes(TAD, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.4) +
  facet_grid(DATA~TRTACT) +
  xgx_scale_y_log10() +
  xgx_scale_x_time_units("hours", breaks = seq(0, 24, by = 6)) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time after dose (hours)", y = "PK concentration", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("Last Dose (2016 h) Conc. Profile")
```

The synthetic profiles are smoother than the source's, and that is the
generator's shape rather than a fault in the run: every one of them is the same
one-compartment curve evaluated at a different draw of clearance, volume and
absorption, with residual error on top. A real cohort's profiles wander in ways
one structural model does not reproduce.

Almost half the source concentrations are below the assay limit, so the honest
comparison in the low arms is how much of each arm is censored rather than where
its curve sits.

```{r blq-table, eval=run}
censored <- function(data, label) {
  rows <- data$EVID == 0 & data$NAME == "PK Concentration" & !is.na(data$LIDV)
  out <- aggregate(list(pct_blq = data$CENS[rows] == 1),
                   list(arm = data$TRTACT[rows]),
                   function(x) round(100 * mean(x)))
  stats::setNames(out, c("arm", paste0("pct_blq_", label)))
}
blq_table <- merge(censored(raw, "source"), censored(synthetic, "synthetic"),
                   by = "arm")
show(blq_table, "Share of concentrations below the assay limit, by arm")
```

The censored share moves in the right direction across the arms and does not
match the source arm for arm: the generator censors more of the 3 mg arm than
the study did and less of the 30 and 100 mg arms. One clearance and one volume
distribution, shared by every arm and evaluated at the arm's dose, cannot
reproduce six arm-specific censoring fractions — the study's low arms are not
simply its high arms divided down. Where the analysis to be developed turns on
how much data is below the limit, read this table before trusting the output.

```{r overlay-pd, eval=plots, fig.height=5}
ggplot(obs[obs$NAME == "PD - Continuous", ],
       aes(TIME, LIDV, group = ID, colour = DATA)) +
  geom_line(alpha = 0.35) +
  xgx_scale_x_time_units(units_dataset = "hours", units_plot = "weeks") +
  facet_grid(DATA~TRTACT) +
  scale_colour_manual(values = comparison_colours) +
  labs(x = "Time (hours)", y = "PD (continuous)", colour = NULL) +
  theme(legend.position = "top") +
  ggtitle("PD Response")
```

The PD endpoint is an exponential in time per subject, which is the whole of
what the shape search offers: a constant, a line, or an exponential, with
variability on the baseline and no exposure term. Its spread across the study
lands close to the source's — the median moves from 123 to 136 and the range is
about as wide — because the shape was fitted to the study rather than asserted.

**What is missing is the dose ordering.** One shape is fitted to the pooled
observations and each subject gets a draw on its baseline, so the arm a subject
was assigned to does not reach its response.

```{r pd-by-arm, eval=run}
late_pd <- function(data, label) {
  rows <- data$NAME == "PD - Continuous" & data$EVID == 0 &
    data$TIME > 1500 & !is.na(data$LIDV)
  out <- aggregate(list(mean_pd = data$LIDV[rows]),
                   list(arm = data$TRTACT[rows]), function(x) round(mean(x)))
  stats::setNames(out, c("arm", paste0("mean_pd_", label)))
}
pd_table <- merge(late_pd(raw, "source"), late_pd(synthetic, "synthetic"),
                  by = "arm")
show(pd_table, "Mean PD response after 1500 h, by arm")
```

The source's arms are ordered by dose. The synthetic ones are not: its placebo
arm can sit above its 300 mg arm. That is the documented boundary of this
generator rather than a bad seed — the PD shape carries no exposure term, so a
dataset whose point is exposure-response is not served by it.
`synpmx_avatar()` keeps that relationship without modelling it, and
`vignette("avatar-demo")` runs it on this study.

## Distributions of Synthetic and Original Data

One panel per endpoint and per baseline covariate, source against synthetic.
`compare_pmx_distributions_height()` sizes the figure, since it grows a row of
panels at a time; `output = "tables"` gives the numbers behind it.

```{r distributions, eval=distributions, fig.height=compare_pmx_distributions_height(raw, roles)}
compare_pmx_distributions(raw, synthetic, roles)
```

## Scoring it

`synpmx_scorecard()` does not know what produced the dataset it scores, so it
reads this output the same way it reads an AVATAR or PCA one.

```{r scorecard, eval=run}
card <- synpmx_scorecard(raw, synthetic, roles)
synpmx_scorecard_datatable(card)
```

Nothing fails. **B4b reads 0, and it is the row that answers the disclosure
question for this generator**: no value any patient measured is reproduced. B4a
asks the same question of the observation times and reads `not applicable`,
because this generator draws attendance per visit from a per-arm probability
rather than reusing anybody's visit set — a match would be a coincidence with a
computable chance rather than a copy.

**Three rows read `not applicable` for a different reason.** B1a, B1b and C2
read a run record that `synpmx_avatar()` writes on its output and this generator
does not, so there is nothing for them to measure. That is a gap in those three
checks rather than a property of this dataset.

**Two rows ask to be read.** A5a reports observations per patient falling from
30.7 to 29, which is the visit model drawing attendance rather than copying it.
D1 reports the standard deviation of `PK Concentration` at 1.3 times the
source's, the furthest of the three numeric variables — a spread that widened,
consistent with the censored arms above.

`vignette("avatar-scorecard")` documents what each row asks and what its pass
criterion is.

## One call

`synpmx_model()` is the two stages together, for when you do not need to look at
the fit first. It is on the result either way, as an attribute.

```{r one-call, eval=FALSE}
synthetic <- synpmx_model(raw, roles, seed = SEED)
attr(synthetic, "pmx_fitted_model")
```

## See also

- `vignette("pmxmodel-algorithm")` — what each step does and why.
- `vignette("pmxmodel-fingerprint")` — every quantity the fit carries out of the
  study, in detail.
- [Evaluating the model generator on public data](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html)
  — the same generator over the public studies the other two surveys use.
- `vignette("pca-demo")` — the same study through the principal-component
  generator.
- `vignette("avatar-demo")` — and through blending.
- `vignette("avatar-scorecard")` — the checks above, in detail.
