---
title: "The PMX model study fingerprint"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The PMX model study fingerprint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_dt <- requireNamespace("DT", quietly = TRUE)
stored <- system.file("extdata", "warfarin-model-fit.rds", package = "synpmx")
fit <- if (nzchar(stored)) readRDS(stored) else NULL
run <- has_data && !is.null(fit)
plots <- run && has_ggplot2 && has_xgxr
tables <- run && has_dt

# Three significant digits on measured quantities. Times and counts are exempt:
# rounding a nominal visit time would move it onto a time the study never had.
exact <- c("time", "ntime", "nomtime", "nominal_time", "cycle", "id", "index",
           "n", "observations", "imputed", "patients", "min_patients",
           "numbers", "planned_cycles", "distinct", "source_doses", "offset",
           "count")
digits3 <- function(x) {
  for (column in names(x)) {
    if (column %in% exact || !is.numeric(x[[column]])) next
    x[[column]] <- signif(x[[column]], 3)
  }
  x
}
show <- function(x, caption = NULL, paged = FALSE) {
  if (!has_dt) return(digits3(as.data.frame(x)))
  DT::datatable(
    digits3(as.data.frame(x)), rownames = FALSE, caption = caption,
    options = list(dom = if (paged) "tip" else "t", paging = paged,
                   pageLength = 10, scrollX = TRUE, ordering = TRUE)
  )
}
```

Every generator in this package reduces a study to a **fingerprint** — a set of
summaries small enough to carry out of the environment that holds the real data,
and complete enough to build a synthetic study from. The fingerprint is the
whole of what a synthetic dataset descends from. Nothing else about the source
reaches it, which is why looking at the fingerprint is how you decide whether
the output can be trusted, and how you satisfy yourself that no patient left the
room.

`synpmx_model_estimate()` returns this generator's fingerprint, as a
`pmx_fitted_model`. `synpmx_model_generate()` reads that object and no patient
row.

This vignette walks through it group by group. It is a reference: read
`vignette("pmxmodel-demo")` first for a run end to end, and
`vignette("pmxmodel-algorithm")` for how each quantity is arrived at and why.
`vignette("pca-fingerprint")` is the same walk through the PCA generator's
fingerprint, and the two are worth reading side by side.

## Two halves, and they are not alike

This fingerprint divides more sharply than the PCA one does.

**The estimated half** is a population model: a structural form, a handful of
fixed effects, a covariance matrix, a residual error. It is perhaps twenty
numbers describing the shape of every profile in the study, which is a very
small fingerprint for a very strong claim. It also looks exactly like the output
of a real population analysis, and it is not one — the candidate set is small
and the covariate model is allometric scaling or nothing.

**The summarized half** is the study's apparatus: who was in which arm, when
doses were planned and how patients departed from that plan, which visits were
attended, what the covariates looked like, where the assay limit sat. None of
this is estimated. It is exactly the apparatus `synpmx_pca_summarize()` builds,
from the same code.

The two halves carry different risks. Twenty fitted numbers are a compact
summary of 32 people; a per-arm dosing model with a cycle grid can be much
larger and is backed by one arm's patients rather than the whole cohort.

## The study

`nlmixr2data::warfarin` — thirty-two patients, a single oral dose, a
concentration endpoint and a pharmacodynamic one.

```{r data, eval=run}
data("warfarin", package = "nlmixr2data")
warfarin <- as.data.frame(warfarin)
warfarin$ntime <- warfarin$time
roles <- pmx_roles(
  id = "id", time = "time", nominal_time = "ntime", dv = "dv", amt = "amt",
  evid = "evid", dvid = "dvid", covariates = c("wt", "age", "sex")
)
```

```{r estimate-shown, eval=FALSE}
fit <- synpmx_model_estimate(warfarin, roles, seed = 1)
```

The chunk above is shown rather than run: fitting compiles a model, so this
document reads a stored fit and `R CMD check` never needs a compiler.

```{r fit, eval=run}
fit
```

## The inventory

`model_report()` is the top-level account, and it separates the two halves
because they answer to different things.

```{r report, eval=run}
model_report(fit)
```

Every section below expands one part of it. The object is a plain list and the
names used are its own, so a quantity can be reached directly when no accessor
covers it.

```{r names, eval=run}
names(fit)
```

## The settings that produced it

```{r settings, eval=run}
unlist(fit$settings)
c(patients = fit$n_source, arms = length(fit$arms$arms))
```

## How the concentration endpoint was decided

Not a released quantity, but the first thing to read: everything downstream is a
model *of* this endpoint, and if the wrong one was chosen nothing else matters.

```{r signals, eval=run}
show(fit$endpoints$signals, "The four signals, per endpoint")
```

`proportional` reads `NA` here because `warfarin` gives every patient the same
dose, so there are no dose levels to compare — the classification rests on the
remaining signals. A column of `NA` under `proportional` is the object telling
you how little evidence it had, and `endpoint_roles` is how you overrule it.

```{r design, eval=run}
fit$design$reason
fit$endpoints$decided_by
```

## The structural model

```{r structural, eval=run}
fit$structural
model_candidates(fit)
```

One row, because the default fits one model. `pk` is what asks for more.

## The fixed effects

The typical parameters, on the natural scale. These are the numbers that look
most like a result and are least entitled to be read as one.

```{r fixed, eval=run}
model_parameters(fit)$fixed
```

## Between-subject variability

A covariance matrix on the log scale, one row per parameter carrying a random
effect. Generation draws each synthetic subject's deviations from this matrix.

```{r omega, eval=run}
model_parameters(fit)$omega
sqrt(diag(model_parameters(fit)$omega))  # as CV on the log scale
```

**No individual estimates.** Empirical Bayes estimates are per-subject
quantities, and an object carrying them would be a description of each real
patient. They are not here. This matrix is what stands in for them, and it is a
statement about the population rather than about anybody in it.

## The residual error

```{r residual, eval=run}
model_parameters(fit)$residual
```

Additive here rather than proportional, because `warfarin` holds concentrations
recorded as zero and a proportional error on zero is zero. The substitution is
recorded on the object rather than being silent.

## What the assay limit cost

Values below the limit are imputed before anything is fitted, and the boundary
is put back when data is generated. That is the intended behaviour — it is what
lets every part of the fingerprint read a latent value rather than a stack of
identical boundary substitutions — but it is an assumption, and its weight is
the share of each endpoint carrying it.

```{r censoring, eval=run}
fit$censoring
```

`warfarin` declares no censoring, so nothing here was imputed. On a study where
it is, this is the line to read before trusting the concentrations:
`xgxr::case1_pkpd` has 46% of its concentrations below the limit and 95% of the
lowest dose arm, and there the fitted parameters are substantially a statement
about the draw.

## The covariate effects

```{r covariates, eval=run}
fit$covariate_effects
```

Allometric scaling on clearance and volume, with the standard exponents. It is
asserted rather than tested — `vignette("pmxmodel-algorithm")` says why — and
`covariate_effects = "none"` removes it.

### What is *not* modelled shows up here

```{r correlations, eval=tables}
show(fit$correlations[order(-abs(fit$correlations$correlation)), ],
     "Each covariate against the individual random effects")
```

This is the most important table in the document, and it is a diagnostic rather
than a released quantity. A covariate that moves with a random effect and is not
in the model above is generated **independently** of the profiles, so the
synthetic data carries no relationship between the two. `synpmx_avatar()`
preserves those relationships without modelling them, because a blended
subject's covariates and profile come from the same donors.

The correlations are computed from the individual random effects and only the
correlations are kept, so no per-subject quantity survives into the object.

## The pharmacodynamic shapes

```{r pd, eval=run}
lapply(fit$pd, function(shape) c(shape = shape$pd, round(shape$typical, 3)))
```

```{r pd-candidates, eval=tables}
show(fit$pd[[1L]]$candidates, "The three shapes, compared on AIC")
```

A time course with no exposure term. A PD endpoint driven by concentration is
reproduced as a curve that resembles the average subject's response, which is
adequate for exercising longitudinal code and is not an exposure-response model.

## The covariate distributions

One distribution per covariate per arm, drawn from independently at generation.

```{r covariate-model, eval=run}
str(fit$covariates, max.level = 3)
```

## The arms

```{r arms, eval=run}
data.frame(arm = fit$arms$arms, patients = as.integer(fit$arms$sizes))
```

## The dosing model

Per arm: a planned schedule, a dose ladder, and three discrete-time hazards.
This is the apparatus `synpmx_pca_summarize()` builds, unchanged.

```{r dosing, eval=tables}
arm <- fit$arms$arms[[1L]]
show(fit$dosing[[arm]]$planned, "The planned schedule")
```

```{r rates, eval=run}
dosing <- fit$dosing[[fit$arms$arms[[1L]]]]
c(levels = length(dosing$levels), reduction = dosing$reduction,
  interruption = dosing$interruption, discontinuation = dosing$discontinuation)
```

`warfarin` is a single dose that everybody received, so the ladder has one level
and all three rates are zero. The model then reproduces the planned schedule
exactly, which is the right answer and is what those zeros say. On a study where
patients reduce, skip cycles or come off treatment, these are the numbers that
carry it — and because the schedule is drawn *before* the profile is computed
from it, a synthetic subject who steps down has the lower exposure that implies.

## The visit model

Per endpoint and per retained nominal time, the fraction of the arm holding an
observation there. Attendance is drawn per visit at generation.

```{r visits, eval=tables}
cells <- fit$cells
cells$probability <- as.numeric(fit$visits[[fit$arms$arms[[1L]]]]$probability)
show(cells[, c("endpoint", "time", "probability")],
     "Attendance, per grid cell", paged = TRUE)
```

```{r visits-plot, eval=plots, fig.height=3.5}
library(ggplot2)
library(xgxr)
xgx_theme_set()

ggplot(cells, aes(time, probability, colour = endpoint)) +
  geom_line() + geom_point(size = 1.5) +
  ylim(0, 1) +
  xgx_scale_x_time_units("hours", breaks = seq(0, 120, by = 24)) +
  labs(x = "Nominal time (hours)", y = "Fraction of the arm observed",
       colour = NULL) +
  theme(legend.position = "top")
```

A cell is kept only where at least `min_arm_patients` distinct patients hold an
observation there. A nominal time one patient attended is that patient, and
generating from it would put them back.

## The schema

What the generated table has to look like to be the same study: the columns and
their classes, the compartment numbers, the assay limit per endpoint, and how
identifiers are written.

```{r schema, eval=run}
fit$schema$columns
fit$schema$cmt_dose
unlist(fit$schema$cmt_obs)
fit$schema$censoring
```

## Generating from it

Everything above, and nothing else, is what the next line reads.

```{r generate, eval=run}
synthetic <- synpmx_model_generate(fit, n_subjects = 32, seed = 7)
c(rows = nrow(synthetic),
  subjects = length(unique(synthetic$id)),
  valid = validate_pmx(synthetic, roles)$valid)
```

## Where to go next

- `vignette("pmxmodel-demo")` — this generator run on a study end to end.
- `vignette("pmxmodel-algorithm")` — how each quantity above is arrived at.
- `vignette("pca-fingerprint")` — the same walk for the PCA generator, which
  carries the same apparatus and a completely different description of the
  profiles.
- `vignette("avatar-scorecard")` — the checks that read a generated dataset
  against its source.
