---
title: "Demo: Using synpmx_prior"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx_prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center",
                      out.width = "100%")
library(synpmx)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

One case worked end to end: a synthetic pyrazinamide dataset, 1200 mg orally
once daily for two weeks, sampled on day 14. Every number and every choice is
shown, including the ones the generator cannot represent.

`synpmx_prior()` reads **no data at all**. It takes a public structural model
and a trial design and simulates from those alone, so there is no `data`
argument to supply and nothing here needs a privacy argument. See
[Privacy](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html) for
where this sits among the method families.

The cardinal rule for this mode: every number below must come from somewhere
that is *not* the study being simulated. The values used here are literature
and protocol values for pyrazinamide, recorded in each object's `source` field.
Fit a model to a real study and paste its estimates in here and the real data
reaches the output through the parameters, making the "nothing was read" claim
false. `synpmx_prior()` cannot detect that, so the discipline is yours to keep.

**Read the last section before using the output.** Three parts of the
specification below cannot be expressed, and none of them fails loudly.

## The specification

| Quantity | Value |
|---|---|
| Structural model | 1-compartment oral, first-order absorption and elimination |
| Ka | 1.31 h^-1^ |
| CL/F | 3.52 L/h |
| V/F | 28.57 L |
| IIV (log-SD) | Ka 0.620, CL 0.174, V 0.181 |
| IOV (log-SD) | V 0.117 |
| Residual error | proportional 0.22, additive SD 2.41 mg/L |
| WT, CD38, GENDER | *not generated — see the last section* |
| Regimen | 1200 mg orally once daily |
| Sampling | after 2 weeks of dosing, at 0.3, 0.9, 2.2, 4.5, 8 h post dose |

Choices the specification does not give, made here and flagged so they are easy
to change:

```{r choices}
N_SUBJECTS <- 60          # cohort size
SEED       <- 20260728    # reproducibility seed
LLOQ       <- 0.2         # mg/L, a typical HPLC assay limit for pyrazinamide
DROPOUT    <- 0           # no discontinuation modelled
N_DOSES    <- 14          # "2 weeks of dosing" at once daily
```

## One unit conversion

`pmx_structural_model(iiv = )` takes a **CV**, while the specification gives
**log-SD**. These are different parameterisations of the same lognormal, so the
values are converted rather than pasted: `CV = sqrt(exp(omega^2) - 1)`.

```{r iiv}
omega_to_cv <- function(omega) sqrt(exp(omega^2) - 1)

iiv_log_sd <- c(ka = 0.620, cl = 0.174, v = 0.181)
iiv_cv     <- omega_to_cv(iiv_log_sd)
knitr::kable(
  data.frame(parameter = names(iiv_log_sd),
             log_sd = iiv_log_sd, cv = round(iiv_cv, 4)),
  row.names = FALSE, caption = "IIV: specification (log-SD) to package (CV)"
)
```

## The public inputs

```{r model}
pza_model <- pmx_structural_model(
  pk      = "1cmt_oral",
  typical = c(cl = 3.52, v = 28.57, ka = 1.31),
  iiv     = iiv_cv,
  residual_cv = 0.22,
  source  = paste(
    "Published pyrazinamide population PK parameters; never fitted to the",
    "study being simulated."
  )
)
pza_model
```

Sampling is requested only after two weeks of dosing, so thirteen doses carry
no samples and the fourteenth carries the full profile. `sampling` takes one
entry per dose, and `NULL` means "no samples after this one".

```{r design}
post_dose <- c(0.3, 0.9, 2.2, 4.5, 8)

pza_design <- pmx_trial_design(
  dose_levels   = 1200,
  cohort_sizes  = N_SUBJECTS,
  sampling      = c(rep(list(NULL), N_DOSES - 1L), list(post_dose)),
  n_doses       = N_DOSES,
  dose_interval = 24,
  source        = "Protocol: 1200 mg orally once daily for 2 weeks."
)
pza_design
```

## Naming the output columns

Every generator in the package takes a `pmx_roles()` declaration and writes its
output under those column names. The three that read a study take it as a
description of what they are reading; this one reads nothing, so the same
object does the opposite job here and says which schema to generate into. Pass
the roles of the study this data has to stand in for, and the output drops
straight into code written against it.

```{r roles}
pza_roles <- pmx_roles(
  id = "USUBJID", time = "TIME", nominal_time = "NTIME", tad = "TAD",
  occasion = "OCC", dv = "DV", amt = "AMT", evid = "EVID", cmt = "CMT",
  mdv = "MDV", cens = "CENS"
)
```

A role left undeclared gets no column, which is why there is no `RATE` or
`DOSE` below. `pmx_generated_roles()` is the default and names every column the
generator can produce.

## Generate

```{r generate}
pza <- synpmx_prior(
  pza_model, pza_design, pza_roles,
  n_subjects = N_SUBJECTS, seed = SEED,
  dropout = DROPOUT, lloq = LLOQ
)

str(pza)
```

```{r head}
knitr::kable(head(pza, 8), digits = 3,
             caption = "First rows of the generated dataset")
```

## Does it match the specification?

The checks worth making are the ones on quantities that were actually
requested.

Two time columns matter here. `NTIME` is the **nominal** (protocol) time and
`TAD` is the **actual** time after dose, which the generator jitters around
nominal by `visit_window` (5% by default) so that sample times look like a real
study rather than a grid. Checking the schedule means checking nominal time;
the last dose falls at hour `24 * (N_DOSES - 1)`.

```{r check-design}
obs <- pza$EVID == 0 & !is.na(pza$DV)

# Nominal time after the qualifying dose, recovered from NTIME and the occasion.
nominal_tad <- pza$NTIME - 24 * (pza$OCC - 1)

knitr::kable(
  data.frame(
    quantity = c("subjects", "dose amount (mg)", "doses per subject",
                 "observations per subject", "sampling occasion",
                 "nominal times after dose (h)"),
    value = c(
      length(unique(pza$USUBJID)),
      paste(unique(pza$AMT[pza$AMT > 0]), collapse = ", "),
      round(mean(table(pza$USUBJID[pza$AMT > 0])), 1),
      round(mean(table(pza$USUBJID[obs])), 1),
      paste(sort(unique(pza$OCC[obs])), collapse = ", "),
      paste(sort(unique(round(nominal_tad[obs], 2))), collapse = ", ")
    )
  ),
  row.names = FALSE, caption = "Design realised in the output"
)
```

```{r check-concentrations}
by_time <- split(pza$DV[obs], round(nominal_tad[obs], 2))
knitr::kable(
  data.frame(
    nominal_tad = names(by_time),
    n           = vapply(by_time, length, integer(1)),
    median      = round(vapply(by_time, median, numeric(1)), 2),
    min         = round(vapply(by_time, min, numeric(1)), 2),
    max         = round(vapply(by_time, max, numeric(1)), 2)
  ),
  row.names = FALSE,
  caption = "Concentration (mg/L) by nominal time after dose (h)"
)
```

```{r profiles, eval=has_ggplot2, fig.height=4}
plot_data <- data.frame(
  id  = as.character(pza$USUBJID[obs]),
  tad = pza$TAD[obs],
  dv  = pza$DV[obs]
)
ggplot2::ggplot(plot_data, ggplot2::aes(tad, dv, group = id)) +
  ggplot2::geom_line(alpha = 0.25, colour = "#1B6CA8") +
  ggplot2::geom_point(alpha = 0.35, size = 0.8, colour = "#1B6CA8") +
  ggplot2::labs(
    x = "Time after dose (h)", y = "Concentration (mg/L)",
    title = "Pyrazinamide 1200 mg once daily, day 14"
  ) +
  ggplot2::theme_minimal()
```

## What this does and does not represent

Prior-only generation exists to produce structurally correct data for pipeline
and model-code development. It is not a reimplementation of the source model,
and three parts of the specification have no representation in it. None of them
fails loudly, so they are listed here rather than left to be discovered.

### 1. Covariate effects, which is why no covariates are generated

The requested WT exponents (0.75 on CL/F, 1 on V/F), the GENDER effect on CL/F
(-0.40), and the CD38 effect on CL/F (-0.22) have nowhere to go.

The reason is not a deep one. The generator draws each subject's parameters as
typical value times lognormal IIV and simulates the whole profile from those.
Covariates, when requested, are drawn separately afterwards and merged onto the
finished table, so by the time a weight exists the concentration beside it has
already been computed. No argument on `pmx_structural_model()` links them, and
no code in the package applies a covariate to a parameter.

The generator will happily add the columns if asked, and
`pmx_covariates_reference()` makes that a single call:

```{r covariates}
pmx_covariates_reference(c(WT = "WT", GENDER = "SEX"))
```

Those are plausible numbers bearing **no relationship to the concentrations in
the same row**, and a reader who did not know that would draw a false
conclusion from the first covariate plot they made. A missing column is honest
and a meaningless one is not, so they are left out here. Add them only if your
pipeline needs columns of the right *type* to run against, and say so where
anyone might read them otherwise.

Note what this is *not*. It is not a privacy constraint: prior-only generation
reads nothing, so a covariate effect would be a public assertion exactly like
`typical` and `iiv`, costing no budget. (In the differentially private engines
the calculus differs, since every covariate consumes a slice of the budget.) It
is simply unimplemented. If you need covariate-parameter relationships,
simulate them explicitly in `rxode2` or `nlmixr2`, or use `synpmx_avatar()` on
real data, where real covariate relationships survive because whole real
subjects do.

### 2. Inter-occasion variability is absent

The requested inter-occasion variability on V (log-SD 0.117) is not
represented; the generator draws one parameter vector per subject for the whole
study. With sampling on a single day this is a small omission here, but it
would matter if the schedule were extended across occasions.

### 3. Residual error is proportional only

`residual_cv = 0.22` carries the proportional part. The additive component (SD
2.41 mg/L) has no argument and is dropped, so low concentrations are less noisy
than specified. With sampling ending at 8 h and concentrations staying well
above the additive scale the practical effect is modest, but at a trough it
would not be.

What the output *is* good for: exercising code that reads, reshapes, plots and
fits a repeated-dose PK dataset with a realistic event structure, nominal and
actual times, occasions, and a censoring flag, without a single real patient
being involved.

## Where to go next

- `vignette("prior-algorithm")` — what the generator does with these inputs,
  and the full catalogue of models and designs it can express.
- [Calibration](https://iamstein.github.io/synpmx/articles/synpmx-methods.html)
  — the same public model, with a small privacy budget spent correcting its
  magnitude against a real study.
- [Building the structural model](https://iamstein.github.io/synpmx/articles/model-elicitation.html)
  and [Describing the trial](https://iamstein.github.io/synpmx/articles/data-elicitation.html)
  — how to produce the two public inputs above without reading data.
- [The synpmx data generation algorithms](https://iamstein.github.io/synpmx/articles/synpmx-methods.html)
  — the other generation modes, and which one to use when.
