---
title: "The synpmx_prior Algorithm"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The synpmx_prior Algorithm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

`synpmx_prior()` simulates a dataset from a public structural model and a
public trial design. It reads no confidential data, so there is nothing to
protect and no budget to spend: this is `epsilon = 0`, the strongest guarantee
the package offers, and the weakest fidelity.

**Out of scope, stated once and bluntly.** The output is exactly as good as the
model asserted. It is not an estimate of anything, it contains no information
about any real study, and the catalogue of models it can express is small and
fixed — the section on what it cannot say is the one to read before relying on
the output. If your question is "does my code handle this dataset shape?", this
is the right tool. If it is "what would this study look like under my model?",
write the model in `rxode2` and simulate it there.

Each numbered step below is an operation the function performs, in order.

## Overview of Algorithm

| Step | Operation | Reads data? |
|---|---|---|
| 1 | Validate the public structural model | No |
| 2 | Validate the public trial design | No |
| 3 | Check the output declaration can be filled | No |
| 4 | Assign subjects to cohorts | No |
| 5 | Draw one parameter vector per subject | No |
| 6 | Place dose and observation events on a clock | No |
| 7 | Evaluate the profile by superposition | No |
| 8 | Apply residual error | No |
| 9 | Apply dropout and the assay limit | No |
| 10 | Draw and merge baseline covariates | No |
| 11 | Name the columns from the declaration | No |

Every row reads no data, which is the whole point: there is no step at which a
confidential value could enter.

## Step 1: Validate the Public Structural Model

`pmx_structural_model()` takes a PK form, a named vector of typical parameter
values, an optional PD form, and a required `source`. The `source` is not
decoration: a model without data-independent provenance cannot be treated as a
public input, and the constructor refuses without one.

The catalogue is small and fixed. Rather than restate it here, where it could
drift, these two chunks ask the package:

```{r pk-catalogue, error=TRUE}
pmx_structural_model(pk = "list the options", typical = c(cl = 1, v = 1),
                     source = "illustrative")
```

```{r pd-catalogue, error=TRUE}
pmx_structural_model(pk = "1cmt_iv", typical = c(cl = 1, v = 1),
                     pd = "list the options", source = "illustrative")
```

Every form is a closed-form analytic solution. There are no ODEs: `rx` accepts
an `rxode2` model and warns that it is never used, because silently returning
an analytic curve for a user-supplied ODE model would be a fidelity claim the
package cannot keep.

Each form requires its own parameters, and the error says which:

```{r required-params, error=TRUE}
pmx_structural_model(pk = "2cmt_oral", typical = c(cl = 1), source = "x")
```

Two forms in that list, `1cmt_mixed` and `2cmt_mixed`, are for a study dosed by
two routes. They resolve to a single-route form per dose, and `f` is the
bioavailability applied to the extravascular one. **This mode refuses them**,
because resolving the form needs to know which dose went by which route and a
public trial design has no way to say:

```{r mixed-refusal, error=TRUE}
synpmx_prior(
  pmx_structural_model("1cmt_mixed", c(cl = 2, v = 10, ka = 0.5, f = 0.7),
                       source = "illustrative"),
  pmx_trial_design(100, 30, sampling = c(1, 3, 7), source = "illustrative")
)
```

Before that refusal existed, these forms ran and returned the plain
intravenous profile with `f` discarded, identical at any value of it. Declare
the route the study used, or use `synpmx_model()`, which reads the route from
the data.

Two further inputs are assertions about variability:

- `iiv`, a CV per parameter, applied as lognormal between-subject variability.
  The default is `c(cl = 0.3, v = 0.2)`.
- `residual_cv`, proportional residual error. The default is `0.15`.

## Step 2: Validate the Public Trial Design

`pmx_trial_design()` describes the protocol, and also requires a `source`. Dose
can be given in one of two ways, and never both:

- `dose_levels` with `cohort_sizes`, for parallel groups.
- `dose_escalation`, for within-subject escalation. One sequence applies to
  every subject; a list of sequences gives one per cohort.

`sampling` is where most of the expressive power sits. A bare vector applies
the same profile after every dose. A **list** with one entry per dose says what
each dose carries, and `NULL` means no samples after that one — which is how a
repeated-dose study that samples richly on the first and last days and takes a
single trough between them is written down. `n_doses` and `dose_interval` place
the doses, `duration` makes them infusions, and `visit_window` is the fraction
by which actual sample times deviate from nominal.

```{r design}
rich <- c(0, 0.5, 1, 2, 4, 8, 24)
design <- pmx_trial_design(
  dose_levels = c(100, 300), cohort_sizes = c(12, 12),
  sampling = list(rich, 0, NULL, rich),
  n_doses = 4, dose_interval = 24,
  source = "illustrative protocol"
)
design
```

## Step 3: Check the Output Declaration Can Be Filled

`roles` is a `pmx_roles()` declaration naming the output columns, defaulting to
`pmx_generated_roles()`. Every generator in the package takes one and writes
its output under those names, so a single declaration serves the generator,
`compare_pmx_distributions()` and `synpmx_scorecard()`. The three that read a
study take it as a description of the input; this one reads nothing, so the
same object prescribes the schema to generate into.

Taking a declaration means honouring it, so a declaration naming a column this
generator cannot produce is refused rather than answered with a table that
fails `validate_pmx()`:

```{r roles-refusal, error=TRUE}
model <- pmx_structural_model(pk = "1cmt_oral",
                              typical = c(cl = 6, v = 35, ka = 1.5),
                              source = "illustrative")
synpmx_prior(model, design, pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  covariates = "WT"
))
```

A role left undeclared simply gets no column.

## Step 4: Assign Subjects to Cohorts

Subjects are split across the declared cohorts in proportion to
`cohort_sizes`, with the rounding remainder distributed so the requested total
is exact. A cohort can end up empty if `n_subjects` is smaller than the number
of cohorts.

## Step 5: Draw One Parameter Vector per Subject

Each subject's parameters are the typical values times a lognormal draw:

```
p_i = typical * exp(N(0, sqrt(log(1 + cv^2))))
```

`typical` is the **median**, equivalently the geometric mean, following the
usual population-PK convention for a lognormal parameter. This also has to
match what the calibrated mode estimates: the released correction is a mean on
the log scale, so it targets the geometric mean. Centring the arithmetic mean
instead would leave a systematic `exp(sigma^2 / 2)` gap between what is fitted
and what is generated, which does not shrink with the number of subjects or
with epsilon.

One vector per subject, for the whole study. There is no inter-occasion
variability.

## Step 6: Place Dose and Observation Events on a Clock

Doses go at the times `n_doses` and `dose_interval` imply. Each observation is
assigned to the dose that qualifies it, and its time after that dose is
jittered by `visit_window`.

The jitter is applied to time **since the qualifying dose**, never to absolute
time. That keeps a predose sample exactly predose, keeps every sample inside
its own occasion, and guarantees a nonnegative time after dose. Jittering
absolute time does none of those things, and doing so was a defect (`SIM-005`).
A jittered time is also clipped so it cannot reach the next dose.

## Step 7: Evaluate the Profile by Superposition

Every built-in model is linear in dose, so a multiple-dose profile is the sum
of single-dose profiles from each dose already given. That is exact here rather
than an approximation, and it is also what lets one subject receive two routes.

The consequence is worth seeing, because it is the one fidelity property this
mode does hold exactly:

```{r superposition}
low  <- synpmx_prior(model, pmx_trial_design(
  dose_levels = 10, cohort_sizes = 1, sampling = c(1, 4, 12),
  source = "illustrative"), n_subjects = 1, seed = 1)
high <- synpmx_prior(model, pmx_trial_design(
  dose_levels = 1000, cohort_sizes = 1, sampling = c(1, 4, 12),
  source = "illustrative"), n_subjects = 1, seed = 1)
obs <- low$EVID == 0
round(high$DV[obs] / low$DV[obs], 6)
```

A hundred times the dose is exactly a hundred times the concentration.

## Step 8: Apply Residual Error

Each observation is multiplied by a lognormal draw with CV `residual_cv`.
Proportional only: there is no additive component and no combined error model,
so low concentrations are less noisy than a real assay would make them.

## Step 9: Apply Dropout and the Assay Limit

`dropout` is the fraction of subjects who discontinue early — a protocol
assumption, not something learned. A subject who drops out has their follow-up
truncated at a uniform point, keeping at least two observations.

`lloq` flags observations below the limit with `CENS = 1` and sets the value to
the limit, following the Monolix convention. It applies to the PK endpoint
only: a signed PD endpoint has no lower limit of quantification, and treating
one as censored would report a value above the uncensored ones.

## Step 10: Draw and Merge Baseline Covariates

Declared covariates are drawn once per subject and merged onto the finished
table. `pmx_covariate()` takes a public `range` for clipping plus a `median`
and `cv` for the distribution inside it; `pmx_covariates_reference()` supplies
those for the usual covariates, and `pmx_covariate_reference_table()` says on
what basis.

**The merge happens after the profile is computed, and that is the whole
story.** By the time a weight exists, the concentration beside it has already
been evaluated. Nothing links them.

## Step 11: Name the Columns from the Declaration

The table is built under the names in `pmx_generated_roles()` and renamed to
whatever `roles` declared, keeping the generator's column order. One collision
is permitted, because NONMEM's `CMT` routinely does two jobs: a declaration
naming the same column as both `cmt` and `dvid` takes the compartment column,
whose values already tell the endpoints apart, and drops the separate endpoint
key rather than writing it twice.

## What This Algorithm Cannot Express

The catalogue is a scope decision rather than a gap waiting to be filled. The
space of pharmacometric models is effectively unbounded and every study wants
something idiosyncratic; chasing that would slowly produce a worse `rxode2`,
which already does the job properly.

- **No covariate-parameter relationship.** No allometric exponent on
  clearance, no sex or biomarker effect. Covariates are columns beside the
  data, not inputs to it.
- **No exposure-driven PD.** This one is worth demonstrating, because the
  output looks plausible and is not. The PD forms are functions of time alone,
  so the PD endpoint does not respond to dose:

```{r pd-invariance}
pkpd <- pmx_structural_model(
  pk = "1cmt_oral", pd = "linear",
  typical = c(cl = 6, v = 35, ka = 1.5, baseline = 100, slope = 2),
  source = "illustrative"
)
two_doses <- pmx_trial_design(dose_levels = c(10, 1000),
                              cohort_sizes = c(60, 60),
                              sampling = c(0, 4, 12, 24),
                              source = "illustrative")
generated <- synpmx_prior(pkpd, two_doses, n_subjects = 120, seed = 7)
rows <- generated$EVID == 0
medians <- function(endpoint) {
  keep <- rows & generated$DVID == endpoint
  round(tapply(generated$DV[keep],
               list(dose = generated$DOSE[keep],
                    time = generated$NTIME[keep]),
               stats::median), 2)
}
medians("cp")
medians("pd")
```

  The concentration column separates the two dose groups by a factor of a
  hundred. The PD column does not separate them at all: both arms rise from the
  baseline on the same time course, and the differences are residual error. A
  dose-response plot of this data would show nothing, and an exposure-response
  analysis of it would be meaningless.

- **No inter-occasion variability**, no absorption lag or transit
  compartments, no time-varying parameters or enzyme induction, no additive or
  combined residual error, no ODE model.

## What Leaves the Source Data

Nothing. There is no source data. Every input is a declaration with a
`source` field recording where it came from, and the only way a real study can
reach the output is if someone fits a model to it and pastes the estimates into
`pmx_structural_model()`. The function cannot detect that, and neither can
anything downstream — the `source` field is where that discipline is recorded,
and keeping it honest is the caller's job.

## Where to go next

- `vignette("prior-demo")` — one specification worked end to end, including the
  parts of it that could not be expressed.
- [Calibration](https://iamstein.github.io/synpmx/articles/synpmx-methods.html)
  — the same public model with a small privacy budget spent correcting its
  magnitude against a real study.
- [Building the structural model](https://iamstein.github.io/synpmx/articles/model-elicitation.html)
  and [Describing the trial](https://iamstein.github.io/synpmx/articles/data-elicitation.html)
  — producing the two public inputs without reading data.
- [The synpmx data generation algorithms](https://iamstein.github.io/synpmx/articles/synpmx-methods.html)
  — the other generation modes, and which one to use when.
