---
title: "Evaluating AVATAR on public data"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Evaluating AVATAR on public data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
# `compare_pmx_distributions()` composes its panels with patchwork and refuses
# to draw without it, rather than degrading the way the rest of the plotting
# does. Guarding those chunks keeps a missing suggested package a missing
# figure instead of a failed build.
has_distributions <- has_ggplot2 &&
  requireNamespace("patchwork", quietly = TRUE)
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")
```

This vignette compares eight source datasets against their synthetic
counterparts generated by `synpmx_avatar()`. 

<details>
<summary>Plotting helpers used throughout this vignette</summary>

```{r plot-helpers}
observed_plot_data <- function(data, roles, dataset,
                               clock = "study_time") {
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  if (!is.null(roles$mdv)) {
    observed <- observed & as.character(data[[roles$mdv]]) %in% c("0", "0.0")
  }
  observed <- observed & !is.na(data[[roles$dv]])
  observation_rows <- which(observed)
  occasion <- rep(1L, length(observation_rows))
  tad <- rep(NA_real_, length(observation_rows))
  if (!is.null(roles$occasion)) {
    declared <- suppressWarnings(as.integer(
      data[[roles$occasion]][observation_rows]
    ))
    valid <- !is.na(declared) & declared >= 1L
    occasion[valid] <- declared[valid]
  }
  if (!is.null(roles$tad)) {
    declared <- suppressWarnings(as.numeric(data[[roles$tad]][observation_rows]))
    valid <- is.finite(declared)
    tad[valid] <- pmax(0, declared[valid])
  }
  subject_values <- data[[roles$id]]
  for (id in unique(subject_values[observation_rows])) {
    subject_rows <- which(!is.na(subject_values) & subject_values == id)
    events <- !(as.character(data[[roles$evid]][subject_rows]) %in%
                  c("0", "0.0"))
    if (!is.null(roles$amt)) {
      events <- events & as.numeric(data[[roles$amt]][subject_rows]) > 0
    }
    positions <- which(subject_values[observation_rows] == id)
    event_rows <- subject_rows[events]
    if (length(event_rows) && !is.null(roles$occasion)) {
      event_occasion <- suppressWarnings(as.integer(
        data[[roles$occasion]][event_rows]
      ))
      for (position in positions) {
        candidates <- event_rows[event_occasion == occasion[position]]
        if (length(candidates) && !is.finite(tad[position])) {
          origin <- min(as.numeric(data[[roles$time]][candidates]))
          tad[position] <-
            as.numeric(data[[roles$time]][observation_rows[position]]) - origin
        }
      }
    } else if (length(event_rows)) {
      dose_times <- sort(unique(as.numeric(data[[roles$time]][event_rows])))
      occasion[positions] <- pmax(1L, findInterval(
        as.numeric(data[[roles$time]][observation_rows[positions]]),
        dose_times
      ))
      occasion[positions] <- pmin(occasion[positions], length(dose_times))
      tad[positions] <-
        as.numeric(data[[roles$time]][observation_rows[positions]]) -
        dose_times[occasion[positions]]
    }
  }
  plotted_time <- if (identical(clock, "tad")) tad else
    as.numeric(data[[roles$time]][observation_rows])
  data.frame(
    dataset = factor(dataset, levels = c("Source", "Synthetic")),
    subject = as.character(data[[roles$id]][observation_rows]),
    time = plotted_time,
    dv = as.numeric(data[[roles$dv]][observation_rows]),
    occasion = occasion,
    endpoint = if (is.null(roles$dvid)) "DV" else
      as.character(data[[roles$dvid]][observation_rows]),
    stringsAsFactors = FALSE
  )
}

# Every dataset below gets the same figure: observation rows only, source beside
# synthetic on a shared y axis, one row per endpoint. `clock = "tad"` plots time
# after dose instead of study time, which is the readable view once a study
# doses more than once.
overlay_plot <- function(source, synthetic, roles, title,
                         clock = "study_time",
                         x_label = "Study time (hours)", y_label = "DV",
                         log_y = FALSE, alpha = 0.3) {
  plotted <- rbind(
    observed_plot_data(source, roles, "Source", clock),
    observed_plot_data(synthetic, roles, "Synthetic", clock)
  )
  # One line per patient on a study-time axis, and one line per occasion on a
  # dose-relative one, where the profiles are meant to lie on top of each other.
  plotted$series <- if (identical(clock, "tad")) {
    interaction(plotted$dataset, plotted$subject, plotted$occasion)
  } else {
    interaction(plotted$dataset, plotted$subject)
  }
  figure <- ggplot2::ggplot(
    plotted,
    ggplot2::aes(time, dv, group = series, colour = dataset)
  ) +
    ggplot2::geom_line(alpha = alpha) +
    ggplot2::geom_point(alpha = alpha, size = 0.7) +
    # Endpoint down the side, source and synthetic across. `facet_grid()` frees
    # a scale per ROW, so this orientation is the one that gives each endpoint
    # its own y while holding source and synthetic on a shared one -- which is
    # the only arrangement the eye can compare. The transpose does the opposite
    # on both counts: it lets the two panels drift onto different axes, and it
    # squeezes every endpoint onto one, so a PK concentration reading single
    # digits flattens to a line next to a PD score in the hundreds.
    #
    # A single-endpoint study needs no row strip; it would print the endpoint's
    # name down the side of the only row, beside the axis title already naming
    # the same thing.
    (if (length(unique(plotted$endpoint)) > 1L) {
      ggplot2::facet_grid(endpoint ~ dataset, scales = "free_y",
                          switch = "y")
    } else {
      ggplot2::facet_wrap(~dataset)
    }) +
    ggplot2::scale_colour_manual(values = comparison_colours) +
    ggplot2::labs(x = x_label, y = y_label, colour = "Dataset", title = title) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "none",
                   strip.placement = "outside")
  if (isTRUE(log_y)) figure + ggplot2::scale_y_log10() else figure
}

# Facet rows are endpoints now, so a five-endpoint study needs five times the
# height a one-endpoint study does. Chunks pass this to `fig.height` rather than
# every figure inheriting the one-endpoint default and arriving unreadable.
overlay_height <- function(data, roles, per_endpoint = 2.2, minimum = 3.4) {
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  endpoints <- if (is.null(roles$dvid)) 1L else {
    length(unique(data[[roles$dvid]][observed]))
  }
  max(minimum, per_endpoint * endpoints)
}

# The masking accounting for one run. `pmx_masking_report()` owns the row labels
# and the explanation next to each one, so this vignette and the study reports
# under `scripts_private/` cannot drift apart; only the caption is local.
#
# `section` is what keeps these tables readable. The whole report is thirty-odd
# rows -- the right size to read once after a run, and the wrong size under a
# paragraph making one point about one mechanism, where the five rows it is
# discussing arrive buried in twenty-eight it is not.
masking_table <- function(source, roles, synthetic, label, section = NULL) {
  knitr::kable(
    as.data.frame(pmx_masking_report(synthetic, source, roles,
                                     section = section)),
    row.names = FALSE, align = c("l", "r", "l"),
    caption = paste0(
      "What ", label, "'s run removed",
      if (is.null(section)) ", and what was left to build on." else
        paste0(", on the ",
               paste0("`", section, "`", collapse = " and "),
               " side. `pmx_masking_report()` without `section` prints all of it.")
    )
  )
}
```

</details>

## The datasets

Every dataset here is public package data from `nlmixr2data` or `xgxr`, both
`Suggests` of this package.

<!-- Pandoc sizes these columns in proportion to the dashes in the row below,
     so the last column's width is set by the dashes under it rather than by
     its contents. Keep them uneven. -->

| Dataset | Package | Patients | Rows | Endpoints | What it is here to exercise |
|----------|-----------|-----|-----|-----|--------------------------------------------------------|
| `case1_pkpd` | xgxr | 180 | 20820 | 2 | Six arms, a declared `NOMTIME`, PK and PD, baseline weight, `CENS`. The shape a study report usually arrives in |
| `mad` | xgxr | 60 | 4160 | 5 | Multiple ascending dose carrying ordinal, count and binary PD alongside continuous PD and PK |
| `theo_md` | nlmixr2data | 12 | 348 | 1 | Seven Q24H oral doses with dense profiles on occasions 1 and 7 only. Dosed by weight. |
| `warfarin` | nlmixr2data | 32 | 515 | 2 | One dose, PK and PD running on different time courses, categorical covariates |
| `wbcSim` | nlmixr2data | 45 | 280 | 1 | Infusion start/stop pairs and a delayed WBC response. Follow-up time varies. |
| `nimoData` | nlmixr2data | 12 | 441 | 1 | Ten roughly weekly infusions, recorded at the times they happened |
| `mavoglurant` | nlmixr2data | 120 | 2678 | 1 | One- and two-period crossover, `TIME` resetting within `OCC`, an occasion-varying assigned dose, infusion rows. |
| `pheno_sd` | nlmixr2data | 59 | 744 | 1 | Neonatal phenobarbital, individualised dosing, sparse irregular sampling, time-varying weight. |

**None of these datasets uses steady state, `ADDL`, or `II`.** Every dose is
written out as its own row, so the handling of compressed dose records is
exercised by the package's tests rather than by anything below.

## Shared workflow

Every example follows the same five steps:

1. declare column meanings with `pmx_roles()`;
2. synthesize with `synpmx_avatar()`;
3. plot the real and synthetic data;
4. plot the observation (DV) and baseline covariate distributions, source
   against synthetic
5. report the score card for the synthetic dataset.

The scorecard marks each check `pass`, `FAIL`, or `review`. A `review` row is
one where the pharmacometrician could evaluate whether the output is acceptable for use.  If the synthetic dataset will not be crossing a trust boundary, items marked `review` are acceptable.

**No dataset here fails a check**, which is a thing to be careful about rather
than reassured by. `nimoData`, on B1b, previously failed, and its section
shows what fixing the issue involved. Every card carries five or six `review` rows
instead, and those are where the eight datasets actually differ. The runs below
are quiet: `synpmx_avatar()` prints an alert for every exposure it could not
remove, and eighteen of those blocks would say what the scorecard and
`pmx_masking_report()` say here in tables.

## case1_pkpd: a declared nominal grid

**Start here: this is what a study report usually looks like.** 180 patients
across six treatment arms, with a declared `NOMTIME`, two endpoints keyed by a
character `NAME` column, a baseline weight, and a `CENS` column — the industry
shape, and the one the package is easiest on. `vignette("demo")` runs this
dataset end to end.

Almost everything passes, and the reason is the declared nominal time: the
protocol grid is written down, so coarsening reads it instead of guessing at
one. Keep that in mind through the six datasets after `mad`, where most of the
difficulty turns out to be its absence.

```{r case1, eval=has_xgxr, message=FALSE}
case1_pkpd <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
case1_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB",
  keep = "STUDY"
)
case1_synth <- suppressWarnings(
  synpmx_avatar(case1_pkpd, case1_roles, seed = 808)
)
```

```{r case1-plot, echo=FALSE, eval=has_xgxr && has_ggplot2, fig.height=overlay_height(case1_pkpd, case1_roles)}
overlay_plot(case1_pkpd, case1_synth, case1_roles,
             "case1_pkpd: PK concentration and continuous PD, six arms",
             y_label = "LIDV", alpha = 0.2)
```

```{r case1-distributions, eval=has_xgxr && has_distributions, fig.height=compare_pmx_distributions_height(case1_pkpd, case1_roles)}
compare_pmx_distributions(case1_pkpd, case1_synth, case1_roles)
```

```{r case1-scorecard, eval=has_xgxr}
scorecard <- synpmx_scorecard(case1_pkpd, case1_synth, case1_roles)
synpmx_scorecard_datatable(scorecard)
```

Nothing fails, and C1 passes: all six treatment arms keep their source size. An
avatar never leaves the arm it was anchored in, because `TRTACT` and `DOSE` are
declared as `strata` and are copied from the anchor. The arm *sizes* match
because of a second mechanism: anchors are sampled with replacement, so
`preserve_strata_balance` (the default) gives each stratum its source share
rather than leaving the balance to the draw. Only a declared stratum gets this;
`warfarin` below has a `sex` covariate that is never declared as one, and so has
no equivalent row.

All 180 patients start with an observation schedule nobody else shares and none
ends with one, and the difference from every dataset that follows is one
declared column. With `NOMTIME` present the grid is `nominal` rather than derived, so
coarsening reads the protocol instead of guessing at it, 180 patients collapse
onto six visit sets, none of them rare, and nothing is discarded. Read this next
to `nimoData`, which is the same mechanism with nothing to work from. The gap
between them is not a tuning difference; it is whether the study recorded a
nominal time.

### Declaring `cens` on this study is refused

```{r case1-cens, eval=has_xgxr}
case1_roles_cens <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME", cens = "CENS",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB"
)
validate_pmx(case1_pkpd, case1_roles_cens)$valid
```

That study's `CENS` is meaningful only for the PK endpoint. The PD effect is
signed, so a left-censored PD row reports a value above uncensored ones and the
flag cannot mean what it means for PK. Running `validate_pmx()` on the data you
are about to generate from is how a mis-declared role is caught before it
becomes a generation bug, and this is the dataset where it fires. Leaving `cens`
undeclared, as the run above does, is one right answer; `vignette("demo")` takes
the other and sets `CENS` to 0 on the PD rows before declaring it.

## mad: five endpoints, including ordinal, count and binary

The second `xgxr` set, and the other one shaped like a study report: 60 subjects
in a multiple-ascending-dose study with a declared `NOMTIME`, five observation
endpoints keyed by `NAME` — PK concentration, continuous PD, and ordinal, count
and binary PD. Every other dataset here has one or two endpoints, and an
endpoint silently vanishing during generation is invisible with fewer than
three.

```{r mad, eval=has_xgxr, message=FALSE}
mad <- as.data.frame(get(utils::data(list = "mad", package = "xgxr")))
mad_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", mdv = "MDV", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"),
  covariates = c("WEIGHTB", "SEX")
)
mad_synth <- suppressWarnings(synpmx_avatar(mad, mad_roles, seed = 909))
```

```{r mad-plot, echo=FALSE, eval=has_xgxr && has_ggplot2, fig.height=overlay_height(mad, mad_roles)}
overlay_plot(mad, mad_synth, mad_roles, "mad: five endpoints on one time axis",
             y_label = "LIDV", alpha = 0.2) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 7))
```

```{r mad-distributions, eval=has_xgxr && has_distributions, fig.height=compare_pmx_distributions_height(mad, mad_roles)}
compare_pmx_distributions(mad, mad_synth, mad_roles)
```

```{r mad-scorecard, eval=has_xgxr}
scorecard <- synpmx_scorecard(mad, mad_synth, mad_roles)
synpmx_scorecard_datatable(scorecard)
```

Nothing fails. A3 reads 5 of 5, and it is a set comparison rather than a row
count: row counts stayed plausible in the defect that motivated the check while
an entire compartment disappeared. This is the second declared `NOMTIME` and it
produces the same clean structural result as `case1_pkpd`.

The `SEX` panel carries the same categorical drift as `warfarin`: an even split
in the source, two thirds to one third in the synthetic cohort — 30 and 30
becoming 40 and 20. It is declared as a covariate rather than as `strata`, so
the anchor draw sets it.

This is also the dataset that shows why the figure switches geometry. A panel
holding eight or fewer distinct values is drawn as proportion bars rather than
as a curve, so `PD - Binary` and `PD - Ordinal` come out as bars: a smoothed
density over `{0, 1}` would be a shape this study does not have. `PD - Count`
has more levels than that and stays a curve.

### Discrete endpoints keep their scale

Three of `mad`'s five endpoints are discrete, and `LIDV` is one numeric column
carrying all five, so restoring the *column's* class restores nothing about any
of them. Blending is a weighted mean, and a weighted mean of five donors' zeros
and ones is a number between them; the subject and residual noise terms then
carry it off the level set entirely. Left alone, this study's 0/1 endpoint comes
back as 600 distinct values spanning -0.13 to 1.08.

`synpmx_avatar()` therefore decides what kind of values each endpoint takes,
from the source, and puts the generated values back on that scale. The decision
is worth reading before the data:

```{r mad-endpoint-types, eval=has_xgxr}
pmx_endpoint_types(mad, mad_roles)
```

```{r mad-discrete, eval=has_xgxr}
discrete <- c("PD - Binary", "PD - Count", "PD - Ordinal")
values_taken <- function(data, endpoint) {
  values <- data$LIDV[data$NAME == endpoint & data$EVID == 0]
  sprintf("%d distinct, %.2f to %.2f", length(unique(values)),
          min(values), max(values))
}
knitr::kable(
  data.frame(
    endpoint = discrete,
    source = vapply(discrete, values_taken, character(1), data = mad),
    synthetic = vapply(discrete, values_taken, character(1), data = mad_synth),
    row.names = NULL
  ),
  caption = "Values taken by the three discrete endpoints."
)
```

A6 in the scorecard above is the check on it, and it reads the finished table
rather than trusting the mechanism. It is on every card; the seven studies
before this one have no discrete endpoint, so theirs read
`no discrete endpoint` and pass.

Two limits. The type is inferred, so an endpoint whose values are whole numbers
for a reason other than being discrete is called discrete anyway — `warfarin`'s
`pca` is a percentage recorded without decimals, and its generated values are
rounded to match. `pmx_roles(endpoint_types = )` overrides the inference in
either direction. And snapping a binary endpoint onto its levels means generated
values are source values: a 0/1 endpoint has no third value to emit, so what
protects a patient here is everything in the masking report, never the
distinctness of the number.

## theo_md: seven oral doses, dosed by weight

Twelve subjects on one oral regimen, seven doses 24 hours apart, with dense
concentration sampling around the first and last dose only.

```{r theo, eval=has_nlmixr2data, message=FALSE}
data("theo_md", package = "nlmixr2data")
theo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", covariates = "WT"
)
theo_synth <- suppressWarnings(synpmx_avatar(theo_md, theo_roles, seed = 303))
```

```{r theo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
overlay_plot(theo_md, theo_synth, theo_roles,
             "theo_md: dose-relative concentration profiles",
             clock = "tad", x_label = "Time after dose (hours)", alpha = 0.25)
```

```{r theo-distributions, eval=has_nlmixr2data && has_distributions, fig.height=compare_pmx_distributions_height(theo_md, theo_roles)}
compare_pmx_distributions(theo_md, theo_synth, theo_roles)
```

```{r theo-scorecard, eval=has_nlmixr2data}
scorecard <- synpmx_scorecard(theo_md, theo_synth, theo_roles)
synpmx_scorecard_datatable(scorecard)
```

Nothing fails. Twelve subjects on one dense protocol leave an obvious visit grid
to find, so coarsening takes the twelve unique observation schedules to zero
without a declared `nominal_time`, and the three visit sets that remain are each
held by several subjects.

## warfarin: two endpoints on different time courses

32 subjects, a single dose, a pharmacokinetic endpoint (`cp`) and a
pharmacodynamic one (`pca`) that run on different time courses, and a lower-case
schema with one categorical covariate.

```{r warfarin, eval=has_nlmixr2data, message=FALSE}
data("warfarin", package = "nlmixr2data")
warfarin_roles <- pmx_roles(
  id = "id", time = "time", dv = "dv", amt = "amt", evid = "evid",
  dvid = "dvid", covariates = c("wt", "age", "sex"),
  dose_covariate = "wt"   # 1.5 mg/kg; say so rather than let it be inferred
)
warfarin_synth <- suppressWarnings(
  synpmx_avatar(warfarin, warfarin_roles, seed = 404)
)
```

```{r warfarin-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2, fig.height=overlay_height(warfarin, warfarin_roles)}
overlay_plot(warfarin, warfarin_synth, warfarin_roles,
             "warfarin: PK (cp) and PD (pca) endpoints",
             x_label = "Time (hours)")
```

```{r warfarin-distributions, eval=has_nlmixr2data && has_distributions, fig.height=compare_pmx_distributions_height(warfarin, warfarin_roles)}
compare_pmx_distributions(warfarin, warfarin_synth, warfarin_roles)
```

```{r warfarin-scorecard, eval=has_nlmixr2data}
scorecard <- synpmx_scorecard(warfarin, warfarin_synth, warfarin_roles)
synpmx_scorecard_datatable(scorecard)
```

Nothing fails. Warfarin is dosed at 1.5 mg/kg to within 0.1%, so `wt` is
declared as the `dose_covariate` and each avatar's `amt` is rebuilt from its own
blended weight rather than copied from its anchor's. The ratio here is tight
enough that inference would have found it unaided, but declaring it is the
habit worth having: inference fails closed, and a copied amount under
weight-based dosing discloses one real patient's weight exactly.

## wbcSim: infusions, and patients too extreme to build on

45 subjects with infusion start/stop pairs and a delayed white-blood-cell
decline, nadir and recovery. Follow-up time varies, and a few subjects are
followed far longer than the rest.

```{r wbc, eval=has_nlmixr2data, message=FALSE}
data("wbcSim", package = "nlmixr2data")
wbc_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID", cmt = "CMT"
)
wbc_synth <- suppressWarnings(synpmx_avatar(wbcSim, wbc_roles, seed = 505))
```

```{r wbc-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
overlay_plot(wbcSim, wbc_synth, wbc_roles,
             "wbcSim: delayed decline, nadir, and recovery",
             y_label = "WBC")
```

```{r wbc-distributions, eval=has_nlmixr2data && has_distributions, fig.height=compare_pmx_distributions_height(wbcSim, wbc_roles)}
compare_pmx_distributions(wbcSim, wbc_synth, wbc_roles)
```

```{r wbc-scorecard, eval=has_nlmixr2data}
scorecard <- synpmx_scorecard(wbcSim, wbc_synth, wbc_roles)
synpmx_scorecard_datatable(scorecard)
```

Nothing fails, and C2 is the row to read: 1 of the 4 source dose regimens is
represented in the output.

### Three re-dosed patients are not in the synthetic cohort

```{r wbc-masking, eval=has_nlmixr2data}
masking_table(wbcSim, wbc_roles, wbc_synth, "wbcSim",
              section = "dose_schedules")
```

Forty-two of the 45 subjects have a single infusion at time 0. The other three
were re-dosed on a schedule nobody else shares, so no avatar can be built on
them without pointing at them, and every synthetic subject carries the one
shared regimen. Doses per patient falls from 1.2 to 1.0 in the A5b row. The three
patients are not removed and still act as donors, but that part of the study
design is not in the output, and C2 is `review` rather than `pass` for exactly
this reason.

Two of those three are also the only patients in this vignette that the anchor
screen removes: they are followed far enough past the rest to exceed twice the
cohort's 90th percentile, so `synpmx_avatar()` builds no avatar on them
(`screen = TRUE`, the default). `section = "anchors"` prints that as a pair:
**43 anchors available, 45 avatars built.** Screening removes a subject
from the pool avatars are drawn from, not from the cohort. The remaining 43 are
sampled with replacement to fill the same 45 slots, and the screened subjects
still contribute measurements as donors. What is lost is the possibility of an
avatar with their distinctive follow-up length. `screen = FALSE` keeps every
subject in the pool.

Dose amounts are not recomputed: `wbcSim` declares no covariates, so there is
nothing to test the amounts against.

## nimoData: ten infusions recorded at actual times

Twelve subjects, ten roughly weekly infusions each, with declared occasion
(`OCC`) and time after dose (`TAD`). The nominal dose group `DOS` is carried
through with `keep`, which copies it from the same subject that supplied the
doses so that it stays coherent with them. The redundant `WGT` column is left
undeclared and is dropped.

```{r nimo, eval=has_nlmixr2data, message=FALSE}
data("nimoData", package = "nlmixr2data")
nimo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
  covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
)
nimo_synth <- suppressWarnings(synpmx_avatar(nimoData, nimo_roles, seed = 606))
```

```{r nimo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
overlay_plot(nimoData, nimo_synth, nimo_roles,
             "nimoData: ten infusions over a long follow-up",
             alpha = 0.2)
```

```{r nimo-distributions, eval=has_nlmixr2data && has_distributions, fig.height=compare_pmx_distributions_height(nimoData, nimo_roles)}
compare_pmx_distributions(nimoData, nimo_synth, nimo_roles)
```

```{r nimo-scorecard, eval=has_nlmixr2data}
scorecard <- synpmx_scorecard(nimoData, nimo_synth, nimo_roles)
synpmx_scorecard_datatable(scorecard)
```

**Nothing fails, and the dataset is still not shippable.** A5b is the row to
read: doses per patient falls from 10 to 1.6, nowhere near the 5% that would
pass it. Ten weekly infusions go in and between one and two come out, because
the only openings any two subjects share are the first dose and occasionally the
second. C2 agrees from the other side — 7 of the 12 source regimens are
represented, none of them at full length. This is the case where fourteen passes
are worth less than one `review` row.

### The inferred grid achieves nothing here

```{r nimo-masking, eval=has_nlmixr2data}
masking_table(nimoData, nimo_roles, nimo_synth, "nimoData",
              section = "visits")
```

12 of 12 subjects are unique before coarsening, 12 of 12 after, and every one of
them on a one-off observation time. Ten roughly weekly infusions over a long
follow-up means no two subjects were ever dosed or sampled close enough together
for an inferred grid to merge them, so every subject is observed at moments that
are theirs alone.

`section = "visit_sets"` shows the same thing from the other side. Twelve
subjects hold twelve distinct visit sets, two of them held by a single patient and discarded,
and **every avatar's arrangement was built from a shape rather than reused,
100%.** The shape is how many visits were missed and whether the misses were
terminal, contiguous, or scattered; which specific visits each patient missed
does not survive.

The dosing is cut back rather than copied, because dose events are copied from
the anchor verbatim and moving a dose is not a protocol-valid edit the way
moving a sample is. Truncating one is. When ten infusions are recorded at actual
times — 165.70, 167.24 and 168.05 for what the protocol called one weekly visit —
no two subjects share an opening past the first dose or two, so that is where
every avatar's course stops. Nothing is disclosed and almost nothing is left.

### What that does to the record

One row per subject, a grey tick for each dose and a coloured dot for each
observation. This is the figure to read before the scorecard:

```{r nimo-schedule, eval=has_nlmixr2data, fig.height=5}
plot_pmx_schedule(nimoData, nimo_roles, main = "nimoData, source")
plot_pmx_schedule(nimo_synth, nimo_roles, main = "nimoData, synthetic")
```

The source has a dose tick standing in front of every cluster of samples, right
across the follow-up. The synthetic cohort has one or two ticks at the far left
and then nothing, while the samples carry on to 2100 h regardless — **the same
follow-up with the exposure taken out from under it.**

The overlay at the top of this section shows what that does to the
concentrations. The source's sawtooth — a peak after each infusion, a decline,
then the next one — is absent from the synthetic panel, because there are no
repeated infusions left to produce it. Both panels span the same follow-up:
maximum `TIME` is 2135 h against the source's 2252 h. What has changed is
everything underneath, and time after dose is where it shows up as a number:

```{r nimo-tad, eval=has_nlmixr2data}
after_last_dose <- function(data) {
  observed <- data[data$EVID == 0, ]
  last <- tapply(data$TIME[data$EVID != 0], data$ID[data$EVID != 0], max)
  c(median_tad = round(median(observed$TAD, na.rm = TRUE), 1),
    past_last_dose = sum(observed$TIME > last[as.character(observed$ID)]),
    observations = nrow(observed))
}
rbind(source = after_last_dose(nimoData), synthetic = after_last_dose(nimo_synth))
```

Median time after dose goes from about four days to about thirty-nine, and
almost every sample ends up with no dose behind it at all. A truncation is the
right edit for a patient who stopped early, and follow-up continuing past the
last dose is then real; here the stopping point is an artefact of masking, and
nothing in the mechanism asks whether the observation record still has exposure
supporting it.

### Constructing a nominal time

All three findings have one cause: there is no protocol grid in this dataset,
only the times things happened to occur at. The design is ten weekly infusions
with a declared occasion and time after dose, so the grid can be written down —
the occasion number times the nominal interval, plus the visit's nominal time
after dose.

Writing it down is a statement about the protocol, and it takes one decision the
arithmetic will not make for you. Rounding time after dose to the nearest day
sends a trough drawn just before the next infusion — `TAD` near 168 h, and 98 of
the 321 observations here are one — onto that infusion's own nominal time, where
it stops being a trough and becomes a time-zero sample. Give the pre-dose sample
a slot of its own instead:

```{r nimo-nominal, eval=has_nlmixr2data, message=FALSE}
nimo_interval <- 168                     # the protocol's weekly infusion
nimo_nominal <- nimoData
last_occasion <- ave(nimo_nominal$OCC, nimo_nominal$ID, FUN = max)

nominal_tad <- round(nimo_nominal$TAD / 24) * 24
# A sample at the end of its interval is the NEXT infusion's pre-dose trough,
# not that infusion's time-zero sample. Only where another infusion follows:
# after the last one, a long time after dose is real follow-up.
pre_dose <- nimo_nominal$EVID == 0 & nominal_tad >= nimo_interval &
  nimo_nominal$OCC < last_occasion
nominal_tad[pre_dose] <- nimo_interval - 1

nimo_nominal$NTIME <- (nimo_nominal$OCC - 1) * nimo_interval +
  ifelse(nimo_nominal$EVID == 0, nominal_tad, 0)
```

`TIME` and `NTIME` side by side, for the first two occasions of one subject.
The recorded times wander; the nominal ones do not, and that is the whole of
what the declaration buys:

```{r nimo-nominal-columns, eval=has_nlmixr2data}
shown <- nimo_nominal[nimo_nominal$ID == nimo_nominal$ID[1] &
                        nimo_nominal$OCC <= 2,
                      c("ID", "OCC", "EVID", "TIME", "TAD", "NTIME")]
knitr::kable(shown, row.names = FALSE, digits = 2)
```

The two infusions were given at 0 and 167.20 h and land on a nominal 0 and 168.
The row to look at is the sample at 167.16 h, four minutes *before* that second
infusion and still carrying occasion 1: it is the trough. It keeps a nominal
time of 167, an hour ahead of the infusion it precedes, instead of collapsing
onto it and being recorded as that infusion's time-zero sample.

```{r nimo-nominal-run, eval=has_nlmixr2data, message=FALSE}
nimo_roles_nominal <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
  nominal_time = "NTIME",
  covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
)
nimo_fixed <- suppressWarnings(
  synpmx_avatar(nimo_nominal, nimo_roles_nominal, seed = 606)
)
scorecard <- synpmx_scorecard(nimo_nominal, nimo_fixed, nimo_roles_nominal)
synpmx_scorecard_datatable(scorecard[scorecard$check %in% c("A5a", "A5b"), ])
```

A5a and A5b are the two rows this construction was for, so only they are shown;
`scorecard` still holds every row. The same figure as before, drawn on the new
cohort:

```{r nimo-fixed-schedule, eval=has_nlmixr2data, fig.height=5}
plot_pmx_schedule(nimo_fixed, nimo_roles_nominal,
                  main = "nimoData, synthetic with a declared NTIME")
```

Every avatar carries the full course of ten infusions, each with its cluster of
samples behind it — the figure the source gives, rather than the two ticks and
2100 h of unsupported follow-up above.

**The dosing comes back in full.** A5b goes from `10 -> 1.6` to `10 -> 10` and
C2 from 7 of 12 to 1 of 1, because on the nominal grid the twelve dose schedules
become one schedule that all twelve subjects share — so there is nothing left to
truncate, and no avatar's course has to stop early. Unique observation schedules
fall from 12 to 5, the twelve visit sets collapse to eight, real sets become
reusable, and the invented-arrangement share falls from 100% to 25%. Two visit
sets are discarded either way, which is why the discard count should never be
read on its own.

**Neither card fails, and only one of them is usable.** A5b is what separates
them: `review` on the recorded times, `pass` on the constructed grid, while the
privacy guarantee is identical on both sides. Stopping at the B rows would have
made the two cards look the same. The samples get their exposure
back with the doses: 61 of 326 observations sit after the last dose, against 59
of 321 in the source, where before it was 317 of 331. Median time after dose
comes back to 72 h against the source's 96 h, where the recorded-time run gave
935 h — and it is the pre-dose slot above that buys most of that. Rounding the
trough onto its next infusion instead leaves the median at 0 and doubles the
share of time-zero samples, which is the A1 gap in
`vignette("scorecard-synthetic-data-checks")` reached by a construction rather
than by the generator.

What the package does not yet do is *notice* the first case: no check asks
whether an avatar's observations still have dosing supporting them, so that card
passes every row while describing a study nobody ran.

### Which path to try

There is no single answer, and the order below is what the datasets in this
vignette suggest rather than a rule.

1. **Declare a `nominal_time` that already exists.** `case1_pkpd` and `mad` ship
   with `NOMTIME`, and both reach zero exposure with nothing else done. If your
   study has the column, this is the whole job.
2. **Construct one, as above.** Available whenever the protocol can be written
   down from what the dataset records — here an occasion number and a time after
   dose. It is a statement about the design, so it belongs in the dataset where a
   reader can check it, not inside the generator where they cannot. The trough
   decision above is why: the arithmetic has a choice in it that only somebody
   who knows the study can make.
3. **Fall back to the inferred grid.** With no `nominal_time`, `coarsen_time`
   (on by default) derives one from the recorded times. This is what `pheno_sd`
   runs on, and it does real work there — but it is a guess at a grid, and on
   `nimoData` it merges nothing at all, which is the whole of this section.
4. **Accept the truncation, and read A5b.** Where dosing is genuinely
   individualised there is no protocol grid to recover, because there was no
   protocol. `pheno_sd` is that case: routine neonatal care, no occasions, and
   nothing to construct. The dosing is then shortened to what several patients
   share, and how much survives is a number to judge rather than a bug to fix.

## mavoglurant: an occasion-reset clock

120 subjects in one- and two-period profiles, with `TIME` resetting within
`OCC`, an occasion-varying assigned `DOSE` carried through with `keep`,
numeric-coded `SEX`, and infusion rows. The reset clock validates within ID and
occasion.

```{r mavo, eval=has_nlmixr2data, message=FALSE}
data("mavoglurant", package = "nlmixr2data")
mavo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  cmt = "CMT", rate = "RATE", mdv = "MDV", occasion = "OCC",
  keep = "DOSE", covariates = c("AGE", "SEX", "WT", "HT")
)
mavo_synth <- suppressWarnings(
  synpmx_avatar(mavoglurant, mavo_roles, seed = 707)
)
```

```{r mavo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
overlay_plot(mavoglurant, mavo_synth, mavo_roles,
             "mavoglurant: dose-relative profiles with occasion reset",
             clock = "tad", x_label = "Time after dose (hours)",
             y_label = "DV (log scale)", log_y = TRUE, alpha = 0.2)
```

```{r mavo-distributions, eval=has_nlmixr2data && has_distributions, fig.height=compare_pmx_distributions_height(mavoglurant, mavo_roles)}
compare_pmx_distributions(mavoglurant, mavo_synth, mavo_roles)
```

```{r mavo-scorecard, eval=has_nlmixr2data}
scorecard <- synpmx_scorecard(mavoglurant, mavo_synth, mavo_roles)
synpmx_scorecard_datatable(scorecard)
```

Nothing fails, and B2 is the row this study used to break. Recorded follow-up
length is bimodal here, because `TIME` restarts within `OCC` and a subject
therefore reads as either about 24 hours or about 36 to 48. Scored on a robust z
against one pooled tolerance — no `strata` are declared, so the cohort is the
group — subjects at both ends came back flagged, 41 of 120 of them. B2 now
requires a patient to be separated from the nearest other patient as well, and
two clusters of sixty single nobody out, so it reads 0. The study's shape is
still bimodal; it was never a property of the synthetic data.

### The largest residual in the vignette

```{r mavo-masking, eval=has_nlmixr2data}
masking_table(mavoglurant, mavo_roles, mavo_synth, "mavoglurant",
              section = "visits")
```

Of the 64 patients still unique after coarsening, eleven have a one-off
observation time and the other 53 share every observation time with somebody,
differing only in which of those visits they attended. Those 53 are the largest
absolute residual here, and no grid setting moves them. This is what the two
sub-rows are for: they separate a problem that declaring `nominal_time` fixes
from one that can only be dropped, remediated, or accepted.

`section = "visit_sets"` is the other half of the picture. Mavoglurant has by far
the most distinct visit sets of any dataset here, 73, and only two of them are held
by a single patient, so only those two are discarded and no arrangement had to
be invented at all. For every one of the 120 avatars a set that several real
patients share was available to reuse. A large number of distinct visit sets is
not by itself a problem; a large number of singleton sets is.

## pheno_sd: individualised dosing in routine care

59 neonates given phenobarbital for seizure prevention, from routine clinical
care rather than from a protocol. Dosing is individualised to the infant,
sampling is sparse and irregular, and weight varies over time. It is the least
protocol-like dataset here.

```{r pheno, eval=has_nlmixr2data, message=FALSE}
data("pheno_sd", package = "nlmixr2data")
pheno_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  covariates = c("WT", "APGR")
)
pheno_synth <- suppressWarnings(
  synpmx_avatar(pheno_sd, pheno_roles, seed = 1010)
)
```

```{r pheno-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
overlay_plot(pheno_sd, pheno_synth, pheno_roles,
             "pheno_sd: sparse, irregular, individualised",
             y_label = "Phenobarbital (ug/mL)")
```

```{r pheno-distributions, eval=has_nlmixr2data && has_distributions, fig.height=compare_pmx_distributions_height(pheno_sd, pheno_roles)}
compare_pmx_distributions(pheno_sd, pheno_synth, pheno_roles)
```

```{r pheno-scorecard, eval=has_nlmixr2data}
scorecard <- synpmx_scorecard(pheno_sd, pheno_synth, pheno_roles)
synpmx_scorecard_datatable(scorecard)
```

**Nothing fails, and the two rows that moved say what reaching that cost.**
Doses per patient falls from 10 to 5.6 in A5b, and 35 of the 56 source dose
regimens are represented in C2. B1b passing here is not each infant's full
course being masked; it is each course being cut short.

### Dosing that can only be masked by shortening it

```{r pheno-masking, eval=has_nlmixr2data}
masking_table(pheno_sd, pheno_roles, pheno_synth, "pheno_sd",
              section = "dose_schedules")
```

One row per infant, a grey tick for each dose and a coloured dot for each
observation. The dose ticks are the story: a forest of them across the whole
follow-up in the source, and in the output the same forest for the first few
days, thinning out and stopping well before the source does.

```{r pheno-schedule, eval=has_nlmixr2data, fig.height=7}
plot_pmx_schedule(pheno_sd, pheno_roles, main = "pheno_sd, source")
plot_pmx_schedule(pheno_synth, pheno_roles, main = "pheno_sd, synthetic")
```

The tail of each course is what this study loses. On the observation side the
mechanism works: B1a holds at 0, so no avatar has a set of attended visits that
belongs to one real infant. On the dose side there is almost no *complete*
schedule safe to copy — the source holds 56 distinct dose schedules across 59
infants — but that is not the only lever. Most infants **open** the same way,
twelve-hourly from time zero, and an opening several infants share can be given
to an avatar as long as no single infant stopped exactly there. So each avatar's
dosing is truncated back to the deepest such opening, which on this study is
about half the course: ten doses a patient becomes 5.6, over 35 of the 56
regimens.

What is lost is specifically the divergent tail — the part of a course where an
individual clinician's decisions have separated one infant from every other, and
therefore exactly the part that identifies them. Dose events are copied from the
anchor verbatim, so there is no way to keep it. Coarsening does not help either:
it acts on the visit grid, while the exposure here is *which days this infant
was dosed*.

A study with this shape accepts a synthetic cohort whose dosing is a shortened
version of the real one, and has to decide whether that is enough for the
purpose. `pheno_sd` declares no arms;
where a study does, the same shape confined to one arm is reported by arm —
`unmaskable_strata()` before generating, and rows B1a and B1b afterwards, which
name the arm that could not be masked. An avatar is never anchored outside the
arm it was allocated to, so an arm whose dosing is individualised fails rather
than borrowing patients from an arm whose dosing is not.

## How well did the obfuscation work?

Everything above shows that the synthetic data *looks* right. This section asks
the other question: how much of each real patient is still visible in it?

### Exposure across the eight datasets

Each dataset above carries its own accounting. This section puts the same
quantities in one place, because the *contrast* between datasets is what shows
how much the answer depends on study design rather than on the package.

```{r exposure-table, eval=has_nlmixr2data && has_xgxr}
exposure_row <- function(label, source, roles, synthetic) {
  before <- skeleton_uniqueness(source, roles)
  after <- attr(synthetic, "pmx_settings")
  data.frame(
    Dataset = label,
    Subjects = nrow(before),
    `Unique before` = attr(before, "n_unique_schedule"),
    `Unique after` = after$unique_schedule_n,
    `Unique observation time` = after$unique_obs_time_n,
    `Unique visit set` = after$unique_visit_set_n,
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

exposure <- rbind(
  exposure_row("case1_pkpd", case1_pkpd, case1_roles, case1_synth),
  exposure_row("mad", mad, mad_roles, mad_synth),
  exposure_row("theo_md", theo_md, theo_roles, theo_synth),
  exposure_row("warfarin", warfarin, warfarin_roles, warfarin_synth),
  exposure_row("wbcSim", wbcSim, wbc_roles, wbc_synth),
  exposure_row("nimoData", nimoData, nimo_roles, nimo_synth),
  exposure_row("mavoglurant", mavoglurant, mavo_roles, mavo_synth),
  exposure_row("pheno_sd", pheno_sd, pheno_roles, pheno_synth)
)
knitr::kable(
  exposure,
  caption = paste(
    "Source subjects holding a visit schedule no other subject shares,",
    "before and after time coarsening. The last two columns split the",
    "'after' count by cause."
  )
)
```

Three datasets reach zero and five do not, and the split is not about cohort
size. `case1_pkpd` (180 patients) and `mad` (60) reach it because they declare
`NOMTIME`, so coarsening reads the protocol grid instead of inferring one;
`theo_md` reaches it on an inferred grid because twelve subjects on one dense
protocol leave an obvious grid to find. Everything else is a study whose
recorded times do not collapse, and `nimoData` and `pheno_sd` are the extreme
cases — real studies where the times are genuinely per-patient.

### What the masking cost

Exposure is only half the story. The mechanisms that reduce it do so by
*removing* things, and the removals are worth seeing next to the numbers above.

```{r exposure-cost, eval=has_nlmixr2data && has_xgxr}
cost_row <- function(label, source, roles, synthetic) {
  settings <- attr(synthetic, "pmx_settings")
  data.frame(
    Dataset = label,
    Patients = settings$source_subjects,
    `Screened out` = settings$anchors_screened_out,
    `Below donor floor` = settings$anchors_route_excluded,
    `Anchors left` = settings$anchors_available,
    `Patterns` = settings$patterns_total,
    `Patterns lost` = settings$patterns_dropped,
    `Patients affected` = settings$subjects_with_dropped_pattern,
    `Dose basis` = ifelse(is.na(settings$dose_basis), "—", settings$dose_basis),
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

knitr::kable(
  rbind(
    cost_row("case1_pkpd", case1_pkpd, case1_roles, case1_synth),
    cost_row("mad", mad, mad_roles, mad_synth),
    cost_row("theo_md", theo_md, theo_roles, theo_synth),
    cost_row("warfarin", warfarin, warfarin_roles, warfarin_synth),
    cost_row("wbcSim", wbcSim, wbc_roles, wbc_synth),
    cost_row("nimoData", nimoData, nimo_roles, nimo_synth),
    cost_row("mavoglurant", mavoglurant, mavo_roles, mavo_synth),
    cost_row("pheno_sd", pheno_sd, pheno_roles, pheno_synth)
  ),
  caption = paste(
    "What the masking removed. The first three counts are patients excluded",
    "from the anchor pool; `Patterns` counts distinct visit sets in",
    "the source and `Patterns lost` those too rare to be reused."
  )
)
```

Three columns in this table need reading with care.

**Only `wbcSim` loses any patient from the anchor pool, and the cohort is still
full size.** Screening removes a subject from the pool avatars are *drawn from*,
not from the cohort: the remaining anchors are sampled with replacement to fill
every slot, and the screened subjects still contribute measurements as donors.
An exclusion costs coverage of a structure, never sample size.

**`Patterns lost` equals `Patients affected` in every row.** That is an identity,
not a coincidence. At the default floor of 2, a pattern is discarded exactly when
*fewer than two* patients share it — that is, when exactly one does. So every
discarded pattern has precisely one holder and the two columns must agree. They
only diverge at a floor of 3 or more, where a pattern held by two patients is
also dropped.

**`Patterns lost` says nothing about how well a dataset did.** `mavoglurant`
loses 2 of 73 and `nimoData` loses 2 of 12, and they are not comparable results:
mavoglurant reused a real visit set for every avatar, while nimoData could reuse
none and invented all twelve from a shape. The discard count has to be read next
to `pattern_generated_fraction`, which `pmx_masking_report()` prints as
"misses placed fresh" and this table does not.

### Is a "pattern" too strictly defined?

Reasonably asked, and the answer is yes, deliberately. A pattern here is the
**exact** set of endpoint-and-time pairs a subject was observed at. Two patients
who each missed exactly one visit have *different* patterns if they missed
different visits. Every pattern held by exactly one patient is discarded, and how
many that is depends entirely on the study: 4 of warfarin's 14 after coarsening,
5 of wbcSim's 25, but only 2 of mavoglurant's 73.

The strictness is what makes the guarantee exact: reusing a pattern that one real
patient holds would reproduce that patient's schedule, so nothing short of exact
matching would support the claim. The cost is that the definition cannot see that
"missed one visit, early" and "missed one visit, late" are the same *kind* of
event.

That is why the draw is two-stage. A **shape** is chosen first — how many visits
were missed, and whether the misses were terminal (at the end of the record),
contiguous (an
interruption), or scattered — and both of those patients share one. Within the
shape a real pattern is reused if one clears the floor; only otherwise is an
arrangement generated, and a generated one is rejected and redrawn if it happens
to land on a pattern too rare to have been reusable.

What the fallback rescues is the *avatar*, not the pattern count. A discarded
pattern stays discarded; what the shape gives is a legal set of visits to hand
the avatar that would otherwise have worn it. `pmx_masking_report()` shows the price in its
"misses placed fresh" row, and it varies as much as everything else here: `nimoData` pays it in full at 100%, `pheno_sd` at 42%, `warfarin` at 22%,
`wbcSim` at 2%, and `mavoglurant`, `theo_md`, `case1_pkpd` and `mad` not at all.

What remains lost is *resolution*: how much missingness there was and what kind
survive; which specific visits each patient missed does not.

### How close the values landed

Everything above concerns *structure*. Blending is the mechanism that protects
the **values**, and `compare_pmx_proximity()` is its measurement: it asks whether
each subject's nearest neighbour lies in its own dataset or the other one. Near
0.5 means a synthetic subject is no more like a real subject than one real
subject is like another, which is the target; toward 0 means memorisation.

```{r proximity, eval=has_nlmixr2data && has_xgxr}
proximity_row <- function(label, source, synthetic, roles) {
  report <- compare_pmx_proximity(source, synthetic, roles, replicates = 30)
  data.frame(
    Dataset = label,
    `Adversarial accuracy` = round(report$adversarial_accuracy, 3),
    `Null lower` = round(report$null_lower, 3),
    `Null upper` = round(report$null_upper, 3),
    `Per side` = report$n_compared,
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

knitr::kable(
  rbind(
    proximity_row("case1_pkpd", case1_pkpd, case1_synth, case1_roles),
    proximity_row("mad", mad, mad_synth, mad_roles),
    proximity_row("theo_md", theo_md, theo_synth, theo_roles),
    proximity_row("warfarin", warfarin, warfarin_synth, warfarin_roles),
    proximity_row("wbcSim", wbcSim, wbc_synth, wbc_roles),
    proximity_row("nimoData", nimoData, nimo_synth, nimo_roles),
    proximity_row("mavoglurant", mavoglurant, mavo_synth, mavo_roles),
    proximity_row("pheno_sd", pheno_sd, pheno_synth, pheno_roles)
  ),
  caption = paste(
    "Nearest-neighbour adversarial accuracy against a split-half null built",
    "from the source cohort itself. 0.5 is the target."
  )
)
```

Read the null intervals before the point estimates. They are **wide** — these
are cohorts of a few dozen, and the statistic is built from nearest-neighbour
comparisons that are noisy at that size. A value inside the interval means
*nothing was detected*, not that nothing is there. What it would catch is a
blatant leak: a synthetic subject sitting on top of a real one drives the
statistic to zero, which the package's own regression test confirms by handing
the function a verbatim copy and requiring it to object.

### Reading the tables across datasets

**The unique-schedule count after coarsening is the number of patients you would
consider dropping**, and the split by cause decides whether dropping is even the
right response. Each dataset
section above works through its own numbers; the pattern across all eight is what
matters here.

The spread runs from **`case1_pkpd`, where 180 patients on a declared nominal
grid go to 0 and the mechanism has nothing left to do**, through **`mavoglurant`,
where the grid does most of what it can and still leaves 53 of 120 patients
unique on their visit set**, to **`nimoData`, where the grid achieves nothing at all**,
and finally to **`pheno_sd`, where the observation side is clean and the *dosing*
survives only as far as the shared opening of each course**. Same package, same defaults, same seed discipline —
the difference is entirely in how the studies were designed and how their times
were recorded.

That is the point of reporting these per dataset rather than quoting a headline
number. There is no "synpmx removes X% of the exposure." There is only what it
removed from *your* study, which is why every run reports it and why the sequence
below starts with a question about your data rather than about a setting.

### What the other mechanisms leave behind

One residual belongs to a mechanism these tables do not have a column for.

**Dose sequence.** The dose *amount* is handled: under dosing proportional to a
baseline covariate, each avatar's amount is recomputed from its own blended
covariate, as the masking reports above show. What is still copied is the
*sequence of levels* an anchor climbed. Where escalation is outcome-adaptive —
driven by a subject's own tolerability — that sequence encodes the subject's
response, and nothing here touches it. No public dataset in this vignette has
that shape, so it does not appear in any table above; a dose-escalation oncology
study would.

**Which visits were attended.** The unique-visit-set row above is what
`min_pattern_share` (default 2) addresses: an avatar's set of attended visits is
drawn from patterns at least that many source subjects share, so no synthetic
patient carries a schedule unique to a real one. No subject is removed from the
cohort to achieve it — a patient with a rare pattern still contributes
measurements as a donor.

What *is* lost is the rare patterns themselves. They are discarded rather than
approximated, so those specific patterns of missing visits and dose
interruptions will not
appear in the output. The loss is a handful of patterns on every dataset here —
two each on `nimoData` and `mavoglurant`, three on `pheno_sd`, four on
`warfarin`, five on `wbcSim` — and it is zero wherever coarsening has already
made the patterns common, on `theo_md` from an inferred grid and on `case1_pkpd`
and `mad` from a declared one.

Every run reports the figures, and `pmx_masking_report()` prints them:
`patterns_total`, `patterns_dropped`, and `subjects_with_dropped_pattern` in the
settings, plus a loud alert. Read them together with
`pattern_generated_fraction`, because a small discard count can be bought with a
large share of invented arrangements — `nimoData` discards only two patterns and
invents the arrangement for every single avatar.

### What to do about it

Dropping 53 of 120 subjects is a real cost, and it is not automatic. Under this
package's governance model — AVATAR output stays inside the source data's own
access controls — a unique schedule is a reason for care, not a blocker. The
sequence worth following is:

1. **Declare `nominal_time` if the study has it.** This is free, and it is the
   only thing that helps the unique-moment column. A study without a
   nominal-time column can often still construct one from an occasion number and
   a time after dose, which the `nimoData` section does: it takes that dataset
   from 1.6 doses per patient back to the full ten.
2. **Then re-run this table.** If the unique-moment count is at or near zero,
   the grid has done its whole job and no amount of tuning will improve it.
3. **Then decide about the pattern column.** These are missing-visit
   patterns. `flag_identifiable_subjects()` and
   `remediate_identifiable_subjects()` can drop or truncate them; whether that
   is worth the lost cohort size is a judgment about who will see the output.

Two limits to keep in view. This counts *schedules*, not dose amounts — a study
with weight-based dosing or per-subject titration leaves subjects unique on dose
regardless of what the grid does, which `skeleton_uniqueness()` reports
separately as `n_share_dosing`. And none of this bounds what an adversary
learns; it reduces the ways a real patient can be singled out without limiting
how often that succeeds. That is what the differentially private modes are for.

## What is preserved in AVATAR, and what is not

Preserved: schema, column classes, factor levels, cohort size, endpoint set,
event structure, coarse regimen and sampling timing, and the broad shape and
magnitude of each endpoint.

Not preserved, by design: exact source distributions, parameter estimates,
covariate-response relationships, and rare individual trajectories. Identifiers
are always freshly generated and never reuse a source value.

Not provided: any formal privacy guarantee. The synthetic data is built by
blending real subject trajectories, so it is appropriate wherever the source
data and synthetic data are accessible by the same users, but — but not for release to anyone outside. 

## References

1. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).

2. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).
