---
title: "Scorecard: Checks of the synthetic data"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Scorecard: Checks of the synthetic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
both <- has_nlmixr2data && has_xgxr
```

You have generated a synthetic dataset. Should you use it?  The scorecard, generated by `synpmx_scorecard()` helps assess this and this Vignette explains it.

## The scorecard

The scorecard contains a table of the following items:

* The check of the dataset that is performed.
* That data is read (source data, synthetic data, or both)
* The resulting calculation from the check
* A verdict for whether the check passed, failed or needs review
* An `explore` function to help assess any checks that failed or need review.

There are four categories of checks, and these are the questions each one asks:

* A. Is the dataset valid
    * **A1**: Is the synthetic table a legal PMX dataset?
    * **A2**: Is the *source* legal under the roles you declared?
    * **A3**: Did every endpoint survive?
    * **A4**: Did the cohort size survive?
    * **A5a**: Did the number of observations per patient survive?
    * **A5b**: Did the number of doses per patient survive?
    * **A6**: Did a discrete endpoint stay discrete?
* B. Is any individual patient identifiable based on the synthetic data
    * **B1a**: Does any avatar have a visit set nobody else shares?
    * **B1b**: Does any avatar have a dose schedule nobody else shares?
    * **B2**: Does any synthetic patient stand out from its own stratum?
    * **B3**: Is any avatar too close to a real patient in value space?
    * **B4a**: Is any generated time vector a copy of an exposed real one?
    * **B4b**: Is any generated `DV` vector a copy of an exposed real one?
    * **B5a**: Does any categorical level sit on a single synthetic patient?
    * **B5b**: Did a level too few *source* patients held reach the output?
* C. Has the synthetic study design significantly changed from the source
    * **C1**: Did every arm keep its endpoints and its size?
    * **C2**: What features of the study were lost?
* D. How much have the covariate and observation distributions changed
    * **D1**: Do the values land in the same range?

**Seven checks can say `FAIL`, and no others.** A1, because the output is not a
legal dataset; A3 and A6, because it is not the study that went in; and B1a,
B1b, B4a and B4b, because each means one real patient's structure was
reproduced verbatim. Everything else is `review`, because everything else can
move for a legitimate reason: a subject dropped for want of donors, an arm that
changed size, a statistic wandering at a small sample size.

One requirement is not in the list, because no function can answer it: **does
the pipeline that will consume the real study run unchanged against this?** It
is the only strictly necessary check, and it is the reason the package exists.

## The two datasets used here

```{r data, eval=both, message = FALSE}
case1_pkpd <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
case1_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB",
  keep = "STUDY"
)
case1_synth <- suppressWarnings(
  synpmx_avatar(case1_pkpd, case1_roles, seed = 808)
)

data("pheno_sd", package = "nlmixr2data")
pheno_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  covariates = c("WT", "APGR")
)
pheno_synth <- suppressWarnings(
  synpmx_avatar(pheno_sd, pheno_roles, seed = 1010)
)
```

`xgxr::case1_pkpd` is the worked example: 180 patients, six treatment arms, a
declared nominal time, two endpoints — a pharmacokinetic (PK) concentration and
a continuous pharmacodynamic (PD) effect — and a baseline weight. It is shaped
like a real study report, and most of these checks pass on it.

`nlmixr2data::pheno_sd` is 59 real neonates given phenobarbital in routine
clinical care, and it is here because **no check on it fails while its full
dosing history does not survive**: the median infant received twelve doses and
the median avatar receives one. A dataset on which everything passes teaches
nothing. Where the two disagree is where a check is doing work.

## The two scorecards

```{r cards, eval=both}
case1_card <- synpmx_scorecard(case1_pkpd, case1_synth, case1_roles)
pheno_card <- synpmx_scorecard(pheno_sd, pheno_synth, pheno_roles)
case1_card
pheno_card
```

Nothing fails on either, and on `pheno_sd` that is not the same as being fine:
A5b reads `10 -> 1.1` doses per patient and C2 represents 3 of its 56 dose
schedules, both `review`. The rest of this vignette is what each check asks,
why its pass criterion is what it is, and how a study can lose its dosing
without a single `FAIL`.

# A. Is the dataset valid

```{r a-validate, eval=both}
validate_pmx(case1_synth, case1_roles)$valid
validate_pmx(pheno_synth, pheno_roles)$valid
```

`validate_pmx()` checks schema and column classes, event grammar, time
monotonicity within subject, censoring flags against their dependent variable
(`DV`), infusion start/stop pairing, baseline covariates that are actually
constant, and strata that do not vary within a subject. Run it on the source as
well as the synthetic table, which is what A1 and A2 are:

```{r a-source, eval=has_xgxr}
case1_roles_cens <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME", cens = "CENS",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB"
)
validate_pmx(case1_pkpd, case1_roles_cens)$valid
```

That study's `CENS` column is meaningful only for the PK endpoint. The PD
effect is signed, so a row flagged as below the limit of quantification (BLOQ)
reports a value *above* uncensored ones. Declaring `cens` here would produce a
coherent-looking dataset with nonsense censoring; leave it undeclared, or
correct the column.

## A2. Time after dose

`validate_pmx()` reports a disagreement in time after dose (`TAD`) as a
*warning*, so it passes `$valid` and still has to be read. A trough sample must
stay a trough, and this is the column that says whether it did.

**`tad` is an output, not an input.** `synpmx_avatar()` recomputes it from the
generated times and dose rows and overwrites whatever the source held.
Declaring the role says which column to overwrite and carry through. The source
values are read in one place only, `validate_pmx()` on the source, and a
disagreement there is a real finding every time:

```{r a-tad, eval=has_nlmixr2data}
nimo <- get(utils::data(list = "nimoData", package = "nlmixr2data"))
nimo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
  covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
)
nimo_checks <- validate_pmx(nimo, nimo_roles)$checks
cat(nimo_checks$message[nimo_checks$check == "tad_agreement"])
```

Forty-five percent of observation rows disagree, by up to 311.9 hours. That
means one of: the study measures `TAD` from the end of an infusion, from a
nominal dose time, or from an assigned occasion rather than the most recent
dose; or the source column is wrong; or the derivation is wrong for this study.
The check cannot tell you which. The synthetic column follows the derivation,
so a silent change of convention is what this check surfaces.

Two limits. A sample taken before any dose is reported at `TAD` 0 because
`validate_pmx()` refuses a negative. And where `addl`/`ii` are declared, the
derivation cannot see the doses they imply, so the check is skipped and says so.
A related gap, whether a sample stayed on the right side of its dose, is in
"What is still missing" at the end.

## A3. Endpoint survival

Compare endpoint *sets*, not row counts. The defect that motivated A3 lost an
entire endpoint, 108 rows in and 60 out, with every observation in one
compartment gone and no warning.

```{r a-endpoints, eval=has_xgxr}
setequal(
  unique(as.character(case1_pkpd$NAME)),
  unique(as.character(case1_synth$NAME))
)
```

## A5a, A5b. Observations and doses per patient

A5a and A5b split rows per patient by event type, because a total hides the
thing worth seeing. On `case1_pkpd` both hold. On `pheno_sd` observations are
intact at 2.6 per patient either side while dosing falls from 10.0 to 1.1:
744 rows become 220, and 589 dose rows become 65.

```{r a-doses, eval=has_nlmixr2data}
doses_per_patient <- function(data) {
  as.numeric(table(data$ID[data$EVID != 0]))
}
summary(doses_per_patient(pheno_sd))
summary(doses_per_patient(pheno_synth))
```

The median real infant received twelve doses. The median avatar receives one.
`plot_pmx_schedule()` draws it, one row per patient, a grey tick for each dose
and a coloured dot for each observation:

```{r a-schedule, eval=has_nlmixr2data, fig.height=7}
plot_pmx_schedule(pheno_sd, pheno_roles, main = "pheno_sd, source")
plot_pmx_schedule(pheno_synth, pheno_roles, main = "pheno_sd, synthetic")
```

The source is a forest of dose ticks running the length of follow-up. The
synthetic cohort has one column of them at time zero, and its observations have
piled onto a handful of shared grid times. The shading answers a different
question from B1a and will not agree with it: it marks patients whose visit set
no other patient *in the drawn table* has, while B1a asks whether an avatar's
set is one a real infant held.

Nothing is invalid here. Every avatar's regimen is one a real infant received,
which is the guarantee section C describes. Dose schedules in this study are
nearly all unique, so truncating each avatar's regimen back to a depth several
patients share collapses most of them to the shortest depth. The masking worked
as designed, and the result is not usable as a dosing dataset. Section B1 is the
other half: the privacy checks read 0, so nothing in that tier reports a
problem. A5b is what catches it.

## A6. Discrete endpoints

Ask what kind of values each endpoint takes before reading anything else about
it. Blending is a weighted mean, so a weighted mean of several patients' zeros
and ones is a number between them, and a 0/1 endpoint comes back continuous
unless the generated values are put back on the scale the source used.

```{r a-endpoint-types, eval=has_xgxr}
pmx_endpoint_types(case1_pkpd, case1_roles)
```

Both endpoints here are continuous, so `synpmx_scorecard()` omits A6 entirely.
The check fires on a study with an ordinal, binary or count endpoint:
`vignette("public-data-examples")` runs it on `xgxr::mad`, which carries all
three.

The type is *inferred* from the source — whether every observed value is a
whole number, and how many distinct ones there are — and
`pmx_roles(endpoint_types = )` overrides it where that inference reads the
study wrongly. Putting a binary endpoint back on its levels means generated
values are source values, since a 0/1 endpoint has no third value to emit. What
protects a patient on a discrete endpoint is section B, never the distinctness
of any one number.

# B. Is any individual patient identifiable based on the synthetic data

## B1. Timing and structure

Who was observed when, and who was dosed when. This is the axis most specific
to longitudinal data, and the one a general-purpose synthetic-data tool will
not have addressed.

```{r b1-uniqueness, eval=has_xgxr}
skeleton_uniqueness(case1_pkpd, case1_roles)
```

Every one of the 180 patients has an observation schedule nobody else has,
before any coarsening. On the visit grid `synpmx_avatar()` actually builds:

```{r b1-coarsened, eval=has_xgxr}
skeleton_uniqueness(case1_pkpd, case1_roles, coarsen_time = TRUE)
```

Zero. The declared `nominal_time` supplies the protocol's grid, so patients
land on it exactly.

The counts to act on are measured on the output, not the source, and both must
be 0:

```{r b1-guarantees, eval=both}
guarantees <- function(synthetic, label) {
  settings <- attr(synthetic, "pmx_settings")
  data.frame(
    dataset = label,
    identifying_visit_sets = settings$identifying_visit_sets,
    identifying_dose_schedules = settings$identifying_dose_schedules
  )
}
knitr::kable(rbind(
  guarantees(case1_synth, "case1_pkpd"),
  guarantees(pheno_synth, "pheno_sd")
))
```

Both are 0, `pheno_sd` included, and `pheno_sd` is still the case worth
studying: passing is not the same as being fine. Neonatal phenobarbital is
dosed to the individual infant on the day a clinician decided, so 55 of its 59
infants have a dose schedule nobody else has:

```{r b1-pheno, eval=has_nlmixr2data}
pheno_unique <- skeleton_uniqueness(pheno_sd, pheno_roles)
sum(pheno_unique$n_share_dosing == 1)
```

Dose events are copied from the anchor verbatim, because moving a dose invents
a regimen the protocol never permitted. The only lever left is to build avatars
on the few patients whose dosing *is* shared, and here the one shared schedule
is a single dose. B1b is 0 because the dosing was truncated away, not because
it was masked.

**Read B1a and B1b next to A5b.** A check that must be 0 says nothing about
what reaching 0 cost, and A5b is where that cost showed up. Where dosing is
individualised, no setting fixes this. It is a property of the study, and the
honest response is to say so rather than ship.

### B1 failures by arm

`pheno_sd` declares no arms, so its dosing problem is the whole cohort's. Where
`strata` are declared, both checks are decided arm by arm, because an avatar is
anchored on one real patient and is never moved out of the arm it was allocated
to. An arm holding nobody who can be masked therefore fails whatever the rest
of the study looks like, and the result cell names it:

```
B1b   Avatars with a dose schedule nobody else shares   6 in ARM8 (6)   FAIL
```

`unmaskable_strata(source, roles)` answers the same question before generating,
one row per arm, splitting the patients no avatar can be built on into
`unmaskable_dosing`, which nothing can substitute for, and `unmaskable_visits`,
which a wider pool usually can. An arm with `safe_anchors = 0` fails on every
seed.

### Near-miss distance

Exact-set equality is a harsh test. A patient one missing sample away from a
neighbour is counted exactly like a patient with an ad-hoc schedule, and the
remedies are different.

```{r b1-nearmiss, eval=has_nlmixr2data}
summary(pheno_unique$nearest_set_diff)
```

`nearest_set_diff` is how many observation slots separate a patient from their
nearest neighbour. A median of a few slots on a sparse study is a different
situation from a median of twenty, and the headline count cannot tell them
apart.

### Which endpoint is responsible

A schedule is only as shared as its least-shared part:

```{r b1-endpoint, eval=has_xgxr}
attr(skeleton_uniqueness(case1_pkpd, case1_roles, coarsen_time = TRUE),
     "by_endpoint")
```

Neither endpoint is responsible here, because neither is exposed. Where the
count is nonzero, this table says whether dense PK sampling or a sparse
biomarker is driving it.

`plot_pmx_schedule()` draws the same thing, and a count that looks alarming and
one that is ordinary look different on the page. The `pheno_sd` figures above
are that picture: 54 of 59 rows shaded is a study where being alone is the
norm, not a study with a few stragglers.

## B2. Extremes

Who stands out from the cohort, on follow-up length, dose count, dose magnitude
or peak `DV`. This runs on the synthetic table and does not read the source.

```{r b2-flag, eval=has_xgxr}
case1_flags <- flag_identifiable_subjects(case1_synth, case1_roles)
sum(case1_flags$flagged)
sort(table(as.character(case1_flags$outlier_axes[case1_flags$flagged])),
     decreasing = TRUE)
```

One flagged, on follow-up time. What makes that the right answer is the
comparison group. `case1_pkpd` is a six-arm dose-ranging study from 3 mg to
300 mg, so dose magnitude is spread by *protocol*:

```{r b2-arms, eval=has_xgxr}
table(case1_flags$max_dose)
```

Those are the six arms, not six kinds of outlier. Scored against the whole
cohort, the top arm sits about 6.4 modified-z units from the median dose, so
all thirty of its patients are flagged for receiving the dose they were
assigned. Taking the strata away shows it:

```{r b2-unstratified, eval=has_xgxr}
case1_roles_flat <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME", covariates = "WEIGHTB"
)
sum(flag_identifiable_subjects(case1_synth, case1_roles_flat)$flagged)
```

`flag_identifiable_subjects()` therefore scores each axis within the declared
`strata`. Within an arm every dose is identical, so nobody stands out on it and
what survives is the patient who is genuinely unusual, while a patient given
twice their own arm's dose is still flagged. "Does this patient stand out?" is
meaningless without saying *from whom*, and a screen that ignores treatment
assignment reports the protocol back to you as a privacy finding. Strata
holding fewer than five patients are scored against the cohort instead, since a
scale estimated from four patients describes those four.

`pheno_sd` has no strata to declare, so it is scored cohort-wide:

```{r b2-pheno, eval=has_nlmixr2data}
pheno_flags <- flag_identifiable_subjects(pheno_synth, pheno_roles)
sum(pheno_flags$flagged)
```

The screen's own statistics need checking too. Its robust scale, the median
absolute deviation (MAD), is exactly zero whenever more than half a cohort
shares one value, which on trial data is the ordinary case rather than a
degenerate one. A zero scale makes everything an outlier. Test a screen like
this on a cohort where half the values are identical.

`remediate_identifiable_subjects()` acts on what survives that reading, by
dropping or truncating the subjects it is given.

## B3. Values and geometry

Whether a synthetic patient's numbers sit too close to a real patient's, taking
covariates and trajectory together.

```{r b3-proximity, eval=both}
proximity <- function(source, synthetic, roles, label) {
  report <- compare_pmx_proximity(source, synthetic, roles, replicates = 30)
  data.frame(
    dataset = label,
    adversarial_accuracy = round(report$adversarial_accuracy, 3),
    null_lower = round(report$null_lower, 3),
    null_upper = round(report$null_upper, 3),
    per_side = report$n_compared
  )
}
knitr::kable(rbind(
  proximity(case1_pkpd, case1_synth, case1_roles, "case1_pkpd"),
  proximity(pheno_sd, pheno_synth, pheno_roles, "pheno_sd")
))
```

Adversarial accuracy asks whether each subject's nearest neighbour lies in its
own dataset or the other one. 0.5 means a synthetic subject is no more like a
real subject than two real subjects are like each other; toward 0 means
memorisation.

**Read the null interval before the point estimate.** It is wide at
pharmacometric cohort sizes, and a value inside it means *nothing was
detected*, never *nothing is there*. What it reliably catches is a blatant
leak: hand this function a verbatim copy of the source and it drives to zero
and objects.

**Read the direction before reacting to a miss.** B3 is `review` whatever it
lands on, because falling outside the interval means two opposite things. Below
it is memorisation, synthetic subjects sitting closer to real ones than real
ones sit to each other, and is the failure this statistic exists to find. Above
it means the two sets have separated, which costs utility and discloses
nothing. At a few dozen patients either can also be the seed.

## B4. Exact copies

Two vectors per subject must not be reproduced: B4a, the observation times, and
B4b, the `DV` values recorded at them. They fail independently. A generator can
invent a schedule and carry a real patient's measurements onto it, or place
real measurements on a new grid, and each is a copy of half the patient.

What counts as a copy is a vector too few real patients share, the same rule
`min_pattern_share` applies to visit sets. Reproducing a vector a whole arm
shares identifies nobody: on a fixed protocol grid every patient is sampled at
the same times, so a check asking "does any synthetic time vector equal some
real one" fails every well-designed study while catching nothing. It has to ask
whether the vector belonged to *one* patient.

```{r b4-copies, eval=both}
case1_card[case1_card$check %in% c("B4a", "B4b"), ]
```

Both zero. Values are sorted before comparison, so a reordering of the same
numbers still counts as a copy, since the reordering is not what protects the
patient. On a study with one protocol grid B4a is close to silent by
construction, so B4b is the check doing the work. The same comparison is worth
running on whole covariate rows, which nothing does for you.

## B5a, B5b. Rare categories

`synpmx_avatar()` treats the two kinds of covariate differently, and only one
of them is blended:

- **Numeric** covariates become a weighted mean of the donors' values plus
  noise. The result is a new number nobody had.
- **Categorical** covariates are `sample()`d from the donors' values. There is
  nothing to average: a synthetic patient's category is always some real
  patient's actual category, copied.
- **`strata`** (arm, dose group) are copied from the anchor exactly, because
  they are protocol facts.

So the question is when a copied category reaches the output.
`vignette("avatar-algorithm")`, under step 8, runs a 40-patient fixture in
which one categorical level is held by a varying number of otherwise-ordinary
patients. **A level held by one patient never reaches the output; a level held
by two does.** By ten holders the output tracks the source frequency closely.
That is geometry rather than protection — the sole holder of a level sits alone
on its own one-hot axis, is nobody's nearest neighbour, and is therefore almost
never chosen as a donor — so nothing enforces it and nothing reports it.

The dangerous shape is a rare level on a *typical* patient. A patient who is
unusual on their covariates is nobody's nearest neighbour and is
self-protecting. A patient who is ordinary in every way except one rare
category, an ultra-rare mutation status say, sits in the middle of the profile
space, is selected as a donor constantly, and their category is copied out.

The disclosure is also not "this value is unique in the dataset". It is that
the value may be rare **in the world**. If five people alive carry a mutation,
a synthetic dataset containing it discloses that someone with that mutation was
in this study, which for a named trial with public inclusion criteria can be
nearly identifying on its own. No cohort size helps.

### B5b. The source-side census

`compare_pmx_rare_levels()` censuses every categorical axis the roles declare,
`strata` and each non-numeric covariate. The level set it reports is the
**union** of both sides rather than the synthetic side alone, because a level
dropped on the way out is a coverage failure that a one-sided census cannot
see.

```{r b5-census, eval=has_xgxr}
compare_pmx_rare_levels(case1_pkpd, case1_synth, case1_roles)
```

A level is **exposed** when fewer than `min_pattern_share` real patients held
it, the same floor that protects visit sets. The row to act on is an exposed
level that `reached` the output, because that level is one real patient's
attribute, copied. On `case1_pkpd` nothing is exposed: every arm is held by 30
patients. A study with a two-patient stratum would show it here and nowhere
else. A `0` in the synthetic column is the other failure, a level the source
had and the output lost.

The remedies are upstream of generation: drop the covariate from `covariates`,
or collapse its rare levels before generating. The synthetic column moves
because anchors are sampled with replacement.

### B5a. What can be checked without the source

The census above reads real patient data and is restricted output. One weaker
question can be asked of the released table alone: does any categorical level
sit on exactly one synthetic patient? That patient is singled out within the
release, whatever the source looked like.

```{r b5-singletons, eval=has_xgxr}
case1_card[case1_card$check == "B5a", ]
```

The result names the column and the level, not just the count, because "1" on
its own gives a reader no way to find the patient it is talking about. Here the
answer is an arm size, thirty, and the pass is any answer above one. `pheno_sd`
declares no categorical covariate and no `strata`, so the check is absent from
its card, which is *not applicable* rather than *clean*.

B5a is the weak form, and both its failure modes are real. A level held by two
source patients can be copied onto ten avatars and pass while the disclosure is
real, since what matters is how many *real* patients held it. A level held by
thirty source patients can land on one avatar by chance and fail harmlessly.
The synthetic-side count is what you can compute on a table that has left the
trusted environment; the source-side count is what the risk is made of.

### Combinations, which nothing enumerates

With `d` covariates there are `2^d` subsets that could serve as a
quasi-identifier — arm x sex x age band x mutation — and an attacker's chosen
combination need not be one you checked. One mitigation is accidentally
present: each covariate is sampled independently, so sex may come from one
donor and race from another, and a rare *joint* combination is less likely to
be reproduced whole than any of its parts. That is a fidelity cost, real
correlations between covariates broken, doing double duty as a weak privacy
benefit.

### Differentially private modes

This is the clearest difference between the two families in this package.
AVATAR's protections are **enumerated** — visit sets, dose schedules, extremes,
proximity — so a risk nobody named is a risk nobody covered, and combinations
are exactly such a risk. An (epsilon, delta) differential privacy (DP)
guarantee is **unenumerated**: it bounds the influence of any one individual on
*any* function of the output, so it holds for combinations nobody thought to
check.

The DP path never copies a category. `pmx_covariate(levels = )` releases a
noisy count vector over the levels, normalises it to probabilities, and
generation draws from that distribution; a level held by two patients of forty
is swamped by noise at any sensible epsilon. Three caveats:

1. **The level set must be public.** A domain derived from the data leaks the
   domain, which is why `pmx_covariate()` requires `levels` and a citation.
2. **Delta matters more than usual.** (epsilon, delta)-DP permits the guarantee
   to fail with probability delta, and for a category held by one or two people
   that is the failure that matters. Keep delta well below 1/n.
3. **DP bounds the mechanism, not the world.** If inclusion criteria are public
   and only a handful of people could qualify, membership may be inferable
   regardless.

The [privacy article](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html) works through the trade-off.

## Deterministic proxies, which nothing checks

A column that is a function of a covariate discloses that covariate exactly.
This is not a scorecard check, because there is nothing to run: the generator
handles the one case it can detect, and the rest is a reading of your own
columns.

The worked case is dosing. Under milligram-per-kilogram dosing, an `AMT` copied
from the anchor reveals the anchor's weight to the gram, and blending the
weight column is irrelevant because the dose gives it back. `synpmx_avatar()`
detects proportional dosing, recomputes the amount from the avatar's own
blended covariate, and reports which path it took:

```{r b6-dose, eval=both}
attr(case1_synth, "pmx_settings")$dose_basis_note
attr(pheno_synth, "pmx_settings")$dose_basis_note
```

Neither study is dosed proportionally to a declared covariate, so inference
declines rather than imposing a rule the study did not use. When it does fire,
`dose_basis` names the covariate.

Detection covers only the dose. Body surface area from height and weight, body
mass index, creatinine clearance, any derived exposure column carried through
with `keep` — each is a covariate written down twice, and masking one copy does
nothing. Look for them by hand.

# C. Has the synthetic study design significantly changed from the source

Failing these endangers nobody. It produces data nobody can develop against,
which is the entire purpose of the package.

Dose counts, intervals and amounts must be ones the protocol permitted, which
is why `synpmx_avatar()` never resamples dose times and only ever truncates a
real schedule at one of its own dose times. A generator that jitters dose times
produces regimens no protocol allows, and it looks fine in every distributional
check.

## C1. Endpoint coverage and arm size

All endpoints present is A3. Whether each endpoint is present *for the right
patients* is C1:

```{r c-coverage, eval=has_xgxr}
observed_by_arm <- function(data, label) {
  observed <- data[data$EVID == 0, ]
  out <- as.data.frame.matrix(table(observed$TRTACT, observed$NAME))
  out$dataset <- label
  out$arm <- rownames(out)
  out[, c("dataset", "arm", setdiff(names(out), c("dataset", "arm")))]
}
knitr::kable(rbind(
  observed_by_arm(case1_pkpd, "source"),
  observed_by_arm(case1_synth, "synthetic")
), row.names = FALSE)
```

The placebo arm has no PK concentration in either table, which is the check
passing: an avatar never leaves the arm it was anchored in, so it cannot
acquire an endpoint its arm never had.

The arm sizes also match, which they did not until this check was written.
Anchors are sampled with replacement, so left alone the arm sizes are whatever
the draw gives: this design of exactly 30 per arm came back between 21 and 39
per arm depending only on the seed, with cohort size and arm membership both
exact and only the *balance* wrong. `preserve_strata_balance` (default `TRUE`)
gives each declared stratum its source share:

```{r c-balance, eval=has_xgxr}
unbalanced <- suppressWarnings(
  synpmx_avatar(case1_pkpd, case1_roles, seed = 808,
                preserve_strata_balance = FALSE)
)
table(unbalanced$TRTACT[!duplicated(unbalanced$ID)])
```

One deliberate exception: a stratum holding fewer than three source patients is
left unbalanced on purpose, because reproducing its size exactly would disclose
that size. `strata_balanced` and `strata_stochastic` in the run settings say how
many strata fell on each side.

## C2. What the masking cost

`pmx_masking_report()` says what each mechanism did and what it cost. The
dosing section is the one C2 reports; drop the `section` argument for the other
six.

```{r c-masking, eval=has_nlmixr2data}
knitr::kable(as.data.frame(
  pmx_masking_report(pheno_synth, pheno_sd, pheno_roles,
                     section = "dose_schedules")
), row.names = FALSE, align = c("l", "r", "l"))
```

A "dose schedule", here and in C2, is the set of dose **times** a patient
received, and nothing else. Amounts are deliberately not part of it: on a
weight-based study every patient's milligrams are their own, so counting
amounts would make every patient their own schedule. Two patients on the same
visit days at different doses are one schedule; a patient who stopped after two
of three doses is another, and if nobody shares it, declining to anchor on them
removes it from the output.

On `pheno_sd` 3 of 56 distinct dose schedules are represented, and a third of
avatars had their dosing re-truncated to reach that. It is the A5b finding
again, from the generator's side.

# D. How much have the covariate and observation distributions changed

## D1. Endpoint and covariate distributions

```{r d-distributions, eval=has_xgxr}
case1_dist <- compare_pmx_distributions(case1_pkpd, case1_synth, case1_roles)
knitr::kable(case1_dist$endpoints, digits = 3,
             caption = "Endpoints, source against synthetic")
knitr::kable(case1_dist$covariates_numeric, digits = 2,
             caption = "Baseline weight, source against synthetic")
```

**Between-subject variability shrinks, necessarily.** An avatar is an average
of several donors, and averaging reduces variance. Baseline weight above has a
source standard deviation of about 20.5 kg and a synthetic one of about 14.1.
D1 reports whichever endpoint or covariate moved furthest, `sd x0.64` on the PD
endpoint here, and is always `review`, since neither direction has a threshold
that would be right on a second study. The quantities that explain how much are
reported with the run:

```{r d-blending, eval=has_xgxr}
unlist(attr(case1_synth, "pmx_settings")[
  c("k", "mean_effective_donors", "max_donor_weight", "cap_binding_fraction")
])
```

`mean_effective_donors` is how many donors an avatar is genuinely averaged over
once the weights are accounted for, under 3 here out of `k = 5`, and
`cap_binding_fraction` is how often the single-donor weight cap bound. Fewer
effective donors means more fidelity and less masking; that trade is the whole
design.

**Plot the data as well.** Everything above is summary statistics, and a
summary statistic cannot see a shape: a single mode and two modes with the same
mean and spread give identical rows. Overlay source and synthetic on the same
axes, `DV` against time and each covariate's distribution, before deciding the
output is usable. No function here draws it, because every group has plotting
code it already trusts for its own study.

**Do the covariate–response relationships survive?** The question a modeller
actually cares about, and the least covered by anything above: a
weight–exposure slope, a dose–response relationship, a treatment effect.
Marginal distributions can match while the relationship between them is
destroyed. The
[covariate and treatment effect article](https://iamstein.github.io/synpmx/articles/example-avatar-PKPD-covariate-treatment-effect.html)
is the worked evidence.

# What these checks cannot tell you

**Nothing here bounds what an adversary learns.** These checks reduce the ways
a real patient can be singled out. They do not limit how often that succeeds,
and they compose no better than the weakest one. That is what the
differentially private modes are for.

**The guarantees are about reproduction, not similarity.**
`identifying_visit_sets` counts avatars with a visit set that is *exactly* one
real patient's. An avatar whose visits sit merely *near* a real patient's is
not covered by it, and `nearest_set_diff` is the only thing that will tell you
how near.

**Most checks read the source, and are therefore restricted output.** They may
not leave the environment the source data lives in. `compare_pmx()` reports
this per component:

```{r e-release, eval=has_xgxr}
compare_pmx(case1_pkpd, case1_synth, case1_roles)$release_status
```

Only two things in this vignette do not read the source: `validate_pmx()` and
`flag_identifiable_subjects()`, both on the synthetic table. Every uniqueness
count, distribution comparison and proximity statistic is computed from real
patient data and inherits its handling obligations.

# What is still missing

**A1: a sample can cross a dose.** Generated observation times move and dose
times do not, so a sample drawn before a dose can land at or after it. On
`nlmixr2data::nimoData` that happens to 6 of 12 avatars, each carrying an
observation labelled occasion 2 at the same time as the occasion-3 dose: a
trough that is now a time-zero sample, and a declared `occasion` column that no
longer runs in step with `time`. Closing it takes a post-generation invariant
over the finished table, checking that each observation keeps the number of
doses that preceded it and that `occasion` is non-decreasing within a subject.

**B5: rare combinations, and rarity in the world.** The source-side census
exists as `compare_pmx_rare_levels()` and B5b. Joint combinations do not, since
with `d` covariates there are `2^d` of them and the census is per column;
closing that takes a decision about which combinations are worth enumerating.
The one that matters most is declaring which levels are rare *in the
population* rather than in this study, which is not measurable from the study
at all. Its realistic form is a declaration on `pmx_roles()`, or not carrying
that covariate. It is also the only gap here that would change generation
rather than reporting, by suppressing or coarsening a level.

Two further gaps are whole measurement approaches the package does not attempt,
both covered in the
[checking literature review](https://iamstein.github.io/synpmx/articles/synthetic-data-checking-review.html):

- **No control group.** Every check that reads the source compares the
  synthetic data against patients the generator was allowed to use. That
  confounds *the generator captured the population* with *the generator
  memorized a patient*, and the standard remedy — holding patients out of
  generation and asking the question twice — is not implemented. It is why B3's
  null interval says less than it appears to.
- **Nothing measures linkability, and nothing is per patient on the privacy
  side.** The AVATAR literature's own per-patient measures, local cloaking and
  hidden rate, apply directly to this generator and are not computed.

And one check cannot be on this list: **does the pipeline that will consume the
real study run unchanged against this?** Only you can answer it. Run your own
code against the output and see whether the joins, reshapes, derivations and
control streams behave.

## Where to go next

- `vignette("demo")` — generating a dataset and reading its scorecard, on one
  study end to end.
- `vignette("avatar-algorithm")` — how the default generator works, and the six
  masking mechanisms these checks are measuring.
- `vignette("public-data-examples")` — these checks run across eight public
  datasets, with the masking cost for each.
- [Literature: checking synthetic data](https://iamstein.github.io/synpmx/articles/synthetic-data-checking-review.html) — the published methods for
  checking synthetic data, what each one asks, and which of them this package
  does not attempt.
- [Literature: generating synthetic data](https://iamstein.github.io/synpmx/articles/synthetic-data-generation-review.html) — the four families of
  generation method, and where `synpmx` sits among them.
- [Privacy](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html) — when the enumerated protections are not enough,
  and what a formal guarantee buys instead.
