---
title: "Checks of the synthetic data"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Checks of the synthetic data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
both <- has_nlmixr2data && has_xgxr
```

You have generated a synthetic dataset. Should you use it?

This vignette is the question list. It is written for two readers: someone
deciding whether a generated dataset is fit to work with, and someone building
their own synthetic-data method who wants the categories rather than this
package's function names. **The categories are what generalize.** A method that
answers all six questions with something is in better shape than one that
answers three of them very well.

The list exists because of how the defects in this package were actually found.
Every one of them was found by *looking at output*, not by reasoning about the
algorithm, and several had been in place for weeks behind mechanisms that were
individually correct. The last section says what that taught, because it is the
part that transfers.

## The scorecard

Every check in this vignette, with the thing to run and **the answer that counts
as passing**. Most of the pass criteria below are stated once, in prose, in the
section that introduces them; collecting them here is what makes the document
usable as a checklist rather than as an essay.

Three columns need a word of explanation.

- **Reads** is where the number comes from, and it decides where the number may
  go. `source` means the check reads real patient data, so its output is
  restricted and stays inside the environment the source lives in — see section
  E. `synthetic` and `run settings` do not read the source. `run settings` is
  the generation run's own record of what it did, carried on the output as
  `attr(synthetic, "pmx_settings")`, rather than a measurement taken from
  either table.
- **Pass** is deliberately not always a number. Some of these are guarantees
  that must hold exactly; others are readings that need judgement, and marking
  those *review* is more honest than inventing a threshold.
- **Gap** in the "What to run" column means the package does not provide it. The
  section says what closing it would take, and the table at the end of the
  vignette collects them.

| # | Question | What to run | Reads | Pass |
|---|---|---|---|---|
| **A1** | Is the synthetic table a legal PMX dataset? | `validate_pmx(synthetic, roles)$valid` | synthetic | `TRUE` |
| **A2** | Is the *source* legal under the roles you declared? | `validate_pmx(source, roles)$valid` | source | `TRUE`, or a refusal you can explain — *review* either way |
| **A3** | Did every endpoint survive? | `setequal()` on the `dvid` levels | both | sets equal |
| **A4** | Did the cohort size survive? | count distinct `id` | both | equal is a pass; a smaller cohort is *review*, since a subject with too few donors is dropped by design |
| **A5** | Did dosing survive? | rows per patient, **split by event type** | both | review — totals hide this |
| **A6** | Did a discrete endpoint stay discrete? | `pmx_endpoint_types()`; `synpmx_scorecard()` | both | every generated value on a binary, ordinal or integer endpoint is one the source could have held |
| **B1a** | Does any avatar wear one real patient's visit set? | `identifying_visit_sets` | run settings | **0** |
| **B1b** | Does any avatar wear one real patient's dose schedule? | `identifying_dose_schedules` | run settings | **0** |
| **B2** | Does any synthetic patient stand out from its own stratum? | `flag_identifiable_subjects()` | synthetic | review each; not necessarily 0 |
| **B3** | Is any avatar too close to a real patient in value space? | `compare_pmx_proximity()` | both | inside the null interval, near 0.5 — always *review*, because outside it means two different things |
| **B4a** | Is any generated time vector a copy of an exposed real one? | `synpmx_scorecard()` | both | **0** |
| **B4b** | Is any generated `DV` vector a copy of an exposed real one? | `synpmx_scorecard()` | both | **0** |
| **B5a** | Does any categorical level sit on a single synthetic patient? | `synpmx_scorecard()`; `table(synthetic$COLUMN)` | synthetic | no level on one patient — *review*, since this is the weak form of the check |
| **B5b** | Did a level too few *source* patients held reach the output? | `compare_pmx_rare_levels()` | both | no exposed level reaching the output — *review* while the check earns its thresholds |
| **C1** | Is time after dose still what it was? | `validate_pmx(source, roles)` `tad_agreement` | source | pass, or a disagreement you can explain |
| **C2** | Do dose and observation stay in order? | *gap* | — | — |
| **C3** | Did every arm keep its endpoints and its size? | arm x endpoint table; `compare_pmx_strata_sizes()` | both | no endpoint where the source had none; every arm matching its source size is a pass, a shifted size is *review* |
| **C4** | What features of the study were lost? | `pmx_masking_report()` | both | every distinct set of dose **times** in the source still represented; the rest of the question has no pass mark |
| **D1** | Do the values land in the same range? | `compare_pmx_distributions()` | both | same magnitude and shape; spread **will** shrink |

**Seven of these rows can say `FAIL`, and no others.** A1, because the output
is not a legal dataset; A3 and A6, because it is not the study that went in —
an endpoint missing, or a value on a discrete endpoint that the source could
not have held; and B1a, B1b, B4a and B4b, because each of them means one real
patient's structure was reproduced verbatim. Every other row is `review` even
when its answer is unwelcome, because every other row can move for a legitimate
reason: a subject dropped for want of donors, an arm that changed size, a
statistic wandering at a small sample size, a real source a validator objects
to. A card of `review`s is a card to read, not a card that failed.

One requirement is deliberately not in the table, because no function can
answer it: **does the pipeline that will consume the real study run unchanged
against this?** It is the only strictly necessary check, and it is the reason
the package exists.

`vignette("demo")` closes by computing this scorecard on a real run, so you can
see what the filled-in version looks like.

## The two datasets used here

Two public datasets, and the second one is the point.

```{r data, eval=both}
case1_pkpd <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
case1_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB",
  keep = "STUDY"
)
case1_synth <- suppressWarnings(
  synpmx_avatar(case1_pkpd, case1_roles, seed = 808)
)

data("pheno_sd", package = "nlmixr2data")
pheno_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  covariates = c("WT", "APGR")
)
pheno_synth <- suppressWarnings(
  synpmx_avatar(pheno_sd, pheno_roles, seed = 1010)
)
```

`xgxr::case1_pkpd` is the worked example: 180 patients, six treatment arms, a
declared nominal time, two endpoints — a pharmacokinetic (PK) concentration and
a continuous pharmacodynamic (PD) effect — and a baseline weight. It is the
only public dataset shaped like a real study report, and most of these checks
pass on it.

`nlmixr2data::pheno_sd` is 59 real neonates given phenobarbital in routine
clinical care. Its checks **fail**, and it is here for that reason: a document
in which everything passes teaches nothing. Where the two disagree is where the
check is doing work.

Both are public, so every diagnostic below can be shown. On your own study most
of them read the source data and are therefore restricted output; section E says
which.

# A. Is it a valid dataset at all?

The tier that is easy to skip and embarrassing to fail. A table that is not a
legal pharmacometric (PMX) dataset cannot be assessed for anything else, so
structure comes before statistics.

```{r a-validate, eval=both}
validate_pmx(case1_synth, case1_roles)$valid
validate_pmx(pheno_synth, pheno_roles)$valid
```

`validate_pmx()` checks schema and column classes, event grammar, time
monotonicity within subject, censoring flags against their dependent variable
(`DV`), infusion start/stop pairing, baseline covariates that are actually
constant, and strata that do not vary within a subject. It does not
assess scientific validity, and it is not a privacy check.

**Run it on the source as well.** A validator that only ever passes is not
evidence of anything, and the source is where a mis-declared role gets caught
before it becomes a generation bug. On `case1_pkpd` it refuses one:

```{r a-source, eval=has_xgxr}
case1_roles_cens <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME", cens = "CENS",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB"
)
validate_pmx(case1_pkpd, case1_roles_cens)$valid
```

That study's `CENS` column is meaningful only for the PK endpoint. The PD effect
is signed, so a row flagged as below the limit of quantification (BLOQ) reports a
value *above* uncensored ones, and the flag cannot mean for PD what it means for
PK. Declaring `cens` here would have produced a coherent-looking dataset with
nonsense censoring; the right answer is to leave it undeclared, which is what
the run above does.

## Endpoint survival

**Compare endpoint *sets*, not row counts.** The defect that motivated this
check lost an entire endpoint — 108 rows in, 60 out, every observation in one
compartment gone, no warning — and a row count alone looked plausible.

```{r a-endpoints, eval=has_xgxr}
setequal(
  unique(as.character(case1_pkpd$NAME)),
  unique(as.character(case1_synth$NAME))
)
```

## Discrete endpoints

**Ask what kind of values each endpoint takes before reading anything else about
it.** Blending is a weighted mean, so a weighted mean of several patients' zeros
and ones is a number between them, and a 0/1 endpoint comes back continuous
unless the generated values are put back on the scale the source used.

```{r a-endpoint-types, eval=has_xgxr}
pmx_endpoint_types(case1_pkpd, case1_roles)
```

Both endpoints here are continuous, so there is nothing to do and
`synpmx_scorecard()` omits the A6 row entirely. The check fires on a study with an
ordinal, binary or count endpoint: `vignette("public-data-examples")`
runs it on `xgxr::mad`, which carries all three.

Two things are worth knowing before you trust the answer. The type is *inferred*
from the source — whether every observed value is a whole number, and how many
distinct ones there are — and `pmx_roles(endpoint_types = )` overrides it where
that inference reads the study wrongly. And putting a binary endpoint back on
its levels means generated values are source values, because a 0/1 endpoint has
no third value to emit. What protects a patient on a discrete endpoint is
section B's machinery, never the distinctness of any one number.

## Counts against the source

Count patients, and count rows **split by event type**, because a total hides
the thing worth seeing:

```{r a-counts, eval=both}
counts <- function(source, synthetic, roles, label) {
  tally <- function(data, which) {
    events <- !as.character(data[[roles$evid]]) %in% c("0", "0.0")
    rows <- if (which == "dose") sum(events) else sum(!events)
    round(rows / length(unique(data[[roles$id]])), 1)
  }
  data.frame(
    dataset = label,
    patients = paste(length(unique(source[[roles$id]])), "->",
                     length(unique(synthetic[[roles$id]]))),
    rows = paste(nrow(source), "->", nrow(synthetic)),
    obs_per_patient = paste(tally(source, "obs"), "->",
                            tally(synthetic, "obs")),
    doses_per_patient = paste(tally(source, "dose"), "->",
                              tally(synthetic, "dose"))
  )
}
knitr::kable(rbind(
  counts(case1_pkpd, case1_synth, case1_roles, "case1_pkpd"),
  counts(pheno_sd, pheno_synth, pheno_roles, "pheno_sd")
), caption = "Source -> synthetic. Cohort size is preserved; nothing else is.")
```

**Patient counts match exactly, and should. Row counts do not, and should not** —
each avatar's set of attended visits is redrawn, so totals move by a few percent.
That is a masking mechanism, not a bug, and a row count matching *exactly* would
be the surprising result.

**But look at `pheno_sd`.** It loses 42% of its rows, and the split says where:
its observations are essentially intact while its **dosing is halved**. The
median real infant received twelve doses; the median avatar receives one.

```{r a-doses, eval=has_nlmixr2data}
doses_per_patient <- function(data) {
  as.numeric(table(data$ID[data$EVID != 0]))
}
summary(doses_per_patient(pheno_sd))
summary(doses_per_patient(pheno_synth))
```

Nothing is invalid here — every avatar's regimen is one a real infant received,
which is the guarantee section C describes. What happened is that dose schedules
in this study are nearly all unique, so truncating each avatar's regimen back to
a depth several patients share collapses most of them to the shortest depths.
**The masking worked as designed and the result is not usable as a dosing
dataset.** Section B1 shows the other half of the same story: it did not even
succeed at masking. This is the check that catches it, and it is three lines of
counting.

# B. Is any single patient singled out?

The privacy tier, and the one worth splitting, because "identifying" is not one
property. A patient can be singled out through any of the six routes below
independently, and fixing one does nothing for the others.

## B1. Timing and structure

Who was observed when, and who was dosed when. This is the axis most specific to
longitudinal data, and the one a general-purpose synthetic-data tool will not
have thought about at all.

```{r b1-uniqueness, eval=has_xgxr}
skeleton_uniqueness(case1_pkpd, case1_roles)
```

Every one of the 180 patients has an observation schedule nobody else has — and
that is *before* any coarsening, which is the state the source arrives in, not a
verdict on the output. Run it again on the visit grid `synpmx_avatar()` actually
builds:

```{r b1-coarsened, eval=has_xgxr}
skeleton_uniqueness(case1_pkpd, case1_roles, coarsen_time = TRUE)
```

Zero. That is what a declared `nominal_time` buys: the grid is the protocol's,
so patients land on it exactly.

**The counts to act on are measured on the output, not the source**, and there
are two of them. Both must be 0:

```{r b1-guarantees, eval=both}
guarantees <- function(synthetic, label) {
  settings <- attr(synthetic, "pmx_settings")
  data.frame(
    dataset = label,
    identifying_visit_sets = settings$identifying_visit_sets,
    identifying_dose_schedules = settings$identifying_dose_schedules
  )
}
knitr::kable(rbind(
  guarantees(case1_synth, "case1_pkpd"),
  guarantees(pheno_synth, "pheno_sd")
))
```

**`pheno_sd` fails the second one**, and this is the failure worth studying. Its
visit sets are fine: no avatar wears a set of attended visits belonging to one
real infant. Its *dosing* is not, because neonatal phenobarbital is dosed to the
individual infant on the day a clinician decided, so most patients have a dose
schedule nobody shares and there is no safe patient to anchor on instead:

```{r b1-pheno, eval=has_nlmixr2data}
pheno_unique <- skeleton_uniqueness(pheno_sd, pheno_roles)
sum(pheno_unique$n_share_dosing == 1)
```

Dose events are copied from the anchor verbatim, because moving a dose invents a
regimen the protocol never permitted (see section C). So where dosing is
individualised, this number cannot be fixed by any setting — it is a property of
the study, and the honest response is to say so rather than ship.

### Read the near-miss distance, not just the count

Exact-set equality is a harsh test. A patient one missing sample away from a
neighbour is counted exactly like a patient with a genuinely ad-hoc schedule,
and the remedies are completely different.

```{r b1-nearmiss, eval=has_nlmixr2data}
summary(pheno_unique$nearest_set_diff)
```

`nearest_set_diff` is how many observation slots separate a patient from their
nearest neighbour. A median of a few slots on a sparse study is a different
situation from a median of twenty, and the headline count cannot tell them
apart. On one real 21-patient study, 15 of 21 patients had a "unique" schedule
and every one of them was a *single* missing sample from somebody else.

### Say which endpoint is responsible

A schedule is only as shared as its least-shared part, so attribute it:

```{r b1-endpoint, eval=has_xgxr}
attr(skeleton_uniqueness(case1_pkpd, case1_roles, coarsen_time = TRUE),
     "by_endpoint")
```

Here neither endpoint is responsible, because neither is exposed. On a study
where the count is nonzero this table is what tells you whether it is the dense
PK sampling or the sparse biomarker driving it — and on one real study the PK
sampling accounted for all of it and the biomarker for none, which made the fix
obvious and cheap.

`plot_pmx_schedule()` draws the same thing, and a count that looks alarming and
one that is ordinary look completely different on the page:

```{r b1-plot, eval=has_nlmixr2data, fig.height=7}
plot_pmx_schedule(pheno_sd, pheno_roles)
```

## B2. Extremes

Who stands out from the cohort, on follow-up length, dose count, dose magnitude,
or peak `DV`. This one runs on the **synthetic** table and does not read the
source.

```{r b2-flag, eval=has_xgxr}
case1_flags <- flag_identifiable_subjects(case1_synth, case1_roles)
sum(case1_flags$flagged)
sort(table(as.character(case1_flags$outlier_axes[case1_flags$flagged])),
     decreasing = TRUE)
```

One flagged, on follow-up time. **What makes that the right answer is the
comparison group**, and this dataset is the one that shows why it matters:
`case1_pkpd` is a six-arm dose-ranging study running from 3 mg to 300 mg, so
dose magnitude is spread by *protocol*.

```{r b2-arms, eval=has_xgxr}
table(case1_flags$max_dose)
```

Those are the six arms, not six kinds of outlier. Scored against the **whole
cohort**, the top arm sits about 6.4 modified-z units from the median dose, so
all thirty of its patients get flagged for receiving the dose they were
assigned. You can see it by taking the strata away:

```{r b2-unstratified, eval=has_xgxr}
case1_roles_flat <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME", covariates = "WEIGHTB"
)
sum(flag_identifiable_subjects(case1_synth, case1_roles_flat)$flagged)
```

**`flag_identifiable_subjects()` therefore scores each axis within the declared
`strata`.** Within an arm every dose is identical, so nobody stands out on it and
what survives is the patient who is genuinely unusual — while a patient given
twice their own arm's dose is still flagged, because that is a real outlier among
people who could have looked like them.

**This part generalizes past this package.** "Does this patient stand out?" is
meaningless without saying *from whom*, and the cohort is the wrong answer as
soon as a study assigns anything. A screen that ignores assignment reports the
protocol back to you as a privacy finding. Strata holding fewer than five
patients are scored against the cohort instead, since a scale estimated from four
patients describes those four rather than the one being screened.

`pheno_sd` has no strata to declare — its dose is a property of the infant, not a
protocol assignment — so it is scored cohort-wide, and its flags land on
follow-up time:

```{r b2-pheno, eval=has_nlmixr2data}
pheno_flags <- flag_identifiable_subjects(pheno_synth, pheno_roles)
sum(pheno_flags$flagged)
```

**A screen's own statistics need checking too.** This one previously flagged
every patient who stopped dosing early, because its robust scale — the median
absolute deviation (MAD) — is exactly **zero** whenever more than half a cohort
shares one value. On trial data that is the ordinary case, not a degenerate one:
most patients complete the same number of doses. A zero scale makes every
deviation infinitely many scale units, so everything is an outlier. If you build
a screen like this, test it on a cohort where half the values are identical.

`remediate_identifiable_subjects()` acts on what survives that reading, by
dropping or truncating the subjects it is given.

## B3. Values and geometry

Whether a synthetic patient's numbers sit too close to a real patient's, taking
covariates and trajectory together.

```{r b3-proximity, eval=both}
proximity <- function(source, synthetic, roles, label) {
  report <- compare_pmx_proximity(source, synthetic, roles, replicates = 30)
  data.frame(
    dataset = label,
    adversarial_accuracy = round(report$adversarial_accuracy, 3),
    null_lower = round(report$null_lower, 3),
    null_upper = round(report$null_upper, 3),
    per_side = report$n_compared
  )
}
knitr::kable(rbind(
  proximity(case1_pkpd, case1_synth, case1_roles, "case1_pkpd"),
  proximity(pheno_sd, pheno_synth, pheno_roles, "pheno_sd")
))
```

The statistic asks whether each subject's nearest neighbour lies in its own
dataset or the other one. 0.5 means a synthetic subject is no more like a real
subject than two real subjects are like each other; toward 0 means memorisation.

**Read the null interval before the point estimate.** It is wide at
pharmacometric cohort sizes, and a value inside it means *nothing was detected*,
never *nothing is there*. What it reliably catches is a blatant leak: hand this
function a verbatim copy of the source and it drives to zero and objects, which
is what the package's own regression test requires of it.

**Read the direction before reacting to the miss.** The scorecard marks this row
*review* whatever it lands on, because falling outside the interval means two
opposite things. **Below** it is memorisation — synthetic subjects sitting
closer to real ones than real ones sit to each other — and is the failure this
statistic exists to find. **Above** it means the two sets have separated, which
costs utility and discloses nothing. At a few dozen patients either can also be
the seed.

## B4. Exact copies

The crudest check, and worth keeping precisely because it catches a whole class
of plumbing mistakes that the subtle checks miss. Two vectors per subject must
not be reproduced: **B4a**, the observation times, and **B4b**, the `DV` values
recorded at them. They fail independently. A generator can invent a schedule
and then carry a real patient's measurements onto it, or place real
measurements on a new grid, and each is a copy of half the patient.

**What counts as a copy is a vector too few real patients share**, which is the
same rule `min_pattern_share` applies to visit sets. Reproducing a vector a
whole arm shares identifies nobody: on a fixed protocol grid every patient is
sampled at the same times, so a check asking merely "does any synthetic time
vector equal some real one" fails every well-designed study while catching
nothing. It has to ask whether the vector belonged to *one* patient.

`synpmx_scorecard()` computes both rows. Values are sorted before comparison, so a
reordering of the same numbers still counts as a copy — the stricter reading,
and the right one, since the reordering is not what protects the patient.

```{r b4-copies, eval=both}
card <- synpmx_scorecard(case1_pkpd, case1_synth, case1_roles)
card[card$check %in% c("B4a", "B4b"), ]
```

Both zero. The same comparison is worth running on whole covariate rows, which
nothing does for you. And on a study with one protocol grid, B4a is close to
silent by construction, so B4b is the row doing the work.

## B5. Rare categories and rare combinations

**Nothing in the package flags a level that too few *source* patients held, and
it is the most valuable thing missing.** `synpmx_scorecard()` reports the weaker
synthetic-side version below. The mechanism is worth stating first, because it
inverts the intuition that blending protects everything.

`synpmx_avatar()` treats the two kinds of covariate completely differently, and
only one of them is blended:

- **Numeric** covariates become a weighted mean of the donors' values plus
  noise. The result is a new number nobody had.
- **Categorical** covariates are `sample()`d from the donors' values. That is not
  blending, and there is nothing to average: a synthetic patient's category is
  always **some real patient's actual category, copied**.
- **`strata`** (arm, dose group) are copied from the anchor, exactly,
  by design — they are protocol facts.

So the question is when a copied category reaches the output, and the answer is
measured rather than assumed. `vignette("avatar-algorithm")`, under step 8, runs
a 40-patient fixture in which one categorical level is held by a varying number
of otherwise-ordinary patients. The result:

**A level held by one patient never reaches the output; a level held by two
does.** By ten holders the output tracks the source frequency closely. The
mechanism is geometric rather than protective — the sole holder of a level sits
alone on its own one-hot axis, is nobody's nearest neighbour, and is therefore
almost never chosen as a donor — so **it is not something to rely on.** Nothing
enforces it, nothing reports it, and it degrades smoothly.

**The dangerous shape is a rare level on *typical* patients.** A patient who is
unusual on their covariates is nobody's nearest neighbour and is
self-protecting; a patient who is ordinary in every way except one rare
category — an ultra-rare mutation status, say — sits in the middle of the
profile space, is selected as a donor constantly, and their category is copied
out. Being an outlier protects you. Being ordinary except in one respect does
not.

**And the disclosure is not "this value is unique in the dataset".** It is that
the value may be rare **in the world**. If five people alive carry a mutation, a
synthetic dataset containing it discloses that someone with that mutation was in
this study, which for a named trial with public inclusion criteria can be nearly
identifying on its own. No cohort size helps.

### The checks to run, most of which do not exist yet

1. **Level census, source against synthetic.** For every categorical covariate,
   tabulate the levels on both sides, and report any level reaching the output
   together with **how many source patients held it**. A level held by one or two
   real patients is the whole risk, and it is invisible in a marginal
   distribution comparison that only checks the proportions look similar.
   `compare_pmx_rare_levels()` does this over `strata` and every non-numeric
   covariate, and the scorecard reports it as **B5b**; it is worked below.

   Half of this exists already:
   `compare_pmx_distributions()$covariates_categorical` is a per-patient level
   census of source against synthetic. Two things keep it from answering the
   question. It reports no threshold, so a one-holder level and a thirty-holder
   level look alike in the table; and it loops over declared `covariates` only,
   so **`strata` are absent from it** — which is the one categorical axis copied
   from the anchor verbatim, and therefore the one where a two-patient cell is
   reproduced exactly. That is why the chunk below has to be written by hand for
   `TRTACT`.
2. **Minimum holders, as a threshold.** Flag any level in the output that fewer
   than `min_pattern_share` source patients held — the same number that already
   protects visit sets, for the same reason.
3. **Cross-tabulation with `strata`.** Arm x category, then arm x
   category x any other categorical. Arms are copied exactly, so any joint cell
   involving one is as rare as its rarest part.
4. **Numeric covariates: is anyone in a cluster of their own?** A
   nearest-neighbour distance in covariate space, source against synthetic.
   `compare_pmx_proximity()` answers this over the *whole* profile and would
   mask a covariate-only effect.
5. **Levels the user declares rare in the population.** The four above can only
   see the dataset. Rarity in the world has to be declared, and nothing in
   `pmx_roles()` accepts such a declaration today. It is also the only one where
   the right answer may be to **suppress or coarsen the level** rather than
   report it: collapse a specific variant to "mutation present", or drop the
   column.

Checks 1 to 4 are mechanical. Check 5 changes what the generator does rather
than what it reports.

Check 1 is `compare_pmx_rare_levels()`. It censuses every categorical axis the
roles declare — `strata` and each non-numeric covariate — and the level set it
reports is the **union** of both sides rather than the synthetic side alone: a
level that was dropped on the way out is a coverage failure, and a census that
cannot see it is only checking one of its two directions.

```{r b5-census, eval=has_xgxr}
compare_pmx_rare_levels(case1_pkpd, case1_synth, case1_roles)
```

Read the source column first. A level is **exposed** when fewer than
`min_pattern_share` real patients held it — the same floor that already protects
visit sets, for the same reason — and the row to act on is an exposed level that
`reached` the output, because that level is one real patient's attribute,
copied. On `case1_pkpd` nothing is exposed: every arm is held by 30 patients, so
nothing in this study is at risk on this axis. A study with a two-patient
stratum would show it here and nowhere else.

The remedies are upstream of generation rather than after it: drop the covariate
from `covariates`, or collapse its rare levels before generating. (The synthetic
column moves because anchors are sampled with replacement — section C.) A `0` in
the synthetic column is the other failure: a level the source had and the output
lost.

### The one line of it that can be checked without the source

The census above reads real patient data and is therefore restricted output.
One weaker question can be asked of the released table alone: **does any
categorical level sit on exactly one synthetic patient?** A level held by a
single synthetic patient singles that patient out within the release, whatever
the source looked like. `synpmx_scorecard()` reports it as B5a, over `strata` and
every non-numeric covariate:

```{r b5-singletons, eval=has_xgxr}
card[card == "B5a", ]
```

The row names the column and the level it found, not just the count, because
"1" on its own gives a reader no way to find the patient it is talking about —
and the column it names is where to look, which is what its `explore` call
tabulates. Here the answer is an arm size, thirty, and the pass is any answer
above one. `pheno_sd` declares no categorical covariate and no `strata`, so the
row is absent from its scorecard entirely — which is *not applicable*, not
*clean*.

**Two things this does not do**, and they are the reason the checks above it
stay on the list. A level held by two source patients can be copied onto ten
avatars and pass this check while the disclosure is real, since what matters is
how many *real* patients held it. And a level held by thirty source patients can
land on one avatar by chance and fail it harmlessly. The synthetic-side count is
what you can compute on a table that has left the trusted environment; the
source-side count is what the risk is actually made of.

### What none of this can do

Enumerate combinations. With `d` covariates there are `2^d` subsets that could
serve as a quasi-identifier — arm x sex x age band x mutation — and an attacker's
chosen combination need not be one you checked. One mitigation is already
accidentally present: each covariate is sampled **independently**, so sex may
come from one donor and race from another, and a rare *joint* combination is less
likely to be reproduced whole than any of its parts. That is a fidelity cost
(real correlations between covariates are broken) doing double duty as a weak
privacy benefit, and it should be read as such rather than relied on.

### This is the strongest argument for the differentially private modes

It is the clearest difference between the two families in this package. AVATAR's
protections are **enumerated** — visit sets, dose schedules, extremes, proximity
— so a risk nobody named is a risk nobody covered, and this section is exactly
such a risk. An (epsilon, delta) differential privacy (DP) guarantee is
**unenumerated**: it bounds the influence of any one individual on *any*
function of the output, so it holds for combinations nobody thought to check,
including ones constructed later with data that does not exist yet.

Concretely, the DP path never copies a category. `pmx_covariate(levels = )`
releases a noisy count vector over the levels, normalises it to probabilities,
and generation draws from that distribution; a level held by two patients of
forty is swamped by noise at any sensible epsilon. Three honest caveats:

1. **The level set must be public.** A domain derived from the data leaks the
   domain, which is why `pmx_covariate()` requires `levels` and a citation for
   where that public knowledge came from. "The mutations observed in this trial"
   is itself disclosive.
2. **Delta matters more than usual here.** (epsilon, delta)-DP permits the
   guarantee to fail with probability delta, and for a category held by one or
   two people that is the failure that matters. Keep delta well below 1/n.
3. **DP bounds the mechanism, not the world.** If inclusion criteria are public
   and only a handful of people could qualify, membership may be inferable
   regardless.

The [privacy article](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html) works through that trade-off.

## Deterministic proxies

A column that is a function of a covariate discloses that covariate exactly, and
every dataset has candidates. This is not a scorecard row, because there is
nothing to run: the generator handles the one case it can detect, and the rest
is a reading of your own columns.

The worked case is dosing. Under milligram-per-kilogram dosing, an `AMT` copied
from the anchor reveals the anchor's weight to the gram — the blending of the
weight column is irrelevant, because the dose gives it back. `synpmx_avatar()`
detects proportional dosing and recomputes the amount from the avatar's own
blended covariate, and reports which path it took:

```{r b6-dose, eval=both}
attr(case1_synth, "pmx_settings")$dose_basis_note
attr(pheno_synth, "pmx_settings")$dose_basis_note
```

Neither study is dosed proportionally to a declared covariate, so inference
declines rather than imposing a rule the study did not use — and says why. When
it does fire, `dose_basis` names the covariate.

**The general rule is worth stating**, because detection covers only the dose:
body surface area from height and weight, body mass index, creatinine clearance,
any derived exposure column carried through with `keep`. Each is a covariate
written down twice, and masking one copy does nothing. Look for them by hand.

# C. Is it still the same study?

The plausibility tier. These are not privacy checks, and failing them endangers
nobody — it produces data nobody can develop against, which is the entire
purpose of the package.

## Semantic ordering

A trough sample must stay a trough. The carrier of this check is time after dose
(`TAD`), and there is something about it that surprises people:

**`tad` is an output, not an input.** `synpmx_avatar()` recomputes it from the
generated times and dose rows and overwrites whatever the source held. Declaring
the role says which column to overwrite and carry through; the source's values
never generate anything.

`validate_pmx()` on the **source** is the one place they are read, and a
disagreement there is a real finding every time:

```{r c-tad, eval=has_nlmixr2data}
nimo <- get(utils::data(list = "nimoData", package = "nlmixr2data"))
nimo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
  covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
)
nimo_checks <- validate_pmx(nimo, nimo_roles)$checks
cat(nimo_checks$message[nimo_checks$check == "tad_agreement"])
```

Forty-five percent of observation rows disagree, by up to 311.9 hours. That
means one of: the study measures `TAD` from the end of an infusion, or from a
nominal dose time, or from an assigned occasion rather than the most recent
dose; or the source column is wrong; or the derivation is wrong for this study.
The check cannot tell you which, and says so rather than guessing — but the
synthetic column will follow the derivation, so a silent change of convention is
exactly what this category exists to surface.

Two limits: a sample taken before any dose is reported at `TAD` 0 because
`validate_pmx()` refuses a negative, not because it is genuinely zero hours after
a dose it precedes; and where `addl`/`ii` are declared the derivation cannot see
the doses they imply, so the check is skipped and says so.

**Still missing:** dose/observation *ordering* and occasion assignment have no
check at all. `TAD` covers the first only indirectly. This is the gap that would
have caught two failed attempts at a dose-authoritative time grid in this
package, both of which put a dose *before* the sample that preceded it.

## Regimen validity

No invented regimen. Dose counts, intervals and amounts must be ones the
protocol permitted, which is why `synpmx_avatar()` never resamples dose times and
only ever truncates a real schedule at one of its own dose times. A generator
that jitters dose times will produce regimens no protocol allows, and it will
look fine in every distributional check.

## Endpoint coverage, and coverage lost

All endpoints present is category A. Whether each endpoint is present *for the
right patients* is this one:

```{r c-coverage, eval=has_xgxr}
observed_by_arm <- function(data, label) {
  observed <- data[data$EVID == 0, ]
  out <- as.data.frame.matrix(table(observed$TRTACT, observed$NAME))
  out$dataset <- label
  out$arm <- rownames(out)
  out[, c("dataset", "arm", setdiff(names(out), c("dataset", "arm")))]
}
knitr::kable(rbind(
  observed_by_arm(case1_pkpd, "source"),
  observed_by_arm(case1_synth, "synthetic")
), row.names = FALSE)
```

Two things to read here. **The placebo arm has no PK concentration in either
table**, which is correct and is the check passing: an avatar never leaves the
arm it was anchored in, so it cannot acquire an endpoint its arm never had.

**And the arm sizes match**, which they did not until this check was written.
Anchors are sampled with replacement, so left alone the arm sizes are whatever
the draw gives: this design of exactly 30 per arm came back between 21 and 39 per
arm depending only on the seed, with cohort size and arm membership both exact
and only the *balance* wrong. `preserve_strata_balance` (default `TRUE`) now
gives each declared stratum its source share. Turning it off shows what the
check was catching:

```{r c-balance, eval=has_xgxr}
unbalanced <- suppressWarnings(
  synpmx_avatar(case1_pkpd, case1_roles, seed = 808,
                preserve_strata_balance = FALSE)
)
table(unbalanced$TRTACT[!duplicated(unbalanced$ID)])
```

One deliberate exception: **a stratum holding fewer than three source patients is
left unbalanced on purpose**, because reproducing its size exactly would disclose
that size — a joint cell with one real patient in it would get exactly one avatar
on every seed. `strata_balanced` and `strata_stochastic` in the run settings say
how many strata fell on each side.

Finally, the masking mechanisms buy safety with fidelity, and the cost is
invisible unless reported:

```{r c-masking, eval=has_nlmixr2data}
knitr::kable(as.data.frame(
  pmx_masking_report(pheno_synth, pheno_sd, pheno_roles)
), row.names = FALSE, align = c("l", "r", "l"))
```

A "dose schedule" here, and in the scorecard's C4 row, is **the set of dose
times a patient received, and nothing else** — the amounts are not part of it.
That is deliberate: on a weight-based study every patient's milligrams are their
own, so counting amounts would make every patient their own schedule and the
number would measure nothing. Two patients on the same visit days at different
doses are one schedule; a patient who stopped after two of three doses is
another, and if nobody shares it, declining to anchor on them is what removes it
from the output.

Three rows of that table are the coverage lost, and on `pheno_sd` they are
severe: **16 of 56 distinct dose schedules are represented**, three visit sets
were discarded as too rare to reuse, and 41% of avatars had their missed visits
placed fresh rather than copied. That is the same finding section A reached by
counting dose rows, arrived at from the generator's side.

The reason this belongs in a report rather than in a reader's head: on one study
a cohort of nineteen patients on three dose levels, one on two and one on one
came out as twenty-one patients on three levels — the two smallest regimens
silently gone — and nothing said so until this report was made to print
"1 of 3 regimens represented".

# D. How close are the distributions?

The utility tier: the covariates and the dependent variables.

```{r d-distributions, eval=has_xgxr}
case1_dist <- compare_pmx_distributions(case1_pkpd, case1_synth, case1_roles)
knitr::kable(case1_dist$endpoints, digits = 3,
             caption = "Endpoints, source against synthetic")
knitr::kable(case1_dist$covariates_numeric, digits = 2,
             caption = "Baseline weight, source against synthetic")
```

**Between-subject variability shrinks, necessarily, and this should be expected
rather than discovered.** An avatar is an average of several donors, and
averaging reduces variance. Baseline weight above has a source standard
deviation of about 20.5 kg and a synthetic one of about 14.7. The quantities
that explain how much are reported with the run:

```{r d-blending, eval=has_xgxr}
unlist(attr(case1_synth, "pmx_settings")[
  c("k", "mean_effective_donors", "max_donor_weight", "cap_binding_fraction")
])
```

`mean_effective_donors` is how many donors an avatar is genuinely averaged over
once the weights are accounted for — under 3 here, out of `k = 5` — and
`cap_binding_fraction` is how often the single-donor weight cap bound. Fewer
effective donors means more fidelity and less masking; that trade is the whole
design.

**Do the covariate–response relationships survive?** The question a modeller
actually cares about, and the least covered by anything above: a weight–exposure
slope, a dose–response relationship, a treatment effect. Marginal distributions
can match while the relationship between them is destroyed. The
[covariate and treatment effect article](https://iamstein.github.io/synpmx/articles/example-avatar-PKPD-covariate-treatment-effect.html)
is the worked evidence.

**Both failure directions matter.** Too close is a privacy failure; too far is a
utility failure. `compare_pmx_proximity()` is the one diagnostic that reports
both from a single statistic — near 0 is memorisation, near 1 means the
synthetic cohort is nothing like the real one, and 0.5 is the target. Most of
these checks have two tails, and a check with only one is usually a check that
has picked a side without saying so.

# E. What these checks cannot tell you

A section, not an afterthought.

**Nothing here bounds what an adversary learns.** These checks reduce the ways a
real patient can be singled out. They do not limit how often that succeeds, and
they compose no better than the weakest one. That is what the differentially
private modes are for.

**The guarantees are about reproduction, not similarity.** `identifying_visit_sets`
counts avatars wearing a visit set that is *exactly* one real patient's. An
avatar whose visits sit merely *near* a real patient's is not covered by it, and
`nearest_set_diff` is the only thing that will tell you how near.

**Passing on every dataset you have is not proof.** One change to this package's
time coarsening measured strictly better on all five public datasets available
at the time and still collapsed three real visits into one on a denser fixture.
Public data is a floor.

**Most of these read the source, and are therefore restricted output.** They may
not leave the environment the source data lives in. `compare_pmx()` reports this
per component rather than leaving it to be inferred:

```{r e-release, eval=has_xgxr}
compare_pmx(case1_pkpd, case1_synth, case1_roles)$release_status
```

Only two things in this vignette do not read the source: `validate_pmx()` on the
synthetic table, and `flag_identifiable_subjects()` on the synthetic table.
Everything else — every uniqueness count, every distribution comparison, every
proximity statistic — is computed from real patient data and inherits its
handling obligations.

# F. Check the output, not the algorithm

If this vignette teaches one thing to somebody building their own generator, it
is this one, and the evidence is a single day of debugging in which every leak
found sat behind a mechanism that was correct on its own terms:

- The routine that places an avatar's missed visits rejected identifying
  placements exactly as designed — and the caller then copied the anchor's own
  set when it returned nothing.
- The pool of reusable visit sets was built exactly as designed — and a schedule
  group with one member never got one, so there was nothing to draw from.
- The routine that applies a drawn visit set returned the untouched skeleton
  when the set could not be built on its anchor, so the caller believed a set had
  been applied while the avatar quietly kept its own.

Each was locally right and jointly wrong. **The fix that generalizes is not any
of the individual patches.** It is asking the question of the *finished table* —
does any synthetic patient hold a visit set that fewer than `min_pattern_share`
real patients share? — and recording the answer as a number that must be zero.
That is what `identifying_visit_sets` is, and it would have caught all three
without knowing anything about the mechanisms that caused them.

Two corollaries, both learned the hard way:

**A check written in the mechanism's own vocabulary inherits its blind spots.**
The first end-to-end check here compared recorded times directly and reported
"nothing identifying" on a dataset where everything is — because generation
resamples the very deviations that coarsening removed, so an avatar in the same
grid cell never holds the same number. The second snapped back to the grid and
produced a false alarm on a different dataset, because those deviations are
comparable to grid spacing on an irregular study. Only the third — recording, per
avatar, what was actually applied — was both exact and quiet.

**Write the check where the risk is, not where the code is.** The three bullets
above are three different functions. The number that catches all of them is
computed after all three have run, from the table they produced, in the
vocabulary of the thing you are trying to prevent.

# What is still missing

Stated plainly, because a checklist that implies it is complete is worse than
one that does not. These are the rows marked *gap* in the scorecard at the top,
with what closing each would take:

| # | Gap | What closing it takes |
|---|---|---|
| B5 | The source-side census (check 1) now exists as `compare_pmx_rare_levels()` and scorecard row B5b. Still missing: joint combinations, since with `d` covariates there are `2^d` of them and the census is per column; and check 5, which is the one that matters most — declaring which levels are rare *in the population* rather than in this study | Joint rarity needs a decision about which combinations are worth enumerating. Population rarity is not measurable from the study at all: it needs domain knowledge, so the realistic form is a declaration on `pmx_roles()`, or simply not carrying that covariate |
| C2 | Dose/observation ordering and occasion assignment have no check | A post-generation invariant over the finished table, in the spirit of section F |
| A5 | Dose-count fidelity is not reported. `pheno_sd` went from a median of twelve doses per patient to one, and no warning fires — the count above is the only way to see it | A row in `pmx_masking_report()` |

All but one are reporting gaps: they change what you know, not what the
generator does. The exception is the fifth B5 check — declaring which levels are
rare in the population — which would change generation itself, by suppressing or
coarsening a level rather than reporting it.

Two further gaps are not in the scorecard because they are not checks the
package half-does — they are whole measurement approaches it does not attempt,
and both are covered in the
[checking literature review](https://iamstein.github.io/synpmx/articles/synthetic-data-checking-review.html):

- **No control group.** Every check above that reads the source compares the
  synthetic data against patients the generator was allowed to use. That
  confounds *the generator captured the population* with *the generator
  memorized a patient*, and the standard remedy — holding patients out of
  generation and asking the question twice — is not implemented. It is the
  reason B3's null interval says less than it appears to.
- **Nothing measures linkability, and nothing is per patient on the privacy
  side.** The AVATAR literature's own per-patient measures, local cloaking and
  hidden rate, apply directly to this generator and are not computed.

And one check is not on this list because it cannot be: **does the pipeline that
will consume the real study run unchanged against this?** That is the reason the
package exists, no function here can answer it, and only you can — run your own
code against the output and see whether the joins, reshapes, derivations and
control streams behave.

## Where to go next

- `vignette("demo")` — the scorecard at the top of this vignette, computed on
  one real run.
- `vignette("avatar-algorithm")` — how the default generator works, and the six
  masking mechanisms these checks are measuring.
- `vignette("public-data-examples")` — these checks run across eight
  public datasets, with the masking cost for each.
- [Literature: checking synthetic data](https://iamstein.github.io/synpmx/articles/synthetic-data-checking-review.html) — the published methods for
  checking synthetic data, what each one asks, and which of them this package
  does not attempt.
- [Literature: generating synthetic data](https://iamstein.github.io/synpmx/articles/synthetic-data-generation-review.html) — the four families of
  generation method, and where `synpmx` sits among them.
- [Privacy](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html) — when the enumerated protections are not enough,
  and what a formal guarantee buys instead.
