---
title: "Demo: Using synpmx AVATAR with 5 datasets"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Demo: Using synpmx AVATAR with 5 datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")
```

This vignette compares five source datasets against their synthetic
counterparts, so it needs a few helpers to pull observed rows into a common
shape and plot them. They are ordinary plotting code, not part of the package
API; expand the block if you want to reuse them.

<details>
<summary>Plotting helpers used throughout this vignette</summary>

```{r plot-helpers}
observed_plot_data <- function(data, roles, dataset,
                               clock = "study_time") {
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  if (!is.null(roles$mdv)) {
    observed <- observed & as.character(data[[roles$mdv]]) %in% c("0", "0.0")
  }
  observed <- observed & !is.na(data[[roles$dv]])
  observation_rows <- which(observed)
  occasion <- rep(1L, length(observation_rows))
  tad <- rep(NA_real_, length(observation_rows))
  if (!is.null(roles$occasion)) {
    declared <- suppressWarnings(as.integer(
      data[[roles$occasion]][observation_rows]
    ))
    valid <- !is.na(declared) & declared >= 1L
    occasion[valid] <- declared[valid]
  }
  if (!is.null(roles$tad)) {
    declared <- suppressWarnings(as.numeric(data[[roles$tad]][observation_rows]))
    valid <- is.finite(declared)
    tad[valid] <- pmax(0, declared[valid])
  }
  subject_values <- data[[roles$id]]
  for (id in unique(subject_values[observation_rows])) {
    subject_rows <- which(!is.na(subject_values) & subject_values == id)
    events <- !(as.character(data[[roles$evid]][subject_rows]) %in%
                  c("0", "0.0"))
    if (!is.null(roles$amt)) {
      events <- events & as.numeric(data[[roles$amt]][subject_rows]) > 0
    }
    positions <- which(subject_values[observation_rows] == id)
    event_rows <- subject_rows[events]
    if (length(event_rows) && !is.null(roles$occasion)) {
      event_occasion <- suppressWarnings(as.integer(
        data[[roles$occasion]][event_rows]
      ))
      for (position in positions) {
        candidates <- event_rows[event_occasion == occasion[position]]
        if (length(candidates) && !is.finite(tad[position])) {
          origin <- min(as.numeric(data[[roles$time]][candidates]))
          tad[position] <-
            as.numeric(data[[roles$time]][observation_rows[position]]) - origin
        }
      }
    } else if (length(event_rows)) {
      dose_times <- sort(unique(as.numeric(data[[roles$time]][event_rows])))
      occasion[positions] <- pmax(1L, findInterval(
        as.numeric(data[[roles$time]][observation_rows[positions]]),
        dose_times
      ))
      occasion[positions] <- pmin(occasion[positions], length(dose_times))
      tad[positions] <-
        as.numeric(data[[roles$time]][observation_rows[positions]]) -
        dose_times[occasion[positions]]
    }
  }
  plotted_time <- if (identical(clock, "tad")) tad else
    as.numeric(data[[roles$time]][observation_rows])
  data.frame(
    dataset = factor(dataset, levels = c("Source", "Synthetic")),
    subject = as.character(data[[roles$id]][observation_rows]),
    time = plotted_time,
    dv = as.numeric(data[[roles$dv]][observation_rows]),
    occasion = occasion,
    endpoint = if (is.null(roles$dvid)) "DV" else
      as.character(data[[roles$dvid]][observation_rows]),
    stringsAsFactors = FALSE
  )
}

tad_plot_data <- function(data, roles, dataset) {
  out <- observed_plot_data(data, roles, dataset, clock = "tad")
  names(out)[names(out) == "time"] <- "tad"
  out
}

demo_design_summary <- function(data, roles, dataset, time_bounds,
                                clock = "study_time") {
  plotted <- observed_plot_data(data, roles, dataset, clock)
  if (!identical(clock, "tad")) {
    plotted <- plotted[
      plotted$time >= time_bounds[1L] & plotted$time <= time_bounds[2L],
      , drop = FALSE
    ]
  }
  cohort_ids <- as.character(unique(data[[roles$id]]))
  pieces <- lapply(sort(unique(plotted$endpoint)), function(endpoint) {
    rows <- plotted[plotted$endpoint == endpoint, , drop = FALSE]
    counts <- table(factor(rows$subject, levels = cohort_ids))
    data.frame(
      dataset = dataset,
      endpoint = endpoint,
      patients = length(cohort_ids),
      patients_with_endpoint = sum(counts > 0),
      observations = nrow(rows),
      mean_time_points_per_patient = mean(counts),
      median_time_points_per_patient = stats::median(counts),
      first_time = min(rows$time),
      last_time = max(rows$time),
      stringsAsFactors = FALSE
    )
  })
  do.call(rbind, pieces)
}

check_demo_similarity <- function(source, synthetic, roles, time_bounds,
                                  label, clock = "study_time") {
  source_subjects <- length(unique(source[[roles$id]]))
  synthetic_subjects <- length(unique(synthetic[[roles$id]]))
  if (source_subjects != synthetic_subjects) {
    stop(label, ": source and synthetic patient counts differ (",
         source_subjects, " versus ", synthetic_subjects, ").")
  }
  source_summary <- demo_design_summary(
    source, roles, "Source", time_bounds, clock
  )
  synthetic_summary <- demo_design_summary(
    synthetic, roles, "Synthetic", time_bounds, clock
  )
  if (!setequal(source_summary$endpoint, synthetic_summary$endpoint)) {
    stop(label, ": source and synthetic endpoint sets differ.")
  }
  paired <- merge(
    source_summary, synthetic_summary, by = "endpoint",
    suffixes = c("_source", "_synthetic")
  )
  point_difference <- abs(
    paired$mean_time_points_per_patient_synthetic -
      paired$mean_time_points_per_patient_source
  )
  point_allowance <- pmax(
    1, 0.25 * paired$mean_time_points_per_patient_source
  )
  if (any(point_difference > point_allowance)) {
    stop(label, ": mean time points per patient differ materially for ",
         paste(paired$endpoint[point_difference > point_allowance],
               collapse = ", "), ".")
  }
  coverage_allowance <- 0.20 * diff(time_bounds)
  bad_coverage <-
    abs(paired$first_time_synthetic - paired$first_time_source) >
      coverage_allowance |
    abs(paired$last_time_synthetic - paired$last_time_source) >
      coverage_allowance
  if (any(bad_coverage)) {
    stop(label, ": source and synthetic time coverage differs materially for ",
         paste(paired$endpoint[bad_coverage], collapse = ", "), ".")
  }
  summary <- rbind(source_summary, synthetic_summary)
  summary$dataset <- factor(
    summary$dataset, levels = c("Source", "Synthetic")
  )
  summary[order(summary$dataset, summary$endpoint), , drop = FALSE]
}

comparison_facets <- function(data) {
  ggplot2::facet_grid(dataset ~ endpoint, scales = "free_y")
}

source_synthetic_preview <- function(source, synthetic, n = 6L) {
  columns <- intersect(names(source), names(synthetic))
  source_rows <- utils::head(source[, columns, drop = FALSE], n)
  synthetic_rows <- utils::head(synthetic[, columns, drop = FALSE], n)
  source_rows$.dataset <- "Source"
  synthetic_rows$.dataset <- "Synthetic"
  out <- rbind(source_rows, synthetic_rows)
  rownames(out) <- NULL
  out[, c(".dataset", columns), drop = FALSE]
}
```

</details>

## What this package is for

`synpmx` creates **synthetic pharmacometric data for model-workflow
exploration**: data to develop and debug cleaning, joins, reshaping, plotting,
control-file plumbing, and repeated-dose or longitudinal analysis code, so that
a pipeline runs unchanged against the real data later.

The package offers four generation modes, and the "Introduction to
synpmx" vignette applies all four to one dataset side by side:

- **AVATAR blending** [1, 2] (`synpmx_avatar()`) — build each synthetic subject
  from real subjects. No elicitation, no formal privacy guarantee.
- **Prior only** (`synpmx_prior()` on a public structural model) — read no data
  at all.
- **Calibration** (`synpmx_calibrated()`) — simulate from a public structural
  model whose magnitude is corrected by a small differentially private release.
- **Empirical** (`synpmx_empirical()`) — release a dense set of differentially
  private summaries and rebuild subjects from them.

"AVATAR" is a method name rather than an initialism, from the patient-centric
*avatarization* literature: the original method is due to Guillaudeux and
colleagues [2], and Destere and colleagues benchmark a modified AVATAR for
population PK [1]. This package implements an AVATAR-*inspired* adaptation for
longitudinal event tables, not published AVATAR software.

**Most of this vignette is about the default, AVATAR blending**, applied to five
public datasets with different structural challenges. For each generated subject
it samples a compatible source subject's event skeleton as a template, then
fills the covariates and endpoint trajectories with a distance-weighted blend of
similar subjects, plus subject and residual noise. The output has the source
schema, fresh identifiers, and the same cohort size. The final section runs the
model-based path on the same theophylline data, as `scripts/demo_nlmixr2data.R`
does for all five.

AVATAR output is **not** anonymous and carries **no formal privacy guarantee**.
It is synthetic data that stays under the source data's access controls and
confidentiality obligations, in the same spirit as Novartis's `synadam` (which
resamples each column marginally from the data). Those obligations follow the
data rather than the machine, so working with it on a local workstation covered
by the same controls is a supported use. It is not appropriate for parameter
estimation, inference, model selection, or clinical decisions. When the
generated data would reach anyone the source data could not, use one of the
differentially private modes instead; `vignette("synpmx-privacy")` works through
that decision and the tradeoff behind it.

## Shared workflow

Every example follows the same three steps:

1. declare column meanings with `pmx_roles()`;
2. synthesize with `synpmx_avatar()`;
3. validate structure with `validate_pmx()` and compare to the source.

Identifiers, schema, classes, and factor levels are restored on output. The
caller's random-number state is left untouched; reproducibility comes from the
`seed` argument.

## Theophylline: repeated dosing, dose-relative PK

`theo_md` has 12 subjects on a repeated oral regimen with a single
concentration endpoint.

```{r theo, eval=has_nlmixr2data}
data("theo_md", package = "nlmixr2data")
theo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", covariates = "WT"
)
theo_synth <- synpmx_avatar(theo_md, theo_roles, seed = 303)
validate_pmx(theo_synth, theo_roles)$valid
knitr::kable(
  source_synthetic_preview(theo_md, theo_synth),
  caption = "Actual Theophylline rows and synthesized rows"
)
```

`compare_pmx_distributions()` summarizes the dependent variable per endpoint and
each baseline covariate, source against synthetic, so you can see at a glance
that the ranges line up. Expect them to be close in magnitude and shape, not
identical — AVATAR does not reproduce the source distribution exactly. Like
every source-derived diagnostic it is marked restricted and stays inside the
source data's access controls.

```{r theo-distributions, eval=has_nlmixr2data}
theo_dist <- compare_pmx_distributions(theo_md, theo_synth, theo_roles)
knitr::kable(theo_dist$endpoints, digits = 2,
             caption = "Concentration distribution, source versus synthetic")
knitr::kable(theo_dist$covariates_numeric, digits = 2,
             caption = "Baseline weight distribution, source versus synthetic")
```

For a per-subject re-identification check, `flag_identifiable_subjects()` scores
each synthetic subject for being a structural outlier, and
`remediate_identifiable_subjects()` removes or shortens the ones it flags.

Those find subjects that are *extreme*. `skeleton_uniqueness()` finds subjects
that are *alone* — run on the **source**, before generating, it reports how many
others share each subject's observation time vector, observation count, and
event signature. That matters because the event skeleton is copied verbatim from
one anchor, so a subject holding the only copy of its visit schedule passes it to
every avatar anchored on it. `synpmx_avatar(coarsen_time = TRUE)`, the default,
is what collapses the schedule case.

```{r theo-uniqueness, eval=has_nlmixr2data}
skeleton_uniqueness(theo_md, theo_roles)
```

```{r theo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
theo_comparison <- rbind(
  observed_plot_data(theo_md, theo_roles, "Source"),
  observed_plot_data(theo_synth, theo_roles, "Synthetic")
)
ggplot2::ggplot(
  theo_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.35) +
  ggplot2::geom_point(alpha = 0.55, size = 0.8) +
  comparison_facets(theo_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "DV", colour = "Dataset",
    title = "Theophylline: source and synthetic data"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

A dose-relative display makes the coarse curve shape easier to see. AVATAR
preserves the rise-and-fall because it blends real subject profiles rather than
inventing a shape.

```{r theo-tad, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
theo_tad <- rbind(
  tad_plot_data(theo_md, theo_roles, "Source"),
  tad_plot_data(theo_synth, theo_roles, "Synthetic")
)
ggplot2::ggplot(
  theo_tad,
  ggplot2::aes(tad, dv, group = interaction(dataset, subject, occasion),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.18) +
  ggplot2::geom_point(alpha = 0.25, size = 0.7) +
  comparison_facets(theo_tad) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Time after dose (hours)", y = "DV", colour = "Dataset",
    title = "Theophylline: dose-relative profiles"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

## Warfarin: separate PK and PD endpoints

`warfarin` has a lower-case schema, one dose, and two endpoints (`cp` and
`pca`) with factor covariates. Both endpoints and all subjects are retained.

```{r warfarin, echo=FALSE, eval=has_nlmixr2data}
data("warfarin", package = "nlmixr2data")
warfarin_roles <- pmx_roles(
  id = "id", time = "time", dv = "dv", amt = "amt", evid = "evid",
  dvid = "dvid", covariates = c("wt", "age", "sex")
)
warfarin_synth <- synpmx_avatar(warfarin, warfarin_roles, seed = 404)
```

```{r warfarin-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
warfarin_comparison <- rbind(
  observed_plot_data(warfarin, warfarin_roles, "Source"),
  observed_plot_data(warfarin_synth, warfarin_roles, "Synthetic")
)
ggplot2::ggplot(
  warfarin_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.3) +
  ggplot2::geom_point(alpha = 0.5, size = 0.8) +
  comparison_facets(warfarin_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Time (hours)", y = "DV", colour = "Dataset",
    title = "Warfarin: PK (cp) and PD (pca) endpoints"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

## WBC: infusion, a delayed response, and structural screening

`wbcSim` has infusion start/stop events and a study-time white-blood-cell
response with a delayed decline, nadir, and recovery. A few subjects are
followed far longer than the rest.

A patient followed much longer than everyone else stands out, so a synthetic
copy built from them could point back to the real person. To lower that risk,
`synpmx_avatar()` by default does not build an avatar from such an extreme
subject, quietly leaving the very-long follow-ups out of the synthetic data
(`screen = TRUE`). This is an informal safeguard against singling someone out,
not a formal privacy guarantee; ordinary long follow-ups are kept, and
`screen = FALSE` keeps every subject.

```{r wbc, echo=FALSE, eval=has_nlmixr2data}
data("wbcSim", package = "nlmixr2data")
wbc_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID", cmt = "CMT"
)
wbc_synth <- suppressWarnings(synpmx_avatar(wbcSim, wbc_roles, seed = 505))
```

```{r wbc-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
wbc_comparison <- rbind(
  observed_plot_data(wbcSim, wbc_roles, "Source"),
  observed_plot_data(wbc_synth, wbc_roles, "Synthetic")
)
ggplot2::ggplot(
  wbc_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.3) +
  ggplot2::geom_point(alpha = 0.4, size = 0.7) +
  comparison_facets(wbc_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "WBC", colour = "Dataset",
    title = "WBC: delayed decline, nadir, and recovery"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

## NimoData: infusions, occasions, and a dose group

`nimoData` has ten approximately weekly infusions, declared OCC/TAD, and a
nominal dose group (`DOS`). The dose group is carried through with `keep`, which
copies it verbatim from the same subject that supplied the doses, so it stays
coherent with them. The redundant `WGT` column is simply left undeclared, so
AVATAR drops it.

```{r nimo, echo=FALSE, eval=has_nlmixr2data}
data("nimoData", package = "nlmixr2data")
nimo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
  covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
)
nimo_synth <- suppressWarnings(synpmx_avatar(nimoData, nimo_roles, seed = 606))
```

```{r nimo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
nimo_comparison <- rbind(
  tad_plot_data(nimoData, nimo_roles, "Source"),
  tad_plot_data(nimo_synth, nimo_roles, "Synthetic")
)
ggplot2::ggplot(
  nimo_comparison,
  ggplot2::aes(tad, dv, group = interaction(dataset, subject, occasion),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.2) +
  ggplot2::geom_point(alpha = 0.35, size = 0.7) +
  comparison_facets(nimo_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Time after dose (hours)", y = "DV", colour = "Dataset",
    title = "NimoData: dose-relative profiles across ten infusions"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

## Mavoglurant: occasion-reset clock and a carried dose column

`mavoglurant` has one- and two-period profiles, a `TIME` axis that resets within
occasion, an occasion-varying `DOSE`, numeric-coded `SEX`, and infusion rows.
The reset clock validates within ID and occasion. AVATAR copies the whole event
template from one anchor, so `DOSE` stays coherent with the doses when carried
through with `keep`.

```{r mavo, echo=FALSE, eval=has_nlmixr2data}
data("mavoglurant", package = "nlmixr2data")
mavo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  cmt = "CMT", rate = "RATE", mdv = "MDV", occasion = "OCC",
  keep = "DOSE", covariates = c("AGE", "SEX", "WT", "HT")
)
mavo_synth <- suppressWarnings(synpmx_avatar(mavoglurant, mavo_roles,
                                              seed = 707))
```

```{r mavo-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
mavo_comparison <- rbind(
  tad_plot_data(mavoglurant, mavo_roles, "Source"),
  tad_plot_data(mavo_synth, mavo_roles, "Synthetic")
)
mavo_plot <- ggplot2::ggplot(
  mavo_comparison,
  ggplot2::aes(tad, dv, group = interaction(dataset, subject, occasion),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.2) +
  ggplot2::geom_point(alpha = 0.35, size = 0.7) +
  comparison_facets(mavo_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Time after dose (hours)", y = "DV (log scale)", colour = "Dataset",
    title = "Mavoglurant: dose-relative profiles with occasion reset"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
# Mavoglurant concentrations span orders of magnitude, so a log y axis reads
# better. Use xgxr's log scale when available, otherwise a plain log10 scale.
mavo_plot + if (requireNamespace("xgxr", quietly = TRUE)) {
  xgxr::xgx_scale_y_log10()
} else {
  ggplot2::scale_y_log10()
}
```

## The model-based path on the same data

Everything above blends real subjects. The alternative is to fit a public model
to the data through an accounted release and simulate from it, which is what
`scripts/demo_nlmixr2data.R` does for all five datasets. Repeating it here for
theophylline shows how much more must be declared, and what is bought.

The empirical engine measures the trajectory shape from the data through noised
summaries. Every clipping range, contribution limit, and budget share is an
explicit public input. `synpmx_empirical()` (like `synpmx_calibrated()`)
refuses to run until `synpmx_enable_dp_engines()` has been called once in the
session, acknowledging the maintenance status covered in
`vignette("synpmx-privacy")`:

```{r enable-dp, eval=has_nlmixr2data}
synpmx_enable_dp_engines()
```

```{r empirical, eval=has_nlmixr2data}
theo_private <- synpmx_empirical(
  data = theo_md, roles = theo_roles,
  endpoints = list(cp = pmx_endpoint(
    alignment = "dose_relative", transform = "log", shape = "occasion", cmt = 2
  )),
  epsilon = 5, delta = 0,
  bounds = pmx_bounds(
    time = c(0, 170), endpoints = list(cp = c(0, 30)), amt = c(0, 500),
    covariates = list(WT = c(40, 130))
  ),
  public_design = pmx_public_design(
    pmx_schema(theo_md), dose_evid = 101, dose_cmt = 1
  ),
  contribution_limits = pmx_contribution_limits(40, 8, 8, 30, 11),
  budget_allocation = pmx_budget_allocation(
    subject_count = 0.10, event = 0.15, timing = 0.15,
    covariates = 0.10, endpoints = 0.50, censoring = 0
  ),
  seed = 707,
  backend = "public", public_source = TRUE   # theo_md is public; no DP claim
)
validate_pmx(theo_private, theo_roles)$valid
sampling_summary(theo_private)
```

The confidential data is read once, at the moment `synpmx_empirical()` is
called. The release it produced travels with the returned dataset, so any
number of further datasets can be drawn from it as post-processing, without
spending more budget:

```{r empirical-again, eval=has_nlmixr2data}
theo_private_2 <- synpmx_generate(theo_private, seed = 708)   # spends nothing
```

Reach for `synpmx_generate()` rather than calling `synpmx_empirical()` a second
time — a second call is a second release, and its budget has to be composed
with the first.

```{r empirical-plot, echo=FALSE, eval=has_nlmixr2data && has_ggplot2}
theo_private_comparison <- rbind(
  observed_plot_data(theo_md, theo_roles, "Source"),
  observed_plot_data(theo_private, theo_roles, "Synthetic")
)
ggplot2::ggplot(
  theo_private_comparison,
  ggplot2::aes(time, dv, group = interaction(dataset, subject),
               colour = dataset)
) +
  ggplot2::geom_line(alpha = 0.35) +
  ggplot2::geom_point(alpha = 0.55, size = 0.8) +
  comparison_facets(theo_private_comparison) +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(
    x = "Study time (hours)", y = "DV", colour = "Dataset",
    title = "Theophylline: model-based private generation"
  ) +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "bottom")
```

Because `theo_md` is already public, this demonstration uses the guarded
public-fixture backend, which is **noiseless and makes no DP claim** — exactly
as the demo script does. A confidential fit uses the default OpenDP backend and
fails closed when it is unavailable. The `synpmx_calibrated()` mode, which
asserts curve shape from a public structural model and privately calibrates only
the magnitude, is the better choice at this cohort size; the introduction
vignette runs it on this same dataset.

## How well did the obfuscation work?

Everything above shows that the synthetic data *looks* right. This section asks
the other question: how much of each real patient is still visible in it?

### What is being counted, and why

`synpmx_avatar()` blends every avatar's **measurements** from at least five real
donors, so no single patient's numbers survive. But it copies the **event
skeleton** — the pattern of when that patient was dosed and sampled — verbatim
from one real anchor. Blending never touches it.

So the question that matters is: **does any real subject have a visit schedule
that no other subject in the cohort shares?** If yes, an avatar built on that
subject wears a schedule belonging to exactly one real person, however carefully
their concentrations were blended. Call such a subject **alone**.

`skeleton_uniqueness()` counts them, and `coarsen_time = TRUE` (the default) is
the mechanism that tries to fix it, by snapping every subject onto a shared visit
grid before any avatar is built. A subject can end up alone for two reasons, and
they have different answers:

- **Seen at a unique moment.** The subject was observed at a time no other
  subject was observed at all — a one-off visit. The grid is supposed to absorb
  this by pulling nearby times together, so a nonzero count here means the
  inferred grid could not find a shared schedule. **The fix is to declare a
  `nominal_time` role**, which snaps to the real protocol grid instead of a
  guessed one.
- **Unique pattern of visits.** Every one of the subject's times is shared with
  somebody — but the particular *combination* they attended is theirs alone.
  This is dropout, early discontinuation, or a missed sample. **No grid can fix
  it**, however fine or coarse, because the times are already shared. Only
  dropping or remediating the subject helps.

```{r exposure-table, eval=has_nlmixr2data}
exposure_row <- function(label, source, roles, synthetic) {
  before <- skeleton_uniqueness(source, roles)
  after <- attr(synthetic, "pmx_settings")
  data.frame(
    Dataset = label,
    Subjects = nrow(before),
    `Alone before` = attr(before, "n_alone"),
    `Alone after` = after$exposure_alone,
    `Unique moment` = after$exposure_unique_moment,
    `Unique pattern` = after$exposure_unique_pattern,
    check.names = FALSE, stringsAsFactors = FALSE
  )
}

exposure <- rbind(
  exposure_row("theo_md", theo_md, theo_roles, theo_synth),
  exposure_row("warfarin", warfarin, warfarin_roles, warfarin_synth),
  exposure_row("wbcSim", wbcSim, wbc_roles, wbc_synth),
  exposure_row("nimoData", nimoData, nimo_roles, nimo_synth),
  exposure_row("mavoglurant", mavoglurant, mavo_roles, mavo_synth)
)
knitr::kable(
  exposure,
  caption = paste(
    "Source subjects holding a visit schedule no other subject shares,",
    "before and after time coarsening. The last two columns split the",
    "'after' count by cause."
  )
)
```

### Reading the table

**"Alone after" is the number of patients you would consider dropping**, and the
last two columns say why — which decides whether dropping is even the right
response.

- **`theo_md` — nothing to do.** All twelve subjects start out alone, because
  their sampling times are recorded to the minute and no two patients were
  bled at exactly the same moment. Coarsening pulls those onto a shared grid and
  the count goes to zero. This is the mechanism working exactly as intended.
- **`warfarin` — consider dropping 12 of 32, but the reason matters.** Only two
  are alone because of a unique moment; the other ten share every one of their
  times and are alone purely on *which* visits they attended. Warfarin has
  subjects with 13, 14, 16, 17, 18, 19 and 25 observations, and the rarer counts
  single those subjects out. Coarsening cannot help them.
- **`wbcSim` — 17 of 45, split 6 and 11.** The six unique-moment subjects are
  the ones to look at first; declaring a nominal-time column would likely clear
  them.
- **`nimoData` — 12 of 12, and coarsening achieved nothing.** All twelve are
  alone on a unique moment. This is a genuine failure of the inferred grid: with
  ten roughly-weekly infusions per subject over a long follow-up, no two
  subjects were ever dosed or sampled close enough together for the grid to
  merge them. This dataset needs `nominal_time` declared, or its avatars should
  be treated as carrying real visit schedules.
- **`mavoglurant` — 59 of 120, but only one from a unique moment.** The grid did
  essentially all it could. The remaining 58 differ in which of the shared
  visits they attended.

### What the other mechanisms leave behind

Two of the six mechanisms act on the columns this table does not show, and it is
worth knowing which residual belongs to which.

**Dose.** Under dosing proportional to a baseline covariate, each avatar's `AMT`
is recomputed from its *own* blended weight rather than copied from its anchor,
so the amount is neither a real patient's real dose nor inconsistent with the
avatar's own covariates. `warfarin` is recognised this way (1.5 mg/kg to within
0.1%); `theo_md` is not, because its dose-to-weight ratio varies by 15% and
detection deliberately fails closed. Where the dose sequence is instead
outcome-adaptive — intra-patient escalation driven by a subject's own
tolerability — nothing here touches it, because the sequence *is* that subject's
response.

**Attendance.** The `Unique pattern` column above is what
`min_pattern_share` (default 3) addresses: an avatar's set of attended visits is
drawn from patterns at least that many source subjects share, so no
one-of-a-kind pattern is reproduced. Nobody is dropped to achieve it — a subject
with a rare pattern still contributes measurements as a donor. The cost is real
and dataset-dependent: on `warfarin` and `theo_md` only one pattern clears a
floor of 3, so every avatar becomes a complete attender and the cohort's
missingness distribution collapses. A floor of 2 keeps all three of `theo_md`'s
patterns and 8 of `wbcSim`'s 25. On `nimoData` no pattern is shared by even two
subjects, so there is no pool at all and the run alerts.

### What to do about it

Dropping 59 of 120 subjects is a real cost, and it is not automatic. Under this
package's governance model — AVATAR output stays inside the source data's own
access controls — a unique schedule is a reason for care, not a blocker. The
sequence worth following is:

1. **Declare `nominal_time` if the study has it.** This is free, and it is the
   only thing that helps the unique-moment column. On `nimoData` it is the
   difference between the mechanism working and not working at all.

   A study that has no nominal-time column can often still *construct* one, and
   `nimoData` is a good example because its inferred grid fails completely. Its
   design is ten roughly-weekly infusions with a declared occasion and
   time-after-dose, so the protocol grid is the occasion number times the
   nominal interval, plus the visit's nominal time after dose:

    ```{r nimo-nominal, eval=has_nlmixr2data}
    nimo_nominal <- nimoData
    nimo_nominal$NTIME <- (nimo_nominal$OCC - 1) * 168 +
      ifelse(nimo_nominal$EVID == 0, round(nimo_nominal$TAD / 24) * 24, 0)

    nimo_roles_nominal <- pmx_roles(
      id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
      rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
      nominal_time = "NTIME",
      covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
    )
    nimo_fixed <- suppressWarnings(
      synpmx_avatar(nimo_nominal, nimo_roles_nominal, seed = 606)
    )
    unlist(attr(nimo_fixed, "pmx_settings")[
      c("time_grid", "exposure_alone", "exposure_unique_moment")
    ])
    ```

    Subjects alone fall from 12 of 12 to 5, all in the long washout tail after
    the last dose, where a follow-up visit at 240 or 624 hours belongs to one
    patient. Coarsening that tail further does not help — at a two-week grid the
    count goes back up, because merging genuinely distinct visits creates new
    *pattern* uniqueness. Just as important, attendance sampling now has a pool
    at all: with the inferred grid no two subjects shared a pattern, so
    `min_pattern_share` had nothing to draw from.
2. **Then re-run this table.** If the unique-moment count is at or near zero,
   the grid has done its whole job and no amount of tuning will improve it.
3. **Then decide about the pattern column.** These are dropout and missed-visit
   patterns. `flag_identifiable_subjects()` and
   `remediate_identifiable_subjects()` can drop or truncate them; whether that
   is worth the lost cohort size is a judgment about who will see the output.

Two limits to keep in view. This counts *schedules*, not dose amounts — a study
with weight-based dosing or per-subject titration leaves subjects unique on dose
regardless of what the grid does, which `skeleton_uniqueness()` reports
separately as `signature_class`. And none of this bounds what an adversary
learns; it reduces the ways a real patient can be singled out without limiting
how often that succeeds. That is what the differentially private modes are for.

## What is preserved in AVATAR, and what is not

Preserved: schema, column classes, factor levels, cohort size, endpoint set,
event structure, coarse regimen and sampling timing, and the broad shape and
magnitude of each endpoint.

Not preserved, by design: exact source distributions, parameter estimates,
covariate-response relationships, and rare individual trajectories. Identifiers
are always freshly generated and never reuse a source value.

Not provided: any formal privacy guarantee. The synthetic data is built by
blending real subject trajectories, so it is appropriate wherever the source
data and synthetic data are accessible by the same users, but — but not for release to anyone outside. 

## References

1. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).

2. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).
