---
title: "The four generation modes"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{The four generation modes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)

# Observed rows only, in a common shape, so five datasets with different
# schemas can be plotted and summarized together.
observations <- function(data, roles, label) {
  keep <- as.character(data[[roles$evid]]) %in% c("0", "0.0") &
    !is.na(data[[roles$dv]])
  if (!is.null(roles$mdv)) {
    keep <- keep & as.character(data[[roles$mdv]]) %in% c("0", "0.0")
  }
  data.frame(
    method = label,
    subject = as.character(data[[roles$id]][keep]),
    time = as.numeric(data[[roles$time]][keep]),
    dv = as.numeric(data[[roles$dv]][keep]),
    stringsAsFactors = FALSE
  )
}
```

## What this package does

`synpmx` builds synthetic pharmacometric (PMX) datasets: dosing and
measurement event tables with the same schema, event grammar, and rough
behavior as a real study, so that data-assembly code, diagnostic plots, and
model-run plumbing can be developed outside the restricted environment that
holds the real data.

That narrow purpose is important. A generated dataset is not anonymous data and
is not a formal proof that source subjects cannot be re-identified. It is also not
suitable for estimation, model selection, inference, dose selection, or clinical
decisions. The method aims for structural usefulness, not scientific
equivalence.

Given that synthetic data is oven used as a way to protect privacy, this package offers four synthetic data generation functions that offer different privacy guarantees.   This vignette introduces all four and
applies each one to the same public dataset.

The four modes, from most faithful to most protective:

1. **AVATAR blending** [1, 2] — build each synthetic subject out of real
   subjects.
2. **Prior only** — read no data at all; simulate from a public model.
3. **Calibration** — simulate from a public model whose magnitude is corrected
   by a small, differentially private release.
4. **Empirical** — release a dense set of differentially private summaries and
   rebuild subjects from them.

"AVATAR" is a method name rather than an initialism, from the patient-centric
*avatarization* literature in which each synthetic record is built from the
local neighborhood of real records. The original method is due to Guillaudeux
and colleagues [2]; Destere and colleagues benchmark a modified AVATAR against
other synthesis algorithms for population PK [1]. This package implements an
AVATAR-*inspired* adaptation for longitudinal event tables, not published AVATAR
software.

This vignette stays at the level of what each mode does and when to use it. The
full AVATAR algorithm — every step, the mathematics, the edge cases, and a
worked example — is in the
[AVATAR Algorithm article](https://iamstein.github.io/synpmx/articles/avatar-algorithm.html).

## The example: theophylline

`theo_md` (from `nlmixr2data`) is a small public dataset — 12 subjects, seven
daily 320 mg oral doses, one concentration endpoint. Twelve subjects is a
deliberately hard case: it is typical of early-phase work and it is where the
differences between the four modes are most visible.

```{r data, eval=has_nlmixr2data}
data("theo_md", package = "nlmixr2data")
theo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", covariates = "WT"
)
str(theo_md)
```

## Mode 1: AVATAR blending

Start here, because it is the least work. `synpmx_avatar()` needs nothing but
the data and a declaration of what the columns mean. For each synthetic subject
it copies a real subject's event skeleton, then fills the covariates and
concentrations with a distance-weighted blend of that subject's nearest
compatible neighbors, plus noise.

```{r avatar, eval=has_nlmixr2data}
avatar <- suppressWarnings(synpmx_avatar(theo_md, theo_roles, seed = 101))
validate_pmx(avatar, theo_roles)$valid
```

Nothing was elicited, nothing was assumed, and the output keeps the cohort size
and every column named in `theo_roles`. `synpmx_avatar()` keeps only role-named
columns and drops the rest — so a stray identifier cannot leak out of a real
subject by being forgotten — and here every `theo_md` column has a role, so
nothing is dropped. On a 12-subject dataset it also emits documented small-group
fallback warnings (some event-pattern groups have only one usable donor); they
are suppressed above and explained in the
[AVATAR Algorithm article](https://iamstein.github.io/synpmx/articles/avatar-algorithm.html).

What you cannot say about this output is that it is anonymous. It is assembled
from real trajectories, so it inherits the source data's handling obligations
wherever it is used — a constraint on who may see it, not on which machine holds
it.

### What AVATAR does to obscure the source

AVATAR gives no formal guarantee, so the honest way to describe its protection
is to list the mechanisms and say what each one does and does not cover. There
are seven: five run by default, and two are measurements you run yourself.

**1. Blending across donors.** No synthetic subject's measurements come from one
real patient. Each avatar's covariates and concentrations are a
distance-weighted blend of at least `k` = 5 compatible real donors, and
`max_donor_weight` = 0.50 caps any single donor at half the blend. This is what
protects the *values*. It is also the only mechanism whose strength grows with
cohort size, and it is why a source with fewer than `k + 1` subjects triggers a
loud alert rather than quietly producing near-copies.

**2. Screening structurally extreme subjects** (`screen = TRUE`). A source
subject whose follow-up length or dose count exceeds twice the cohort's 90th
percentile is never used as an anchor, so no avatar inherits a conspicuous
skeleton. Only those two axes are screened at generation, because dose magnitude
is noisy under weight-based dosing and DV is blended rather than copied. After
generation, `flag_identifiable_subjects()` screens four axes — follow-up time,
dose count, dose magnitude, and peak DV — and
`remediate_identifiable_subjects()` truncates, drops, or replaces what it finds.
This catches subjects who are *extreme*.

**3. Coarsening the visit grid, then re-refining it** (`coarsen_time = TRUE`).
Source times are collapsed onto a shared visit grid, and the per-visit
deviations are pooled across the cohort and resampled independently onto each
avatar. This is what protects the *schedule*, which blending does not touch: the
event skeleton is copied verbatim from one anchor, and under actual recorded
times almost every subject holds the only copy of their visit vector. The order
is what makes it work — snapping is many-to-one and destroys the deviation,
where perturbing the original time in place would leave it recoverable. The grid
is the `nominal_time` role where one is declared, and inferred from the pooled
times otherwise; the inferred case is best-effort and alerts loudly when it
cannot collapse a subject. The cost is timing fidelity: an avatar's deviation
from nominal is drawn from the cohort rather than inherited.

**4. Recomputing the dose from a blended covariate** (automatic). When the dose
is a fixed multiple of a baseline covariate within each assigned stratum — mg/kg,
mg/m² — that multiplier is a protocol property the stratum shares, while the
covariate is individual and already blended. So the multiplier is kept and the
amount recomputed from the avatar's *own* blended weight. Previously `AMT` was
copied verbatim while covariates were blended, which both disclosed the anchor's
weight exactly and left every avatar violating its own protocol: a cohort dosed
at exactly 5 mg/kg produced avatars from 4.4 to 5.3. Several dose levels are
found by clustering the observed ratios, so a 1/2/3 mg/kg escalation is
recognised without declaring the arm — and so is *intra-patient* escalation,
where the level changes within a subject. Detection fails closed where the dose
is unrelated to any covariate.

**5. Sampling the attendance pattern** (`min_pattern_share`, default 2). Once
coarsening has put every subject on a shared visit grid, what remains of a
schedule is which of those visits each subject attended. Every time is then
shared, so no single visit is identifying — the *combination of absences* is,
and a patient who missed weeks 2 and 3 is singled out by a fingerprint made of
gaps. No grid fixes this at any resolution, because the grid decides where the
visits are, not which ones a subject has. So the pattern is drawn from ones at
least `min_pattern_share` subjects hold. Nobody leaves the cohort: a subject with
a rare pattern still contributes measurements as a donor, only their distinctive
absences stop being reproduced. Dose events are never sampled, since that could
emit a regimen no protocol permits.

The default of 2 states exactly one thing: no synthetic patient carries a
schedule unique to a real patient. Matching *exact* patterns alone would discard
almost everything, because two patients who each missed one visit count as
different patterns if they missed different visits. So the draw is two-stage: a
**shape** first — how many visits were missed and whether the misses were
terminal, contiguous, or scattered — then a real pattern of that shape if one
clears the floor, and only otherwise a generated arrangement, which is rejected
and redrawn if it lands on a pattern too rare to have been reusable. On
`warfarin` that takes the loss from 12 patterns to 2. What is lost is
resolution: how much missingness and what kind survive, which specific visits
does not. Every run reports the figures, so the trade can be judged per study
rather than assumed.

**7. Measuring how close the values landed** (`compare_pmx_proximity()`,
manual). The measurement for mechanism 1. Blending protects the values, and this
asks whether they landed too close to somebody real: each subject's nearest
neighbour is either in its own dataset or the other one, and under the ideal that
is a coin flip. The null comes from splitting the source cohort in half and
running the identical statistic, so small-sample artefacts cancel. Wide at
pharmacometric cohort sizes — it catches a blatant leak, not a subtle one.

**6. Checking for unique event skeletons** (`skeleton_uniqueness()`, manual).
Reports, per subject, how many others share its observation time vector, its
observation count, and its event signature. Run it on the source before
generating. It is a measurement, not a mitigation — what to do about what it
finds is mechanism 2 or 3 depending on which class is exposed. Observation
*times* are coarsening's job; observation *counts* are the screen's, because no
grid can change a count and what survives coarsening is dropout and missed
visits; dose amount is neither's, so weight-based dosing leaves a cohort unique
on signature regardless. `scripts/measure_skeleton_uniqueness.R` runs the before
and after over every public dataset.

Two things none of these do. They do not bound what an adversary learns, which
is what differential privacy provides and why the trust-boundary question below
decides the mode rather than the mechanism list. And they do not defend against
an attacker who already holds a suspected participant's record and only wants to
confirm membership — the mechanisms reduce the ways that attack succeeds without
limiting how often it does.

## Mode 2: prior only

The opposite extreme. Declare a public structural model and a public protocol,
and simulate. No confidential data is read, so there is nothing to protect and
no budget to spend: this is `epsilon = 0`, the strongest possible guarantee.

The typical parameter values must come from somewhere that is not the data —
allometric scaling from preclinical work, a published model for the compound
class, or the reasoning that set the starting dose.

Before reaching for this mode as a simulator, read
[what the built-in models can and cannot express](#the-built-in-models-are-illustrative-and-deliberately-so).
The catalogue is small on purpose: no covariate effects, no inter-occasion
variability, no additive residual error, no ODE models. It produces a
structurally correct dataset to develop code against, not a faithful rendering
of an arbitrary pharmacometric model.

Others are working in parallel to develop a data simulation 
SKILL.md file to support development of such synthetic data.  This work is
important, but outside the scope of this package.

```{r prior, eval=has_nlmixr2data}
theo_model <- pmx_structural_model(
  pk = "1cmt_oral",
  typical = c(cl = 6, v = 35, ka = 1.5),          # deliberately imperfect
  source = "illustrative allometric scaling; never fitted to theo_md"
)
# theo_md samples richly on the first and last days and takes a single trough
# in between, which is the usual shape for a repeated-dose study. `sampling`
# takes one entry per dose so that can be said directly; a bare vector would
# apply the rich profile after all seven doses and oversample the study
# three-fold.
rich <- c(0, 0.25, 0.5, 1, 2, 4, 7, 9, 12, 24)
theo_design <- pmx_trial_design(
  dose_levels = 320, cohort_sizes = 12,
  sampling = list(rich, 0, NULL, NULL, NULL, NULL, rich),
  n_doses = 7, dose_interval = 24,
  source = "illustrative protocol"
)
theo_design
prior_only <- synpmx_prior(theo_model, theo_design, n_subjects = 12, seed = 202)
head(prior_only, 3)
```

The generated table uses the package's own generated schema
(`pmx_generated_roles()`), including nominal time, time after dose, and
occasion columns. The clearance of 6 L/h assumed above is about twice the truth
for theophylline, and the output shows it: concentrations run low. That is the
honest cost of spending no budget — the data is exactly as good as the prior.

## Mode 3: calibration

The middle path, and the recommended one when a formal guarantee is needed and
the cohort is small. Both `synpmx_calibrated()` and `synpmx_empirical()` refuse
to run until `synpmx_enable_dp_engines()` has been called once in the session —
a deliberate speed bump.

```{r enable-dp, eval=has_nlmixr2data}
synpmx_enable_dp_engines()
```

Keep the public model's *shape*, and spend a small privacy
budget correcting only its *magnitude*.

Each subject is reduced to a bounded multiplicative correction of the model's
own prediction, clipped to a public prior range, and released with calibrated
noise. Only two numbers leave the data: the correction and a noised subject
count.

```{r preflight, eval=has_nlmixr2data}
priors <- pmx_priors(pk = pmx_prior(c(1 / 4, 4), source = "scaling literature"))
pmx_preflight(priors, epsilon = 1, n_subjects = 12)
```

`pmx_preflight()` costs nothing and reads no data: it answers "is this release
worth its budget?" before any budget is spent.

```{r calibrate, eval=has_nlmixr2data}
calibrated_data <- synpmx_calibrated(
  data = theo_md, roles = theo_roles, model = theo_model,
  design = theo_design, priors = priors, epsilon = 1, seed = 303,
  backend = "public", public_source = TRUE   # theo_md is public; no DP claim
)
```

The correction pulled the assumed clearance toward the data. The release that
produced this dataset travels with it, so the accounting is always at hand:

```{r calibrate-report, eval=has_nlmixr2data}
attr(calibrated_data, "synpmx_release")
```

Generation from that release is post-processing, so further datasets cost
nothing. Draw them with `synpmx_generate()` rather than by calling
`synpmx_calibrated()` again — a second fit would spend the budget a second time.

```{r calibrate-again, eval=has_nlmixr2data}
another <- synpmx_generate(calibrated_data, seed = 304)   # spends nothing
```

```{r opendp, eval=FALSE}
calibrated <- synpmx_calibrated(
  data = confidential, roles = theo_roles, model = theo_model,
  design = theo_design, priors = priors, epsilon = 1,
  backend = "opendp"
)
```

```{r backend, eval=requireNamespace("opendp", quietly = TRUE)}
dp_backend_status()
```

At 12 subjects a genuine DP release of this correction is noisy enough
that it is often censored at the prior boundary — the package warns when that
happens, because the generated data then reflects the prior, not the study.
`vignette("synpmx-privacy")` covers when the release is worth making.

## Mode 4: empirical

The general-purpose private engine. Rather than asserting the curve shape, it
measures it: it releases noised summaries for the subject count, event and
regimen structure, observation timing, endpoint trajectories, baseline
covariates, and censoring, then rebuilds subjects from those summaries. This
buys realism that the public model does not contain, and pays for it by
splitting one epsilon across many released quantities.

It also needs the most declaration: every clipping range, contribution limit,
and budget share is an explicit public input.

```{r empirical, eval=has_nlmixr2data}
empirical_data <- synpmx_empirical(
  data = theo_md, roles = theo_roles,
  endpoints = list(cp = pmx_endpoint(
    alignment = "dose_relative", transform = "log", shape = "occasion", cmt = 2
  )),
  epsilon = 5, delta = 0,
  bounds = pmx_bounds(
    time = c(0, 170), endpoints = list(cp = c(0, 30)), amt = c(0, 500),
    covariates = list(WT = c(40, 130))
  ),
  public_design = pmx_public_design(
    pmx_schema(theo_md), dose_evid = 101, dose_cmt = 1
  ),
  contribution_limits = pmx_contribution_limits(40, 8, 8, 30, 11),
  budget_allocation = pmx_budget_allocation(
    subject_count = 0.10, event = 0.15, timing = 0.15,
    covariates = 0.10, endpoints = 0.50, censoring = 0
  ),
  seed = 404,
  backend = "public", public_source = TRUE   # theo_md is public; no DP claim
)
privacy_report(empirical_data)
```

Note the requested `epsilon = 5`, five times the calibrated fit's budget, for a
worse result at this cohort size. That is not a bug: the same budget is being
split six ways over dozens of released coordinates. This engine earns its keep
on large pooled datasets, not on twelve subjects.

## The four side by side

```{r compare, eval=has_nlmixr2data}
generated_roles <- pmx_generated_roles()
all_observations <- rbind(
  observations(theo_md, theo_roles, "Source"),
  observations(avatar, theo_roles, "1. AVATAR"),
  observations(prior_only, generated_roles, "2. Prior only"),
  observations(calibrated_data, generated_roles, "3. Calibration"),
  # The empirical engine restores the source schema, so it uses source roles.
  observations(empirical_data, theo_roles, "4. Empirical")
)
summaries <- do.call(rbind, lapply(
  split(all_observations$dv, all_observations$method),
  function(dv) {
    data.frame(
      n_observations = length(dv),
      median = stats::median(dv),
      p10 = stats::quantile(dv, 0.10, names = FALSE),
      p90 = stats::quantile(dv, 0.90, names = FALSE)
    )
  }
))
knitr::kable(
  summaries, digits = 2,
  caption = "Observed concentrations by generation mode"
)
```

```{r compare-plot, eval=has_nlmixr2data && has_ggplot2, fig.height=6}
ggplot2::ggplot(
  all_observations,
  ggplot2::aes(time, dv, group = subject)
) +
  ggplot2::geom_line(alpha = 0.4, colour = "#1B6CA8") +
  ggplot2::geom_point(alpha = 0.5, size = 0.7, colour = "#1B6CA8") +
  ggplot2::facet_wrap(~ method, ncol = 2) +
  ggplot2::labs(
    x = "Study time (hours)", y = "Concentration",
    title = "One dataset, four generation modes"
  ) +
  ggplot2::theme_minimal()
```

AVATAR tracks the source most closely, because it is made of it. The
prior-only data has the right structure and the wrong level. Calibration moves
the level toward the truth for a small budget. The empirical release recovers
more of the real timing and spread, but at this cohort size the noise is
visible.

## Choosing a mode

```{r modes-table, echo=FALSE}
modes <- data.frame(
  Mode = c("1. AVATAR blending", "2. Prior only", "3. Calibration",
           "4. Empirical"),
  Function = c("synpmx_avatar()", "synpmx_prior()",
               "synpmx_calibrated()", "synpmx_empirical()"),
  `Output built from` = c(
    "Real subject templates and blended real trajectories",
    "A public model and protocol only",
    "A public model, magnitude corrected by 2 private releases",
    "Dozens of noised population summaries"
  ),
  Guarantee = c(
    "None; governance only",
    "epsilon = 0 (no data read)",
    "(epsilon, delta) DP",
    "(epsilon, delta) DP"
  ),
  `Cohort size` = c("Any, from ~5", "Any (data-independent)",
                    "~20 and up", "~200 and up"),
  `Elicitation needed` = c(
    "None", "Structural model + protocol",
    "Model, protocol, prior ranges",
    "Endpoints, bounds, limits, budget split"
  ),
  check.names = FALSE, stringsAsFactors = FALSE
)
knitr::kable(modes)
```

Where each mode belongs:

| Environment | Appropriate modes | Why |
|---|---|---|
| Inside the validated environment holding the source data; you are the only consumer | **AVATAR**, or any other | Access control and governance already bound the risk. A formal guarantee defends against an adversary who cannot reach the output, so it buys nothing and costs utility. |
| Shared with a partner, vendor, or contract research organization (CRO) | **Calibration** or **Empirical**, with an approved epsilon | The output leaves your controls. A contract is not a mathematical bound; DP is what survives a determined recipient. |
| Published, posted to a repository, or shipped inside a package or teaching material | **Prior only**, or **Calibration** with a small approved epsilon | Anyone may inspect it, forever, alongside side information you cannot anticipate. Prior-only data reads no patient record at all and is the safest thing to publish. |
| Software testing where only schema and event grammar matter | **Prior only** | Fidelity is irrelevant; a data-independent generator removes the question entirely. |

Two rules of thumb behind the table:

- **The trust boundary decides the level of privacy needed.** Ask whether the generated
  data can reach anyone the source data could not. A workstation under the same access controls reaches no one new. If no one new, AVATAR is more useful and its lack of a formal guarantee costs
  nothing. If someone new, only an accounted release holds up.
- **The cohort size decides which differential privacy mode is usable.** Epsilon buys accuracy in
  proportion to the number of subjects and in inverse proportion to how many
  quantities you release. At 12 subjects, releasing two numbers can work and
  releasing fifty cannot.

Epsilon and delta are governance decisions, not defaults. For anything public
facing they should be set and justified by whoever owns the data, and recorded:
every fit carries a release ledger, and `privacy_report()` prints the realized
accounting.


## Why AVATAR is the default

Novartis's `synadam` generates synthetic ADaM (Analysis Data Model) datasets by
resampling **each column** marginally from the real data: a uniform draw over
the observed range for continuous columns, a proportional resample for
categorical ones, with no differential privacy. It preserves each column's
marginal support and relies on governance rather than a mathematical guarantee.
That is standard, accepted practice.

AVATAR blending is the same governance-based idea applied at a different
granularity: it resamples and blends **whole subject trajectories** rather than
individual columns, because a pharmacometric endpoint is a correlated
time-course that would be destroyed by independent per-column resampling. If
`synadam`'s privacy model is acceptable for its use, AVATAR's is acceptable for
the same use — output kept under the source data's own access controls, reaching
no one the source data could not.

One caveat follows from the difference in granularity. A resampled
covariate value (a weight of 72 kg) is weakly identifying because many people
share it. A resampled subject trajectory is more strongly identifying because a full sampling-and-response pattern is more nearly
unique. Blending several donors, adding noise, and removing outlying patients mitigates this, but not
formally. AVATAR therefore depends on the governance context somewhat more than
`synadam`'s column resampling does.

## What the model-based modes replace

Modes 2, 3, and 4 do not blend anything. They replace the AVATAR pipeline
entirely: there is no anchor subject, no donor neighborhood, and no event
template. The trial structure comes from a **declared public protocol**
(`pmx_trial_design()` or `pmx_public_design()`) rather than from a source
subject's rows.

Both differentially private engines are **aggregate-based**: no source
subject's rows, template, or trajectory reaches the output. They read the
confidential data only through per-subject contributions clipped to publicly
declared ranges, release those aggregates with calibrated noise, and generate
from the noised numbers alone. They differ in what supplies the curve shape.

`synpmx_calibrated()` takes shape from a **public structural model** — a
closed-form one-compartment or two-compartment PK model, evaluated analytically
rather than by solving ordinary differential equations (ODEs) — and spends
budget only on correcting its magnitude. The supported PK shapes are
`"1cmt_iv"`, `"1cmt_oral"`, `"1cmt_infusion"`, `"2cmt_iv"`, and `"2cmt_oral"`.
Optional PD shapes are `"constant"`, `"linear"`, and `"exponential"`, with no
exposure dependence. Between-subject variability (`iiv`) and residual error
(`residual_cv`) are public assumptions and consume no budget.

Because only a handful of numbers are released (`d = 2` for a single PK
correction plus the count), the noise per released quantity stays small, which
is why this engine remains usable at 20 to 60 subjects. The tradeoff is that
everything not calibrated is *asserted*: curve shape, variability, and residual
error come from the public model, so the output is only as realistic as that
model. It cannot reveal a structural feature the model does not contain.

`synpmx_empirical()` instead reconstructs shape from a denser set of noised
summaries. It asserts less — trajectory shape is measured rather than assumed —
but it releases far more numbers, so the same epsilon is split many ways.
Utility therefore collapses below a few hundred subjects.

### The built-in models are illustrative, and deliberately so

It is worth being blunt about the ceiling here, because the phrase "structural
model" invites an expectation this package does not meet. The model catalogue is
small and fixed:

| | Supported |
|---|---|
| PK | `1cmt_iv`, `1cmt_oral`, `1cmt_infusion`, `2cmt_iv`, `2cmt_oral` — closed form, no ODEs |
| PD | `constant`, `linear`, `exponential`, with no exposure dependence |
| Variability | Lognormal between-subject variability on the typical parameters |
| Residual error | Proportional only |

And that is the whole of it. There are **no covariate–parameter relationships**
(no allometric exponent on clearance, no sex or biomarker effect), no
inter-occasion variability, no additive or combined residual error, no
absorption lag or transit compartments, no enzyme induction or time-varying
parameters, no exposure-driven PD, and no user-supplied ODE model. Declared
covariates appear in the output as columns, but nothing links them to the
concentrations beside them.

This is a scope decision rather than a gap waiting to be filled. The space of
pharmacometric models is effectively unbounded, and every study wants something
idiosyncratic; a package that chased that would slowly become a worse `rxode2`,
which already does the job properly. What the built-in models are *for* is
narrow and useful: giving the DP engines a public backbone whose magnitude can
be corrected under a budget, and producing structurally correct event tables to
develop pipeline code against.

So if your question is **"does my code handle this dataset shape?"**, these
modes are the right tool. If it is **"what would this study look like under my
model?"**, write that model in `rxode2` and simulate it there —
that is the right tool, and an LLM is a capable assistant for producing the
model code.

## Where to go next

- `vignette("synpmx-demo")` — the practical workflow across five public
  datasets, with structural checks.
- `vignette("synpmx-privacy")` — what differential privacy guarantees, what it
  does not, the trust-boundary decision rule, and how epsilon trades against
  utility.
- [The AVATAR Algorithm](https://iamstein.github.io/synpmx/articles/avatar-algorithm.html)
  — the default generator step by step, with the worked example.
- [Model elicitation](https://iamstein.github.io/synpmx/articles/model-elicitation.html)
  and
  [data elicitation](https://iamstein.github.io/synpmx/articles/data-elicitation.html)
  — how to produce the public model and protocol that modes 2 to 4 need.
- [Feasibility by cohort size](https://iamstein.github.io/synpmx/articles/feasibility.html)
  — the measured evidence for what each private mode can deliver at your N.

## References

1. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).

2. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).
