---
title: "Privacy"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Privacy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synpmx)
```

## Scope

`synpmx` generates synthetic pharmacometric (PMX) datasets — dosing and
measurement event tables used for a population pharmacokinetic (PK) or
pharmacodynamic analysis. It offers three families of method that protect
patients in fundamentally different ways. This vignette explains what each one
actually is, what differential privacy does and does not promise, how to choose
between methods and if a private mode is chosen, how to choose an epsilon.

The companion vignettes are `vignette("synpmx-4-methods")` (the four generation
modes) and `vignette("synpmx-demo")` (the practical workflow).

## The three method families

They differ on one axis above all: **what of the real data reaches the output.**
AVATAR sends real trajectories, blended. The DP engines send noised aggregates.
Public-model simulation sends nothing, because it never opens the file.

### AVATAR blending: synthetic data built from real subjects

The **default** method in this package is `synpmx_avatar()`, AVATAR-style
blending [1, 2]. "AVATAR" is a method name rather than an initialism: it comes
from the patient-centric *avatarization* literature, in which each synthetic
record ("avatar") is constructed from the local neighborhood of real records
rather than from a fitted parametric model. The original method is due to
Guillaudeux and colleagues [2]; Destere and colleagues benchmark a modified
AVATAR for population PK [1]. This package implements an AVATAR-*inspired*
adaptation for longitudinal event tables, not published AVATAR software.

Mechanically, for each synthetic subject it:

1. samples a real source subject as an **anchor** and keeps that subject's event
   skeleton — dosing rows, observation times, missing-value pattern — so the
   generated subject is structurally a real trial subject;
2. finds the **k nearest compatible neighbors** of the anchor in a standardized,
   principal-component (PCA) profile space (the default is `k = 5`);
3. fills covariates and endpoint trajectories with a randomized,
   distance-weighted **blend** of those donors, with no single donor allowed
   more than 50% of the weight; and
4. adds subject-level and within-trajectory random noise.

The output looks like trial data because it is assembled from trial data. That
is the source of both its utility and its risk: it works at any cohort size,
preserves real covariate correlations and real trajectory shapes for free, needs
no elicited model or priors — and makes **no formal privacy guarantee**. Its
safety rests on governance: the data stays inside a trusted, access-controlled
environment. This is the same footing as Novartis's `synadam`, an ADaM (Analysis
Data Model) synthetic-data package that resamples each column marginally from
the real data with no privacy accounting at all.

### Differential privacy: a bound on what any one subject can change

The **second** family is the differentially private (DP) engines,
`synpmx_calibrated()` and `synpmx_empirical()`. These never copy or blend a
subject's data into the output. Instead they:

1. compute a small number of **aggregate** statistics from the source data,
   with every per-subject contribution clipped into a publicly declared range;
2. add **calibrated random noise** to each aggregate, with the noise scale
   derived from how much one subject could have moved that number; and
3. generate the synthetic dataset from the noised aggregates and a public structural
   model.

Only the noised aggregates ever touch the generated data, so the release carries
a mathematical guarantee.

### Public-model simulation: synthetic data that never reads the real data

The **third** family is `synpmx_prior()`, and it does not answer the privacy
question so much as step around it. Give it a public structural model and a
trial design and it simulates subjects from those alone. It never opens the
source data — its arguments do not even include a dataset.

That buys a property neither other family can have. AVATAR has no formal
guarantee; the DP engines have a proven one. Public-model simulation needs
neither, because nothing about any real patient is ever read. There is no
disclosure to bound, no budget to spend, and epsilon simply does not apply.

**But that property belongs to the model, not to the function — and this is the
one thing to get right.** Every parameter must be fixed *before* anyone looks at
the source data: allometric scaling from preclinical work, a published model for
the compound class, the reasoning that set the starting dose. If instead you fit
a model to the real study and feed those estimates to `synpmx_prior()`, the real
data reaches the output through the parameters, and the "nothing was read" claim
is simply false. The generated rows would carry real information about real
patients while looking, to every downstream reader, like pure simulation.

`synpmx_prior()` cannot detect this. It sees a model object and has no way to
know where the numbers in it came from. `pmx_structural_model()` and
`pmx_trial_design()` each take a `source` argument to record that provenance in
writing, but it is a statement by the author, not a check by the package. The
discipline is yours to keep.

This is the same failure mode as a data-derived clipping range in the DP engines
— an input that quietly smuggles the data past the accounting — and it deserves
the same suspicion. If a number in your model was chosen because it made the
synthetic data look more like the real data, it came from the data.

The cost is the exact mirror of the benefit. Everything in the output is
*asserted* rather than learned, so the fidelity of the result is precisely the
fidelity of the model, design, and priors you supplied. It cannot discover a
covariate relationship you did not put in, and it will reproduce a wrong
assumption faithfully and without complaint. Its natural uses are exercising
pipeline code before any data exists, and cases where the real data cannot be
touched at all.

Producing those public inputs without reading data is its own skill; see the
[model elicitation](https://iamstein.github.io/synpmx/articles/model-elicitation.html)
and
[data elicitation](https://iamstein.github.io/synpmx/articles/data-elicitation.html)
articles.

## What differential privacy actually is

Differential privacy is a property of the **release procedure**, not of the
released dataset. You cannot inspect a table and check whether it is
differentially private; you can only check the mechanism that produced it.

A randomized mechanism `M` is `(epsilon, delta)`-differentially private if, for
every pair of **neighboring** datasets `D` and `D'` — identical except that one
complete subject is added or removed — and every set of possible outputs `S`:

$$\Pr[M(D) \in S] \;\le\; e^{\varepsilon} \, \Pr[M(D') \in S] \;+\; \delta$$

Read it as a promise made to one patient: *whatever the analyst concludes from
this release, they would have concluded almost the same thing had you never
enrolled.* The bound holds against an adversary with unlimited computing power
and arbitrary side information, including one who already knows every other
subject in the study.

Making that true requires two ingredients, both visible in the package:

- **Sensitivity** — the most that adding or removing one subject can move the
  released quantity. It exists only because per-subject values are clipped into
  a range declared *without looking at the data* so that an extreme value does not
  have a large impact on the synthetic data. A data-derived range is itself
  a leak, and the accounting will not catch it.
- **A noise scale tied to that sensitivity.** The engines use the Laplace
  mechanism: noise of scale `sensitivity / epsilon`. Smaller epsilon means a
  stronger promise and more noise.

The [privacy background article](https://iamstein.github.io/synpmx/articles/privacy-background.html)
works the arithmetic through with examples.

## The three families are fundamentally different

It is tempting to rank the methods as "less private" and "more private" but the difference is more substantial.

| | AVATAR blending | Differentially private engines | Public-model simulation |
|---|---|---|---|
| Function | `synpmx_avatar()` | `synpmx_calibrated()`, `synpmx_empirical()` | `synpmx_prior()` |
| Reads the source data | Yes, whole trajectories | Yes, as clipped aggregates | **Never** — and neither may the model it is given |
| What produces the output | Real subject trajectories, blended and perturbed | Noised aggregate statistics plus a public structural model | A public structural model and trial design alone |
| Privacy claim | None | Proven `(epsilon, delta)` bound on one subject's influence | Nothing is disclosed |
| Holds against a determined adversary | Not established | Yes, by construction | Yes, there is nothing to attack |
| Rests on | Governance and access control | Mathematics, plus correct declared ranges | Discipline: that every parameter was fixed before anyone looked at the data |
| Utility at small N | Good  | Degrades sharply below a few hundred | Independent of N — it depends on your model, not your cohort |
| What you must supply | Nothing but the source data | A public structural model, priors, and clipping ranges — all worked out without looking at the data | A public structural model and a trial design |

A caveat about AVATAR is that a resampled subject **trajectory** is much closer to a
fingerprint for a patient than a randomly chosen covariate. 
A trajectory contains particular sampling times, missed visits, noise pattern, and
curve shape can be nearly unique. Blending several donors and adding noise
mitigates this, but not formally, and pushing the noise high enough to defeat a
nearest-neighbor linkage attack would destroy the same signal a DP mechanism
would have destroyed — without the accounting to prove it. AVATAR therefore
leans on the governance context more heavily than column-wise resampling like
`synadam` does. This package has undergone no attack-based privacy validation.

## What each family asks of you

The quickest way to feel the difference is to ask what you have to supply.
Ordered below, the amount **you must assert** falls and the amount **taken from
the data** rises.

**1. Public-model simulation.** You specify everything: the structural model,
every typical parameter value, the dose levels, and the full sampling schedule.
All of it has to come from somewhere that is not this study — allometric scaling
from preclinical work, a published model for the compound class, the reasoning
behind the starting dose. You never look at the data; the function does not even
take a dataset. The most work of the three, and in exchange the privacy question
never arises.

**2. Differential privacy.** You still assert the model, the design, and the
prior ranges, and then buy a small correction toward reality with epsilon. For
`synpmx_calibrated()` exactly **two numbers leave the data**: a bounded
multiplicative correction to the model's magnitude, and a noised subject count.
Everything else — curve shape, between-subject variability, residual error,
covariate relationships — stays asserted from the public model and costs
nothing. `synpmx_empirical()` sits further along the same axis: it buys many
more numbers and asserts less, so the same epsilon splits many more ways, which
is why it needs a far larger cohort to be worth it.

**3. AVATAR.** You name the columns — which one is the subject ID, which is
time, which is the measurement, which is the dose — and it works out the rest
from the data itself. No model, no design, no priors, no budget. Least work by a
wide margin, and for the same reason the least to stand on: there is no asserted
model to point at and no formal guarantee either.

The pattern is worth stating plainly, because it is not a coincidence: **the
work you did not do is the guarantee you do not get.** Each step down this list
trades an assertion you would have had to defend for a fact taken from real
patients, and the privacy claim weakens by exactly that much.

For all three run end to end on `theo_md`, with code and output, see
`vignette("synpmx-4-methods")`.

## The decision rule

Two questions, in order.

> **First: does the output need to reflect the real data at all?**

- **No** — you are exercising pipeline code, the study has not read out yet, or
  the data cannot be touched → use **public-model simulation**
  (`synpmx_prior()`). Nothing is read, so the privacy question never arises. You
  are trading away all fidelity to the real cohort in exchange for having no
  disclosure to argue about.
- **Yes** — the synthetic data has to resemble *these* patients → ask the second
  question.

> **Second: does the generated data reach anyone the source data could not?**

- **No** — the same organization, under the same access controls and
  confidentiality obligations → use **AVATAR** (`synpmx_avatar()`).
- **Yes** — shared with a partner or vendor, published, or moved to a system
  outside those obligations → use a **DP** engine. A formal guarantee is the
  only thing that survives a determined adversary.

Differential privacy is more complex precisely because it defends against someone
who wants to break it. Use it when the output will be handed to strangers; skip
it when the synthetic data does not reach anyone outside the trusted organization.

Note that the first question is about *utility*, not privacy. That is the point:
public-model simulation is the only mode whose privacy story is settled before
you start, so the only thing left to weigh is whether a model-derived cohort is
good enough for what you are doing.

## The differentially private engines

Both make a subject-level guarantee: neighboring datasets differing by one
complete subject produce nearly indistinguishable output, so no one person's
participation can be inferred from the release.

- **`synpmx_calibrated()`** (structural correction). Asserts curve shape from a
  public structural model and privately calibrates only the exposure magnitude
  by a small correction factor. Because it releases very few quantities, it
  remains viable at small cohorts. This is the recommended DP path for
  early-phase studies.
- **`synpmx_empirical()`** (dense grid). Releases a larger set of noised
  population summaries. Retained for large pooled corpora where its cost is
  affordable.

Both are built on Laplace releases, which spend no delta, so the realized
accounting reports `delta = 0` — a pure epsilon guarantee — even though the
requested `delta` is carried as slack in the contract. Both fail closed when the
validated OpenDP backend is unavailable, and neither ever substitutes ordinary
random noise for a calibrated mechanism:

```{r backend, eval=requireNamespace("opendp", quietly = TRUE)}
dp_backend_status()
```

### The unaudited status is enforced, not just documented

Both engines are complete and tested, but not under active development, carry
known open findings tracked in `design/REVIEW_BACKLOG.md`, and have not been
independently privacy-audited. Treat them as a principled demonstration of the
privacy/utility tradeoff, not as a production release mechanism — a real
regulated release needs specialist review.

That distinction used to live in documentation alone, which made the risky
path exactly as easy to call as `synpmx_avatar()`. Both functions now refuse
to run until `synpmx_enable_dp_engines()` has been called once in the current
session:

```r
synpmx_enable_dp_engines()
```

The acknowledgment does not persist across sessions, script runs, or CI jobs —
each one must call it again. `backend = "public"` calls make no DP claim
already and are exempt from the gate.

### Budget is spent once, per call

Both functions read the confidential data exactly once and return a synthetic
dataset with the resulting release attached to it. Generating from that release
again is post-processing: it reads nothing and costs nothing.

```r
syn  <- synpmx_calibrated(data, roles, model, design, priors, epsilon = 1)
syn2 <- synpmx_generate(syn, seed = 202)   # free
privacy_report(syn)                        # the realized accounting
```

The distinction matters because calling `synpmx_calibrated()` or
`synpmx_empirical()` a second time is a **second release**, whose budget must be
composed with the first — two calls at `epsilon = 1` spend 2 in total, not 1.
Use `synpmx_generate()` for further datasets, or ask for several at once with
`n_datasets`. A repeated fit against the same data warns with the running total,
but the accounting is a governance obligation, not something the package can
enforce for you.

## What the guarantee does and does not mean

- **Epsilon** is the one-person influence limit: smaller is stronger. It is not
  a re-identification probability.
- **Delta** is a small additive slack in the probability bound. It is not the
  fraction of unprotected patients.
- Differential privacy bounds the information attributable to one person's
  participation. It does **not** establish legal anonymity, authorize release,
  or secure the environment.
- The guarantee is only as good as the declared clipping ranges. Ranges chosen
  by looking at the data break it silently.


## Choosing an epsilon

Everything below applies to the **differentially private** modes only. If your
synthetic data stays within the source data's own access controls and
obligations, use `synpmx_avatar()` and ignore epsilon entirely.

Epsilon is the one-person influence limit: smaller means stronger privacy and
more noise, larger means weaker privacy and less noise. There is no universal
default; an approved value must come from governance and threat modeling, not
from whichever number makes a plot look best.

For the structural-correction engine (`synpmx_calibrated()`), the usable
accuracy is captured by one quantity:

$$f \;=\; \frac{d}{\varepsilon N}$$

where `d` is the number of released quantities, `N` the number of subjects, and
`f` the fraction of the prior's width that survives as noise. The
`pmx_preflight()` helper reports it before any budget is spent:

```{r preflight}
priors <- pmx_priors(pk = pmx_prior(c(1 / 4, 4), source = "example"))
pmx_preflight(priors, epsilon = 0.5, n_subjects = 60)
```

The decision rule is not "is the error small" but "does the release beat the
prior": when `f` is near or above 1 the release conveys nothing the prior did
not, and generating from public inputs alone is strictly better.

### The measured frontier

Utility degrades sharply in small cohorts, and this is a property of
differential privacy, not of the implementation. Measured total fold-error on
clearance, for the structural engine with a correction-factor prior:

| N | epsilon 0.25 | epsilon 1 |
|---:|---:|---:|
| 6 | 2.3-fold | 1.7-fold |
| 20 | 1.6-fold | 1.3-fold |
| 60 | 1.4-fold | 1.1-fold |
| 300 | 1.08-fold | 1.06-fold |

From about 60 subjects upward the error is limited by estimator bias rather than
by the privacy mechanism, so a smaller epsilon is the better buy: it strengthens
the guarantee at no real cost in accuracy.

Epsilon and delta are governance decisions, not defaults. For anything public
facing they should be set and justified by whoever owns the data, and recorded:
every fit carries a release ledger, and `privacy_report()` prints the realized
accounting.

## Where to read more

- [Privacy background](https://iamstein.github.io/synpmx/articles/privacy-background.html)
  — how the arithmetic works: `d`, `f`, epsilon, and the error law, with worked
  examples.
- [Mechanism-level privacy argument](https://iamstein.github.io/synpmx/articles/privacy-argument.html)
  — the formal argument, for a reviewer.
- [Feasibility by cohort size](https://iamstein.github.io/synpmx/articles/feasibility.html)
  — the complete measured frontier, and why small cohorts are hard for any
  formal method.
- `vignette("synpmx-4-methods")` — the four generation modes and why AVATAR is the
  default. Note that it splits public-model simulation and the two DP engines
  into three separate *modes*, where this vignette groups the DP engines into
  one *family*; four modes, three families, same functions.
- [The AVATAR Algorithm](https://iamstein.github.io/synpmx/articles/avatar-algorithm.html)
  — the default generator in detail.
- [Model elicitation](https://iamstein.github.io/synpmx/articles/model-elicitation.html)
  and
  [data elicitation](https://iamstein.github.io/synpmx/articles/data-elicitation.html)
  — how to produce the public model, priors, and design that public-model
  simulation and the DP engines both need.

## References

1. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).

2. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).
