---
title: "AVATAR mathematics: the generator step by step"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  out.width = "100%"
)
library(synpmx)
```

This article is the full specification of the default AVATAR generator: every
step from role declaration to the restored output schema, with the mathematics,
the edge cases, and a worked example.

It is the companion to `vignette("synpmx-method")`, which introduces all four
generation modes at a high level and says which one to reach for. Read that
first. Come here when you need to defend, debug, or review what
`synpmx_avatar()` actually did.

## Step 1: declare the meaning of the columns

The package never guesses critical PMX roles from column names. The user must
declare them with `pmx_roles()`:

```{r roles-example}
roles <- pmx_roles(
  id = "ID",
  time = "TIME",
  dv = "DV",
  amt = "AMT",
  evid = "EVID",
  cmt = "CMT",
  dvid = "DVID",
  mdv = "MDV",
  rate = "RATE",
  cens = "CENS",           # below-limit indicator, if the study has BLQ data
  limit = "LIMIT",         # the other end of an interval, when used
  covariates = c("WT", "AGE", "SEX"),
  keep = "ARM"             # carried verbatim; see the allowlist below
)
roles
```

`id`, `time`, `dv`, and `evid` are required; the rest are optional. `amt`,
`cmt`, `mdv`, `rate`, `cens`, and `limit` each name one column. `dvid` names one
column, or several that encode the same endpoint (see below). `covariates` and
`keep` each take any number of columns. One source column cannot be assigned to
two roles. The roles `subject_properties`, `assigned_dose`, and `exclude` belong
to the differentially private engines only; `synpmx_avatar()` rejects them and
points at the role that does the job here.

### The role declaration is the allowlist

`synpmx_avatar()` keeps only columns a role names, and **drops every other
column**, reporting which ones. This is deliberate. AVATAR builds each synthetic
subject from one real *anchor* subject, and any column it retains that it does
not actively synthesize is copied from that anchor verbatim. A secondary
identifier — `USUBJID`, a site number, a randomization date — left undeclared
would otherwise ride a real subject's real value straight into the synthetic
data. Dropping by default means a column you forget fails safe, toward removal,
rather than leaking.

That leaves four ways a column is treated, and the difference matters:

| Declared as | What AVATAR does to it | Use for |
|---|---|---|
| `dv` (with `cens`/`limit`) | Blended across donors into a new trajectory | the measurement |
| `covariates` | Blended/resampled across donors into a new value | baselines you want *synthesized* — weight, age |
| `keep` | Copied verbatim from the one anchor subject | assigned values you want kept faithful to that subject's dosing — arm, dose group, a redundant endpoint label |
| *(undeclared)* | Dropped | anything you do not need |

`covariates` and `keep` are opposites, and choosing wrongly is the common
mistake. A **covariate is blended**: the synthetic subject gets a genuinely new
value interpolated from several neighbours, decoupled from any one person. A
**kept column is copied**: the synthetic subject gets one real subject's real
value, unchanged. For a treatment arm, a dose group, or a randomization sequence
— anything that must stay consistent with the doses AVATAR copied from the same
anchor — `keep` is correct, precisely because it never leaves that anchor's
side. Because a kept value is a real subject's real value, `keep` is for the
trusted environment only.

Redundant endpoint labels have their own handling. A dataset that carries both a
numeric `YTYPE` and a character `NAME` for the same endpoint declares **both**
as `dvid`: `dvid = c("YTYPE", "NAME")`. The first is the grouping key; validation
checks that the rest are a consistent 1:1 mapping with it — catching a source
where the two labels disagree, which `keep` would carry through silently — and
AVATAR carries all of them into the output, aligned.

Immediately before generation, `synpmx_avatar()` calls `validate_pmx(...,
strict = TRUE)`. Among other checks, validation requires:

- nonmissing IDs and EVID values;
- finite numeric TIME values and nondecreasing TIME within subject;
- numeric DV and finite nonmissing observations;
- an endpoint value on every eligible observation when DVID is declared;
- multiple `dvid` columns to be a consistent 1:1 mapping;
- every declared baseline covariate to be constant within subject; and
- when `cens` is declared, each CENS flag to agree with its DV — a
  left-censored value at the limit cannot exceed an ordinary measurement of the
  same endpoint. A pooled dataset with two assays at different limits will trip
  this legitimately; synthesize one study at a time, or drop the `cens` role.

Every error names the role and the column it maps to, and the count and an
example of the offending rows, so a failure says exactly what to fix.

Rows are eligible observations when EVID is zero and, if MDV is declared, MDV
is also zero. A nonmissing DV is additionally required when a value is used to
build a profile or donor trajectory. In symbols, with bracket notation for a
logical condition,

$$
O_{ij} = [\mathrm{EVID}_{ij}=0]
         [\mathrm{MDV}_{ij}=0\;\text{if MDV is declared}]
         [Y_{ij}\;\text{is present}].
$$

Rows with EVID equal to zero but missing DV, or with nonzero MDV, remain in the
sampled template but do not contribute a measurement value.

## Step 2: split structure from values

Let subject $i$ have source rows $D_i$. For every requested synthetic subject, the
generator samples an **anchor** $a$ with replacement and begins with

$$
S^{(0)} = D_a.
$$

Thus $S^{(0)}$ contains the anchor's full row count and row-wise combination of
TIME, EVID, AMT, RATE, CMT, DVID, and any other source columns. This is the
event skeleton. By copying these rows together, the method avoids impossible
records that can result from generating event fields independently.

Only a few parts of this skeleton are subsequently changed:

- TIME may be jittered if `time_jitter > 0`;
- declared baseline covariates are replaced for all rows of the synthetic subject;
- eligible, originally present observation DVs are synthesized;
- standard MDV may be re-derived;
- ID is replaced; and
- rows are sorted by TIME, retaining original order within ties.

Event-row DV values and undeclared columns otherwise remain anchor-template
values. Consequently, users should not place a semantically important column
in the input and assume the package understands it merely because it is
present.

## Step 3: align time without pretending every subject was sampled identically

### First-event-relative time

Neighbor profiles and donor trajectories are constructed on an internal
aligned time. For subject $i$, the preferred origin is the first nonzero-EVID
row having a positive AMT:

$$
t_{0i} = \min\{t_{ij}: \mathrm{EVID}_{ij}\ne0,\ \mathrm{AMT}_{ij}>0\}.
$$

If no positive-AMT event exists, the first nonzero-EVID row is used. If the
subject has no event row, TIME is left unchanged. The internal aligned time is

$$
t^*_{ij}=t_{ij}-t_{0i}.
$$

This removes a subject-specific calendar or study-time offset. It does **not**
compute time after each dose.

### TIME is not TAD

A conventional time-after-dose (TAD) variable resets after every dose. AVATAR
does not use a declared `tad` role for alignment and never recomputes a TAD
column. With multiple doses, $t^*$ continues to increase from the first
qualifying dose; there is no later reset. A declared `tad` column is retained
and copied from the anchor template unchanged; an undeclared one is dropped, as
any undeclared column is.

This has two practical consequences:

1. Use the actual analysis-time column as `time`. Do not expect a TAD column to
   drive neighbor construction or interpolation.
2. If `time_jitter > 0`, a declared TAD column is copied, not recalculated, and
   can become inconsistent with the jittered TIME. Recompute TAD downstream if
   the workflow requires it.

### Different observation times across subjects

For endpoint $e$, the package pools finite aligned observation times and makes
a common grid $G_e$. If there are at most 15 unique times, all are retained. If
there are more, 15 type-8 empirical quantiles are used and duplicate quantiles
are removed.

Each subject's transformed trajectory is linearly interpolated onto $G_e$.
There is no extrapolation during profile construction, so grid locations
outside a subject's observed window become missing profile features. Tied
times are averaged by `stats::approx(..., ties = mean)`.

During final synthesis, donors are first interpolated directly to the anchor's
aligned observation times. Suppose target time $t$ lies outside donor $r$'s
window. Rather than extrapolating indefinitely, the method maps the target's
relative position in the whole anchor observation window to the donor window:

$$
q(t)=\frac{t-a_{\mathrm{target}}}
           {b_{\mathrm{target}}-a_{\mathrm{target}}},
\qquad
u_r(t)=a_r+q(t)(b_r-a_r),
$$

where $[a_{\mathrm{target}},b_{\mathrm{target}}]$ and $[a_r,b_r]$ are the
target and donor ranges. The donor trajectory is evaluated at $u_r(t)$. This
normalized-window fallback preserves progress through a trajectory, but it is
not a mechanistic warping model. A donor with only one unique time supplies its
mean at every requested target time.

### Optional coherent time jitter

The default `time_jitter = 0` copies template times exactly. Otherwise, one
normal perturbation is drawn for each unique template time $u_l$,

$$
u'_l=u_l+\delta_l, \qquad
\delta_l\sim\mathcal{N}(0,\sigma_t^2),
$$

and constrained between the midpoints of adjacent nominal times. Nonnegative
time is enforced at the first time. All rows tied at $u_l$ receive the same
$u'_l$, so a dose and observation tied at a nominal time remain tied and time
ordering cannot cross.

## Step 4: choose an endpoint transformation

Transformations are selected separately for each DVID endpoint, or once for
the implicit endpoint `"DV"` when DVID is absent. After discarding nonfinite
values, an endpoint is considered positive-like when it has at least one
positive value, its median is positive, and at most 1% of values are negative.

For a positive-like endpoint, define

$$
c_e=\max\left\{\frac{1}{2}\min(Y_e[Y_e>0]),\sqrt{\epsilon_{\mathrm{mach}}}\right\}
$$

and use the offset-log pair

$$
g_e(y)=\log\{\max(y,0)+c_e\},
\qquad
g_e^{-1}(z)=\max\{\exp(z)-c_e,0\}.
$$

Other endpoints use $g_e(y)=y$. The truncations mean a positive-like endpoint
cannot generate a negative DV. Transformation choices and offsets are stored
in `attr(synthetic, "pmx_settings")$endpoint_transforms`.

## Step 5: build one numeric profile per source subject

A profile concatenates baseline and longitudinal information:

$$
\mathbf{x}_i=
\left[
  \mathbf{c}_i,\
  \mathbf{y}_{i1}(G_1),\ldots,\
  \mathbf{y}_{iE}(G_E)
\right].
$$

Here $\mathbf{c}_i$ contains the first nonmissing value of each declared
baseline covariate. Numeric covariates contribute one feature. Factors and
character covariates contribute one indicator for every factor level or
observed character value. The vector $\mathbf{y}_{ie}(G_e)$ is subject $i$'s
transformed endpoint-$e$ trajectory interpolated on its common grid.

This profile is used only for distance calculations. It does not replace the
source data and is never returned as a synthetic patient.

### Missing features and scaling

An entirely missing feature is removed. Otherwise, a missing profile entry is
replaced with its across-subject feature median $m_l$:

$$
x^{\mathrm{imp}}_{il}=
\begin{cases}
x_{il}, & x_{il}\text{ is finite},\\
m_l, & \text{otherwise}.
\end{cases}
$$

This imputation exists only inside neighbor finding. It does not fill source or
output DV rows. Features whose standard deviation is not greater than
$\sqrt{\epsilon_{\mathrm{mach}}}$ are removed. The remaining columns are
standardized as

$$
z_{il}=\frac{x^{\mathrm{imp}}_{il}-\bar{x}^{\mathrm{imp}}_l}{s_l}.
$$

### Rank-safe PCA

When there are at least two subjects and two usable features, PCA is applied to
the standardized matrix with maximum rank

$$
r_{\max}=\min(n-1,p).
$$

Let $\lambda_h$ be the variance of principal component $h$. The retained
dimension $H$ is the smallest value satisfying

$$
\frac{\sum_{h=1}^{H}\lambda_h}
     {\sum_{h=1}^{r_{\max}}\lambda_h}
\ge v,
$$

where $v$ is `pca_variance` (default 0.90). The subject coordinate is
$\boldsymbol{\xi}_i=(\mathrm{PC}_{i1},\ldots,\mathrm{PC}_{iH})$. If PCA is not
available but standardized features exist, those features are used directly.
If none exist, all subjects receive the same one-dimensional zero coordinate.

The term **rank-safe** means the attempted PCA rank cannot exceed either the
feature rank or $n-1$. It does not imply that sparse or nearly duplicated
profiles contain useful separation.

## Below-the-limit-of-quantification (BLOQ) data and a Monolix-style CENS column

Declare `cens` and, if the source uses one, `limit` in `pmx_roles()`, and AVATAR
reconstructs censoring rather than copying it. The mechanism has three parts.

**The boundary is read from the source.** AVATAR makes no formal privacy claim,
so it may recover the assay limit from the data itself: under the Monolix
convention a censored row reports the boundary in `DV`, with `LIMIT` carrying
the other end of an interval when there is one. Left, right, and interval
censoring are each recognised. A differentially private engine may *not* do
this — there the limit must be a declared public input, because a boundary
inferred from confidential records is itself a release. That asymmetry is the
reason the two engines derive boundaries differently.

If a study used more than one assay limit, the most conservative boundary is
taken rather than inventing a per-row rule the source cannot support.

**Censored donor values are imputed before blending.** Every censored source row
reports the same substituted number, so blending those values directly would put
a floor under the synthetic data that the real study does not have. Each
censored row is instead replaced by a draw inside its censoring region — uniform
below the limit for a left-censored row, uniform within the interval for an
interval-censored one — before the trajectory is transformed and blended. A
uniform draw rather than a fixed LLOQ/2 avoids replacing one artificial spike
with another.

**`DV`, `CENS`, and `LIMIT` are reconstructed together.** The back-transformed
blend is treated as a *latent* value: what the subject would have measured with
no assay limit. The boundary is then applied to that latent, and all three
columns are written from the result, so they cannot disagree. A generated value
below the limit produces `CENS = 1` with `DV` exactly at the limit, reproducing
the spike of identical values that characterises real censored data.

Two honest limitations. The censored *fraction* is reproduced approximately, not
exactly: blending shrinks variance toward the middle while the imputation draws
uniformly across the censoring region, so the synthetic rate typically runs
somewhat above the source's. And if a `cens` role is declared but no boundary
can be read from it — rows flagged with `DV` missing, for instance — AVATAR
warns and carries the flag through untouched rather than guessing.

## Step 6: restrict neighbors to compatible event patterns

Similarity in PCA space is not sufficient. A subject receiving a single oral
dose should not donate a trajectory to an incompatible infusion or repeat-dose
template. Each source subject therefore receives an event signature containing:

- the ordered EVID values on event rows;
- optional CMT and DVID values on those rows;
- the sign and rounded numeric magnitude of optional AMT and RATE;
- the number of positive-dose starts, or all events if no positive AMT exists;
- successive start-time gaps rounded to two significant digits; and
- the set of endpoints on eligible observation rows.

Notice what is not in the signature: the observation-time schedule. Event
values still come from the anchor template, while unequal observation times are
handled by interpolation. Including dose magnitude prevents profiles from
being blended across incompatible exposure scales.

For anchor $a$, only non-anchor subjects with the same signature are candidate
donors. Euclidean distance in retained profile coordinates is

$$
d_{ar}=\left\|\boldsymbol{\xi}_a-
                   \boldsymbol{\xi}_r\right\|_2
=\sqrt{\sum_{h=1}^{H}(\xi_{ah}-\xi_{rh})^2}.
$$

The closest `k` candidates are kept. If fewer exist, `k` is reduced and a
warning is recorded. If none exist, the anchor itself becomes the sole donor;
only subsequent randomized perturbation can then move its measurements.

## Step 7: randomize and cap donor weights

Suppose $K$ donors have been selected. Let $E_r\sim\mathrm{Exponential}(1)$
and let $R_r$ be a random permutation of $1,\ldots,K$. Importantly, $R_r$ is a
randomized rank, not the distance order. The raw donor weight is

$$
q_r=\frac{E_r}{\max(d_{ar},\varepsilon)}2^{-R_r},
$$

where

$$
\varepsilon=
\begin{cases}
\max\{10^{-8},10^{-6}\operatorname{median}(d_{ar}:d_{ar}>0)\},
  & \text{if a positive distance exists},\\
10^{-8}, & \text{otherwise}.
\end{cases}
$$

The first normalization is $w_r=q_r/\sum_s q_s$. When more than one donor is
available, if one weight exceeds 0.80,
the dominant weight is set to 0.80 and all other weights are rescaled
proportionally to sum to 0.20. A final normalization protects against floating
point drift. With exactly one donor, its weight is necessarily 1; the cap
cannot create a second donor. If the raw weights are nonfinite or have a
nonpositive total, the implementation replaces them with equal raw values
before normalization.

The same selected subjects and weights are used for every declared covariate
and endpoint. When donor $r$ lacks a usable value at target location $j$, the
available weights are renormalized locally:

$$
w_{rj}^{*}=\frac{w_r I_{rj}}
                 {\sum_s w_s I_{sj}},
$$

where $I_{rj}=1$ when the value is finite and zero otherwise.

## Step 8: synthesize baseline covariates

For a numeric covariate, let $c_r$ be donor $r$'s first nonmissing value. When
all available values are positive and

$$
\frac{\max_r c_r}{\operatorname{median}_r(c_r)}>3,
$$

the blend is formed on the log scale; otherwise it is formed on the original
scale. With working-scale values $h(c_r)$, the donor center is

$$
\mu_c=\sum_r w_r^* h(c_r).
$$

Let $s_c$ be the ordinary sample standard deviation of the available
working-scale donor values. If it is unavailable or zero, the fallback is
$\max(0.05|\mu_c|,0.01)$. The generated working-scale covariate is

$$
c^*=\mu_c+\eta_c,
\qquad
\eta_c\sim\mathcal{N}(0,(\sigma_{\mathrm{subj}}s_c)^2),
$$

where `subject_noise_sd` is $\sigma_{\mathrm{subj}}$ (default 0.15). A log-scale
result is exponentiated, and any all-positive covariate is floored above zero.
The final value is repeated on every row of that synthetic subject.

For factor, character, and logical covariates, one available donor category is
sampled using the locally normalized weights. Categories are not averaged.

### A covariate and an endpoint are blended independently

Covariates (this step) and endpoint trajectories (the next) are generated in
separate passes, each with its own donor weighting and its own perturbation.
That is fine when they are unrelated, but some datasets carry a **baseline
covariate that is the same quantity as a longitudinal endpoint** — a `B0`
baseline B-cell count beside a B-cell kinetic endpoint, a baseline biomarker
beside its own time course. AVATAR does not know they are linked, so a synthetic
subject's baseline covariate need not equal the baseline of its own generated
trajectory. The two are individually plausible but not mutually consistent.

This is a known limitation (`REV-022`). It is usually harmless for
workflow-development use, where each column need only be realistic on its own.
If your analysis relies on a covariate agreeing with an endpoint's baseline,
reconcile them after generation, or do not declare the redundant covariate and
read the baseline from the trajectory instead.

## Step 9: synthesize each endpoint trajectory

Only positions that were eligible, nonmissing observations in the anchor
template receive synthesized values. Thus the anchor supplies the number and
placement of observations, including its missing-DV pattern.

For endpoint $e$, donor trajectories are transformed with $g_e$ and
interpolated to the anchor observation times. Let $z_{rj}$ be donor $r$'s
transformed value at target position $j$. The deterministic blend is

$$
\bar z_j=\sum_r w_{rj}^{*}z_{rj}.
$$

If every donor is unavailable at a target, the transformed dataset median for
that endpoint is used and a warning is stored.

The endpoint noise scale $s_e$ is 1 on the offset-log scale. On the identity
scale it is the source endpoint standard deviation, with fallback
$\max(0.1|\operatorname{median}(Y_e)|,0.01)$ when necessary. A single
subject-level shift is shared across all positions of the endpoint:

$$
b_e\sim\mathcal{N}(0,(\sigma_{\mathrm{subj}}s_e)^2).
$$

Within-endpoint residual perturbations follow a stationary-scale first-order
autoregressive, AR(1), process in **observation order**:

$$
\begin{aligned}
\epsilon_{e1} &\sim
  \mathcal{N}(0,(\sigma_{\mathrm{res}}s_e)^2),\\
\epsilon_{ej} &= \phi\epsilon_{e,j-1}+\nu_{ej},\\
\nu_{ej} &\sim
  \mathcal{N}\left(0,
    (\sigma_{\mathrm{res}}s_e)^2(1-\phi^2)\right).
\end{aligned}
$$

Defaults are `residual_noise_sd = 0.05` and `residual_phi = 0.6`. The generated
DV is

$$
Y^*_{ej}=g_e^{-1}(\bar z_j+b_e+\epsilon_{ej}).
$$

Endpoints are processed separately, so PK and PD values cannot be blended
with one another. They nevertheless use the same donor subjects and initial
subject weights, preserving a limited form of cross-endpoint subject
coherence. This is a synthesis device, not a fitted multivariate residual
model.

## A transparent worked example

The small dataset below has six subjects, one dose each, one positive endpoint,
two baseline covariates, and deliberately different observation times. It is
large enough to have non-anchor neighbors but small enough to inspect.

```{r worked-source}
make_subject <- function(id, observation_time, wt, sex, peak) {
  time <- c(0, observation_time)
  evid <- c(1L, rep(0L, length(observation_time)))
  data.frame(
    ID = as.integer(id),
    TIME = time,
    DV = c(0, peak * exp(-0.55 * observation_time) + 0.03 * id),
    AMT = c(100, rep(0, length(observation_time))),
    EVID = evid,
    CMT = c(1L, rep(2L, length(observation_time))),
    MDV = ifelse(evid == 0L, 0L, 1L),
    WT = rep(c(58, 64, 70, 76, 82, 88)[id], length(time)),
    SEX = factor(
      rep(sex, length(time)),
      levels = c("female", "male")
    )
  )
}

worked_source <- do.call(rbind, list(
  make_subject(1, c(0.5, 1.0, 2.0, 4.0), 58, "female", 9.0),
  make_subject(2, c(0.25, 1.5, 3.0, 6.0), 64, "male", 10.0),
  make_subject(3, c(0.75, 1.25, 2.5, 5.0), 70, "female", 11.0),
  make_subject(4, c(0.4, 1.8, 3.5, 5.5), 76, "male", 12.0),
  make_subject(5, c(0.6, 1.4, 2.8, 4.5), 82, "female", 13.0),
  make_subject(6, c(0.3, 1.1, 3.2, 6.5), 88, "male", 14.0)
))
rownames(worked_source) <- NULL

worked_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", mdv = "MDV",
  covariates = c("WT", "SEX")
)

knitr::kable(
  worked_source[worked_source$ID %in% 1:2, ],
  digits = 3,
  caption = "The first two source subjects"
)
```

The next chunk uses internal helpers solely to expose a reproducible teaching
trace. These helpers are not public API. Ordinary analysis code should call
`synpmx_avatar()` rather than depend on them.

```{r worked-trace, echo=FALSE}
worked_seed <- 2026L
worked_k <- 3L
worked_profiles <- synpmx:::.build_profiles(
  worked_source, worked_roles, pca_variance = 0.90
)

set.seed(worked_seed)
anchor_index <- sample.int(
  length(worked_profiles$subjects), 1L, replace = TRUE
)
trace_warnings <- synpmx:::.warning_collector()
worked_donors <- synpmx:::.select_donors(
  anchor_index, worked_profiles, worked_k, trace_warnings
)

trace_summary <- data.frame(
  role = c("anchor", rep("donor", length(worked_donors$indices))),
  source_ID = c(
    worked_profiles$subjects[anchor_index],
    worked_profiles$subjects[worked_donors$indices]
  ),
  profile_distance = c(NA_real_, worked_donors$distances),
  weight = c(NA_real_, worked_donors$weights),
  stringsAsFactors = FALSE
)
```

With `seed = 2026`, the anchor and its selected compatible donors are:

```{r worked-neighbors, echo=FALSE}
knitr::kable(trace_summary, digits = 4)
```

The anchor contributes the event skeleton. Donors contribute transformed DVs
after interpolation to the anchor times. The following table shows the exact
pre-noise blend used by the implementation.

```{r worked-blend, echo=FALSE}
anchor_skeleton <- worked_source[
  worked_profiles$subject_rows[[anchor_index]], , drop = FALSE
]
target_rows <- which(
  synpmx:::.observation_rows(
    anchor_skeleton, worked_roles, require_present = TRUE
  )
)
target_time <- synpmx:::.aligned_time(
  anchor_skeleton, worked_roles
)[target_rows]
worked_transform <- worked_profiles$transforms[["DV"]]

worked_donor_matrix <- vapply(worked_donors$indices, function(donor) {
  trajectory <- synpmx:::.donor_trajectory(
    worked_source,
    worked_roles,
    worked_profiles$subject_rows[[donor]],
    "DV",
    worked_transform
  )
  synpmx:::.interpolate_trajectory(trajectory, target_time)
}, numeric(length(target_time)))
if (is.null(dim(worked_donor_matrix))) {
  worked_donor_matrix <- matrix(worked_donor_matrix, ncol = 1L)
}
colnames(worked_donor_matrix) <- paste0(
  "donor_", worked_profiles$subjects[worked_donors$indices], "_z"
)
worked_blend <- apply(
  worked_donor_matrix, 1L,
  synpmx:::.weighted_available,
  weights = worked_donors$weights
)

worked_synthetic <- synpmx_avatar(
  worked_source,
  worked_roles,
  n_subjects = 1,
  seed = worked_seed,
  k = worked_k
)
worked_synthetic_rows <- synpmx:::.observation_rows(
  worked_synthetic, worked_roles, require_present = TRUE
)

blend_table <- data.frame(
  anchor_TIME = target_time,
  worked_donor_matrix,
  blended_z = worked_blend,
  deterministic_DV = synpmx:::.inverse_dv(
    worked_blend, worked_transform
  ),
  final_synthetic_DV = worked_synthetic$DV[worked_synthetic_rows],
  check.names = FALSE
)
knitr::kable(
  blend_table,
  digits = 3,
  caption = paste(
    "Interpolation and blending for the anchor endpoint; z is the",
    "endpoint working scale"
  )
)
```

`deterministic_DV` is the back-transformed donor blend. `final_synthetic_DV` also
contains the one subject shift and AR(1) residual sequence. The event rows,
new ID, and subject-constant synthesized covariates can be inspected directly:

```{r worked-result}
knitr::kable(worked_synthetic, digits = 3)
```

The result records enough settings to audit the public generator call:

```{r worked-settings, echo=FALSE}
worked_settings <- attr(worked_synthetic, "pmx_settings")
settings_table <- data.frame(
  setting = c(
    "seed", "n_subjects", "k", "pca_variance",
    "subject_noise_sd", "residual_noise_sd", "residual_phi",
    "time_jitter", "alignment", "compatible_event_groups",
    "max_donor_weight", "warnings"
  ),
  value = c(
    worked_settings$seed,
    worked_settings$n_subjects,
    worked_settings$k,
    worked_settings$pca_variance,
    worked_settings$subject_noise_sd,
    worked_settings$residual_noise_sd,
    worked_settings$residual_phi,
    worked_settings$time_jitter,
    worked_settings$alignment,
    worked_settings$compatible_event_groups,
    worked_settings$max_donor_weight,
    if (length(worked_settings$warnings)) {
      paste(worked_settings$warnings, collapse = "; ")
    } else {
      "<none>"
    }
  ),
  stringsAsFactors = FALSE
)
transform_table <- do.call(rbind, lapply(
  names(worked_settings$endpoint_transforms),
  function(endpoint) {
    transform <- worked_settings$endpoint_transforms[[endpoint]]
    data.frame(
      endpoint = endpoint,
      method = transform$method,
      offset = transform$offset,
      positive = transform$positive,
      stringsAsFactors = FALSE
    )
  }
))
knitr::kable(settings_table, caption = "Recorded generator settings")
knitr::kable(
  transform_table,
  digits = 5,
  caption = "Recorded endpoint transformation"
)
```

## Step 10: restore PMX schema and deterministic conventions

After all synthetic subjects are row-bound, the original column order is restored.
Each result column is converted toward its source class:

- factors recover source levels and ordering; new factor IDs add new levels;
- integer columns are rounded and converted to integer;
- double, logical, character, Date, and POSIXct columns are restored; and
- a non-`data.frame` source class, such as a tibble class vector, is reapplied.

New numeric IDs start above the source maximum. New character and factor IDs
use labels such as `syn_001`, with a prefix added repeatedly if needed to
avoid collision.

If MDV is declared and the entire source obeys the standard relationship

$$
[\mathrm{MDV}=0]=[\mathrm{EVID}=0\ \text{and DV is present}],
$$

MDV is re-derived from that rule in the synthetic data and restored to its original
class. For any nonstandard source-wide MDV convention, anchor values are left
as copied. Missing observation positions are not invented or removed.

Finally, `validate_pmx(..., strict = TRUE)` is run on the assembled result. A
generation call cannot silently return an output that fails the package's own
structural requirements.

## Reproducibility and diagnostics

All random operations occur inside a local seed scope. `seed` must be an
integer between zero and `.Machine$integer.max`. The caller's RNG kind and
`.Random.seed` are restored on exit, whether or not `.Random.seed` existed
before the call. Consequently, identical inputs and arguments produce an
identical complete result without consuming the surrounding analysis RNG
stream.

`attr(synthetic, "pmx_settings")` records arguments, explicit roles, endpoint
transformations, the alignment description, number of compatible event groups,
the 0.80 cap, and unique fallback warnings. It does not currently record the
anchor ID, donor IDs, distances, or realized weights for each synthetic subject.

Use the public checks for different questions:

- `validate_pmx()` asks whether one dataset is structurally usable under the
  declared roles.
- `compare_pmx()` reports source/synthetic row counts, subjects, event counts,
  endpoints, column classes, validations, and exploratory trajectories when
  `ggplot2` is installed.

Neither function establishes distributional equivalence, privacy, or model
fidelity.

## Edge cases and documented fallbacks

The implementation favors an explicit warning and structurally valid output
over a silent incompatible blend.

```{r edge-cases-table, echo=FALSE}
edge_cases <- data.frame(
  Situation = c(
    "No compatible non-anchor donor",
    "Fewer compatible donors than k",
    "All selected distances essentially zero",
    "One donor time only",
    "No donor value at a target",
    "No variable profile feature",
    "Identical donors with all noise disabled",
    "Nonstandard MDV convention"
  ),
  Behavior = c(
    "Use anchor as sole donor; randomized noise is the only trajectory change",
    "Reduce k and record a warning",
    "Use epsilon-stabilized randomized weights",
    "Repeat that donor's mean at every target time",
    "Use transformed endpoint median and record a warning",
    "Use a shared zero coordinate; distance cannot separate subjects",
    "A source-shaped trajectory may be mathematically unavoidable; warn",
    "Copy anchor MDV values rather than re-derive them"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(edge_cases, align = c("l", "l"))
```

Other boundaries are deliberate. `synpmx_avatar()` supports only
`event_method = "template"` and `dv_method = "avatar_blend"`. It does not fit
a structural model, reconstruct covariate-parameter relationships, simulate
occasions, enforce a dosing grammar beyond the sampled template, or provide a
censoring model. Fitting an explicit structural model is the job of
`synpmx_calibrated()`, described below.

## Relationship to AVATAR

The method is **AVATAR-inspired**, not an implementation of published AVATAR
software. "AVATAR" is a method name rather than an initialism, from the
patient-centric *avatarization* literature in which each synthetic record is
built from the local neighborhood of real records. The shared family resemblance
is the use of standardized subject features, PCA, local neighbors,
inverse-distance influence, exponential randomization, and rank attenuation.
Destere and colleagues describe a modified AVATAR benchmark for longitudinal
population pharmacokinetic (PopPK) data that first widened the data by subject
and then used PCA, K-nearest neighbors, inverse Euclidean distance, exponential
stochastic weights, and rank attenuation before reverse transformation [1]. The original patient-centric AVATAR method was reported by
Guillaudeux and colleagues [2].

`synpmx` adapts those ideas to a PMX event table in several material ways:

- it preserves a sampled longitudinal event template rather than generating
  all fields from a wide vector;
- compatibility is restricted by event and schedule signatures;
- endpoints are transformed and interpolated separately;
- the same donors support covariates and endpoints;
- a dominant multi-donor weight is capped at 0.80;
- subject and AR(1) perturbations are added on the endpoint working scale; and
- there is no inverse-PCA reconstruction of a full synthetic patient vector.

Destere et al. also demonstrate why method labels are not validation: their
benchmark found materially different PK parameter and residual-error biases
across synthesis algorithms [1]. That paper is a 2026 medRxiv preprint and was
not peer reviewed at the version supplied with this package repository. Its
privacy analyses do not transfer to `synpmx`; this package has undergone
no corresponding attack-based privacy validation.

## Algorithm summary

For each call to `synpmx_avatar()`:

1. Validate source data, roles, and generator arguments.
2. Compute first-event-relative time and endpoint-specific transformations.
3. Assemble baseline covariates and interpolate transformed endpoint
   trajectories into one profile per source subject.
4. Median-impute profile features, remove unusable features, standardize, and
   retain enough PCA components to meet `pca_variance` when possible.
5. Construct event/schedule signatures.
6. Sample anchor subjects with replacement.
7. For each anchor, copy its whole event template.
8. Find up to `k` closest compatible non-anchor donors and randomize their
   inverse-distance weights, applying the multi-donor 0.80 cap.
9. Generate subject-constant covariates from those donors.
10. For each endpoint, interpolate donor trajectories to anchor times, blend,
    add a subject shift and AR(1) perturbations, and back-transform.
11. Re-derive standard MDV when applicable, assign a new ID, preserve tie
    order, restore the source schema, and validate the assembled result.
12. Attach `pmx_settings` and emit any collected fallback warnings.

## References

1. Destere A, Lombardi R, Labriffe M, et al. *Can synthetic data overcome the
   privacy and fidelity bottleneck in Pharmacometrics? A comparative benchmark
   using a daptomycin population pharmacokinetic model.* medRxiv preprint,
   posted June 2, 2026. doi:
   [10.64898/2026.05.30.26354512](https://doi.org/10.64898/2026.05.30.26354512).

2. Guillaudeux M, Rousseau O, Petot J, et al. Patient-centric synthetic data
   generation, no reason to risk re-identification in biomedical data analysis.
   *npj Digital Medicine.* 2023;6.
   doi: [10.1038/s41746-023-00771-5](https://doi.org/10.1038/s41746-023-00771-5).
