---
title: "Example: are PK/PD, covariate, and treatment relationships preserved?"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center",
                      out.width = "100%")
library(synpmx)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

This article builds a source dataset with four relationships
put in deliberately --- a covariate effect, a dose effect, a treatment effect,
and an exposure--response link --- runs `synpmx_avatar()` over it, and measures
how much of each one comes out the other side.

AVATAR does not explicityl model any of these relationships. 
There is no covariate model in it, no dose--response
model, no PK/PD link, and no notion of a treatment arm. Nothing in the generator
knows that weight ought to affect clearance. Whatever survives does so as a
**side effect of how the method works**; whole real subjects are blended
together, and a subject carries all of its properties at once. 
So the right expectation is a *tendency* to keep relationships, 
but there is not a guarantee.  

## The model the source data comes from

Everything here is simulated, so the truth is known exactly and no real patient
is involved. The four relationships being tested are the four places a covariate
or a dose enters these equations --- nothing else in the article puts a
relationship in.

**Pharmacokinetics.** One-compartment oral, first-order absorption and
elimination, for subject $i$ given dose $D_i$:

$$
C_i(t)=\frac{D_i}{V_i}\cdot\frac{k_a}{k_a-CL_i/V_i}
        \left(e^{-\frac{CL_i}{V_i}t}-e^{-k_a t}\right),
\qquad k_a = 1.2\ \mathrm{h^{-1}}
$$

**Allometric scaling --- relationship 1.** This is the covariate effect, and the
exponents are the numbers to remember:

$$
CL_i = 5 \left(\frac{WT_i}{70}\right)^{\mathbf{0.75}} e^{\eta_{CL,i}},
\qquad
V_i = 40 \left(\frac{WT_i}{70}\right)^{\mathbf{1.0}} e^{\eta_{V,i}}
$$

with $\eta_{CL}\sim N(0, 0.20^2)$ and $\eta_V\sim N(0, 0.15^2)$ on the log scale.

**Dose --- relationship 2.** The `low` arm gets $D=300$ and the `high` arm
$D=600$. The model is linear in dose, so exposure should scale exactly with it.

**Pharmacodynamics --- relationships 3 and 4.** A direct, linear inhibitory
effect on a baseline:

$$
E_i(t) = \bigl(E_{0,i} - 3.0\, C_i(t)\bigr)\,e^{\varepsilon},
\qquad E_{0,i} = 100\, e^{\eta_{E_0,i}}
$$

The slope of $3.0$ is what creates both the **treatment effect** (the high arm
reaches higher concentrations, so its response is driven further down) and the
**exposure--response** link (whatever a subject's exposure, response tracks it).
There is no separate "treatment" term: the arm matters only through the dose it
implies.

### What that predicts, before any data is generated

Two of these have closed-form consequences worth writing down, because they are
what the table later checks against:

Since $AUC_{0-\infty} = D/CL$ and $CL \propto WT^{0.75}$,

$$
\log\frac{AUC_i}{D_i} = \text{constant} - 0.75\log WT_i
$$

so **regressing dose-normalised log AUC on log weight should recover a slope of
$-0.75$** --- the allometric exponent itself, negative because a heavier subject
clears faster and is therefore exposed *less*. And because the model is linear
in dose, the **high/low arm AUC ratio should be $600/300 = 2$**.

Those two numbers, $0.75$ and $2$, are the ground truth. Everything below asks
how much of them survives.

## Generating the source dataset

Eighty subjects, forty per arm.

```{r source}
set.seed(2026)

make_subject <- function(id, arm, dose) {
  wt <- rnorm(1, 70, 12)
  # RELATIONSHIP 1 -- the covariate effect. The 0.75 and the 1.0 exponents are
  # the only place weight ever influences anything.
  cl <- 5  * (wt / 70)^0.75 * exp(rnorm(1, 0, 0.20))
  v  <- 40 * (wt / 70)^1.0  * exp(rnorm(1, 0, 0.15))
  ka <- 1.2
  pk_time <- c(0.5, 1, 2, 4, 8, 12, 24)
  pd_time <- c(0, 4, 12, 24)
  # RELATIONSHIP 2 -- `dose` enters here and nowhere else, so the arm affects
  # exposure only through the dose it implies.
  conc <- function(t) {
    dose / v * ka / (ka - cl / v) * (exp(-cl / v * t) - exp(-ka * t))
  }
  e0 <- 100 * exp(rnorm(1, 0, 0.10))
  rbind(
    data.frame(ID = id, TIME = 0, DV = NA_real_, AMT = dose, EVID = 1L,
               CMT = 1L, DVID = "PK", WT = wt, ARM = arm),
    data.frame(ID = id, TIME = pk_time,
               DV = conc(pk_time) * exp(rnorm(length(pk_time), 0, 0.08)),
               AMT = 0, EVID = 0L, CMT = 2L, DVID = "PK", WT = wt, ARM = arm),
    # RELATIONSHIPS 3 and 4 -- the `- 3 * conc(...)` term. It makes response
    # track exposure, and therefore makes the high-dose arm respond more.
    data.frame(ID = id, TIME = pd_time,
               DV = (e0 - 3 * conc(pd_time)) *
                 exp(rnorm(length(pd_time), 0, 0.05)),
               AMT = 0, EVID = 0L, CMT = 3L, DVID = "PD", WT = wt, ARM = arm)
  )
}

source_data <- do.call(rbind, c(
  lapply(1:40,  make_subject, arm = "low",  dose = 300),
  lapply(41:80, make_subject, arm = "high", dose = 600)
))
source_data <- source_data[
  order(source_data$ID, source_data$TIME, source_data$EVID == 0L), ]
rownames(source_data) <- NULL

roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "DVID",
  covariates = "WT",     # blended across donors into a new value
  keep = "ARM"           # copied verbatim from the one anchor subject
)
```

Note the two different treatments of the two non-PK columns, because it matters
for what follows. `WT` is a **covariate**: the synthetic subject gets a new
weight blended from its donors. `ARM` is **kept**: the synthetic subject gets
one real subject's arm label, copied unchanged from its anchor. Those are
different subjects --- the anchor supplies the skeleton and the kept columns,
the donors supply the values --- which is why the treatment effect surviving at
all is not obvious in advance.

## Measuring the four relationships

Each measurement is chosen so its answer can be compared against a number we
already know from the equations above.

**The dose normalisation matters.** To recover the allometric exponent, log AUC
must be regressed on log weight *after dividing out the dose*. Skip that and the
regression is confounded: the arms differ twofold in exposure, so any chance
imbalance between weight and arm assignment leaks straight into the slope. In
this cohort the unadjusted slope reads -0.90 while the dose-normalised one reads
-0.75, and only the second is the allometric exponent.

```{r measure}
relationships <- function(d) {
  obs <- d$EVID == 0 & !is.na(d$DV)
  pk  <- d[obs & d$DVID == "PK", ]
  pd  <- d[obs & d$DVID == "PD", ]
  ids <- as.character(pk$ID)

  auc <- tapply(seq_len(nrow(pk)), ids, function(i) {      # AUC 0-24 h, trapezoid
    t <- pk$TIME[i]; y <- pk$DV[i]; o <- order(t)
    sum(diff(t[o]) * (head(y[o], -1) + tail(y[o], -1)) / 2)
  })
  wt   <- tapply(pk$WT, ids, function(v) v[1])
  arm  <- tapply(as.character(pk$ARM), ids, function(v) v[1])
  dose <- tapply(d$AMT[d$EVID == 1], as.character(d$ID[d$EVID == 1]),
                 function(v) v[1])[names(auc)]
  nadir <- tapply(pd$DV, as.character(pd$ID), min)[names(auc)]
  ok <- is.finite(auc) & auc > 0 & is.finite(wt) & is.finite(nadir)

  c(# 1. allometric exponent on CL; truth = 0.75. Negated so it reads positive.
    allometric = unname(-coef(lm(log(auc[ok] / dose[ok]) ~ log(wt[ok])))[2]),
    # 2. dose effect; truth = 600/300 = 2 for linear PK.
    dose_ratio = unname(mean(auc[ok & arm == "high"]) /
                          mean(auc[ok & arm == "low"])),
    # 3. treatment effect: how much deeper the high arm's response goes.
    pd_effect  = unname(mean(nadir[ok & arm == "low"]) -
                          mean(nadir[ok & arm == "high"])),
    # 4. exposure-response: response against the exposure that drove it.
    er_slope   = unname(coef(lm(nadir[ok] ~ log(auc[ok])))[2]))
}

truth <- relationships(source_data)
round(truth, 3)
```

The first two land where the equations said they would: an allometric exponent
of about 0.75, and an arm ratio of about 2. That is the check that the source
dataset really contains what we think it contains, before AVATAR is asked to
preserve it.

## What the source data looks like

```{r source-plot, eval=has_ggplot2, echo=FALSE, fig.height=3.6}
library(ggplot2)
obs <- source_data$EVID == 0 & !is.na(source_data$DV)
plot_df <- data.frame(
  id = as.character(source_data$ID[obs]), time = source_data$TIME[obs],
  dv = source_data$DV[obs], endpoint = source_data$DVID[obs],
  arm = source_data$ARM[obs]
)
ggplot(plot_df, aes(time, dv, group = id, colour = arm)) +
  geom_line(alpha = 0.35) +
  facet_wrap(~ endpoint, scales = "free_y") +
  scale_colour_manual(values = c(low = "#1B6CA8", high = "#D95F02")) +
  labs(x = "Time (h)", y = "Value", colour = "Arm",
       title = "Source data: PK concentration and PD response by arm") +
  theme_minimal()
```

The high arm reaches higher concentrations and its response is driven further
down --- relationships 2 and 3, visible directly.

## Run AVATAR, several times

One synthetic dataset tells you little, because each run draws its own anchors
and donors. Thirty runs show both the central tendency and the spread, and the
spread is as much the point as the average.

```{r avatar}
set.seed(7)
replicates <- replicate(30, {
  synthetic <- suppressWarnings(
    synpmx_avatar(source_data, roles, seed = sample.int(1e6, 1))
  )
  relationships(synthetic)
})
```

```{r results, echo=FALSE}
knitr::kable(
  data.frame(
    relationship = c(
      "allometric exponent on CL (truth = 0.75)",
      "dose effect: AUC ratio high/low (truth = 2.0)",
      "treatment effect: PD nadir difference (low - high)",
      "exposure-response: PD nadir per log AUC"
    ),
    source    = round(truth, 3),
    synthetic = round(apply(replicates, 1, mean), 3),
    run_to_run_sd = round(apply(replicates, 1, sd), 3),
    retained  = paste0(round(100 * apply(replicates, 1, mean) / truth), "%")
  ),
  row.names = FALSE,
  caption = "Relationships in the source and in AVATAR output, 30 runs"
)
```

Every relationship survives, and all four land within roughly 10% of the source
value. The direction is consistently mild **dilution** rather than distortion:
the allometric exponent softens from 0.75 to about 0.68, and the
exposure--response slope loses about a tenth of its steepness. Blending pulls
each subject toward its neighbours, and a relationship measured across subjects
flattens slightly as a result.

The `run_to_run_sd` column is the honest part of the table. The *average* over
thirty runs sits close to the truth, but a single synthetic dataset can land
well off it --- the allometric exponent has a run-to-run SD of about 0.18 on a
value of 0.68, so one draw can easily read 0.5 or 0.9.

## Seeing it, rather than reading it

The two relationships that are easiest to misjudge from a table are the
covariate effect and the exposure--response link. Both are visible directly.

```{r relationship-plots, eval=has_ggplot2, echo=FALSE, fig.height=3.6}
subject_frame <- function(d, label) {
  obs <- d$EVID == 0 & !is.na(d$DV)
  pk <- d[obs & d$DVID == "PK", ]; pd <- d[obs & d$DVID == "PD", ]
  ids <- as.character(pk$ID)
  auc <- tapply(seq_len(nrow(pk)), ids, function(i) {
    t <- pk$TIME[i]; y <- pk$DV[i]; o <- order(t)
    sum(diff(t[o]) * (head(y[o], -1) + tail(y[o], -1)) / 2)
  })
  dose <- tapply(d$AMT[d$EVID == 1], as.character(d$ID[d$EVID == 1]),
                 function(v) v[1])[names(auc)]
  data.frame(
    dataset = label,
    wt    = as.numeric(tapply(pk$WT, ids, function(v) v[1])),
    auc   = as.numeric(auc),
    dose  = as.numeric(dose),
    arm   = as.character(tapply(as.character(pk$ARM), ids, function(v) v[1])),
    nadir = as.numeric(tapply(pd$DV, as.character(pd$ID), min)[names(auc)])
  )
}
one_syn <- suppressWarnings(synpmx_avatar(source_data, roles, seed = 4242))
both <- rbind(subject_frame(source_data, "Source"),
              subject_frame(one_syn, "Synthetic"))
both$dataset <- factor(both$dataset, levels = c("Source", "Synthetic"))

ggplot(both, aes(wt, auc / dose, colour = arm)) +
  geom_point(alpha = 0.7, size = 1.6) +
  geom_smooth(method = "lm", se = FALSE, colour = "black",
              linewidth = 0.6, formula = y ~ x) +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~ dataset) +
  scale_colour_manual(values = c(low = "#1B6CA8", high = "#D95F02")) +
  labs(x = "Weight (kg, log scale)", y = "Dose-normalised AUC (log scale)",
       colour = "Arm",
       title = "Relationship 1: the allometric weight effect",
       subtitle = "Black line is the fitted slope; its negative is the exponent") +
  theme_minimal()
```

Heavier subjects sit lower on both panels --- higher clearance, lower exposure
--- and the fitted line has a similar downward tilt in each. Note also that the
two arms overlay one another once AUC is divided by dose, which is what dose
proportionality looks like.

```{r er-plot, eval=has_ggplot2, echo=FALSE, fig.height=3.6}
ggplot(both, aes(auc, nadir, colour = arm)) +
  geom_point(alpha = 0.7, size = 1.6) +
  geom_smooth(method = "lm", se = FALSE, colour = "black",
              linewidth = 0.6, formula = y ~ x) +
  scale_x_log10() +
  facet_wrap(~ dataset) +
  scale_colour_manual(values = c(low = "#1B6CA8", high = "#D95F02")) +
  labs(x = "AUC (log scale)", y = "PD nadir",
       colour = "Arm",
       title = "Relationships 3 and 4: treatment effect and exposure-response",
       subtitle = "High arm sits right and low; response tracks exposure across both arms") +
  theme_minimal()
```

The treatment effect is the separation between the two colours, and the
exposure--response link is the downward slope running through both. Both are
present in the synthetic panel, with the cloud slightly tighter --- that
tightening is the blending, and it is why the fitted slope flattens a little.

## Is the treatment arm coherent with the dose?

`ARM` is copied from the anchor and the concentrations are blended from the
donors, so nothing structurally prevents a "high" label from arriving with
"low" data. Check it directly.

```{r arm-purity}
one_run <- suppressWarnings(synpmx_avatar(source_data, roles, seed = 99))
table(ARM = one_run$ARM[one_run$EVID == 1], dose = one_run$AMT[one_run$EVID == 1])
```

Every synthetic subject's arm label matches its dose. That is not luck, and it
is worth understanding why, because it is the mechanism the whole article rests
on.

## Why anything survives at all

Two independent mechanisms, neither of which was designed for this purpose.

**1. The event signature separates the arms.** Donors are first sought among
subjects with an identical event signature, and that signature *includes the
dose amount*. The 300 mg and 600 mg subjects therefore fall into different donor
groups automatically, and with 40 subjects per arm the floor of `k = 5` is met
without ever borrowing across them. The arm label copied from the anchor lands
on concentrations blended from same-arm donors.

**2. The profile distance separates them again.** Even where dose does not
distinguish two arms, the donor search ranks candidates by distance in a profile
space built from the covariates *and the DV trajectory*. A subject with higher
concentrations sits near other subjects with higher concentrations. The effect
you are trying to preserve is itself part of what selects the donors.

And underneath both: `.synthesize_covariates()` and `.synthesize_trajectories()`
are handed **the same donors and the same weights**. A synthetic subject whose
concentrations came mostly from one donor also gets mostly that donor's weight.
The relationship between them is never taken apart, so it does not have to be
put back together.

## Where it weakens

The second mechanism above also says where this breaks down: it works because
the effect is visible in the trajectory. An effect that is small relative to
between-subject variability does not organise the donor neighbourhoods, donors
get drawn from both groups, and the difference dilutes further than the mild
flattening seen here.

That is the boundary to keep in mind. Everything measured on this page had a
large, clean effect and a dose that separated the arms exactly.

## What to take from this

- Relationships put into a source dataset come out of AVATAR at close to their
  original size --- within about 10% here --- without anything in the method
  being told about them.
- That is a **tendency, not a specification**. It rests on donors being chosen
  by profile similarity and on covariates, endpoints, and the anchor's kept
  columns being tied to the same donor draw. Nothing declares it, nothing
  enforces it, and no argument controls its strength.
- It is strongest when the effect is large relative to between-subject
  variability, and when the arms differ in something the event signature already
  separates --- dose, schedule, route.
- Expect mild dilution. Every relationship here came back a little weaker than
  it went in, so a size read off synthetic data is a slight underestimate, not
  an inflation.
- Nothing here licenses estimating parameters from synthetic data. These
  measurements say the *structure* survives well enough to develop and debug
  analysis code against. A parameter estimate from AVATAR output is an estimate
  from blended data, and this article does not make it trustworthy.

The numbers on this page come from one simulated source with clean, strong
relationships and 80 subjects. They are an illustration of the mechanism, not a
general guarantee about your dataset. Run the same measurement on your own
source if the answer matters --- the `relationships()` function above is the
whole method, and it takes any data frame.
