---
title: "Example: are PK/PD, covariate, and treatment relationships preserved?"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center",
                      out.width = "100%")
library(synpmx)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

A synthetic dataset is only useful if the things you were going to *look for*
are still in it. This article builds a source dataset with four relationships
put in deliberately --- a covariate effect, a dose effect, a treatment effect,
and an exposure--response link --- runs `synpmx_avatar()` over it, and measures
how much of each one comes out the other side.

**Read this first, because it frames every number below.** AVATAR does not model
any of these relationships. There is no covariate model in it, no dose--response
model, no PK/PD link, and no notion of a treatment arm. Nothing in the generator
knows that weight ought to affect clearance. Whatever survives does so as a
**side effect of how the method works** --- whole real subjects are blended
together, and a subject carries all of its properties at once --- and not
because anything was specified. So the right expectation is a *tendency*, strong
under some conditions and weaker under others, rather than a guarantee. The last
two sections measure exactly where it weakens.

## A source dataset with known relationships

Everything here is simulated, so the truth is known and no real patient is
involved. Eighty subjects, two treatment arms at different doses, a
one-compartment oral PK model with allometric weight effects on clearance and
volume, and a direct-effect PD endpoint driven by concentration.

```{r source}
set.seed(2026)

make_subject <- function(id, arm, dose) {
  wt <- rnorm(1, 70, 12)
  cl <- 5  * (wt / 70)^0.75 * exp(rnorm(1, 0, 0.20))   # allometric on CL
  v  <- 40 * (wt / 70)      * exp(rnorm(1, 0, 0.15))   # allometric on V
  ka <- 1.2
  pk_time <- c(0.5, 1, 2, 4, 8, 12, 24)
  pd_time <- c(0, 4, 12, 24)
  conc <- function(t) {
    dose / v * ka / (ka - cl / v) * (exp(-cl / v * t) - exp(-ka * t))
  }
  e0 <- 100 * exp(rnorm(1, 0, 0.10))
  rbind(
    data.frame(ID = id, TIME = 0, DV = NA_real_, AMT = dose, EVID = 1L,
               CMT = 1L, DVID = "PK", WT = wt, ARM = arm),
    data.frame(ID = id, TIME = pk_time,
               DV = conc(pk_time) * exp(rnorm(length(pk_time), 0, 0.08)),
               AMT = 0, EVID = 0L, CMT = 2L, DVID = "PK", WT = wt, ARM = arm),
    data.frame(ID = id, TIME = pd_time,
               DV = (e0 - 3 * conc(pd_time)) *
                 exp(rnorm(length(pd_time), 0, 0.05)),
               AMT = 0, EVID = 0L, CMT = 3L, DVID = "PD", WT = wt, ARM = arm)
  )
}

source_data <- do.call(rbind, c(
  lapply(1:40,  make_subject, arm = "low",  dose = 300),
  lapply(41:80, make_subject, arm = "high", dose = 600)
))
source_data <- source_data[
  order(source_data$ID, source_data$TIME, source_data$EVID == 0L), ]
rownames(source_data) <- NULL

roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "DVID",
  covariates = "WT",     # blended across donors into a new value
  keep = "ARM"           # copied verbatim from the one anchor subject
)
```

Note the two different treatments of the two non-PK columns, because it matters
for what follows. `WT` is a **covariate**: the synthetic subject gets a new
weight blended from its donors. `ARM` is **kept**: the synthetic subject gets
one real subject's arm label, copied unchanged from its anchor. Those are
different subjects --- the anchor supplies the skeleton and the kept columns,
the donors supply the values --- which is why the treatment effect surviving at
all is not obvious in advance.

## The four relationships, measured the same way on any dataset

```{r measure}
relationships <- function(d) {
  obs <- d$EVID == 0 & !is.na(d$DV)
  pk  <- d[obs & d$DVID == "PK", ]
  pd  <- d[obs & d$DVID == "PD", ]
  ids <- as.character(pk$ID)

  auc <- tapply(seq_len(nrow(pk)), ids, function(i) {
    t <- pk$TIME[i]; y <- pk$DV[i]; o <- order(t)
    sum(diff(t[o]) * (head(y[o], -1) + tail(y[o], -1)) / 2)
  })
  wt     <- tapply(pk$WT, ids, function(v) v[1])
  arm    <- tapply(as.character(pk$ARM), ids, function(v) v[1])
  pd_min <- tapply(pd$DV, as.character(pd$ID), min)[names(auc)]
  ok     <- is.finite(auc) & auc > 0 & is.finite(wt) & is.finite(pd_min)

  c(wt_slope   = unname(coef(lm(log(auc[ok]) ~ log(wt[ok])))[2]),
    dose_ratio = unname(mean(auc[ok & arm == "high"]) /
                          mean(auc[ok & arm == "low"])),
    pd_effect  = unname(mean(pd_min[ok & arm == "low"]) -
                          mean(pd_min[ok & arm == "high"])),
    er_slope   = unname(coef(lm(pd_min[ok] ~ log(auc[ok])))[2]))
}

truth <- relationships(source_data)
```

## Run AVATAR, several times

One synthetic dataset tells you little, because each run draws its own anchors
and donors. Thirty runs show both the central tendency and the spread, and the
spread is as much the point as the average.

```{r avatar}
set.seed(7)
replicates <- replicate(30, {
  synthetic <- suppressWarnings(
    synpmx_avatar(source_data, roles, seed = sample.int(1e6, 1))
  )
  relationships(synthetic)
})
```

```{r results, echo=FALSE}
knitr::kable(
  data.frame(
    relationship = c(
      "WT to exposure (slope of log AUC on log WT)",
      "dose to exposure (AUC ratio, high/low arm)",
      "treatment effect (PD nadir difference, low - high)",
      "exposure to response (slope of PD nadir on log AUC)"
    ),
    source    = round(truth, 3),
    synthetic = round(apply(replicates, 1, mean), 3),
    run_to_run_sd = round(apply(replicates, 1, sd), 3),
    retained  = paste0(round(100 * apply(replicates, 1, mean) / truth), "%")
  ),
  row.names = FALSE,
  caption = "Relationships in the source and in AVATAR output, 30 runs"
)
```

Every relationship survives --- none collapses to nothing --- but read the
`retained` column carefully, because the three rows do **not** behave the same
way, and the differences are informative rather than noise.

- **The dose and treatment effects come through almost exactly** (104%, 99%).
- **The exposure--response slope is diluted** (89%). It relates two quantities
  that were *both* blended, and blending pulls each toward its local mean, so
  the slope between them flattens.
- **The covariate effect is amplified** (143%). That is not a rounding artifact:
  across 30 runs the standard error of that mean is about 0.08, so a source
  slope of -0.90 becoming -1.28 is a real shift, and it is in the *opposite*
  direction to the dilution above.

The amplification has a specific cause, and it is worth understanding because it
generalises to any covariate you declare. `WT` is one of the features used to
build the profile space that selects donors, so a subject's donors tend to
resemble it in weight *and* in exposure at the same time. Blending then averages
away the part of a subject's exposure that weight does not explain, while
leaving the part it does. The relationship comes out cleaner than it went in.

**So a declared covariate can look more predictive in AVATAR output than it is
in reality.** That is the single most important caveat on this page: it is the
opposite of the failure people expect from synthetic data, and it runs in the
direction that would make you over-confident.

## Is the treatment arm coherent with the dose?

`ARM` is copied from the anchor and the concentrations are blended from the
donors, so nothing structurally prevents a "high" label from arriving with
"low" data. Check it directly.

```{r arm-purity}
one_run <- suppressWarnings(synpmx_avatar(source_data, roles, seed = 99))
table(ARM = one_run$ARM[one_run$EVID == 1], dose = one_run$AMT[one_run$EVID == 1])
```

Every synthetic subject's arm label matches its dose. That is not luck, and it
is worth understanding why, because it is the mechanism the whole article rests
on.

## Why anything survives at all

Two independent mechanisms, neither of which was designed for this purpose.

**1. The event signature separates the arms.** Donors are first sought among
subjects with an identical event signature, and that signature *includes the
dose amount*. The 300 mg and 600 mg subjects therefore fall into different donor
groups automatically, and with 40 subjects per arm the floor of `k = 5` is met
without ever borrowing across them. The arm label copied from the anchor lands
on concentrations blended from same-arm donors.

**2. The profile distance separates them again.** Even where dose does not
distinguish two arms, the donor search ranks candidates by distance in a profile
space built from the covariates *and the DV trajectory*. A subject with higher
concentrations sits near other subjects with higher concentrations. The effect
you are trying to preserve is itself part of what selects the donors.

And underneath both: `.synthesize_covariates()` and `.synthesize_trajectories()`
are handed **the same donors and the same weights**. A synthetic subject whose
concentrations came mostly from one donor also gets mostly that donor's weight.
The relationship between them is never taken apart, so it does not have to be
put back together.

## Where it weakens

The second mechanism suggests the failure mode: it works because the effect is
visible in the trajectory. When an effect is small relative to between-subject
variability, the arms overlap in profile space, donors are drawn from both, and
the difference dilutes.

To show this without the dose doing the separating, here are two arms given the
**same** dose, differing only in relative bioavailability --- swept across effect
sizes and levels of variability in clearance.

```{r sweep}
make_f_subject <- function(id, arm, relF, iiv_cl) {
  wt <- rnorm(1, 70, 12)
  cl <- 5 * (wt / 70)^0.75 * exp(rnorm(1, 0, iiv_cl))
  v  <- 40 * (wt / 70) * exp(rnorm(1, 0, 0.15)); ka <- 1.2
  tt <- c(0.5, 1, 2, 4, 8, 12, 24)
  conc <- relF * 450 / v * ka / (ka - cl / v) *
    (exp(-cl / v * tt) - exp(-ka * tt))
  rbind(
    data.frame(ID = id, TIME = 0, DV = NA_real_, AMT = 450, EVID = 1L,
               CMT = 1L, WT = wt, ARM = arm),
    data.frame(ID = id, TIME = tt, DV = conc * exp(rnorm(7, 0, 0.08)),
               AMT = 0, EVID = 0L, CMT = 2L, WT = wt, ARM = arm)
  )
}
f_roles <- pmx_roles(id = "ID", time = "TIME", dv = "DV", amt = "AMT",
                     evid = "EVID", cmt = "CMT", covariates = "WT",
                     keep = "ARM")

auc_ratio <- function(d) {
  obs <- d$EVID == 0 & !is.na(d$DV); pk <- d[obs, ]; ids <- as.character(pk$ID)
  auc <- tapply(seq_len(nrow(pk)), ids, function(i) {
    t <- pk$TIME[i]; y <- pk$DV[i]; o <- order(t)
    sum(diff(t[o]) * (head(y[o], -1) + tail(y[o], -1)) / 2)
  })
  arm <- tapply(as.character(pk$ARM), ids, function(v) v[1])
  mean(auc[arm == "test"]) / mean(auc[arm == "ref"])
}

sweep <- do.call(rbind, lapply(
  list(c(1.15, 0.20), c(1.15, 0.40), c(1.80, 0.20)),
  function(cfg) {
    relF <- cfg[1]; iiv <- cfg[2]
    set.seed(2026)
    src <- do.call(rbind, c(
      lapply(1:40,  make_f_subject, arm = "ref",  relF = 1,    iiv_cl = iiv),
      lapply(41:80, make_f_subject, arm = "test", relF = relF, iiv_cl = iiv)))
    src <- src[order(src$ID, src$TIME, src$EVID == 0L), ]
    rownames(src) <- NULL
    observed <- auc_ratio(src)
    set.seed(5)
    got <- replicate(15, auc_ratio(suppressWarnings(
      synpmx_avatar(src, f_roles, seed = sample.int(1e6, 1)))))
    data.frame(
      effect = paste0(relF, "x"), iiv_cl = iiv,
      source = round(observed, 3), synthetic = round(mean(got), 3),
      retained = paste0(round(100 * (mean(got) - 1) / (observed - 1)), "%")
    )
  }))
```

```{r sweep-table, echo=FALSE}
knitr::kable(sweep, row.names = FALSE,
             caption = "Retention of an arm effect when dose cannot separate the arms")
```

The pattern is the one the mechanism predicts. A large effect is preserved
almost completely, because it is loud enough to organise the donor
neighbourhoods by itself. A small effect against large variability is retained
least --- the arms blur together, donors cross between them, and the difference
shrinks toward the cohort mean.

Here retention is at or below 100% throughout, so an effect that the donor
search cannot see gets *diluted*. Put beside the covariate result above, the
rule is about what the donor search knows:

| The relationship is with... | Effect on the estimate | Why |
|---|---|---|
| a declared **covariate** | **amplified** | it helps choose the donors, so blending sharpens it |
| something in the **event signature** (dose, schedule, route) | preserved | donor groups line up with it exactly |
| a **kept** label the signature cannot see | diluted | donors cross the groups, and the difference averages out |
| another **blended** quantity (exposure to response) | diluted | both sides are pulled toward their local means |

## What to take from this

- Relationships put into a source dataset largely come out of AVATAR, without
  anything in the method being told about them --- but not at their original
  size. Expect covariate effects to be somewhat overstated and effects between
  blended quantities to be understated.
- That is a **tendency, not a specification**. It rests on donors being chosen
  by profile similarity and on covariates, endpoints, and the anchor's kept
  columns being tied to the same donor draw. Nothing declares it, nothing
  enforces it, and no argument controls its strength.
- It is strongest when the effect is large relative to between-subject
  variability, and when the arms differ in something the event signature already
  separates --- dose, schedule, route.
- It is weakest for subtle effects in noisy data, which is exactly the situation
  where you would most want to trust a synthetic dataset. Do not use AVATAR
  output to decide whether a small effect is real --- and do not read a strong
  covariate relationship in synthetic data as evidence that one exists, because
  the method sharpens exactly those.
- Nothing here licenses estimating parameters from synthetic data. These
  measurements say the *structure* survives well enough to develop and debug
  analysis code against. A parameter estimate from AVATAR output is an estimate
  from blended data, and this article does not make it trustworthy.

The numbers on this page come from one simulated source with clean, strong
relationships and 80 subjects. They are an illustration of the mechanism, not a
general guarantee about your dataset. Run the same measurement on your own
source if the answer matters --- the `relationships()` function above is the
whole method, and it takes any data frame.
