---
title: "Privacy Protections in the PMX Model Generator"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Privacy Protections in the PMX Model Generator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
# Every fit is read from the installed package, so this article fits nothing
# and needs neither nlmixr2 nor a compiler.
stored_fit <- function(file) {
  path <- system.file("extdata", file, package = "synpmx")
  if (nzchar(path)) readRDS(path) else NULL
}
fit <- stored_fit("warfarin-model-fit.rds")
misdosed_fit <- stored_fit("warfarin-misdosed-model-fit.rds")
misdosed_kept_fit <- stored_fit("warfarin-misdosed-kept-model-fit.rds")
extreme_fit <- stored_fit("warfarin-extreme-model-fit.rds")
run <- has_nlmixr2data && !is.null(fit) && !is.null(misdosed_fit) &&
  !is.null(misdosed_kept_fit) && !is.null(extreme_fit)
plots <- run && has_ggplot2
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")

# Three significant digits on measured quantities, written out rather than
# with a shared exponent, since one column can hold a clearance beside a
# variance a thousand times smaller. Counts are exempt.
digits3 <- function(x) {
  x <- as.data.frame(x)
  attr(x, "who") <- NULL
  for (column in names(x)) {
    if (is.numeric(x[[column]]) && !column %in% c("patients")) {
      x[[column]] <- vapply(x[[column]], function(v) {
        format(signif(v, 3), scientific = FALSE, drop0trailing = TRUE,
               trim = TRUE)
      }, character(1))
    }
  }
  x
}
```

```{r membership-simulation, include=FALSE, file="pmxmodel-privacy-simulation.R"}
```

```{r membership-tables, include=FALSE}
# Computed by the simulation article's code, from the same seed, so the numbers
# quoted here are the ones that article explains.
tables <- membership_tables()
population_table <- tables$population
attendance_table <- tables$attendance
rounding_change <- sprintf("%.3f", max(abs(population_table$unrounded -
                                             population_table$two_figures)))
rounding_gain <- sprintf("%.2f", max(population_table$unrounded -
                                       population_table$released))
floor_change <- sprintf("%.2f", max(abs(attendance_table$without_floor -
                                          attendance_table$released)))
population_auc <- function(patients, off_by) {
  sprintf("%.2f", population_table$released[
    population_table$patients == patients & population_table$off_by == off_by])
}
attendance_auc <- function(visits, patients, correlation = 0) {
  sprintf("%.2f", attendance_table$released[
    attendance_table$visits == visits & attendance_table$patients == patients &
      attendance_table$correlation == correlation])
}
```

## Introduction

The pharmacometric (PMX) model generator, `synpmx_model()`, fits a set of models to a study and then simulates from these models a synthetic dataset, for the purpose of transferring the data outside the GxP environment, but staying within the secure system ecosystem. The models that are fit are:

1.  PopPK model fit to the PK data
2.  PD model fit to each PD observation type, pooled across cohorts
3.  An attendance model, to allow for simulation of missed visits
4.  A dosing model to simulate dose reductions and missed doses

This article clarifies what information from a study (via the model parameters) can leave the GxP computing environment, and what risk remains. While this approach offers many privacy protections which reduce how much any one patient can impact the synthetic data, the algorithm does not offer a formal differential privacy (DP) guarantee.

## Why a data generation method that does not formally guarantee DP was chosen {#why-not-dp}

DP limits how much any one person's data can change what is released, whatever else an attacker knows \[1\]. A DP mechanism achieves this limit by adding calibrated noise to every released number.

For a Phase 1 study (12-60 patients), a DP guarantee would require adding so much noise to the synthetic dataset that it would no longer meaningfully resemble the clinical data. Moreover, the DP guarantee would be protecting the data from an attacker that the use does not face. That use is synthetic data for developing analysis code, shared with people who work under the study's own access controls or an agreement like them. The case rests on three points:

1.  **The release holds what the study's own reports ultimately make public, more coarsely.** A population PK report gives each typical value, its between-subject variability and the residual error, with their standard errors, and regulators ask for exactly that \[18, 19\]. The study's main publication gives each arm's baseline characteristics and how many patients discontinued \[20\]. This release holds the same kinds of number at one significant figure, pooled over the arms and the visits, without standard errors, and a few dozen at most ([Released Parameters at a Glance](#released-parameters-at-a-glance)).
2.  **Any attacker already holds the answer they are seeking.** Membership inference needs the candidate's own measurements to assess whether that candidate was in the study; measurements such as concentrations, visit attendance, and dose changes. Outside the company sponsoring the study, this information exists only at the sites and in routine care of the patient. Thus whoever holds a candidate's trial record knows the candidate was in the trial, and learning it again from the release discloses nothing.
3.  **DP at small cohort sizes replaces the study with noise**, as [the appendix](#dp-noise-small-study) computes, and for this use noise is worse than no data. The [calibrated generator's evaluation](https://iamstein.github.io/synpmx/articles/calibrated-public-data-examples.html) finds a DP release at phase 1 sizes worse than generating from a public prior, which reads nothing.

## Privacy Assessment Summary

### What the Generator Releases

A population PK/PD model describes how the concentration of a drug in the blood, and a response to the drug, change over time across a population of patients. It is a short list of numbers: a **typical value** for each parameter, such as how fast the body clears the drug; a **between-subject variance** for each, which says how far patients spread around the typical value; and a **residual error**, the scatter of single measurements around a patient's own curve. `synpmx_model()` estimates these from the study's patients with `nlmixr2`, an R package for fitting population models.

The generator also summarizes the study's design as it was carried out: the number of patients in each treatment group (**arm**), each arm's dose schedule, how often doses were reduced, skipped or stopped, the share of scheduled visits at which patients were measured, for each kind of measurement, the distribution of each baseline covariate such as body weight, age or sex, and, for an endpoint recorded in levels such as a yes/no response or a toxicity grade, how often each level was recorded. Everything in that list read from how patients behaved, rather than from the protocol, is one set of numbers for the whole study, pooled over the arms and the visits, as the PD time courses are pooled over the arms.

What leaves the environment that holds the study is the **release**: these numbers, assembled by `model_release()` from a list of permitted fields, together with a description of the table to generate. Synthetic data is simulated from the release and from random numbers alone, so it discloses nothing the release does not, and the release travels with the synthetic data as an attribute. A reviewer can read every number in a release with `str(unclass(release))`, and [Released Parameters at a Glance](#released-parameters-at-a-glance) counts them for each public study.

### What Is Protected, and What Is Not

**✅ No patient record, identifier or per-patient estimate leaves the study.** The release holds numbers about the population and about arms. Synthetic subject identifiers are new. Each patient's estimated deviation from the typical values is discarded.

**✅ No released number is one patient's value, and every released frequency rests on at least three patients on each side.** Minima, maxima, medians and modes are replaced by quantities no one patient decides. A frequency, a category level, or a value copied onto every patient of an arm is released only where the patients on each side of it number zero or at least three.

**✅ A patient who moves a released estimate by 15% or more is left out of the estimates.** Estimation reads how far each patient moves each model parameter estimate, leaves out anyone who moves one by 15% or more (for a between-subject spread, by 0.15 in its standard deviation (SD) on the log scale), and estimates again. A patient left out of one estimation step is still used in the others. The reading is approximate and does not cover every estimate; its limits are among the outstanding risks.

**⚠️ Membership inference remains possible for an attacker who holds a candidate's own trial data.** In simulation, an attacker with a candidate's own measurements and exact reference values for the population ranks a study member above a non-member with probability `r population_auc(12, 0)` when the population model is fitted to 12 patients and `r population_auc(60, 0)` when it is fitted to 60. This probability is the area under the curve (AUC), where 0.5 is chance and 1 is certainty. A reference off by 20% brings the two to `r population_auc(12, 0.2)` and `r population_auc(60, 0.2)`. The release holds no table per visit, because one would be more exposed: a share of patients sampled at each of 20 visits gives `r attendance_auc(20, 10)` when 10 patients stand behind each share and `r attendance_auc(20, 100)` when 100 do. Attendance is one rate per endpoint instead, and each binary or ordinal endpoint one set of level frequencies. A rate exposes about as much as one visit's share, so five endpoints' rates behind 10 patients give about `r attendance_auc(5, 10)`. Rounding the estimates to one significant figure lowers these numbers by up to `r rounding_gain`, and the three-patient floor moves them by at most `r floor_change`; cohort size and the number of released numbers decide them. [Membership inference against the PMX model release](https://iamstein.github.io/synpmx/articles/pmxmodel-privacy-simulation.html) shows how each number is computed.

**❌ No formal guarantee.** Nothing bounds what an attacker who knows every other patient learns about the remaining one.

## The Privacy Protections {#the-protections}

### 1. An Allowlist Decides What Leaves

`model_release()` copies the fields Step 6 of the [algorithm](https://iamstein.github.io/synpmx/articles/pmxmodel-algorithm.html) reads. Inside the PK models, the PD shapes, the dosing models, the endpoint list and the settings, a second list names the parts it copies; the schema, the arm, visit and covariate models and the remaining fields are copied whole. A field added to the fit later stays behind unless it is added to a list. The pattern is the glimpse-then-simulate split of Novartis's `synadam` package \[7\], which writes the summary that synthetic data is generated from to one file a reviewer can read.

### 2. No Per-Patient Quantity

The empirical Bayes estimates (EBEs) are each patient's own random effects: how far that patient's parameters sit from the typical values, estimated from the patient's own measurements. They are read twice at estimation, for the correlations between covariates and random effects and for the influence reading of protection 8, and then dropped. Generation draws new random effects from the between-subject covariance matrix.

### 3. No Identifier

Synthetic subject identifiers (IDs) are minted at generation. No source label is kept, including the levels of an ID column stored as a factor. A numeric ID keeps only the largest source ID rounded up to a power of ten, which stops a synthetic ID colliding with a real one without storing the real one. `synadam` refuses to run when an ID column is left undeclared, for the same reason \[7\].

### 4. No Single Patient's Value

Every summary that could have been one patient's value is replaced:

| Quantity | Would have been | Is |
|------------------------|------------------------|------------------------|
| Emission floor, where no assay limit is declared | half the smallest value reported | half the lowest value at least three patients reached, rounded down to 1, 2 or 5 times a power of ten |
| Allometric reference weight | the cohort median | the median to one significant figure |
| Planned dose amount or infusion rate, where no value is shared | the most common value, which is then one patient's | the arm's mean to two significant figures |
| Continuous covariate | a mean and SD over every patient, and a median | a trimmed mean and SD (protection 6), no median |
| Endpoint type description | text quoting the count of levels and the lowest and highest value observed | the type, without the text |

Statistical disclosure control (SDC), the field that protects published statistical tables, treats extreme values the same way, by top-coding and coarsening them \[8\].

### 5. Several Patients Behind Every Frequency

A frequency resting on one patient is that patient. The threshold rule of SDC publishes a count only where at least *k* units stand behind it \[8\], and `synadam` masks values held once \[7\]. The rule applies to both sides of a frequency: attendance at a visit that all but one patient kept says as much about the one who missed it as attendance by one patient says about the one who came. Here *k* is 3 by default:

- arms below `min_arm_patients` are dropped before anything is fitted;
- a nominal visit slot, a dose level and a planned cycle are kept only where at least `min_arm_patients` patients reached them, and an arm has a visit only where at least that many of its patients were observed there;
- attendance is one rate per endpoint, and is 1 where fewer than `min_arm_patients` patients missed any of the endpoint's visits;
- a dose-change rate, pooled over every arm, whose events fewer than `min_arm_patients` patients had is zero;
- a categorical covariate level, and a binary or ordinal endpoint's level, below `min_category_patients` patients is excluded, or folded into the most common level;
- a `keep` value of any type, copied from the arm's first patient as `keep` declares, is written as missing where fewer than `min_category_patients` patients in the arm hold it;
- the description of the table to generate keeps, for a factor column, only the levels its arms carry or at least `min_category_patients` patients hold, and for a binary or ordinal endpoint only the levels that many patients recorded, so the label of an arm too small to keep, or a grade one patient reached, stays behind.

The rule stops a frequency from singling out one or two patients. It does not stop membership inference, which adds up small signals over many frequencies ([Membership Inference Against the Release](#membership-inference-against-the-release)). Pooling is what weakens that: attendance is one rate per endpoint, the dose-change rates are three, and a binary or ordinal endpoint is one set of level frequencies, each pooled over the arms and the visits rather than resting on one arm at one visit. `pd_by_arm = TRUE` reads the endpoint frequencies per arm again, with the PD time courses.

### 6. Covariate Summaries Without Their Extremes

A continuous covariate is summarized by a mean and SD taken after the highest and lowest 5% of patients are set aside, at least one at each end. The SD of what remains understates the spread, because the tails are what was removed, so it is divided by the SD a standard normal keeps under the same trimming. With a share $\alpha$ removed from each end and $z = \Phi^{-1}(1-\alpha)$,

$$\mathrm{SD} = \frac{\mathrm{SD}_\text{trimmed}}{\sqrt{1 - \dfrac{2 z \phi(z)}{1-2\alpha}}}.$$

This is the trimmed mean of robust statistics \[9\]. Trimming is also how DP mean estimators bound one person's influence \[10\], since no single value can then move the result by more than the trimmed range allows.

### 7. One Significant Figure

Every released PK and PD estimate is rounded to one significant figure: PK fixed effects, between-subject variances, residual error and PD shape parameters. Covariate summaries are rounded to two, because one would put a mean height of 172 cm at 200. Rounding is the other standard coarsening of SDC \[8\]. One significant figure moves an estimate by 9% on average and by 33% at worst, against 1% and 5% for two, and in exchange it weakens membership inference where two figures do not: its step is as wide as an estimate's sampling variation, so the released value is usually the same with a given patient in the study or out ([Membership Inference Against the Release](#membership-inference-against-the-release)). Rounding does not hide an outlier's effect on a variance, which protection 8 measures. The full-precision estimates stay on the fit, outside the release, for whoever ran it. The frequencies are not rounded: each is an exact count over the study, held to the floor of protection 5 instead.

### 8. How Far One Patient Moves Each Estimate

DP calls the most one person can move a released number its *sensitivity*, taken over every possible dataset \[1\]. The same question asked of the study at hand, leaving each patient out in turn and recording the largest move, is *local sensitivity*, and in regression the same idea is Cook's distance \[11\]. It is computed at estimation without refitting:

- A population typical value on the log scale sits close to the mean of the patients' individual log parameters, so leaving out patient $i$ moves it by about $\eta_i/(n-1)$, where $\eta_i$ is that patient's random effect. The PD typical baseline is read the same way, from each patient's own baseline.
- The between-subject variance sits close to the mean of the squared random effects, so leaving out patient $i$ scales it by $\frac{(S-\eta_i^2)/(n-1)}{S/n}$, with $S=\sum_j \eta_j^2$.
- The spread of the PD baselines and each covariate summary are recomputed with the patient left out, which needs no approximation.

The largest move over all patients is kept for each estimate: in percent for a typical value, and in points of the between-subject SD on the log scale, close to a coefficient of variation (CV) in percent, for a spread. A move of 15 is a review; a move of 30 fails, warns at estimation, and names the patient on the console. The name is printed in the environment that holds the study and is not stored: the fit keeps the largest move per estimate, never anything per patient. On a parameter whose between-subject SD is 0.27, a patient 3 SD from the mean in a cohort of 32 moves the typical value by about 3% and the spread by about 4 points. Shrinkage, the pull of a random effect toward zero when the patient's own measurements say little about their parameters, makes the reading understate the move, so it is reported beside it.

A patient who moves an estimate by 15 or more is then left out of the estimates, and estimation runs again. The patient is left out of the PK fit, the PD fits and the covariate summaries, and stays in the dosing, visit and arm models, so the cohort and arm sizes the release describes do not change. Rounds repeat until the reading flags nobody, and stop before a tenth of the cohort has been left out. Each patient left out is named on the console; the fit records how many, and which estimates they moved. `drop_influential = FALSE` keeps every patient and leaves the reading as a review or a failure.

The rounds are repeated because leaving out the most influential patients makes the next ones the most influential, which Carlini and colleagues call the privacy onion effect \[12\]. The DP counterpart of the reading is propose-test-release, which checks that the study at hand is far from any study in which one patient could move an estimate far, and only then releases it with noise scaled to that bound \[13\].

## The Five Checks

`model_privacy_checks()` runs five checks on a fit, each with its pass criterion. The first four ask whether the release holds something about one patient that it should not. The fifth is the reading of protection 8.

| Check | Question | Passes when |
|------------------------|------------------------|------------------------|
| P1 | No per-patient table is released | the release holds only fields on the allowlist, none of them per patient |
| P2 | No source identifier is released | no source ID label, and no source ID value as the synthetic ID offset |
| P3 | No single patient's value is released | floors on the 1-2-5 series, the reference weight rounded, no covariate median, no factor level that no arm carries |
| P4 | Every released frequency rests on several patients, on both sides | each side of every frequency is none or at least `min_arm_patients` patients (`min_category_patients` for categorical levels and values) |
| P5 | No single patient moves a released estimate far | every estimate moves less than 15, after any patient who moved one further has been left out |

P1 to P3 are structural and read the release. They pass by construction on a fit from this version, and exist to catch a release assembled some other way or a fit stored before the protections existed. P3 reads the emission floors, the covariate summaries, the reference weight and the factor levels of the columns copied onto arms. P4 recounts, at estimation, the smallest group of patients behind each kind of released frequency after the rules of protection 5 have run, including every factor and endpoint level the schema stores, and P5 is the reading of protection 8. A release carries the verdicts of P4 and P5, and the fit the counts behind them.

```{r checks, eval=run}
checks <- model_privacy_checks(fit)
as.data.frame(checks)[, c("check", "verdict", "result")]
```

The reading behind P5 is the `influence` attribute of the full fit's checks, one row per released estimate: the released value, the largest move any one patient causes, its unit, the number of patients it was read over, the shrinkage of the random effect it came from, and the verdict. No row is about a patient.

```{r influence, eval=run}
digits3(attr(checks, "influence"))
```

The recount behind P4 is the `frequencies` attribute: for each kind of released frequency, the smallest group of patients behind any one of them, the floor it is held to, and how many values the rules changed to meet it.

```{r frequencies, eval=run}
attr(checks, "frequencies")[, c("quantity", "smallest", "threshold",
                                "adjusted")]
```

The same checks on the release report P5 as a verdict only. A passing release states that every estimate moved less than 15, and nothing more, so the one data-dependent thing it carries about the reading is the verdict. A release that does not pass names its worst estimate and the size of the move, which is computed from one patient's data; such a release is not meant to leave.

```{r release-checks, eval=run}
as.data.frame(model_privacy_checks(release))[, c("check", "verdict", "result")]
```

## Three Constructed Cases

Each case is `warfarin` with one or more patients changed, built once by `synpmx_model_estimate()` and stored with the package. None is a real study.

```{r constructed-data, eval=run}
warfarin <- as.data.frame(nlmixr2data::warfarin)
warfarin$ntime <- warfarin$time
roles <- pmx_roles(
  id = "id", time = "time", nominal_time = "ntime", dv = "dv", amt = "amt",
  evid = "evid", dvid = "dvid", covariates = c("wt", "age", "sex")
)
patients <- sort(unique(warfarin$id))

misdosed <- warfarin
on_cp <- misdosed$id == patients[[5L]] & misdosed$evid == 0 &
  misdosed$dvid == "cp"
misdosed$dv[on_cp] <- misdosed$dv[on_cp] * 1000

extreme <- warfarin
extreme$wt[extreme$id == patients[[3L]]] <- 250
extreme$age[extreme$id == patients[[8L]]] <- 95
extreme$sex <- as.character(extreme$sex)
extreme$sex[extreme$id == patients[[12L]]] <- "not recorded"
extreme$sex <- factor(extreme$sex)
```

```{r constructed-fits, eval=FALSE}
misdosed_fit <- synpmx_model_estimate(misdosed, roles, seed = 1)
extreme_fit <- synpmx_model_estimate(extreme, roles, seed = 1)
```

### A Patient Given a Thousand Times the Recorded Dose

One patient received a thousand times the dose the record shows, so a linear model sees concentrations a thousand times higher than the recorded dose implies. The case is fitted twice: as estimation runs by default, and with `drop_influential = FALSE`, which keeps the patient so that the reading can be seen to catch them.

```{r misdosed-fits, eval=FALSE}
misdosed_fit <- synpmx_model_estimate(misdosed, roles, seed = 1)
misdosed_kept_fit <- synpmx_model_estimate(misdosed, roles, seed = 1,
                                           drop_influential = FALSE)
```

Kept, the patient fails P5. The variance terms absorb the excess exposure, and estimation warned that one patient moved a released estimate by 30 or more, naming the patient on the console. The stored fit names nobody.

```{r misdosed-kept, eval=run}
kept_checks <- model_privacy_checks(misdosed_kept_fit)
as.data.frame(kept_checks)[, c("check", "verdict", "result")]
digits3(attr(kept_checks, "influence"))
```

By default, estimation left the patient out of the estimates and ran again. The fit records how many patients were left out and which estimates they moved, and the checks pass:

```{r misdosed-default, eval=run}
misdosed_fit$privacy$left_out$patients
digits3(misdosed_fit$privacy$left_out$because)
as.data.frame(model_privacy_checks(misdosed_fit))[, c("check", "verdict",
                                                      "result")]
```

The synthetic data shows the difference. Kept, the patient's exposure becomes between-subject variability, and the synthetic patients spread wider than the rest of the source cohort. Left out, they do not:

```{r misdosed-plot, eval=plots, fig.height=4}
concentration <- function(data, label) {
  rows <- data$evid == 0 & data$dvid == "cp" & data$dv > 0
  data.frame(dataset = label, id = as.character(data$id[rows]),
             time = data$time[rows], dv = data$dv[rows])
}
kept_synthetic <- suppressWarnings(
  synpmx_model_generate(misdosed_kept_fit, seed = 11))
synthetic <- suppressWarnings(synpmx_model_generate(misdosed_fit, seed = 11))
plotted <- rbind(concentration(misdosed, "Source"),
                 concentration(kept_synthetic, "Synthetic, patient kept"),
                 concentration(synthetic, "Synthetic, patient left out"))
plotted$dataset <- factor(plotted$dataset, levels = unique(plotted$dataset))
ggplot2::ggplot(plotted, ggplot2::aes(time, dv, group = id,
                                      colour = dataset)) +
  ggplot2::geom_line(alpha = 0.4) +
  ggplot2::facet_wrap(~dataset) +
  ggplot2::scale_y_log10() +
  ggplot2::scale_colour_manual(values = c(unname(comparison_colours),
                                          "#7570B3")) +
  ggplot2::labs(x = "Time (hours)", y = "Warfarin concentration") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")
```

The scorecard reads each release's verdict as row E3, so the finding reaches anyone who scores the synthetic data:

```{r misdosed-card, eval=run}
kept_card <- synpmx_scorecard(misdosed, kept_synthetic, roles)
card <- synpmx_scorecard(misdosed, synthetic, roles)
data.frame(fit = c("patient kept", "patient left out"),
           E3 = c(kept_card$verdict[kept_card$check == "E3"],
                  card$verdict[card$check == "E3"]),
           result = c(kept_card$result[kept_card$check == "E3"],
                      card$result[card$check == "E3"]))
```

Leaving the patient out protects the release and keeps the synthetic data plausible. The record is still wrong, and correcting the patient's dosing is the remedy for any analysis of the real study.

### Extreme Covariates

One patient weighs 250 kg and another is 95 years old. Neither enters the covariate summaries, because each is the most extreme patient on their side and trimming sets them aside:

```{r extreme-covariates, eval=run}
summary_of <- function(model) {
  data.frame(covariate = c("wt", "age"),
             geometric_mean = exp(c(model$wt$meanlog, model$age$meanlog)),
             sd_log = c(model$wt$sdlog, model$age$sdlog))
}
rbind(cbind(study = "warfarin", summary_of(fit$covariates)),
      cbind(study = "with the two extreme patients",
            summary_of(extreme_fit$covariates)))
```

The same two summaries taken over every patient, without trimming, move with the extreme patients:

```{r untrimmed, eval=run}
first <- function(data) data[!duplicated(data$id), ]
untrimmed <- function(data) {
  data.frame(covariate = c("wt", "age"),
             geometric_mean = exp(c(mean(log(first(data)$wt)),
                                    mean(log(first(data)$age)))),
             sd_log = c(stats::sd(log(first(data)$wt)),
                        stats::sd(log(first(data)$age))))
}
digits3(rbind(cbind(study = "warfarin", untrimmed(warfarin)),
              cbind(study = "with the two extreme patients",
                    untrimmed(extreme))))
```

P5 reads the covariate rows of the influence table: leaving out any one patient barely moves them, the extreme patients included.

```{r extreme-influence, eval=run}
extreme_influence <- attr(model_privacy_checks(extreme_fit), "influence")
digits3(extreme_influence[extreme_influence$group == "covariate", ])
```

### A Category Nobody Else Holds

A third changed patient has a sex recorded as a level no other patient holds. The level is excluded from the covariate distribution and from the schema the generated table is written in, so no synthetic patient can carry it:

```{r rare-level, eval=run}
extreme_fit$covariates$sex$levels
levels(extreme_fit$schema$prototypes$sex)
```

## The Checks Across the Public Studies

The checks on every stored fit of the [public-data evaluation](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html), and the estimate each study's most influential remaining patient moves furthest.

```{r survey, eval=run}
files <- c(warfarin = "warfarin-model-fit.rds",
           theo_md = "theo-md-model-fit.rds",
           mad = "mad-model-fit.rds",
           case1_pkpd = "case1-pkpd-model-fit.rds",
           wbcSim = "wbcsim-model-fit.rds",
           mavoglurant = "mavoglurant-model-fit.rds",
           nimoData = "nimo-model-fit.rds",
           pheno_sd = "pheno-model-fit.rds",
           mixroute_sim = "mixroute-sim-model-fit.rds",
           onc_sim = "onc-sim-model-fit.rds")
rows <- lapply(names(files), function(study) {
  one <- stored_fit(files[[study]])
  if (!inherits(one, "pmx_fitted_model")) return(NULL)
  checks <- model_privacy_checks(one)
  verdicts <- stats::setNames(checks$verdict, checks$check)
  data.frame(study = study, patients = one$n_source,
             left_out = if (is.null(one$privacy$left_out)) 0L else
               one$privacy$left_out$patients, t(verdicts),
             largest = sub("; after leaving.*$", "",
                           checks$result[checks$check == "P5"]),
             check.names = FALSE)
})
do.call(rbind, rows[!vapply(rows, is.null, logical(1))])
```

`left_out` counts the patients each study's estimation left out for moving an estimate by 15 or more, and `largest` is the largest move among the patients that remain. A move is a fraction of the cohort's spread divided by the cohort size, so the same patient moves a twelve-patient fit about five times as far as a sixty-patient one.

## Checks on the Synthetic Table

The checks above read the fit. `synpmx_scorecard()` reads the synthetic table against its source, and its section B asks whether any synthetic patient reproduces a real one. Four of its rows do not apply to this generator, because no synthetic patient is assembled from a real patient's record; B3, B4b and B5 do.

```{r card, eval=run}
synthetic <- synpmx_model_generate(fit, seed = 11)
card <- synpmx_scorecard(warfarin, synthetic, roles)
as.data.frame(card)[substr(card$check, 1, 1) %in% c("B", "E"),
                    c("check", "question", "result", "verdict")]
```

B3 compares how close synthetic patients sit to real ones against how close real patients sit to each other, B4b asks whether any generated vector of values copies a real patient's, and B5 asks whether a level fewer than three real patients held reached the output. [The scorecard](https://iamstein.github.io/synpmx/articles/scorecard.html) documents each row and its pass criterion.

## Membership Inference Against the Release {#membership-inference-against-the-release}

A membership-inference attack gives each candidate a score that is high when the release has moved, away from what the population would give, toward the candidate's own values. Members of the study score higher on average than people who were not in it, and the attack's strength is read as the AUC, the probability that a randomly chosen member scores above a randomly chosen non-member: 0.5 is chance and 1 is certain identification. The attacker modelled holds the candidate's own measurements and reference values for the population, and does not know the other patients. The release is reduced to its two kinds of number: a population model of five parameters, each released as a typical value and a between-subject variance, and the share of patients sampled at each scheduled visit. [Membership inference against the PMX model release](https://iamstein.github.io/synpmx/articles/pmxmodel-privacy-simulation.html) works through the calculation on one small study, shows both tables in full, and places each public study's release against them. Three readings:

- **Cohort size decides the population model's exposure.** With an exact reference the AUC falls from `r population_auc(12, 0)` at 12 patients to `r population_auc(60, 0)` at 60, and a reference off by 20% brings the two to `r population_auc(12, 0.2)` and `r population_auc(60, 0.2)`.
- **One significant figure lowers it; two would not.** Scored on the same simulated studies, rounding to one figure, as the release does, lowers the AUC by up to `r rounding_gain`, while two figures move it by at most `r rounding_change`, because a one-figure step is as wide as an estimate's sampling variation and a two-figure step is not.
- **A table per visit would be the most exposed part of a release, so the release holds none.** A share of patients sampled at each of 20 visits gives `r attendance_auc(20, 10)` with 10 patients behind each share, `r attendance_auc(20, 30)` with 30 and `r attendance_auc(20, 100)` with 100, and the three-patient rule moves it by at most `r floor_change`, because the rule acts only on shares near 0 or 1. Attendance is one rate per endpoint instead, and a binary or ordinal endpoint one set of level frequencies, each pooled over the arms and the visits. A rate exposes about as much as one visit's share, so a study's attendance rates are about as exposed as a table with one visit for each endpoint: `r attendance_auc(5, 10)` for five endpoints behind 10 patients.

An AUC averages over members, and the most exposed member is the one far from the typical values: that patient moves the release furthest toward themselves, which is what protection 8 reads. The membership-inference literature on machine-learning models reports success against such patients separately for this reason \[14, 15\], and Stadler and colleagues found that synthetic data from generative models leaves them exposed \[16\].

The attacker modelled here is strong. The measurements it needs, a candidate's random effects estimated from their concentrations or their outcome at every visit, are made by the study, and outside it exist only where the same quantities are measured for the same patient: at a study site, in a related study, or in routine care such as drug-level monitoring.

## Outstanding Risks

These are the ways a release from this generator can still disclose something about a patient, most serious first. Each is inherent to releasing estimates without noise, or a limit of the checks.

- **An attacker who holds a candidate's own trial data can infer membership.** [Membership Inference Against the Release](#membership-inference-against-the-release) measures how well. The release holds no per-visit table, so what remains is chiefly the population model of a small cohort and one attendance rate per endpoint, which only noise would reduce.
- **An attacker who knows every other patient is not stopped.** Such an attacker can compute the exact effect of the one patient they do not know, so any move at all is detectable, however far under a threshold it sits. That attacker is the one DP is defined against, and only a DP mechanism stops them.
- **Releases compose, and their counts are exact.** The cohort size, the arm sizes and every frequency are exact counts or exact ratios of counts. Two releases from overlapping data, such as an interim and a final analysis, or the same study before and after one patient withdraws, differ by exactly what the patients between them contributed. Nothing here accounts for that.
- **Three patients is the smallest threshold in common use.** Every released frequency rests on at least three patients on each side, which is the floor the threshold rule of SDC starts from. A release held to a stricter rule needs `min_arm_patients` and `min_category_patients` raised to match.
- **The influence reading describes this study only, and its thresholds are empirical.** How far one patient moves an estimate is itself a fact about the data, which is why DP calibrates its noise to *smooth* sensitivity rather than local sensitivity \[17\]. The review and failure thresholds were set against the public studies above, so that ordinary patients pass and a gross data error fails; they are not derived from a privacy target. Shrinkage makes the reading understate the move, and where shrinkage is close to 1 the reading says little while its verdict still reads pass.
- **Not every estimate is read.** The residual error, a parameter estimated without a between-subject term (bioavailability in a mixed-route model), the PD shape parameters other than the baseline, and a planned dose taken as the arm's mean where no amount is shared have no row.
- **Leaving out stops at a tenth of the cohort.** A study with more patients than that whom the influence reading flags keeps the rest, and P5 then reads a review or a failure.
- **The full fit holds diagnostics the release does not.** The Akaike information criterion (AIC) of every candidate model, the starting values and the correlations between covariates and the individual random effects stay on the object `synpmx_model_estimate()` returns. Share the release or the synthetic data, never the fit.
- **The estimation output names patients.** A patient left out of the estimates, and a patient who fails the influence reading, is named on the console, so logs and rendered reports from the environment that holds the study carry that identifier.
- **Rarity in the world is not measured.** A covariate level held by many patients in this study can still identify someone if few people alive hold it.

## Released Parameters at a Glance {#released-parameters-at-a-glance}

Every number in a release that was estimated from patients, counted once, for the stored fit of each public study. A rate pooled over the arms is listed in the release once per arm and counts once here, and the planned dose schedule, which is the protocol, is counted apart.

```{r parameters, eval=run}
# An arm's dosing model, or one per drug where the study declared which doses
# drive which endpoint.
per_drug <- function(entry) if (!is.null(entry$planned)) list(entry) else entry
parameter_count <- function(release) {
  arms <- names(release$arms$sizes)
  shape <- function(s) length(s$typical) + 2
  rates <- unique(unlist(lapply(release$dosing, function(entry) {
    lapply(per_drug(entry), function(d) {
      paste(d$reduction, d$interruption, d$discontinuation,
            paste(d$levels, collapse = "/"))
    })
  })))
  # One rate per endpoint, listed at each of its visits in each arm.
  attendance <- unique(unlist(lapply(release$visits, function(v) {
    p <- as.numeric(v$probability)
    paste(release$cells$endpoint, p)[p > 0 & p < 1]
  })))
  marginals <- unique(Filter(Negate(is.null),
                             unlist(release$discrete, recursive = FALSE)))
  counts <- c(
    `PK model` = sum(vapply(release$pk_models, function(m) {
      length(m$parameters$fixed) + nrow(m$parameters$omega) + 1
    }, numeric(1))),
    `PD time courses` = sum(vapply(release$pd, function(s) {
      if (length(s$arms)) sum(vapply(s$arms, shape, numeric(1))) else shape(s)
    }, numeric(1))),
    covariates = sum(vapply(release$covariates, function(c) {
      switch(c$kind, lognormal = 2, normal = 2,
             categorical = length(c$levels) - 1, 0)
    }, numeric(1))),
    `dose changes` = sum(vapply(strsplit(rates, " "), function(r) {
      3 + length(strsplit(r[[4]], "/")[[1]]) - 1
    }, numeric(1))),
    attendance = length(attendance),
    `discrete endpoints` = sum(vapply(marginals, function(m) {
      max(0, length(m$levels) - 1)
    }, numeric(1))),
    `counts and floors` = length(arms) + 1 +
      length(release$quantification_floor) +
      length(unique(unlist(lapply(release$covariate_effects,
                                  function(e) e$reference)))))
  c(counts, total = sum(counts),
    `planned cycles (protocol)` = sum(vapply(release$dosing, function(entry) {
      sum(vapply(per_drug(entry), function(d) nrow(d$planned), numeric(1)))
    }, numeric(1))))
}
studies <- c(warfarin = "warfarin-model-fit.rds",
             theo_md = "theo-md-model-fit.rds",
             mad = "mad-model-fit.rds",
             case1_pkpd = "case1-pkpd-model-fit.rds",
             wbcSim = "wbcsim-model-fit.rds",
             mavoglurant = "mavoglurant-model-fit.rds",
             nimoData = "nimo-model-fit.rds",
             pheno_sd = "pheno-model-fit.rds",
             mixroute_sim = "mixroute-sim-model-fit.rds",
             onc_sim = "onc-sim-model-fit.rds")
counted <- do.call(rbind, lapply(names(studies), function(study) {
  release <- model_release(stored_fit(studies[[study]]))
  data.frame(study = study, patients = release$n_source,
             t(parameter_count(release)), check.names = FALSE)
}))
knitr::kable(counted, row.names = FALSE)
```

What each column holds:

- **PK model**: a typical value and a between-subject variance for each PK parameter, and one residual error: 11 numbers for the five parameters of a two-compartment oral model. A parameter fitted without between-subject variability, such as a bioavailability, adds its typical value alone.
- **PD time courses**: for each other continuous endpoint, one to three shape parameters, the spread of the patients' baselines, and a residual error.
- **Covariates**: a trimmed mean and SD for each continuous covariate, and one fewer frequency than levels for each categorical one.
- **Dose changes**: the three pooled rates of reducing, skipping and stopping, and each level of the dose ladder below the full dose, a fraction of the starting dose to one significant figure.
- **Attendance**: one rate per endpoint, and none for an endpoint whose rate is 1, as where every patient was observed at every visit of it their arm had.
- **Discrete endpoints**: one fewer level frequency than levels, for each binary or ordinal endpoint.
- **Counts and floors**: the size of each arm and of the cohort, an emission floor for each endpoint without a declared assay limit, and the reference weight where body-weight scaling was requested.
- **Planned cycles**: each arm's planned dose times and amounts, which are the protocol rather than estimates. The nominal visit grid and the description of the table to generate are the protocol too, and are not counted.

## Appendix: The Noise a DP Guarantee Needs in a Small Study {#dp-noise-small-study}

Two routes give a population model a DP guarantee. The per-patient route estimates each patient's parameters from that patient's own measurements, clips them to a public range and releases their mean with noise, which is how `synpmx_calibrated()` releases its one number. Subsample-and-aggregate splits the patients into disjoint groups of about twenty, fits the model in each, and releases the mean of the groups' clipped estimates with noise \[17, 21\].

With $\varepsilon = 1$ split evenly over *d* released numbers, each clipped to a public range spanning sixteenfold on the log scale (fourfold either side of a prior, as `synpmx_calibrated()` uses), each number gets Laplace noise of scale $b = dS/(n\varepsilon)$ on the log scale by the per-patient route, and $b = dS/(k\varepsilon)$ with $k = n/20$ groups by subsample-and-aggregate. The table gives the median fold error of one released number, $e^{b \ln 2}$, and the noise's spread against the uncertainty an estimate already has from sampling, for a between-subject SD of 0.3:

```{r dp-noise}
span <- log(16)
dp_scale <- function(patients, released) released * span / patients
dp_fold <- function(patients, released) {
  exp(log(2) * dp_scale(patients, released))
}
dp_over_sampling <- function(patients, released) {
  sqrt(2) * dp_scale(patients, released) / (0.3 / sqrt(patients))
}
grid <- expand.grid(patients = c(30, 60, 200, 1000), released = c(11, 30))
fold_text <- function(x) {
  ifelse(x > 100, "over 100-fold", sprintf("%.2f-fold", x))
}
knitr::kable(data.frame(
  patients = grid$patients,
  released_numbers = grid$released,
  per_patient_median_error = fold_text(dp_fold(grid$patients, grid$released)),
  noise_over_sampling_error = sprintf("%.0f times", dp_over_sampling(
    grid$patients, grid$released)),
  subsample_median_error = fold_text(dp_fold(grid$patients / 20,
                                             grid$released))))
```

Eleven numbers are a two-compartment population model alone, and thirty a whole release. At 60 patients the per-patient route leaves half of the population model's numbers off by more than `r sprintf("%.2f-fold", dp_fold(60, 11))`, with noise `r sprintf("%.0f", dp_over_sampling(60, 11))` times the sampling error each estimate already carries, and a whole release off by `r sprintf("%.1f-fold", dp_fold(60, 30))`. At 30 patients the population model alone is off by `r sprintf("%.1f-fold", dp_fold(30, 11))`. Even at 1,000 patients the noise is `r sprintf("%.1f", dp_over_sampling(1000, 11))` times the sampling error. Subsample-and-aggregate, which keeps the fit as it is, is unusable below about a thousand patients. Gaussian noise with tighter accounting grows with the square root of the count rather than the count, and does not change this picture at these cohort sizes. At the 12 to 200 patients this generator sees, a DP release would say less about the study than a public prior does. Neither route is implemented beyond the one number `synpmx_calibrated()` releases.

## References

1.  Dwork C, McSherry F, Nissim K, Smith A. Calibrating noise to sensitivity in private data analysis. *Theory of Cryptography Conference (TCC).* 2006.
2.  Homer N, Szelinger S, Redman M, et al. Resolving individuals contributing trace amounts of DNA to highly complex mixtures using high-density SNP genotyping microarrays. *PLoS Genetics.* 2008;4(8):e1000167.
3.  Sankararaman S, Obozinski G, Jordan MI, Halperin E. Genomic privacy and limits of individual detection in a pool. *Nature Genetics.* 2009;41:965–967.
4.  Dwork C, Smith A, Steinke T, Ullman J, Vadhan S. Robust traceability from trace amounts. *IEEE Symposium on Foundations of Computer Science (FOCS).*
    2015. 
5.  Fredrikson M, Lantz E, Jha S, Lin S, Page D, Ristenpart T. Privacy in pharmacogenetics: an end-to-end case study of personalized warfarin dosing. *USENIX Security Symposium.* 2014.
6.  Dinur I, Nissim K. Revealing information while preserving privacy. *ACM Symposium on Principles of Database Systems (PODS).* 2003.
7.  Novartis. `synadam`: generate synthetic ADaM datasets. R package version 0.3.4. <https://github.com/Novartis/synadam>.
8.  Hundepool A, Domingo-Ferrer J, Franconi L, et al. *Statistical Disclosure Control.* Wiley; 2012.
9.  Huber PJ, Ronchetti EM. *Robust Statistics.* 2nd ed. Wiley; 2009.
10. Bun M, Steinke T. Average-case averages: private algorithms for smooth sensitivity and mean estimation. *Advances in Neural Information Processing Systems (NeurIPS).* 2019.
11. Cook RD. Detection of influential observation in linear regression. *Technometrics.* 1977;19(1):15–18.
12. Carlini N, Jagielski M, Zhang C, Papernot N, Terzis A, Tramèr F. The privacy onion effect: memorization is relative. *Advances in Neural Information Processing Systems (NeurIPS).* 2022.
13. Dwork C, Lei J. Differential privacy and robust statistics. *ACM Symposium on Theory of Computing (STOC).* 2009.
14. Shokri R, Stronati M, Song C, Shmatikov V. Membership inference attacks against machine learning models. *IEEE Symposium on Security and Privacy.*
    2017. 
15. Carlini N, Chien S, Nasr M, Song S, Terzis A, Tramèr F. Membership inference attacks from first principles. *IEEE Symposium on Security and Privacy.* 2022.
16. Stadler T, Oprisanu B, Troncoso C. Synthetic data – anonymisation groundhog day. *USENIX Security Symposium.* 2022.
17. Nissim K, Raskhodnikova S, Smith A. Smooth sensitivity and sampling in private data analysis. *ACM Symposium on Theory of Computing (STOC).* 2007.
18. U.S. Food and Drug Administration. *Population Pharmacokinetics: Guidance for Industry.* 2022.
19. European Medicines Agency. *Guideline on Reporting the Results of Population Pharmacokinetic Analyses.* CHMP/EWP/185990/06. 2007.
20. Schulz KF, Altman DG, Moher D. CONSORT 2010 Statement: updated guidelines for reporting parallel group randomised trials. *BMJ.* 2010;340:c332.
21. Smith A. Privacy-preserving statistical estimation with optimal convergence rates. *ACM Symposium on Theory of Computing (STOC).* 2011.
22. Chaudhuri K, Monteleoni C, Sarwate AD. Differentially private empirical risk minimization. *Journal of Machine Learning Research.* 2011;12:1069–1109.