---
title: "Privacy Protections in the PMX Model Generator"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Privacy Protections in the PMX Model Generator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
# Every fit is read from the installed package, so this article fits nothing
# and needs neither nlmixr2 nor a compiler.
stored_fit <- function(file) {
  path <- system.file("extdata", file, package = "synpmx")
  if (nzchar(path)) readRDS(path) else NULL
}
fit <- stored_fit("warfarin-model-fit.rds")
misdosed_fit <- stored_fit("warfarin-misdosed-model-fit.rds")
misdosed_kept_fit <- stored_fit("warfarin-misdosed-kept-model-fit.rds")
extreme_fit <- stored_fit("warfarin-extreme-model-fit.rds")
run <- has_nlmixr2data && !is.null(fit) && !is.null(misdosed_fit) &&
  !is.null(misdosed_kept_fit) && !is.null(extreme_fit)
plots <- run && has_ggplot2
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")

# Three significant digits on measured quantities, written out rather than
# with a shared exponent, since one column can hold a clearance beside a
# variance a thousand times smaller. Counts are exempt.
digits3 <- function(x) {
  x <- as.data.frame(x)
  attr(x, "who") <- NULL
  for (column in names(x)) {
    if (is.numeric(x[[column]]) && !column %in% c("patients")) {
      x[[column]] <- vapply(x[[column]], function(v) {
        format(signif(v, 3), scientific = FALSE, drop0trailing = TRUE,
               trim = TRUE)
      }, character(1))
    }
  }
  x
}
```

```{r membership-simulation, include=FALSE, file="pmxmodel-privacy-simulation.R"}
```

```{r membership-tables, include=FALSE}
# Computed by the simulation article's code, from the same seed, so the numbers
# quoted here are the ones that article explains.
tables <- membership_tables()
population_table <- tables$population
attendance_table <- tables$attendance
rounding_change <- sprintf("%.3f", max(abs(population_table$unrounded -
                                             population_table$released)))
floor_change <- sprintf("%.2f", max(abs(attendance_table$without_floor -
                                          attendance_table$released)))
population_auc <- function(patients, off_by) {
  sprintf("%.2f", population_table$released[
    population_table$patients == patients & population_table$off_by == off_by])
}
attendance_auc <- function(visits, patients, correlation = 0) {
  sprintf("%.2f", attendance_table$released[
    attendance_table$visits == visits & attendance_table$patients == patients &
      attendance_table$correlation == correlation])
}
```

The pharmacometric (PMX) model generator, `synpmx_model()`, fits a population
pharmacokinetic (PK) and pharmacodynamic (PD) model to a clinical study inside
the environment that holds the study's data, and generates synthetic patients
from the fit. This article sets out what of the study leaves that environment,
how each number that leaves is protected, how the protections are checked, and
what risk remains. The first section summarizes all four for a privacy
reviewer, and the sections after it give the evidence behind each statement in
it.

**No formal privacy guarantee is made.** Nothing here bounds what an attacker
can learn in the sense of differential privacy (DP), a property of a release
procedure that limits how much any one person's data can change what is
released [1]. The protections below reduce what is released and measure how
much one patient moves it. A release that has to withstand a determined
attacker needs a DP mechanism, and the last section says what one would take
for this generator.

## Summary for a Privacy Reviewer

### What the Generator Releases

A population PK/PD model describes how the concentration of a drug in the
blood, and a response to the drug, change over time across a population of
patients. It is a short list of numbers: a **typical value** for each
parameter, such as how fast the body clears the drug; a **between-subject
variance** for each, which says how far patients spread around the typical
value; and a **residual error**, the scatter of single measurements around a
patient's own curve. `synpmx_model()` estimates these from the study's patients
with `nlmixr2`, an R package for fitting population models.

The generator also summarizes the study's design as it was carried out: the
number of patients in each treatment group (**arm**), each arm's dose schedule,
how often doses were reduced, skipped or stopped, the share of scheduled visits
at which patients were measured, the distribution of each baseline covariate
such as body weight, age or sex, and, for an endpoint recorded in levels such
as a yes/no response or a toxicity grade, how often each level was recorded.
Everything in that list read from how patients behaved, rather than from the
protocol, is one set of numbers for the whole study, pooled over the arms and
the visits, as the PD time courses are pooled over the arms.

What leaves the environment that holds the study is the **release**: these
numbers, assembled by `model_release()` from a list of permitted fields,
together with a description of the table to generate. Synthetic data is
simulated from the release and from random numbers alone, so it discloses
nothing the release does not, and the release travels with the synthetic data
as an attribute. A reviewer can read every number in a release with
`str(unclass(release))`, and
[Released Parameters at a Glance](#released-parameters-at-a-glance) counts them
for each public study.

### What Is Protected, and What Is Not

**✅ No patient record, identifier or per-patient estimate leaves the study.**
The release holds numbers about the population and about arms. Synthetic
subject identifiers are new. Each patient's own estimated deviation from the
typical values is used during estimation and then discarded.

**✅ No released number is one patient's value, and every released frequency
rests on at least three patients on each side.** Minima, maxima, medians and
modes are replaced by quantities no one patient decides. A frequency, a
category level, or a value copied onto every patient of an arm is released only
where the patients on each side of it number zero or at least three.

**✅ A patient who moves a released estimate by 15% or more is left out of the
estimates.** Estimation reads how far each patient moves each estimate, leaves
out anyone who moves one by 15% or more (for a between-subject spread, by 0.15
in its standard deviation (SD) on the log scale), and estimates again, up to a
tenth of the cohort.
A patient left out still counts in the arm sizes, the dosing and the visit
attendance. The reading is approximate and does not cover every estimate; its
limits are among the outstanding risks.

**⚠️ Membership inference remains possible for an attacker who holds a
candidate's own trial data.** In simulation, an attacker with a candidate's own
measurements and exact reference values for the population ranks a study member
above a non-member with probability `r population_auc(12, 0)` when the
population model is fitted to 12 patients and `r population_auc(60, 0)` when it
is fitted to 60. This probability is the area under the curve (AUC), where 0.5
is chance and 1 is certainty. A reference off by 20% brings the two to
`r population_auc(12, 0.2)` and `r population_auc(60, 0.2)`. The release holds
no table per visit, because one would be more exposed: a share of patients
sampled at each of 20 visits gives `r attendance_auc(20, 10)` when 10 patients
stand behind each share and `r attendance_auc(20, 100)` when 100 do.
Attendance is one rate for the study instead, and each binary or ordinal
endpoint one set of level frequencies. Rounding moves these numbers by at most
`r rounding_change` and the three-patient floor by at most `r floor_change`;
cohort size and the number of released numbers decide them.
[Membership inference against the PMX model release](https://iamstein.github.io/synpmx/articles/pmxmodel-privacy-simulation.html)
shows how each number is computed.

**❌ No formal guarantee.** Nothing bounds what an attacker who knows every
other patient learns about the remaining one. That attacker is the one DP is
defined against, and no DP mechanism is implemented for this generator.

### How to Share a Release

**Share the release or the synthetic data, never the fit.** The object
`synpmx_model_estimate()` returns also holds diagnostics: the comparison of
candidate models, the starting values, and the correlations between covariates
and each patient's random effects. Save `model_release(fit)`.

**Keep the estimation's console output and rendered reports inside the study
environment.** A patient left out of the estimates, or one who moves an estimate
far, is named there by their study identifier.

**Use a method with a formal guarantee where membership in the study is itself
sensitive** and a recipient could hold a candidate's measurements of the kind
the study made: a site holding the patient's records, a partner holding a
related study, or routine care that measures the same quantities, such as
drug-level monitoring.
[What differential privacy does and does not guarantee](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html)
sets out that choice.

## How Summary Statistics Disclose Patients

A fitted model holds no patient rows, but the DP literature shows that
statistics computed from patients can still disclose them. Four mechanisms
apply here.

**Membership inference** decides whether a given person was in the study. The
attacker holds that person's data and the release, and asks whether the release
fits them better than it would fit someone who was not in the study. Homer and
colleagues did this from published allele frequencies alone [2], after which
public access to that kind of aggregate genetic data was withdrawn. The power
of the attack grows with the number of released statistics and with their
precision, and falls as the number of people behind each statistic grows
[3, 4].

**Model inversion** infers a hidden attribute from an estimated relationship.
Fredrikson and colleagues took a published warfarin dosing regression, a
patient's demographics and that patient's stable dose, and inferred the
patient's genotype nearly as well as a model built to predict genotype [5].

**Reconstruction** recovers individual records from enough accurate aggregate
answers: answering too many questions too precisely gives the dataset away
[6].

**A single patient's value** is the simplest case. A minimum, a maximum, the
median of an odd number of patients, and the most common value where every
value is held once are each one patient's value, whatever the summary is
called. Releasing one releases that patient's value.

What an attacker already holds decides which of these succeed:

| Attacker | Holds | What the release gives them |
|---|---|---|
| An outsider | a candidate's demographics: age, sex, body weight | the covariates as trimmed, rounded means and SDs, and as level frequencies held by at least three patients: the kind of summary a trial's table of baseline characteristics reports |
| An attacker with trial data | a candidate's own measurements of the kind the study made, and reference values for the population | an advantage at membership inference, measured under [Membership Inference Against the Release](#membership-inference-against-the-release) |
| An attacker who knows every other patient | the whole study except one patient | that patient's contribution to every released number, which only a DP mechanism hides |

Three properties of a summary lower all four risks: fewer released numbers,
coarser released numbers, and numbers no single patient can move far. Each
protection below is one of these.

## The Release, Field by Field

A fitted model carries two kinds of content. Generation reads one of them. The
other is diagnostics for whoever ran the fit, and stays with it.
`model_release()` builds the part generation reads, and
`synpmx_model_generate()` attaches that part, never the fit, to the data it
returns. Generating from a fit and from its release gives the same data. The
worked example is `nlmixr2data::warfarin`: 32 patients, one oral dose, a
warfarin concentration and a prothrombin activity.

The diagnostics that stay with the fit:

```{r release, eval=run}
release <- model_release(fit)
setdiff(names(fit), names(release))
```

Every field of the release, what it holds, and how many finite numbers it
holds. `parameters` repeats the first entry of `pk_models` and is left out.

```{r count, eval=run, echo=FALSE}
count_numbers <- function(x) {
  if (is.null(x) || is.function(x) || is.environment(x)) return(0L)
  if (is.data.frame(x)) {
    return(sum(vapply(x, function(column) {
      if (is.numeric(column)) sum(is.finite(column)) else 0L
    }, integer(1))))
  }
  if (is.list(x)) return(sum(vapply(x, count_numbers, integer(1))))
  if (is.numeric(x)) sum(is.finite(x)) else 0L
}
holds <- c(
  structural = "the structural PK model selected",
  pk_models = paste("per concentration endpoint: typical values,",
                    "between-subject variances, residual error, any",
                    "body-weight scaling"),
  pd = paste("per other continuous endpoint: a time-course shape and its",
             "parameters, the spread of patients' baselines, residual error"),
  arms = "the arms and the number of patients in each",
  dosing = paste("per arm: the planned dose at each scheduled time; pooled",
                 "over arms: the levels doses were reduced to, and the rates",
                 "of reducing, skipping and stopping"),
  visits = paste("the visits each arm has, and one attendance rate for the",
                 "study, listed at each of them"),
  cells = "the scheduled visits kept, as endpoint and nominal time",
  discrete = paste("per binary or ordinal endpoint: one set of level",
                   "frequencies for the study, listed at each visit"),
  covariates = paste("per baseline covariate: a trimmed mean and SD, or level",
                     "frequencies"),
  covariate_effects = "body-weight scaling, where requested: a reference weight",
  schema = paste("the table to generate: columns and their types, compartment",
                 "numbers, assay limits, values copied onto each arm, each",
                 "endpoint's value type"),
  roles = "which column holds what",
  endpoints = "which endpoints are concentrations, time courses or levels",
  quantification_floor = paste("per endpoint without a declared assay limit,",
                               "the value nothing is generated below"),
  n_source = "the number of patients",
  settings = "the frequency floors used",
  privacy = "the verdicts of checks P4 and P5"
)
fields <- setdiff(names(release), "parameters")
knitr::kable(data.frame(
  field = fields,
  holds = ifelse(fields %in% names(holds), holds[fields], "(undescribed)"),
  numbers = vapply(unclass(release)[fields], count_numbers, integer(1)),
  row.names = NULL))
```

Membership inference gains power with the number of released numbers [3, 4].
Most of the count is the visit and dosing apparatus rather than the population
model, which is in `pk_models` and `pd`. Part of the apparatus is the protocol
rather than a measurement: the nominal visit times in `cells` and the planned
dose times in `dosing` are public in the sense the protocol is.

## The Protections

Each protection is applied by `synpmx_model_estimate()` without being asked for,
and each names the work it rests on.

### 1. An Allowlist Decides What Leaves

`model_release()` copies the fields Step 6 of the
[algorithm](https://iamstein.github.io/synpmx/articles/pmxmodel-algorithm.html)
reads. Inside the PK models, the PD shapes, the dosing models, the endpoint
list and the settings, a second list names the parts it copies; the schema, the
arm, visit and covariate models and the remaining fields are copied whole. A
field added to the fit later stays behind unless it is added to a list. The
pattern is the glimpse-then-simulate split of Novartis's `synadam` package [7],
which writes the summary that synthetic data is generated from to one file a
reviewer can read.

### 2. No Per-Patient Quantity

The empirical Bayes estimates (EBEs) are each patient's own random effects: how
far that patient's parameters sit from the typical values, estimated from the
patient's own measurements. They are read twice at estimation, for the
correlations between covariates and random effects and for the influence
reading of protection 8, and then dropped. Generation draws new random effects
from the between-subject covariance matrix.

### 3. No Identifier

Synthetic subject identifiers (IDs) are minted at generation. No source label is
kept, including the levels of an ID column stored as a factor. A numeric ID
keeps only the largest source ID rounded up to a power of ten, which stops a
synthetic ID colliding with a real one without storing the real one. `synadam`
refuses to run when an ID column is left undeclared, for the same reason [7].

### 4. No Single Patient's Value

Every summary that could have been one patient's value is replaced:

| Quantity | Would have been | Is |
|---|---|---|
| Emission floor, where no assay limit is declared | half the smallest value reported | half the lowest value at least three patients reached, rounded down to 1, 2 or 5 times a power of ten |
| Allometric reference weight | the cohort median | the median to one significant figure |
| Planned dose amount or infusion rate, where no value is shared | the most common value, which is then one patient's | the arm's mean to two significant figures |
| Continuous covariate | a mean and SD over every patient, and a median | a trimmed mean and SD (protection 6), no median |
| Endpoint type description | text quoting the count of levels and the lowest and highest value observed | the type, without the text |

Statistical disclosure control (SDC), the field that protects published
statistical tables, treats extreme values the same way, by top-coding and
coarsening them [8].

### 5. Several Patients Behind Every Frequency

A frequency resting on one patient is that patient. The threshold rule of SDC
publishes a count only where at least *k* units stand behind it [8], and
`synadam` masks values held once [7]. The rule applies to both sides of a
frequency: attendance at a visit that all but one patient kept says as much
about the one who missed it as attendance by one patient says about the one who
came. Here *k* is 3 by default:

- arms below `min_arm_patients` are dropped before anything is fitted;
- a nominal visit slot, a dose level and a planned cycle are kept only where at
  least `min_arm_patients` patients reached them, and an arm has a visit only
  where at least that many of its patients were observed there;
- attendance is one rate for the study, and is 1 where fewer than
  `min_arm_patients` patients missed any visit;
- a dose-change rate, pooled over every arm, whose events fewer than
  `min_arm_patients` patients had is zero;
- a categorical covariate level, and a binary or ordinal endpoint's level,
  below `min_category_patients` patients is excluded, or folded into the most
  common level;
- a `keep` value of any type, copied from the arm's first patient as `keep`
  declares, is written as missing where fewer than `min_category_patients`
  patients in the arm hold it;
- the description of the table to generate keeps, for a factor column, only the
  levels its arms carry or at least `min_category_patients` patients hold, and
  for a binary or ordinal endpoint only the levels that many patients recorded,
  so the label of an arm too small to keep, or a grade one patient reached,
  stays behind.

The rule stops a frequency from singling out one or two patients. It does not
stop membership inference, which adds up small signals over many frequencies
([Membership Inference Against the Release](#membership-inference-against-the-release)).
Pooling is what weakens that: attendance is one rate, the dose-change rates are
three, and a binary or ordinal endpoint is one set of level frequencies, each
resting on every patient rather than on one arm at one visit. `pd_by_arm =
TRUE` reads the endpoint frequencies per arm again, with the PD time courses.

### 6. Covariate Summaries Without Their Extremes

A continuous covariate is summarized by a mean and SD taken after the highest
and lowest 5% of patients are set aside, at least one at each end. The SD of
what remains understates the spread, because the tails are what was removed,
so it is divided by the SD a standard normal keeps under the same trimming.
With a share $\alpha$ removed from each end and $z = \Phi^{-1}(1-\alpha)$,

$$\mathrm{SD} = \frac{\mathrm{SD}_\text{trimmed}}{\sqrt{1 - \dfrac{2 z \phi(z)}{1-2\alpha}}}.$$

This is the trimmed mean of robust statistics [9]. Trimming is also how DP
mean estimators bound one person's influence [10], since no single value can
then move the result by more than the trimmed range allows.

### 7. Two Significant Figures

Every released estimate is rounded to two significant figures: PK fixed
effects, between-subject variances, residual error, PD shape parameters and
covariate summaries. Rounding is the other standard coarsening of SDC [8].
Leaving out a typical patient moves a typical value by less than one rounding
step, so for most patients the released typical value is the same with or
without them. Rounding does not hide an outlier's effect on a variance, which
protection 8 measures, and it does not weaken membership inference, because an
estimate's sampling variation is larger than one rounding step
([Membership Inference Against the Release](#membership-inference-against-the-release)).
The frequencies are not rounded: each is an exact count over an arm, held to the
floor of protection 5 instead.

### 8. How Far One Patient Moves Each Estimate

DP calls the most one person can move a released number its *sensitivity*,
taken over every possible dataset [1]. The same question asked of the study at
hand, leaving each patient out in turn and recording the largest move, is
*local sensitivity*, and in regression the same idea is Cook's distance [11].
It is computed at estimation without refitting:

- A population typical value on the log scale sits close to the mean of the
  patients' individual log parameters, so leaving out patient $i$ moves it by
  about $\eta_i/(n-1)$, where $\eta_i$ is that patient's random effect. The PD
  typical baseline is read the same way, from each patient's own baseline.
- The between-subject variance sits close to the mean of the squared random
  effects, so leaving out patient $i$ scales it by
  $\frac{(S-\eta_i^2)/(n-1)}{S/n}$, with $S=\sum_j \eta_j^2$.
- The spread of the PD baselines and each covariate summary are recomputed with
  the patient left out, which needs no approximation.

The largest move over all patients is kept for each estimate: in percent for a
typical value, and in points of the between-subject SD on the log scale, close
to a coefficient of variation (CV) in percent, for a spread. A move of 15 is a
review; a move of 30 fails, warns at estimation, and names the patient on the
console. The name is printed in the environment that holds the study and is
not stored: the fit keeps the largest move per estimate, never anything per
patient. On a parameter whose between-subject SD is 0.27, a patient 3 SD from
the mean in a cohort of 32 moves the typical value by about 3% and the spread
by about 4 points. Shrinkage, the pull of a random effect toward zero when the
patient's own measurements say little about their parameters, makes the
reading understate the move, so it is reported beside it.

A patient who moves an estimate by 15 or more is then left out of the
estimates, and estimation runs again. The patient is left out of the PK fit,
the PD fits and the covariate summaries, and stays in the dosing, visit and arm
models, so the cohort and arm sizes the release describes do not change.
Rounds repeat until the reading flags nobody, and stop before a tenth of the
cohort has been left out. Each patient left out is named on the console; the
fit records how many, and which estimates they moved. `drop_influential =
FALSE` keeps every patient and leaves the reading as a review or a failure.

The rounds are repeated because leaving out the most influential patients makes
the next ones the most influential, which Carlini and colleagues call the
privacy onion effect [12]. The DP counterpart of the reading is
propose-test-release, which checks that the study at hand is far from any study
in which one patient could move an estimate far, and only then releases it with
noise scaled to that bound [13].

## The Five Checks

`model_privacy_checks()` runs five checks on a fit, each with its pass
criterion. The first four ask whether the release holds something about one
patient that it should not. The fifth is the reading of protection 8.

| Check | Question | Passes when |
|---|---|---|
| P1 | No per-patient table is released | the release holds only fields on the allowlist, none of them per patient |
| P2 | No source identifier is released | no source ID label, and no source ID value as the synthetic ID offset |
| P3 | No single patient's value is released | floors on the 1-2-5 series, the reference weight rounded, no covariate median, no factor level that no arm carries |
| P4 | Every released frequency rests on several patients, on both sides | each side of every frequency is none or at least `min_arm_patients` patients (`min_category_patients` for categorical levels and values) |
| P5 | No single patient moves a released estimate far | every estimate moves less than 15, after any patient who moved one further has been left out |

P1 to P3 are structural and read the release. They pass by construction on a
fit from this version, and exist to catch a release assembled some other way
or a fit stored before the protections existed. P3 reads the emission floors,
the covariate summaries, the reference weight and the factor levels of the
columns copied onto arms. P4 recounts, at estimation, the smallest group of
patients behind each kind of released frequency after the rules of protection
5 have run, including every factor and endpoint level the schema stores, and P5
is the reading of protection 8. A release carries the verdicts of P4 and P5,
and the fit the counts behind them.

```{r checks, eval=run}
checks <- model_privacy_checks(fit)
as.data.frame(checks)[, c("check", "verdict", "result")]
```

The reading behind P5 is the `influence` attribute of the full fit's checks,
one row per released estimate: the released value, the largest move any one
patient causes, its unit, the number of patients it was read over, the
shrinkage of the random effect it came from, and the verdict. No row is about
a patient.

```{r influence, eval=run}
digits3(attr(checks, "influence"))
```

The recount behind P4 is the `frequencies` attribute: for each kind of
released frequency, the smallest group of patients behind any one of them, the
floor it is held to, and how many values the rules changed to meet it.

```{r frequencies, eval=run}
attr(checks, "frequencies")[, c("quantity", "smallest", "threshold",
                                "adjusted")]
```

The same checks on the release report P5 as a verdict only. A passing release
states that every estimate moved less than 15, and nothing more, so the one
data-dependent thing it carries about the reading is the verdict. A release
that does not pass names its worst estimate and the size of the move, which is
computed from one patient's data; such a release is not meant to leave.

```{r release-checks, eval=run}
as.data.frame(model_privacy_checks(release))[, c("check", "verdict", "result")]
```

## Three Constructed Cases

Each case is `warfarin` with one or more patients changed, built once by
`synpmx_model_estimate()` and stored with the package. None is a real study.

```{r constructed-data, eval=run}
warfarin <- as.data.frame(nlmixr2data::warfarin)
warfarin$ntime <- warfarin$time
roles <- pmx_roles(
  id = "id", time = "time", nominal_time = "ntime", dv = "dv", amt = "amt",
  evid = "evid", dvid = "dvid", covariates = c("wt", "age", "sex")
)
patients <- sort(unique(warfarin$id))

misdosed <- warfarin
on_cp <- misdosed$id == patients[[5L]] & misdosed$evid == 0 &
  misdosed$dvid == "cp"
misdosed$dv[on_cp] <- misdosed$dv[on_cp] * 1000

extreme <- warfarin
extreme$wt[extreme$id == patients[[3L]]] <- 250
extreme$age[extreme$id == patients[[8L]]] <- 95
extreme$sex <- as.character(extreme$sex)
extreme$sex[extreme$id == patients[[12L]]] <- "not recorded"
extreme$sex <- factor(extreme$sex)
```

```{r constructed-fits, eval=FALSE}
misdosed_fit <- synpmx_model_estimate(misdosed, roles, seed = 1)
extreme_fit <- synpmx_model_estimate(extreme, roles, seed = 1)
```

### A Patient Given a Thousand Times the Recorded Dose

One patient received a thousand times the dose the record shows, so a linear
model sees concentrations a thousand times higher than the recorded dose
implies. The case is fitted twice: as estimation runs by default, and with
`drop_influential = FALSE`, which keeps the patient so that the reading can be
seen to catch them.

```{r misdosed-fits, eval=FALSE}
misdosed_fit <- synpmx_model_estimate(misdosed, roles, seed = 1)
misdosed_kept_fit <- synpmx_model_estimate(misdosed, roles, seed = 1,
                                           drop_influential = FALSE)
```

Kept, the patient fails P5. The variance terms absorb the excess exposure, and
estimation warned that one patient moved a released estimate by 30 or more,
naming the patient on the console. The stored fit names nobody.

```{r misdosed-kept, eval=run}
kept_checks <- model_privacy_checks(misdosed_kept_fit)
as.data.frame(kept_checks)[, c("check", "verdict", "result")]
digits3(attr(kept_checks, "influence"))
```

By default, estimation left the patient out of the estimates and ran again.
The fit records how many patients were left out and which estimates they
moved, and the checks pass:

```{r misdosed-default, eval=run}
misdosed_fit$privacy$left_out$patients
digits3(misdosed_fit$privacy$left_out$because)
as.data.frame(model_privacy_checks(misdosed_fit))[, c("check", "verdict",
                                                      "result")]
```

The synthetic data shows the difference. Kept, the patient's exposure becomes
between-subject variability, and the synthetic patients spread wider than the
rest of the source cohort. Left out, they do not:

```{r misdosed-plot, eval=plots, fig.height=4}
concentration <- function(data, label) {
  rows <- data$evid == 0 & data$dvid == "cp" & data$dv > 0
  data.frame(dataset = label, id = as.character(data$id[rows]),
             time = data$time[rows], dv = data$dv[rows])
}
kept_synthetic <- suppressWarnings(
  synpmx_model_generate(misdosed_kept_fit, seed = 11))
synthetic <- suppressWarnings(synpmx_model_generate(misdosed_fit, seed = 11))
plotted <- rbind(concentration(misdosed, "Source"),
                 concentration(kept_synthetic, "Synthetic, patient kept"),
                 concentration(synthetic, "Synthetic, patient left out"))
plotted$dataset <- factor(plotted$dataset, levels = unique(plotted$dataset))
ggplot2::ggplot(plotted, ggplot2::aes(time, dv, group = id,
                                      colour = dataset)) +
  ggplot2::geom_line(alpha = 0.4) +
  ggplot2::facet_wrap(~dataset) +
  ggplot2::scale_y_log10() +
  ggplot2::scale_colour_manual(values = c(unname(comparison_colours),
                                          "#7570B3")) +
  ggplot2::labs(x = "Time (hours)", y = "Warfarin concentration") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")
```

The scorecard reads each release's verdict as row E3, so the finding reaches
anyone who scores the synthetic data:

```{r misdosed-card, eval=run}
kept_card <- synpmx_scorecard(misdosed, kept_synthetic, roles)
card <- synpmx_scorecard(misdosed, synthetic, roles)
data.frame(fit = c("patient kept", "patient left out"),
           E3 = c(kept_card$verdict[kept_card$check == "E3"],
                  card$verdict[card$check == "E3"]),
           result = c(kept_card$result[kept_card$check == "E3"],
                      card$result[card$check == "E3"]))
```

Leaving the patient out protects the release and keeps the synthetic data
plausible. The record is still wrong, and correcting the patient's dosing is
the remedy for any analysis of the real study.

### Extreme Covariates

One patient weighs 250 kg and another is 95 years old. Neither enters the
covariate summaries, because each is the most extreme patient on their side
and trimming sets them aside:

```{r extreme-covariates, eval=run}
summary_of <- function(model) {
  data.frame(covariate = c("wt", "age"),
             geometric_mean = exp(c(model$wt$meanlog, model$age$meanlog)),
             sd_log = c(model$wt$sdlog, model$age$sdlog))
}
rbind(cbind(study = "warfarin", summary_of(fit$covariates)),
      cbind(study = "with the two extreme patients",
            summary_of(extreme_fit$covariates)))
```

The same two summaries taken over every patient, without trimming, move with
the extreme patients:

```{r untrimmed, eval=run}
first <- function(data) data[!duplicated(data$id), ]
untrimmed <- function(data) {
  data.frame(covariate = c("wt", "age"),
             geometric_mean = exp(c(mean(log(first(data)$wt)),
                                    mean(log(first(data)$age)))),
             sd_log = c(stats::sd(log(first(data)$wt)),
                        stats::sd(log(first(data)$age))))
}
digits3(rbind(cbind(study = "warfarin", untrimmed(warfarin)),
              cbind(study = "with the two extreme patients",
                    untrimmed(extreme))))
```

P5 reads the covariate rows of the influence table: leaving out any one
patient barely moves them, the extreme patients included.

```{r extreme-influence, eval=run}
extreme_influence <- attr(model_privacy_checks(extreme_fit), "influence")
digits3(extreme_influence[extreme_influence$group == "covariate", ])
```

### A Category Nobody Else Holds

A third changed patient has a sex recorded as a level no other patient holds.
The level is excluded from the covariate distribution and from the schema the
generated table is written in, so no synthetic patient can carry it:

```{r rare-level, eval=run}
extreme_fit$covariates$sex$levels
levels(extreme_fit$schema$prototypes$sex)
```

## The Checks Across the Public Studies

The checks on every stored fit of the
[public-data evaluation](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html),
and the estimate each study's most influential remaining patient moves
furthest.

```{r survey, eval=run}
files <- c(warfarin = "warfarin-model-fit.rds",
           theo_md = "theo-md-model-fit.rds",
           mad = "mad-model-fit.rds",
           case1_pkpd = "case1-pkpd-model-fit.rds",
           wbcSim = "wbcsim-model-fit.rds",
           mavoglurant = "mavoglurant-model-fit.rds",
           nimoData = "nimo-model-fit.rds",
           pheno_sd = "pheno-model-fit.rds",
           mixroute_sim = "mixroute-sim-model-fit.rds",
           onc_sim = "onc-sim-model-fit.rds")
rows <- lapply(names(files), function(study) {
  one <- stored_fit(files[[study]])
  if (!inherits(one, "pmx_fitted_model")) return(NULL)
  checks <- model_privacy_checks(one)
  verdicts <- stats::setNames(checks$verdict, checks$check)
  data.frame(study = study, patients = one$n_source,
             left_out = if (is.null(one$privacy$left_out)) 0L else
               one$privacy$left_out$patients, t(verdicts),
             largest = sub("; after leaving.*$", "",
                           checks$result[checks$check == "P5"]),
             check.names = FALSE)
})
do.call(rbind, rows[!vapply(rows, is.null, logical(1))])
```

`left_out` counts the patients each study's estimation left out for moving an
estimate by 15 or more, and `largest` is the largest move among the patients
that remain. A move is a fraction of the cohort's spread divided by the cohort
size, so the same patient moves a twelve-patient fit about five times as far as
a sixty-patient one.

## Checks on the Synthetic Table

The checks above read the fit. `synpmx_scorecard()` reads the synthetic table
against its source, and its section B asks whether any synthetic patient
reproduces a real one. Four of its rows do not apply to this generator,
because no synthetic patient is assembled from a real patient's record; B3,
B4b and B5 do.

```{r card, eval=run}
synthetic <- synpmx_model_generate(fit, seed = 11)
card <- synpmx_scorecard(warfarin, synthetic, roles)
as.data.frame(card)[substr(card$check, 1, 1) %in% c("B", "E"),
                    c("check", "question", "result", "verdict")]
```

B3 compares how close synthetic patients sit to real ones against how close
real patients sit to each other, B4b asks whether any generated vector of
values copies a real patient's, and B5 asks whether a level fewer than three
real patients held reached the output.
[The scorecard](https://iamstein.github.io/synpmx/articles/scorecard.html)
documents each row and its pass criterion.

## Membership Inference Against the Release

A membership-inference attack gives each candidate a score that is high when
the release has moved, away from what the population would give, toward the
candidate's own values. Members of the study score higher on average than
people who were not in it, and the attack's strength is read as the AUC, the
probability that a randomly chosen member scores above a randomly chosen
non-member: 0.5 is chance and 1 is certain identification. The attacker
modelled holds the candidate's own measurements and reference values for the
population, and does not know the other patients. The release is reduced to its
two kinds of number: a population model of five parameters, each released as a
typical value and a between-subject variance, and the share of patients
sampled at each scheduled visit.
[Membership inference against the PMX model release](https://iamstein.github.io/synpmx/articles/pmxmodel-privacy-simulation.html)
works through the calculation on one small study, shows both tables in full,
and places each public study's release against them. Three readings:

- **Cohort size decides the population model's exposure.** With an exact
  reference the AUC falls from `r population_auc(12, 0)` at 12 patients to
  `r population_auc(60, 0)` at 60, and a reference off by 20% brings the two to
  `r population_auc(12, 0.2)` and `r population_auc(60, 0.2)`.
- **Two significant figures do not change it.** Scored on the same simulated
  studies before and after rounding, the AUC moves by at most
  `r rounding_change`, because an estimate's sampling variation is larger than
  one rounding step.
- **A table per visit would be the most exposed part of a release, so the
  release holds none.** A share of patients sampled at each of 20 visits gives
  `r attendance_auc(20, 10)` with 10 patients behind each share,
  `r attendance_auc(20, 30)` with 30 and `r attendance_auc(20, 100)` with 100,
  and the three-patient rule moves it by at most `r floor_change`, because the
  rule acts only on shares near 0 or 1. Attendance is one rate for the study
  instead, and a binary or ordinal endpoint one set of level frequencies, each
  resting on every patient.

An AUC averages over members, and the most exposed member is the one far from
the typical values: that patient moves the release furthest toward themselves,
which is what protection 8 reads. The membership-inference literature on
machine-learning models reports success against such patients separately for
this reason [14, 15], and Stadler and colleagues found that synthetic data from
generative models leaves them exposed [16].

The attacker modelled here is strong. The measurements it needs, a candidate's
random effects estimated from their concentrations or their outcome at every
visit, are made by the study, and outside it exist only where the same
quantities are measured for the same patient: at a study site, in a related
study, or in routine care such as drug-level monitoring.

## Outstanding Risks

These are the ways a release from this generator can still disclose something
about a patient, most serious first. Each is inherent to releasing estimates
without noise, or a limit of the checks.

- **An attacker who holds a candidate's own trial data can infer membership.**
  [Membership Inference Against the Release](#membership-inference-against-the-release)
  measures how well. The release holds no per-visit table, so what remains is
  chiefly the population model of a small cohort, which only noise would
  reduce.
- **An attacker who knows every other patient is not stopped.** Such an
  attacker can compute the exact effect of the one patient they do not know, so
  any move at all is detectable, however far under a threshold it sits. That
  attacker is the one DP is defined against, and only a DP mechanism stops them.
- **Releases compose, and their counts are exact.** The cohort size, the arm
  sizes and every frequency are exact counts or exact ratios of counts. Two
  releases from overlapping data, such as an interim and a final analysis, or
  the same study before and after one patient withdraws, differ by exactly what
  the patients between them contributed. Nothing here accounts for that.
- **Three patients is the smallest threshold in common use.** Every released
  frequency rests on at least three patients on each side, which is the floor
  the threshold rule of SDC starts from. A release held to a stricter rule needs
  `min_arm_patients` and `min_category_patients` raised to match.
- **The influence reading describes this study only, and its thresholds are
  empirical.** How far one patient moves an estimate is itself a fact about the
  data, which is why DP calibrates its noise to *smooth* sensitivity rather
  than local sensitivity [17]. The review and failure thresholds were set
  against the public studies above, so that ordinary patients pass and a gross
  data error fails; they are not derived from a privacy target. Shrinkage makes
  the reading understate the move, and where shrinkage is close to 1 the
  reading says little while its verdict still reads pass.
- **Not every estimate is read.** The residual error, a parameter estimated
  without a between-subject term (bioavailability in a mixed-route model), the
  PD shape parameters other than the baseline, and a planned dose taken as the
  arm's mean where no amount is shared have no row.
- **Leaving out stops at a tenth of the cohort.** A study with more patients
  than that whom the influence reading flags keeps the rest, and P5 then reads
  a review or a failure.
- **The full fit holds diagnostics the release does not.** The Akaike
  information criterion (AIC) of every candidate model, the starting values and
  the correlations between covariates and the individual random effects stay on
  the object `synpmx_model_estimate()` returns. Share the release or the
  synthetic data, never the fit.
- **The estimation output names patients.** A patient left out of the
  estimates, and a patient who fails the influence reading, is named on the
  console, so logs and rendered reports from the environment that holds the
  study carry that identifier.
- **Rarity in the world is not measured.** A covariate level held by many
  patients in this study can still identify someone if few people alive hold
  it.

## What a Formal Guarantee Would Take

The DP method that fits a generator built on `nlmixr2` is
subsample-and-aggregate [17, 18], because it treats the estimator as a black
box:

1. Split the patients into *k* disjoint groups and fit the population model in
   each.
2. Clip each group's estimate of each log parameter to a public range of width
   *S*.
3. Release the mean over the groups plus Laplace noise of scale
   $b = dS/(\varepsilon k)$ per parameter, where *d* parameters share the
   privacy budget $\varepsilon$.

One patient is in one group, so the guarantee holds whatever the fit inside a
group does. The textbook DP methods for regression perturb a convex loss with
bounded gradients [19], and a population likelihood is neither, which rules
them out here. The cost is the error law of the
[privacy background](https://iamstein.github.io/synpmx/articles/privacy-background.html)
with the cohort size replaced by the number of groups. At about twenty patients
per group, and a median Laplace error of $b \ln 2$:

```{r subsample-aggregate, eval=TRUE}
grid <- expand.grid(patients = c(60, 200, 1000), released = c(3, 11))
grid$groups <- grid$patients / 20
scale <- grid$released * log(10) / (1 * grid$groups)
grid$median_fold_error <- signif(exp(log(2) * scale), 3)
grid
```

That is with $\varepsilon = 1$ and a public range spanning tenfold for each
parameter. With all eleven parameters of a two-compartment model released, the
error stays above 40% even at a thousand patients; with three released and the
rest taken from a public prior, it reaches about 10% there. At the cohort sizes
this generator usually sees, the
[calibrated generator's evaluation](https://iamstein.github.io/synpmx/articles/calibrated-public-data-examples.html)
finds that a release is worse than generating from the public prior alone.
Subsample-and-aggregate is not implemented.

## Released Parameters at a Glance

Every number in a release that was estimated from patients, counted once, for
the stored fit of each public study. A rate pooled over the arms is listed in
the release once per arm and counts once here, and the planned dose schedule,
which is the protocol, is counted apart.

```{r parameters, eval=run}
# An arm's dosing model, or one per drug where the study declared which doses
# drive which endpoint.
per_drug <- function(entry) if (!is.null(entry$planned)) list(entry) else entry
parameter_count <- function(release) {
  arms <- names(release$arms$sizes)
  shape <- function(s) length(s$typical) + 2
  rates <- unique(unlist(lapply(release$dosing, function(entry) {
    lapply(per_drug(entry), function(d) {
      paste(d$reduction, d$interruption, d$discontinuation,
            paste(d$levels, collapse = "/"))
    })
  })))
  attendance <- unique(unlist(lapply(release$visits, function(v) {
    as.numeric(v$probability)
  })))
  marginals <- unique(Filter(Negate(is.null),
                             unlist(release$discrete, recursive = FALSE)))
  counts <- c(
    `PK model` = sum(vapply(release$pk_models, function(m) {
      length(m$parameters$fixed) + nrow(m$parameters$omega) + 1
    }, numeric(1))),
    `PD time courses` = sum(vapply(release$pd, function(s) {
      if (length(s$arms)) sum(vapply(s$arms, shape, numeric(1))) else shape(s)
    }, numeric(1))),
    covariates = sum(vapply(release$covariates, function(c) {
      switch(c$kind, lognormal = 2, normal = 2,
             categorical = length(c$levels) - 1, 0)
    }, numeric(1))),
    `dose changes` = sum(vapply(strsplit(rates, " "), function(r) {
      3 + length(strsplit(r[[4]], "/")[[1]]) - 1
    }, numeric(1))),
    attendance = sum(attendance > 0 & attendance < 1),
    `discrete endpoints` = sum(vapply(marginals, function(m) {
      max(0, length(m$levels) - 1)
    }, numeric(1))),
    `counts and floors` = length(arms) + 1 +
      length(release$quantification_floor) +
      length(unique(unlist(lapply(release$covariate_effects,
                                  function(e) e$reference)))))
  c(counts, total = sum(counts),
    `planned cycles (protocol)` = sum(vapply(release$dosing, function(entry) {
      sum(vapply(per_drug(entry), function(d) nrow(d$planned), numeric(1)))
    }, numeric(1))))
}
studies <- c(warfarin = "warfarin-model-fit.rds",
             theo_md = "theo-md-model-fit.rds",
             mad = "mad-model-fit.rds",
             case1_pkpd = "case1-pkpd-model-fit.rds",
             wbcSim = "wbcsim-model-fit.rds",
             mavoglurant = "mavoglurant-model-fit.rds",
             nimoData = "nimo-model-fit.rds",
             pheno_sd = "pheno-model-fit.rds",
             mixroute_sim = "mixroute-sim-model-fit.rds",
             onc_sim = "onc-sim-model-fit.rds")
counted <- do.call(rbind, lapply(names(studies), function(study) {
  release <- model_release(stored_fit(studies[[study]]))
  data.frame(study = study, patients = release$n_source,
             t(parameter_count(release)), check.names = FALSE)
}))
knitr::kable(counted, row.names = FALSE)
```

What each column holds:

- **PK model**: a typical value and a between-subject variance for each PK
  parameter, and one residual error: 11 numbers for the five parameters of a
  two-compartment oral model. A parameter fitted without between-subject
  variability, such as a bioavailability, adds its typical value alone.
- **PD time courses**: for each other continuous endpoint, one to three shape
  parameters, the spread of the patients' baselines, and a residual error.
- **Covariates**: a trimmed mean and SD for each continuous covariate, and one
  fewer frequency than levels for each categorical one.
- **Dose changes**: the three pooled rates of reducing, skipping and stopping,
  and each level of the dose ladder below the full dose.
- **Attendance**: one rate for the study, or none where every patient attended
  every visit their arm had.
- **Discrete endpoints**: one fewer level frequency than levels, for each
  binary or ordinal endpoint.
- **Counts and floors**: the size of each arm and of the cohort, an emission
  floor for each endpoint without a declared assay limit, and the reference
  weight where body-weight scaling was requested.
- **Planned cycles**: each arm's planned dose times and amounts, which are the
  protocol rather than estimates. The nominal visit grid and the description of
  the table to generate are the protocol too, and are not counted.

## References

1. Dwork C, McSherry F, Nissim K, Smith A. Calibrating noise to sensitivity in
   private data analysis. *Theory of Cryptography Conference (TCC).* 2006.
2. Homer N, Szelinger S, Redman M, et al. Resolving individuals contributing
   trace amounts of DNA to highly complex mixtures using high-density SNP
   genotyping microarrays. *PLoS Genetics.* 2008;4(8):e1000167.
3. Sankararaman S, Obozinski G, Jordan MI, Halperin E. Genomic privacy and
   limits of individual detection in a pool. *Nature Genetics.*
   2009;41:965–967.
4. Dwork C, Smith A, Steinke T, Ullman J, Vadhan S. Robust traceability from
   trace amounts. *IEEE Symposium on Foundations of Computer Science (FOCS).*
   2015.
5. Fredrikson M, Lantz E, Jha S, Lin S, Page D, Ristenpart T. Privacy in
   pharmacogenetics: an end-to-end case study of personalized warfarin dosing.
   *USENIX Security Symposium.* 2014.
6. Dinur I, Nissim K. Revealing information while preserving privacy.
   *ACM Symposium on Principles of Database Systems (PODS).* 2003.
7. Novartis. `synadam`: generate synthetic ADaM datasets. R package version
   0.3.4. <https://github.com/Novartis/synadam>.
8. Hundepool A, Domingo-Ferrer J, Franconi L, et al. *Statistical Disclosure
   Control.* Wiley; 2012.
9. Huber PJ, Ronchetti EM. *Robust Statistics.* 2nd ed. Wiley; 2009.
10. Bun M, Steinke T. Average-case averages: private algorithms for smooth
    sensitivity and mean estimation. *Advances in Neural Information
    Processing Systems (NeurIPS).* 2019.
11. Cook RD. Detection of influential observation in linear regression.
    *Technometrics.* 1977;19(1):15–18.
12. Carlini N, Jagielski M, Zhang C, Papernot N, Terzis A, Tramèr F. The
    privacy onion effect: memorization is relative. *Advances in Neural
    Information Processing Systems (NeurIPS).* 2022.
13. Dwork C, Lei J. Differential privacy and robust statistics. *ACM Symposium
    on Theory of Computing (STOC).* 2009.
14. Shokri R, Stronati M, Song C, Shmatikov V. Membership inference attacks
    against machine learning models. *IEEE Symposium on Security and Privacy.*
    2017.
15. Carlini N, Chien S, Nasr M, Song S, Terzis A, Tramèr F. Membership
    inference attacks from first principles. *IEEE Symposium on Security and
    Privacy.* 2022.
16. Stadler T, Oprisanu B, Troncoso C. Synthetic data – anonymisation groundhog
    day. *USENIX Security Symposium.* 2022.
17. Nissim K, Raskhodnikova S, Smith A. Smooth sensitivity and sampling in
    private data analysis. *ACM Symposium on Theory of Computing (STOC).* 2007.
18. Smith A. Privacy-preserving statistical estimation with optimal
    convergence rates. *ACM Symposium on Theory of Computing (STOC).* 2011.
19. Chaudhuri K, Monteleoni C, Sarwate AD. Differentially private empirical
    risk minimization. *Journal of Machine Learning Research.*
    2011;12:1069–1109.
