---
title: "Membership Inference Against the PMX Model Release"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    fig_width: 9
    fig_height: 3.5
vignette: >
  %\VignetteIndexEntry{Membership Inference Against the PMX Model Release}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
stored_fit <- function(file) {
  path <- system.file("extdata", file, package = "synpmx")
  if (nzchar(path)) readRDS(path) else NULL
}
two <- function(x) sprintf("%.2f", x)
```

[Privacy protections in the PMX model generator](https://iamstein.github.io/synpmx/articles/pmxmodel-privacy.html)
reports how well an attacker can tell whether a given patient was in a study
from what `synpmx_model()` releases. This article shows how those numbers are
computed: what the attacker holds, the score it gives each person, and how the
area under the curve (AUC) summarizes how well that score separates patients
who were in the study from patients who were not. Every number is computed in
this article.

**The simulated release is a simplified model of the real one, and is not the
package's code.** It keeps the two kinds of number that membership inference
works on, a population model's typical values and variances, and the share of
patients sampled at each scheduled visit, and reduces each to what the attack
needs. The code is shown in full at the end.

```{r functions, file="pmxmodel-privacy-simulation.R", include=FALSE}
```

## What the Attacker Holds

Membership inference asks one question about one person, the **candidate**: was
this person in the study? The attacker holds three things:

1. the release, which `synpmx_model()` attaches to every synthetic dataset it
   generates;
2. the candidate's own measurements of the kind the study made, such as which
   scheduled visits they were sampled at, or their **random effects**, which say
   how far that person's own drug clearance and volume sit from the typical
   values and which their concentrations imply;
3. reference values for the population the study drew from, such as the
   average attendance at each scheduled visit.

The attacker does not see the other patients. A patient in the study
contributed to every number in the release, so the release leans slightly
toward them; a patient who was not in it contributed nothing. The attack
measures that lean.

## One Study, Five Visits

The smallest case shows every step. Ten patients each have a concentration
sample scheduled at 5 visits, and each sample is either taken (1) or missed
(0). The average attendance differs by visit, as it does in a sparse sampling
design, where each patient is sampled at some of the scheduled times. Each row
is a patient:

```{r example-study}
set.seed(3)
average <- c(0.3, 0.5, 0.7, 0.4, 0.6)
study <- draw_attendance(10, average, correlation = 0)
dimnames(study) <- list(paste("patient", 1:10), paste("visit", 1:5))
study
```

The simulated release holds the share of patients sampled at each visit,
after the floor that rounds a share with one or two patients on either side to
0 or 1:

```{r example-release}
released <- release_shares(study, floor = 3)
rbind(average = average, released_share = released)
```

```{r example-floor, include=FALSE}
held <- colSums(study)
floored <- which(released != held / 10)
```

`r if (length(floored)) paste0("At visit ", floored[[1L]], ", ", held[[floored[[1L]]]], " of the 10 patients were sampled, which leaves ", 10 - held[[floored[[1L]]]], " who missed it, so the floor released the share as ", released[[floored[[1L]]]], ".") else "No share in this study had one or two patients on either side, so the floor changed nothing."`

There are two candidates: patient 1, who is in the study, and a new patient
from the same population, who is not.

```{r example-candidates}
member <- unname(study[1, ])
non_member <- draw_attendance(1, average, correlation = 0)[1, ]
```

**The score asks whether the visits the candidate missed are the visits whose
released attendance came out below average.** At each visit it multiplies how
far the candidate sits from the average, 1 minus the average for a visit they
were sampled at and minus the average for one they missed, by how far the
released share sits from the average, and adds the products. A product is
positive when the candidate and the release lean the same way: sampled where
the share came out above average, or missed where it came out below.

```{r example-score}
data.frame(
  visit = 1:5, average = average, share = unname(released),
  member = member, member_term = (member - average) * (released - average),
  non_member = non_member,
  non_member_term = (non_member - average) * (released - average),
  row.names = NULL)
scores <- c(member = attendance_score(member, average, released),
            non_member = attendance_score(non_member, average, released))
scores
```

Patient 1 is one of the ten patients behind every share, so each share leans
one tenth of the way toward patient 1's own attendance: a visit patient 1
missed comes out a tenth lower than it would have with them sampled there. The
non-member's attendance had no part in the shares. In this one study the member
scores
`r if (scores[["member"]] > scores[["non_member"]]) "higher" else "lower"` than
the non-member, but one study shows little, because the shares also vary by
chance from one study to the next, by more than one patient's tenth. The lean
shows over many studies.

## From Scores to the AUC

The experiment is repeated: a new study is drawn, one member and one non-member
are scored, and this is done 2,000 times. Each panel shows the two resulting
piles of scores, for 10 patients scheduled at 5, 20 or 60 visits:

```{r score-piles}
set.seed(11)
piles <- do.call(rbind, lapply(c(5, 20, 60), function(visits) {
  one <- attendance_scores(10, visits, correlation = 0, floor = 3,
                           draws = 2000)
  rbind(data.frame(visits = visits, candidate = "member",
                   score = one$member),
        data.frame(visits = visits, candidate = "non-member",
                   score = one$non_member))
}))
piles$panel <- factor(paste(piles$visits, "visits"),
                      levels = paste(c(5, 20, 60), "visits"))
```

```{r score-distributions, eval=has_ggplot2, fig.height=3.2}
ggplot2::ggplot(piles, ggplot2::aes(score, fill = candidate)) +
  ggplot2::geom_density(alpha = 0.5, colour = NA) +
  ggplot2::facet_wrap(~panel, scales = "free") +
  ggplot2::scale_fill_manual(values = c(member = "#D95F02",
                                        `non-member` = "#1B6CA8")) +
  ggplot2::labs(x = "Attacker's score", y = NULL, fill = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
                 legend.position = "top")
```

**The AUC is the probability that a randomly chosen member scores above a
randomly chosen non-member [1].** An AUC of 0.5 means the two piles are the
same, and the score is a coin toss. An AUC of 1 means every member scores above
every non-member. The name comes from the receiver operating characteristic
curve, which plots the share of members flagged against the share of
non-members flagged as a threshold on the score moves; the area under it equals
that probability.

```{r panel-auc}
panel_auc <- vapply(c(5, 20, 60), function(visits) {
  auc(piles$score[piles$visits == visits & piles$candidate == "member"],
      piles$score[piles$visits == visits & piles$candidate == "non-member"])
}, numeric(1))
data.frame(visits = c(5, 20, 60), auc = two(panel_auc))
```

An AUC is not the share of patients identified. An AUC of
`r two(panel_auc[[2]])` means that, shown one
patient who was in the study and one who was not, the attacker picks the right
one that often. The piles still overlap, so for any single person the attacker
can be wrong; what the AUC measures is how far the release, taken over many
people, gives membership away.

## The Attendance Table, Row by Row

```{r tables}
tables <- membership_tables()
attendance <- tables$attendance
data.frame(visits = attendance$visits,
           patients_behind_share = attendance$patients,
           correlation_within_patient = attendance$correlation,
           auc_without_floor = two(attendance$without_floor),
           auc_released = two(attendance$released))
```

Each row is one setting of the experiment above, repeated 1,500 times, with the
average attendance at each visit drawn between 0.2 and 0.8, which is where the
in-between shares of the sparse-sampling public studies sit (the last table
below):

- `visits` is how many scheduled visits a patient can be sampled at, and so how
  many shares the release holds.
- `patients_behind_share` is how many patients each share is a share of. The
  release pools a share over the arms that have the visit, so this is usually
  the whole cohort.
- `correlation_within_patient` is 0 where a patient's visits are independent,
  and 0.6 where a patient who misses one visit is much more likely to miss the
  next, as a patient drifting out of the study would.
- `auc_without_floor` releases every share as it is.
- `auc_released` rounds a share with one or two patients on either side to 0
  or 1, as `synpmx_model()` releases it. Both columns score the same simulated
  studies, so they differ only by what the floor does.

Read across one row: with `r attendance$patients[[2]]` patients behind each
share and `r attendance$visits[[2]]` independently attended visits, an attacker
who knows which of the `r attendance$visits[[2]]` visits one candidate was
sampled at, and the average attendance at each, picks the member over the
non-member with probability `r two(attendance$released[[2]])`.

Three things move the AUC, and the floor barely does. **More visits raise
it**, because each visit adds the same small lean toward every member: the
leans add up in proportion to the number of visits, while the chance variation
grows only with its square root. **More patients behind each share lower
it**, because each patient is a smaller part of every share. **Correlation
within a patient lowers
it**, because correlated visits repeat the same information. **The floor
barely moves it**, because it changes only shares near 0 or 1.

**The release does not hold this table.** Its visit model is one attendance rate
per endpoint, applied at every visit of the endpoint an arm has, so an attacker
who knows which visits a candidate was sampled at has one number per endpoint
to match rather than a share per visit. Pooling an endpoint's visits into one
rate shrinks the member's lean on it and its chance variation alike, so each
rate exposes about as much as one visit's share, and a study with five
endpoints sits near the five-visit rows. A binary or ordinal endpoint
is released the same way, as one set of level frequencies over the whole
study, and dose changes as three rates.

## The Population-Model Table

The population model works the same way with different numbers. Each patient
has a random effect on each of five parameters of a two-compartment oral model:
clearance, the central and peripheral volumes, the flow between them, and the
absorption rate. The released typical value of a parameter sits close to the
true typical value times the exponential of the patients' mean random effect,
and the released between-subject variance close to the mean of their squared
random effects. A patient in the study therefore pulls each typical value
toward their own random effect and each variance toward their own squared one,
by one part in the cohort size.

The attacker holds the candidate's own random effects and reference values for
the population, and the score adds two kinds of product, each positive when the
release leans toward the candidate: the candidate's random effect times how far
the released typical value sits from the reference, and the candidate's squared
random effect, relative to the variance, times how far the released variance
sits from the reference variance.

```{r population}
population <- tables$population
data.frame(patients = population$patients,
           reference_off_by = sprintf("%.0f%%", 100 * population$off_by),
           auc_unrounded = two(population$unrounded),
           auc_two_figures = two(population$two_figures),
           auc_released = two(population$released))
```

- `patients` is the cohort the model is fitted to.
- `reference_off_by` is the error in the attacker's reference values. At 0% the
  attacker knows the population's true values. At 10% or 20% the reference
  comes from another study of the same drug, whose population differs from this
  one's.
- `auc_unrounded` releases every estimate at full precision.
- `auc_two_figures` rounds every estimate to two significant figures.
- `auc_released` rounds every estimate to one significant figure, as
  `synpmx_model()` releases them. All three columns score the same simulated
  studies.

Read across one row: for a model fitted to `r population$patients[[1]]`
patients, an attacker who holds a candidate's random effects and the
population's true values picks the member over the non-member with probability
`r two(population$released[[1]])`, and with a reference off by 20%,
`r two(population$released[population$patients == population$patients[[1]] & population$off_by == 0.2])`.
The AUC falls with cohort size and with the reference's error. Rounding to two
significant figures leaves it where it was, because an estimate's sampling
variation from one study to the next is larger than a two-figure step. One
figure lowers it, because a one-figure step, from a tenth of the value to all
of it, is as wide as that variation, so the released value is often the same
with the candidate in the study or out; that is why the release rounds to
one.

## What Each Public Study Releases

The numbers the attack could use in the release of each stored fit in the
[public-data evaluation](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html):
the random effects of the population model, the attendance rate of each
endpoint, and the free level frequencies of the binary and ordinal endpoints,
each of which rests on every patient in the study or in the arms that measured
the endpoint. An endpoint with several levels has one fewer free frequency than
levels, because they add up to 1.

```{r public-studies}
files <- c(warfarin = "warfarin-model-fit.rds",
           theo_md = "theo-md-model-fit.rds",
           mad = "mad-model-fit.rds",
           case1_pkpd = "case1-pkpd-model-fit.rds",
           wbcSim = "wbcsim-model-fit.rds",
           mavoglurant = "mavoglurant-model-fit.rds",
           nimoData = "nimo-model-fit.rds",
           pheno_sd = "pheno-model-fit.rds",
           mixroute_sim = "mixroute-sim-model-fit.rds",
           onc_sim = "onc-sim-model-fit.rds")
rows <- lapply(names(files), function(study) {
  fit <- stored_fit(files[[study]])
  if (!inherits(fit, "pmx_fitted_model")) return(NULL)
  release <- model_release(fit)
  # One rate per endpoint, listed at each of its visits in each arm.
  rate <- unique(unlist(lapply(release$visits, function(v) {
    p <- as.numeric(v$probability)
    paste(release$cells$endpoint, p)[p > 0 & p < 1]
  })))
  marginals <- unique(Filter(Negate(is.null),
                             unlist(release$discrete, recursive = FALSE)))
  data.frame(study = study, patients = release$n_source,
             random_effects = sum(vapply(release$pk_models, function(model) {
               nrow(model$parameters$omega)
             }, integer(1))),
             attendance_rates = length(rate),
             discrete_frequencies = sum(vapply(marginals, function(m) {
               max(0, length(m$levels) - 1)
             }, numeric(1))))
})
do.call(rbind, rows[!vapply(rows, is.null, logical(1))])
```

An attendance rate of 1 tells the attacker nothing, so `attendance_rates`
counts the endpoints whose rate is below 1, each a single number for the
patients behind it and each worth about one visit in the attendance table.
For the population model, the
cohort size places a study among the rows of the second table; a study fitted
with more random effects than five gives the attacker more numbers to add up.

## What the Simulation Leaves Out

- **The attacker is strong.** Its reference is exact except where the error is
  set, and it holds the candidate's own trial measurements, which outside the
  study exist only where the same quantities are measured for the same patient:
  at a study site, in a related study, or in routine care.
- **The AUC is an average over members.** A patient far from the typical
  values, or with an unusual run of outcomes, moves the release further toward
  themselves and is more exposed than the average member, which is why
  membership-inference studies report the attack's success on such patients
  separately [2, 3].
- **The release is simplified.** The visits are exchangeable apart from their
  rates, the random effects are normal and independent across parameters, and
  the real release holds further numbers this simulation leaves out: the dosing
  model, the covariate summaries and the PD time courses.
- **The score is not the most powerful one possible.** It is of the kind used
  to trace individuals in pooled genetic data [4, 5, 6]. An attacker who can
  simulate studies from the same population can calibrate each candidate's
  score, as the strongest attacks on machine-learning models do [3], so these
  AUCs are a lower bound for an attacker holding what this one holds.

## The Simulation Code

Every function used above, as run:

```{r code, file="pmxmodel-privacy-simulation.R", eval=FALSE}
```

## References

1. Hanley JA, McNeil BJ. The meaning and use of the area under a receiver
   operating characteristic (ROC) curve. *Radiology.* 1982;143(1):29–36.
2. Shokri R, Stronati M, Song C, Shmatikov V. Membership inference attacks
   against machine learning models. *IEEE Symposium on Security and Privacy.*
   2017.
3. Carlini N, Chien S, Nasr M, Song S, Terzis A, Tramèr F. Membership
   inference attacks from first principles. *IEEE Symposium on Security and
   Privacy.* 2022.
4. Homer N, Szelinger S, Redman M, et al. Resolving individuals contributing
   trace amounts of DNA to highly complex mixtures using high-density SNP
   genotyping microarrays. *PLoS Genetics.* 2008;4(8):e1000167.
5. Sankararaman S, Obozinski G, Jordan MI, Halperin E. Genomic privacy and
   limits of individual detection in a pool. *Nature Genetics.*
   2009;41:965–967.
6. Dwork C, Smith A, Steinke T, Ullman J, Vadhan S. Robust traceability from
   trace amounts. *IEEE Symposium on Foundations of Computer Science (FOCS).*
   2015.
