---
title: "Membership Inference Against the PMX Model Release"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    fig_width: 9
    fig_height: 3.5
vignette: >
  %\VignetteIndexEntry{Membership Inference Against the PMX Model Release}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
stored_fit <- function(file) {
  path <- system.file("extdata", file, package = "synpmx")
  if (nzchar(path)) readRDS(path) else NULL
}
two <- function(x) sprintf("%.2f", x)
```

[Privacy protections in the PMX model generator](https://iamstein.github.io/synpmx/articles/pmxmodel-privacy.html)
reports how well an attacker can tell whether a given patient was in a study
from what `synpmx_model()` releases. This article shows how those numbers are
computed: what the attacker holds, the score it gives each person, and how the
area under the curve (AUC) summarizes how well that score separates patients
who were in the study from patients who were not. Every number is computed in
this article.

**The simulated release is a simplified model of the real one, and is not the
package's code.** It keeps the two kinds of number that membership inference
works on, a population model's typical values and variances, and the share of
patients with an outcome at each visit, and reduces each to what the attack
needs. The code is shown in full at the end.

```{r functions, file="pmxmodel-privacy-simulation.R", include=FALSE}
```

## What the Attacker Holds

Membership inference asks one question about one person, the **candidate**: was
this person in the study? The attacker holds three things:

1. the release, which `synpmx_model()` attaches to every synthetic dataset it
   generates;
2. the candidate's own measurements of the kind the study made, such as their
   yes/no outcome at each visit, or their **random effects**, which say how far
   that person's own drug clearance and volume sit from the typical values and
   which their concentrations imply;
3. reference values for the population the study drew from, such as the true
   rate of a yes at each visit.

The attacker does not see the other patients. A patient in the study
contributed to every number in the release, so the release leans slightly
toward them; a patient who was not in it contributed nothing. The attack
measures that lean.

## One Arm, Five Visits

The smallest case shows every step. Ten patients record a yes/no outcome at 5
visits, such as whether a symptom was present, and the true rate of a yes
differs by visit. Each row is a patient:

```{r example-study}
set.seed(3)
rate <- c(0.3, 0.5, 0.7, 0.4, 0.6)
arm <- draw_outcomes(10, rate, correlation = 0)
dimnames(arm) <- list(paste("patient", 1:10), paste("visit", 1:5))
arm
```

The release holds the share of patients with a yes at each visit, after the
floor that rounds a share with one or two patients on either side to 0 or 1:

```{r example-release}
released <- release_shares(arm, floor = 3)
rbind(true_rate = rate, released_share = released)
```

```{r example-floor, include=FALSE}
held <- colSums(arm)
floored <- which(released != held / 10)
```

`r if (length(floored)) paste0("At visit ", floored[[1L]], ", ", held[[floored[[1L]]]], " of the 10 patients had a yes, which leaves ", 10 - held[[floored[[1L]]]], " with a no, so the floor released the share as ", released[[floored[[1L]]]], ".") else "No share in this study had one or two patients on either side, so the floor changed nothing."`

There are two candidates: patient 1, who is in the study, and a new patient
from the same population, who is not.

```{r example-candidates}
member <- unname(arm[1, ])
non_member <- draw_outcomes(1, rate, correlation = 0)[1, ]
```

**The score multiplies, at each visit, how far the candidate's outcome sits from
the true rate by how far the released share sits from it, and adds the
products.** A product is positive when the candidate and the release lean the
same way: a yes where the share came out above the true rate, or a no where it
came out below.

```{r example-score}
data.frame(
  visit = 1:5, true_rate = rate, share = unname(released),
  member = member, member_term = (member - rate) * (released - rate),
  non_member = non_member,
  non_member_term = (non_member - rate) * (released - rate),
  row.names = NULL)
scores <- c(member = frequency_score(member, rate, released),
            non_member = frequency_score(non_member, rate, released))
scores
```

Patient 1 is one of the ten patients behind every share, so each share leans
one tenth of the way toward patient 1's own outcome. The non-member's outcomes
had no part in the shares. In this one study the member scores
`r if (scores[["member"]] > scores[["non_member"]]) "higher" else "lower"` than
the non-member, but one study shows little, because the shares also vary by
chance from one study to the next, by more than one patient's tenth. The lean
shows over many studies.

## From Scores to the AUC

The experiment is repeated: a new study is drawn, one member and one non-member
are scored, and this is done 2,000 times. Each panel shows the two resulting
piles of scores, for 10 patients measured at 5, 20 or 60 visits:

```{r score-piles}
set.seed(11)
piles <- do.call(rbind, lapply(c(5, 20, 60), function(visits) {
  one <- frequency_scores(10, visits, correlation = 0, floor = 3,
                          draws = 2000)
  rbind(data.frame(visits = visits, candidate = "member",
                   score = one$member),
        data.frame(visits = visits, candidate = "non-member",
                   score = one$non_member))
}))
piles$panel <- factor(paste(piles$visits, "visits"),
                      levels = paste(c(5, 20, 60), "visits"))
```

```{r score-distributions, eval=has_ggplot2, fig.height=3.2}
ggplot2::ggplot(piles, ggplot2::aes(score, fill = candidate)) +
  ggplot2::geom_density(alpha = 0.5, colour = NA) +
  ggplot2::facet_wrap(~panel, scales = "free") +
  ggplot2::scale_fill_manual(values = c(member = "#D95F02",
                                        `non-member` = "#1B6CA8")) +
  ggplot2::labs(x = "Attacker's score", y = NULL, fill = NULL) +
  ggplot2::theme_minimal() +
  ggplot2::theme(axis.text.y = ggplot2::element_blank(),
                 legend.position = "top")
```

**The AUC is the probability that a randomly chosen member scores above a
randomly chosen non-member [1].** An AUC of 0.5 means the two piles are the
same, and the score is a coin toss. An AUC of 1 means every member scores above
every non-member. The name comes from the receiver operating characteristic
curve, which plots the share of members flagged against the share of
non-members flagged as a threshold on the score moves; the area under it equals
that probability.

```{r panel-auc}
panel_auc <- vapply(c(5, 20, 60), function(visits) {
  auc(piles$score[piles$visits == visits & piles$candidate == "member"],
      piles$score[piles$visits == visits & piles$candidate == "non-member"])
}, numeric(1))
data.frame(visits = c(5, 20, 60), auc = two(panel_auc))
```

An AUC is not the share of patients identified. An AUC of
`r two(panel_auc[[2]])` means that, shown one
patient who was in the study and one who was not, the attacker picks the right
one that often. The piles still overlap, so for any single person the attacker
can be wrong; what the AUC measures is how far the release, taken over many
people, gives membership away.

## The Per-Visit Table, Row by Row

```{r tables}
tables <- membership_tables()
frequency <- tables$frequency
data.frame(visits = frequency$visits,
           patients_behind_share = frequency$patients,
           correlation_within_patient = frequency$correlation,
           auc_without_floor = two(frequency$without_floor),
           auc_released = two(frequency$released))
```

Each row is one setting of the experiment above, repeated 1,500 times, with
the rate at each visit drawn between 0.2 and 0.8:

- `visits` is how many visits record the outcome, and so how many shares the
  release holds for the arm.
- `patients_behind_share` is how many patients each share is a share of. The
  release pools a share over the arms that have the visit, so this is usually
  the whole cohort.
- `correlation_within_patient` is 0 where a patient's visits are independent,
  and 0.6 where a patient with a yes at one visit is much more likely to have
  one at the next, as a chronic symptom would be.
- `auc_without_floor` releases every share as it is.
- `auc_released` rounds a share with one or two patients on either side to 0
  or 1, as `synpmx_model()` releases it. Both columns score the same simulated
  studies, so they differ only by what the floor does.

Read across one row: with `r frequency$patients[[2]]` patients behind each
share and a yes/no outcome at `r frequency$visits[[2]]` independent visits, an
attacker who
holds one candidate's `r frequency$visits[[2]]` outcomes and the true rate at
each visit picks the member over the non-member with probability
`r two(frequency$released[[2]])`.

Three things move the AUC, and the floor barely does. **More visits raise
it**, because each visit adds the same small lean toward every member: the
leans add up in proportion to the number of visits, while the chance variation
grows only with its square root. **More patients behind each share lower
it**, because each patient is a smaller part of every share. **Correlation
within a patient lowers
it**, because correlated visits repeat the same information. **The floor
barely moves it**, because it changes only shares near 0 or 1.

In the release, two fields are tables of this kind. `visits` holds the share of
patients measured at each scheduled visit, where the candidate's outcome is
whether they attended; `discrete` holds the share of patients at each level of
a binary or ordinal endpoint at each visit. Both are pooled over the arms that
have the visit, so the patients behind a share are those arms' patients
together. A share of 0 or 1 says the same about every patient behind it and
gives the attacker nothing, so the shares that count are those strictly between
0 and 1.

## The Population-Model Table

The population model works the same way with different numbers. Each patient
has a random effect on each of five parameters of a two-compartment oral model:
clearance, the central and peripheral volumes, the flow between them, and the
absorption rate. The released typical value of a parameter sits close to the
true typical value times the exponential of the patients' mean random effect,
and the released between-subject variance close to the mean of their squared
random effects. A patient in the study therefore pulls each typical value
toward their own random effect and each variance toward their own squared one,
by one part in the cohort size.

The attacker holds the candidate's own random effects and reference values for
the population, and the score adds two kinds of product, each positive when the
release leans toward the candidate: the candidate's random effect times how far
the released typical value sits from the reference, and the candidate's squared
random effect, relative to the variance, times how far the released variance
sits from the reference variance.

```{r population}
population <- tables$population
data.frame(patients = population$patients,
           reference_off_by = sprintf("%.0f%%", 100 * population$off_by),
           auc_unrounded = two(population$unrounded),
           auc_released = two(population$released))
```

- `patients` is the cohort the model is fitted to.
- `reference_off_by` is the error in the attacker's reference values. At 0% the
  attacker knows the population's true values. At 10% or 20% the reference
  comes from another study of the same drug, whose population differs from this
  one's.
- `auc_unrounded` releases every estimate at full precision.
- `auc_released` rounds every estimate to two significant figures, as
  `synpmx_model()` releases them. Both columns score the same simulated
  studies.

Read across one row: for a model fitted to `r population$patients[[1]]`
patients, an attacker who holds a candidate's random effects and the
population's true values picks the member over the non-member with probability
`r two(population$released[[1]])`, and with a reference off by 20%,
`r two(population$released[population$patients == population$patients[[1]] & population$off_by == 0.2])`.
The AUC falls with cohort size and with the reference's error. Rounding to two
significant figures leaves it where it was, because an estimate's sampling
variation from one study to the next is larger than one rounding step.

## Each Public Study's Release Against the Tables

The features that place a study among the rows above, read from the release of
each stored fit in the
[public-data evaluation](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html):
the number of random effects, and for `visits` and `discrete` the number of
per-visit shares strictly between 0 and 1 with the fewest patients behind any
of them. At a visit where a binary or ordinal endpoint has several levels,
their shares add up to 1, so the visit counts one fewer than its number of
levels: one for a yes/no outcome, as in the simulation.

```{r public-studies}
files <- c(warfarin = "warfarin-model-fit.rds",
           theo_md = "theo-md-model-fit.rds",
           mad = "mad-model-fit.rds",
           case1_pkpd = "case1-pkpd-model-fit.rds",
           wbcSim = "wbcsim-model-fit.rds",
           mavoglurant = "mavoglurant-model-fit.rds",
           nimoData = "nimo-model-fit.rds",
           pheno_sd = "pheno-model-fit.rds",
           mixroute_sim = "mixroute-sim-model-fit.rds",
           onc_sim = "onc-sim-model-fit.rds")
rows <- lapply(names(files), function(study) {
  fit <- stored_fit(files[[study]])
  if (!inherits(fit, "pmx_fitted_model")) return(NULL)
  release <- model_release(fit)
  sizes <- release$arms$sizes
  arms <- names(sizes)
  # Each visit's share once, with the patients of every arm that has it.
  attendance <- vapply(seq_len(nrow(release$cells)), function(i) {
    p <- vapply(arms, function(arm) {
      as.numeric(release$visits[[arm]]$probability)[[i]]
    }, numeric(1))
    c(free = as.numeric(max(p) > 0 && max(p) < 1),
      behind = sum(sizes[p > 0]))
  }, numeric(2))
  levels <- vapply(seq_len(nrow(release$cells)), function(i) {
    marginals <- lapply(arms, function(arm) release$discrete[[arm]][[i]])
    held <- !vapply(marginals, is.null, logical(1))
    if (!any(held)) return(c(free = 0, behind = 0))
    p <- marginals[[which(held)[[1L]]]]$probability
    c(free = max(0, sum(p > 0 & p < 1) - 1), behind = sum(sizes[held]))
  }, numeric(2))
  fewest <- function(counts) {
    behind <- counts["behind", counts["free", ] > 0]
    if (length(behind)) min(behind) else NA
  }
  data.frame(study = study, patients = sum(sizes),
             random_effects = sum(vapply(release$pk_models, function(model) {
               nrow(model$parameters$omega)
             }, integer(1))),
             attendance_shares = sum(attendance["free", ]),
             fewest_behind_attendance = fewest(attendance),
             discrete_shares = sum(levels["free", ]),
             fewest_behind_discrete = fewest(levels))
})
do.call(rbind, rows[!vapply(rows, is.null, logical(1))])
```

A study with few patients and many shares between 0 and 1 sits among the
per-visit rows with the highest AUC, and a large study with a few such shares
among the lowest. For the population model, the cohort size places a study
among the rows of the second table; a study fitted with more random effects
than five gives the attacker more numbers to add up.

## What the Simulation Leaves Out

- **The attacker is strong.** Its reference is exact except where the error is
  set, and it holds the candidate's own trial measurements, which outside the
  study exist only where the same quantities are measured for the same patient:
  at a study site, in a related study, or in routine care.
- **The AUC is an average over members.** A patient far from the typical
  values, or with an unusual run of outcomes, moves the release further toward
  themselves and is more exposed than the average member, which is why
  membership-inference studies report the attack's success on such patients
  separately [2, 3].
- **The release is simplified.** The visits are exchangeable apart from their
  rates, the random effects are normal and independent across parameters, and
  the real release holds further numbers this simulation leaves out: the dosing
  model, the covariate summaries and the PD time courses.
- **The score is not the most powerful one possible.** It is of the kind used
  to trace individuals in pooled genetic data [4, 5, 6]. An attacker who can
  simulate studies from the same population can calibrate each candidate's
  score, as the strongest attacks on machine-learning models do [3], so these
  AUCs are a lower bound for an attacker holding what this one holds.

## The Simulation Code

Every function used above, as run:

```{r code, file="pmxmodel-privacy-simulation.R", eval=FALSE}
```

## References

1. Hanley JA, McNeil BJ. The meaning and use of the area under a receiver
   operating characteristic (ROC) curve. *Radiology.* 1982;143(1):29–36.
2. Shokri R, Stronati M, Song C, Shmatikov V. Membership inference attacks
   against machine learning models. *IEEE Symposium on Security and Privacy.*
   2017.
3. Carlini N, Chien S, Nasr M, Song S, Terzis A, Tramèr F. Membership
   inference attacks from first principles. *IEEE Symposium on Security and
   Privacy.* 2022.
4. Homer N, Szelinger S, Redman M, et al. Resolving individuals contributing
   trace amounts of DNA to highly complex mixtures using high-density SNP
   genotyping microarrays. *PLoS Genetics.* 2008;4(8):e1000167.
5. Sankararaman S, Obozinski G, Jordan MI, Halperin E. Genomic privacy and
   limits of individual detection in a pool. *Nature Genetics.*
   2009;41:965–967.
6. Dwork C, Smith A, Steinke T, Ullman J, Vadhan S. Robust traceability from
   trace amounts. *IEEE Symposium on Foundations of Computer Science (FOCS).*
   2015.
