---
title: "Evaluating the PMX model generator on public data"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Evaluating the PMX model generator on public data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_xgxr <- requireNamespace("xgxr", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
# Fitting compiles a model. Every fit below was built once by
# `scripts/build-model-fits.R` and is read from the installed package, so this
# article -- and the site that renders it -- needs neither nlmixr2 nor a
# compiler.
stored_fit <- function(file) {
  path <- system.file("extdata", file, package = "synpmx")
  if (nzchar(path)) readRDS(path) else NULL
}
has_fits <- !is.null(stored_fit("case1-pkpd-model-fit.rds"))
run <- has_nlmixr2data && has_xgxr && has_fits
plots <- run && has_ggplot2
# `compare_pmx_distributions()` composes its panels with patchwork and refuses
# to draw without it, so those chunks are guarded separately.
has_distributions <- has_ggplot2 &&
  requireNamespace("patchwork", quietly = TRUE)
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")
```

This article runs `synpmx_model()` over the same eight public datasets that
[Evaluating AVATAR on public data](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html)
and
[Evaluating PCA on public data](https://iamstein.github.io/synpmx/articles/pca-public-data-examples.html)
run their generators over, so the three can be read against each other on
identical studies.

This generator asks more of a study than the other two, and **what separates
these datasets is whether the study can identify a population model at all.**
Three requirements, each of which refuses rather than guesses:

- **A declared nominal grid**, as `synpmx_pca()` needs — the visit model is
  built on it.
- **At least twenty subjects.** A covariance matrix fitted to fewer describes
  those subjects rather than a population, so `min_subjects` refuses below that
  and says so.
- **An endpoint that behaves like a drug concentration**: absent before the
  first dose and rising with dose. Without one there is nothing to put a
  structural model on.

Five studies clear all three. `theo_md` clears the first and third and is shown
below the floor on purpose. `nimoData` and `pheno_sd` are refused, for
different reasons, and the refusals are the most useful thing in this article
if your own study looks like either.

Every fit below reads patient data once, through `synpmx_model_estimate()`.
Everything after that — every figure, every table, every scorecard — is
computed from the fit and never touches the source again.

<details>
<summary>Plotting and reporting helpers used throughout this vignette</summary>

```{r helpers}
# Observation rows only, source beside synthetic on a shared y axis, one row per
# endpoint. The same figure the PCA survey draws, so the two can be compared
# panel for panel.
overlay_plot <- function(source, synthetic, roles, title,
                         y_label = "DV", log_y = FALSE, alpha = 0.3,
                         max_time = Inf) {
  frame <- function(data, label) {
    observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0") &
      !is.na(data[[roles$dv]])
    if (!is.null(roles$mdv)) {
      observed <- observed & as.character(data[[roles$mdv]]) %in% c("0", "0.0")
    }
    observed <- observed & as.numeric(data[[roles$time]]) <= max_time
    data.frame(
      dataset = factor(label, levels = c("Source", "Synthetic")),
      subject = as.character(data[[roles$id]][observed]),
      occasion = if (is.null(roles$occasion)) "1" else
        as.character(data[[roles$occasion]][observed]),
      time = as.numeric(data[[roles$time]][observed]),
      dv = as.numeric(data[[roles$dv]][observed]),
      endpoint = if (is.null(roles$dvid)) "DV" else
        as.character(data[[roles$dvid]][observed]),
      stringsAsFactors = FALSE
    )
  }
  plotted <- rbind(frame(source, "Source"), frame(synthetic, "Synthetic"))
  figure <- ggplot2::ggplot(
    plotted,
    ggplot2::aes(time, dv,
                 group = interaction(dataset, subject, occasion),
                 colour = dataset)
  ) +
    ggplot2::geom_line(alpha = alpha) +
    ggplot2::geom_point(alpha = alpha, size = 0.7) +
    (if (length(unique(plotted$endpoint)) > 1L) {
      ggplot2::facet_grid(endpoint ~ dataset, scales = "free_y", switch = "y")
    } else {
      ggplot2::facet_wrap(~dataset)
    }) +
    ggplot2::scale_colour_manual(values = comparison_colours) +
    ggplot2::labs(x = "Time (hours)", y = y_label, title = title) +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "none", strip.placement = "outside")
  if (isTRUE(log_y)) figure + ggplot2::scale_y_log10() else figure
}

overlay_height <- function(data, roles, per_endpoint = 2.2, minimum = 3.4) {
  observed <- as.character(data[[roles$evid]]) %in% c("0", "0.0")
  endpoints <- if (is.null(roles$dvid)) 1L else
    length(unique(data[[roles$dvid]][observed]))
  max(minimum, per_endpoint * endpoints)
}

# Every dataset is run the same way and its numbers collected in one place, so
# the cross-dataset tables at the end cannot drift from the sections above them.
runs <- list()
model_run <- function(label, source, roles, file, seed) {
  fit <- stored_fit(file)
  synthetic <- synpmx_model_generate(fit, seed = seed)
  runs[[label]] <<- list(label = label, source = source, roles = roles,
                         fit = fit, synthetic = synthetic,
                         card = synpmx_scorecard(source, synthetic, roles))
  invisible(runs[[label]])
}

# Nearest-neighbour snapping onto a stated design grid, as in the PCA survey:
# which grid to snap to is a statement about the study and is written out at
# each call.
snap_to <- function(x, grid) {
  grid[max.col(-abs(outer(x, grid, "-")), ties.method = "first")]
}
```

</details>

## case1_pkpd: six arms and a censored endpoint

180 patients across six treatment arms, two endpoints keyed by a character
`NAME` column, a baseline weight and a `CENS` column. This is the study the
three demos share.

```{r case1, eval=run, message=FALSE}
case1_pkpd <- as.data.frame(
  get(utils::data(list = "case1_pkpd", package = "xgxr"))
)
# That study's CENS is meaningful only for PK: the PD effect is signed, so a
# left-censored PD row would report a value above the uncensored ones.
case1_pkpd$CENS <- ifelse(case1_pkpd$NAME == "PD - Continuous", 0,
                          case1_pkpd$CENS)
case1_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", cens = "CENS", amt = "AMT",
  evid = "EVID", cmt = "CMT", dvid = "NAME", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"), covariates = "WEIGHTB", keep = "STUDY"
)
case1 <- model_run("case1_pkpd", case1_pkpd, case1_roles,
                   "case1-pkpd-model-fit.rds", seed = 808)
case1$fit
```

```{r case1-plot, echo=FALSE, eval=plots, fig.height=overlay_height(case1_pkpd, case1_roles)}
overlay_plot(case1$source, case1$synthetic, case1_roles,
             "case1_pkpd: PK concentration and continuous PD, six arms",
             y_label = "LIDV", alpha = 0.2)
```

```{r case1-distributions, eval=run && has_distributions, fig.height=compare_pmx_distributions_height(case1_pkpd, case1_roles)}
compare_pmx_distributions(case1$source, case1$synthetic, case1_roles)
```

```{r case1-scorecard, eval=run}
synpmx_scorecard_datatable(case1$card)
```

Nothing fails. A5a reads observations per patient falling from 30.7 to 29,
which is the visit model drawing attendance rather than copying it, and D1
puts the concentration's standard deviation at 1.3 times the source's.
`vignette("pmxmodel-demo")` works this study end to end, including what its 46%
censoring and its PD endpoint cost.

Four rows read `not applicable` on every card in this article, so they are
worth reading once here. B1a, B1b and C2 need a run record that
`synpmx_avatar()` writes and this generator does not. B4a asks whether a
generated set of observation times copies a real one, which is a disclosure
question only where the set was taken from somebody: this generator decides
each visit independently from a per-arm probability, so a match is a
coincidence with a computable chance. **B4b is the row that carries the claim
here, and it is 0 on every dataset below**: no value any patient measured is
reproduced.

## mad: five endpoints, only one of them a concentration

60 subjects in a multiple-ascending-dose study with a declared `NOMTIME` and
five endpoints — PK concentration, continuous PD, and ordinal, count and binary
PD. The generator fits a structural model to the concentration and a shape to
each continuous PD endpoint; the discrete ones are a question this generator
does not answer.

```{r mad, eval=run, message=FALSE}
mad <- as.data.frame(get(utils::data(list = "mad", package = "xgxr")))
mad_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "LIDV", amt = "AMT", evid = "EVID",
  cmt = "CMT", dvid = "NAME", mdv = "MDV", nominal_time = "NOMTIME",
  strata = c("TRTACT", "DOSE"), covariates = c("WEIGHTB", "SEX")
)
mad_run <- model_run("mad", mad, mad_roles, "mad-model-fit.rds", seed = 909)
mad_run$fit
```

```{r mad-plot, echo=FALSE, eval=plots, fig.height=overlay_height(mad, mad_roles)}
overlay_plot(mad_run$source, mad_run$synthetic, mad_roles,
             "mad: five endpoints on one time axis", y_label = "LIDV",
             alpha = 0.2) +
  ggplot2::theme(strip.text = ggplot2::element_text(size = 7))
```

```{r mad-scorecard, eval=run}
synpmx_scorecard_datatable(mad_run$card)
```

Nothing fails, and A3 reads 5 of 5: every endpoint survives, including the
three discrete ones, which are drawn from the level frequencies their arm holds
at each visit rather than modelled. D1 is the only row to read, at 0.81 times
the source's spread on the concentration.

This is also the study that found a defect: a visit where every patient
recorded the same level of an ordinal endpoint left the draw with a single
level, and the draw read that level as a count of levels rather than as the
level itself. Fixed, and pinned by a regression test.

## warfarin: a single dose, and no dose levels to compare

32 subjects, a single oral dose, a PK endpoint (`cp`) and a PD one (`pca`) on
different time courses. Its 16 recorded observation times are the protocol's,
so declaring `nominal_time` from `time` states something true about this study
rather than constructing a grid.

```{r warfarin, eval=run, message=FALSE}
data("warfarin", package = "nlmixr2data")
warfarin <- as.data.frame(warfarin)
warfarin$ntime <- warfarin$time
warfarin_roles <- pmx_roles(
  id = "id", time = "time", nominal_time = "ntime", dv = "dv", amt = "amt",
  evid = "evid", dvid = "dvid", covariates = c("wt", "age", "sex")
)
warfarin_run <- model_run("warfarin", warfarin, warfarin_roles,
                          "warfarin-model-fit.rds", seed = 404)
model_report(warfarin_run$fit)
```

```{r warfarin-plot, echo=FALSE, eval=plots, fig.height=overlay_height(warfarin, warfarin_roles)}
overlay_plot(warfarin_run$source, warfarin_run$synthetic, warfarin_roles,
             "warfarin: concentration and prothrombin complex activity",
             alpha = 0.35)
```

```{r warfarin-distributions, eval=run && has_distributions, fig.height=compare_pmx_distributions_height(warfarin, warfarin_roles)}
compare_pmx_distributions(warfarin_run$source, warfarin_run$synthetic,
                          warfarin_roles)
```

```{r warfarin-scorecard, eval=run}
synpmx_scorecard_datatable(warfarin_run$card)
```

Read the design line in the report above rather than the parameters. It says
the median profile peaks at 9 h and that **31% of subjects do too** — 22 of the
32 are first sampled at 24 h, well past the peak, so most individual profiles
only decline. The route was decided from the pooled median rather than by a
per-subject vote for exactly that reason, and the low percentage is the fit
telling you how much it had to go on. `proportional` reads `NA` for the same
kind of reason: every patient gets the same dose, so there are no dose levels
to compare and the residual error falls back to additive.

Nothing fails; D1 at 1.2 times the source's spread on `cp` is the only row to
read.

## wbcSim: an infusion and a delayed response

45 subjects with infusion start/stop pairs and a delayed white-blood-cell
decline, nadir and recovery. The concentration this generator looks for is not
here: what is recorded is the response.

```{r wbc, eval=run, message=FALSE}
data("wbcSim", package = "nlmixr2data")
wbcSim <- as.data.frame(wbcSim)
wbcSim$NTIME <- wbcSim$TIME
wbc_roles <- pmx_roles(
  id = "ID", time = "TIME", nominal_time = "NTIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", rate = "RATE"
)
wbc_run <- model_run("wbcSim", wbcSim, wbc_roles, "wbcsim-model-fit.rds",
                     seed = 505)
model_report(wbc_run$fit)
```

```{r wbc-plot, echo=FALSE, eval=plots}
overlay_plot(wbc_run$source, wbc_run$synthetic, wbc_roles,
             "wbcSim: white blood cell count", max_time = 700)
```

```{r wbc-scorecard, eval=run}
synpmx_scorecard_datatable(wbc_run$card)
```

**This is the study where the generator's endpoint test is wrong, and the
scorecard only partly catches it.** `wbcSim` records a white blood cell count,
not a drug concentration. The test asks whether an endpoint is absent before
the first dose and rises and falls afterwards, and a delayed myelosuppression
answers yes to both, so a one-compartment infusion model was fitted to a cell
count. It converged, and the report says `1cmt_infusion` with a clearance of
0.012 as though that meant something.

The output shows it. A cell count recovers to its baseline between infusions
and a concentration decays toward zero, so the generated profiles run far below
the source's — a median of 0 to 3.8 against the source's 3.8 to 10.4 over the
same window — and the source's nadir at 216 h and its recovery are absent.
A4 reads 45 -> 37, because eight subjects drew no observations at all, and A5a
and A5b fall with it.

Nothing here `FAIL`s, which is the honest report of what the scorecard checks:
it asks whether the output is a legal dataset in the study's shape and whether
it copies anybody, not whether the structural model was the right one to fit.
`endpoint_roles` cannot help — the endpoint really is the one modelled. What
this study needs is a generator that does not assert a PK shape, and both
`synpmx_pca()` and `synpmx_avatar()` reproduce its nadir and recovery, as their
own surveys show.

## mavoglurant: an occasion-reset clock

120 subjects in one- and two-period profiles, with `TIME` resetting inside
`OCC` so it is already dose-relative. The grid is the one the PCA survey writes
down, reused unchanged.

```{r mavo, eval=run, message=FALSE}
data("mavoglurant", package = "nlmixr2data")
mavoglurant <- as.data.frame(mavoglurant)
mavo_design <- c(0, 0.25, 0.5, 0.75, 1, 1.5, 2, 3, 4, 6, 8, 10, 12, 24, 36, 48)
mavoglurant$NTIME <- ifelse(mavoglurant$EVID == 0,
                            snap_to(mavoglurant$TIME, mavo_design),
                            mavoglurant$TIME)
mavo_roles <- pmx_roles(
  id = "ID", time = "TIME", nominal_time = "NTIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", rate = "RATE", mdv = "MDV", occasion = "OCC",
  keep = "DOSE", covariates = c("AGE", "SEX", "WT", "HT")
)
mavo_run <- model_run("mavoglurant", mavoglurant, mavo_roles,
                      "mavoglurant-model-fit.rds", seed = 707)
mavo_run$fit
```

```{r mavo-plot, echo=FALSE, eval=plots}
overlay_plot(mavo_run$source, mavo_run$synthetic, mavo_roles,
             "mavoglurant: occasion-relative concentration profiles",
             log_y = TRUE, alpha = 0.25, max_time = 30)
```

```{r mavo-scorecard, eval=run}
synpmx_scorecard_datatable(mavo_run$card)
```

Three rows to read, and they are one finding. A5b reports occasions per patient
falling from 1.65 to 1, A5a observations per patient from 20.2 to 11.3, and D1
the concentration's spread at 0.34 of the source's. `mavoglurant` is one- and
two-period, and the dosing model carries one planned schedule per arm: the
second period is not in it, so the patients who had one do not get it back.
A study whose periods matter needs them declared as arms, or a generator that
keeps each patient's own schedule.

## theo_md: twelve subjects, which is below the floor

12 subjects, seven doses exactly 24 h apart, dense sampling around the first
and last dose. The grid is constructible and the endpoint is a concentration,
so the only thing standing between this study and a fit is the cohort size.

```{r theo-refusal, eval=has_nlmixr2data, error=TRUE}
data("theo_md", package = "nlmixr2data")
theo_md <- as.data.frame(theo_md)
theo_doses <- seq(0, 144, by = 24)
theo_samples <- c(0, 0.25, 0.5, 1, 2, 3, 4, 5, 7, 9, 12, 24)
interval <- pmax(1L, findInterval(theo_md$TIME, theo_doses))
theo_md$NTIME <- ifelse(
  theo_md$EVID == 0,
  theo_doses[interval] +
    snap_to(theo_md$TIME - theo_doses[interval], theo_samples),
  theo_md$TIME
)
theo_roles <- pmx_roles(
  id = "ID", time = "TIME", nominal_time = "NTIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", covariates = "WT"
)
synpmx_model_estimate(theo_md, theo_roles, seed = 1)
```

The floor is an argument rather than a constant, so it can be lowered where the
consequence is understood. The fit below was built with `min_subjects = 12L`,
and what it costs is stated in the row it fills in the closing tables: a
covariance matrix estimated from twelve subjects describes those twelve.

```{r theo, eval=run, message=FALSE}
theo_run <- model_run("theo_md", theo_md, theo_roles, "theo-md-model-fit.rds",
                      seed = 303)
theo_run$fit
```

```{r theo-plot, echo=FALSE, eval=plots}
overlay_plot(theo_run$source, theo_run$synthetic, theo_roles,
             "theo_md: seven Q24H oral doses", alpha = 0.35)
```

```{r theo-scorecard, eval=run}
synpmx_scorecard_datatable(theo_run$card)
```

Nothing fails, and the card looks like the others. That is the uncomfortable
part: the scorecard cannot see that a covariance matrix came from twelve
subjects, because the question it asks is whether this output copies anybody or
changed the study's shape, and a twelve-subject fit does neither. The floor
exists because nothing downstream of it can tell.

## nimoData: refused twice

12 subjects, ten roughly weekly infusions, with dose times recorded as actuals.
The grid construction is the one the AVATAR and PCA surveys work out, reused
unchanged — and the study is still refused, first for its cohort size and then,
with the floor lowered, for a second reason worth reading.

```{r nimo, eval=has_nlmixr2data, error=TRUE}
data("nimoData", package = "nlmixr2data")
nimoData <- as.data.frame(nimoData)
nimo_interval <- 168
last_occasion <- ave(nimoData$OCC, nimoData$ID, FUN = max)
nominal_tad <- round(nimoData$TAD / 24) * 24
pre_dose <- nimoData$EVID == 0 & nominal_tad >= nimo_interval &
  nimoData$OCC < last_occasion
nominal_tad[pre_dose] <- nimo_interval - 1
nimoData$NTIME <- (nimoData$OCC - 1) * nimo_interval +
  ifelse(nimoData$EVID == 0, nominal_tad, 0)
nimo_roles <- pmx_roles(
  id = "ID", time = "TIME", nominal_time = "NTIME", dv = "DV", amt = "AMT",
  evid = "EVID", rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
  covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
)
synpmx_model_estimate(nimoData, nimo_roles, seed = 1, min_subjects = 12L)
```

Every subject in `nimoData` receives the same dose, so there are no dose levels
to compare and nothing in the table says that this endpoint scales with dose.
The detection fails closed rather than assuming: `endpoint_roles = c(pk = "DV")`
names the concentration by hand where you know it is one, and the refusal is
the generator declining to make that statement on your behalf.

## pheno_sd: no grid and no dose-proportional endpoint

59 neonates on phenobarbital in routine care, with individualised dosing and
sparse irregular sampling. There is no protocol grid because there was no
protocol, which refuses this study for `synpmx_pca()` too, and the
concentration test fails for the same reason it fails on `nimoData`.

```{r pheno, eval=has_nlmixr2data, error=TRUE}
data("pheno_sd", package = "nlmixr2data")
pheno_sd <- as.data.frame(pheno_sd)
pheno_sd$NTIME <- pheno_sd$TIME     # a declaration this study cannot support
pheno_roles <- pmx_roles(
  id = "ID", time = "TIME", nominal_time = "NTIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", covariates = "WT"
)
synpmx_model_estimate(pheno_sd, pheno_roles, seed = 1)
```

`synpmx_avatar()` is the generator for this study, and the
[AVATAR evaluation](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html)
runs it.

## What the six runs held

```{r inventory, eval=run}
inventory <- do.call(rbind, lapply(runs, function(entry) {
  fit <- entry$fit
  data.frame(
    Dataset = entry$label,
    Patients = length(unique(entry$source[[entry$roles$id]])),
    Model = fit$structural,
    `Fixed effects` = paste(names(fit$parameters$fixed),
                            signif(fit$parameters$fixed, 3),
                            collapse = ", "),
    # A proportional error reports a coefficient of variation and an additive
    # one a standard deviation, so the field is named by the kind.
    Residual = paste(fit$parameters$residual$kind,
                     signif(if (is.null(fit$parameters$residual$cv))
                              fit$parameters$residual$sd else
                              fit$parameters$residual$cv, 3)),
    check.names = FALSE, stringsAsFactors = FALSE
  )
}))
knitr::kable(inventory, row.names = FALSE,
             caption = "What each fit carries out of its study.")
```

```{r verdicts, eval=run}
verdicts <- do.call(rbind, lapply(runs, function(entry) {
  card <- as.data.frame(entry$card)
  counts <- table(factor(card$verdict,
                         levels = c("pass", "review", "FAIL",
                                    "not applicable")))
  data.frame(Dataset = entry$label, as.list(counts),
             Failing = paste(card$check[card$verdict == "FAIL"],
                             collapse = ", "),
             check.names = FALSE, stringsAsFactors = FALSE)
}))
knitr::kable(verdicts, row.names = FALSE,
             caption = "Scorecard verdicts across the six runs.")
```

No card fails on any of the six. Four rows on each read `not applicable` for
the reasons given under `case1_pkpd`, and the rows that ask to be read are
`A5a`, `A5b` and `D1` — how many observations and occasions each patient kept,
and how far a spread moved.

## What is preserved, and what is not

Preserved on all six: the schema and event grammar, the cohort and arm sizes,
the nominal grid the study declared, one dose schedule per arm, the covariate
marginals, and the guarantee B4b measures — no value any patient measured is
reproduced.

Not preserved, with the study that shows each:

- **Visit-to-visit measurement noise.** Every profile is one structural model
  at a different parameter draw, so generated profiles are smoother than real
  ones. Visible on every dataset here.
- **Exposure-response.** The PD shapes are fitted to the pooled observations
  with no exposure term, so a synthetic patient's arm does not reach their
  response. `case1_pkpd` is the worked case, in the demo.
- **Per-patient dose schedules, and multiple periods.** One planned schedule
  per arm. `mavoglurant` is the worked case: occasions per patient fall from
  1.65 to 1.
- **Arm-specific censoring.** One parameter distribution evaluated at each
  arm's dose cannot reproduce six arm-specific fractions below the limit;
  `case1_pkpd`, again in the demo.
- **Any endpoint that is not a concentration.** `wbcSim` is the worked case,
  and the one to read before trusting this generator on a response variable.

## Where to go next

- `vignette("pmxmodel-algorithm")` — what each step of the generator does.
- `vignette("pmxmodel-demo")` — one study end to end, with the fit inspected.
- `vignette("pmxmodel-fingerprint")` — every quantity a fit carries out of a
  study.
- [Evaluating PCA on public data](https://iamstein.github.io/synpmx/articles/pca-public-data-examples.html)
  and
  [Evaluating AVATAR on public data](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html)
  — the same eight studies through the other two generators.
