---
title: "Privacy Protections in the PMX Model Generator"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_width: 9
    fig_height: 4.5
vignette: >
  %\VignetteIndexEntry{Privacy Protections in the PMX Model Generator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
# Every fit is read from the installed package, so this article fits nothing
# and needs neither nlmixr2 nor a compiler.
stored_fit <- function(file) {
  path <- system.file("extdata", file, package = "synpmx")
  if (nzchar(path)) readRDS(path) else NULL
}
fit <- stored_fit("warfarin-model-fit.rds")
misdosed_fit <- stored_fit("warfarin-misdosed-model-fit.rds")
extreme_fit <- stored_fit("warfarin-extreme-model-fit.rds")
run <- has_nlmixr2data && !is.null(fit) && !is.null(misdosed_fit) &&
  !is.null(extreme_fit)
plots <- run && has_ggplot2
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")

# Three significant digits on measured quantities, written out rather than
# with a shared exponent, since one column can hold a clearance beside a
# variance a thousand times smaller. Counts are exempt.
digits3 <- function(x) {
  x <- as.data.frame(x)
  attr(x, "who") <- NULL
  for (column in names(x)) {
    if (is.numeric(x[[column]]) && !column %in% c("patients")) {
      x[[column]] <- vapply(x[[column]], function(v) {
        format(signif(v, 3), scientific = FALSE, drop0trailing = TRUE,
               trim = TRUE)
      }, character(1))
    }
  }
  x
}
```

The pharmacometric (PMX) model generator, `synpmx_model()`, fits a population
pharmacokinetic (PK) and pharmacodynamic (PD) model to a study inside the
environment that holds it, and generates synthetic patients from the fit. This
article sets out what of the study leaves that environment, how each released
number is protected, and five checks that measure the protections, run on
public studies and on three cases built to need them.

**No formal privacy guarantee is made.** Nothing here bounds what an adversary
can learn in the sense of differential privacy (DP), a property of a release
procedure that limits how much any one person's data can change what is
released. The protections below reduce what is released and measure how much
one patient moves it. A release that has to survive a determined adversary
needs a DP mechanism, and the last section says what one would take for this
generator.

## Why Releasing Only Parameters Is Not Enough

A fitted model holds no patient rows, but the DP literature shows that
statistics computed from patients can still disclose them. Four mechanisms
matter here.

**Membership inference** decides whether a given person was in the study. The
adversary holds that person's data and the release, and asks whether the
release fits them better than it would fit someone who was not in the study.
Homer and colleagues did this from published allele frequencies alone [1],
after which public access to that kind of aggregate genetic data was
withdrawn. The power of the attack grows with the number of released
statistics, with their precision, and as the number of people behind them
falls [2, 3].

**Model inversion** infers a hidden attribute from an estimated relationship.
Fredrikson and colleagues took a published warfarin dosing regression, a
patient's demographics and that patient's stable dose, and inferred the
patient's genotype nearly as well as a model built to predict genotype [4].

**Reconstruction** recovers individual records from enough accurate aggregate
answers: answering too many questions too precisely gives the dataset away
[5].

**A single patient's value** is the simplest case. A minimum, a maximum, the
median of an odd number of patients, and the most common value where every
value is held once are each one patient's value, whatever the summary is
called. Releasing one releases that patient's value.

Three properties of a summary lower all four risks: fewer released numbers,
coarser released numbers, and numbers no single patient can move far. Each
protection below is one of these.

## What Leaves the Study

A fit carries two kinds of content. Generation reads one of them. The other is
diagnostics for whoever ran the fit, and stays with it. `model_release()`
builds the part generation reads from an allowlist, and
`synpmx_model_generate()` attaches that part, never the fit, to the data it
returns. The worked example is `nlmixr2data::warfarin`: 32 patients, one oral
dose, a concentration and a prothrombin activity.

```{r release, eval=run}
release <- model_release(fit)
names(release)
setdiff(names(fit), names(release))
```

How many numbers each released field holds. Membership inference gains power
with this count, and most of it is the visit and dosing apparatus rather than
the population model, which is in `pk_models` and `pd`. The top-level
`parameters` repeat the first PK model's and are left out of the count.

```{r count, eval=run}
count_numbers <- function(x) {
  if (is.null(x) || is.function(x) || is.environment(x)) return(0L)
  if (is.data.frame(x)) {
    return(sum(vapply(x, function(column) {
      if (is.numeric(column)) sum(is.finite(column)) else 0L
    }, integer(1))))
  }
  if (is.list(x)) return(sum(vapply(x, count_numbers, integer(1))))
  if (is.numeric(x)) sum(is.finite(x)) else 0L
}
counts <- vapply(unclass(release)[setdiff(names(release), "parameters")],
                 count_numbers, integer(1))
data.frame(field = names(counts), numbers = as.integer(counts),
           row.names = NULL)[counts > 0, ]
```

The visit grid's nominal times and the planned dose times are the protocol's,
not measurements, and are public in the sense the protocol is.

## The Protections

Each is applied by `synpmx_model_estimate()` without being asked for, and each
names the work it rests on.

### 1. An Allowlist Decides What Leaves

`model_release()` copies the fields Step 6 of the
[algorithm](https://iamstein.github.io/synpmx/articles/pmxmodel-algorithm.html)
reads, and inside each only the parts it reads. A field added to the fit later
stays behind unless it is added to the list. The pattern is the
glimpse-then-simulate split of Novartis's `synadam` package [13], which writes
the summary that synthetic data is generated from to one file a reviewer can
read.

### 2. No Per-Patient Quantity

The empirical Bayes estimates (EBEs), each patient's individual random effects,
are read twice at estimation, for the covariate correlations and for the
influence reading in protection 8, and dropped. Generation draws new random
effects from the between-subject covariance matrix.

### 3. No Identifier

Synthetic subject IDs are minted at generation. No source label is kept,
including the levels of an ID column stored as a factor. A numeric ID keeps
only the largest source ID rounded up to a power of ten, which stops a
synthetic ID colliding with a real one without storing the real one. `synadam`
refuses to run when an ID column is left undeclared, for the same reason [13].

### 4. No Single Patient's Value

Every summary that could have been one patient's value is replaced:

| Quantity | Would have been | Is |
|---|---|---|
| Emission floor, where no assay limit is declared | half the smallest value reported | half the lowest value at least three patients reached, rounded down to 1, 2 or 5 times a power of ten |
| Allometric reference weight | the cohort median | the median to one significant figure |
| Planned dose amount or infusion rate, where no value is shared | the most common value, which is then one patient's | the arm's mean to two significant figures |
| Continuous covariate | a mean and standard deviation (SD) over every patient, and a median | a trimmed mean and SD (protection 6), no median |
| Endpoint type description | a count of levels and the lowest and highest value observed | the type alone |

Statistical disclosure control (SDC), the field that protects published
statistical tables, treats extreme values the same way, by top-coding and
coarsening them [9].

### 5. Several Patients Behind Every Frequency

A frequency resting on one patient is that patient. The threshold rule of SDC
publishes a count only where at least *k* units stand behind it [9], and
`synadam` masks values held once [13]. Here *k* is 3 by default:

- arms below `min_arm_patients` are dropped before anything is fitted;
- a nominal visit slot, a dose level and a planned cycle are kept only where at
  least `min_arm_patients` patients reached them;
- a categorical covariate level, and a binary or ordinal endpoint's level at a
  visit, below `min_category_patients` patients is excluded, or folded into the
  most common level.

### 6. Covariate Summaries Without Their Extremes

A continuous covariate is summarized by a mean and SD taken after the highest
and lowest 5% of patients are set aside, at least one at each end. The SD of
what remains understates the spread, because the tails are what was removed,
so it is divided by the SD a standard normal keeps under the same trimming.
With a share $\alpha$ removed from each end and $z = \Phi^{-1}(1-\alpha)$,

$$\mathrm{SD} = \frac{\mathrm{SD}_\text{trimmed}}{\sqrt{1 - \dfrac{2 z \phi(z)}{1-2\alpha}}}.$$

This is the trimmed mean of robust statistics [11]. Trimming is also how DP
mean estimators bound one person's influence [12], since no single value can
then move the result by more than the trimmed range allows.

### 7. Two Significant Figures

Every released estimate is rounded to two significant figures: PK fixed
effects, between-subject variances, residual error, PD shape parameters and
covariate summaries. Rounding is the other standard coarsening of SDC [9].
Leaving out a typical patient moves a typical value by less than one rounding
step, so for most patients the released number is identical with or without
them. It does not hide an outlier's effect on a variance, which is what
protection 8 measures.

### 8. How Far One Patient Moves Each Estimate

DP calls the most one person can move a released number its *sensitivity*,
taken over every possible dataset [6]. The same question asked of the study at
hand, leaving each patient out in turn and recording the largest move, is
*local sensitivity*, and in regression the same idea is Cook's distance [10].
It is computed at estimation without refitting:

- A population typical value on the log scale sits close to the mean of the
  patients' individual log parameters, so leaving out patient $i$ moves it by
  about $\eta_i/(n-1)$, where $\eta_i$ is that patient's random effect.
- The between-subject variance sits close to the mean of the squared random
  effects, so leaving out patient $i$ scales it by
  $\frac{(S-\eta_i^2)/(n-1)}{S/n}$, with $S=\sum_j \eta_j^2$.
- The PD baseline and its spread are recomputed from each subject's own
  baseline with the patient left out, and so is each covariate summary. Neither
  needs an approximation.

The largest move over all patients is kept for each estimate: in percent for a
typical value, and in points of the between-subject SD on the log scale, close
to a coefficient of variation in percent, for a spread. A move of 15 is a
review; a move of 30 fails, warns at estimation, and names the patient on the
console. The name is printed in the environment that holds the study and is
not stored: the fit keeps the largest move per estimate, never anything per
patient. On a parameter whose between-subject SD is 0.27, a patient 3 SD from
the mean in a cohort of 32 moves the typical value by about 3% and the spread
by about 4 points. Shrinkage pulls every random effect toward zero and
makes the reading understate the move, so it is reported beside it.

## The Five Checks

`model_privacy_checks()` runs five checks on a fit, each with its pass
criterion. The first four ask whether the release holds something about one
patient that it should not. The fifth is the reading of protection 8.

| Check | Question | Passes when |
|---|---|---|
| P1 | No per-patient table is released | the release holds only fields on the allowlist, none of them per patient |
| P2 | No source identifier is released | no source ID label, and no source ID value as the synthetic ID offset |
| P3 | No single patient's value is released | floors on the 1-2-5 series, the reference weight rounded, no covariate median |
| P4 | Every released frequency rests on several patients | every arm holds at least `min_arm_patients` patients |
| P5 | No single patient moves a released estimate far | every estimate moves less than 15 |

P1 to P4 are structural and read the release. They pass by construction on a
fit from this version, and exist to catch a release assembled some other way
or a fit stored before the protections existed. P5 is read from the fit, and a
release carries its verdict.

```{r checks, eval=run}
checks <- model_privacy_checks(fit)
as.data.frame(checks)[, c("check", "verdict", "result")]
```

The reading behind P5 is the `influence` attribute of the full fit's checks,
one row per released estimate: the released value, the largest move any one
patient causes, its unit, the number of patients it was read over, the
shrinkage of the random effect it came from, and the verdict. No row is about
a patient.

```{r influence, eval=run}
digits3(attr(checks, "influence"))
```

The same checks on the release report P5 as a verdict only. A passing release
states that every estimate moved less than 15, and nothing more, so the one
data-dependent thing it carries about the reading is the verdict.

```{r release-checks, eval=run}
as.data.frame(model_privacy_checks(release))[, c("check", "verdict", "result")]
```

## Three Cases Built to Need the Checks

Each is `warfarin` with one or more patients changed, built once by
`synpmx_model_estimate()` and stored with the package. None is a real study.

```{r constructed-data, eval=run}
warfarin <- as.data.frame(nlmixr2data::warfarin)
warfarin$ntime <- warfarin$time
roles <- pmx_roles(
  id = "id", time = "time", nominal_time = "ntime", dv = "dv", amt = "amt",
  evid = "evid", dvid = "dvid", covariates = c("wt", "age", "sex")
)
patients <- sort(unique(warfarin$id))

misdosed <- warfarin
on_cp <- misdosed$id == patients[[5L]] & misdosed$evid == 0 &
  misdosed$dvid == "cp"
misdosed$dv[on_cp] <- misdosed$dv[on_cp] * 1000

extreme <- warfarin
extreme$wt[extreme$id == patients[[3L]]] <- 250
extreme$age[extreme$id == patients[[8L]]] <- 95
extreme$sex <- as.character(extreme$sex)
extreme$sex[extreme$id == patients[[12L]]] <- "not recorded"
extreme$sex <- factor(extreme$sex)
```

```{r constructed-fits, eval=FALSE}
misdosed_fit <- synpmx_model_estimate(misdosed, roles, seed = 1)
extreme_fit <- synpmx_model_estimate(extreme, roles, seed = 1)
```

### A Patient Given a Thousand Times the Recorded Dose

One patient received a thousand times the dose the record shows. A linear
model sees concentrations a thousand times higher than the recorded dose
implies. Estimation fits the patient anyway, and the variance terms absorb the
excess. The estimation call warned that one patient moved a released estimate by
30 or more, and named the patient. The stored fit names nobody.

```{r misdosed-checks, eval=run}
misdosed_checks <- model_privacy_checks(misdosed_fit)
as.data.frame(misdosed_checks)[, c("check", "verdict", "result")]
digits3(attr(misdosed_checks, "influence"))
```

The synthetic data shows what the fit absorbed. The misdosed patient's
exposure has become between-subject variability, so the synthetic patients
spread wider than the rest of the source cohort does:

```{r misdosed-plot, eval=plots, fig.height=4}
synthetic <- suppressWarnings(
  synpmx_model_generate(misdosed_fit, seed = 11))
concentration <- function(data, label) {
  rows <- data$evid == 0 & data$dvid == "cp" & data$dv > 0
  data.frame(dataset = label, id = as.character(data$id[rows]),
             time = data$time[rows], dv = data$dv[rows])
}
plotted <- rbind(concentration(misdosed, "Source"),
                 concentration(synthetic, "Synthetic"))
ggplot2::ggplot(plotted, ggplot2::aes(time, dv, group = id,
                                      colour = dataset)) +
  ggplot2::geom_line(alpha = 0.4) +
  ggplot2::facet_wrap(~dataset) +
  ggplot2::scale_y_log10() +
  ggplot2::scale_colour_manual(values = comparison_colours) +
  ggplot2::labs(x = "Time (hours)", y = "Warfarin concentration") +
  ggplot2::theme_minimal() +
  ggplot2::theme(legend.position = "none")
```

The scorecard reads the release's verdict as row E3, so the same finding
reaches anyone who scores the synthetic data:

```{r misdosed-card, eval=run}
card <- synpmx_scorecard(misdosed, synthetic, roles)
as.data.frame(card)[card$check %in% c("E1", "E2", "E3"),
                    c("check", "result", "verdict")]
```

The remedy is to correct the data: fix the patient's dosing record, or leave
the patient out, and fit again.

### Extreme Covariates

One patient weighs 250 kg and another is 95 years old. Neither enters the
covariate summaries, because each is the most extreme patient on their side
and trimming sets them aside:

```{r extreme-covariates, eval=run}
summary_of <- function(model) {
  data.frame(covariate = c("wt", "age"),
             geometric_mean = exp(c(model$wt$meanlog, model$age$meanlog)),
             sd_log = c(model$wt$sdlog, model$age$sdlog))
}
rbind(cbind(study = "warfarin", summary_of(fit$covariates)),
      cbind(study = "with the two extreme patients",
            summary_of(extreme_fit$covariates)))
```

The same two summaries taken over every patient, as they were before the
trimming, move with the extreme patients:

```{r untrimmed, eval=run}
first <- function(data) data[!duplicated(data$id), ]
untrimmed <- function(data) {
  data.frame(covariate = c("wt", "age"),
             geometric_mean = exp(c(mean(log(first(data)$wt)),
                                    mean(log(first(data)$age)))),
             sd_log = c(stats::sd(log(first(data)$wt)),
                        stats::sd(log(first(data)$age))))
}
digits3(rbind(cbind(study = "warfarin", untrimmed(warfarin)),
              cbind(study = "with the two extreme patients",
                    untrimmed(extreme))))
```

P5 reads the covariate rows of the influence table: leaving out any one
patient barely moves them, extreme patients included.

```{r extreme-influence, eval=run}
extreme_influence <- attr(model_privacy_checks(extreme_fit), "influence")
digits3(extreme_influence[extreme_influence$group == "covariate", ])
```

### A Category Nobody Else Holds

A third changed patient has a sex recorded as a level no other patient holds.
The level is excluded from the covariate distribution and from the schema the
generated table is written in, so no synthetic patient can carry it:

```{r rare-level, eval=run}
extreme_fit$covariates$sex$levels
levels(extreme_fit$schema$prototypes$sex)
```

## Across the Public Studies

The checks on every stored fit of the
[public-data evaluation](https://iamstein.github.io/synpmx/articles/pmxmodel-public-data-examples.html),
and the estimate each study's most influential patient moves furthest.

```{r survey, eval=run}
files <- c(warfarin = "warfarin-model-fit.rds",
           theo_md = "theo-md-model-fit.rds",
           mad = "mad-model-fit.rds",
           case1_pkpd = "case1-pkpd-model-fit.rds",
           wbcSim = "wbcsim-model-fit.rds",
           mavoglurant = "mavoglurant-model-fit.rds",
           nimoData = "nimo-model-fit.rds",
           pheno_sd = "pheno-model-fit.rds",
           mixroute_sim = "mixroute-sim-model-fit.rds",
           onc_sim = "onc-sim-model-fit.rds")
rows <- lapply(names(files), function(study) {
  one <- stored_fit(files[[study]])
  if (!inherits(one, "pmx_fitted_model")) return(NULL)
  checks <- model_privacy_checks(one)
  verdicts <- stats::setNames(checks$verdict, checks$check)
  data.frame(study = study, patients = one$n_source, t(verdicts),
             largest = checks$result[checks$check == "P5"],
             check.names = FALSE)
})
do.call(rbind, rows[!vapply(rows, is.null, logical(1))])
```

A `review` there is a study in which one patient moves an estimate by 15 or
more, and the influence table of its fit says which estimate. A move is a
fraction of the cohort's spread divided by the cohort size, so the same patient
moves a twelve-patient fit about five times as far as a sixty-patient one.

## Checks on the Synthetic Table

The checks above read the fit. `synpmx_scorecard()` reads the synthetic table
against its source, and its section B asks whether any synthetic patient
reproduces a real one. Four of its rows do not apply to this generator,
because no synthetic patient is assembled from a real patient's record; B3,
B4b and B5 do.

```{r card, eval=run}
synthetic <- synpmx_model_generate(fit, seed = 11)
card <- synpmx_scorecard(warfarin, synthetic, roles)
as.data.frame(card)[substr(card$check, 1, 1) %in% c("B", "E"),
                    c("check", "question", "result", "verdict")]
```

B3 compares how close synthetic patients sit to real ones against how close
real patients sit to each other, B4b asks whether any generated vector of
values copies a real patient's, and B5 asks whether a level fewer than three
real patients held reached the output.
[The scorecard](https://iamstein.github.io/synpmx/articles/scorecard.html)
documents each row and its pass criterion.

## What These Checks Do Not Establish

- **An adversary who knows every other patient is not stopped.** Such an
  adversary can compute the exact effect of the one patient they do not know,
  so any move at all is detectable, however far under a threshold it sits.
  That adversary is the one DP is defined against.
- **The influence reading describes this study only.** How far one patient
  moves an estimate is itself a fact about the data, which is why DP calibrates
  its noise to *smooth* sensitivity rather than local sensitivity [7].
  Shrinkage makes the reading understate the move.
- **Not every estimate is read.** The residual error, a parameter estimated
  without a between-subject term (bioavailability in a mixed-route model), and
  the PD shape parameters other than the baseline have no row.
- **Per-arm rates and attendance can rest on fewer than three patients'
  events.** A dose reduction one patient had makes its arm's reduction rate
  positive. These are treated as trial-level realized design, which a clinical
  study report tabulates in the same form.
- **Columns declared in `keep` are copied from one patient per arm.** Declare
  there only values that are constant within an arm.
- **Rarity in the world is not measured.** A covariate level held by many
  patients in this study can still identify someone if few people alive hold
  it.
- **Releases compose.** Two releases from overlapping data disclose more than
  either alone, and nothing here accounts for that.

## What a Formal Guarantee Would Take

The DP method that fits a generator built on `nlmixr2` is
subsample-and-aggregate [7, 8], because it treats the estimator as a black box:

1. Split the patients into *k* disjoint groups and fit the population model in
   each.
2. Clip each group's estimate of each log parameter to a public range of width
   *S*.
3. Release the mean over the groups plus Laplace noise of scale
   $b = dS/(\varepsilon k)$ per parameter, where *d* parameters share the
   privacy budget $\varepsilon$.

One patient is in one group, so the guarantee holds whatever the fit inside a
group does. The textbook DP methods for regression perturb a convex loss with
bounded gradients [14], and a population likelihood is neither, which rules
them out here. The cost is the error law of the
[privacy background](https://iamstein.github.io/synpmx/articles/privacy-background.html)
with the cohort size replaced by the number of groups. At about twenty patients
per group, and a median Laplace error of $b \ln 2$:

```{r subsample-aggregate, eval=TRUE}
grid <- expand.grid(patients = c(60, 200, 1000), released = c(3, 11))
grid$groups <- grid$patients / 20
scale <- grid$released * log(10) / (1 * grid$groups)
grid$median_fold_error <- signif(exp(log(2) * scale), 3)
grid
```

That is with $\varepsilon = 1$ and a public range spanning tenfold for each
parameter. With all eleven parameters of a two-compartment model released, the
error stays above 40% even at a thousand patients; with three released and the
rest taken from a public prior, it reaches about 10% there. At the cohort sizes this generator usually sees, the
[calibrated generator's evaluation](https://iamstein.github.io/synpmx/articles/calibrated-public-data-examples.html)
finds that a release is worse than generating from the public prior alone.
Subsample-and-aggregate is not implemented.

## References

1. Homer N, Szelinger S, Redman M, et al. Resolving individuals contributing
   trace amounts of DNA to highly complex mixtures using high-density SNP
   genotyping microarrays. *PLoS Genetics.* 2008;4(8):e1000167.
2. Sankararaman S, Obozinski G, Jordan MI, Halperin E. Genomic privacy and
   limits of individual detection in a pool. *Nature Genetics.*
   2009;41:965–967.
3. Dwork C, Smith A, Steinke T, Ullman J, Vadhan S. Robust traceability from
   trace amounts. *IEEE Symposium on Foundations of Computer Science (FOCS).*
   2015.
4. Fredrikson M, Lantz E, Jha S, Lin S, Page D, Ristenpart T. Privacy in
   pharmacogenetics: an end-to-end case study of personalized warfarin dosing.
   *USENIX Security Symposium.* 2014.
5. Dinur I, Nissim K. Revealing information while preserving privacy.
   *ACM Symposium on Principles of Database Systems (PODS).* 2003.
6. Dwork C, McSherry F, Nissim K, Smith A. Calibrating noise to sensitivity in
   private data analysis. *Theory of Cryptography Conference (TCC).* 2006.
7. Nissim K, Raskhodnikova S, Smith A. Smooth sensitivity and sampling in
   private data analysis. *ACM Symposium on Theory of Computing (STOC).* 2007.
8. Smith A. Privacy-preserving statistical estimation with optimal
   convergence rates. *ACM Symposium on Theory of Computing (STOC).* 2011.
9. Hundepool A, Domingo-Ferrer J, Franconi L, et al. *Statistical Disclosure
   Control.* Wiley; 2012.
10. Cook RD. Detection of influential observation in linear regression.
    *Technometrics.* 1977;19(1):15–18.
11. Huber PJ, Ronchetti EM. *Robust Statistics.* 2nd ed. Wiley; 2009.
12. Bun M, Steinke T. Average-case averages: private algorithms for smooth
    sensitivity and mean estimation. *Advances in Neural Information
    Processing Systems (NeurIPS).* 2019.
13. Novartis. `synadam`: generate synthetic ADaM datasets. R package version
    0.3.4. <https://github.com/Novartis/synadam>.
14. Chaudhuri K, Monteleoni C, Sarwate AD. Differentially private empirical
    risk minimization. *Journal of Machine Learning Research.*
    2011;12:1069–1109.
