---
title: "Evaluating prior-only generation on public data"
author: "Andrew Stein"
output:
  rmarkdown::html_vignette:
    toc: true
    fig_width: 9
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{Evaluating prior-only generation on public data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
library(synpmx)
has_nlmixr2data <- requireNamespace("nlmixr2data", quietly = TRUE)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
run <- has_nlmixr2data
plots <- run && has_ggplot2
has_distributions <- has_ggplot2 &&
  requireNamespace("patchwork", quietly = TRUE)
comparison_colours <- c(Source = "#1B6CA8", Synthetic = "#D95F02")
```

`synpmx_prior()` reads no data, so this article measures something different
from the other three surveys. There, a poor comparison is the generator's
fault. Here every mismatch is the asserted model's fault, and the number that
matters is how far the output sits from the study the model was meant to
describe.

**Read this as a measurement of elicitation, not of an algorithm.** The
generator has no parameters to tune and nothing to learn. Ask it for a study
under a correct model and it produces one; ask under a model that is twice
wrong and it produces something twice wrong, quietly and without complaint.
The two runs below bracket that, and the sensitivity section puts a number on
the slope between them.

```{r data, eval=run}
data("theo_md", package = "nlmixr2data")
theo_md <- as.data.frame(theo_md)
mixroute <- as.data.frame(mixroute_sim)
```

## Why two datasets and not ten

The other surveys run the same generator across every public study in the
package's collection. That is not available here, because prior-only
generation needs the protocol stated as a `pmx_trial_design()`, and most of
these studies have a design a public declaration cannot express. The
obstruction is almost always the dosing rather than the endpoint.

| Dataset | Distinct dose amounts | What a public design cannot state |
|---|---|---|
| `mixroute_sim` | 1 | The two administration routes; `1cmt_mixed` is refused because a design cannot say which dose went by which route |
| `theo_md` | 1 | — |
| `case1_pkpd` | 5 levels | A placebo arm (`dose_levels` must be positive), and samples at negative nominal times before the first dose |
| `mad` | 5 levels | The same placebo arm and pre-dose samples, plus four of its five endpoints: ordinal, count and binary have no form |
| `mavoglurant` | 3 levels, 4 infusion rates | A per-subject infusion rate |
| `nimoData` | 4 | A per-subject infusion rate, over ten infusions recorded as actuals |
| `warfarin` | 20 | A per-subject dose amount, and its PD endpoint is turnover |
| `pheno_sd` | 54 | A per-subject dose amount, in a neonatal cohort |
| `wbcSim` | 28 | A per-subject dose amount, and myelosuppression is an ODE model |
| `onc_sim` | — | Dosing compressed into `ADDL`/`II`, a dose that changes within a patient from that patient's own response, and an ODE tumour model |

Two studies have a single protocol dose that a public design states exactly,
and they are the two below. Everything else would need a dose amount invented
per subject, which puts the invention rather than the model under test.

That table is also the quickest way to answer "can I use this mode on my
study?". A protocol dose and a stated sampling schedule are what it needs; a
study whose dose follows the patient is a study for `synpmx_avatar()` or
`synpmx_model()`.

## Shared workflow

Both examples follow the same five steps:

1. declare the column meanings with `pmx_roles()`, which for this mode names
   the schema to generate **into** rather than one to read;
2. state the public model and design, with where each came from;
3. generate with `synpmx_prior()`;
4. plot the real and synthetic observations together;
5. report the scorecard and the ratio of median concentrations.

`synpmx_scorecard_datatable(card, report = "minimal")` prints the tally and
only the rows that are not a pass, which is the same call the other surveys
use. Five rows read `not applicable` on both cards below, for two different
reasons. **B1a**, **B1b** and **C2** read a run record that `synpmx_avatar()`
writes and this generator does not. **E1** and **E2** ask whether a fitted
model's parameters moved off their starting values and whether its
between-subject terms were estimated — questions for `synpmx_model()`, and
this mode fits nothing. Both are limitations of those rows on this generator
rather than properties of either dataset, and they read the same on both
cards.

Note which questions are *not* marked inapplicable. **B4a** and **B4b** ask
directly whether a generated time vector or value vector copies an exposed
real one, and both are scored here as on any other card. Both pass at zero,
which is the answer a mode that reads nothing has to give — and a card that
skipped the question could not have shown it.

```{r helpers, eval=run}
# Observation rows only, source beside synthetic, on a shared axis.
observed <- function(data, roles, label) {
  keep <- data[[roles$evid]] == 0
  value <- suppressWarnings(as.numeric(data[[roles$dv]][keep]))
  data.frame(
    set = label,
    subject = paste(label, data[[roles$id]][keep]),
    time = suppressWarnings(as.numeric(data[[roles$time]][keep])),
    dv = value
  )[is.finite(value) & value > 0, ]
}

# The number this article is about: where the asserted model put the level,
# against where the study has it.
level_ratio <- function(source, synthetic, roles) {
  med <- function(d) {
    v <- suppressWarnings(as.numeric(d[[roles$dv]][d[[roles$evid]] == 0]))
    stats::median(v[is.finite(v) & v > 0])
  }
  c(source = med(source), synthetic = med(synthetic),
    ratio = med(synthetic) / med(source))
}
```

## mixroute_sim: the model that generated it is public

This fixture's own documentation states the parameters it was simulated from —
clearance 2 L/day, volume 10 L, absorption 0.5 /day and bioavailability 0.7,
with clearance and volume scaled allometrically on weight. Handing those back
as the public model is the best case this mode can have: the prior is not an
estimate or a guess, it is the truth, and whatever gap remains is the part of
the truth the catalogue cannot hold.

Three things it cannot hold here. The allometric scaling, because no covariate
reaches a parameter. The bioavailability, because `1cmt_oral` has no `f` and
the mixed form that does is refused for a public design. And the three arms,
because one asserted model covers one route.

```{r mixroute-inputs, eval=run}
mixroute_roles <- pmx_roles(
  id = "ID", time = "TIME", nominal_time = "NTIME", dv = "DV", amt = "AMT",
  evid = "EVID", cmt = "CMT", cens = "CENS", covariates = "WT"
)
mixroute_model <- pmx_structural_model(
  pk = "1cmt_oral", typical = c(cl = 2, v = 10, ka = 0.5),
  source = "the documented generating truth of the mixroute_sim fixture"
)
mixroute_design <- pmx_trial_design(
  dose_levels = 100, cohort_sizes = 90, sampling = c(0, 1, 2, 3, 7),
  n_doses = 3, dose_interval = 7,
  source = "the fixture's protocol: 100 mg on days 0, 7 and 14"
)
mixroute_covariates <- pmx_covariates(WT = pmx_covariate(
  range = c(40, 130), median = 72, cv = 0.17,
  source = "the fixture's documented weight distribution"
))
mixroute_syn <- synpmx_prior(
  mixroute_model, mixroute_design, mixroute_roles,
  n_subjects = 90, seed = 1, covariates = mixroute_covariates
)
```

```{r mixroute-figure, eval=plots, fig.height=4}
frame <- rbind(observed(mixroute, mixroute_roles, "Source"),
               observed(mixroute_syn, mixroute_roles, "Synthetic"))
ggplot2::ggplot(frame, ggplot2::aes(time, dv, group = subject, colour = set)) +
  ggplot2::geom_line(alpha = 0.3) +
  ggplot2::geom_point(alpha = 0.35, size = 0.7) +
  ggplot2::facet_wrap(~ set) +
  ggplot2::scale_colour_manual(values = comparison_colours, guide = "none") +
  ggplot2::labs(x = "Study time (days)", y = "Concentration (mg/L)",
                title = "mixroute_sim under its own generating model") +
  ggplot2::theme_minimal()
```

```{r mixroute-level, eval=run}
round(level_ratio(mixroute, mixroute_syn, mixroute_roles), 3)
```

```{r mixroute-card, eval=run}
synpmx_scorecard_datatable(
  synpmx_scorecard(mixroute, mixroute_syn, mixroute_roles), report = "minimal"
)
```

The median sits within about a quarter of the study's, which flatters it. The
figure shows why, and so does the scorecard's one substantive review row: the
synthetic profiles are tighter than the real ones. One asserted model gives
every subject the same route, and the route is what separates this study's
arms.

```{r mixroute-arms, eval=run}
spread <- function(data) {
  v <- suppressWarnings(as.numeric(data$DV[data$EVID == 0]))
  v <- v[is.finite(v) & v > 0]
  c(median = stats::median(v), sd = stats::sd(v),
    p90 = stats::quantile(v, 0.9, names = FALSE))
}
by_arm <- tapply(mixroute$DV[mixroute$EVID == 0],
                 mixroute$ARM[mixroute$EVID == 0],
                 function(v) stats::median(v, na.rm = TRUE))
round(by_arm, 2)
round(rbind(source = spread(mixroute), synthetic = spread(mixroute_syn)), 2)
```

The three arms sit a factor of two and a half apart, because an intravenous
dose is all of the dose and a subcutaneous one is seventy per cent of it. The
synthetic cohort lands between them — it is the average of arms it has no way
to tell apart — and its spread is roughly half the study's, the missing part
being that between-arm separation plus the allometric scaling on weight that no
covariate can reach.

**This is the ceiling, and it is worth being precise about what it is a ceiling
on.** Handed the model that generated the data, this mode reproduces the event
structure exactly, the central level to within a quarter, and the dispersion to
about a half. Nothing a better prior could fix remains: what is left out is
what the catalogue cannot say.

## theo_md: an illustrative prior, and what being wrong costs

Nothing published was used here. The model is the one the methods article uses
for the same study — round numbers from allometric scaling, never fitted to
`theo_md` — and its clearance is deliberately too fast.

```{r theo-inputs, eval=run}
theo_roles <- pmx_roles(
  id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
  cmt = "CMT", covariates = "WT"
)
theo_model <- pmx_structural_model(
  pk = "1cmt_oral", typical = c(cl = 6, v = 35, ka = 1.5),
  source = "illustrative allometric scaling; never fitted to theo_md"
)
rich <- c(0, 0.25, 0.5, 1, 2, 4, 7, 9, 12, 24)
theo_design <- pmx_trial_design(
  dose_levels = 320, cohort_sizes = 12,
  sampling = list(rich, 0, NULL, NULL, NULL, NULL, rich),
  n_doses = 7, dose_interval = 24,
  source = "illustrative protocol: 320 mg daily, sampled on days 1 and 7"
)
theo_covariates <- pmx_covariates(WT = pmx_covariate(
  range = c(40, 130), median = 70, cv = 0.15,
  source = "illustrative; a healthy-volunteer cohort"
))
theo_syn <- synpmx_prior(theo_model, theo_design, theo_roles,
                         n_subjects = 12, seed = 1,
                         covariates = theo_covariates)
```

```{r theo-figure, eval=plots, fig.height=4}
frame <- rbind(observed(theo_md, theo_roles, "Source"),
               observed(theo_syn, theo_roles, "Synthetic"))
ggplot2::ggplot(frame, ggplot2::aes(time, dv, group = subject, colour = set)) +
  ggplot2::geom_line(alpha = 0.4) +
  ggplot2::geom_point(alpha = 0.4, size = 0.8) +
  ggplot2::facet_wrap(~ set) +
  ggplot2::scale_colour_manual(values = comparison_colours, guide = "none") +
  ggplot2::labs(x = "Study time (hours)", y = "Concentration (mg/L)",
                title = "theo_md under a deliberately imperfect prior") +
  ggplot2::theme_minimal()
```

```{r theo-level, eval=run}
round(level_ratio(theo_md, theo_syn, theo_roles), 3)
```

```{r theo-card, eval=run}
synpmx_scorecard_datatable(
  synpmx_scorecard(theo_md, theo_syn, theo_roles), report = "minimal"
)
```

The concentrations run low, by about half. The event structure, the sampling
schedule, the occasions and the censoring flag are all correct; only the level
is wrong, and it is wrong because the clearance assumed was wrong. The
scorecard is the same shape as the run above, which is the point worth noticing
about the scorecard here: it reads structure, and structure is what this mode
gets right whatever the parameters say.

## How wrong can the prior be

The two runs above bracket the answer. This sweep fills in between, holding
everything but the assumed clearance fixed and reading off the level each one
produces.

```{r sensitivity, eval=run}
sweep <- do.call(rbind, lapply(c(1.5, 2.2, 3, 4.5, 6, 9, 13.5), function(cl) {
  model <- pmx_structural_model(
    pk = "1cmt_oral", typical = c(cl = cl, v = 35, ka = 1.5),
    source = "illustrative sweep"
  )
  syn <- synpmx_prior(model, theo_design, theo_roles, n_subjects = 12,
                      seed = 1, covariates = theo_covariates)
  level <- level_ratio(theo_md, syn, theo_roles)
  data.frame(assumed_cl = cl, synthetic_median = round(level[["synthetic"]], 2),
             ratio_to_source = round(level[["ratio"]], 2))
}))
knitr::kable(sweep, row.names = FALSE,
             caption = "Assumed clearance against the level it produces")
```

```{r sensitivity-figure, eval=plots, fig.height=3.6}
ggplot2::ggplot(sweep, ggplot2::aes(assumed_cl, ratio_to_source)) +
  ggplot2::geom_hline(yintercept = 1, linetype = "dashed",
                      colour = "#888888") +
  ggplot2::geom_line(colour = "#1B6CA8") +
  ggplot2::geom_point(colour = "#1B6CA8") +
  ggplot2::scale_x_log10() + ggplot2::scale_y_log10() +
  ggplot2::labs(x = "Assumed clearance (L/h, log scale)",
                y = "Synthetic / source median (log scale)",
                title = "The output level is the elicitation error") +
  ggplot2::theme_minimal()
```

Two things to take from this. The level the study actually sits at corresponds
to an assumed clearance of a little over 2 L/h, and the illustrative model's 6
is about three times that — hence the halved concentrations.

And the slope is **shallower than one**. Clearance across this sweep spans a
factor of nine, and the resulting level spans a factor of four. A profile
sampled richly through absorption is not purely clearance-determined: early
concentrations are absorption-limited, so an error in clearance is diluted in
the median. That cuts both ways. A prior wrong by a factor of two costs less
than a factor of two in the output, and a level that looks close is weaker
evidence about the clearance than it appears.

## What the two runs held

```{r summary-table, eval=run}
levels_tbl <- rbind(
  c(dataset = "mixroute_sim", prior = "documented generating truth",
    round(level_ratio(mixroute, mixroute_syn, mixroute_roles), 2)),
  c(dataset = "theo_md", prior = "illustrative allometric scaling",
    round(level_ratio(theo_md, theo_syn, theo_roles), 2))
)
knitr::kable(levels_tbl, row.names = FALSE,
             caption = "Median concentration, source against synthetic")
```

A correct prior put the level within a quarter; a prior three times off in
clearance halved it. Neither run failed a structural check, because nothing
about the event table depends on the parameters being right — which is exactly
what makes this mode useful for developing code and useless for answering
questions about a drug.

## Where to go next

- `vignette("prior-demo")` — one specification worked end to end, including the
  parts of it the generator cannot express.
- `vignette("prior-algorithm")` — every step, and the full catalogue enumerated
  from the code.
- [Evaluating AVATAR on public data](https://iamstein.github.io/synpmx/articles/avatar-public-data-examples.html)
  — the same collection of studies under a generator that reads them, which is
  what the table at the top of this article is implicitly recommending for most
  of them.
- [The synpmx data generation algorithms](https://iamstein.github.io/synpmx/articles/synpmx-methods.html)
  — the other modes, and which one to use when.
