---
title: "Example: prior-only synthetic PK for pyrazinamide"
format: html
---

One concrete case worked end to end: a synthetic pyrazinamide dataset, 1200 mg
orally once daily for two weeks, sampled on day 14. Every number and every
choice is shown, including the ones the generator cannot represent.

Prior-only generation reads **no data at all** --- it takes a public structural
model and a trial design and simulates from those alone --- so nothing here
needs a privacy argument, and there is no `data` argument to supply. See
[Privacy](https://iamstein.github.io/synpmx/articles/synpmx-privacy.html) for
where this sits among the three method families.

The cardinal rule for this mode: every number below must come from somewhere
that is *not* the study being simulated. The values used here are literature /
protocol values for pyrazinamide, recorded in each object's `source` field. If
you were to fit a model to a real study and paste its estimates in here, the
real data would reach the output through the parameters and the "nothing was
read" claim would be false --- `synpmx_prior()` cannot detect that, so the
discipline is yours to keep.

**Covariates are deliberately not generated here.** The original specification
included WT, CD38, and GENDER along with their effects on CL/F and V/F. This
generator has no way to link a covariate to a parameter (see the last section),
so those columns would carry no relationship to the concentrations beside them
--- decoration that invites exactly the wrong conclusion. They are left out
rather than shipped as plausible-looking noise.

**Read the last section before using the output.** Three further parts of the
specification cannot be expressed either, and the script does not pretend
otherwise.

```{r}
#| label: setup
#| message: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center",
                      out.width = "100%")
library(synpmx)
has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
```

## The specification

| Quantity | Value |
|---|---|
| Structural model | 1-compartment oral, first-order absorption and elimination |
| Ka | 1.31 h^-1^ |
| CL/F | 3.52 L/h |
| V/F | 28.57 L |
| IIV (log-SD) | Ka 0.620, CL 0.174, V 0.181 |
| IOV (log-SD) | V 0.117 |
| Residual error | proportional 0.22, additive SD 2.41 mg/L |
| WT, CD38, GENDER | *not generated --- see above* |
| Regimen | 1200 mg orally once daily |
| Sampling | after 2 weeks of dosing, at 0.3, 0.9, 2.2, 4.5, 8 h post dose |

Choices not given in the specification, made here and flagged so they are easy
to change:

```{r}
#| label: choices
N_SUBJECTS <- 60          # cohort size
SEED       <- 20260728    # reproducibility seed
LLOQ       <- 0.2         # mg/L, a typical HPLC assay limit for pyrazinamide
DROPOUT    <- 0           # no discontinuation modelled
N_DOSES    <- 14          # "2 weeks of dosing" at once daily
```

## One unit conversion

`pmx_structural_model(iiv = )` takes a **CV**, while the specification gives
**log-SD**. These are different parameterisations of the same lognormal, so the
values are converted rather than pasted: `CV = sqrt(exp(omega^2) - 1)`.

```{r}
#| label: iiv
omega_to_cv <- function(omega) sqrt(exp(omega^2) - 1)

iiv_log_sd <- c(ka = 0.620, cl = 0.174, v = 0.181)
iiv_cv     <- omega_to_cv(iiv_log_sd)
knitr::kable(
  data.frame(parameter = names(iiv_log_sd),
             log_sd = iiv_log_sd, cv = round(iiv_cv, 4)),
  row.names = FALSE, caption = "IIV: specification (log-SD) to package (CV)"
)
```

## The public inputs

```{r}
#| label: model
pza_model <- pmx_structural_model(
  pk      = "1cmt_oral",
  typical = c(cl = 3.52, v = 28.57, ka = 1.31),
  iiv     = iiv_cv,
  residual_cv = 0.22,
  source  = paste(
    "Published pyrazinamide population PK parameters; never fitted to the",
    "study being simulated."
  )
)
pza_model
```

Sampling is requested only after two weeks of dosing, so thirteen doses carry no
samples and the fourteenth carries the full profile. `sampling` takes one entry
per dose, and `NULL` means "no samples after this one".

```{r}
#| label: design
post_dose <- c(0.3, 0.9, 2.2, 4.5, 8)

pza_design <- pmx_trial_design(
  dose_levels   = 1200,
  cohort_sizes  = N_SUBJECTS,
  sampling      = c(rep(list(NULL), N_DOSES - 1L), list(post_dose)),
  n_doses       = N_DOSES,
  dose_interval = 24,
  source        = "Protocol: 1200 mg orally once daily for 2 weeks."
)
pza_design
```

## Generate

```{r}
#| label: generate
pza <- synpmx_prior(
  pza_model, pza_design,
  n_subjects = N_SUBJECTS, seed = SEED,
  dropout = DROPOUT, lloq = LLOQ
)

str(pza)
```

```{r}
#| label: head
knitr::kable(head(pza, 8), digits = 3,
             caption = "First rows of the generated dataset")
```

## Does it match the specification?

The checks worth making are the ones on quantities that were actually
requested.

Two time columns matter here. `NTIME` is the **nominal** (protocol) time and
`TAD` is the **actual** time after dose, which the generator jitters around
nominal by `visit_window` (5% by default) so that sample times look like a real
study rather than a grid. Checking the schedule means checking nominal time; the
last dose falls at hour `24 * (N_DOSES - 1)`.

```{r}
#| label: check-design
roles <- pmx_generated_roles()
obs   <- pza[[roles$evid]] == 0 & !is.na(pza[[roles$dv]])

# Nominal time after the qualifying dose, recovered from NTIME and the occasion.
nominal_tad <- pza[[roles$nominal_time]] - 24 * (pza[[roles$occasion]] - 1)

knitr::kable(
  data.frame(
    quantity = c("subjects", "dose amount (mg)", "doses per subject",
                 "observations per subject", "sampling occasion",
                 "nominal times after dose (h)"),
    value = c(
      length(unique(pza[[roles$id]])),
      paste(unique(pza[[roles$amt]][pza[[roles$amt]] > 0]), collapse = ", "),
      round(mean(table(pza[[roles$id]][pza[[roles$amt]] > 0])), 1),
      round(mean(table(pza[[roles$id]][obs])), 1),
      paste(sort(unique(pza[[roles$occasion]][obs])), collapse = ", "),
      paste(sort(unique(round(nominal_tad[obs], 2))), collapse = ", ")
    )
  ),
  row.names = FALSE, caption = "Design realised in the output"
)
```

```{r}
#| label: check-concentrations
by_time <- split(pza[[roles$dv]][obs], round(nominal_tad[obs], 2))
knitr::kable(
  data.frame(
    nominal_tad = names(by_time),
    n           = vapply(by_time, length, integer(1)),
    median      = round(vapply(by_time, median, numeric(1)), 2),
    min         = round(vapply(by_time, min, numeric(1)), 2),
    max         = round(vapply(by_time, max, numeric(1)), 2)
  ),
  row.names = FALSE,
  caption = "Concentration (mg/L) by nominal time after dose (h)"
)
```

```{r}
#| label: profiles
#| eval: !expr has_ggplot2
#| fig-height: 4
plot_data <- data.frame(
  id   = as.character(pza[[roles$id]][obs]),
  tad  = pza[[roles$tad]][obs],
  dv   = pza[[roles$dv]][obs]
)
ggplot2::ggplot(plot_data, ggplot2::aes(tad, dv, group = id)) +
  ggplot2::geom_line(alpha = 0.25, colour = "#1B6CA8") +
  ggplot2::geom_point(alpha = 0.35, size = 0.8, colour = "#1B6CA8") +
  ggplot2::labs(
    x = "Time after dose (h)", y = "Concentration (mg/L)",
    title = "Pyrazinamide 1200 mg once daily, day 14"
  ) +
  ggplot2::theme_minimal()
```

## What this does and does not represent

Prior-only generation exists to produce structurally correct data for pipeline
and model-code development. It is not a reimplementation of the source model,
and three parts of the specification have no representation in it. None of them
fails loudly, so they are listed here rather than left to be discovered.

**1. Covariate effects cannot be expressed, which is why no covariates are
generated.** The requested WT exponents (0.75 on CL/F, 1 on V/F), the GENDER
effect on CL/F (-0.40), and the CD38 effect on CL/F (-0.22) have nowhere to go.

The reason is worth stating precisely, because it is not a deep one. The
generator draws each subject's parameters as typical value times lognormal IIV
and simulates the whole profile from those; covariates, when requested, are
drawn separately afterwards and merged onto the finished table, so by the time a
weight exists the concentration beside it has already been computed. There is no
argument on `pmx_structural_model()` that could link them, and no code anywhere
in the package that applies a covariate to a parameter.

The generator *will* happily add `WT`, `CD38`, and `GENDER` columns if asked --
`synpmx_prior(covariates = pmx_covariates(...))`. They would be plausible
numbers bearing **no relationship to the concentrations in the same row**, and a
reader who did not know that would draw a false conclusion from the first
covariate plot they made. That is the reason they are omitted here: a missing
column is honest, a meaningless one is not. Add them only if your pipeline needs
columns of the right *type* to run against, and label them as such.

Note what this is *not*. It is not a privacy constraint: prior-only generation
reads nothing, so a covariate effect would be a public assertion exactly like
`typical` and `iiv`, costing no budget. (In the differentially private engines
the calculus differs, since every covariate consumes a slice of the budget.) It
is simply unimplemented. If you need covariate--parameter relationships in
synthetic data, simulate them explicitly in `rxode2` or `nlmixr2`, or use
`synpmx_avatar()` on real data, where real covariate relationships survive
because whole real subjects do.

**2. IOV is absent.** The requested inter-occasion variability on V (log-SD
0.117) is not represented; the generator draws one parameter vector per subject
for the whole study. With sampling on a single day this is a small omission
here, but it would matter if the schedule were extended across occasions.

**3. Residual error is proportional only.** `residual_cv = 0.22` carries the
proportional part. The additive component (SD 2.41 mg/L) has no argument and is
dropped, so low concentrations are less noisy than specified. With sampling
ending at 8 h and concentrations staying well above the additive scale, the
practical effect is modest --- but at a trough it would not be.

What the output *is* good for: exercising code that reads, reshapes, plots, and
fits a repeated-dose PK dataset with a realistic event structure, nominal and
actual times, occasions, and a censoring flag --- without a single real patient
being involved.
