---
title: "Visit-time masking on nimoData: a failure, diagnosed, then fixed"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synpmx)
data("nimoData", package = "nlmixr2data")
```

`nimoData` is the one public dataset where `coarsen_time = TRUE` — on by
default — accomplishes **nothing**. All twelve subjects hold a set of
observation times no other subject shares, before coarsening and after, so every
avatar carries one real patient's exact visit schedule.

This walks through the failure, diagnoses it with measurements, and then fixes
it by constructing the nominal time the dataset does not record. The fix takes
subjects-alone from 12 to 5, and the last section is honest about why it does
not reach 0.

```{r roles}
nimo_roles <- function(nominal = NULL) {
  pmx_roles(
    id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
    rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
    nominal_time = nominal,
    covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
  )
}

exposure <- function(data, roles, label) {
  fit <- suppressWarnings(suppressMessages(
    synpmx_avatar(data, roles, seed = 606)
  ))
  s <- attr(fit, "pmx_settings")
  data.frame(
    Run = label,
    Grid = s$time_grid,
    Alone = s$exposure_alone,
    `Unique moment` = s$exposure_unique_moment,
    `Unique visit set` = s$exposure_unique_pattern,
    `Visit sets` = s$patterns_total,
    `Invented` = paste0(round(100 * s$pattern_generated_fraction), "%"),
    check.names = FALSE
  )
}
```

## 1. Two kinds of exposure, in plain terms

The measurements in an avatar are blended from several real donors, so nobody's
concentrations survive. What is copied verbatim from one real patient is the
**schedule**: when they were dosed and when they were sampled. Two different
things can make a schedule identifying, and they need different fixes.

**Seen at a unique moment.** The patient was observed at a clock time nobody
else was observed at. Sampling recorded to the minute does this to everybody:
if you were bled at 1.28 hours and nobody else was bled at exactly 1.28 hours,
your time vector is yours alone. *Fixable*, by placing everyone on a shared
visit grid — that is what `coarsen_time` does.

**Unique set of visits attended.** Once everyone *is* on a shared grid, every
individual time is shared, so no single visit gives anyone away. What can still
give them away is the **combination of visits they attended and missed** —
dropout, an interrupted course, a skipped week. If one patient is the only one
who missed weeks 2 and 3, that combination points at them. *Not fixable by any
grid*, because the times are already shared; the grid decides where the visits
are, not which ones a patient turned up for.

Throughout this document, "visit set" means that combination: which of the
study's visits a patient actually had.

## 2. The failure

```{r failure}
knitr::kable(exposure(nimoData, nimo_roles(), "default, inferred grid"),
             row.names = FALSE)
```

Twelve of twelve alone, every one of them on **unique moment**. The grid found
nothing to share. Note the last column too — every avatar's visit set had to be
invented rather than reused, which is the same failure seen from the other side.

Why would a study look like this?

```{r shape}
observed <- nimoData$EVID == 0 & nimoData$MDV == 0
times <- nimoData$TIME[observed]
subject <- as.character(nimoData$ID[observed])

c(subjects = length(unique(subject)),
  observations = length(times),
  distinct_times = length(unique(times)),
  obs_per_subject = round(mean(table(subject)), 1))
```

**311 distinct observation times in a cohort of twelve.** If the study had a
shared visit grid you would expect roughly 26 — one per protocol visit. That
number is the whole diagnosis: these patients were not sampled at common times.

## 3. Why the grid cannot fix it

`.derive_time_grid()` pools every observation time and merges them into visit
cells agglomeratively, closest pair first. A merge is refused for either of two
reasons:

1. **The cell would hold one patient twice.** Collapsing two of subject 7's own
   visits into one time destroys a real visit rather than hiding a schedule.
2. **The gap is wider than typical within-patient spacing**, guarding against an
   outlier's lone late observation dragging an early visit to their mean. The
   threshold is the median within-patient gap.

Both come from the data, so both can be checked:

```{r thresholds}
gaps_within <- unlist(lapply(split(times, subject),
                             function(v) diff(sort(unique(v)))))
gaps_within <- gaps_within[is.finite(gaps_within) & gaps_within > 0]
gaps_between <- diff(sort(unique(times)))

c(merge_threshold_hours = stats::median(gaps_within),
  cohort_gaps_under_threshold = sum(gaps_between <= stats::median(gaps_within)),
  cohort_gaps_total = length(gaps_between))
```

**Constraint 2 is barely binding** — 294 of 310 gaps are under the threshold. So
the blocking is almost entirely constraint 1, and that is the finding that
matters.

```{r grid-built}
grid <- synpmx:::.derive_time_grid(times, subject)
snapped <- synpmx:::.snap_to_grid(times, grid)
occupancy <- rowSums(table(snapped, subject) > 0)

c(distinct_times_before = length(unique(times)),
  grid_cells_after = length(grid),
  cells_held_by_one_patient = sum(occupancy == 1),
  cells_held_by_two = sum(occupancy == 2),
  cells_held_by_three_plus = sum(occupancy >= 3))
```

Coarsening **does** run — 311 times collapse to about 98 cells. It is simply not
enough, because about a quarter of the cells still hold exactly one patient, and
a patient needs only one such cell to be alone.

And this is structural, not a tuning problem. Twelve patients contribute 24–28
observations each. For the grid to reach ~26 shared cells, each cell would have
to hold roughly one observation from each patient — and getting there requires
repeatedly merging times that belong to the *same* patient, which constraint 1
forbids. **The algorithm stops where it does because going further deletes real
visits.**

## 4. "So why not just coarsen further?"

The obvious question, and worth answering with a measurement rather than an
argument. Rounding `TIME` to a fixed resolution and declaring it as
`nominal_time` forces exactly the coarser grid the default declined to build.

```{r sweep}
rounded <- function(hours) {
  out <- nimoData
  out$NTIME <- round(out$TIME / hours) * hours
  out
}

knitr::kable(rbind(
  exposure(nimoData, nimo_roles(), "default, inferred grid"),
  do.call(rbind, lapply(c(6, 12, 24, 48, 168, 336), function(h) {
    exposure(rounded(h), nimo_roles("NTIME"), sprintf("round to %g h", h))
  }))
), row.names = FALSE)
```

**It helps, then stops helping, and it is not monotone.** A one-week grid takes
unique-moment exposure to zero — but `Alone` does not follow it to zero, and
both a two-day and a two-week grid are *worse* than the default.

The reason is the trade between the two middle columns. Merging times removes
unique-moment exposure by making every time shared, but it also changes which
cells each patient occupies, and a patient whose merged occupancy is unique
becomes alone on **unique visit set** instead — which no grid can fix. The
exposure moves columns rather than disappearing.

That non-monotonicity is why `synpmx` does not simply retry at a coarser
resolution until the number improves. There is no dial where coarser is reliably
better.

### And coarser grids are not free

```{r cost}
cost_row <- function(label, values) {
  data.frame(
    Grid = label,
    `Same-patient collisions` = sum(duplicated(paste(subject, values))),
    `Distinct times per patient` = round(
      mean(tapply(values, subject, function(v) length(unique(v)))), 1),
    check.names = FALSE
  )
}

knitr::kable(rbind(
  cost_row("source TIME", times),
  cost_row("inferred grid", snapped),
  do.call(rbind, lapply(c(24, 48, 168, 336), function(h) {
    cost_row(sprintf("round to %g h", h), round(times / h) * h)
  }))
), row.names = FALSE)
```

A "same-patient collision" is two of one patient's own observations forced onto
the same time. The one-week grid that zeroed unique-moment exposure buys it with
**164 collisions**, cutting distinct sampling times per patient from 26.8 to
13.1 — half the PK profile thrown away. The two-week grid leaves 7 of 27.

**The default is not being timid. It is refusing to buy a better privacy number
with data destruction.**

## 5. The fix: construct the nominal time

The study *has* a protocol grid. It just is not recorded as a column — it is
implicit in two columns that are.

### What OCC and TAD contain

```{r structure}
knitr::kable(
  head(nimoData[nimoData$ID == 1,
                c("ID", "TIME", "TAD", "OCC", "EVID", "AMT", "DV")], 11),
  row.names = FALSE, digits = 2,
  caption = "Subject 1: the first three infusion occasions."
)
```

Read the `OCC` and `TAD` columns together and the design is visible. `OCC` is
the infusion number, 1 to 10. `TAD` is time since that infusion started, and it
resets to 0 at each one. Subject 1's first occasion is sampled at TAD 1.28,
23.16, 46.91, 95.21, 144.13, 167.16 — which is days 0, 1, 2, 4, 6, 7, recorded
to the minute.

**So the protocol is: infusions a week apart, sampled on nominal days within
each week.** Neither the week nor the day is stored, but both are recoverable.

### The drift that defeats the inferred grid

```{r drift}
doses <- nimoData[nimoData$EVID != 0 & nimoData$AMT > 0, ]
drift <- do.call(rbind, lapply(1:5, function(occasion) {
  actual <- doses$TIME[doses$OCC == occasion]
  data.frame(Occasion = occasion,
             `Nominal hour` = (occasion - 1) * 168,
             `Earliest actual` = round(min(actual), 1),
             `Latest actual` = round(max(actual), 1),
             `Spread (h)` = round(max(actual) - min(actual), 1),
             check.names = FALSE)
}))
knitr::kable(drift, row.names = FALSE)
```

There is the mechanism. Occasion 1 starts at hour 0 for everyone, but by
occasion 4 the actual start times are spread over almost a day, and the spread
grows with each occasion as small delays accumulate. Sampling times inherit that
drift and add their own. By late occasions, two patients at the *same protocol
visit* are hundreds of hours apart on the clock — far beyond anything the merge
threshold would join, and hopelessly entangled with genuinely different visits.

**The inferred grid never had a chance, because on the actual clock these
patients really were not at the same place at the same time.** The shared
structure exists only relative to each patient's own infusion schedule.

### Building it

```{r build-nominal}
nimo_fixed <- nimoData
nimo_fixed$NTIME <-
  (nimo_fixed$OCC - 1) * 168 +                                  # which week
  ifelse(nimo_fixed$EVID == 0, round(nimo_fixed$TAD / 24) * 24, 0)  # which day

c(distinct_TIME = length(unique(nimo_fixed$TIME)),
  distinct_NTIME = length(unique(nimo_fixed$NTIME)),
  median_deviation_h = stats::median(abs(nimo_fixed$TIME - nimo_fixed$NTIME)))
```

Two terms. `(OCC - 1) * 168` places the visit in its protocol week. `round(TAD /
24) * 24` places it on its nominal day within that week, and applies only to
observations — a dose row is by definition at TAD 0.

**413 distinct clock times become 33 nominal ones**, and the median observation
sits 3 hours from its nominal slot. That is the shared grid the inferred method
could not see.

### What it achieves

```{r fixed}
knitr::kable(rbind(
  exposure(nimoData, nimo_roles(), "default, inferred grid"),
  exposure(nimo_fixed, nimo_roles("NTIME"), "constructed nominal time")
), row.names = FALSE)
```

- **Alone falls from 12 to 5**, and the grid is now reported as `nominal` rather
  than `derived` — snapping is exact instead of guessed.
- **Unique-visit-set exposure is zero.** Every patient who is still alone is
  alone for the one reason a grid can still address, not the one it cannot.
- **Invented visit sets fall from 100% to about 17%.** Once patients share a
  grid, real visit sets become reusable, so most avatars get a real one.

That last point is worth dwelling on, because the discard count alone would have
hidden it: two visit sets are discarded either way. What changed is that the
default had to *fabricate* every avatar's visit set from a shape, and the fix
does not.

### Why five patients are still exposed

```{r residual}
last_visit <- tapply(nimo_fixed$NTIME[nimo_fixed$EVID == 0],
                     nimo_fixed$ID[nimo_fixed$EVID == 0], max)
knitr::kable(
  data.frame(Subject = names(sort(last_visit)),
             `Last nominal visit (h)` = as.integer(sort(last_visit)),
             `In weeks` = round(sort(last_visit) / 168, 1),
             check.names = FALSE),
  row.names = FALSE,
  caption = "Follow-up length after the protocol grid is applied."
)
```

The dosing weeks are shared; the **washout tail after the last infusion is
not**. Patients stop at 1992, 2016, 2112, and 2136 hours, and the ones at the
uncommon end are alone on that final visit no matter how the earlier weeks are
gridded. Coarsening the tail further does not help — it merges genuinely
different follow-up lengths, which converts moment-exposure into visit-set
exposure and, as the sweep showed, can leave the total worse.

**Zero is probably not reachable on this study, and that is a fact about the
study rather than about the package.** Twelve patients with individual follow-up
lengths do not contain twelve indistinguishable schedules. From here the options
are to remediate the five with `flag_identifiable_subjects()` and
`remediate_identifiable_subjects()`, or to accept that these avatars carry
schedule information and keep the output inside the source data's access
controls — which is the governance model the package assumes anyway.

## 6. Did the fix damage the data?

A privacy improvement that wrecks the profiles is not an improvement. The
nominal grid moves every observation by a median of 3 hours, so it is worth
checking what that did.

```{r utility}
fitted <- suppressWarnings(suppressMessages(
  synpmx_avatar(nimo_fixed, nimo_roles("NTIME"), seed = 606)
))
knitr::kable(
  compare_pmx_distributions(nimo_fixed, fitted, nimo_roles("NTIME"))$endpoints,
  digits = 2, row.names = FALSE,
  caption = "Concentration distribution, source versus synthetic."
)
```

```{r profiles, echo=FALSE, fig.width=9, fig.height=4.5}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  panel <- function(data, label) {
    keep <- data$EVID == 0 & data$MDV == 0
    data.frame(dataset = label, subject = as.character(data$ID[keep]),
               tad = data$TAD[keep], dv = data$DV[keep])
  }
  both <- rbind(panel(nimo_fixed, "Source"), panel(fitted, "Synthetic"))
  print(
    ggplot2::ggplot(both, ggplot2::aes(tad, dv, group = subject,
                                       colour = dataset)) +
      ggplot2::geom_line(alpha = 0.3) +
      ggplot2::geom_point(alpha = 0.4, size = 0.7) +
      ggplot2::facet_wrap(~dataset) +
      ggplot2::scale_colour_manual(
        values = c(Source = "#1B6CA8", Synthetic = "#D95F02")) +
      ggplot2::labs(x = "Time after dose (hours)", y = "DV",
                    title = "nimoData with a constructed nominal time") +
      ggplot2::theme_minimal() +
      ggplot2::theme(legend.position = "none")
  )
}
```

## 7. What to take from this

1. **An inferred grid can fail completely, and it says so.** The alert is not
   decoration — 12 of 12 alone means every avatar wears a real patient's
   schedule.
2. **The failure is a property of the study.** Individually drifting visit times
   mean there is no shared clock to find. No amount of tuning invents one.
3. **Coarsening harder is not the answer.** It is non-monotone, and it pays for
   the improvement by collapsing a patient's own visits together.
4. **Recovering the protocol grid is the answer, and it is usually possible.**
   `OCC` and `TAD` were enough here. Most studies carry something equivalent
   even when they do not carry a nominal-time column outright.
5. **Check the residual rather than assuming success.** Five patients remain
   exposed, all through follow-up length, which is a different problem needing a
   different response.

## Appendix: a discrepancy found while writing this

The merge enforces that a cell never holds one patient twice, but the snap that
follows assigns each time to the **nearest cell centre**, which is a separate
decision. A time near a boundary can land in a neighbouring cell and break the
constraint the merge just enforced:

```{r snap-mismatch}
collisions <- paste(subject, snapped)[duplicated(paste(subject, snapped))]
example <- which(paste(subject, snapped) == collisions[1])
data.frame(subject = subject[example], source_time = times[example],
           snapped_to = round(snapped[example], 3))
```

Two observations of one patient, recorded less than an hour apart, pushed onto a
single time — 35 such collisions here. Small next to the 164 a one-week grid
causes, and it changes no conclusion above, but the guarantee in the code
comments is stronger than what the code delivers. Recorded as `REV-031`.
