---
title: "Visit-time masking on nimoData: a failure, diagnosed, then fixed"
format:
  html:
    toc: true
    toc-depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(synpmx)
data("nimoData", package = "nlmixr2data")
```

## The one term you need first

Everything below counts patients with a **unique sampling schedule**, so here is
what that means before anything else.

> A patient has a **unique sampling schedule** when the list of times they were
> observed at is not shared by any other patient in the study.

Say a study enrols ten patients. Nine of them had PK sampled at hours 1, 4, 24
and 48. The tenth was sampled at hours 1, 4, 24 and 49. That tenth patient is
the one with a unique sampling schedule: their list of times belongs to nobody
else, so it works as a fingerprint.
Anyone
holding that patient's real visit records can pick their synthetic counterpart
out of the cohort — because `synpmx` copies the *schedule* verbatim from one
real patient even though it blends the *measurements* from several.

So **this is the number to drive to zero**, and it is counted on the real data
before any synthetic data is made. Zero means no synthetic patient can be traced
back to a particular real one through their visit times. Twelve out of twelve
means every single one can.

Two things it does **not** mean:

- It does not mean anyone's measurements leaked. Concentrations are blended
  across at least five donors and capped, so no real patient's numbers appear.
- It does not mean an attack has happened. Uniqueness is a property of the
  source data — it says the schedule *would* work as an identifier.

## What this document does

`nimoData` is the one public dataset where `coarsen_time = TRUE` — the default
mechanism for fixing this — accomplishes **nothing**. All twelve patients have a
unique sampling schedule before coarsening, and all twelve still do after.

This walks through that failure, diagnoses it with measurements, and then fixes
it by constructing the nominal visit time the dataset does not record. **The fix
takes the count from 12 to 5**, and the last section is honest about why it
cannot reach 0.

```{r roles}
nimo_roles <- function(nominal = NULL) {
  pmx_roles(
    id = "ID", time = "TIME", dv = "DV", amt = "AMT", evid = "EVID",
    rate = "RATE", mdv = "MDV", tad = "TAD", occasion = "OCC",
    nominal_time = nominal,
    covariates = c("BSA", "AGE", "HGT"), keep = "DOS"
  )
}

exposure <- function(data, roles, label) {
  fit <- suppressWarnings(suppressMessages(
    synpmx_avatar(data, roles, seed = 606)
  ))
  s <- attr(fit, "pmx_settings")
  data.frame(
    Run = label,
    Grid = s$time_grid,
    `Unique schedule` = s$unique_schedule_n,
    `Unique sample time` = s$unique_sample_time_n,
    `Unique visit set` = s$unique_visit_set_n,
    `Visit sets` = s$patterns_total,
    `Invented` = paste0(round(100 * s$pattern_generated_fraction), "%"),
    check.names = FALSE
  )
}
```

## 1. The two reasons a schedule ends up unique

A schedule can be unique for two different reasons. Every table below splits the
count between them, because they call for opposite responses.

**Seen at a unique moment.** The patient was observed at a clock time nobody
else was observed at. Sampling recorded to the minute does this to everybody —
PK sampled at 1.28 hours when nobody else was sampled at exactly 1.28 hours, and
the list is already unique. **Fixable**: place everyone on a shared visit grid,
so 1.28 and 1.31 and 1.19 all become "the 1-hour visit". That is exactly what
`coarsen_time` does, and on most datasets it works.

**Unique set of visits attended.** Once everyone *is* on a shared grid, every
individual time is shared, so no single visit gives anyone away. What can still
give them away is the **combination of visits they attended and missed** —
dropout, an interrupted course, a skipped week. If one patient is the only one
who missed weeks 2 and 3, that combination points at them just as well as an odd
sampling time would. **Not fixable by any grid**, because the times are already
shared; a grid decides where the visits are, not which ones a patient turned up
for. This one needs a different mechanism (`min_pattern_share`) or a decision to
drop the patient.

Throughout, **"visit set"** is the shorthand for that combination: which of the
study's shared visits one patient actually had.

The distinction is the practical one in this whole document. A high
unique-moment count means *try harder on the grid*. A high unique-visit-set
count means *the grid is finished; the remaining exposure is dropout, and you
have a different decision to make.*

## 2. The failure

Run `synpmx_avatar()` on `nimoData` with everything at its defaults:

```{r failure}
knitr::kable(exposure(nimoData, nimo_roles(), "default, inferred grid"),
             row.names = FALSE)
```

### What each column is

This table appears repeatedly below, so it is worth reading the columns once.

| Column | Meaning | Want |
|---|---|---|
| **Unique schedule** | Patients out of 12 whose list of observation times nobody else shares. The headline number. | 0 |
| **Unique sample time** | Of those, how many are unique because they were sampled at a clock time nobody else was. | 0 — a grid can fix these |
| **Unique visit set** | Of those, how many share every individual time with somebody but attended a combination of visits nobody else did. | no grid fixes these |
| **Visit sets** | How many distinct combinations of attended visits exist in the study, after coarsening. | — |
| **Invented** | Share of avatars whose missed visits were placed fresh instead of copied from a real patient. | — (see below) |

`Unique schedule` is always `Unique sample time` + `Unique visit set`.

### Reading this row

**All twelve patients have a unique sampling schedule.** Every patient has a list
of observation times that no other patient shares, so every avatar built from
this data wears a schedule belonging to exactly one real person. That is the
failure, and it is total.

**All twelve are unique on sample time.** Not one of them was rescued by the
grid: each was observed at clock times nobody else was observed at, and the
coarsening that exists to merge such times found nothing to merge. Because
`Unique visit set` is 0, the entire problem here is the *fixable* kind — which
is the one encouraging thing in the row, and the reason the rest of this
document is worth reading.

**`Invented` at 100% is a symptom, not a harm.** This is worth being precise
about, because the column is easy to misread. An invented visit set means no
real patient's combination of missed visits was reused — which is *good* for
privacy, not bad. The reason to look at it is diagnostic: reuse is only possible
when two patients attended the same set of visits, so **100% invented means no
two patients in this study had anything in common to reuse.** It is the same
underlying problem as the `Alone` column, measured a different way, and later it
becomes the most sensitive indicator that the fix worked.

If you only track one number, track **unique schedule**.

### Why would a study look like this?

One count explains the whole thing.

```{r shape}
observed <- nimoData$EVID == 0 & nimoData$MDV == 0
times <- nimoData$TIME[observed]
subject <- as.character(nimoData$ID[observed])

c(patients = length(unique(subject)),
  observations = length(times),
  observations_per_patient = round(mean(table(subject)), 1),
  distinct_times = length(unique(times)))
```

Each patient is sampled about **27** times. So if the twelve patients were
sampled at the *same* protocol visits, the whole study would contain about **27
distinct times** — everyone's hour-1 sample recorded as hour 1, everyone's
week-2 sample as week 2.

Instead it contains **311**, out of 321 observations.

That is the diagnosis, and it is worth stating bluntly: **almost every single
observation in this study sits at a moment of its own.** Not approximately
shared, not shared after rounding — its own. Here is the same fact from the
sharing side:

```{r sharing}
attendance <- table(times, subject) > 0
patients_per_time <- rowSums(attendance)

c(distinct_times = length(patients_per_time),
  times_shared_by_two_or_more = sum(patients_per_time >= 2),
  observations_at_a_shared_time = sum(patients_per_time[as.character(times)] >= 2),
  observations_total = length(times))
```

**Nine.** Nine times in the entire study are attended by more than one patient,
and 19 of 321 observations sit at one. Ninety-four percent of the observations
are at a moment no other patient was observed at.

### The same fact, as a picture

```{r visitmap, echo=FALSE, fig.width=9, fig.height=5.5}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  data("theo_md", package = "nlmixr2data")
  strip <- function(d, lab, tmax = Inf) {
    o <- d$EVID == 0 & d$TIME <= tmax
    if ("MDV" %in% names(d)) o <- o & d$MDV == 0
    tt <- d$TIME[o]
    counts <- table(tt)
    data.frame(study = lab, time = tt,
               shared = as.vector(counts[as.character(tt)] > 1))
  }
  both <- rbind(
    strip(theo_md, "theo_md — 32% of times shared by 2+ patients"),
    strip(nimoData, "nimoData — 3% of times shared by 2+ patients", tmax = 340))
  print(
    ggplot2::ggplot(both, ggplot2::aes(time, 0, colour = shared)) +
      ggplot2::geom_jitter(height = 0.35, width = 0, size = 1.5, alpha = 0.8) +
      ggplot2::facet_wrap(~study, scales = "free_x", ncol = 1) +
      ggplot2::scale_colour_manual(
        values = c("TRUE" = "#1B6CA8", "FALSE" = "#D95F02"),
        labels = c("TRUE" = "shared with another patient",
                   "FALSE" = "this patient alone here"),
        name = NULL) +
      ggplot2::labs(x = "Recorded observation time (hours)", y = NULL) +
      ggplot2::theme_minimal(base_size = 11) +
      ggplot2::theme(axis.text.y = ggplot2::element_blank(),
                     panel.grid.major.y = ggplot2::element_blank(),
                     panel.grid.minor = ggplot2::element_blank(),
                     legend.position = "bottom")
  )
}
```

Every recorded observation time, coloured by whether any other patient was also
observed at it. In `theo_md` the blue points are the shared protocol visits a
grid can be built from. In `nimoData` there are essentially none — the times are
orange almost everywhere, which is the same "9 out of 311" from above, seen
rather than counted.

A grid works by finding times that different patients have in common and pulling
them together. **Here there is essentially nothing in common to find** — which
is why all twelve schedules are unique, and why no amount of tuning
the grid algorithm will change that. The problem is not that the grid is set too
finely. The problem is that on the recorded clock, these twelve patients were
never in the same place at the same time.

Sections 3 and 4 show what the grid does try, and why pushing it harder makes
things worse. Section 5 finds the shared structure that *does* exist, in a place
the clock does not show it.

## 3. Why the grid cannot fix it

### What coarsening actually does

Coarsening has one job: **decide which recorded times are really the same
visit, and give them one shared time.** If three patients were sampled at 1.28,
1.31 and 1.19 hours, those are the same protocol visit written down three ways,
and replacing all three with "1.26 hours" hides who was who without changing the
data in any way that matters.

The way it decides is simple and worth picturing. Start with all 311 recorded
times as 311 separate visits. Then repeatedly:

> Find the two closest times still separate, and **join** them into one visit.

Join the closest pair, then the next closest, and so on. Each join makes the
grid coarser. The only question is when to stop — and rather than stopping at
some fixed resolution, the algorithm refuses individual joins for two reasons:

**Rule 1 — never put the same patient in one visit twice.** If patient 7 was
sampled at 1.2 hours *and* at 1.4 hours, those two times cannot be joined,
because patient 7 genuinely attended two separate visits there. Joining them
would not be masking; it would be **deleting one of patient 7's real
observations**. This is the rule that does almost all the work here.

**Rule 2 — never join across a gap wider than a typical within-patient gap.**
Guards against an odd, isolated late observation dragging a whole early visit
towards it. Set to the median gap between one patient's consecutive visits.

Since both thresholds come from the data, both can be checked directly:

```{r thresholds}
gaps_within <- unlist(lapply(split(times, subject),
                             function(v) diff(sort(unique(v)))))
gaps_within <- gaps_within[is.finite(gaps_within) & gaps_within > 0]
gaps_between <- diff(sort(unique(times)))

c(merge_threshold_hours = stats::median(gaps_within),
  cohort_gaps_under_threshold = sum(gaps_between <= stats::median(gaps_within)),
  cohort_gaps_total = length(gaps_between))
```

**Rule 2 is barely doing anything** — 294 of the 310 gaps between adjacent times
are under the threshold, so it permits almost every join it is asked about. The
blocking is therefore almost entirely Rule 1: *the same patient is already in
that visit.*

Here is what the algorithm manages before it runs out of legal joins.

```{r grid-built}
grid <- synpmx:::.derive_time_grid(times, subject)
snapped <- synpmx:::.snap_to_grid(times, grid)
occupancy <- rowSums(table(snapped, subject) > 0)

c(distinct_times_before = length(unique(times)),
  grid_cells_after = length(grid),
  cells_held_by_one_patient = sum(occupancy == 1),
  cells_held_by_two = sum(occupancy == 2),
  cells_held_by_three_plus = sum(occupancy >= 3))
```

Coarsening **does** run: 311 separate times become about 98 visits. It is simply
not nearly enough. About a quarter of those visits are still attended by exactly
one patient, and **a patient needs only one such visit for their whole schedule to be unique.**

And the wall it hits is structural, not a setting. Work backwards from the
target: to get down to ~27 shared visits, each visit would have to hold roughly
one observation from each of the twelve patients. But every patient contributes
about 27 observations spread across the study, so continuing to join would very
soon mean joining two times belonging to the *same* patient — which is exactly
what Rule 1 forbids, and forbids for a good reason.

**The algorithm stops where it does because the next join would start deleting
real observations rather than masking them.**

### What a "visit set" and a "shape" actually are

Two terms appear in the tables and are worth pinning down, because they decide
what comes out of the generator.

A **visit set** is which of the shared visits one patient actually had. Taking
`warfarin`, whose PK visits land on a 13-point grid:

```
grid          0.5  1.2  2  3  6  9  12  24  36  48  72  96  120
patient 1     0.5  1.2  2  3  6  9  12  24  36  48  72
patient 2                                  24  36  48  72  96  120
patient 3     0.5       2  3  6     12  24  36  48  72  96  120
```

Three patients, three different visit sets. Patient 2 was not sampled until
hour 24; patient 3 missed the 1.2 and 9 hour draws.

A **shape** is that same information with the specific visits forgotten — how
many were missed and how they were arranged:

| shape | meaning |
|---|---|
| `complete` | attended everything |
| `trailing` | every miss is at the end — dropout |
| `block` | the misses are one contiguous run — an interruption |
| `scattered` | anything else |

So patient 3 above is `scattered|2`. Shape exists because exact matching is
almost always too strict: two patients who each missed one visit have different
*visit sets* if they missed different visits, but the same *shape*, and two of
them together clear a floor that neither clears alone.

### So what does an avatar actually get?

When an avatar needs a visit set, `synpmx` draws a shape first, then:

- **if a real visit set of that shape is shared by enough patients, reuse it**
  verbatim; otherwise
- **place that many misses on the grid in that arrangement, choosing which
  visits to miss fresh** — and reject the draw if it happens to land on a real
  visit set too rare to have been reusable.

The second case is what the `Invented` column counts. Continuing the example,
here are three avatars from that same `warfarin` run:

```
AVATAR 34     24  36  48  72  96  120        <- patient 2's set, reused exactly
AVATAR 36     3  6  12  24  36  72  96  120  <- a set no real patient had
```

Avatar 36 is the invented case. **It is still an ordinary, complete dataset** —
real grid times, blended concentrations, valid dose and event structure, and it
passes `validate_pmx()` like any other. Nothing about it is half-formed. The
only synthetic part is *which* visits it attended: it missed as many visits, of
the same kind, as some real patient did, but not the same ones.

That is the whole trade. What survives is how much missingness a study has and
of what kind. What is deliberately lost is which specific visits any one patient
missed — because that is the part that points back at a person.

## 4. "So why not just coarsen further?"

The obvious question, and worth answering with a measurement rather than an
argument. Rounding `TIME` to a fixed resolution and declaring it as
`nominal_time` forces exactly the coarser grid the default declined to build.

```{r sweep}
rounded <- function(hours) {
  out <- nimoData
  out$NTIME <- round(out$TIME / hours) * hours
  out
}

knitr::kable(rbind(
  exposure(nimoData, nimo_roles(), "default, inferred grid"),
  do.call(rbind, lapply(c(6, 12, 24, 48, 168, 336), function(h) {
    exposure(rounded(h), nimo_roles("NTIME"), sprintf("round to %g h", h))
  }))
), row.names = FALSE)
```

**Coarsening more sometimes helps and sometimes makes things worse.** A one-week
grid takes unique-sample-time exposure to zero — but the headline count does not
follow it to zero, and both a two-day and a two-week grid come out *worse* than
the default. There is no "coarser is better" rule to follow.

The reason is the trade between the two middle columns. Merging times removes
unique-moment exposure by making every time shared, but it also changes which
cells each patient occupies, and a patient whose merged occupancy is unique
becomes unique on **set of visits** instead — which no grid can fix. The exposure
moves from one column to the other rather than disappearing.

That is why `synpmx` does not simply retry at a coarser resolution until the
number improves: it has no way to know whether the next step will help or hurt,
and the step that helps the count can be the one that ruins the data.

### And coarser grids are not free

```{r cost}
cost_row <- function(label, values) {
  data.frame(
    Grid = label,
    `Same-patient collisions` = sum(duplicated(paste(subject, values))),
    `Distinct times per patient` = round(
      mean(tapply(values, subject, function(v) length(unique(v)))), 1),
    check.names = FALSE
  )
}

knitr::kable(rbind(
  cost_row("source TIME", times),
  cost_row("inferred grid", snapped),
  do.call(rbind, lapply(c(24, 48, 168, 336), function(h) {
    cost_row(sprintf("round to %g h", h), round(times / h) * h)
  }))
), row.names = FALSE)
```

A "same-patient collision" is two of one patient's own observations forced onto
the same time. The one-week grid that zeroed unique-moment exposure buys it with
**164 collisions**, cutting distinct sampling times per patient from 26.8 to
13.1 — half the PK profile thrown away. The two-week grid leaves 7 of 27.

**The default is not leaving easy gains on the table. It is refusing to buy a
better privacy number by destroying data.**

## 5. The fix: construct the nominal time

The study *has* a protocol grid. It just is not recorded as a column — it is
implicit in two columns that are.

### What OCC and TAD contain

```{r structure}
knitr::kable(
  head(nimoData[nimoData$ID == 1,
                c("ID", "TIME", "TAD", "OCC", "EVID", "AMT", "DV")], 11),
  row.names = FALSE, digits = 2,
  caption = "Subject 1: the first three infusion occasions."
)
```

Read the `OCC` and `TAD` columns together and the design is visible. `OCC` is
the infusion number, 1 to 10. `TAD` is time since that infusion started, and it
resets to 0 at each one. Subject 1's first occasion is sampled at TAD 1.28,
23.16, 46.91, 95.21, 144.13, 167.16 — which is days 0, 1, 2, 4, 6, 7, recorded
to the minute.

**So the protocol is: infusions a week apart, sampled on nominal days within
each week.** Neither the week nor the day is stored, but both are recoverable.

### The drift that defeats the inferred grid

```{r drift}
doses <- nimoData[nimoData$EVID != 0 & nimoData$AMT > 0, ]
drift <- do.call(rbind, lapply(1:5, function(occasion) {
  actual <- doses$TIME[doses$OCC == occasion]
  data.frame(Occasion = occasion,
             `Nominal hour` = (occasion - 1) * 168,
             `Earliest actual` = round(min(actual), 1),
             `Latest actual` = round(max(actual), 1),
             `Spread (h)` = round(max(actual) - min(actual), 1),
             check.names = FALSE)
}))
knitr::kable(drift, row.names = FALSE)
```

There is the mechanism. Occasion 1 starts at hour 0 for everyone, but by
occasion 4 the actual start times are spread over almost a day, and the spread
grows with each occasion as small delays accumulate. Sampling times inherit that
drift and add their own. By late occasions, two patients at the *same protocol
visit* are hundreds of hours apart on the clock — far beyond anything the merge
threshold would join, and hopelessly entangled with genuinely different visits.

**The inferred grid never had a chance, because on the actual clock these
patients really were not at the same place at the same time.** The shared
structure exists only relative to each patient's own infusion schedule.

### Building it

```{r build-nominal}
nimo_fixed <- nimoData
nimo_fixed$NTIME <-
  (nimo_fixed$OCC - 1) * 168 +                                  # which week
  ifelse(nimo_fixed$EVID == 0, round(nimo_fixed$TAD / 24) * 24, 0)  # which day

c(distinct_TIME = length(unique(nimo_fixed$TIME)),
  distinct_NTIME = length(unique(nimo_fixed$NTIME)),
  median_deviation_h = stats::median(abs(nimo_fixed$TIME - nimo_fixed$NTIME)))
```

Two terms. `(OCC - 1) * 168` places the visit in its protocol week. `round(TAD /
24) * 24` places it on its nominal day within that week, and applies only to
observations — a dose row is by definition at TAD 0.

**413 distinct clock times become 33 nominal ones**, and the median observation
sits 3 hours from its nominal slot. That is the shared grid the inferred method
could not see.

### What it achieves

```{r fixed}
knitr::kable(rbind(
  exposure(nimoData, nimo_roles(), "default, inferred grid"),
  exposure(nimo_fixed, nimo_roles("NTIME"), "constructed nominal time")
), row.names = FALSE)
```

- **Unique schedules fall from 12 to 5**, and the grid is reported as `nominal` rather
  than `derived` — snapping is exact instead of guessed.
- **Unique-visit-set exposure is zero.** Every patient still exposed is exposed
  for the one reason a grid can address, not the one it cannot.
- **Invented visit sets fall from 100% to about 17%.** Remember this one is a
  symptom rather than a harm: it fell because patients now share visit sets with
  each other, which is the thing that was missing.

That last point is the most sensitive indicator here, and it is the reason to
watch it at all. **Two visit sets are discarded either way** — the discard count
is 2 before the fix and 2 after, so on its own it would suggest nothing changed.
The invented share is what reveals that the study went from "no two patients had
a comparable schedule" to "most of them did". A number that does not move can
hide a change this large, which is a general warning about reading any one of
these columns alone.

### Why five patients are still exposed

```{r residual}
last_visit <- tapply(nimo_fixed$NTIME[nimo_fixed$EVID == 0],
                     nimo_fixed$ID[nimo_fixed$EVID == 0], max)
knitr::kable(
  data.frame(Subject = names(sort(last_visit)),
             `Last nominal visit (h)` = as.integer(sort(last_visit)),
             `In weeks` = round(sort(last_visit) / 168, 1),
             check.names = FALSE),
  row.names = FALSE,
  caption = "Follow-up length after the protocol grid is applied."
)
```

The dosing weeks are shared; the **washout tail after the last infusion is
not**. Patients stop at 1992, 2016, 2112, and 2136 hours, and the ones at the
uncommon end are alone on that final visit no matter how the earlier weeks are
gridded. Coarsening the tail further does not help — it merges genuinely
different follow-up lengths, which converts moment-exposure into visit-set
exposure and, as the sweep showed, can leave the total worse.

**Zero is probably not reachable on this study, and that is a fact about the
study rather than about the package.** Twelve patients with individual follow-up
lengths do not contain twelve indistinguishable schedules. From here the options
are to remediate the five with `flag_identifiable_subjects()` and
`remediate_identifiable_subjects()`, or to accept that these avatars carry
schedule information and keep the output inside the source data's access
controls — which is the governance model the package assumes anyway.

## 6. Did the fix damage the data?

A privacy improvement that wrecks the profiles is not an improvement. The
nominal grid moves every observation by a median of 3 hours, so it is worth
checking what that did.

```{r utility}
fitted <- suppressWarnings(suppressMessages(
  synpmx_avatar(nimo_fixed, nimo_roles("NTIME"), seed = 606)
))
knitr::kable(
  compare_pmx_distributions(nimo_fixed, fitted, nimo_roles("NTIME"))$endpoints,
  digits = 2, row.names = FALSE,
  caption = "Concentration distribution, source versus synthetic."
)
```

```{r profiles, echo=FALSE, fig.width=9, fig.height=4.5}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  panel <- function(data, label) {
    keep <- data$EVID == 0 & data$MDV == 0
    data.frame(dataset = label, subject = as.character(data$ID[keep]),
               tad = data$TAD[keep], dv = data$DV[keep])
  }
  both <- rbind(panel(nimo_fixed, "Source"), panel(fitted, "Synthetic"))
  print(
    ggplot2::ggplot(both, ggplot2::aes(tad, dv, group = subject,
                                       colour = dataset)) +
      ggplot2::geom_line(alpha = 0.3) +
      ggplot2::geom_point(alpha = 0.4, size = 0.7) +
      ggplot2::facet_wrap(~dataset) +
      ggplot2::scale_colour_manual(
        values = c(Source = "#1B6CA8", Synthetic = "#D95F02")) +
      ggplot2::labs(x = "Time after dose (hours)", y = "DV",
                    title = "nimoData with a constructed nominal time") +
      ggplot2::theme_minimal() +
      ggplot2::theme(legend.position = "none")
  )
}
```

## 7. What to take from this

1. **An inferred grid can fail completely, and it says so.** The alert is not
   decoration — 12 of 12 alone means every avatar wears a real patient's
   schedule.
2. **The failure is a property of the study.** Individually drifting visit times
   mean there is no shared clock to find. No amount of tuning invents one.
3. **Coarsening harder is not the answer.** It is non-monotone, and it pays for
   the improvement by collapsing a patient's own visits together.
4. **Recovering the protocol grid is the answer, and it is usually possible.**
   `OCC` and `TAD` were enough here. Most studies carry something equivalent
   even when they do not carry a nominal-time column outright.
5. **Check the residual rather than assuming success.** Five patients remain
   exposed, all through follow-up length, which is a different problem needing a
   different response.

## Appendix: a discrepancy found while writing this

The merge enforces that a cell never holds one patient twice, but the snap that
follows assigns each time to the **nearest cell centre**, which is a separate
decision. A time near a boundary can land in a neighbouring cell and break the
constraint the merge just enforced:

```{r snap-mismatch}
collisions <- paste(subject, snapped)[duplicated(paste(subject, snapped))]
example <- which(paste(subject, snapped) == collisions[1])
data.frame(subject = subject[example], source_time = times[example],
           snapped_to = round(snapped[example], 3))
```

Two observations of one patient, recorded less than an hour apart, pushed onto a
single time — 35 such collisions here. Small next to the 164 a one-week grid
causes, and it changes no conclusion above, but the guarantee in the code
comments is stronger than what the code delivers. Recorded as `REV-031`.
