---
title: "AVATAR synthesis on your own modeling data"
format: html
---

AVATAR blending is the default `synpmx` method. Use it when the synthetic data
*stays inside* your trusted computing environment (you are the only consumer;
governance and access controls apply). It resamples and blends whole source
subject trajectories -- faithful, and far less ceremony than the differentially
private (DP) path -- but makes **no** formal privacy guarantee. If the data may cross
a trust boundary (e.g. externally shared data), use `try_dp_calibrated.R` instead or explore
the other algorithms as well (Prior, Empirical); see the `synpmx-privacy` vignette for the decision rule.

Prefer a plain script? `knitr::purl("try_avatar.qmd")` writes the R chunks out
as a `.R` file. There is one template; the script is generated from it.

To use: edit the two configuration chunks -- the data path, and the column
roles -- and run. You should not need to touch anything below them.

```{r}
#| label: setup
#| message: false
library(synpmx)
library(ggplot2)
# If synpmx is not installed here: devtools::load_all("/path/to/synpmx")
```

## Configuration 1 --- where the data is

```{r}
#| label: config
DATA_PATH  <- "data/your_modeling_dataset.csv"
OUT_DIR    <- "output"
SEED       <- 1234        # reproducibility seed; the caller's RNG is untouched
N_SUBJECTS <- NULL        # NULL keeps the source cohort size
```

## Configuration 2 --- what the columns mean

Declare only the roles your dataset has. Every column **not** named by a role is
dropped -- the run prints which -- so a stray identifier cannot leak out of a
real subject by being forgotten.

```{r}
#| label: roles
ROLES <- pmx_roles(
  id       = "ID",
  time     = "TIME",
  dv       = "DV",
  amt      = "AMT",
  evid     = "EVID",
  dvid     = "DVID",   # one endpoint-key column, or several that agree:
                       #   dvid = c("YTYPE", "NAME") declares a numeric key and
                       #   a character label; both are checked and carried out.
  cmt      = "CMT",
  mdv      = "MDV",
  rate     = NULL,     # infusion RATE column, if any
  occasion = NULL,     # occasion column, if TIME resets by occasion
  cens     = NULL,     # below-limit indicator (Monolix: 1 below, -1 above), if
  limit    = NULL,     #   the study has BLQ data; limit is the interval's far end

  # BLENDED into genuinely new values across neighbours -- use for baselines you
  # want synthesized (weight, age). Do NOT list a baseline that is itself an
  # endpoint's starting value (e.g. baseline B cells beside a B-cell endpoint):
  # covariates and endpoints are blended independently, so the two would not
  # agree (see the AVATAR mathematics article, and REV-022).
  covariates = c("WT", "AGE", "SEX"),

  # COPIED verbatim from the subject that supplied the doses, so it stays
  # coherent with them -- treatment arm, dose group, study id, a character
  # endpoint label. A kept value is one real subject's real value.
  keep     = NULL
)
```

## Read and validate the source

```{r}
#| label: read-validate
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(DATA_PATH)) {
  stop("DATA_PATH does not exist: ",
       normalizePath(DATA_PATH, mustWork = FALSE),
       "\nEdit DATA_PATH in the configuration chunk.", call. = FALSE)
}
raw   <- read.csv(DATA_PATH, stringsAsFactors = FALSE)
roles <- ROLES

report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
cat("Source is valid:", length(unique(raw[[roles$id]])), "subjects.\n")
```

## Synthesize the data

```{r}
#| label: synthesize
synthetic <- synpmx_avatar(raw, roles, n_subjects = N_SUBJECTS, seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)

fresh_ids <- length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0L
cat("Generated", nrow(synthetic), "rows for",
    length(unique(synthetic[[roles$id]])), "subjects.\n")
cat("Every synthetic ID is new:", fresh_ids, "\n")

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"),
          row.names = FALSE)
```

## Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r}
#| label: diagnostic
#| fig-width: 9
#| fig-height: 6
# The plot adapts to however many endpoints there are -- one PK, one PD, two,
# more -- because faceting by endpoint grows a column per endpoint on its own.
# You do not have to know the count in advance.
#
# Facet strips read best from a character label. If you declared several `dvid`
# columns, the last one (often a character NAME) is used; otherwise the single
# dvid column; otherwise a constant.
label_col <- if (length(roles$dvid) > 1L) {
  roles$dvid[[length(roles$dvid)]]
} else if (!is.null(roles$dvid)) roles$dvid[[1L]] else NULL

frame <- function(d, label) {
  keep <- d[[roles$evid]] == 0 & !is.na(d[[roles$dv]])   # observations only
  data.frame(
    dataset  = label,
    id       = d[[roles$id]][keep],
    time     = as.numeric(d[[roles$time]][keep]),
    dv       = as.numeric(d[[roles$dv]][keep]),
    endpoint = if (is.null(label_col)) "DV" else
      as.character(d[[label_col]][keep]),
    stringsAsFactors = FALSE
  )
}
compare <- rbind(frame(raw, "source"), frame(synthetic, "synthetic"))

ggplot(compare, aes(time, dv, group = id)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.4, size = 0.6) +
  # rows: source vs synthetic. columns: one per endpoint, however many.
  facet_grid(dataset ~ endpoint, scales = "free_y") +
  labs(title = "RESTRICTED: source vs AVATAR synthetic",
       subtitle = "do not export", x = "Time", y = "DV") +
  theme_bw()
# Concentrations often read better on a log axis. Add `+ scale_y_log10()` above
# if every DV is positive; a log scale drops zeros and negatives.
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories and is for trusted-environment use only.
