---
title: "AVATAR synthesis on CPIT565A1 modeling data"
format: html
---

Try the AVATAR blending generator on a real modeling dataset.

This is the **default** method. Use it when the synthetic data *reaches no one
the source data could not* -- the same access controls and confidentiality
obligations still apply, which includes carrying it to your own workstation for
model development. It resamples and blends whole source subject
trajectories -- more faithful and far less ceremony than the differentially
private path -- but makes **no** formal privacy guarantee. If the data may reach
someone new, use `try_dp_calibrated.R` instead. See the `synpmx-privacy`
vignette for the decision rule (which method when).

This document runs in the safe computing environment, in a gitignored folder so
nothing here can be committed.

Workflow:

1. fill in the CONFIG chunk (just the column roles)
2. compare the synthetic table to the source

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(xgxr)
xgx_theme_set()

# If not installed here: devtools::load_all("/path/to/synpmx")
# This load_all line only needs to be run once, so you may comment out if you're 
# Trying this out on your own dataset

devtools::load_all()
library(synpmx)

```

## Configuration

Everything you edit is in this chunk.

```{r}
#| label: config
user <- Sys.getenv("USER")
gps_prefix <- paste0("/view/", user, "_view/vob/")
filename   <- file.path(gps_prefix, "CPIT565A1/mas/mas_1/data/output_001/", "mt96714_poppkpd.csv")
data_in    <- read.csv(filename, stringsAsFactors = FALSE)

DATA       <- data_in |> filter(NAME %in% c("PIT565 Dose", "PIT565 PK", "CRP"), STUDYID == "CPIT565A12101")
OUT_DIR   <- "output_pit565a"
SEED      <- 1234        # reproducibility seed; the caller's RNG is untouched

# plot initial data
ggplot(DATA |> filter(NAME != "PIT565 Dose"), aes(x = TIME, y = LIDV, group = ID, color = TRT)) + 
  geom_point(alpha = 0.5) + 
  geom_line(alpha = 0.5) + 
  facet_wrap(~NAME) + 
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10()
```

Map your columns onto PMX roles. Only the roles the real dataset has. AVATAR
needs no model, priors, design, or budget -- just the column meanings. List
baseline covariates in `covariates` and they are blended from compatible donors
like everything else.

```{r}
#| label: roles
ROLES <- pmx_roles(
  id       = "USUBJID",
  time     = "TIME",
  dv       = "LIDV",
  amt      = "AMT",
  evid     = "EVID",
  # Two columns encode the same endpoint: numeric YTYPE and character NAME.
  # Declare both. YTYPE (first) is the grouping key; validation checks NAME is a
  # consistent 1:1 mapping with it and errors if not, and both are carried out.
  dvid     = c("YTYPE", "NAME"),
  cmt      = "CMT",
  mdv      = "MDV",

  # DECLARE NOMINAL TIME IF THE STUDY HAS IT. `coarsen_time = TRUE` (the
  # default) collapses subjects onto a shared visit grid so no avatar carries one
  # real subject's exact visit schedule. With `nominal_time` declared it snaps to
  # the protocol grid exactly; without it the grid is inferred from the pooled
  # times, which is best-effort and alerts loudly when it cannot collapse a
  # subject. Under ACTUAL recorded times this is the difference between the
  # mechanism working and not working -- see `skeleton_uniqueness()`.
  # Gaps are fine: rows with no nominal time fall back to the inferred grid,
  # and the run reports the grid as "mixed". A column that is entirely missing
  # should be left undeclared instead.
  nominal_time = NULL, # e.g. "NTIME"
  tad      = NULL,     # time-after-dose column; recomputed after generation
  # ASSIGNED STRATUM -- treatment arm, dose group, cohort. Carried verbatim like
  # `keep`, and additionally used to group the pool of attendance patterns an
  # avatar may draw from, so an avatar only inherits a visit pattern from its own
  # arm. Not a blending barrier: donors are still borrowed across strata to reach
  # the donor floor, because only route of administration is absolute.
  #
  # Whether to declare an arm here is a real trade, not an obvious win. Declaring
  # it is more faithful -- arms with different visit schedules stay separate --
  # but it splits the pattern pools, so fewer patterns clear `min_pattern_share`
  # and more real dropout patterns are discarded. The run reports how many were
  # lost either way; try it both ways and compare.
  # Gaps are fine here too: subjects with no recorded value form their own
  # stratum and the run warns. A column that *varies* within a subject is an
  # error -- it is not that subject's assignment. A column that does not exist
  # in this study is also an error, naming itself: the two PIT565 parts do not
  # carry identical columns, so check this block against the study you are
  # running rather than assuming it carries over.
  subject_properties = NULL,   # e.g. "TRT", or c("TRT", "COHORT")
  rate     = NULL,           # set if you have infusion RATE rows
  occasion = NULL,           # set if TIME resets by occasion
  # If the data has below-limit-of-quantification (BLQ) records, declare cens
  # (and limit, if intervals are used). AVATAR then reconstructs censoring
  # coherently -- DV at the limit, CENS/LIMIT agreeing -- instead of copying a
  # flag that no longer matches the blended DV. Leave NULL if there is no BLQ.
  cens     = "CENS",           # e.g. "CENS"  (Monolix: 1 = below, -1 = above)
  limit    = NULL,           # e.g. "LIMIT" (other end of an interval)
  # BLENDED into new values across donors. B0 (baseline B cells) is deliberately
  # left out: it is the same quantity as the Circ B Cells endpoint, and AVATAR
  # blends covariates and endpoints independently, so a synthetic B0 would not
  # match its own generated trajectory (REV-022). Read the baseline off the
  # trajectory instead if you need it.
  covariates = c("WEIGHT0", "AGE0", "SEX", "RACE"),
  # keep = copied verbatim from the subject that supplied the doses, so it stays
  # coherent with them. Use it for pass-through values with no other meaning --
  # a study id, a character label. An assigned *stratum* (treatment arm, dose
  # group) is better declared in `subject_properties` above, which carries it
  # the same way and also groups the attendance-pattern pools.
  keep     = c("STUDYIDN", "STUDYID", "TRT", "TRTN")
  # Every column NOT named by a role above is dropped. That is deliberate: a
  # secondary identifier you forget cannot leak out of a real subject. The run
  # prints exactly which columns it dropped.
)

# Number of synthetic subjects. NULL keeps the source cohort size.
N_SUBJECTS <- NULL
```

## Workflow

You should not need to edit below here.

```{r}
#| label: read-source
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
raw   <- DATA
roles <- ROLES
```

### Source structural validation

```{r}
#| label: validate-source
report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
message("  valid; ", length(unique(raw[[roles$id]])), " subjects")
```

### Source skeleton exposure

Run before generating. The event skeleton is copied verbatim from one anchor, so
a subject holding the only copy of its observation time vector hands that
schedule to every avatar anchored on it.

Read `alone` first, then why:

- `min_time_share == 1` -- observed at a moment no other subject was. This is
  what `coarsen_time` collapses, and a residue here means the *inferred* grid
  found no shared schedule. Declaring `nominal_time` is the fix.
- `min_time_share >= 2` but `alone` -- every time is shared and only the
  *pattern* of attended visits is unique. Dropout and missed visits. No grid
  touches this; `min_pattern_share` is what stops it being reproduced.
- `n_obs_class == 1` -- a sample count nobody else has.
- `signature_class == 1` -- dose structure or amount nobody else has.

Record these in `scripts_private/README.md`'s study inventory.

```{r}
#| label: skeleton-uniqueness
exposure <- skeleton_uniqueness(raw, roles)
exposure
```

### Synthesize (AVATAR blending)

```{r}
#| label: synthesize
synthetic <- synpmx_avatar(raw, roles, seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)
message("  generated ", nrow(synthetic), " rows for ",
        length(unique(synthetic[[roles$id]])), " subjects; new identifiers: ",
        length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0)

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"), row.names = FALSE)
```

### What the mechanisms actually did

Every masking mechanism records what it did, and two of them record what it
cost. Copy these into the study inventory in `scripts_private/README.md`.

```{r}
#| label: mechanism-report
settings <- attr(synthetic, "pmx_settings")
data.frame(
  quantity = c(
    "time grid used",
    "subjects still alone after coarsening",
    "  ... seen at a unique moment (declare nominal_time to fix)",
    "  ... unique attendance pattern (dropout; min_pattern_share's job)",
    "dose basis detected",
    "dose levels found",
    "attendance patterns in the source",
    "  ... discarded as too rare to reuse",
    "  ... real subjects who held those",
    "avatars given a sampled pattern",
    "  ... of those, arrangement generated not reused",
    "donor floor: mean effective donors",
    "donor cap binding fraction"
  ),
  value = c(
    settings$time_grid,
    settings$exposure_alone,
    settings$exposure_unique_moment,
    settings$exposure_unique_pattern,
    ifelse(is.na(settings$dose_basis), "none (amounts left alone)",
           settings$dose_basis),
    ifelse(is.na(settings$dose_basis), "-",
           paste(signif(settings$dose_levels, 4), collapse = ", ")),
    settings$patterns_total,
    settings$patterns_dropped,
    settings$subjects_with_dropped_pattern,
    settings$pattern_sampled_fraction,
    round(settings$pattern_generated_fraction, 2),
    round(settings$mean_effective_donors, 2),
    round(settings$cap_binding_fraction, 2)
  ),
  stringsAsFactors = FALSE
) |> knitr::kable()
```

At the default floor of 2, `discarded` and `real subjects who held those` are
necessarily equal: a pattern is discarded exactly when fewer than two patients
share it, so every discarded pattern has exactly one holder. They diverge only at
a floor of 3 or more.

`discarded as too rare to reuse` is the one to look at hardest. Those are real
dropout and dose-interruption patterns that will not appear in the synthetic
data, because too few patients shared them. Discarding them is what stops an
avatar carrying a schedule traceable to one person, but if this study's
interruptions matter, lower `min_pattern_share` (2 is the lowest value that
still guarantees no synthetic patient has a schedule unique to a real one) or
set it to `1` to keep every pattern and give up that guarantee.

### Restricted diagnostic: distribution summary

Source-derived. Keep it in the safe environment; do not export these tables.

`compare_pmx_distributions()` summarizes the dependent variable per endpoint and
each baseline covariate, source against synthetic -- a quick numeric range check
before the plot. Expect the summaries to line up in magnitude and shape, not to
match exactly, since AVATAR does not reproduce the source distribution.

```{r}
#| label: distribution-summary
#| results: asis
# Rendered with knitr::kable() so the tables come out as tables when this
# document is knitted, and stay readable when run chunk-by-chunk. Rounding is
# for display only; the returned object keeps the exact values.
kable_distributions <- function(x) {
  section <- function(df, caption) {
    if (is.null(df)) return(invisible())
    numeric_cols <- vapply(df, is.numeric, logical(1))
    df[numeric_cols] <- lapply(df[numeric_cols], signif, digits = 4)
    print(knitr::kable(df, caption = caption))
    cat("\n")
  }
  section(x$endpoints,
          "RESTRICTED -- endpoints (dependent variable on observation rows)")
  section(x$covariates_numeric,
          "RESTRICTED -- continuous covariates (baseline, per subject)")
  section(x$covariates_categorical,
          "RESTRICTED -- categorical covariates (baseline, per subject)")
  invisible(x)
}

kable_distributions(compare_pmx_distributions(raw, synthetic, roles))
```

### Restricted diagnostic: how close did the values land?

Everything above concerns structure. Blending is what protects the **values**,
and this is its measurement: does each subject's nearest neighbour lie in its own
dataset or the other one? Near 0.5 means a synthetic subject is no more like a
real subject than one real subject is like another -- the target. Toward 0 means
memorisation.

Read the null interval before the point estimate. It is built by splitting *this*
source cohort in half and running the identical statistic, so every small-sample
artefact cancels -- but at a few dozen subjects it is wide. Inside the interval
means **nothing was detected**, not that nothing is there.

```{r}
#| label: proximity
compare_pmx_proximity(raw, synthetic, roles)
```

### Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r}
#| label: diagnostic

  raw_keep = raw[,names(synthetic)]

  compare = rbind(raw_keep |> mutate(DATA = "source"), 
                  synthetic |> mutate(DATA = "synthetic"))
  
  g = ggplot(compare |> filter(NAME != "PIT565 Dose"), aes(x = TIME, y = LIDV, group = USUBJID, color = TRT)) + 
    geom_point(alpha = 0.5) + 
    geom_line(alpha = 0.5) + 
    facet_wrap(DATA~NAME) + 
    xgx_scale_x_time_units("days") +
    xgx_scale_y_log10()

  g
  
  #ggsave(file.path(OUT_DIR, "RESTRICTED_source_vs_synthetic.pdf"), g, 
  #       width = 7, height = 6, dpi = 110)
  
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories, so it stays under the source data's own access
controls and obligations wherever it is used — your own workstation included, a
partner or vendor not.
