---
title: "AVATAR synthesis on your own modeling data"
format: html
---

AVATAR blending is the default `synpmx` method. Use it when the synthetic data
*reaches no one the source data could not* -- the same access controls and
confidentiality obligations still apply, which includes carrying it to your own
workstation for model development. It resamples and blends whole source
subject trajectories -- faithful, and far less ceremony than the differentially
private (DP) path -- but makes **no** formal privacy guarantee. If the data may reach
someone new (e.g. externally shared data), use `try_dp_calibrated.R` instead or explore
the other algorithms as well (Prior, Empirical); see the `synpmx-privacy` vignette for the decision rule.

Prefer a plain script? `knitr::purl("try_avatar.qmd")` writes the R chunks out
as a `.R` file. There is one template; the script is generated from it.

To use: edit the two configuration chunks -- the data path, and the column
roles -- and run. You should not need to touch anything below them.

```{r}
#| label: setup
#| message: false
library(synpmx)
library(ggplot2)
# If synpmx is not installed here: devtools::load_all("/path/to/synpmx")
```

## Configuration 1 --- where the data is

```{r}
#| label: config
DATA_PATH  <- "data/your_modeling_dataset.csv"
OUT_DIR    <- "output"
SEED       <- 1234        # reproducibility seed; the caller's RNG is untouched
N_SUBJECTS <- NULL        # NULL keeps the source cohort size
```

## Configuration 2 --- what the columns mean

Declare only the roles your dataset has. Every column **not** named by a role is
dropped -- the run prints which -- so a stray identifier cannot leak out of a
real subject by being forgotten.

```{r}
#| label: roles
ROLES <- pmx_roles(
  id       = "ID",
  time     = "TIME",
  dv       = "DV",
  amt      = "AMT",
  evid     = "EVID",
  dvid     = "DVID",   # one endpoint-key column, or several that agree:
                       #   dvid = c("YTYPE", "NAME") declares a numeric key and
                       #   a character label; both are checked and carried out.
  cmt      = "CMT",
  mdv      = "MDV",

  # DECLARE NOMINAL TIME IF THE STUDY HAS IT. `coarsen_time = TRUE` (the
  # default) collapses subjects onto a shared visit grid so no avatar carries one
  # real subject's exact visit schedule. With `nominal_time` declared it snaps to
  # the protocol grid exactly; without it the grid is inferred from the pooled
  # times, which is best-effort and alerts loudly when it cannot collapse a
  # subject. Under ACTUAL recorded times this is the difference between the
  # mechanism working and not working -- see `skeleton_uniqueness()`.
  # Gaps are fine: rows with no nominal time fall back to the inferred grid,
  # and the run reports the grid as "mixed". A column that is entirely missing
  # should be left undeclared instead.
  nominal_time = NULL, # e.g. "NTIME"
  tad      = NULL,     # time-after-dose column; recomputed after generation
  # ASSIGNED STRATUM -- treatment arm, dose group, cohort. Carried verbatim like
  # `keep`, and additionally used to group the pool of attendance patterns an
  # avatar may draw from, so an avatar only inherits a visit pattern from its own
  # arm. Not a blending barrier: donors are still borrowed across strata to reach
  # the donor floor, because only route of administration is absolute.
  #
  # Whether to declare an arm here is a real trade, not an obvious win. Declaring
  # it is more faithful -- arms with different visit schedules stay separate --
  # but it splits the pattern pools, so fewer patterns clear `min_pattern_share`
  # and more real dropout patterns are discarded. The run reports how many were
  # lost either way; try it both ways and compare.
  # Gaps are fine here too: subjects with no recorded value form their own
  # stratum and the run warns. A column that *varies* within a subject is an
  # error -- it is not that subject's assignment. A column that does not exist
  # in this study is also an error, naming itself: the two PIT565 parts do not
  # carry identical columns, so check this block against the study you are
  # running rather than assuming it carries over.
  subject_properties = NULL,   # e.g. "TRT", or c("TRT", "COHORT")
  rate     = NULL,     # infusion RATE column, if any
  occasion = NULL,     # occasion column, if TIME resets by occasion
  cens     = NULL,     # below-limit indicator (Monolix: 1 below, -1 above), if
  limit    = NULL,     #   the study has BLQ data; limit is the interval's far end

  # BLENDED into genuinely new values across neighbours -- use for baselines you
  # want synthesized (weight, age). Do NOT list a baseline that is itself an
  # endpoint's starting value (e.g. baseline B cells beside a B-cell endpoint):
  # covariates and endpoints are blended independently, so the two would not
  # agree (see the AVATAR mathematics article, and REV-022).
  covariates = c("WT", "AGE", "SEX"),

  # COPIED verbatim from the subject that supplied the doses, so it stays
  # coherent with them -- a study id, a character endpoint label. A kept value is
  # one real subject's real value. An assigned *stratum* (treatment arm, dose
  # group) belongs in `subject_properties` above instead: carried the same way,
  # and used to group the attendance-pattern pools.
  keep     = NULL
)
```

## Read and validate the source

```{r}
#| label: read-validate
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(DATA_PATH)) {
  stop("DATA_PATH does not exist: ",
       normalizePath(DATA_PATH, mustWork = FALSE),
       "\nEdit DATA_PATH in the configuration chunk.", call. = FALSE)
}
raw   <- read.csv(DATA_PATH, stringsAsFactors = FALSE)
roles <- ROLES

report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
cat("Source is valid:", length(unique(raw[[roles$id]])), "subjects.\n")
```

## How exposed is the source skeleton?

Run this BEFORE generating. `synpmx_avatar()` copies each avatar's event
skeleton verbatim from one anchor, so a source subject holding the only copy of
its observation time vector hands that schedule to every avatar anchored on it.

Read `alone` first, then why:

- `min_time_share == 1` -- observed at a moment no other subject was. This is
  what `coarsen_time` collapses, and a residue here means the *inferred* grid
  found no shared schedule. Declaring `nominal_time` is the fix.
- `min_time_share >= 2` but `alone` -- every time is shared and only the
  *pattern* of attended visits is unique. Dropout and missed visits. No grid
  touches this; `min_pattern_share` is what stops it being reproduced.
- `n_obs_class == 1` -- a sample count nobody else has.
- `signature_class == 1` -- dose structure or amount nobody else has.

Record the numbers in `scripts_private/README.md`'s study inventory.

```{r}
#| label: skeleton-uniqueness
exposure <- skeleton_uniqueness(raw, roles)
exposure
```

## Synthesize the data

```{r}
#| label: synthesize
synthetic <- synpmx_avatar(raw, roles, n_subjects = N_SUBJECTS, seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)

fresh_ids <- length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0L
cat("Generated", nrow(synthetic), "rows for",
    length(unique(synthetic[[roles$id]])), "subjects.\n")
cat("Every synthetic ID is new:", fresh_ids, "\n")

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"),
          row.names = FALSE)
```

### What the mechanisms actually did

Every masking mechanism records what it did, and two of them record what it
cost. Copy these into the study inventory in `scripts_private/README.md`.

```{r}
#| label: mechanism-report
settings <- attr(synthetic, "pmx_settings")
data.frame(
  quantity = c(
    "time grid used",
    "subjects still alone after coarsening",
    "  ... seen at a unique moment (declare nominal_time to fix)",
    "  ... unique attendance pattern (dropout; min_pattern_share's job)",
    "dose basis detected",
    "dose levels found",
    "attendance patterns in the source",
    "  ... discarded as too rare to reuse",
    "  ... real subjects who held those",
    "avatars given a sampled pattern",
    "  ... of those, arrangement generated not reused",
    "donor floor: mean effective donors",
    "donor cap binding fraction"
  ),
  value = c(
    settings$time_grid,
    settings$exposure_alone,
    settings$exposure_unique_moment,
    settings$exposure_unique_pattern,
    ifelse(is.na(settings$dose_basis), "none (amounts left alone)",
           settings$dose_basis),
    ifelse(is.na(settings$dose_basis), "-",
           paste(signif(settings$dose_levels, 4), collapse = ", ")),
    settings$patterns_total,
    settings$patterns_dropped,
    settings$subjects_with_dropped_pattern,
    settings$pattern_sampled_fraction,
    round(settings$pattern_generated_fraction, 2),
    round(settings$mean_effective_donors, 2),
    round(settings$cap_binding_fraction, 2)
  ),
  stringsAsFactors = FALSE
) |> knitr::kable()
```

At the default floor of 2, `discarded` and `real subjects who held those` are
necessarily equal: a pattern is discarded exactly when fewer than two patients
share it, so every discarded pattern has exactly one holder. They diverge only at
a floor of 3 or more.

`discarded as too rare to reuse` is the one to look at hardest. Those are real
dropout and dose-interruption patterns that will not appear in the synthetic
data, because too few patients shared them. Discarding them is what stops an
avatar carrying a schedule traceable to one person, but if this study's
interruptions matter, lower `min_pattern_share` (2 is the lowest value that
still guarantees no synthetic patient has a schedule unique to a real one) or
set it to `1` to keep every pattern and give up that guarantee.

## Restricted diagnostic: distribution summary

Source-derived. Keep it in the safe environment; do not export these tables.

Numbers before pictures: `compare_pmx_distributions()` summarizes the dependent
variable per endpoint and each baseline covariate, source against synthetic.
Read it as a range check -- expect the summaries to line up in magnitude and
shape, not to match exactly, since AVATAR does not reproduce the source
distribution.

```{r}
#| label: distribution-summary
#| results: asis
# Rendered with knitr::kable() so the tables come out as tables when this
# document is knitted, and stay readable when run chunk-by-chunk. Rounding is
# for display only; the returned object keeps the exact values.
kable_distributions <- function(x) {
  section <- function(df, caption) {
    if (is.null(df)) return(invisible())
    numeric_cols <- vapply(df, is.numeric, logical(1))
    df[numeric_cols] <- lapply(df[numeric_cols], signif, digits = 4)
    print(knitr::kable(df, caption = caption))
    cat("\n")
  }
  section(x$endpoints,
          "RESTRICTED -- endpoints (dependent variable on observation rows)")
  section(x$covariates_numeric,
          "RESTRICTED -- continuous covariates (baseline, per subject)")
  section(x$covariates_categorical,
          "RESTRICTED -- categorical covariates (baseline, per subject)")
  invisible(x)
}

kable_distributions(compare_pmx_distributions(raw, synthetic, roles))
```

### Restricted diagnostic: how close did the values land?

Everything above concerns structure. Blending is what protects the **values**,
and this is its measurement: does each subject's nearest neighbour lie in its own
dataset or the other one? Near 0.5 means a synthetic subject is no more like a
real subject than one real subject is like another -- the target. Toward 0 means
memorisation.

Read the null interval before the point estimate. It is built by splitting *this*
source cohort in half and running the identical statistic, so every small-sample
artefact cancels -- but at a few dozen subjects it is wide. Inside the interval
means **nothing was detected**, not that nothing is there.

```{r}
#| label: proximity
compare_pmx_proximity(raw, synthetic, roles)
```

## Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r}
#| label: diagnostic
#| fig-width: 9
#| fig-height: 6
# The plot adapts to however many endpoints there are -- one PK, one PD, two,
# more -- because faceting by endpoint grows a column per endpoint on its own.
# You do not have to know the count in advance.
#
# Facet strips read best from a character label. If you declared several `dvid`
# columns, the last one (often a character NAME) is used; otherwise the single
# dvid column; otherwise a constant.
label_col <- if (length(roles$dvid) > 1L) {
  roles$dvid[[length(roles$dvid)]]
} else if (!is.null(roles$dvid)) roles$dvid[[1L]] else NULL

frame <- function(d, label) {
  keep <- d[[roles$evid]] == 0 & !is.na(d[[roles$dv]])   # observations only
  data.frame(
    dataset  = label,
    id       = d[[roles$id]][keep],
    time     = as.numeric(d[[roles$time]][keep]),
    dv       = as.numeric(d[[roles$dv]][keep]),
    endpoint = if (is.null(label_col)) "DV" else
      as.character(d[[label_col]][keep]),
    stringsAsFactors = FALSE
  )
}
compare <- rbind(frame(raw, "source"), frame(synthetic, "synthetic"))

ggplot(compare, aes(time, dv, group = id)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.4, size = 0.6) +
  # rows: source vs synthetic. columns: one per endpoint, however many.
  facet_grid(dataset ~ endpoint, scales = "free_y") +
  labs(title = "RESTRICTED: source vs AVATAR synthetic",
       subtitle = "do not export", x = "Time", y = "DV") +
  theme_bw()
# Concentrations often read better on a log axis. Add `+ scale_y_log10()` above
# if every DV is positive; a log scale drops zeros and negatives.
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories, so it stays under the source data's own access
controls and obligations wherever it is used — your own workstation included, a
partner or vendor not.
