---
title: "AVATAR synthesis on ECI modeling data"
format:
  html:
    toc: true
    # Everything except the configuration and the generation call is reporting
    # code, and reading it is not the point of this document. Folded, not
    # hidden: click "Show code" on any block to see exactly what produced a
    # number. The chunks worth reading unfolded set `code-fold: false`.
    code-fold: true
    code-summary: "Show code"
---

Try the AVATAR blending generator on a real modeling dataset.

This is the **default** method. Use it when the synthetic data *reaches no one
the source data could not* -- the same access controls and confidentiality
obligations still apply, which includes carrying it to your own workstation for
model development. It resamples and blends whole source subject
trajectories -- more faithful and far less ceremony than the differentially
private path -- but makes **no** formal privacy guarantee. If the data may reach
someone new, use `try_dp_calibrated.R` instead. See the `synpmx-privacy`
vignette for the decision rule (which method when).

This document runs in the safe computing environment, in a gitignored folder so
nothing here can be committed.

Workflow:

1. fill in the CONFIG chunk (just the column roles)
2. compare the synthetic table to the source

```{r}
#| label: setup
#| code-fold: false
#| message: false
library(tidyverse)
library(xgxr)
xgx_theme_set()

# Only needed once per session; comment out if synpmx is installed.

devtools::load_all()
library(synpmx)

```

## Configuration

Everything you edit is in this chunk.

```{r}
#| label: config
#| code-fold: false
user <- Sys.getenv("USER")
gps_prefix <- paste0("/view/", user, "_view/vob/")
filename   <- file.path(gps_prefix, "/CECI830A1/mas/mas_1/data/output_001/MT28454_R.csv")
data_in    <- read.csv(filename, stringsAsFactors = FALSE)

DATA       <- data_in
OUT_DIR   <- "output_eci"
SEED      <- 1234        # reproducibility seed; the caller's RNG is untouched

# plot initial data
# Observation rows only.
ggplot(DATA |> filter(EVID == 0, !is.na(LIDV)), aes(x = TIME, y = LIDV, group = ID, color = TRT)) + 
  geom_point(alpha = 0.5) + 
  geom_line(alpha = 0.5) + 
  facet_wrap(~NAME) + 
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10()
```

Map your columns onto PMX roles. Only the roles the real dataset has. AVATAR
needs no model, priors, design, or budget -- just the column meanings. List
baseline covariates in `covariates` and they are blended from compatible donors
like everything else.

```{r}
#| label: roles
#| code-fold: false
ROLES <- pmx_roles(
  id       = "ID",
  time     = "TIME",
  dv       = "LIDV",
  amt      = "AMT",
  evid     = "EVID",
  dvid     = c("YTYPE", "NAME"),  # first is the key; the rest checked 1:1 against it
  cmt      = "CMT",
  mdv      = "MDV",
  # Declare it if the study has it. Rows that are missing one fall back to an
  # inferred grid, and the run then reports the grid as "mixed".
  nominal_time = NULL, # e.g. "NTIME"
  tad      = NULL,     # time-after-dose column; recomputed after generation
  # Treatment arm / dose group. Also splits the pool of visit sets an avatar may
  # draw from: declaring it is more faithful, but more rare visit sets are then
  # too thinly held to reuse. Try it both ways and compare the run reports.
  strata = c("TRT", "TRTN"),  # NULL pools the arms
  rate     = NULL,           # set if you have infusion RATE rows
  occasion = NULL,           # set if TIME resets by occasion
  # Declared, censoring is rebuilt to agree with the blended DV; left NULL, a
  # BLQ flag would survive that no longer matches its own value.
  cens     = "CENS",           # Monolix: 1 = below, -1 = above
  limit    = NULL,           # e.g. "LIMIT" (other end of an interval)
  # Blended across donors. Leave out any baseline that is the same quantity as
  # an endpoint -- the two are blended independently, so they would disagree
  # (REV-022). Read that baseline off the generated trajectory instead.
  covariates = c("WEIGHT0", "AGE0", "SEX", "RACE"),
  # DECLARE THIS IF THE STUDY DOSES ON WEIGHT OR BSA. Names the covariate the
  # amount is a fixed multiple of, so each avatar's dose is rebuilt from its own
  # blended value. Left NULL the run tries to infer it, and infers conservatively
  # -- vial rounding or patients escalating by different factors defeat the test,
  # and then amounts are copied verbatim and still encode one real patient's
  # weight. Must also appear in `covariates` above. The run report says which
  # path was taken.
  dose_covariate = NULL,   # e.g. "WEIGHT0"
  keep     = c("STUDYIDN", "STUDYID")  # pass-through only; a stratum belongs above
  # Any column no role names is dropped, so a forgotten identifier cannot ride
  # out of a real patient. The run prints which columns went.
)

# Number of synthetic subjects. NULL keeps the source cohort size.
N_SUBJECTS <- NULL
```

## Workflow

You should not need to edit below here.

```{r}
#| label: read-source
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
raw   <- DATA
roles <- ROLES
```

### Source structural validation

```{r}
#| label: validate-source
report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
message("  valid; ", length(unique(raw[[roles$id]])), " subjects")
```

### How identifying is this study's visit schedule?

`synpmx_avatar()` copies its anchor's event skeleton **verbatim**, so a patient
holding the only copy of their observation times hands that schedule to every
avatar anchored on them. Run it twice: on the times as given, and on the grid
the generator actually builds. The second is the one that matters, and the
difference between them is what coarsening bought.

```{r}
#| label: skeleton-uniqueness-before
exposure_before <- skeleton_uniqueness(raw, roles)
exposure_before
```

```{r}
#| label: skeleton-uniqueness-after
exposure <- skeleton_uniqueness(raw, roles, coarsen_time = TRUE)
exposure
```

Split the headline by cause before deciding it is bad. **A one-off observation
time** is a real problem and a fixable one -- declare `nominal_time`. **A
unique set of attended visits** means visits are missing rather than moved --
a missed visit, a discontinuation, or follow-up that has not reached them yet.
No grid at any resolution touches that, and `min_pattern_share` handles it
downstream. A large count there with a small count above is the normal result
for a study with any of those, not a failure.

The picture answers the same question faster. Vertical stripes mean patients
share visits, a ragged right edge is follow-up ending, and a red bar in the
lower panel
is a moment only one patient was ever sampled at -- exactly what would be
copied onto an avatar.

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map
plot_pmx_schedule(raw, roles)
```

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map-raw
plot_pmx_schedule(raw, roles, coarsen_time = FALSE)
```

Record these in `scripts_private/README.md`'s study inventory.

### Synthesize (AVATAR blending)

```{r}
#| label: synthesize
#| code-fold: false
synthetic <- synpmx_avatar(raw, roles, n_subjects = N_SUBJECTS,
                           seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)
message("  generated ", nrow(synthetic), " rows for ",
        length(unique(synthetic[[roles$id]])), " subjects; new identifiers: ",
        length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0)

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"), row.names = FALSE)
```

### What the mechanisms actually did

Each row carries the sentence that says what its number means. Copy this into
the study inventory in `scripts_private/README.md`.

```{r}
#| label: mechanism-report
pmx_masking_report(synthetic, raw, roles)
```

Three rows to stop on.

**Amounts recomputed from a covariate** answers "did it find the weight-based
dosing?" outright, and when the answer is no it names which covariates were
tried and what failed. Detection is deliberately conservative -- it refuses
rather than rewrite amounts in a study that is not dose-proportional. A "no" on
a study that *is* weight-based means `AMT` is copied verbatim from the anchor,
so each avatar's implied mg/kg is wrong and the amount still encodes one real
patient's weight.

**Shared by too few patients, so not reused** is the fidelity cost: real
patterns of missing visits and interruptions that will not appear in the
output. Refusing
them is what stops an avatar carrying a schedule traceable to one person.
`min_pattern_share = 2` is the lowest value that keeps that guarantee; `1` gives
it up. At 2 this row necessarily equals the one below it, since a set is
discarded exactly when one patient holds it.

**Avatars keeping their anchor's own visit set** undoes the row above when it is
high: those avatars carry one real patient's absences. The run alerts past 10%,
and the fix is `nominal_time` or a lower `min_pattern_share`.

### Restricted diagnostic: distribution summary

Source-derived. Keep it in the safe environment; do not export these tables.

`compare_pmx_distributions()` summarizes the dependent variable per endpoint and
each baseline covariate, source against synthetic -- a quick numeric range check
before the plot. Expect the summaries to line up in magnitude and shape, not to
match exactly, since AVATAR does not reproduce the source distribution.

```{r}
#| label: distribution-summary
compare_pmx_distributions(raw, synthetic, roles)
```

### Restricted diagnostic: how close did the values land?

Everything above concerns structure. Blending protects the **values**, and this
measures it: is a synthetic patient closer to a real patient than real patients
are to each other?

Compare two numbers rather than interpreting one. **Measured** is the run;
`0.5` is the target, not the maximum -- it means a synthetic patient is no more
like a real patient than one real patient is like another. Toward `0` is
memorisation, the failure; toward `1` the sets have separated, which costs
utility. **Expected** is what that statistic comes out at when nothing is
wrong, and it is not assumed: the same calculation runs repeatedly on two
halves of the real cohort, each half the size of the synthetic set being tested
(that is the "patients per half"), so small-sample artefacts cancel.

At a few dozen patients the expected range is wide. Inside it means **nothing
was detected**, never that nothing is there.

```{r}
#| label: proximity
compare_pmx_proximity(raw, synthetic, roles)
```

### Post-generation screen (optional, not run for you)

`flag_identifiable_subjects()` scores each **synthetic** patient on four axes --
follow-up time, dose count, dose magnitude, peak DV -- and flags robust
outliers. It is a **plausibility check rather than a masking step**: the anchor
screen has already kept structurally extreme *source* patients from anchoring an
avatar (the "excluded as structurally extreme" row of the report above), and a
conspicuous *blended* value identifies nobody, because it belongs to no one.

Read what it flags before acting on it. Its modified z-score reacts to the shape
of the distribution as well as to genuine outliers: on a two-humped cohort such
as `mavoglurant`, where half the patients have one dosing period and half have
two, it flags 17 of 120 avatars on the same axes the anchor screen had already
handled. If this study has two clearly different groups, expect some of that and
do not simply drop what it returns.

```{r}
#| label: flag-identifiable
flagged <- flag_identifiable_subjects(synthetic, roles)
sum(flagged$flagged)
knitr::kable(utils::head(flagged[flagged$flagged, ], 10))
```

`remediate_identifiable_subjects()` will truncate, drop, or replace them and
refill the cohort. Note `REV-028`: replacements are screened against their own
batch rather than the merged cohort, so a replacement can be clean in its batch
and an outlier once joined.

### Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r}
#| label: diagnostic
#| fig-width: 14
#| fig-height: 5
# The CONFIG plot again, with source and synthetic overlaid and coloured by
# which. `raw[, names(synthetic)]` lines the two up: generation keeps exactly
# the columns the roles named, so this compares every column that survived.
# Source and synthetic IDs are disjoint by construction, so grouping on the ID
# alone never joins a real patient's line to an avatar's.
compare <- rbind(
  raw[, names(synthetic)] |> mutate(DATA = "source"),
  synthetic               |> mutate(DATA = "synthetic")
)

ggplot(compare |> filter(EVID == 0, !is.na(LIDV)),
       aes(x = TIME, y = LIDV, group = ID, color = DATA)) +
  geom_point(alpha = 0.5) +
  geom_line(alpha = 0.5) +
  facet_grid(NAME ~ TRT, scales = "free_y", switch = "y") +
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10()
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories, so it stays under the source data's own access
controls and obligations wherever it is used — your own workstation included, a
partner or vendor not.
