---
title: "AVATAR synthesis on CPIT565B1 modeling data"
format:
  html:
    toc: true
    # Everything except the configuration and the generation call is reporting
    # code, and reading it is not the point of this document. Folded, not
    # hidden: click "Show code" on any block to see exactly what produced a
    # number. The chunks worth reading unfolded set `code-fold: false`.
    code-fold: true
    code-summary: "Show code"
---

Try the `synpmx_avatar()` on a real modeling dataset using the default AVATAR method.

Workflow:
1. Fill in the CONFIG chunks: the dataset location and the column roles
2. Explore the assessment of patient profiles and uniqueness. 
3. Compare the synthetic data to the actual data (tables and plots)

```{r}
#| label: setup
#| message: false
#| code-fold: false
library(tidyverse)
library(xgxr)
xgx_theme_set()

# If not installed here: devtools::load_all("/path/to/synpmx")
# This load_all line only needs to be run once, so you may comment out if you're 
# Trying this out on your own dataset

devtools::load_all()
library(synpmx)

```

## Configuration

Everything you edit is in this chunk.

```{r, fig.width = 14, fig.height = 5}
#| label: config
#| code-fold: false
user <- Sys.getenv("USER")
gps_prefix <- paste0("/view/", user, "_view/vob/")
filename   <- file.path(gps_prefix, "CPIT565A1/mas/mas_1/data/output_001/", "mt96714_poppkpd.csv")
data_in    <- read.csv(filename, stringsAsFactors = FALSE)

DATA       <- data_in |> filter(NAME %in% c("PIT565 Dose", "PIT565 PK", "Circ B Cells"), STUDYID == "CPIT565B12101")
OUT_DIR   <- "output_pit565b"
SEED      <- 1234        # reproducibility seed; the caller's RNG is untouched

# plot initial data
# Observation rows only. This used to exclude the dose endpoint by name, which
# is the previous study's name on a copied template; EVID says the same thing
# for any study.
ggplot(DATA |> filter(EVID == 0, !is.na(LIDV)), aes(x = TIME, y = LIDV, group = ID, color = TRT)) + 
  geom_point(alpha = 0.5) + 
  geom_line(alpha = 0.5) + 
  facet_grid(NAME~TRT, scales = "free_y", switch = "y") + 
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10()
```

Map your columns onto PMX roles. Only the roles the real dataset has. AVATAR
needs no model, priors, design, or budget -- just the column meanings. List
baseline covariates in `covariates` and they are blended from compatible donors
like everything else.

```{r}
#| label: roles
#| code-fold: false
ROLES <- pmx_roles(
  id       = "USUBJID",
  time     = "TIME",
  dv       = "LIDV",
  amt      = "AMT",
  evid     = "EVID",
  dvid     = c("YTYPE", "NAME"),
  cmt      = "CMT",
  mdv      = "MDV",
  nominal_time = "NOMTIME", # e.g. "NTIME"
  tad      = NULL,     # time-after-dose column; recomputed after generation
  subject_properties = c("TRT", "TRTN"),  # set NULL to compare pooled pattern pools
  rate     = NULL,           # set if you have infusion RATE rows
  occasion = NULL,           # set if TIME resets by occasion
  cens     = "CENS",         # e.g. "CENS"  (Monolix: 1 = below, -1 = above)
  limit    = NULL,           # e.g. "LIMIT" (other end of an interval)
  covariates = c("WEIGHT0", "AGE0", "SEX", "RACE"),
  keep     = c("STUDYIDN", "STUDYID")
)

# Number of synthetic subjects. NULL keeps the source cohort size.
N_SUBJECTS <- NULL
```

## Workflow

You should not need to edit below here.

```{r}
#| label: read-source
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
raw   <- DATA
roles <- ROLES
```

### Source structural validation

```{r}
#| label: validate-source
report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
message("  valid; ", length(unique(raw[[roles$id]])), " subjects")
```

### How identifying is this study's visit schedule?

Assessed by function `skeleton_uniqueness()` and run twice generating the synthetic data, in order to assess the uniqueness of each patient's visit schedule.  If many patients have unique schedules, a coarsening step is run first, snapping patients to a nominal time grid to reduce identifiability by the observation time.  

The visit schedule is assessed twice

- **as given** -- the recorded times, before anything is done to them.
- **after coarsening** -- the shared visit grid `synpmx_avatar()` actually
  builds and generates on. *These are the numbers that matter.* The difference
  between the two is what coarsening bought.

Every column is the same kind of number: **how many patients share this
property with me, me included**, so `1` always means "nobody else". Read the
first table's headline row, then its two sub-rows, which have opposite
remedies.

```{r}
#| label: skeleton-uniqueness-before
exposure_before_coarsening <- skeleton_uniqueness(raw, roles)
exposure_before_coarsening
```

```{r}
#| label: skeleton-uniqueness-after
exposure_after_coarsening <- skeleton_uniqueness(raw, roles, coarsen_time = TRUE)
exposure_after_coarsening
```

**Is a high percentage bad?** Not on its own, and this is the part that is easy
to misread. Split it by cause:

- **a one-off observation time** -- this patient was sampled at a moment
  nobody else was. It *is* a problem, and it is fixable: declaring
  `nominal_time` snaps visits onto the real protocol grid. A count above zero
  here means the inferred grid could not find a shared schedule.
- **the set of visits attended** -- every one of this patient's times is shared
  with somebody; what singles them out is which of those visits thave observations. This is due to either drop-out or that the patient's follow-up time was not long enough, and **no grid at any resolution can change it**. It is
  handled downstream by `min_pattern_share`, which refuses to give any avatar a
  visit set that only one real patient had. A large count here with a small
  count above is the *normal* result for a study with dropout, not a failure.

The picture says in one look what the counts say in numbers -- whether this is
a protocol grid with a few stragglers or an ad-hoc schedule the grid could not
rescue.

Read the map top to bottom: vertical stripes mean patients share visits, and a
ragged right edge is dropout. In the lower panel, a bar of height one is a
moment only one patient was ever sampled at -- exactly what would be copied
onto an avatar -- and those are drawn in red. Compare against the uncoarsened
version to see what the grid merged.

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map-raw
plot_pmx_schedule(raw, roles, coarsen_time = FALSE)
```

### Synthesize (AVATAR blending)

```{r}
#| label: synthesize
#| code-fold: false
synthetic <- synpmx_avatar(raw, roles, n_subjects = N_SUBJECTS,
                           seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)
message("  generated ", nrow(synthetic), " rows for ",
        length(unique(synthetic[[roles$id]])), " subjects; new identifiers: ",
        length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0)

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"), row.names = FALSE)
```

### What the mechanisms actually did

Every masking mechanism records what it did, and several record what it cost.
Each row carries the sentence that says what its number means, because none of
them mean anything on their own. Copy this into the study inventory in
`scripts_private/README.md`.

```{r}
#| label: mechanism-report
pmx_masking_report(synthetic, raw, roles)
```

Four rows are worth stopping on.

**Did it detect weight-based dosing?** The **Dose** section says so outright:
either "**yes**, from `WEIGHT0`" with the protocol levels it found, or
"**no**" together with the reason it declined. If this study doses on weight
and the answer is no, the reason names which covariates were tried and what
failed -- most often that the dose-to-covariate ratio did not collapse onto a
small number of protocol levels, which is the conservative test the detector
uses to avoid rewriting amounts in a study that is not dose-proportional. When
the answer is no, `AMT` is copied verbatim from the anchor, which means each
avatar's implied mg/kg is wrong and the amount still encodes one real patient's
weight.

**Shared by too few patients, so not reused.** These are real dropout and
dose-interruption patterns that will not appear in the synthetic data. Refusing
them is what stops an avatar carrying a schedule traceable to one person. If
this study's interruptions matter, lower `min_pattern_share`; `2` is the lowest
value that still guarantees no synthetic patient has a visit set unique to a
real one, and `1` keeps every pattern and gives up the guarantee.

At the default floor of 2, that row and **real patients holding those** are
necessarily equal: a set is discarded exactly when fewer than two patients
share it, so every discarded set has exactly one holder. They diverge only at a
floor of 3 or more.

**Avatars keeping their anchor's own visit set.** Read this against the row
above it. A high percentage means the schedule groups had nothing shareable to
draw from, so those avatars carry one real patient's pattern of absences --
which undoes what discarding the rare sets was for. If it is high, the run also
raises the "no shared visit set to draw from" alert, and the fix is the same:
declare `nominal_time`, or lower `min_pattern_share`.

### Restricted diagnostic: distribution summary

Source-derived. Keep it in the safe environment; do not export these tables.

`compare_pmx_distributions()` summarizes the dependent variable per endpoint and
each baseline covariate, source against synthetic -- a quick numeric range check
before the plot. Expect the summaries to line up in magnitude and shape, not to
match exactly, since AVATAR does not reproduce the source distribution.

```{r}
#| label: distribution-summary
compare_pmx_distributions(raw, synthetic, roles)
```

### Restricted diagnostic: how close did the values land?

Everything above concerns structure. Blending is what protects the **values**,
and this is its measurement. It answers one question: **is a synthetic patient
closer to a real patient than real patients are to each other?**

The way to read it is to compare two numbers, not to interpret one:

- **Measured** -- the statistic on this run. `0.5` is the target, not the
  maximum: it means a synthetic patient is no more like a real patient than one
  real patient is like another. Toward `0` means memorisation, which is the
  failure. Toward `1` means the two sets have separated, which costs utility
  rather than privacy.
- **Expected** -- what that statistic comes out at when nothing is wrong. It is
  not assumed. The same calculation is run repeatedly on two halves of the real
  cohort, so every small-sample artefact is present in the null and in the
  measurement alike and cancels. Each half has the same number of patients as
  the synthetic set being tested, which is what the *"patients per half"*
  figure is: the size of one side of the comparison, matched on all three arms
  so the counts are comparable.

At a few dozen patients that expected range is wide. Inside it means **nothing
was detected**, never that nothing is there.

```{r}
#| label: proximity
compare_pmx_proximity(raw, synthetic, roles)
```

### Post-generation screen (optional, not run for you)

`flag_identifiable_subjects()` scores each **synthetic** patient on four axes --
follow-up time, dose count, dose magnitude, peak DV -- and flags robust
outliers. It is a **plausibility check rather than a masking step**: the anchor
screen has already kept structurally extreme *source* patients from anchoring an
avatar (the "excluded as structurally extreme" row of the report above), and a
conspicuous *blended* value identifies nobody, because it belongs to no one.

Read what it flags before acting on it. Its median/MAD statistic behaves badly
on a bimodal cohort -- on `mavoglurant`, where half the patients have one dosing
period and half have two, it flags 75 of 120 avatars on the same axes the anchor
screen had already handled. If this study has two clearly different groups,
expect the same and do not simply drop what it returns.

```{r}
#| label: flag-identifiable
flagged <- flag_identifiable_subjects(synthetic, roles)
sum(flagged$flagged)
knitr::kable(utils::head(flagged[flagged$flagged, ], 10))
```

`remediate_identifiable_subjects()` will truncate, drop, or replace them and
refill the cohort. Note `REV-028`: replacements are screened against their own
batch rather than the merged cohort, so a replacement can be clean in its batch
and an outlier once joined.

### Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r, fig.width = 14, fig.height = 5}
#| label: diagnostic
#| fig-width: 14
#| fig-height: 5
# Read from `roles`, not from literal column names. This chunk was copied
# between studies and kept the previous one's columns: the ECI report grouped
# by `USUBJID` where its `id` role is `ID`, and filtered out `"PIT565 Dose"` on
# a study that has no such endpoint. Naming roles instead cannot drift, and the
# facets grow a column per endpoint on their own, whatever this study has.
label_col <- if (length(roles$dvid) > 1L) {
  roles$dvid[[length(roles$dvid)]]
} else if (!is.null(roles$dvid)) roles$dvid[[1L]] else NULL

frame <- function(d, label) {
  keep <- d[[roles$evid]] == 0 & !is.na(d[[roles$dv]])   # observations only
  out <- data.frame(
    DATA     = label,
    id       = d[[roles$id]][keep],
    time     = as.numeric(d[[roles$time]][keep]),
    dv       = as.numeric(d[[roles$dv]][keep]),
    endpoint = if (is.null(label_col)) "DV" else
      as.character(d[[label_col]][keep]),
    stringsAsFactors = FALSE
  )
  # Colour by the first declared subject property (treatment arm) when there is
  # one, so this works unchanged on a study with no arms.
  arm <- roles$subject_properties[[1L]]
  out$arm <- if (is.null(arm)) "all" else as.character(d[[arm]][keep])
  out
}
compare <- rbind(frame(raw, "source"), frame(synthetic, "synthetic"))

g <- ggplot(compare, aes(x = time, y = dv, group = id, color = arm)) +
  geom_point(alpha = 0.5) +
  geom_line(alpha = 0.5) +
  facet_grid(DATA ~ endpoint, scales = "free_y") +
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10() +
  labs(title = "RESTRICTED: source vs AVATAR synthetic",
       subtitle = "do not export", x = "Time", y = "DV", color = "Arm")
g
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories, so it stays under the source data's own access
controls and obligations wherever it is used — your own workstation included, a
partner or vendor not.
