---
title: "AVATAR synthesis on CPIT565B1 modeling data"
format:
  html:
    toc: true
    # Everything except the configuration and the generation call is reporting
    # code, and reading it is not the point of this document. Folded, not
    # hidden: click "Show code" on any block to see exactly what produced a
    # number. The chunks worth reading unfolded set `code-fold: false`.
    code-fold: true
    code-summary: "Show code"
---

Try the `synpmx_avatar()` on a real modeling dataset using the default AVATAR method.

Workflow:
1. Fill in the CONFIG chunks: the dataset location and the column roles
2. Explore the assessment of patient profiles and uniqueness. 
3. Compare the synthetic data to the actual data (tables and plots)

```{r}
#| label: setup
#| message: false
#| code-fold: false
library(tidyverse)
library(xgxr)
xgx_theme_set()

# Only needed once per session; comment out if synpmx is installed.

devtools::load_all()
library(synpmx)

```

## Configuration

Everything you edit is in this chunk.

```{r, fig.width = 14, fig.height = 5}
#| label: config
#| code-fold: false
user <- Sys.getenv("USER")
gps_prefix <- paste0("/view/", user, "_view/vob/")
filename   <- file.path(gps_prefix, "CPIT565A1/mas/mas_1/data/output_001/", "mt96714_poppkpd.csv")
data_in    <- read.csv(filename, stringsAsFactors = FALSE)

DATA       <- data_in |> filter(NAME %in% c("PIT565 Dose", "PIT565 PK", "Circ B Cells"), STUDYID == "CPIT565B12101")
OUT_DIR   <- "output_pit565b"
SEED      <- 1234        # reproducibility seed; the caller's RNG is untouched

# plot initial data
# Observation rows only.
ggplot(DATA |> filter(EVID == 0, !is.na(LIDV)), aes(x = TIME, y = LIDV, group = ID, color = TRT)) + 
  geom_point(alpha = 0.5) + 
  geom_line(alpha = 0.5) + 
  facet_grid(NAME~TRT, scales = "free_y", switch = "y") + 
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10()
```

Map your columns onto PMX roles. Only the roles the real dataset has. AVATAR
needs no model, priors, design, or budget -- just the column meanings. List
baseline covariates in `covariates` and they are blended from compatible donors
like everything else.

```{r}
#| label: roles
#| code-fold: false
ROLES <- pmx_roles(
  id       = "USUBJID",
  time     = "TIME",
  dv       = "LIDV",
  amt      = "AMT",
  evid     = "EVID",
  dvid     = c("YTYPE", "NAME"),  # first is the key; the rest checked 1:1 against it
  cmt      = "CMT",
  mdv      = "MDV",
  # Rows missing one fall back to an inferred grid; the run then reports the
  # grid as "mixed" and says how many rows that was.
  nominal_time = "NOMTIME",
  tad      = NULL,           # time-after-dose column; recomputed after generation
  # Treatment arm / dose group. Also splits the pool of visit sets an avatar may
  # draw from: declaring it is more faithful, but more rare visit sets are then
  # too thinly held to reuse. Try it both ways and compare the run reports.
  subject_properties = c("TRT", "TRTN"),  # NULL pools the arms
  rate     = NULL,           # set if you have infusion RATE rows
  occasion = NULL,           # set if TIME resets by occasion
<<<<<<< HEAD
  cens     = "CENS",         # e.g. "CENS"  (Monolix: 1 = below, -1 = above)
=======
  # Declared, censoring is rebuilt to agree with the blended DV; left NULL, a
  # BLQ flag would survive that no longer matches its own value.
  cens     = "CENS",         # Monolix: 1 = below, -1 = above
>>>>>>> 9d8fd18d0ebca1232ac8896596db1843ade6ddbf
  limit    = NULL,           # e.g. "LIMIT" (other end of an interval)
  # Blended across donors. Leave out any baseline that is the same quantity as
  # an endpoint -- the two are blended independently, so they would disagree
  # (REV-022). Read that baseline off the generated trajectory instead.
  covariates = c("WEIGHT0", "AGE0", "SEX", "RACE"),
  keep     = c("STUDYIDN", "STUDYID")  # pass-through only; a stratum belongs above
  # Any column no role names is dropped, so a forgotten identifier cannot ride
  # out of a real patient. The run prints which columns went.
)

# Number of synthetic subjects. NULL keeps the source cohort size.
N_SUBJECTS <- NULL
```

## Workflow

You should not need to edit below here.

```{r}
#| label: read-source
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
raw   <- DATA
roles <- ROLES
```

### Source structural validation

```{r}
#| label: validate-source
report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
message("  valid; ", length(unique(raw[[roles$id]])), " subjects")
```

### How identifying is this study's visit schedule?

<<<<<<< HEAD
Assessed by function `skeleton_uniqueness()` and run twice generating the synthetic data, in order to assess the uniqueness of each patient's visit schedule.  If many patients have unique schedules, a coarsening step is run first, snapping patients to a nominal time grid to reduce identifiability by the observation time.  

The visit schedule is assessed twice

- **as given** -- the recorded times, before anything is done to them.
- **after coarsening** -- the shared visit grid `synpmx_avatar()` actually
  builds and generates on. *These are the numbers that matter.* The difference
  between the two is what coarsening bought.

Every column is the same kind of number: **how many patients share this
property with me, me included**, so `1` always means "nobody else". Read the
first table's headline row, then its two sub-rows, which have opposite
remedies.
=======
`synpmx_avatar()` copies its anchor's event skeleton **verbatim**, so a patient
holding the only copy of their observation times hands that schedule to every
avatar anchored on them. Run it twice: on the times as given, and on the grid
the generator actually builds. The second is the one that matters, and the
difference between them is what coarsening bought.
>>>>>>> 9d8fd18d0ebca1232ac8896596db1843ade6ddbf

```{r}
#| label: skeleton-uniqueness-before
exposure_before_coarsening <- skeleton_uniqueness(raw, roles)
exposure_before_coarsening
```

```{r}
#| label: skeleton-uniqueness-after
exposure_after_coarsening <- skeleton_uniqueness(raw, roles, coarsen_time = TRUE)
exposure_after_coarsening
```

Split the headline by cause before deciding it is bad. **A one-off observation
time** is a real problem and a fixable one -- declare `nominal_time`. **A
unique set of attended visits** is dropout: no grid at any resolution touches
it, and `min_pattern_share` handles it downstream. A large count there with a
small count above is the normal result for a study with dropout, not a failure.

<<<<<<< HEAD
- **a one-off observation time** -- this patient was sampled at a moment
  nobody else was. It *is* a problem, and it is fixable: declaring
  `nominal_time` snaps visits onto the real protocol grid. A count above zero
  here means the inferred grid could not find a shared schedule.
- **the set of visits attended** -- every one of this patient's times is shared
  with somebody; what singles them out is which of those visits thave observations. This is due to either drop-out or that the patient's follow-up time was not long enough, and **no grid at any resolution can change it**. It is
  handled downstream by `min_pattern_share`, which refuses to give any avatar a
  visit set that only one real patient had. A large count here with a small
  count above is the *normal* result for a study with dropout, not a failure.

The picture says in one look what the counts say in numbers -- whether this is
a protocol grid with a few stragglers or an ad-hoc schedule the grid could not
rescue.

Read the map top to bottom: vertical stripes mean patients share visits, and a
ragged right edge is dropout. In the lower panel, a bar of height one is a
moment only one patient was ever sampled at -- exactly what would be copied
onto an avatar -- and those are drawn in red. Compare against the uncoarsened
version to see what the grid merged.
=======
The picture answers the same question faster. Vertical stripes mean patients
share visits, a ragged right edge is dropout, and a red bar in the lower panel
is a moment only one patient was ever sampled at -- exactly what would be
copied onto an avatar.

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map
plot_pmx_schedule(raw, roles)
```
>>>>>>> 9d8fd18d0ebca1232ac8896596db1843ade6ddbf

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map-raw
plot_pmx_schedule(raw, roles, coarsen_time = FALSE)
```

### Synthesize (AVATAR blending)

```{r}
#| label: synthesize
#| code-fold: false
synthetic <- synpmx_avatar(raw, roles, n_subjects = N_SUBJECTS,
                           seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)
message("  generated ", nrow(synthetic), " rows for ",
        length(unique(synthetic[[roles$id]])), " subjects; new identifiers: ",
        length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0)

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"), row.names = FALSE)
```

### What the mechanisms actually did

Each row carries the sentence that says what its number means. Copy this into
the study inventory in `scripts_private/README.md`.

```{r}
#| label: mechanism-report
pmx_masking_report(synthetic, raw, roles)
```

Three rows to stop on.

**Amounts recomputed from a covariate** answers "did it find the weight-based
dosing?" outright, and when the answer is no it names which covariates were
tried and what failed. Detection is deliberately conservative -- it refuses
rather than rewrite amounts in a study that is not dose-proportional. A "no" on
a study that *is* weight-based means `AMT` is copied verbatim from the anchor,
so each avatar's implied mg/kg is wrong and the amount still encodes one real
patient's weight.

**Shared by too few patients, so not reused** is the fidelity cost: real
dropout and interruption patterns that will not appear in the output. Refusing
them is what stops an avatar carrying a schedule traceable to one person.
`min_pattern_share = 2` is the lowest value that keeps that guarantee; `1` gives
it up. At 2 this row necessarily equals the one below it, since a set is
discarded exactly when one patient holds it.

**Avatars keeping their anchor's own visit set** undoes the row above when it is
high: those avatars carry one real patient's absences. The run alerts past 10%,
and the fix is `nominal_time` or a lower `min_pattern_share`.

### Restricted diagnostic: distribution summary

Source-derived. Keep it in the safe environment; do not export these tables.

`compare_pmx_distributions()` summarizes the dependent variable per endpoint and
each baseline covariate, source against synthetic -- a quick numeric range check
before the plot. Expect the summaries to line up in magnitude and shape, not to
match exactly, since AVATAR does not reproduce the source distribution.

```{r}
#| label: distribution-summary
compare_pmx_distributions(raw, synthetic, roles)
```

### Restricted diagnostic: how close did the values land?

Everything above concerns structure. Blending protects the **values**, and this
measures it: is a synthetic patient closer to a real patient than real patients
are to each other?

Compare two numbers rather than interpreting one. **Measured** is the run;
`0.5` is the target, not the maximum -- it means a synthetic patient is no more
like a real patient than one real patient is like another. Toward `0` is
memorisation, the failure; toward `1` the sets have separated, which costs
utility. **Expected** is what that statistic comes out at when nothing is
wrong, and it is not assumed: the same calculation runs repeatedly on two
halves of the real cohort, each half the size of the synthetic set being tested
(that is the "patients per half"), so small-sample artefacts cancel.

At a few dozen patients the expected range is wide. Inside it means **nothing
was detected**, never that nothing is there.

```{r}
#| label: proximity
compare_pmx_proximity(raw, synthetic, roles)
```

### Post-generation screen (optional, not run for you)

`flag_identifiable_subjects()` scores each **synthetic** patient on four axes --
follow-up time, dose count, dose magnitude, peak DV -- and flags robust
outliers. It is a **plausibility check rather than a masking step**: the anchor
screen has already kept structurally extreme *source* patients from anchoring an
avatar (the "excluded as structurally extreme" row of the report above), and a
conspicuous *blended* value identifies nobody, because it belongs to no one.

Read what it flags before acting on it. Its median/MAD statistic behaves badly
on a bimodal cohort -- on `mavoglurant`, where half the patients have one dosing
period and half have two, it flags 75 of 120 avatars on the same axes the anchor
screen had already handled. If this study has two clearly different groups,
expect the same and do not simply drop what it returns.

```{r}
#| label: flag-identifiable
flagged <- flag_identifiable_subjects(synthetic, roles)
sum(flagged$flagged)
knitr::kable(utils::head(flagged[flagged$flagged, ], 10))
```

`remediate_identifiable_subjects()` will truncate, drop, or replace them and
refill the cohort. Note `REV-028`: replacements are screened against their own
batch rather than the merged cohort, so a replacement can be clean in its batch
and an outlier once joined.

### Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r, fig.width = 14, fig.height = 5}
#| label: diagnostic
#| fig-width: 14
#| fig-height: 5
# Named roles, not literal columns, so this survives being copied to another
# study. The facets grow a column per endpoint on their own.
label_col <- if (length(roles$dvid) > 1L) {
  roles$dvid[[length(roles$dvid)]]
} else if (!is.null(roles$dvid)) roles$dvid[[1L]] else NULL

frame <- function(d, label) {
  keep <- d[[roles$evid]] == 0 & !is.na(d[[roles$dv]])   # observations only
  out <- data.frame(
    DATA     = label,
    id       = d[[roles$id]][keep],
    time     = as.numeric(d[[roles$time]][keep]),
    dv       = as.numeric(d[[roles$dv]][keep]),
    endpoint = if (is.null(label_col)) "DV" else
      as.character(d[[label_col]][keep]),
    stringsAsFactors = FALSE
  )
  # Colour by the first declared subject property (treatment arm) when there is
  # one, so this works unchanged on a study with no arms.
  arm <- roles$subject_properties[[1L]]
  out$arm <- if (is.null(arm)) "all" else as.character(d[[arm]][keep])
  out
}
compare <- rbind(frame(raw, "source"), frame(synthetic, "synthetic"))

g <- ggplot(compare, aes(x = time, y = dv, group = id, color = arm)) +
  geom_point(alpha = 0.5) +
  geom_line(alpha = 0.5) +
  facet_grid(DATA ~ endpoint, scales = "free_y") +
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10() +
  labs(title = "RESTRICTED: source vs AVATAR synthetic",
       subtitle = "do not export", x = "Time", y = "DV", color = "Arm")
g
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories, so it stays under the source data's own access
controls and obligations wherever it is used — your own workstation included, a
partner or vendor not.
