---
title: "AVATAR synthesis on CPIT565A1 modeling data"
format: html
---

Try the AVATAR blending generator on a real modeling dataset.

This is the **default** method. Use it when the synthetic data *stays inside*
your trusted computing environment (you are the only consumer, governance and
access controls apply). It resamples and blends whole source subject
trajectories -- more faithful and far less ceremony than the differentially
private path -- but makes **no** formal privacy guarantee. If the data may cross
a trust boundary, use `try_dp_calibrated.R` instead. See `README.md` and
`design/METHOD_DISCUSSION.md` for the decision rule.

This document runs in the safe computing environment, in a gitignored folder so
nothing here can be committed.

Workflow:

1. fill in the CONFIG chunk (just the column roles)
2. compare the synthetic table to the source

```{r}
#| label: setup
#| message: false
library(synpmx)
# If not installed here: devtools::load_all("/path/to/synpmx")
```

## Configuration

Everything you edit is in this chunk.

```{r}
#| label: config
user <- Sys.getenv("USER")
gps_prefix <- paste0("/view/", user, "_view/vob/")
DATA_PATH <- file.path(gps_prefix, "CPIT565A1/mas/mas_1/data/output_001/", "mt96714_poppkpd.csv")
OUT_DIR   <- "output_pit565"
SEED      <- 1234        # reproducibility seed; the caller's RNG is untouched
```

Map your columns onto PMX roles. Only the roles the real dataset has. AVATAR
needs no model, priors, design, or budget -- just the column meanings. List
baseline covariates in `covariates` and they are blended from compatible donors
like everything else.

```{r}
#| label: roles
ROLES <- pmx_roles(
  id       = "USUBJID",
  time     = "TIME",
  dv       = "LIDV",
  amt      = "AMT",
  evid     = "EVID",
  dvid     = "YTYPE",         # NULL if a single endpoint
  cmt      = "CMT",
  mdv      = "MDV",
  rate     = NULL,           # set if you have infusion RATE rows
  occasion = NULL,           # set if TIME resets by occasion
  covariates = c("WT", "AGE", "SEX"),
  exclude  = NULL            # columns to drop before synthesizing
)

# Number of synthetic subjects. NULL keeps the source cohort size.
N_SUBJECTS <- NULL

# Optional per-dataset cleanup: fix units, drop screening rows, recode. Receives
# the raw data frame, returns a cleaned one.
PREP <- function(d) {
  d
}
```

## Workflow

You should not need to edit below here.

```{r}
#| label: read-source
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

read_source <- function() {
  if (!file.exists(DATA_PATH)) {
    stop("DATA_PATH not found: ", normalizePath(DATA_PATH, mustWork = FALSE))
  }
  utils::read.csv(DATA_PATH, stringsAsFactors = FALSE)
}

raw   <- PREP(read_source())
roles <- ROLES
```

### Source structural validation

```{r}
#| label: validate-source
report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report$checks[report$checks$status == "error", c("check", "message")])
  stop("Source failed structural validation; fix the role mapping first.")
}
message("  valid; ", length(unique(raw[[roles$id]])), " subjects")
```

### Synthesize (AVATAR blending)

```{r}
#| label: synthesize
synthetic <- synpmx_avatar(raw, roles, n_subjects = N_SUBJECTS, seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)
message("  generated ", nrow(synthetic), " rows for ",
        length(unique(synthetic[[roles$id]])), " subjects; new identifiers: ",
        length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0)

utils::write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"),
                 row.names = FALSE)
```

### Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r}
#| label: diagnostic
if (requireNamespace("ggplot2", quietly = TRUE)) {
  obs <- function(d, label) {
    keep <- d[[roles$evid]] == 0 & !is.na(d[[roles$dv]])
    data.frame(dataset = label,
               time = as.numeric(d[[roles$time]][keep]),
               dv   = as.numeric(d[[roles$dv]][keep]),
               endpoint = if (is.null(roles$dvid)) "DV" else
                 as.character(d[[roles$dvid]][keep]))
  }
  compare <- rbind(obs(raw, "source"), obs(synthetic, "synthetic"))
  p <- ggplot2::ggplot(compare, ggplot2::aes(time, dv, colour = dataset)) +
    ggplot2::geom_point(alpha = 0.4, size = 0.7) +
    ggplot2::facet_grid(dataset ~ endpoint, scales = "free_y") +
    ggplot2::labs(title = "RESTRICTED: source vs AVATAR synthetic",
                  subtitle = "do not export", x = "Time", y = "DV") +
    ggplot2::theme_minimal() +
    ggplot2::theme(legend.position = "none")
  ggplot2::ggsave(file.path(OUT_DIR, "RESTRICTED_source_vs_synthetic.png"), p,
                  width = 7, height = 6, dpi = 110)
  message("  wrote ", file.path(OUT_DIR, "RESTRICTED_source_vs_synthetic.png"),
          " (restricted; keep in the safe environment)")
  p
} else {
  message("  ggplot2 not available; skipping the diagnostic plot")
}
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories and is for trusted-environment use only.
