---
title: "AVATAR synthesis on CPIT565B1 modeling data"
format: html
---

Try the AVATAR blending generator on a real modeling dataset using the default AVATAR method. 

Workflow:
1. fill in the CONFIG chunk (just the column roles)
2. compare the synthetic table to the source

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(xgxr)
xgx_theme_set()

# If not installed here: devtools::load_all("/path/to/synpmx")
# This load_all line only needs to be run once, so you may comment out if you're 
# Trying this out on your own dataset

devtools::load_all()
library(synpmx)

```

## Configuration

Everything you edit is in this chunk.

```{r, fig.width = 14, fig.height = 5}
#| label: config
user <- Sys.getenv("USER")
gps_prefix <- paste0("/view/", user, "_view/vob/")
filename   <- file.path(gps_prefix, "CPIT565A1/mas/mas_1/data/output_001/", "mt96714_poppkpd.csv")
data_in    <- read.csv(filename, stringsAsFactors = FALSE)

DATA       <- data_in |> filter(NAME %in% c("PIT565 Dose", "PIT565 PK", "Circ B Cells"), STUDYID == "CPIT565B12101")
OUT_DIR   <- "output_pit565b"
SEED      <- 1234        # reproducibility seed; the caller's RNG is untouched

# plot initial data
ggplot(DATA |> filter(NAME != "PIT565 Dose"), aes(x = TIME, y = LIDV, group = ID, color = TRT)) + 
  geom_point(alpha = 0.5) + 
  geom_line(alpha = 0.5) + 
  facet_grid(NAME~TRT, scales = "free_y", switch = "y") + 
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10()
```

Map your columns onto PMX roles. Only the roles the real dataset has. AVATAR
needs no model, priors, design, or budget -- just the column meanings. List
baseline covariates in `covariates` and they are blended from compatible donors
like everything else.

```{r}
#| label: roles
ROLES <- pmx_roles(
  id       = "USUBJID",
  time     = "TIME",
  dv       = "LIDV",
  amt      = "AMT",
  evid     = "EVID",
  dvid     = c("YTYPE", "NAME"),
  cmt      = "CMT",
  mdv      = "MDV",
  nominal_time = "NOMTIME", # e.g. "NTIME"
  tad      = NULL,     # time-after-dose column; recomputed after generation
  subject_properties = c("TRT", "TRTN"),  # set NULL to compare pooled pattern pools
  rate     = NULL,           # set if you have infusion RATE rows
  occasion = NULL,           # set if TIME resets by occasion
  cens     = "CENS",           # e.g. "CENS"  (Monolix: 1 = below, -1 = above)
  limit    = NULL,           # e.g. "LIMIT" (other end of an interval)
  covariates = c("WEIGHT0", "AGE0", "SEX", "RACE"),
  keep     = c("STUDYIDN", "STUDYID")
)

# Number of synthetic subjects. NULL keeps the source cohort size.
N_SUBJECTS <- NULL
```

## Workflow

You should not need to edit below here.

```{r}
#| label: read-source
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
raw   <- DATA
roles <- ROLES
```

### Source structural validation

```{r}
#| label: validate-source
report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
message("  valid; ", length(unique(raw[[roles$id]])), " subjects")
```

### Source skeleton exposure

Run before generating. The event skeleton is copied verbatim from one anchor, so
a subject holding the only copy of its observation time vector hands that
schedule to every avatar anchored on it.

Read `alone` first, then why:

- `min_time_share == 1` -- observed at a moment no other subject was. This is
  what `coarsen_time` collapses, and a residue here means the *inferred* grid
  found no shared schedule. Declaring `nominal_time` is the fix.
- `min_time_share >= 2` but `alone` -- every time is shared and only the
  *pattern* of attended visits is unique. Dropout and missed visits. No grid
  touches this; `min_pattern_share` is what stops it being reproduced.
- `n_obs_class == 1` -- a sample count nobody else has.
- `signature_class == 1` -- dose structure or amount nobody else has.

Record these in `scripts_private/README.md`'s study inventory.

```{r}
#| label: skeleton-uniqueness
exposure <- skeleton_uniqueness(raw, roles)
exposure
```

### Synthesize (AVATAR blending)

```{r}
#| label: synthesize
synthetic <- synpmx_avatar(raw, roles, seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)
message("  generated ", nrow(synthetic), " rows for ",
        length(unique(synthetic[[roles$id]])), " subjects; new identifiers: ",
        length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0)

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"), row.names = FALSE)
```

### What the mechanisms actually did

Every masking mechanism records what it did, and two of them record what it
cost. Copy these into the study inventory in `scripts_private/README.md`.

```{r}
#| label: mechanism-report
settings <- attr(synthetic, "pmx_settings")
data.frame(
  quantity = c(
    "source patients",
    "  ... excluded as structural outliers (M2, screen)",
    "  ... excluded below the donor floor (M2, route arm)",
    "  ... left available to anchor on",
    "route arms / compatible event groups",
    "time grid used",
    "patients with a unique observation schedule, after coarsening",
    "  ... unique observation time (declare nominal_time to fix)",
    "  ... unique visit set (dropout; M3's job)",
    "dose basis detected",
    "dose levels found",
    "distinct visit sets in the source",
    "  ... discarded as too rare to reuse",
    "  ... real subjects who held those",
    "avatars given a real patient's visit set",
    "  ... of those, which visits were missed was invented",
    "donor floor: mean effective donors",
    "donor cap binding fraction"
  ),
  value = c(
    settings$source_subjects,
    settings$anchors_screened_out,
    settings$anchors_route_excluded,
    settings$anchors_available,
    paste(settings$routes, "/", settings$compatible_event_groups),
    settings$time_grid,
    settings$unique_schedule_n,
    settings$unique_obs_time_n,
    settings$unique_visit_set_n,
    ifelse(is.na(settings$dose_basis), "none (amounts left alone)",
           settings$dose_basis),
    ifelse(is.na(settings$dose_basis), "-",
           paste(signif(settings$dose_levels, 4), collapse = ", ")),
    settings$patterns_total,
    settings$patterns_dropped,
    settings$subjects_with_dropped_pattern,
    settings$pattern_sampled_fraction,
    round(settings$pattern_generated_fraction, 2),
    round(settings$mean_effective_donors, 2),
    round(settings$cap_binding_fraction, 2)
  ),
  stringsAsFactors = FALSE
) |> knitr::kable()
```

At the default floor of 2, `discarded` and `real subjects who held those` are
necessarily equal: a pattern is discarded exactly when fewer than two patients
share it, so every discarded pattern has exactly one holder. They diverge only at
a floor of 3 or more.

`discarded as too rare to reuse` is the one to look at hardest. Those are real
dropout and dose-interruption patterns that will not appear in the synthetic
data, because too few patients shared them. Discarding them is what stops an
avatar carrying a schedule traceable to one person, but if this study's
interruptions matter, lower `min_pattern_share` (2 is the lowest value that
still guarantees no synthetic patient has a schedule unique to a real one) or
set it to `1` to keep every pattern and give up that guarantee.

### Restricted diagnostic: distribution summary

Source-derived. Keep it in the safe environment; do not export these tables.

`compare_pmx_distributions()` summarizes the dependent variable per endpoint and
each baseline covariate, source against synthetic -- a quick numeric range check
before the plot. Expect the summaries to line up in magnitude and shape, not to
match exactly, since AVATAR does not reproduce the source distribution.

```{r}
#| label: distribution-summary
#| results: asis
# Rendered with knitr::kable() so the tables come out as tables when this
# document is knitted, and stay readable when run chunk-by-chunk. Rounding is
# for display only; the returned object keeps the exact values.
kable_distributions <- function(x) {
  section <- function(df, caption) {
    if (is.null(df)) return(invisible())
    numeric_cols <- vapply(df, is.numeric, logical(1))
    df[numeric_cols] <- lapply(df[numeric_cols], signif, digits = 4)
    print(knitr::kable(df, caption = caption))
    cat("\n")
  }
  section(x$endpoints,
          "RESTRICTED -- endpoints (dependent variable on observation rows)")
  section(x$covariates_numeric,
          "RESTRICTED -- continuous covariates (baseline, per subject)")
  section(x$covariates_categorical,
          "RESTRICTED -- categorical covariates (baseline, per subject)")
  invisible(x)
}

kable_distributions(compare_pmx_distributions(raw, synthetic, roles))
```

### Restricted diagnostic: how close did the values land?

Everything above concerns structure. Blending is what protects the **values**,
and this is its measurement: does each subject's nearest neighbour lie in its own
dataset or the other one? Near 0.5 means a synthetic subject is no more like a
real subject than one real subject is like another -- the target. Toward 0 means
memorisation.

Read the null interval before the point estimate. It is built by splitting *this*
source cohort in half and running the identical statistic, so every small-sample
artefact cancels -- but at a few dozen subjects it is wide. Inside the interval
means **nothing was detected**, not that nothing is there.

```{r}
#| label: proximity
compare_pmx_proximity(raw, synthetic, roles)
```

### Post-generation screen (optional, not run for you)

`flag_identifiable_subjects()` scores each **synthetic** patient on four axes --
follow-up time, dose count, dose magnitude, peak DV -- and flags robust
outliers. It is a **plausibility check rather than a masking step**: M2 has
already kept structurally extreme *source* patients out of the anchor pool, and
a conspicuous blended value identifies nobody because it belongs to no one.

Read what it flags before acting on it. Its median/MAD statistic behaves badly
on a bimodal cohort -- on `mavoglurant`, where half the patients have one dosing
period and half have two, it flags 75 of 120 avatars on axes M2 already
screened. If this study has two clearly different groups, expect the same and do
not simply drop what it returns.

```{r}
#| label: flag-identifiable
flagged <- flag_identifiable_subjects(synthetic, roles)
sum(flagged$flagged)
knitr::kable(utils::head(flagged[flagged$flagged, ], 10))
```

`remediate_identifiable_subjects()` will truncate, drop, or replace them and
refill the cohort. Note `REV-028`: replacements are screened against their own
batch rather than the merged cohort, so a replacement can be clean in its batch
and an outlier once joined.

### Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r, fig.width = 14, fig.height = 5}
#| label: diagnostic

  raw_keep = raw[,names(synthetic)]

  compare = rbind(raw_keep |> mutate(DATA = "source"), 
                  synthetic |> mutate(DATA = "synthetic"))
  
  g = ggplot(compare |> filter(NAME != "PIT565 Dose"), aes(x = TIME, y = LIDV, group = USUBJID, color = DATA)) + 
    geom_point(alpha = 0.5) + 
    geom_line(alpha = 0.5) + 
    facet_grid(NAME~TRT, scales = "free_y", switch = "y") + 
    xgx_scale_x_time_units("days") +
    xgx_scale_y_log10()
    g
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories, so it stays under the source data's own access
controls and obligations wherever it is used — your own workstation included, a
partner or vendor not.
