---
title: "AVATAR synthesis on your own modeling data"
format:
  html:
    toc: true
    # Everything except the configuration and the generation call is reporting
    # code, and reading it is not the point of this document. Folded, not
    # hidden: click "Show code" on any block to see exactly what produced a
    # number. The chunks worth reading unfolded set `code-fold: false`.
    code-fold: true
    code-summary: "Show code"
---

AVATAR blending is the default `synpmx` method. Use it when the synthetic data
*reaches no one the source data could not* -- the same access controls and
confidentiality obligations still apply, which includes carrying it to your own
workstation for model development. It resamples and blends whole source
subject trajectories -- faithful, and far less ceremony than the differentially
private (DP) path -- but makes **no** formal privacy guarantee. If the data may reach
someone new (e.g. externally shared data), use `try_dp_calibrated.R` instead or explore
the other algorithms as well (Prior, Empirical); see the `synpmx-privacy` vignette for the decision rule.

Prefer a plain script? `knitr::purl("try_avatar.qmd")` writes the R chunks out
as a `.R` file. There is one template; the script is generated from it.

To use: edit the two configuration chunks -- the data path, and the column
roles -- and run. You should not need to touch anything below them.

```{r}
#| label: setup
#| code-fold: false
#| message: false
library(synpmx)
library(ggplot2)
# If synpmx is not installed here: devtools::load_all("/path/to/synpmx")
```

## Configuration 1 --- where the data is

```{r}
#| label: config
#| code-fold: false
DATA_PATH  <- "data/your_modeling_dataset.csv"
OUT_DIR    <- "output"
SEED       <- 1234        # reproducibility seed; the caller's RNG is untouched
N_SUBJECTS <- NULL        # NULL keeps the source cohort size
```

## Configuration 2 --- what the columns mean

Declare only the roles your dataset has. Every column **not** named by a role is
dropped -- the run prints which -- so a stray identifier cannot leak out of a
real subject by being forgotten.

```{r}
#| label: roles
#| code-fold: false
ROLES <- pmx_roles(
  id       = "ID",
  time     = "TIME",
  dv       = "DV",
  amt      = "AMT",
  evid     = "EVID",
  dvid     = "DVID",   # one endpoint-key column, or several that agree:
                       #   dvid = c("YTYPE", "NAME") declares a numeric key and
                       #   a character label; both are checked and carried out.
  cmt      = "CMT",
  mdv      = "MDV",

  # DECLARE NOMINAL TIME IF THE STUDY HAS IT. `coarsen_time = TRUE` (the
  # default) collapses subjects onto a shared visit grid so no avatar carries one
  # real subject's exact visit schedule. With `nominal_time` declared it snaps to
  # the protocol grid exactly; without it the grid is inferred from the pooled
  # times, which is best-effort and alerts loudly when it cannot collapse a
  # subject. Under ACTUAL recorded times this is the difference between the
  # mechanism working and not working -- see `skeleton_uniqueness()`.
  # Gaps are fine: rows with no nominal time fall back to the inferred grid,
  # and the run reports the grid as "mixed". A column that is entirely missing
  # should be left undeclared instead.
  nominal_time = NULL, # e.g. "NTIME"
  tad      = NULL,     # time-after-dose column; recomputed after generation
  # ASSIGNED STRATUM -- treatment arm, dose group, cohort. Carried verbatim like
  # `keep`, and additionally used to group the pool of visit sets an
  # avatar may draw from, so an avatar only inherits a visit pattern from its own
  # arm. Not a blending barrier: donors are still borrowed across strata to reach
  # the donor floor, because only route of administration is absolute.
  #
  # Whether to declare an arm here is a real trade, not an obvious win. Declaring
  # it is more faithful -- arms with different visit schedules stay separate --
  # but it splits the visit-set pools, so fewer visit sets clear `min_pattern_share`
  # and more real patterns of missing visits are discarded. The run reports how many
  # lost either way; try it both ways and compare.
  # Gaps are fine here too: subjects with no recorded value form their own
  # stratum and the run warns. A column that *varies* within a subject is an
  # error -- it is not that subject's assignment. A column that does not exist
  # in this study is also an error, naming itself: the two PIT565 parts do not
  # carry identical columns, so check this block against the study you are
  # running rather than assuming it carries over.
  subject_properties = c("TRT", "TRTN"),  # set NULL to compare pooled pattern pools
  rate     = NULL,     # infusion RATE column, if any
  occasion = NULL,     # occasion column, if TIME resets by occasion
  cens     = NULL,     # below-limit indicator (Monolix: 1 below, -1 above), if
  limit    = NULL,     #   the study has BLQ data; limit is the interval's far end

  # BLENDED into genuinely new values across neighbours -- use for baselines you
  # want synthesized (weight, age). Do NOT list a baseline that is itself an
  # endpoint's starting value (e.g. baseline B cells beside a B-cell endpoint):
  # covariates and endpoints are blended independently, so the two would not
  # agree (see the AVATAR mathematics article, and REV-022).
  covariates = c("WT", "AGE", "SEX"),
  # DECLARE THIS IF THE STUDY DOSES ON WEIGHT OR BSA. Names the covariate the
  # amount is a fixed multiple of, so each avatar's dose is rebuilt from its own
  # blended value. Left NULL the run tries to infer it, and infers conservatively
  # -- vial rounding or patients escalating by different factors defeat the test,
  # and then amounts are copied verbatim and still encode one real patient's
  # weight. Must also appear in `covariates` above. The run report says which
  # path was taken.
  dose_covariate = NULL,   # e.g. "WEIGHT0"

  # COPIED verbatim from the subject that supplied the doses, so it stays
  # coherent with them -- a study id, a character endpoint label. A kept value is
  # one real subject's real value. An assigned *stratum* (treatment arm, dose
  # group) belongs in `subject_properties` above instead: carried the same way,
  # and used to group the attendance-pattern pools.
  keep     = NULL
)
```

## Read and validate the source

```{r}
#| label: read-validate
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

if (!file.exists(DATA_PATH)) {
  stop("DATA_PATH does not exist: ",
       normalizePath(DATA_PATH, mustWork = FALSE),
       "\nEdit DATA_PATH in the configuration chunk.", call. = FALSE)
}
raw   <- read.csv(DATA_PATH, stringsAsFactors = FALSE)
roles <- ROLES

report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
cat("Source is valid:", length(unique(raw[[roles$id]])), "subjects.\n")
```

## How identifying is this study's visit schedule?

`synpmx_avatar()` copies its anchor's event skeleton **verbatim**, so a patient
holding the only copy of their observation times hands that schedule to every
avatar anchored on them. Run it twice: on the times as given, and on the grid
the generator actually builds. The second is the one that matters, and the
difference between them is what coarsening bought.

```{r}
#| label: skeleton-uniqueness-before
exposure_before <- skeleton_uniqueness(raw, roles)
exposure_before
```

```{r}
#| label: skeleton-uniqueness-after
exposure <- skeleton_uniqueness(raw, roles, coarsen_time = TRUE)
exposure
```

Split the headline by cause before deciding it is bad. **A one-off observation
time** is a real problem and a fixable one -- declare `nominal_time`. **A
unique set of attended visits** means visits are missing rather than moved --
a missed visit, a discontinuation, or follow-up that has not reached them yet.
No grid at any resolution touches that, and `min_pattern_share` handles it
downstream. A large count there with a small count above is the normal result
for a study with any of those, not a failure.

The picture answers the same question faster. Vertical stripes mean patients
share visits, a ragged right edge is follow-up ending, and a red bar in the
lower panel
is a moment only one patient was ever sampled at -- exactly what would be
copied onto an avatar.

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map
plot_pmx_schedule(raw, roles)
```

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map-raw
plot_pmx_schedule(raw, roles, coarsen_time = FALSE)
```

Record these in `scripts_private/README.md`'s study inventory.

## Synthesize the data

```{r}
#| label: synthesize
#| code-fold: false
synthetic <- synpmx_avatar(raw, roles, n_subjects = N_SUBJECTS, seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)

fresh_ids <- length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0L
cat("Generated", nrow(synthetic), "rows for",
    length(unique(synthetic[[roles$id]])), "subjects.\n")
cat("Every synthetic ID is new:", fresh_ids, "\n")

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"),
          row.names = FALSE)
```

## What the mechanisms actually did

Each row carries the sentence that says what its number means. Copy this into
the study inventory in `scripts_private/README.md`.

```{r}
#| label: mechanism-report
pmx_masking_report(synthetic, raw, roles)
```

Three rows to stop on.

**Amounts recomputed from a covariate** answers "did it find the weight-based
dosing?" outright, and when the answer is no it names which covariates were
tried and what failed. Detection is deliberately conservative -- it refuses
rather than rewrite amounts in a study that is not dose-proportional. A "no" on
a study that *is* weight-based means `AMT` is copied verbatim from the anchor,
so each avatar's implied mg/kg is wrong and the amount still encodes one real
patient's weight.

**Shared by too few patients, so not reused** is the fidelity cost: real
patterns of missing visits and interruptions that will not appear in the
output. Refusing
them is what stops an avatar carrying a schedule traceable to one person.
`min_pattern_share = 2` is the lowest value that keeps that guarantee; `1` gives
it up. At 2 this row necessarily equals the one below it, since a set is
discarded exactly when one patient holds it.

**Avatars keeping their anchor's own visit set** undoes the row above when it is
high: those avatars carry one real patient's absences. The run alerts past 10%,
and the fix is `nominal_time` or a lower `min_pattern_share`.

## Restricted diagnostic: distribution summary

Source-derived. Keep it in the safe environment; do not export these tables.

Numbers before pictures: `compare_pmx_distributions()` summarizes the dependent
variable per endpoint and each baseline covariate, source against synthetic.
Read it as a range check -- expect the summaries to line up in magnitude and
shape, not to match exactly, since AVATAR does not reproduce the source
distribution.

```{r}
#| label: distribution-summary
compare_pmx_distributions(raw, synthetic, roles)
```

## Restricted diagnostic: how close did the values land?

Everything above concerns structure. Blending protects the **values**, and this
measures it: is a synthetic patient closer to a real patient than real patients
are to each other?

Compare two numbers rather than interpreting one. **Measured** is the run;
`0.5` is the target, not the maximum -- it means a synthetic patient is no more
like a real patient than one real patient is like another. Toward `0` is
memorisation, the failure; toward `1` the sets have separated, which costs
utility. **Expected** is what that statistic comes out at when nothing is
wrong, and it is not assumed: the same calculation runs repeatedly on two
halves of the real cohort, each half the size of the synthetic set being tested
(that is the "patients per half"), so small-sample artefacts cancel.

At a few dozen patients the expected range is wide. Inside it means **nothing
was detected**, never that nothing is there.

```{r}
#| label: proximity
compare_pmx_proximity(raw, synthetic, roles)
```

## Post-generation screen (optional, not run for you)

`flag_identifiable_subjects()` scores each **synthetic** patient on four axes --
follow-up time, dose count, dose magnitude, peak DV -- and flags robust
outliers. It is a **plausibility check rather than a masking step**: the anchor
screen has already kept structurally extreme *source* patients from anchoring an
avatar (the "excluded as structurally extreme" row of the report above), and a
conspicuous *blended* value identifies nobody, because it belongs to no one.

Read what it flags before acting on it. Its modified z-score reacts to the shape
of the distribution as well as to genuine outliers: on a two-humped cohort such
as `mavoglurant`, where half the patients have one dosing period and half have
two, it flags 17 of 120 avatars on the same axes the anchor screen had already
handled. If this study has two clearly different groups, expect some of that and
do not simply drop what it returns.

```{r}
#| label: flag-identifiable
flagged <- flag_identifiable_subjects(synthetic, roles)
sum(flagged$flagged)
knitr::kable(utils::head(flagged[flagged$flagged, ], 10))
```

`remediate_identifiable_subjects()` will truncate, drop, or replace them and
refill the cohort. Note `REV-028`: replacements are screened against their own
batch rather than the merged cohort, so a replacement can be clean in its batch
and an outlier once joined.

## Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r}
#| label: diagnostic
#| fig-width: 9
#| fig-height: 6
# The plot adapts to however many endpoints there are -- one PK, one PD, two,
# more -- because faceting by endpoint grows a column per endpoint on its own.
# You do not have to know the count in advance.
#
# Facet strips read best from a character label. If you declared several `dvid`
# columns, the last one (often a character NAME) is used; otherwise the single
# dvid column; otherwise a constant.
label_col <- if (length(roles$dvid) > 1L) {
  roles$dvid[[length(roles$dvid)]]
} else if (!is.null(roles$dvid)) roles$dvid[[1L]] else NULL

frame <- function(d, label) {
  keep <- d[[roles$evid]] == 0 & !is.na(d[[roles$dv]])   # observations only
  data.frame(
    dataset  = label,
    id       = d[[roles$id]][keep],
    time     = as.numeric(d[[roles$time]][keep]),
    dv       = as.numeric(d[[roles$dv]][keep]),
    endpoint = if (is.null(label_col)) "DV" else
      as.character(d[[label_col]][keep]),
    stringsAsFactors = FALSE
  )
}
compare <- rbind(frame(raw, "source"), frame(synthetic, "synthetic"))

ggplot(compare, aes(time, dv, group = id)) +
  geom_line(alpha = 0.3) +
  geom_point(alpha = 0.4, size = 0.6) +
  # rows: source vs synthetic. columns: one per endpoint, however many.
  facet_grid(dataset ~ endpoint, scales = "free_y") +
  labs(title = "RESTRICTED: source vs AVATAR synthetic",
       subtitle = "do not export", x = "Time", y = "DV") +
  theme_bw()
# Concentrations often read better on a log axis. Add `+ scale_y_log10()` above
# if every DV is positive; a log scale drops zeros and negatives.
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories, so it stays under the source data's own access
controls and obligations wherever it is used — your own workstation included, a
partner or vendor not.
