---
title: "AVATAR synthesis on CPIT565A1 modeling data"
format:
  html:
    toc: true
    # Everything except the configuration and the generation call is reporting
    # code, and reading it is not the point of this document. Folded, not
    # hidden: click "Show code" on any block to see exactly what produced a
    # number. The chunks worth reading unfolded set `code-fold: false`.
    code-fold: true
    code-summary: "Show code"
---

Try the AVATAR blending generator on a real modeling dataset.

This is the **default** method. Use it when the synthetic data *reaches no one
the source data could not* -- the same access controls and confidentiality
obligations still apply, which includes carrying it to your own workstation for
model development. It resamples and blends whole source subject
trajectories -- more faithful and far less ceremony than the differentially
private path -- but makes **no** formal privacy guarantee. If the data may reach
someone new, use `try_dp_calibrated.R` instead. See the `synpmx-privacy`
vignette for the decision rule (which method when).

This document runs in the safe computing environment, in a gitignored folder so
nothing here can be committed.

Workflow:

1. fill in the CONFIG chunk (just the column roles)
2. compare the synthetic table to the source

```{r}
#| label: setup
#| code-fold: false
#| message: false
library(tidyverse)
library(xgxr)
xgx_theme_set()

# If not installed here: devtools::load_all("/path/to/synpmx")
# This load_all line only needs to be run once, so you may comment out if you're 
# Trying this out on your own dataset

devtools::load_all()
library(synpmx)

```

## Configuration

Everything you edit is in this chunk.

```{r, fig.width = 10}
#| label: config
#| code-fold: false
user <- Sys.getenv("USER")
gps_prefix <- paste0("/view/", user, "_view/vob/")
filename   <- file.path(gps_prefix, "CPIT565A1/mas/mas_1/data/output_001/", "mt96714_poppkpd.csv")
data_in    <- read.csv(filename, stringsAsFactors = FALSE)

DATA       <- data_in |> filter(NAME %in% c("PIT565 Dose", "PIT565 PK", "CRP"), STUDYID == "CPIT565A12101")
OUT_DIR   <- "output_pit565a"
SEED      <- 1234        # reproducibility seed; the caller's RNG is untouched

# plot initial data
ggplot(DATA |> filter(NAME != "PIT565 Dose"), aes(x = TIME, y = LIDV, group = ID, color = TRT)) + 
  geom_point(alpha = 0.5) + 
  geom_line(alpha = 0.5) + 
  facet_wrap(~NAME) + 
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10()
```

Map your columns onto PMX roles. Only the roles the real dataset has. AVATAR
needs no model, priors, design, or budget -- just the column meanings. List
baseline covariates in `covariates` and they are blended from compatible donors
like everything else.

```{r}
#| label: roles
#| code-fold: false
ROLES <- pmx_roles(
  id       = "USUBJID",
  time     = "TIME",
  dv       = "LIDV",
  amt      = "AMT",
  evid     = "EVID",
  # Two columns encode the same endpoint: numeric YTYPE and character NAME.
  # Declare both. YTYPE (first) is the grouping key; validation checks NAME is a
  # consistent 1:1 mapping with it and errors if not, and both are carried out.
  dvid     = c("YTYPE", "NAME"),
  cmt      = "CMT",
  mdv      = "MDV",

  # DECLARE NOMINAL TIME IF THE STUDY HAS IT. `coarsen_time = TRUE` (the
  # default) collapses subjects onto a shared visit grid so no avatar carries one
  # real subject's exact visit schedule. With `nominal_time` declared it snaps to
  # the protocol grid exactly; without it the grid is inferred from the pooled
  # times, which is best-effort and alerts loudly when it cannot collapse a
  # subject. Under ACTUAL recorded times this is the difference between the
  # mechanism working and not working -- see `skeleton_uniqueness()`.
  # Gaps are fine: rows with no nominal time fall back to the inferred grid,
  # and the run reports the grid as "mixed". A column that is entirely missing
  # should be left undeclared instead.
  nominal_time = NULL, # e.g. "NTIME"
  tad      = NULL,     # time-after-dose column; recomputed after generation
  # ASSIGNED STRATUM -- treatment arm, dose group, cohort. Carried verbatim like
  # `keep`, and additionally used to group the pool of visit sets an
  # avatar may draw from, so an avatar only inherits a visit pattern from its own
  # arm. Not a blending barrier: donors are still borrowed across strata to reach
  # the donor floor, because only route of administration is absolute.
  #
  # Whether to declare an arm here is a real trade, not an obvious win. Declaring
  # it is more faithful -- arms with different visit schedules stay separate --
  # but it splits the visit-set pools, so fewer visit sets clear `min_pattern_share`
  # and more real dropout patterns are discarded. The run reports how many were
  # lost either way; try it both ways and compare.
  # Gaps are fine here too: subjects with no recorded value form their own
  # stratum and the run warns. A column that *varies* within a subject is an
  # error -- it is not that subject's assignment. A column that does not exist
  # in this study is also an error, naming itself: the two PIT565 parts do not
  # carry identical columns, so check this block against the study you are
  # running rather than assuming it carries over.
  subject_properties = c("TRT", "TRTN"),  # set NULL to compare pooled pattern pools
  rate     = NULL,           # set if you have infusion RATE rows
  occasion = NULL,           # set if TIME resets by occasion
  # If the data has below-limit-of-quantification (BLQ) records, declare cens
  # (and limit, if intervals are used). AVATAR then reconstructs censoring
  # coherently -- DV at the limit, CENS/LIMIT agreeing -- instead of copying a
  # flag that no longer matches the blended DV. Leave NULL if there is no BLQ.
  cens     = "CENS",           # e.g. "CENS"  (Monolix: 1 = below, -1 = above)
  limit    = NULL,           # e.g. "LIMIT" (other end of an interval)
  # BLENDED into new values across donors. B0 (baseline B cells) is deliberately
  # left out: it is the same quantity as the Circ B Cells endpoint, and AVATAR
  # blends covariates and endpoints independently, so a synthetic B0 would not
  # match its own generated trajectory (REV-022). Read the baseline off the
  # trajectory instead if you need it.
  covariates = c("WEIGHT0", "AGE0", "SEX", "RACE"),
  # keep = copied verbatim from the subject that supplied the doses, so it stays
  # coherent with them. Use it for pass-through values with no other meaning --
  # a study id, a character label. An assigned *stratum* (treatment arm, dose
  # group) is better declared in `subject_properties` above, which carries it
  # the same way and also groups the attendance-pattern pools.
  keep     = c("STUDYIDN", "STUDYID")
  # Every column NOT named by a role above is dropped. That is deliberate: a
  # secondary identifier you forget cannot leak out of a real subject. The run
  # prints exactly which columns it dropped.
)

# Number of synthetic subjects. NULL keeps the source cohort size.
N_SUBJECTS <- NULL
```

## Workflow

You should not need to edit below here.

```{r}
#| label: read-source
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
raw   <- DATA
roles <- ROLES
```

### Source structural validation

```{r}
#| label: validate-source
report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
message("  valid; ", length(unique(raw[[roles$id]])), " subjects")
```

### How identifying is this study's visit schedule?

Run before generating. `synpmx_avatar()` copies its anchor's event skeleton
**verbatim**, so a patient holding the only copy of their observation time
vector hands that schedule to every avatar anchored on them. This section asks
how many such patients there are, and whether that is a problem worth acting
on.

Two runs, because the answer differs:

- **as given** -- the recorded times, before anything is done to them.
- **after coarsening** -- the shared visit grid `synpmx_avatar()` actually
  builds and generates on. *These are the numbers that matter.* The difference
  between the two is what coarsening bought.

Every column is the same kind of number: **how many patients share this
property with me, me included**, so `1` always means "nobody else". Read the
first table's headline row, then its two sub-rows, which have opposite
remedies.

```{r}
#| label: skeleton-uniqueness-before
exposure_before <- skeleton_uniqueness(raw, roles)
exposure_before
```

```{r}
#| label: skeleton-uniqueness-after
exposure <- skeleton_uniqueness(raw, roles, coarsen_time = TRUE)
exposure
```

**Is a high percentage bad?** Not on its own, and this is the part that is easy
to misread. Split it by cause:

- **a one-off observation time** -- this patient was sampled at a moment
  nobody else was. It *is* a problem, and it is fixable: declaring
  `nominal_time` snaps visits onto the real protocol grid. A count above zero
  here means the inferred grid could not find a shared schedule.
- **the set of visits attended** -- every one of this patient's times is shared
  with somebody; what singles them out is which of those visits they turned up
  for. This is dropout, and **no grid at any resolution can change it**. It is
  handled downstream by `min_pattern_share`, which refuses to give any avatar a
  visit set that only one real patient had. A large count here with a small
  count above is the *normal* result for a study with dropout, not a failure.

The picture says in one look what the counts say in numbers -- whether this is
a protocol grid with a few stragglers or an ad-hoc schedule the grid could not
rescue.

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map
plot_pmx_schedule(raw, roles)
```

Read the map top to bottom: vertical stripes mean patients share visits, and a
ragged right edge is dropout. In the lower panel, a bar of height one is a
moment only one patient was ever sampled at -- exactly what would be copied
onto an avatar -- and those are drawn in red. Compare against the uncoarsened
version to see what the grid merged.

```{r, fig.width = 12, fig.height = 9}
#| label: schedule-map-raw
plot_pmx_schedule(raw, roles, coarsen_time = FALSE)
```

Record these in `scripts_private/README.md`'s study inventory.

### Synthesize (AVATAR blending)

```{r}
#| label: synthesize
#| code-fold: false
synthetic <- synpmx_avatar(raw, roles, seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)
message("  generated ", nrow(synthetic), " rows for ",
        length(unique(synthetic[[roles$id]])), " subjects; new identifiers: ",
        length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0)

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"), row.names = FALSE)
```

### What the mechanisms actually did

Every masking mechanism records what it did, and several record what it cost.
Each row carries the sentence that says what its number means, because none of
them mean anything on their own. Copy this into the study inventory in
`scripts_private/README.md`.

```{r}
#| label: mechanism-report
pmx_masking_report(synthetic, raw, roles)
```

Four rows are worth stopping on.

**Did it detect weight-based dosing?** The **Dose** section says so outright:
either "**yes**, from `WEIGHT0`" with the protocol levels it found, or
"**no**" together with the reason it declined. If this study doses on weight
and the answer is no, the reason names which covariates were tried and what
failed -- most often that the dose-to-covariate ratio did not collapse onto a
small number of protocol levels, which is the conservative test the detector
uses to avoid rewriting amounts in a study that is not dose-proportional. When
the answer is no, `AMT` is copied verbatim from the anchor, which means each
avatar's implied mg/kg is wrong and the amount still encodes one real patient's
weight.

**Shared by too few patients, so not reused.** These are real dropout and
dose-interruption patterns that will not appear in the synthetic data. Refusing
them is what stops an avatar carrying a schedule traceable to one person. If
this study's interruptions matter, lower `min_pattern_share`; `2` is the lowest
value that still guarantees no synthetic patient has a visit set unique to a
real one, and `1` keeps every pattern and gives up the guarantee.

At the default floor of 2, that row and **real patients holding those** are
necessarily equal: a set is discarded exactly when fewer than two patients
share it, so every discarded set has exactly one holder. They diverge only at a
floor of 3 or more.

**Avatars keeping their anchor's own visit set.** Read this against the row
above it. A high percentage means the schedule groups had nothing shareable to
draw from, so those avatars carry one real patient's pattern of absences --
which undoes what discarding the rare sets was for. If it is high, the run also
raises the "no shared visit set to draw from" alert, and the fix is the same:
declare `nominal_time`, or lower `min_pattern_share`.

### Restricted diagnostic: distribution summary

Source-derived. Keep it in the safe environment; do not export these tables.

`compare_pmx_distributions()` summarizes the dependent variable per endpoint and
each baseline covariate, source against synthetic -- a quick numeric range check
before the plot. Expect the summaries to line up in magnitude and shape, not to
match exactly, since AVATAR does not reproduce the source distribution.

```{r}
#| label: distribution-summary
#| results: asis
# Rendered with knitr::kable() so the tables come out as tables when this
# document is knitted, and stay readable when run chunk-by-chunk. Rounding is
# for display only; the returned object keeps the exact values.
kable_distributions <- function(x) {
  section <- function(df, caption) {
    if (is.null(df)) return(invisible())
    numeric_cols <- vapply(df, is.numeric, logical(1))
    df[numeric_cols] <- lapply(df[numeric_cols], signif, digits = 4)
    print(knitr::kable(df, caption = caption))
    cat("\n")
  }
  section(x$endpoints,
          "RESTRICTED -- endpoints (dependent variable on observation rows)")
  section(x$covariates_numeric,
          "RESTRICTED -- continuous covariates (baseline, per subject)")
  section(x$covariates_categorical,
          "RESTRICTED -- categorical covariates (baseline, per subject)")
  invisible(x)
}

kable_distributions(compare_pmx_distributions(raw, synthetic, roles))
```

### Restricted diagnostic: how close did the values land?

Everything above concerns structure. Blending is what protects the **values**,
and this is its measurement. It answers one question: **is a synthetic patient
closer to a real patient than real patients are to each other?**

The way to read it is to compare two numbers, not to interpret one:

- **Measured** -- the statistic on this run. `0.5` is the target, not the
  maximum: it means a synthetic patient is no more like a real patient than one
  real patient is like another. Toward `0` means memorisation, which is the
  failure. Toward `1` means the two sets have separated, which costs utility
  rather than privacy.
- **Expected** -- what that statistic comes out at when nothing is wrong. It is
  not assumed. The same calculation is run repeatedly on two halves of the real
  cohort, so every small-sample artefact is present in the null and in the
  measurement alike and cancels. Each half has the same number of patients as
  the synthetic set being tested, which is what the *"patients per half"*
  figure is: the size of one side of the comparison, matched on all three arms
  so the counts are comparable.

At a few dozen patients that expected range is wide. Inside it means **nothing
was detected**, never that nothing is there.

```{r}
#| label: proximity
compare_pmx_proximity(raw, synthetic, roles)
```

### Post-generation screen (optional, not run for you)

`flag_identifiable_subjects()` scores each **synthetic** patient on four axes --
follow-up time, dose count, dose magnitude, peak DV -- and flags robust
outliers. It is a **plausibility check rather than a masking step**: M2 has
already kept structurally extreme *source* patients out of the anchor pool, and
a conspicuous blended value identifies nobody because it belongs to no one.

Read what it flags before acting on it. Its median/MAD statistic behaves badly
on a bimodal cohort -- on `mavoglurant`, where half the patients have one dosing
period and half have two, it flags 75 of 120 avatars on axes M2 already
screened. If this study has two clearly different groups, expect the same and do
not simply drop what it returns.

```{r}
#| label: flag-identifiable
flagged <- flag_identifiable_subjects(synthetic, roles)
sum(flagged$flagged)
knitr::kable(utils::head(flagged[flagged$flagged, ], 10))
```

`remediate_identifiable_subjects()` will truncate, drop, or replace them and
refill the cohort. Note `REV-028`: replacements are screened against their own
batch rather than the merged cohort, so a replacement can be clean in its batch
and an outlier once joined.

### Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r}
#| label: diagnostic

  raw_keep = raw[,names(synthetic)]

  compare = rbind(raw_keep |> mutate(DATA = "source"), 
                  synthetic |> mutate(DATA = "synthetic"))
  
  g = ggplot(compare |> filter(NAME != "PIT565 Dose"), aes(x = TIME, y = LIDV, group = USUBJID, color = TRT)) + 
    geom_point(alpha = 0.5) + 
    geom_line(alpha = 0.5) + 
    facet_wrap(DATA~NAME) + 
    xgx_scale_x_time_units("days") +
    xgx_scale_y_log10()

  g
  
  #ggsave(file.path(OUT_DIR, "RESTRICTED_source_vs_synthetic.pdf"), g, 
  #       width = 7, height = 6, dpi = 110)
  
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories, so it stays under the source data's own access
controls and obligations wherever it is used — your own workstation included, a
partner or vendor not.
