---
title: "AVATAR synthesis on CPIT565A1 modeling data"
format: pdf
---

Try the AVATAR blending generator on a real modeling dataset.

This is the **default** method. Use it when the synthetic data *stays inside*
your trusted computing environment (you are the only consumer, governance and
access controls apply). It resamples and blends whole source subject
trajectories -- more faithful and far less ceremony than the differentially
private path -- but makes **no** formal privacy guarantee. If the data may cross
a trust boundary, use `try_dp_calibrated.R` instead. See the `synpmx-privacy`
vignette for the decision rule (which method when).

This document runs in the safe computing environment, in a gitignored folder so
nothing here can be committed.

Workflow:

1. fill in the CONFIG chunk (just the column roles)
2. compare the synthetic table to the source

```{r}
#| label: setup
#| message: false
library(tidyverse)
library(xgxr)
xgx_theme_set()

# If not installed here: devtools::load_all("/path/to/synpmx")
# This load_all line only needs to be run once, so you may comment out if you're 
# Trying this out on your own dataset

devtools::load_all()
library(synpmx)

```

## Configuration

Everything you edit is in this chunk.

```{r}
#| label: config
user <- Sys.getenv("USER")
gps_prefix <- paste0("/view/", user, "_view/vob/")
filename   <- file.path(gps_prefix, "CPIT565A1/mas/mas_1/data/output_001/", "mt96714_poppkpd.csv")
data_in    <- read.csv(filename, stringsAsFactors = FALSE)

DATA       <- data_in |> filter(NAME %in% c("PIT565 Dose", "PIT565 PK", "Circ B Cells"), STUDYID == "CPIT565B12101")
OUT_DIR   <- "output_pit565"
SEED      <- 1234        # reproducibility seed; the caller's RNG is untouched

# plot initial data
ggplot(DATA |> filter(NAME != "PIT565 Dose"), aes(x = TIME, y = LIDV, group = ID, color = TRT)) + 
  geom_point(alpha = 0.5) + 
  geom_line(alpha = 0.5) + 
  facet_wrap(~NAME) + 
  xgx_scale_x_time_units("days") +
  xgx_scale_y_log10()
```

Map your columns onto PMX roles. Only the roles the real dataset has. AVATAR
needs no model, priors, design, or budget -- just the column meanings. List
baseline covariates in `covariates` and they are blended from compatible donors
like everything else.

```{r}
#| label: roles
ROLES <- pmx_roles(
  id       = "USUBJID",
  time     = "TIME",
  dv       = "LIDV",
  amt      = "AMT",
  evid     = "EVID",
  # Two columns encode the same endpoint: numeric YTYPE and character NAME.
  # Declare both. YTYPE (first) is the grouping key; validation checks NAME is a
  # consistent 1:1 mapping with it and errors if not, and both are carried out.
  dvid     = c("YTYPE", "NAME"),
  cmt      = "CMT",
  mdv      = "MDV",
  rate     = NULL,           # set if you have infusion RATE rows
  occasion = NULL,           # set if TIME resets by occasion
  # If the data has below-limit-of-quantification (BLQ) records, declare cens
  # (and limit, if intervals are used). AVATAR then reconstructs censoring
  # coherently -- DV at the limit, CENS/LIMIT agreeing -- instead of copying a
  # flag that no longer matches the blended DV. Leave NULL if there is no BLQ.
  cens     = "CENS",           # e.g. "CENS"  (Monolix: 1 = below, -1 = above)
  limit    = NULL,           # e.g. "LIMIT" (other end of an interval)
  # BLENDED into new values across donors. B0 (baseline B cells) is deliberately
  # left out: it is the same quantity as the Circ B Cells endpoint, and AVATAR
  # blends covariates and endpoints independently, so a synthetic B0 would not
  # match its own generated trajectory (REV-022). Read the baseline off the
  # trajectory instead if you need it.
  covariates = c("WEIGHT0", "AGE0", "SEX", "RACE"),
  # keep = copied verbatim from the subject that supplied the doses, so it stays
  # coherent with them. Put assigned, subject-defining values here -- treatment
  # arm, dose group -- e.g. keep = "ARM". List only columns your data has.
  keep     = c("STUDYIDN", "STUDYID", "TRT", "TRTN")
  # Every column NOT named by a role above is dropped. That is deliberate: a
  # secondary identifier you forget cannot leak out of a real subject. The run
  # prints exactly which columns it dropped.
)

# Number of synthetic subjects. NULL keeps the source cohort size.
N_SUBJECTS <- NULL
```

## Workflow

You should not need to edit below here.

```{r}
#| label: read-source
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
raw   <- DATA
roles <- ROLES
```

### Source structural validation

```{r}
#| label: validate-source
report <- validate_pmx(raw, roles, strict = FALSE)
if (!report$valid) {
  print(report)   # names each problem, the role, and the column it maps to
  stop("Source failed structural validation. See the numbered problems above; ",
       "each says which role and column to fix in `pmx_roles()`.", call. = FALSE)
}
message("  valid; ", length(unique(raw[[roles$id]])), " subjects")
```

### Synthesize (AVATAR blending)

```{r}
#| label: synthesize
synthetic <- synpmx_avatar(raw, roles, seed = SEED)
stopifnot(validate_pmx(synthetic, roles)$valid)
message("  generated ", nrow(synthetic), " rows for ",
        length(unique(synthetic[[roles$id]])), " subjects; new identifiers: ",
        length(intersect(synthetic[[roles$id]], raw[[roles$id]])) == 0)

write.csv(synthetic, file.path(OUT_DIR, "avatar_synthetic.csv"), row.names = FALSE)
```

### Restricted diagnostic: source vs synthetic

Source-derived. Keep it in the safe environment; do not export the figure.

```{r}
#| label: diagnostic

  raw_keep = raw[,names(synthetic)]

  compare = rbind(raw_keep |> mutate(DATA = "source"), 
                  synthetic |> mutate(DATA = "synthetic"))
  
  g = ggplot(compare |> filter(NAME != "PIT565 Dose"), aes(x = TIME, y = LIDV, group = USUBJID, color = TRT)) + 
    geom_point(alpha = 0.5) + 
    geom_line(alpha = 0.5) + 
    facet_wrap(DATA~NAME) + 
    xgx_scale_x_time_units("days") +
    xgx_scale_y_log10()

  g
  
  #ggsave(file.path(OUT_DIR, "RESTRICTED_source_vs_synthetic.pdf"), g, 
  #       width = 7, height = 6, dpi = 110)
  
```

The synthetic table carries no formal privacy guarantee; it is built by blending
real subject trajectories and is for trusted-environment use only.
